### 2021-09-16 presentation/glmeiv_talk.tex
update pres

\documentclass{beamer}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{/Users/timbarry/optionFiles/mymacros}
%Information to be included in the title page:
\title{A generalized errors-in-variables model, with application to single-cell CRISPR screens}
\author{Tim Barry$^1$, Eugene Katsevich$^2$, Kathyrn Roeder$^1$}
\institute{$^1$CMU Statistics and Data Science\\ $^2$Wharton Statistics and Data Science}
\date{September 2021}

\begin{document}

\frame{\titlepage}

\begin{frame}
\frametitle{Overview}
\begin{itemize}
\item \textbf{Background}
\item Analysis objective and challenges
\item Existing approach
\item Proposed method
\item Simulation results
\item Real data results
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Single-cell CRISPR screens are a powerful technology for mapping the regulatory wiring of the genome.}

\begin{figure}
	\centering
	\includegraphics[width=1\linewidth]{extra_figs/regulatory_wiring}
	\label{regulatorywiring}
\end{figure}
\end{frame}

\begin{frame}
	\frametitle{Single-cell CRISPR screens entail sequencing gRNAs and mRNAs in individual cells.}

	\begin{figure}
		\centering
		\includegraphics[width=1\linewidth]{extra_figs/experimental_protocol}
	\end{figure}
\end{frame}

\begin{frame}
\frametitle{Overview}
\begin{itemize}
	\item Background
	\item\textbf{Analysis Challenges}
	\item Existing approach
	\item Proposed method
	\item Simulation results
	\item Real data results
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{There are several challenges to the analysis of single-cell CRISPR screen data:}

\begin{itemize}
\item[1.] The perturbation is unobserved.
\item[2.] Technical factors, such as batch and sequencing depth, explain variability in mRNA and gRNA counts.
\item[3.] Unperturbed cells exhibit ``background gRNA reads.'' \cite{Schraivogel2020}
\item[4.] The mRNA and gRNA expression data are highly discrete.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{(1) The perturbation is unobserved, and (2) technical factors are present.}
\begin{figure}
	\centering
	\includegraphics[width=1\linewidth]{../figures/fig1/dag}
	\label{dag}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{(3) Unperturbed cells exhibit background gRNA reads.}
\begin{figure}
	\centering
	\includegraphics[width=0.75\linewidth]{../figures/fig1/density_plot_annotated.png}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{(4) The count data are highly discrete.}
\begin{figure}
	\centering
	\includegraphics[width=0.75\linewidth]{../figures/fig1/mRNA_count_hist_annotated.png}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Overview}
\begin{itemize}
	\item Background
	\item Analysis Challenges
	\item \textbf{Existing approach}
	\item Proposed method
	\item Simulation results
	\item Real data results
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Data and notation}

\begin{itemize}
\item Observe $n \approx 100,000 - 250,000$ cells.
\item Consider a given mRNA and gRNA of interest.
\item For cell $i \in \{1, \dots, n\}$, let 
\begin{itemize}
\item $p_i \in \{0, 1\}$ indicate whether a perturbation occurred. 
\item $m_i \in \mathbb{N}$ be the mRNA count.
\item $g_i \in \mathbb{N}$ be the gRNA count.
\item $l^m_i \in \mathbb{N}$ be the mRNA library size.
\item $z_i \in \R^{d-1}$ be a vector of technical factors, possibly including an intercept term.
\end{itemize}
\item $\approx5,000$ genes, $\approx500 - 5,000$ gRNAs
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{The ``thresholding method''}

\begin{itemize}
\item[1.] For given threshold $c \in \N$, estimate $P_i$ by $$ \begin{cases} \hat{P}_i = 0 \textrm{ if } g_i \geq c, \\ \hat{P}_i = 1 \textrm{ if } g_i < c \end{cases}.$$
\item[2.] Fit the regression model \cite{Sarkar2021}
$$ m_i | \left( z_i, l^m_i \right) \sim \textrm{NB}_\theta(\mu_i),$$ where $\theta >0$ is the NB size parameter and $$\log\left(\mu_i\right) = \beta_m \hat{P}_i + \gamma^T_m z_i + \log\left( l_i^m\right).$$
\item[3.] Obtain an estimate $\hat{\beta}_m$ of $\beta_m$ and compute a CI for $\beta_m$.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Problem 1: There is no clear location at which to draw the threshold.}
\begin{figure}
	\centering
	\includegraphics[width=1\linewidth]{../figures/fig2/histograms}
\end{figure}

\end{frame}

\begin{frame}
\frametitle{Problem 2: Thresholding leads to attenuation bias in many measurement error models \cite{Stefanski2000}.}

\end{frame}

\begin{frame}
\frametitle{Research question}
Does modeling the gRNA count distribution directly (thereby bypassing thresholding) improve estimation and inference?
\end{frame}

\begin{frame}
\frametitle{Overview}
\begin{itemize}
	\item Background
	\item Analysis Challenges
	\item Existing approach
	\item \textbf{Proposed method}
	\item Simulation results
	\item Real data results
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{A bit more notation}
For cell $i \in \{ 1, \dots, n\}$, let $l^g_i$ be the gRNA library size.
\end{frame}

\begin{frame}
\frametitle{We extend the parametric model of \cite{Sarkar2021} to model both mRNA \textit{and} gRNA counts.}
\begin{itemize}
\item[1.] \textbf{mRNA}: $m_i | (z_i, l_i^m) \sim \textrm{NB}_\theta(\mu^m_i),$ where $$ \log\left( \mu^m_i \right) = \beta_m P_i + \gamma_m^T z_i + \log(l^m_i).$$
\item[2.] \textbf{gRNA}: $g_i | (z_i, l_i^g) \sim \textrm{NB}_\theta (\mu_i^g),$ where $$\log(\mu_i^g) = \beta_g P_i + \gamma_g^T z_i + \log(l_i^g)$$
\item[3.] \textbf{Perturbation}: $P_i \sim \textrm{Bern}(\pi)$, where $\pi \in [0, 1/2)$. $P_i$ is latent.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Generalizing the NB model to arbitrary exponential family response distribution and link function yields the ``GLM-EIV'' (GLM errors-in-variables) model.}

This extension is important, because authors have used
 \begin{itemize}
 \item \textcolor{blue}{Negative binomial}, \cite{Choudhary2021}
 \item \textcolor{blue}{Poisson}, \cite{Schraivogel2020}
 \item and \textcolor{blue}{Gaussian} \cite{Lin2021}
 \end{itemize}
distributions to model single-cell data.
\end{frame}

\begin{frame}
\frametitle{Generalizing the NB model to arbitrary exponential family response distribution and link function yields the ``GLM-EIV'' (GLM errors-in-variables) model.}

\begin{itemize}
\item[1.] \textbf{mRNA density}: $$f_m(m_i; \eta^m_i) = \exp\left\{m_i \eta_i^m - \psi_m(\eta_i^m) + c_m(m_i) \right\}.$$
\item[2.] \textbf{gRNA density}: $$f_g(g_i; \eta^g_i) = \exp\left\{ g_i \eta_i^g - \psi_g(\eta_i^g) + c_g(g_i) \right\}.$$
\item[3.] \textbf{Perturbation density}: $$ f(p_i) = \pi^{p_i} (1-\pi)^{1-p_i}.$$
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Generalizing the NB model to arbitrary exponential family response distribution and link function yields the ``GLM-EIV'' (GLM errors-in-variables) model.}

The canonical parameter of the mRNA distribution for the $i$th cell, $\eta^m_i$, is given by $$\eta_i^m = h_m\left( \beta_m P_i + \gamma^T_m z_i + \log\left( l_i^m \right) \right).$$ If the canonical link function is used, then $h_m$ is the identity. The case for $\eta_i^g$ is similar.
\end{frame}

\begin{frame}
\frametitle{We derive an EM algorithm to fit the model.}

\textbf{E step}:
\begin{itemize}
\item Compute membership probabilities $T_1, \dots, T_n$ using the model.
\end{itemize}

\textbf{M step}:
\begin{itemize}
\item Augment count vectors $m \rightarrow [m, m], g \rightarrow [g, g]$.
\item Augment offset vectors $l_m \rightarrow [l_m, l_m], l_g \rightarrow [l_g, l_g]$.
\item Augment covariate matrix  $Z \rightarrow [Z, Z];$ append column of $1$s and $0$s for perturbation indicators.
\item Fit weighted GLM to both modalities using membership probabilities $[T_1, \dots, T_n, 1 - T_1, \dots, 1 - T_n]$ as weights.
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{We derive an analytic expression for the observed information matrix to enable fast inference (CIs, $p$-values).}

\begin{multline*}
J(\hat{\theta}; m, g) = -\E \left[\nabla^2 \mathcal{L}(\theta; m, g, p) | g, m, \hat{\theta} \right] \\ + \E\left[ \nabla \mathcal{L}(\theta; m, g, p) |  g, m, \hat{\theta} \right] \cdot \E\left[ \nabla \mathcal{L}(\theta; m, g, p) | g, m, \hat{\theta}\right]^T \\ - \E\left[ \nabla\mathcal{L}(\theta; m, g, p) \nabla \mathcal{L}(\theta; m, g, p)^T | g, m, \hat{\theta} \right].
\end{multline*}

\begin{figure}
\centering
\includegraphics[width=0.4\linewidth]{extra_figs/fig3_crop}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{We implement several statistical accelerations to make the method fast.}
\end{frame}

% However, the model gives us many insights into single-cell CRISPR screen datasets.
% What is the fold change in gRNA expression?
% What is the perturbation probability?
%

% Practical strategies:
% Is thresholding a reasonable strategy for these data?
% How can I choose a good threshold (cell-specific threshold; use only cells with high or low posterior membership probabilities, and throw all others out)?

\bibliography{/Users/timbarry/optionFiles/glmeiv.bib}
\bibliographystyle{apalike}

\end{document}

### 2021-09-16 R_scripts/plotting/figure2.R

@@ -1,0 +1,56 @@
+# set figure 2 dir
+fig_dir <- paste0(.get_config_path("LOCAL_CODE_DIR"), "glmeiv-manuscript/figures/")
+fig2_dir <- paste0(fig_dir, "fig2/")
+if (!dir.exists(fig2_dir)) dir.create(fig2_dir, recursive = TRUE)
+
+# load plotting objects
+plotting_objects <- paste0(.get_config_path("LOCAL_CODE_DIR"), "glmeiv-manuscript/figures/plotting_objects.R")
+source(plotting_objects)
+
+
+set.seed(2)
+gRNA_ids_gasp <- sample(x = get_feature_ids(gRNA_odm_gasp), size = 2, replace = FALSE)
+gRNA_ids_xie <- sample(x = get_feature_ids(gRNA_odm_xie), size = 2, replace = FALSE)
+
+to_plot <- rbind(data.frame(count = as.numeric(gRNA_odm_gasp[[gRNA_ids_gasp[1],]]), gRNA_id = gRNA_ids_gasp[1], dataset = "Gasp"),
+                 data.frame(count = as.numeric(gRNA_odm_gasp[[gRNA_ids_gasp[2],]]), gRNA_id = gRNA_ids_gasp[2], dataset = "Gasp"),
+                 data.frame(count = as.numeric( gRNA_odm_xie[[gRNA_ids_xie[1],]]), gRNA_id = gRNA_ids_xie[1], dataset = "Xie"),
+                 data.frame(count = as.numeric( gRNA_odm_xie[[gRNA_ids_xie[2],]]), gRNA_id = gRNA_ids_xie[2], dataset = "Xie")) %>% 
+  dplyr::mutate(lab = factor(x = gRNA_id,
+                             levels = c(gRNA_ids_gasp, gRNA_ids_xie),
+                             labels = c("Gasperini gRNA 1", "Gasperini gRNA 2", "Xie gRNA 1", "Xie gRNA 2"))) %>% dplyr::filter(count >= 1)
+
+# compute the original thresholds
+thresh_df <- to_plot %>% dplyr::group_by(lab) %>% dplyr::group_modify(.f = function(tbl, key) {
+  dataset <- tbl$dataset[1]
+  if (dataset == "Gasp") {
+    thresh <- 5
+  } else {
+    curr_count <- tbl$count
+    curr_count <- curr_count[curr_count >= 1]
+    thresh <- mean(curr_count)
+  }
+  return(tibble::tibble(thresh = thresh))
+})
+
[…]

### 2021-09-16 bash_scripts/set_gasp_args.sh
update pres

@@ -51,3 +51,4 @@ fi
 
 # vii. results directory
 result_dir=$LOCAL_GLMEIV_DATA_DIR"public/gasperini/results"
+result_file_name="glmeiv_result.rds"

### 2021-09-16 bash_scripts/set_xie_args.sh

@@ -33,8 +33,7 @@ m_offsets=$processed_data_dir"m_offset.rds"
 g_offsets=$processed_data_dir"g_offset.rds"
 # v. family strings
 m_fam="nb"
-g_fam="nb"
-g_theta=10
+g_fam="poisson"
 # vi. pairs to analyze and pod sizes
 if [ $trial = true ]
 then
@@ -51,4 +50,4 @@ fi
 
 # vii. results directory
 result_dir=$LOCAL_GLMEIV_DATA_DIR"public/xie/results"
-result_file_name="glmeiv_result_nb10.rds"
+result_file_name="glmeiv_result.rds"

### 2021-09-16 presentation/glmeiv_talk.tex
update pres

@@
 \date{September 2021}
+\DeclareMathOperator*{\argmin}{arg\,min}
 
@@
 \end{itemize}
-\item $\approx5,000$ genes, $\approx500 - 5,000$ gRNAs
+\item We measure $\approx5,000$ genes, $\approx500 - 5,000$ gRNAs
 \end{itemize}
@@
 \begin{itemize}
-\item[1.] For given threshold $c \in \N$, estimate $P_i$ by $$ \begin{cases} \hat{P}_i = 0 \textrm{ if } g_i \geq c, \\ \hat{P}_i = 1 \textrm{ if } g_i < c \end{cases}.$$
+\item[1.] For given threshold $c \in \N$, estimate $p_i$ by $$ \begin{cases} \hat{p}_i = 0 \textrm{ if } g_i \geq c, \\ \hat{p}_i = 1 \textrm{ if } g_i < c \end{cases}.$$
 \item[2.] Fit the regression model \cite{Sarkar2021}
-$$ m_i | \left( z_i, l^m_i \right) \sim \textrm{NB}_\theta(\mu_i),$$ where $\theta >0$ is the NB size parameter and $$\log\left(\mu_i\right) = \beta_m \hat{P}_i + \gamma^T_m z_i + \log\left( l_i^m\right).$$
+$$ m_i | \left( z_i, l^m_i \right) \sim \textrm{NB}_\theta(\mu_i),$$ where $\theta >0$ is the NB size parameter, and $$\log\left(\mu_i\right) = \beta_m \hat{p}_i + \gamma^T_m z_i + \log\left( l_i^m\right).$$
 \item[3.] Obtain an estimate $\hat{\beta}_m$ of $\beta_m$ and compute a CI for $\beta_m$.
@@
 \begin{frame}
-\frametitle{Problem 1: There is no clear location at which to draw the threshold.}
+\frametitle{Problem 1: There is no clear location in the data at which to draw the threshold.}
 \begin{figure}
@@
 \begin{frame}
-\frametitle{Problem 2: Thresholding leads to attenuation bias in many measurement error models \cite{Stefanski2000}.}
+\frametitle{Problem 2: Thresholding can lead to attenuation bias.}
+
+As a simple example, suppose
+$$
+\begin{cases}
+p_1, \dots, p_n \sim \textrm{Bern}(\pi) \\
+y_i = \beta_m p_i + \ep_i \\
+x_i = \beta_g p_i + \tau_i,
+\end{cases}
+$$
+where $\ep_i \indep \tau_i$ and $\ep_i, \tau_i \sim N(0,1)$. We observe
+$$ \{(x_1, y_1), \dots, (x_n, y_n)\},$$ and we want to estimate $\beta_m$ using the thresholding method.
+Assume $\beta_g$ and $\pi$ are known.
+We select the threshold $c$ so as to minimize the misclassification rate, i.e., $$c = \argmin_{c \in \R} \frac{1}{n} \sum_{i=1}^n \mathbb{I}(\hat{P}_i \neq P_i).$$
+
+\end{frame}
+
+\begin{frame}
+\frametitle{The thresholding method shows clear attenuation bias.}
+\begin{itemize}
+\item We set $\beta_m$ to 1.
+\end{itemize}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.75\linewidth]{../figures/fig2/att_bias}
+\end{figure}
 
@@
 \begin{itemize}
-\item[1.] \textbf{mRNA}: $m_i | (z_i, l_i^m) \sim \textrm{NB}_\theta(\mu^m_i),$ where $$ \log\left( \mu^m_i \right) = \beta_m P_i + \gamma_m^T z_i + \log(l^m_i).$$
-\item[2.] \textbf{gRNA}: $g_i | (z_i, l_i^g) \sim \textrm{NB}_\theta (\mu_i^g),$ where $$\log(\mu_i^g) = \beta_g P_i + \gamma_g^T z_i + \log(l_i^g)$$
-\item[3.] \textbf{Perturbation}: $P_i \sim \textrm{Bern}(\pi)$, where $\pi \in [0, 1/2)$. $P_i$ is latent.
+\item[1.] \textbf{mRNA}: $m_i | (z_i, l_i^m) \sim \textrm{NB}_\theta(\mu^m_i),$ where $$ \log\left( \mu^m_i \right) = \beta_m p_i + \gamma_m^T z_i + \log(l^m_i).$$
+\item[2.] \textbf{gRNA}: $g_i | (z_i, l_i^g) \sim \textrm{NB}_\theta (\mu_i^g),$ where $$\log(\mu_i^g) = \beta_g p_i + \gamma_g^T z_i + \log(l_i^g)$$
+\item[3.] \textbf{Perturbation}: $p_i \sim \textrm{Bern}(\pi)$, where $\pi \in [0, 1/2)$. $p_i$ is latent.
 \end{itemize}
@@
 
-The canonical parameter of the mRNA distribution for the $i$th cell, $\eta^m_i$, is given by $$\eta_i^m = h_m\left( \beta_m P_i + \gamma^T_m z_i + \log\left( l_i^m \right) \right).$$ If the canonical link function is used, then $h_m$ is the identity.
-The case for $\eta_i^g$ is similar.
+Consider the mRNA model.
+\begin{itemize}
+\item Let $g_m : \R \to \R$ be the link function, i.e.
+$$g_m(\mu_i) = \beta_m p_i + \gamma^T_m z_i + \log(l_i^m).$$
+\item The canonical parameter for the $i$th cell, $\eta^m_i$, is given by
+$$ \eta_i^m = \left[ \psi_m' \right]^{-1} \left( \mu_i \right) = \left[ \psi'_m \right]^{-1} \left( g_m^{-1} \left( \beta_m p_i + \gamma_m^T z_i + \log(l_i^m) \right)\right).$$
+\item Thus, the model is defined by (i) the cumulant-generating function $\psi_m$, and (ii) the link function $g_m$. 
+\end{itemize}
 \end{frame}
@@
 \frametitle{We derive an EM algorithm to fit the model.}
-
 \textbf{E step}:
@@
 \end{itemize}
+\end{frame}
 
+\begin{frame}
+\frametitle{We use statistical tricks to produce an accurate pilot estimate of the parameters, enabling us to run the EM algorithm using only one restart.}
+\begin{itemize}
+	\item Naive approach (random parameter initialization): $$\left( \textrm{15 EM restarts} \right) \left( \frac{ \textrm{20 iterations} }{\textrm{EM restart}} \right) \left( \frac{\textrm{2 GLMs}}{\textrm{iteration}}\right) \approx \textrm{600 GLMs}.$$
+	\item GLM-EIV approach: 
+	$$ \left(\textrm{1 EM restart}\right) \left( \frac{ \textrm{3 iterations}}{\textrm{EM restart}} \right)\left( \frac{ \textrm{2 GLMs} }{ \textrm{iteration} } \right) \approx \textrm{6 GLMs}.$$
+\end{itemize}
 \end{frame}
@@
 \frametitle{We derive an analytic expression for the observed information matrix to enable fast inference (CIs, $p$-values).}
+\begin{multline*}
+J(\hat{\theta}; m, g) = -\E \left[\nabla^2 \mathcal{L}(\theta; m, g, p) | g, m, \hat{\theta} \right] \\ + \E\left[ \nabla \mathcal{L}(\theta; m, g, p) |  g, m, \hat{\theta} \right] \E\left[ \nabla \mathcal{L}(\theta; m, g, p) | g, m, \hat{\theta}\right]^T \\ - \E\left[ \nabla\mathcal{L}(\theta; m, g, p) \nabla \mathcal{L}(\theta; m, g, p)^T | g, m, \hat{\theta} \right].
+\end{multline*}
+\end{frame}
 
-\begin{multline*}
-J(\hat{\theta}; m, g) = -\E \left[\nabla^2 \mathcal{L}(\theta; m, g, p) | g, m, \hat{\theta} \right] \\ + \E\left[ \nabla \mathcal{L}(\theta; m, g, p) |  g, m, \hat{\theta} \right] \cdot \E\left[ \nabla \mathcal{L}(\theta; m, g, p) | g, m, \hat{\theta}\right]^T \\ - \E\left[ \nabla\mathcal{L}(\theta; m, g, p) \nabla \mathcal{L}(\theta; m, g, p)^T | g, m, \hat{\theta} \right].
-\end{multline*}
+\begin{frame}
+\frametitle{We develop a pipeline to deploy the method across hundreds or thousands of processors on HPC and cloud.}
 
@@
 	\centering
-\includegraphics[width=0.4\linewidth]{extra_figs/fig3_crop}
+	\includegraphics[width=0.6\linewidth]{extra_figs/computing}
 \end{figure}
@@
 \begin{frame}
-\frametitle{We implement several statistical accelerations to make the method fast.}
+\frametitle{Method summary}
+\begin{itemize}
+\item We extend the parametric model of \cite{Sarkar2021} to model both mRNA counts and gRNA counts in single-cell CRISPR screen experiments.
+\begin{itemize}
+\item Arbitrary exponential family distributions and link functions are supported. 
+\end{itemize}
+\item We (i) develop a fast EM algorithm to fit the model, (ii) derive an analytic expression for the observed information matrix, and (iii) implement a computational pipeline to deploy the method on HPC and cloud.
+\begin{itemize}
+\item These features enable GLM-EIV to scale to large datasets.
+\end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Overview}
+\begin{itemize}
+	\item Background
+	\item Analysis objective and challenges
+	\item Existing approach
+	\item Proposed method
+	\item \textbf{Simulation results}
+	\item Real data results
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Overview}
+\begin{itemize}
+	\item Background
+	\item Analysis objective and challenges
+	\item Existing approach
+	\item Proposed method
+	\item Simulation results
+	\item\textbf{Real data results}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Real data analysis details}
+\begin{itemize}
+\item Quality control
+\begin{itemize}
+\item Lowly-expressed genes removed
+\item Cells with library sizes below 5th percentile or above 95th percentile removed
+\end{itemize}
+\item mRNA model
+\begin{itemize}
+\item negative binomial distribution (with log link)
+\item size parameter $\theta$ estimated from data.
+\end{itemize}
+\item gRNA model
+\begin{itemize}
+\item Poisson distribution (with log link) \cite{Choudhary2021}
+\end{itemize}
+\end{itemize}
 \end{frame}
@@
 % What is the perturbation probability?
-%
 

### 2021-09-16 R_scripts/plotting/figure1.R

@@ -31,7 +31,7 @@ p1 <- ggplot(data = to_plot, mapping = aes(x = x, y = y, col = Status)) +
   theme(legend.position="bottom") + theme(legend.position="bottom", legend.title=element_blank()) +
   geom_vline(data = dplyr::filter(to_plot, modality == "mRNA"), mapping = aes(xintercept = -0.4), col = "darkred") +
   geom_vline(data = dplyr::filter(to_plot, modality == "mRNA"), mapping = aes(xintercept = 0.0), col = "darkred") +
-  scale_color_manual(values = my_cols) +  geom_line(lwd = 1.1)
+  scale_color_manual(values = my_cols) + geom_line(lwd = 1.1)
 ggsave(filename = paste0(fig1_dir, "density_plot.pdf"), plot = p1, device = "pdf", scale = 1, width = 4, height = 4)
 
 

### 2021-09-16 R_scripts/plotting/figure2.R

@@ -2,12 +2,13 @@
 fig_dir <- paste0(.get_config_path("LOCAL_CODE_DIR"), "glmeiv-manuscript/figures/")
 fig2_dir <- paste0(fig_dir, "fig2/")
 if (!dir.exists(fig2_dir)) dir.create(fig2_dir, recursive = TRUE)
+load_all("~/research_code/glmeiv/")
 
 # load plotting objects
 plotting_objects <- paste0(.get_config_path("LOCAL_CODE_DIR"), "glmeiv-manuscript/figures/plotting_objects.R")
 source(plotting_objects)
 
-
+# a. gRNA count histogram
 set.seed(2)
 gRNA_ids_gasp <- sample(x = get_feature_ids(gRNA_odm_gasp), size = 2, replace = FALSE)
 gRNA_ids_xie <- sample(x = get_feature_ids(gRNA_odm_xie), size = 2, replace = FALSE)
@@ -37,7 +38,7 @@ my_cols <- c("firebrick3", "dodgerblue3")
 
 p1 <- ggplot(data = to_plot %>% dplyr::filter(count >= 1, count <= 40), mapping = aes(x = count)) +
   facet_wrap(lab~., scales = "free_y") + geom_histogram(binwidth = 1, col = "black", fill = "gray") +
-  scale_y_continuous(trans='log10', expand = c(0,NA)) + xlab("gRNA count") + ylab("") +
+  scale_y_continuous(trans='log10', expand = c(0, NA)) + xlab("gRNA count") + ylab("") +
   geom_vline(data = dplyr::filter(to_plot, lab == "Gasperini gRNA 1"),
              mapping = aes(xintercept = dplyr::filter(thresh_df, lab == "Gasperini gRNA 1") %>% dplyr::pull(thresh)), 
              col = my_cols[1], lwd = 1) +
@@ -52,5 +53,23 @@ p1 <- ggplot(data = to_plot %>% dplyr::filter(count >= 1, count <= 40), mapping 
              col = my_cols[2], lwd = 1) + theme_bw()
 ggsave(filename = paste0(fig2_dir, "histograms.pdf"), plot = p1, device = "pdf", scale = 1, width = 6.5, height = 3.75)
 
-
-glmeiv::get_tresholding_estimator_bias(m_perturbation = 1, g_perturbation = 0.1, pi = 0.3)
+# b. theoretical bias of thresholding estimator
+xgrid <- seq(0.2, 6, by = 0.1)
[…]

### 2021-09-17 presentation/glmeiv_talk.tex
finish presentation draft

@@
 	\frametitle{Single-cell CRISPR screens entail sequencing gRNAs and mRNAs in individual cells.}
-
 	\begin{figure}
@@
 		\includegraphics[width=1\linewidth]{extra_figs/experimental_protocol}
+	\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{We investigate single-cell CRISPR screen datasets produced by \cite{Gasperini2019} and \cite{Xie2019}.}
+\begin{figure}
+	\centering
+	\includegraphics[width=1.0\linewidth]{extra_figs/papers}
 \end{figure}
@@
 Assume $\beta_g$ and $\pi$ are known.
-We select the threshold $c$ so as to minimize the misclassification rate, i.e., $$c = \argmin_{c \in \R} \frac{1}{n} \sum_{i=1}^n \mathbb{I}(\hat{P}_i \neq P_i).$$
+We select the threshold $c$ so as to minimize the misclassification rate, i.e., $$c = \argmin_{c \in \R} \frac{1}{n} \sum_{i=1}^n \mathbb{I}(\hat{p}_i \neq p_i).$$
 
@@
 \frametitle{Research question}
-Does modeling the gRNA count distribution directly (thereby bypassing thresholding) improve estimation and inference?
+Does modeling the gRNA count distribution (thereby bypassing thresholding) improve estimation and inference?
 \end{frame}
@@
 \begin{itemize}
-\item We extend the parametric model of \cite{Sarkar2021} to model both mRNA counts and gRNA counts in single-cell CRISPR screen experiments.
+\item We extend the parametric model of \cite{Sarkar2021} to model both mRNA counts and gRNA counts.
 \begin{itemize}
@@
 \end{itemize}
-\item We (i) develop a fast EM algorithm to fit the model, (ii) derive an analytic expression for the observed information matrix, and (iii) implement a computational pipeline to deploy the method on HPC and cloud.
+\item We (i) propose a fast EM algorithm to fit the model, (ii) derive an analytic expression for the observed information matrix, and (iii) implement a computational pipeline to deploy the method on HPC and cloud.
 \begin{itemize}
@@
 \begin{frame}
+\frametitle{Simulation setup}
+\begin{itemize}
+\item No covariates
+\item No offset terms (i.e., library size fixed at one)
+\item Intercept terms fixed
+\item $\pi$ fixed
+\item $\beta_m$ fixed (and of moderate size)
+\item $\beta_g$ varied over an interval
+\item Gaussian, Poisson, and negative binomial distributions
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\begin{figure}
+	\centering
+	\includegraphics[width=1\linewidth]{../figures/fig3/simulation_warmup_arm_g}
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{GLM-EIV outperforms the thresholding method on the simulated data for two main reasons:}
+\begin{itemize}
+\item[1.] GLM-EIV leverages information from \textit{both} modalities to assign perturbation identities to cells.
+\item[2.] GLM-EIV generates \textit{soft} rather than \textit{hard} assignments, capturing the inherent uncertainty in whether a perturbation occurred.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
 \frametitle{Overview}
@@
 \begin{frame}
-\frametitle{Real data analysis details}
+\frametitle{We followed recommendations of \cite{Choudhary2021} for quality control and modeling.}
+
 \begin{itemize}
@@
 \begin{itemize}
-\item Lowly-expressed genes removed
-\item Cells with library sizes below 5th percentile or above 95th percentile removed
+\item Lowly-expressed genes filtered
+\item Cells with library sizes below 5th percentile or above 95th percentile filtered
 \end{itemize}
@@
 \begin{itemize}
-\item negative binomial distribution (with log link)
-\item size parameter $\theta$ estimated from data.
+\item used negative binomial distribution (with log link)
+\item size parameter $\theta$ estimated from data
 \end{itemize}
@@
 \begin{itemize}
-\item Poisson distribution (with log link) \cite{Choudhary2021}
+\item used Poisson distribution (with log link)
 \end{itemize}
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{The estimate $\hat{\beta}_g$ for $\beta_g$ was large on both datasets across all site types.}
+\begin{center}
+\textbf{Gasperini}
+\end{center}
+\begin{center}
+\begin{tabular}{|c|c|}
+	\hline 
+	Site type & Mean $\exp\left(\beta_g\right)$ 95\% CI \\ 
+	\hline 
+ 	 \textcolor{blue}{Candidate \textit{cis}} & $(4453, 5353)$ \\ 
+	\hline 
+	\textcolor{blue}{Negative control} &  $(4484, 5408)$ \\ 
+	\hline
+	\textcolor{blue}{TSS-targeting} & $(3605, 4205)$ \\
+	\hline
+\end{tabular} 
+\end{center}
+
+\begin{center}
+	\textbf{Xie}
+\end{center}
+\begin{center}
+	\begin{tabular}{|c|c|}
+		\hline 
+		Site type & Mean $\exp\left(\beta_g\right)$ 95\% CI \\ 
+		\hline 
+		\textcolor{blue}{Candidate \textit{cis}} & $(307, 324)$ \\ 
+		\hline 
+		\textcolor{blue}{Negative control} &  $(299, 316)$ \\
+		\hline
+	\end{tabular} 
+\end{center}
+
+\end{frame}
+
+\begin{frame}
+\frametitle{GLM-EIV and thresholding exhibited similar CI coverage rates on the negative control pairs.}
+\begin{center}
+\begin{tabular}{|c|c|c|}
+	\hline 
+	Dataset & GLM-EIV & Thresholding \\ 
+	\hline 
+	\textcolor{blue}{Xie} & 93.7\% & 93.2\% \\ 
+	\hline 
+	\textcolor{blue}{Gasperini} & 90.6\% & 91.4\%  \\ 
+	\hline 
+\end{tabular}
+\end{center}
+\end{frame}
+
+\begin{frame}
+\frametitle{Estimates and CIs produced by GLM-EIV and the thresholding method were similar.}
+\begin{figure}
+	\centering
+	\includegraphics[width=1\linewidth]{../figures/fig5/comparison}
+\end{figure}
+\begin{center}
+Candidate \textit{cis} pairs in Xie et al.\ data.
+\end{center}
+\end{frame}
+
+\begin{frame}
+\frametitle{We now can answer our core research question.}
+\textbf{Research question}: Does modeling the gRNA count distribution (thereby bypassing thresholding) improve estimation and inference?
+\begin{itemize}
+\item Yes, if the problem is in a ``sufficiently challenging'' setting.
+\item The real data that we analyzed, surprisingly, were in an ``easy'' setting.
+\item Therefore, GLM-EIV and the thresholding method performed similarly on the real data.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{The proposed method could help solve practical challenges.}
+\begin{itemize}
+\item Selecting cell-specific thresholds
+\item Identifying ``problem difficulty'' and thus whether thresholding is appropriate
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Together, GLM-EIV and SCEPTRE shed light on core analysis challenges posed by single-cell CRISPR screens, paving the way for the development of new methods.}
+\begin{itemize}
+	\item[1.] Perturbation unobserved
+	\item[2.] Confounders and nuisance variables
+	\item[3.] Possible model mispecification
+	\item[4.] Background reads
+	\item[5.] Highly discrete data
+	\item[6.] Ineffective gRNAs
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Thank you.}
+Acknowledgments:
+\begin{itemize}
+\item Xuran Wang helped with the Xie data preprocessing.
+\item All analyses were run on Pittsburgh Supercomputer.
 \end{itemize}
@@
 % What is the perturbation probability?
-
 % Practical strategies:

### 2021-09-17 R_scripts/analysis/gasperini_analysis.R

@@ -1,0 +1,29 @@
+library(magrittr)
+
+# load the results
+gasp_result_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/gasperini/results/")
+glmeiv_res <- readRDS(paste0(gasp_result_dir, "result_glmeiv.rds"))
+thresh_res <- readRDS(paste0(gasp_result_dir, "result_thresholding.rds"))
+
+# first, examine g_perturbation for glmeiv in the different categories
+glmeiv_res %>% dplyr::filter(parameter == "g_perturbation",
+                             target %in% c("confint_lower", "confint_upper", "estimate"),
+                             site_type %in% c("DHS", "NTC", "positive_ctrl", "selfTSS")) %>% 
+  dplyr::select(target, value, pair_id, site_type) %>%
+  tidyr::pivot_wider(id_cols = c("pair_id" , "site_type"), names_from = "target", values_from = "value") %>%
+  dplyr::group_by(site_type) %>% dplyr::summarize(m_est = mean(estimate),
+                                                  m_confint_lower = mean(confint_lower),
+                                                  m_confint_upper = mean(confint_upper))
+
+# examine CI coverage rate for negative control pairs for both methods
+get_negctrl_ci_coverage_rate <- function(df) {
+  df %>% dplyr::filter(parameter == "m_perturbation", site_type == "NTC",
+                               target %in% c("confint_lower", "confint_upper", "estimate")) %>%
+    dplyr::select(target, value, pair_id) %>%
+    tidyr::pivot_wider(id_cols = "pair_id", names_from = "target", values_from = "value") %>%
+    dplyr::mutate(covered = confint_lower < 1 & confint_upper > 1) %>%
+    dplyr::summarize(mean(covered, na.rm = TRUE))
+}
+
+get_negctrl_ci_coverage_rate(glmeiv_res)
+get_negctrl_ci_coverage_rate(thresh_res %>% dplyr::filter(threshold == 5))

### 2021-09-17 R_scripts/analysis/xie_analysis.R

@@ -1,0 +1,50 @@
+library(magrittr)
+library(ggplot2)
+
+# load the results
+xie_result_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/xie/results/")
+glmeiv_res <- readRDS(paste0(xie_result_dir, "result_glmeiv.rds")) %>% dplyr::distinct() # ensure ahead of time rows are distinct; check gRNA_gene_pairs data frame.
+thresh_res <- readRDS(paste0(xie_result_dir, "result_thresholding.rds")) %>% dplyr::distinct()
+
+# examine g_perturbation for glmeiv in the different categories
+glmeiv_res %>% dplyr::filter(parameter == "g_perturbation",
+                             target %in% c("confint_lower", "confint_upper", "estimate"),
+                             type %in% c("neg_control", "cis")) %>%
+  dplyr::select(target, value, pair_id, type) %>%
+  tidyr::pivot_wider(id_cols = c("pair_id", "type"),
+                     names_from = "target", values_from = "value") %>%
+  dplyr::group_by(type) %>% dplyr::summarize(m_est = mean(estimate),
+                                             m_confint_lower = mean(confint_lower, na.rm = TRUE),
+                                             m_confint_upper = mean(confint_upper, na.rm = TRUE))
+
+# examine CI coverage rate for negative control pairs for both methods
+get_negctrl_ci_coverage_rate <- function(df) {
+  df %>% dplyr::filter(parameter == "m_perturbation",
+                               target %in% c("confint_lower", "confint_upper", "estimate"),
+                               type %in% c("neg_control")) %>%
+    dplyr::select(target, value, pair_id, type) %>%
+    tidyr::pivot_wider(id_cols = c("pair_id", "type"),
+                       names_from = "target", values_from = "value") %>%
+    dplyr::mutate(covered = confint_upper > 1 & confint_lower < 1) %>%
+    dplyr::summarize(coverage_rate = mean(covered, na.rm = TRUE) * 100) 
+}
+
+get_negctrl_ci_coverage_rate(glmeiv_res)
+get_negctrl_ci_coverage_rate(thresh_res)
+
+# Compare estimates on candidate cis pairs
+thresh_df <- thresh_res %>% dplyr::filter(parameter == "m_perturbation",
+                             target %in% c("estimate", "confint_lower", "confint_upper"), type == "cis") %>%
+  dplyr::select(target, value, pair_id)
[…]

### 2021-09-17 R_scripts/plotting/figure3.R

@@ -1,0 +1,99 @@
+# Figure 3 displays the results of the "warmup" simulation -- no covariates, no offsets, no biological interpretation.
+
+# set figure 3 dir
+fig_dir <- paste0(.get_config_path("LOCAL_CODE_DIR"), "glmeiv-manuscript/figures/")
+fig3_dir <- paste0(fig_dir, "fig3/")
+if (!dir.exists(fig3_dir)) dir.create(fig3_dir, recursive = TRUE)
+
+# load packages and set sim_dir location
+sim_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "private/simulations")
+library(ggplot2)
+library(magrittr)
+
+######################
+# Distribution 1: Gaus
+######################
+sim_spec <- readRDS(paste0(sim_dir, "/sim_spec_4.rds")) # simulatr specifier object
+sim_res <- readRDS(paste0(sim_dir, "/raw_result_4.rds")) # raw results; note: PSC failed for certain settings in
+
+# possibly subset the sim_res according to some criteria
+id_classifications <- glmeiv::obtain_valid_ids(sim_res = sim_res, pi_upper = 0.4)
+valid_ids <- id_classifications$valid_ids
+sim_res_sub <- dplyr::filter(sim_res, id %in% valid_ids)
+
+summarized_results_gaus <- simulatr::summarize_results(sim_spec = sim_spec, sim_res = sim_res_sub,
+                                                       metrics = c("bias", "mse", "coverage"),
+                                                       parameters = c("m_perturbation"),
+                                                       threshold = 0.1) %>% tibble::as_tibble() %>% dplyr::filter(arm_arm_pi_big)
+pi <- summarized_results_gaus$pi[1]  
+m_pert <- sim_spec@fixed_parameters$m_perturbation
+my_funct <- function(x) glmeiv::get_tresholding_estimator_bias(m_perturbation = m_pert, g_perturbation = x, pi = pi)
+my_cols <- c("firebrick3", "dodgerblue3")
+
+density_dfs_and_thresholds <- glmeiv::get_theoretical_densities_and_thresholds(sim_spec = sim_spec,
+                                                                               xgrid = seq(-10, 10, 0.1))
+idx <- 1
+glmeiv::plot_mixture(density_df = density_dfs_and_thresholds$g_dfs[[idx]],
+                     thresh = density_dfs_and_thresholds$g_thresholds[idx],
+                     x_max = 5, x_min = 0, points = FALSE)
+
[…]

### 2021-09-17 R_scripts/plotting/figure5.R

@@ -1,0 +1,55 @@
+# set figure 3 dir
+fig_dir <- paste0(.get_config_path("LOCAL_CODE_DIR"), "glmeiv-manuscript/figures/")
+fig5_dir <- paste0(fig_dir, "fig5/")
+if (!dir.exists(fig5_dir)) dir.create(fig5_dir, recursive = TRUE)
+
+# load packages and results
+xie_result_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/xie/results/")
+glmeiv_res <- readRDS(paste0(xie_result_dir, "result_glmeiv.rds")) %>% dplyr::distinct() # ensure ahead of time rows are distinct; check gRNA_gene_pairs data frame.
+thresh_res <- readRDS(paste0(xie_result_dir, "result_thresholding.rds")) %>% dplyr::distinct()
+library(ggplot2)
+library(magrittr)
+
+# examine g_perturbation for glmeiv in the different categories
+glmeiv_res %>% dplyr::filter(parameter == "g_perturbation",
+                             target %in% c("confint_lower", "confint_upper", "estimate"),
+                             type %in% c("neg_control", "cis")) %>%
+  dplyr::select(target, value, pair_id, type) %>%
+  tidyr::pivot_wider(id_cols = c("pair_id", "type"),
+                     names_from = "target", values_from = "value") %>%
+  dplyr::group_by(type) %>% dplyr::summarize(m_est = mean(estimate),
+                                             m_confint_lower = mean(confint_lower, na.rm = TRUE),
+                                             m_confint_upper = mean(confint_upper, na.rm = TRUE))
+
+# examine CI coverage rate for negative control pairs for both methods
+get_negctrl_ci_coverage_rate <- function(df) {
+  df %>% dplyr::filter(parameter == "m_perturbation",
+                       target %in% c("confint_lower", "confint_upper", "estimate"),
+                       type %in% c("neg_control")) %>%
+    dplyr::select(target, value, pair_id, type) %>%
+    tidyr::pivot_wider(id_cols = c("pair_id", "type"),
+                       names_from = "target", values_from = "value") %>%
+    dplyr::mutate(covered = confint_upper > 1 & confint_lower < 1) %>%
+    dplyr::summarize(coverage_rate = mean(covered, na.rm = TRUE) * 100) 
+}
+
+get_negctrl_ci_coverage_rate(glmeiv_res)
+get_negctrl_ci_coverage_rate(thresh_res)
+
[…]

### 2021-10-11 text/outline/outline.tex
start text outline

\documentclass[12pt]{article}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{graphicx} 
\usepackage{float}
\usepackage[caption = false]{subfig}
\usepackage{/Users/timbarry/optionFiles/mymacros}

\begin{document}
\noindent
Tim B.
\begin{center}
\textbf{A generalized errors-in-variables model, with application to single-cell CRISPR screens}
\end{center}

\section{Introduction}
CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new therapies for diseases \cite{Rothgangl2021,Musunuru2021} and accelerating basic biology research \cite{Przybyla2021}. Recently, scientists have paired CRISPR genome engineering with single-cell sequencing \cite{Dixit2016,Datlinger2017}. The resultant assays, known as a ``single-cell CRISPR screens,'' link specific genetic perturbations to changes in cellular phenotypes, illuminating regulatory networks underlying human diseases and other traits \cite{Morris2021a}.

Despite their extraordinary promise, single-cell CRISPR screens present substantial statistical and computational challenges.

% These assays, called ``single-cell CRISPR screens,'' could 
	 % These assays, called single-cell CRISPR screens, promise to transform our understanding of disease by enabling 
	% CRISPR promises to transform clinical treatment of diseases and basic biology research alike: 
%	\begin{itemize}
	% a biotechnology that enables scientists to precisely and efficiently edit genomes.
	% \item CRISPR technology also can be used to drive basic biological discovery. 
	% \end{itemize}
%	\item \textbf{P2}:  \cite{Przybyla2021}
%	\item \textbf{P3}: Despite their considerable promise, single-cell CRISPR screens pose substantial statistical and computational challenges.
%	\item \textbf{P4}: In this work we make two main contributions.
%	\begin{itemize}
%	\item First, we study the thresholding method from both theoretical and empirical perspectives, characterizing its properties and illuminating the settings in which it is expected to fail.
%	\item Second, we propose 
%	\end{itemize}

	% \item In summary we couple a novel statistical method -- GLM-EIV -- to powerful technologies from genomics and computer science -- CRISPR genome engineering, single-cell sequencing, clound computing, and high-performance computing -- to yield new statistical and biological insights.
% \end{itemize}

\section{Background and analysis challenges}

\begin{itemize}
\item 
\end{itemize}

\section{Thresholding method}

\begin{itemize}
\item 
\item 
\end{itemize}

\subsection{Theoretical analysis}

\subsection{Empirical analysis}

\section{GLM-EIV}

\section{Simulation studies}

\section{Real data analysis}

\section{Discussion}

\section{Appendix}

% \subsection{Dictionary of symbols}

\subsection{Proofs of theoretical results for thresholding estimator}

\subsection{Derivation of EM algorithm}

\subsection{Derivation of observed information matrix}

\subsection{Implementation using R family objects}

\subsection{Statistical accelerations to GLM-EIV}

\subsection{Additional simulation results}

\bibliographystyle{unsrt}
\bibliography{/Users/timbarry/optionFiles/glmeiv.bib}

\end{document}

### 2021-10-11 R_scripts/analysis/threshold_sensitivity_analysis.R

@@ -1,0 +1,55 @@
+library(magrittr)
+library(ggplot2)
+library(cowplot)
+
+fig_dir <- paste0(.get_config_path("LOCAL_CODE_DIR"), "glmeiv-manuscript/figures/")
+fig1_dir <- paste0(fig_dir, "fig1/")
+if (!dir.exists(fig1_dir)) dir.create(fig1_dir, recursive = TRUE)
+
+
+# load the results
+gasp_result_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/gasperini/results/")
+thresh_res <- readRDS(paste0(gasp_result_dir, "result_thresholding.rds"))
+
+# study how choice of threshold influences results on Gasperini self TSS pairs
+thresh_res_pc <- thresh_res %>% dplyr::filter(site_type == "selfTSS", parameter == "m_perturbation")
+
+# Question 1. How do thresh = 1 and thresh = 5 compare on estimation?
+thresh_res_pc_wide <- thresh_res_pc %>% dplyr::select(target, value, threshold, pair_id) %>%
+  tidyr::pivot_wider(id_cols = c("pair_id", "target"),
+                     names_from = "threshold",
+                     values_from = "value", names_prefix = "threshold_")
+
+to_plot <- thresh_res_pc_wide %>% dplyr::filter(target == "estimate") %>% dplyr::select(threshold_1, threshold_5)
+p1 <- ggplot(data = to_plot, mapping = aes(x = threshold_5, y = threshold_1)) + geom_point(alpha = 0.7, col = my_cols[2]) +
+  xlab("Threshold = 5 (est)") + ylab("Threshold = 1 (est)") + geom_abline(slope = 1, intercept = 0) + theme_bw(base_size = 10)
+# threshold = 1 ests are bigger than (i.e., closer to one than) threshold = 5; this is attenuation bias due to the fact that excessively many unperturbed cells have been classified as perturbed.
+
+my_cols <- c("firebrick3", "dodgerblue3", "orchid4")
+# Question 2. How do CI widths and p-values change for thresh = 5 vs thresh = 20?
+
+# a) est. vs. est.
+to_plot <- thresh_res_pc_wide %>% dplyr::filter(target == "estimate") %>% dplyr::select(threshold_5, threshold_20)
[…]

### 2021-10-11 R_scripts/plotting/old/figure3.R

@@ -28,7 +28,6 @@ summarized_results_gaus <- simulatr::summarize_results(sim_spec = sim_spec, sim_
 pi <- summarized_results_gaus$pi[1]  
 m_pert <- sim_spec@fixed_parameters$m_perturbation
 my_funct <- function(x) glmeiv::get_tresholding_estimator_bias(m_perturbation = m_pert, g_perturbation = x, pi = pi)
-my_cols <- c("firebrick3", "dodgerblue3")
 
 density_dfs_and_thresholds <- glmeiv::get_theoretical_densities_and_thresholds(sim_spec = sim_spec,
                                                                                xgrid = seq(-10, 10, 0.1))

### 2021-10-11 R_scripts/plotting/old/figure4.R

@@ -1,0 +1,58 @@
+library(magrittr)
+library(ggplot2)
+library(cowplot)
+
+fig_dir <- paste0(.get_config_path("LOCAL_CODE_DIR"), "glmeiv-manuscript/figures/")
+fig4_dir <- paste0(fig_dir, "fig4/")
+if (!dir.exists(fig4_dir)) dir.create(fig4_dir, recursive = TRUE)
+
+# load the results
+gasp_result_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/gasperini/results/")
+thresh_res <- readRDS(paste0(gasp_result_dir, "result_thresholding.rds"))
+
+# study how choice of threshold influences results on Gasperini self TSS pairs
+thresh_res_pc <- thresh_res %>% dplyr::filter(site_type == "selfTSS", parameter == "m_perturbation")
+
+# Question 1. How do thresh = 1 and thresh = 5 compare on estimation?
+thresh_res_pc_wide <- thresh_res_pc %>% dplyr::select(target, value, threshold, pair_id) %>%
+  tidyr::pivot_wider(id_cols = c("pair_id", "target"),
+                     names_from = "threshold",
+                     values_from = "value", names_prefix = "threshold_")
+
+to_plot <- thresh_res_pc_wide %>% dplyr::filter(target == "estimate") %>% dplyr::select(threshold_1, threshold_5)
+p1 <- ggplot(data = to_plot, mapping = aes(x = threshold_5, y = threshold_1)) + geom_point(alpha = 0.7, col = my_cols[2], size = 1) +
+  xlab("Threshold = 5 (est)") + ylab("Threshold = 1 (est)") + geom_abline(slope = 1, intercept = 0) + theme_cowplot(font_size = 11)
+# threshold = 1 ests are bigger than (i.e., closer to one than) threshold = 5; this is attenuation bias due to the fact that excessively many unperturbed cells have been classified as perturbed.
+
+my_cols <- c("firebrick3", "dodgerblue3", "orchid4")
+# Question 2. How do CI widths and p-values change for thresh = 5 vs thresh = 20?
+
+# a) est. vs. est.
+to_plot <- thresh_res_pc_wide %>% dplyr::filter(target == "estimate") %>% dplyr::select(threshold_5, threshold_20)
[…]

### 2021-10-11 R_scripts/plotting/thresholding_empirical.R

@@ -1,0 +1,50 @@
+library(magrittr)
+library(ggplot2)
+library(cowplot)
+
+# load the results
+gasp_result_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/gasperini/results/")
+thresh_res <- readRDS(paste0(gasp_result_dir, "result_thresholding.rds"))
+
+# study how choice of threshold influences results on Gasperini self TSS pairs
+thresh_res_pc <- thresh_res %>% dplyr::filter(site_type == "selfTSS", parameter == "m_perturbation")
+
+# Question 1. How do thresh = 1 and thresh = 5 compare on estimation?
+thresh_res_pc_wide <- thresh_res_pc %>% dplyr::select(target, value, threshold, pair_id) %>%
+  tidyr::pivot_wider(id_cols = c("pair_id", "target"),
+                     names_from = "threshold",
+                     values_from = "value", names_prefix = "threshold_")
+
+to_plot <- thresh_res_pc_wide %>% dplyr::filter(target == "estimate") %>% dplyr::select(threshold_1, threshold_5)
+p1 <- ggplot(data = to_plot, mapping = aes(x = threshold_5, y = threshold_1)) + geom_point(alpha = 0.7, col = my_cols[2]) +
+  xlab("Threshold = 5 (est)") + ylab("Threshold = 1 (est)") + geom_abline(slope = 1, intercept = 0) + theme_bw(base_size = 10)
+# threshold = 1 ests are bigger than (i.e., closer to one than) threshold = 5; this is attenuation bias due to the fact that excessively many unperturbed cells have been classified as perturbed.
+
+my_cols <- c("firebrick3", "dodgerblue3", "orchid4")
+# Question 2. How do CI widths and p-values change for thresh = 5 vs thresh = 20?
+
+# a) est. vs. est.
+to_plot <- thresh_res_pc_wide %>% dplyr::filter(target == "estimate") %>% dplyr::select(threshold_5, threshold_20)
+p2 <- ggplot(data = to_plot, mapping = aes(x = threshold_5, y = threshold_20)) + geom_point(alpha = 0.9, col = my_cols[2]) +
+  xlab("Threshold = 5 (est)") + ylab("Threshold = 20 (est)") + geom_abline(slope = 1, intercept = 0, lwd = 0.75) + theme_bw(base_size = 10)
+
[…]

### 2021-10-11 R_scripts/plotting/thresholding_theoretical.R

@@ -1,0 +1,45 @@
+load_all("~/research_code/glmeiv/")
+my_cols <- c("firebrick3", "dodgerblue3")
+library(magrittr)
+library(ggplot2)
+library(cowplot)
+
+# a) estimate vs. threshold for different values of beta_g
+x_grid <- seq(0, 7, length.out = 501)
+beta_gs <- c(1, qnorm(0.75) * 2, 1.5)
+
+to_plot_a <- lapply(X = beta_gs, function(curr_beta_g) {
+  y <- get_tresholding_estimator_bias(m_perturbation = 1,
+                                      g_perturbation = curr_beta_g,
+                                      pi = 0.5,
+                                      c = x_grid,
+                                      return_bias = FALSE)
+  data.frame(Estimate = y, Threshold = x_grid, beta_g = curr_beta_g)
+}) %>% do.call(what = "rbind", args = .) %>%
+  dplyr::mutate(beta_g_disp = paste0("beta[g] == ", round(beta_g, 2)) %>% factor())
+
+# subplot a) bias in pi = 1/2 model
+p_a <- ggplot(data = to_plot_a, mapping = aes(x = Threshold, y = Estimate)) +
+  geom_hline(yintercept = 0.5, col = "darkred") +
+  facet_wrap(.~beta_g_disp, labeller = label_parsed, scales = "free_y") +
+  theme_cowplot(font_size = 11) + scale_x_continuous(expand = c(NA, 0)) +
+  geom_vline(data = dplyr::filter(to_plot_a, beta_g_disp == "beta[g] == 1"),
+             mapping = aes(xintercept = beta_gs[1]/2), col = my_cols[2]) +
+  geom_vline(data = dplyr::filter(to_plot_a, beta_g_disp == "beta[g] == 1.35"),
+             mapping = aes(xintercept = beta_gs[2]/2), col = my_cols[2]) +
+  geom_vline(data = dplyr::filter(to_plot_a, beta_g_disp == "beta[g] == 1.5"),
+             mapping = aes(xintercept = beta_gs[3]/2), col = my_cols[2]) +
+  geom_line()
+
+# subplot b) bias as c -> infinity for pi in [0,1/2]
+pi <- seq(0.0, 0.5, length.out = 11)
+Estimate <- 1 - pi
+to_plot_b <- data.frame(pi = pi, Estimate = Estimate)
+
[…]

### 2021-10-11 bash_scripts/setup.sh

@@ -19,4 +19,5 @@ $public_dir"simulations/results" \
 $public_dir"gasperini/data" \
 $public_dir"gasperini/results" \
 $public_dir"xie/data" \
-$public_dir"xie/results"
+$public_dir"xie/results" \
+$public_dir"aux"

### 2021-10-12 text/outline/outline.tex
update ms

@@
 \noindent
-Tim B.
+Tim, Gene, Kathryn
 \begin{center}
-\textbf{A generalized errors-in-variables model, with application to single-cell CRISPR screens}
+\textbf{A generalized errors-in-variables model, with application to CRISPR genome editing and single-cell sequencing}
 \end{center}
@@
 \section{Introduction}
-CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new therapies for diseases \cite{Rothgangl2021,Musunuru2021} and accelerating basic biology research \cite{Przybyla2021}.
+CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new medical therapies \cite{Rothgangl2021,Musunuru2021} and transforming basic biology research \cite{Przybyla2021}.
 Recently, scientists have paired CRISPR genome engineering with single-cell sequencing \cite{Dixit2016,Datlinger2017}.
-The resultant assays, known as a ``single-cell CRISPR screens,'' link specific genetic perturbations to changes in cellular phenotypes, illuminating regulatory networks underlying human diseases and other traits \cite{Morris2021a}.
+The resulting assays, known as a ``single-cell CRISPR screens,'' link genetic perturbations in individual cells to changes in gene expression, illuminating regulatory networks underlying human diseases and other traits \cite{Morris2021a}.
 
-Despite their extraordinary promise, single-cell CRISPR screens present substantial statistical and computational challenges.
+Despite their extraordinary promise, single-cell CRISPR screens present substantial statistical challenges.
+A major difficulty is that CRISPR perturbations are (i) unobserved and (ii) assigned randomly to cells.
+These facts together imply that the analyst cannot know with certainty which cells were perturbed in a single-cell CRISPR screen experiment.
+The analyst instead must leverage an indirect, noisy proxy of perturbation presence or absence -- namely, transcribed barcode counts -- to ``guess'' which cells were perturbed.
+Using these imputed perturbation assignments, the analyst can attempt to estimate the effect of the perturbation on gene expression.
+The standard approach, which we call the ``thresholding method,'' is to assign perturbation identities to cells by thresholding the barcode counts.
 
-% These assays, called ``single-cell CRISPR screens,'' could 
-	 % These assays, called single-cell CRISPR screens, promise to transform our understanding of disease by enabling 
-	% CRISPR promises to transform clinical treatment of diseases and basic biology research alike: 
-%	\begin{itemize}
-	% a biotechnology that enables scientists to precisely and efficiently edit genomes.
-	% \item CRISPR technology also can be used to drive basic biological discovery. 
-	% \end{itemize}
-%	\item \textbf{P2}:  \cite{Przybyla2021}
-%	\item \textbf{P3}: Despite their considerable promise, single-cell CRISPR screens pose substantial statistical and computational challenges.
-%	\item \textbf{P4}: In this work we make two main contributions.
-%	\begin{itemize}
-%	\item First, we study the thresholding method from both theoretical and empirical perspectives, characterizing its properties and illuminating the settings in which it is expected to fail.
-%	\item Second, we propose 
-%	\end{itemize}
+We study estimation and inference in single-cell CRISPR screen experiments  from a statistical perspective, casting the data generating mechanism as a new class of errors-in-variables (or measurement error) models.
+We assume that the response variable $y$ is a GLM of an underlying predictor variable $x^*$.
+We do not observe $x^*$ directly; rather, we observe a noisy version $x$ of $x^*$ that itself is a GLM of $x^*$.
+The goal of the analysis is to estimate (and perform inference on) the effect of $x^*$ on $y$ using the observed data $(x, y)$.
+In the context of the biological application, $x^*$, $y$, and $x$ are perturbations, gene expressions, and barcode counts, respectively.
+
+% First, within the proposed GLM-EIV
 

### 2021-10-12 text/outline/outline.tex
update intro

@@
 
-Despite their extraordinary promise, single-cell CRISPR screens present substantial statistical challenges.
+Despite their promise, single-cell CRISPR screens present substantial statistical challenges.
 A major difficulty is that CRISPR perturbations are (i) unobserved and (ii) assigned randomly to cells.
-These facts together imply that the analyst cannot know with certainty which cells were perturbed in a single-cell CRISPR screen experiment.
-The analyst instead must leverage an indirect, noisy proxy of perturbation presence or absence -- namely, transcribed barcode counts -- to ``guess'' which cells were perturbed.
+The analyst cannot know with certainty which cells were perturbed and instead must leverage an indirect, noisy proxy of perturbation presence or absence -- namely, transcribed barcode counts -- to ``guess'' which cells were perturbed.
 Using these imputed perturbation assignments, the analyst can attempt to estimate the effect of the perturbation on gene expression.
@@
 
-We study estimation and inference in single-cell CRISPR screen experiments  from a statistical perspective, casting the data generating mechanism as a new class of errors-in-variables (or measurement error) models.
+We study estimation and inference in single-cell CRISPR screens from a statistical perspective, formulating the data generating mechanism as a new class of errors-in-variables (or measurement error) models.
 We assume that the response variable $y$ is a GLM of an underlying predictor variable $x^*$.
 We do not observe $x^*$ directly; rather, we observe a noisy version $x$ of $x^*$ that itself is a GLM of $x^*$.
-The goal of the analysis is to estimate (and perform inference on) the effect of $x^*$ on $y$ using the observed data $(x, y)$.
-In the context of the biological application, $x^*$, $y$, and $x$ are perturbations, gene expressions, and barcode counts, respectively.
+The goal of the analysis is to estimate the effect of $x^*$ on $y$ using the observed data $(x , y)$ only.
+In the context of the biological application, $x^*$, $y$, and $x$ are CRISPR perturbations, gene expressions, and barcode counts, respectively.
 
-% First, within the proposed GLM-EIV
+Within this framework we make two contributions.
+First, we study the thresholding method from theoretical and empirical perspectives.
+We characterize the properties of the estimator and uncover settings in which
 
-	% \item In summary we couple a novel statistical method -- GLM-EIV -- to powerful technologies from genomics and computer science -- CRISPR genome engineering, single-cell sequencing, clound computing, and high-performance computing -- to yield new statistical and biological insights.
-% \end{itemize}
+% Next, we propose a new method
+
+GLM-EIV (GLM-based errors in variables).
+GLM-EIV implicitly estimates the probability that each cell was perturbed, thereby obviating the need to explicitly impute perturbation identities via thresholding or some other heuristic.
+
+% Overall, we show how coupling novel, principled statistical approaches to current, powerful technologies from genomics and computer science -- namely, CRISPR genome engineering, single-cell sequencing, cloud computing, and high-performance computing -- can drive new biological insights.
 
@@
 \end{itemize}
+
+\section{Related work}
 

### 2021-10-12 text/outline/outline.tex
finish introduction draft

@@
 \begin{center}
-\textbf{A generalized errors-in-variables model, with application to CRISPR genome editing and single-cell sequencing}
+\textbf{Applying a new class of measurement error models to CRISPR genome editing and single-cell sequencing}
 \end{center}
@@
 A major difficulty is that CRISPR perturbations are (i) unobserved and (ii) assigned randomly to cells.
-The analyst cannot know with certainty which cells were perturbed and instead must leverage an indirect, noisy proxy of perturbation presence or absence -- namely, transcribed barcode counts -- to ``guess'' which cells were perturbed.
+The analyst does not know with certainty which cells were perturbed and instead must leverage an indirect, noisy proxy of perturbation presence or absence -- namely, transcribed barcode counts -- to ``guess'' which cells were perturbed.
 Using these imputed perturbation assignments, the analyst can attempt to estimate the effect of the perturbation on gene expression.
@@
 
-Within this framework we make two contributions.
-First, we study the thresholding method from theoretical and empirical perspectives.
-We characterize the properties of the estimator and uncover settings in which
+Within this framework we make two main contributions.
+First, we carefully study the thresholding method from empirical and theoretical perspectives.
+Notably, we demonstrate the existence of a bias-variance tradeoff for the thresholding method on real data, and we recover this phenomenon in precise mathematical terms in an idealized Gaussian model.
+Second, we introduce a new method for estimation and inference in single-cell CRISPR screens that accounts for measurement error inherent in the experiment.
+The method, called \textit{GLM-EIV} (generalized linear model with errors-in-variables), implicitly estimates the probability that each cell received a perturbation, obviating the need to explicitly impute perturbation identities via thresholding or some other heuristic.
+Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in vast regions of the parameter space.
 
-% Next, we propose a new method
+We implement several statistical accelerations to bring the cost of GLM-EIV to within an order of magnitude of the thresholding method.
+Finally, we develop a computational infrastructure to deploy GLM-EIV at-scale across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters.
+Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, high multiplicity-of-infection single-cell CRISPR screen datasets, yielding new biological and statistical insights.
 
-GLM-EIV (GLM-based errors in variables).
-GLM-EIV implicitly estimates the probability that each cell was perturbed, thereby obviating the need to explicitly impute perturbation identities via thresholding or some other heuristic.
-
-% Overall, we show how coupling novel, principled statistical approaches to current, powerful technologies from genomics and computer science -- namely, CRISPR genome engineering, single-cell sequencing, cloud computing, and high-performance computing -- can drive new biological insights.
+% Overall, our work shows how coupling a novel statistical approach to powerful technologies from genomics and computer science -- CRISPR genome engineering, single-cell sequencing, cloud computing, and high-performance computing -- can drive new biological insights.
 
@@
 
-\begin{itemize}
-\item 
-\end{itemize}
+\subsection{Background on experimental protocol}
+
+\subsection{Analysis challenges}
 

### 2021-10-14 text/outline/outline.tex
update background section

--- 2021-10-12 text/outline/outline.tex
@@
 \usepackage{graphicx}
+\usepackage{xcolor}
+\newcommand{\blue}[1]{\textcolor{blue}{#1}}
 \usepackage{float}
@@
 \begin{center}
-\textbf{A generalized errors-in-variables model, with application to CRISPR genome editing and single-cell sequencing}
+\textbf{A new class of measurement error models, with application to CRISPR genome editing and single-cell sequencing}
 \end{center}
@@
 Despite their promise, single-cell CRISPR screens present substantial statistical challenges.
-A major difficulty is that CRISPR perturbations are (i) unobserved and (ii) assigned randomly to cells.
-The analyst cannot know with certainty which cells were perturbed and instead must leverage an indirect, noisy proxy of perturbation presence or absence -- namely, transcribed barcode counts -- to ``guess'' which cells were perturbed.
-Using these imputed perturbation assignments, the analyst can attempt to estimate the effect of the perturbation on gene expression.
-The standard approach, which we call the ``thresholding method,'' is to assign perturbation identities to cells by thresholding the barcode counts.
+A major difficulty is that CRISPR perturbations are unobservable and assigned stochastically to cells.
+As a consequence, we cannot know with certainty which cells were perturbed.
+Instead, we must leverage an indirect, noisy proxy of perturbation presence or absence -- namely, transcribed barcode counts -- to ``guess'' which cells were perturbed.
+Using these imputed perturbation assignments, we can attempt to estimate the effect of the perturbation on gene expression.
+The standard approach, which we call the ``thresholding method,'' is to assign perturbation identities to cells by simply thresholding the barcode counts.
 
-We study estimation and inference in single-cell CRISPR screens from a statistical perspective, formulating the data generating mechanism as a new class of errors-in-variables (or measurement error) models.
+We study estimation and inference in single-cell CRISPR screens from a statistical perspective, formulating the data generating mechanism using a new class of errors-in-variables (or measurement error) models.
 We assume that the response variable $y$ is a GLM of an underlying predictor variable $x^*$.
@@
 
-Within this framework we make two contributions.
-First, we study the thresholding method from theoretical and empirical perspectives.
-We characterize the properties of the estimator and uncover settings in which
+Our work makes two main contributions.
+First, we study the thresholding method from empirical and theoretical perspectives.
+Notably, we demonstrate on real data that the thresholding method exhibits a bias-variance tradeoff as a function of the selected threshold, and we recover this phenomenon in precise mathematical terms in an idealized Gaussian model.
+Second, we introduce a new method for estimation and inference in single-cell CRISPR screens that accounts for the measurement error inherent in the experiment.
+The method, called \textit{GLM-EIV} (generalized linear model with errors-in-variables), implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic.
+Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.
 
-% Next, we propose a new method
-
-GLM-EIV (GLM-based errors in variables).
-GLM-EIV implicitly estimates the probability that each cell was perturbed, thereby obviating the need to explicitly impute perturbation identities via thresholding or some other heuristic.
-
-% Overall, we show how coupling novel, principled statistical approaches to current, powerful technologies from genomics and computer science -- namely, CRISPR genome engineering, single-cell sequencing, cloud computing, and high-performance computing -- can drive new biological insights.
+We implement several statistical accelerations to bring the cost of GLM-EIV down to within an order of magnitude of the thresholding method.
+Finally, we develop a computational infrastructure to deploy GLM-EIV at-scale across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters.
+Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, high multiplicity-of-infection single-cell CRISPR screen datasets, yielding new biological and statistical insights.
 
@@
 
-\begin{itemize}
-\item 
-\end{itemize}
+Our focus in this work is on high multiplicity-of-infection (MOI), enhancer-targeting, single-cell CRISPR screens. In this section we cover relevant biological background and motivation.
+\\ \\
+\textbf{P1}: The human genome consists of genes (segments of DNA that code for proteins), enhancers (segments of DNA that regulate the expression of one or more genes), and other genomic regions.
+Genome-wide association studies have revealed that the majority ($>95\%$) of variants associated with diseases lie outside genes and (very likely) inside enhancers.
+These noncoding variants contribute to disease by modulating the expression one or more genes, which in turn encode proteins that affect the phenotype.
+A central open challenge in genetics, therefore, is to link enhancers that harbor disease-associated variants to the genes that they target at genome-wide scale.
+\\ \\
+\textbf{P2}: High MOI single-cell CRISPR screens are the most promising biotechnology for solving this problem.
+\blue{Describe the experimental protocol here.
+Explain that we use the terms ``barcodes'' and ``gRNAs'' interchangeably, as polyadenylated gRNAs serve as barcodes in CROP-seq.
+Link to Figure \ref{analysis_challenges}.}
+\\ \\
+\textbf{P3}: Single-cell CRISPR screens pose several core analysis challenges.
+\blue{Describe the analysis challenges here: (i) unobserved perturbation; (ii) existence of background reads; (iii) highly discrete count data; (iv) nuisance variables}.
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=0.9\linewidth]{analysis_challenges}
+	\caption{\textbf{Overview of experimental design and analysis challenges}: \textbf{a,} Experimental design.
+For a given perturbation (e.g., the perturbation represented in yellow), we partition the cells into two groups: those that received the perturbation, and those that did not receive the perturbation.
+For a given gene, we conduct a differential expression analysis across the two groups of cells, yielding an estimate of the impact of the given perturbation on the given gene.
+\textbf{b,} DAG representing the variables in the analysis.
+The perturbation (unobserved) affects both gene expression and gRNA expression; technical factors (e.g., batch, sequencing depth, etc.) act as nuisance variables.
+The target of inference is the effect of the perturbation on gene expression (denoted with question mark).
+\textbf{c,} Schematic illustrating ``background reads.'' The gRNA modality has a nonzero, ``background read'' distribution even in the absence of a perturbation, complicating the assignment of perturbations to cells.
+\textbf{d}, Example data for a given perturbation-gene pair.
+Notice that (i) the perturbations are unobserved, and (ii) the gene and gRNA expression data take the form of discrete counts.}
+	\label{analysis_challenges}
+\end{figure}
 
 \section{Related work}
+
+Motivated (in part) by the count nature of single-cell data, several authors have generalized statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions.
+Lin et al.\ \cite{Lin2021} developed eSVD, an extension of SVD to exponential family and curved Gaussian response distributions.
+eSVD is better-suited than SVD to model the positive relationship between the mean and variance of a gene's expression level, which is induced by the countedness of the data \cite{Lause2021}.
+Along similar lines, Townes et al.\ \cite{Townes2019} proposed GLM-PCA, an extension of PCA that directly models gene expression counts while regressing out technical factors.
+We see our work as a continuation of this broad effort to ``port'' common statistical methods and models to single-cell count data.
+Our focus, however, is on regression rather than dimension reduction: we extend the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.
+
+The closest parallels to our work the statistical methodology literature are \cite{Grun2008} \cite{Ibrahim1990}.
 

### 2021-10-14 R_scripts/plotting/thresholding_theoretical.R

@@ -36,10 +36,28 @@ pi <- seq(0.0, 0.5, length.out = 11)
 Estimate <- 1 - pi
 to_plot_b <- data.frame(pi = pi, Estimate = Estimate)
 
-# subplot c) variance as c -> infinity
-ggplot(data = to_plot_b, mapping = aes(x = pi, y = Estimate)) + geom_line() +
+p_b <- ggplot(data = to_plot_b, mapping = aes(x = pi, y = Estimate)) + geom_line() +
   scale_x_continuous(trans = "reverse", expand = c(NA, 0), limits = c(0.5, -0.02)) + theme_cowplot(font_size = 11) +
   geom_segment(aes(x=0.0, xend=0.5, y=1, yend=1), lwd = 0.3) + scale_y_continuous(expand = c(0, NA), limits = c(0.5, 1.02)) +
   xlab(expression(pi))
 
-# 
+# subplot c) bias-variance tradeoff
+xgrid <- seq(1, 3.9, length.out = 1001)
+g_beta <- 1
+pi <- 0.25
+m_beta <- 1
+n <- 10000
+
+bias_sq <- (1 - get_thresholding_estimator_est_no_int(c = xgrid, g_beta = g_beta, pi = pi, m_beta = m_beta))^2
+var <- get_thresholding_estimator_var_no_int(c = xgrid, g_beta = g_beta, pi = pi, m_beta = m_beta, n = n)
+to_plot_c <- data.frame(Metric = rep(c("Bias^2", "Variance", "MSE"), each = length(xgrid)),
+                        value = c(bias_sq, var, bias_sq + var),
+                        threshold = rep(xgrid, times = 3))
+p_c <- ggplot(data = to_plot_c, mapping = aes(x = threshold, y = value, col = Metric)) + geom_line() + theme_cowplot(font_size = 11) + xlab("Threshold") + ylab("") +
+  scale_color_manual(values = c("black", my_cols[1], my_cols[2]), labels = c("MSE", expression(Bias^2), "Variance"), breaks = c("MSE", expression(Bias^2), "Variance"))
+
+
+# combine
+p_bottom <- plot_grid(p_b, p_c, align = "h", ncol = 2, rel_widths = c(1,2), labels = c("b", "c"))
+p <- plot_grid(p_a, p_bottom, ncol = 1, labels = c("a", ""))
+

### 2021-10-14 text/outline/outline.tex
finish related works section

@@
 A major difficulty is that CRISPR perturbations are unobservable and assigned stochastically to cells.
-As a consequence, we cannot know with certainty which cells were perturbed.
-Instead, we must leverage an indirect, noisy proxy of perturbation presence or absence -- namely, transcribed barcode counts -- to ``guess'' which cells were perturbed.
-Using these imputed perturbation assignments, we can attempt to estimate the effect of the perturbation on gene expression.
+As a consequence, one cannot know with certainty which cells were perturbed.
+Instead, one must leverage an indirect, noisy proxy of perturbation presence or absence -- namely, transcribed barcode counts -- to ``guess'' which cells were perturbed.
+Using these imputed perturbation assignments, one can attempt to estimate the effect of the perturbation on gene expression.
 The standard approach, which we call the ``thresholding method,'' is to assign perturbation identities to cells by simply thresholding the barcode counts.
@@
 First, we study the thresholding method from empirical and theoretical perspectives.
-Notably, we demonstrate on real data that the thresholding method exhibits a bias-variance tradeoff as a function of the selected threshold, and we recover this phenomenon in precise mathematical terms in an idealized Gaussian model.
-Second, we introduce a new method for estimation and inference in single-cell CRISPR screens that accounts for the measurement error inherent in the experiment.
+Notably, we demonstrate on real data that the thresholding method exhibits a bias-variance tradeoff as a function of the selected threshold, and we recover this phenomenon in precise mathematical terms in an Gaussian model.
+Second, we introduce a new method for estimation and inference in single-cell CRISPR screens that accounts for the measurement error inherent in the data generating process.
 The method, called \textit{GLM-EIV} (generalized linear model with errors-in-variables), implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic.
@@
 
-Motivated (in part) by the count nature of single-cell data, several authors have generalized statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions.
-Lin et al.\ \cite{Lin2021} developed eSVD, an extension of SVD to exponential family and curved Gaussian response distributions.
-eSVD is better-suited than SVD to model the positive relationship between the mean and variance of a gene's expression level, which is induced by the countedness of the data \cite{Lause2021}.
+Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broad class of exponential family distributions.
+For example, Lin, Lei, and Roeder  \cite{Lin2021} developed eSVD, an extension of SVD to exponential family and curved Gaussian response distributions.
+Unlike SVD, eSVD captures the relationship between the mean and variance of a gene's expression level, a phenomenon induced by the countedness and sparsity of single-cell data \cite{Lause2021}.
 Along similar lines, Townes et al.\ \cite{Townes2019} proposed GLM-PCA, an extension of PCA that directly models gene expression counts while regressing out technical factors.
@@
 
-The closest parallels to our work the statistical methodology literature are \cite{Grun2008} \cite{Ibrahim1990}.
+% Several applied methods for linking perturbations to changes in gene expression in single-cell CRISPR screens ; however, these methods focus on hypothesis test (rather than estimation) 
+% SCEPTRE \cite{Barry2020}
+
+The closest parallels to our work in the statistical methodology literature are Gr\"{u}n \& Leisch \cite{Grun2008} and Ibrahim \cite{Ibrahim1990}.
+Gr\"{u}n \& Leisch consider estimation and inference in a $k$-component mixture of GLMs.
+While we prefer to view the GLM-EIV model as a kind of errors-in-variables model, the GLM-EIV model is equivalent to a two-component mixture of \textit{products} of GLM densities.
+Ibrahim derives a method for fitting GLMs in the presence of missing-at-random covariates.
+Our method, by contrast, involves fitting two conditionally independent GLMs in the presence of a totally latent covariate.
+Thus, while Ibrahim is helpful reference, our estimation and inference tasks are more complex than his.
+
+The genomics literature has produced several applied methods for linking perturbations to changes in gene expression in single-cell CRISPR screens: SCEPTRE \cite{Barry2020}, MIMOSCA \cite{Dixit2016}, and scMAGeCK \cite{Yang2019}.
+These methods in general are focused on hypothesis testing rather than estimation; none, for instance, produces a confidence interval for the effect size of a perturbation on change in gene expression.
+Additionally, two of these methods (MIMOSCA and scMAGeCK) use linear models to model gene expressions, disregarding the countedness and sparsity of the data.
 

### 2021-10-17 text/outline/outline.tex
work on thresholding section

@@
 \usepackage{/Users/timbarry/optionFiles/mymacros}
+\DeclareMathOperator*{\argmin}{arg\,min}
 
@@
 \begin{center}
-\textbf{A new class of measurement error models, with application to CRISPR genome editing and single-cell sequencing}
+\textbf{CRISPR genome editing, single-cell sequencing, and exponential family measurement error models}
 \end{center}
+
+\begin{abstract}
+CRISPR genome engineering and single-cell sequencing have transformed biological discovery.
+Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases.
+In this work we study single-cell CRISPR screens from a statistical perspective.
+First, we demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter.
+We recover these phenomena in precise theoretical terms in an idealized Gaussian setting.
+Next, we introduce GLM-EIV (“generalized linear model with errors-in-variables”), a new method for single-cell CRISPR screen analysis.
+GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed, overcoming limitations of thresholded regression.
+We develop a computational infrastructure to deploy GLM-EIV across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters.
+Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, yielding new biological insights.
+\end{abstract}
 
@@
 Using these imputed perturbation assignments, one can attempt to estimate the effect of the perturbation on gene expression.
-The standard approach, which we call the ``thresholding method,'' is to assign perturbation identities to cells by simply thresholding the barcode counts.
+The standard approach, which we call ``thresholded regression'' or the ``thresholding method,'' is to assign perturbation identities to cells by simply thresholding the barcode counts.
 
@@
 First, we study the thresholding method from empirical and theoretical perspectives.
-Notably, we demonstrate on real data that the thresholding method exhibits a bias-variance tradeoff as a function of the selected threshold, and we recover this phenomenon in precise mathematical terms in an Gaussian model.
-Second, we introduce a new method for estimation and inference in single-cell CRISPR screens that accounts for the measurement error inherent in the data generating process.
+Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian model.
+Second, we introduce a new method for estimation and inference in single-cell CRISPR screens that accounts for the measurement error inherent in the experiment.
 The method, called \textit{GLM-EIV} (generalized linear model with errors-in-variables), implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic.
@@
 
-We implement several statistical accelerations to bring the cost of GLM-EIV down to within an order of magnitude of the thresholding method.
+We implement several statistical accelerations (that possibly are of independent utility) to bring the cost of GLM-EIV down to within an order of magnitude of the thresholding method.
 Finally, we develop a computational infrastructure to deploy GLM-EIV at-scale across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters.
@@
 
-Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broad class of exponential family distributions.
-For example, Lin, Lei, and Roeder  \cite{Lin2021} developed eSVD, an extension of SVD to exponential family and curved Gaussian response distributions.
-Unlike SVD, eSVD captures the relationship between the mean and variance of a gene's expression level, a phenomenon induced by the countedness and sparsity of single-cell data \cite{Lause2021}.
-Along similar lines, Townes et al.\ \cite{Townes2019} proposed GLM-PCA, an extension of PCA that directly models gene expression counts while regressing out technical factors.
+Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions.
+For example, Lin, Lei, and Roeder \cite{Lin2021} developed eSVD, an extension of SVD to exponential family and curved Gaussian responses.
+Unlike SVD, eSVD models the relationship between the mean and variance of a gene's expression level, a phenomenon induced by the countedness of single-cell data \cite{Lause2021}.
+Similarly, Townes et al.\ \cite{Townes2019} proposed GLM-PCA, an extension of PCA that directly models Poisson- or negative binomially-distributed gene expression counts.
 We see our work as a continuation of this broad effort to ``port'' common statistical methods and models to single-cell count data.
@@
 The closest parallels to our work in the statistical methodology literature are Gr\"{u}n \& Leisch \cite{Grun2008} and Ibrahim \cite{Ibrahim1990}.
-Gr\"{u}n \& Leisch consider estimation and inference in a $k$-component mixture of GLMs.
-While we prefer to view the GLM-EIV model as a kind of errors-in-variables model, the GLM-EIV model is equivalent to a two-component mixture of \textit{products} of GLM densities.
-Ibrahim derives a method for fitting GLMs in the presence of missing-at-random covariates.
+Gr\"{u}n \& Leisch considered estimation and inference in a $k$-component mixture of GLMs.
+While we prefer to view the GLM-EIV model as an errors-in-variables model,  the GLM-EIV model is equivalent to a two-component mixture of \textit{products} of GLM densities.
+Ibrahim derived a method for fitting GLMs in the presence of missing-at-random covariates.
 Our method, by contrast, involves fitting two conditionally independent GLMs in the presence of a totally latent covariate.
-Thus, while Ibrahim is helpful reference, our estimation and inference tasks are more complex than his.
+Thus, while Ibrahim is a helpful reference, our estimation and inference tasks are more complex.
 
 The genomics literature has produced several applied methods for linking perturbations to changes in gene expression in single-cell CRISPR screens: SCEPTRE \cite{Barry2020}, MIMOSCA \cite{Dixit2016}, and scMAGeCK \cite{Yang2019}.
-These methods in general are focused on hypothesis testing rather than estimation; none, for instance, produces a confidence interval for the effect size of a perturbation on change in gene expression.
-Additionally, two of these methods (MIMOSCA and scMAGeCK) use linear models to model gene expressions, disregarding the countedness and sparsity of the data.
+These methods in general are focused on hypothesis testing rather than estimation; none, for instance, produces a confidence interval for the effect size of a perturbation on gene expression change.
+Additionally, two of these methods (MIMOSCA and scMAGeCK) use (possibly penalized) linear models to model gene expressions, thereby disregarding the countedness and sparsity of the data.
 
@@
 
+In this section we study the thresholding method from empirical and theoretical perspectives. First, we define some notation.
+
+\blue{Let $n \approx 100,000 - 250,000$ be the number of cells in the experiment.
+Consider a single perturbation-gene pair.
+For cell $i \in \{ 1, \dots, n \},$ let $p_i \in \{ 0,1 \}$ indicate whether the cell was perturbed, $m_i \in \N$ be the number of observed gene UMIs, $g_i \in \N$ be the number of observed gRNA UMIs, $l^m_i \in \N$ be the gene library size, and $z_i \in \R^{d-1}$ be a vector of technical factors (e.g., batch, percent mitochondrial reads, etc.).
+The thresholding method is defined as follows:}
+\blue{
 \begin{itemize}
-\item 
-\item 
-\end{itemize}
+\item[1.] For given threshold $c \in \N$, calculate the imputed value $\hat{p}_i$ of $p_i$ by $$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i \geq c, \\ \hat{p}_i = 1 \textrm{ if } g_i < c.
+\end{cases}$$
+\item[2.] Fit the regression model \cite{Sarkar2021}
+$$ m_i | \left(\hat{p}_i, z_i, l^m_i \right) \sim \textrm{NB}_\theta(\mu_i),$$ where $\theta >0$ is the NB size parameter, and $$\log\left(\mu_i\right) = \beta_m^0 + \beta_m \hat{p}_i + \gamma^T_m z_i + \log\left( l_i^m\right).$$
+\item[3.] Fit a GLM to obtain an estimate $\hat{\beta}_m$ of $\beta_m$. Compute a confidence interval and $p$-value for $\beta_m$.
+\end{itemize}}
+
+\subsection{Empirical analysis}
+
+\blue{To investigate the impact of threshold selection on the thresholding method, we applied the thresholding method to the set of positive control (i.e., gene-targeting) perturbation-gene pairs in the Gasperini data using three different choices for the threshold: 1, 5, and 20.}
+
+\blue{Description of panels a-d; threshold = 1 leads to considerable attenuation bias (a); threshold = 5 and threshold = 20 produce similar estimates, though the effect sizes are slightly greater for threshold = 20, suggesting threshold = 5 yields mild attenuation bias (b); the threshold = 20 estimates are more variable than than those of threshold = 5, as evidenced by the smaller $p$-values and wider confidence intervals.}
+
+\blue{Description of panels e-f: these are the empirical gRNA count distributions for randomly selected gRNAs from the Gasperini and Xie datasets.
+There does not appear to be a clear location at which to draw the threshold; aside from the initial spike at one (zero not shown), the histograms gradually decrease.}
+
+\blue{Take-home message: (i) the threshold is a tuning parameter that substantially affects the result; (ii) as the threshold increases, bias seems to decrease and variance seems to increase; (iii) it is not clear where to draw the threshold from gRNA counts alone.}
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=1\linewidth]{../../figures/thresholding_empirical/plot}
+	\caption{\textbf{Empirical challenges of thresholded regression.}}
+	\label{thresholding_empirical}
+\end{figure}
+\newpage
 
@@
 
-\subsection{Empirical analysis}
+\blue{\textbf{Bias}: We use the following theoretical model to study bias of the thresholding method: 
+$$
+\begin{cases}
+m_i = \beta^m_0 + \beta^m_1 p_i + \ep_i \\
+g_i = \beta^g_0 + \beta^g_1 p_i + \tau_i \\
+p_i \sim \textrm{Bern}(\pi) \\
+\ep_i, \tau_i \sim N(0,1) \\
+p_i \indep \tau_i \indep \ep_i
+\end{cases}.
+$$}
+We can solve for the limit in probability of $\hat{\beta}_m$.
+We have that $$\hat{\beta}^m_1 \xrightarrow{P} \beta^m_1 \left(\frac{ \pi( \omega - \E[ \hat{p}_i ])}{ \E[\hat{p}_i] (1 - \E[\hat{p}_i])}\right),$$ where
+$$\begin{cases}
+\omega = \pi\left( 1 - \Phi(c - \beta_1^g - \beta_0^g) \right),\\ \zeta = (1-\pi)(1 - \Phi\left( c - \beta^g_0 \right)), \\ \E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi. 
+\end{cases}$$
+Let $\gamma: \R^4 \to \R$ be defined by
+$$ \gamma( \beta^g_0, \beta^g_1, c, \pi ) = \frac{\pi (\omega - \E[\hat{p}_i])}{\E[\hat{p}_i] (1 - \E[\hat{p}_i]) }.$$ We call $\gamma$ the attenuation function.
+Observe that (i) $\gamma$ does not depend on $\beta^m_1$ or $\beta^m_0$, and (ii) $ \hat{\beta}^m_1 \xrightarrow{P} [\gamma(\beta_0^g, \beta_1^g, c, \pi)] \beta^m_1.$ Let $b: \R^4 \to \R$ be the asymptotic bias of $\hat{\beta}^m_1$.
+We can express $b$ in terms of $\gamma$ as $$b\left(\beta^g_0, \beta^g_1, c, \pi \right) = \beta^m_1 - \gamma( \beta_0^g, \beta_1^g, c, \pi) \beta^m_1 = \beta^m_1 \left[ 1- \gamma(\beta^g_0, \beta^g_1, c, \pi) \right].$$ The asymptotic bias vanishes when the attenuation function equals $1$.
+
+\begin{center}
+\textbf{Bias as a function of threshold (Panel a)}
+\end{center}
+
+To investigate the basic question of ``What is a good threshold selection strategy?'', we studied the relationship between the asymptotic bias $b$ of $\hat{\beta}^m_1$ and the selected threshold $c$.
+For simplicity, we set the perturbation probability $\pi$ to $1/2$.
+Additionally, without loss of generality, we set the target of inference $\beta^m_1$ to $1$.
+Let $c_\textrm{bayes} \in \R$ be the Bayes-optimal decision boundary for classifying cells as perturbed or unperturbed, i.e.
+$$c_\textrm{bayes} = \argmin_{c \in \R} \P(\hat{p}_i \neq p_i).$$ Simple algebra shows that $c_\textrm{bayes} = \beta_0^g + (1/2) \beta^g_1.$ Below, we state several results relating the limiting bias $b$ of $\hat{\beta}^m_1$ to $c_\textrm{bayes}$ and $\beta^g_1$, deferring proofs to the appendix.
+
+\begin{itemize}
+\item[1.] The Bayes-optimal threshold $c_\textrm{bayes}$ is a critical value of the bias function $b$, i.e. 
+$$ \frac{\partial b( \beta^g_0, \beta^g_1, c, 1/2)}{\partial c} \vline_{c = c_\textrm{bayes}} = 0.$$
+\item[2.] For certain values of $\beta^g_0$ and $\beta^g_1$, $c_\textrm{bayes}$ is a maximizer of the bias (Figure \ref{thresholding_theoretical}a, left); for other values of $\beta^g_0$ and $\beta^g_1$, $c_\textrm{bayes}$ is a minimizer of the bias (Figure \ref{thresholding_theoretical}a, right).
+\item[3.] The limit of the bias in $c$ is $1/2$, i.e. 
+$$\lim_{c \to \infty} b(\beta^g_0, \beta^g_1, c, 1/2) = 1/2.$$ In other words, we always can set the threshold to a large number and achieve a bias of $1/2$ (Figure \ref{thresholding_theoretical}a, all panels).
+This fact establishes an upper bound on the bias of thresholded regression (under optimal threshold selection).
+\item[4.] For $\beta^g_1 < 2 \Phi(3/4) \approx 1.35$, $c = c_\textrm{bayes}$ yields a \textit{larger} bias than $c = \infty$ (Figure \ref{thresholding_theoretical}a, left); for $\beta^g_1 > 2 \Phi(3/4)$, $c_\textrm{bayes}$ yields a \textit{smaller} bias than $c = \infty$ (Figure \ref{thresholding_theoretical}a, right); finally, for $\beta^g_1 = 2 \Phi(3/4)$, $c = c_\textrm{bayes}$ and $c = \infty$ both yield a bias of $1/2$ (Figure \ref{thresholding_theoretical}a, middle).
+\item[5.] The bias decreases monotonically as $\beta_1^g$ increases (Figure \ref{thresholding_theoretical}a; fix a threshold and scan the panels from left to right).
+This is consistent with the intuition that the problem becomes easier as the gRNA mixture distribution becomes increasingly well-separated.
+\item[6.] For all $\beta^g_0, \beta^g_1,$ and $c \in \R,$ the bias is strictly greater than $0$ and less than $2$.
+Therefore, the thresholding method suffers from attenuation bias, a common phenomenon in errors-in-variables models.
+\end{itemize}
+
+These results are subtle, but we can distill them as follows.
+First, selecting a threshold that minimizes the bias is deceptively challenging, as there is no rule of thumb that we can apply universally for this purpose (e.g., ``always choose the Bayes-optimal decision boundary'' or ``always choose a large number'').
+Furthermore, the optimal threshold depends on unknown parameters.
+Finally, even \textit{if} we have selected a good threshold, the thresholding method incurs a strict attenuation bias.
+
+\begin{center}
+\textbf{Bias of large threshold strategy vs.\ $\pi$ (Panel b)}
+\end{center}
+
+Intrigued by the finding that selecting a large threshold when $\pi = 1/2$ yields a bias of $1/2$, we investigated the relationship between bias and other values of $\pi$ under the large threshold selection strategy.
+
+\begin{center}
+	\textbf{Bias-variance tradeoff (Panel c)}
+\end{center}
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=1\linewidth]{../../figures/thresholding_theoretical/plot}
+	\caption{\textbf{Theoretical challenges of thresholded regression.}}
+	\label{thresholding_theoretical}
+\end{figure}
+
+% These difficulties motivate our core research question: does bypassing thresholding altogether lead to better estimation and inference in single-cell CRISPR screens?
+To answer this question, we propose GLM-EIV, method for single-cell CRISPR screen analysis that models gRNA counts directly.
 
@@
 
-% \subsection{Dictionary of symbols}
-
-\subsection{Proofs of theoretical results for thresholding estimator}
+\subsection{Theoretical results for thresholding estimator}
 
@@
 \bibliographystyle{unsrt}
+\newpage
 \bibliography{/Users/timbarry/optionFiles/glmeiv.bib}

### 2021-10-17 R_scripts/plotting/thresholding_empirical.R

@@ -2,6 +2,10 @@ library(magrittr)
 library(ggplot2)
 library(cowplot)
 
+my_cols <- c("firebrick3", "dodgerblue3", "orchid4", "lightskyblue3")
+fig_dir <- paste0(.get_config_path("LOCAL_CODE_DIR"), "glmeiv-manuscript/figures/thresholding_empirical")
+if (!dir.exists(fig_dir)) dir.create(fig_dir)
+
 # load the results
 gasp_result_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/gasperini/results/")
 thresh_res <- readRDS(paste0(gasp_result_dir, "result_thresholding.rds"))
@@ -17,21 +21,18 @@ thresh_res_pc_wide <- thresh_res_pc %>% dplyr::select(target, value, threshold, 
 
 to_plot <- thresh_res_pc_wide %>% dplyr::filter(target == "estimate") %>% dplyr::select(threshold_1, threshold_5)
 p1 <- ggplot(data = to_plot, mapping = aes(x = threshold_5, y = threshold_1)) + geom_point(alpha = 0.7, col = my_cols[2]) +
-  xlab("Threshold = 5 (est)") + ylab("Threshold = 1 (est)") + geom_abline(slope = 1, intercept = 0) + theme_bw(base_size = 10)
+  xlab("Thresh = 5 (est)") + ylab("Thresh = 1 (est)") + geom_abline(slope = 1, intercept = 0) + theme_bw(base_size = 10)
 # threshold = 1 ests are bigger than (i.e., closer to one than) threshold = 5; this is attenuation bias due to the fact that excessively many unperturbed cells have been classified as perturbed.
-
-my_cols <- c("firebrick3", "dodgerblue3", "orchid4")
-# Question 2. How do CI widths and p-values change for thresh = 5 vs thresh = 20?
 
 # a) est. vs. est.
 to_plot <- thresh_res_pc_wide %>% dplyr::filter(target == "estimate") %>% dplyr::select(threshold_5, threshold_20)
 p2 <- ggplot(data = to_plot, mapping = aes(x = threshold_5, y = threshold_20)) + geom_point(alpha = 0.9, col = my_cols[2]) +
[…]

### 2021-10-17 R_scripts/plotting/thresholding_theoretical.R

@@ -4,6 +4,9 @@ library(magrittr)
 library(ggplot2)
 library(cowplot)
 
+fig_dir <- paste0(.get_config_path("LOCAL_CODE_DIR"), "glmeiv-manuscript/figures/thresholding_theoretical")
+if (!dir.exists(fig_dir)) dir.create(fig_dir)
+
 # a) estimate vs. threshold for different values of beta_g
 x_grid <- seq(0, 7, length.out = 501)
 beta_gs <- c(1, qnorm(0.75) * 2, 1.5)
@@ -53,11 +56,13 @@ var <- get_thresholding_estimator_var_no_int(c = xgrid, g_beta = g_beta, pi = pi
 to_plot_c <- data.frame(Metric = rep(c("Bias^2", "Variance", "MSE"), each = length(xgrid)),
                         value = c(bias_sq, var, bias_sq + var),
                         threshold = rep(xgrid, times = 3))
-p_c <- ggplot(data = to_plot_c, mapping = aes(x = threshold, y = value, col = Metric)) + geom_line() + theme_cowplot(font_size = 11) + xlab("Threshold") + ylab("") +
+p_c <- ggplot(data = to_plot_c %>% dplyr::mutate(Metric = factor(Metric, levels = c("Bias^2", "Variance", "MSE"),
+                                                                 labels = c("Bias^2", "Variance", "MSE"))), mapping = aes(x = threshold, y = value, col = Metric)) + geom_line() + theme_cowplot(font_size = 10) + xlab("Threshold") + ylab("") +
   scale_color_manual(values = c("black", my_cols[1], my_cols[2]), labels = c("MSE", expression(Bias^2), "Variance"), breaks = c("MSE", expression(Bias^2), "Variance"))
-
 
 # combine
 p_bottom <- plot_grid(p_b, p_c, align = "h", ncol = 2, rel_widths = c(1,2), labels = c("b", "c"))
 p <- plot_grid(p_a, p_bottom, ncol = 1, labels = c("a", ""))
 
+# save
+ggsave(filename = paste0(fig_dir, "/plot.jpeg"), plot = p, device = "jpeg", scale = 1.25, width = 6, height = 4)

### 2021-10-18 text/outline/outline.tex
update text in theoretical section

@@
 
-We implement several statistical accelerations (that possibly are of independent utility) to bring the cost of GLM-EIV down to within an order of magnitude of the thresholding method.
+We implement several statistical accelerations (that likely are of independent utility) to bring the cost of GLM-EIV down to within an order of magnitude of the thresholding method.
 Finally, we develop a computational infrastructure to deploy GLM-EIV at-scale across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters.
@@
 
-\blue{\textbf{Bias}: We use the following theoretical model to study bias of the thresholding method: 
-$$
+We study the thresholding method from a theoretical perspective in an idealized Gaussian setting. Suppose we observe data $(g_1, m_1), \dots, (g_n, m_n)$ from the following model:
+
+\begin{equation}\label{theoretical_model}
 \begin{cases}
@@
 \ep_i, \tau_i \sim N(0,1) \\
-p_i \indep \tau_i \indep \ep_i
-\end{cases}.
-$$}
-We can solve for the limit in probability of $\hat{\beta}_m$.
-We have that $$\hat{\beta}^m_1 \xrightarrow{P} \beta^m_1 \left(\frac{ \pi( \omega - \E[ \hat{p}_i ])}{ \E[\hat{p}_i] (1 - \E[\hat{p}_i])}\right),$$ where
+p_i \indep \tau_i \indep \ep_i.
+\end{cases}
+\end{equation}
+For a given threshold $c \in \R$, let $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_0$ for $\beta^m_0$ is given by $$\hat{\beta}^m_0 = \frac{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}}) (m_i - \overline{m})}{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})}.$$ The limit in probability (as $n \to \infty$) of $\hat{\beta}^m_0$ is
+$$\hat{\beta}^m_1 \xrightarrow{P} \beta^m_1 \left(\frac{ \pi( \omega - \E[ \hat{p}_i ])}{ \E[\hat{p}_i] (1 - \E[\hat{p}_i])}\right),$$ where
 $$\begin{cases}
@@
 Let $\gamma: \R^4 \to \R$ be defined by
-$$ \gamma( \beta^g_0, \beta^g_1, c, \pi ) = \frac{\pi (\omega - \E[\hat{p}_i])}{\E[\hat{p}_i] (1 - \E[\hat{p}_i]) }.$$ We call $\gamma$ the attenuation function.
-Observe that (i) $\gamma$ does not depend on $\beta^m_1$ or $\beta^m_0$, and (ii) $ \hat{\beta}^m_1 \xrightarrow{P} [\gamma(\beta_0^g, \beta_1^g, c, \pi)] \beta^m_1.$ Let $b: \R^4 \to \R$ be the asymptotic bias of $\hat{\beta}^m_1$.
+$$ \gamma( \beta^g_0, \beta^g_1, c, \pi ) = \frac{\pi (\omega - \E[\hat{p}_i])}{\E[\hat{p}_i] (1 - \E[\hat{p}_i]) }.$$ We call $\gamma$ the \textit{attenuation function}. Observe that \begin{itemize}
+\item[i.] $\gamma$ does not depend on $\beta^m_1$ or $\beta^m_0$, and
+\item[ii.] $\hat{\beta}^m_1 \xrightarrow{P} [\gamma(\beta_0^g, \beta_1^g, c, \pi)] \beta^m_1.$
+\end{itemize}
+Let $b: \R^4 \to \R$ be the asymptotic bias of $\hat{\beta}^m_1$.
 We can express $b$ in terms of $\gamma$ as $$b\left(\beta^g_0, \beta^g_1, c, \pi \right) = \beta^m_1 - \gamma( \beta_0^g, \beta_1^g, c, \pi) \beta^m_1 = \beta^m_1 \left[ 1- \gamma(\beta^g_0, \beta^g_1, c, \pi) \right].$$ The asymptotic bias vanishes when the attenuation function equals $1$.
@@
 
-To investigate the basic question of ``What is a good threshold selection strategy?'', we studied the relationship between the asymptotic bias $b$ of $\hat{\beta}^m_1$ and the selected threshold $c$.
+To investigate the basic question of ``What is a good threshold selection strategy?'', we study the relationship between the asymptotic bias $b$ of $\hat{\beta}^m_1$ and the selected threshold $c$.
 For simplicity, we set the perturbation probability $\pi$ to $1/2$.
@@
 Let $c_\textrm{bayes} \in \R$ be the Bayes-optimal decision boundary for classifying cells as perturbed or unperturbed, i.e.
-$$c_\textrm{bayes} = \argmin_{c \in \R} \P(\hat{p}_i \neq p_i).$$ Simple algebra shows that $c_\textrm{bayes} = \beta_0^g + (1/2) \beta^g_1.$ Below, we state several results relating the limiting bias $b$ of $\hat{\beta}^m_1$ to $c_\textrm{bayes}$ and $\beta^g_1$, deferring proofs to the appendix.
+$$c_\textrm{bayes} = \argmin_{c \in \R} \P(\hat{p}_i \neq p_i).$$ Simple algebra shows that $c_\textrm{bayes} = \beta_0^g + (1/2) \beta^g_1.$ Below, we state several results relating the limiting bias $b$ of $\hat{\beta}^m_1$ to $c_\textrm{bayes}$ and $\beta^g_1$.
+We defer proofs to the appendix.
 
 \begin{itemize}
-\item[1.] The Bayes-optimal threshold $c_\textrm{bayes}$ is a critical value of the bias function $b$, i.e. 
+\item[1.] The Bayes-optimal threshold $c_\textrm{bayes}$ is a critical value of the bias function $b$ (Figure \ref{thresholding_theoretical}a, all panels), i.e. 
 $$ \frac{\partial b( \beta^g_0, \beta^g_1, c, 1/2)}{\partial c} \vline_{c = c_\textrm{bayes}} = 0.$$
@@
 \item[4.] For $\beta^g_1 < 2 \Phi(3/4) \approx 1.35$, $c = c_\textrm{bayes}$ yields a \textit{larger} bias than $c = \infty$ (Figure \ref{thresholding_theoretical}a, left); for $\beta^g_1 > 2 \Phi(3/4)$, $c_\textrm{bayes}$ yields a \textit{smaller} bias than $c = \infty$ (Figure \ref{thresholding_theoretical}a, right); finally, for $\beta^g_1 = 2 \Phi(3/4)$, $c = c_\textrm{bayes}$ and $c = \infty$ both yield a bias of $1/2$ (Figure \ref{thresholding_theoretical}a, middle).
-\item[5.] The bias decreases monotonically as $\beta_1^g$ increases (Figure \ref{thresholding_theoretical}a; fix a threshold and scan the panels from left to right).
+\item[5.] The bias decreases monotonically as $\beta_1^g$ increases (Figure \ref{thresholding_theoretical}a; fix threshold and scan panels from left to right).
 This is consistent with the intuition that the problem becomes easier as the gRNA mixture distribution becomes increasingly well-separated.
-\item[6.] For all $\beta^g_0, \beta^g_1,$ and $c \in \R,$ the bias is strictly greater than $0$ and less than $2$.
-Therefore, the thresholding method suffers from attenuation bias, a common phenomenon in errors-in-variables models.
+\item[6.] For all $\beta^g_0, \beta^g_1,$ and $c \in \R,$ the bias is strictly greater than $0$ and less than $2$ (Figure \ref{thresholding_theoretical}a, all panels).
+Therefore, the thresholding method suffers from attenuation bias.% a common phenomenon in errors-in-variables models.
 \end{itemize}
 
-These results are subtle, but we can distill them as follows.
-First, selecting a threshold that minimizes the bias is deceptively challenging, as there is no rule of thumb that we can apply universally for this purpose (e.g., ``always choose the Bayes-optimal decision boundary'' or ``always choose a large number'').
-Furthermore, the optimal threshold depends on unknown parameters.
-Finally, even \textit{if} we have selected a good threshold, the thresholding method incurs a strict attenuation bias.
+These results are subtle, but we can summarize them as follows.
+First, selecting a threshold that minimizes the bias is deceptively challenging, as there is no rule of thumb that we can apply universally (e.g., ``always choose the Bayes-optimal decision boundary'' or ``always choose a large number'').
+Second, even if we \textit{have} selected a good threshold, the thresholding method incurs a strict attenuation bias.
 
 \begin{center}
-\textbf{Bias of large threshold strategy vs.\ $\pi$ (Panel b)}
+\textbf{Generalizing to $\pi \in [0,1/2]$ (Panel b)}
 \end{center}
 
-Intrigued by the finding that selecting a large threshold when $\pi = 1/2$ yields a bias of $1/2$, we investigated the relationship between bias and other values of $\pi$ under the large threshold selection strategy.
+Next, we generalize the expression for bias to arbitrary $\pi \in [0,1/2]$. The following simple relationship holds:
+$$ \lim_{ c \to \infty } b(\beta^g_0, \beta^g_1, c, \pi) = \pi.$$ In other words, when the perturbation probability is equal to $\pi$, setting the threshold to a large number yields a bias of $\pi$ (Figure \ref{thresholding_theoretical}b).
+We can build intuition for this result by considering an extreme example.
+When $\pi$ is very small (e.g., $\pi = 0.01$), almost all cells are unperturbed.
+Therefore, in selecting a large threshold, one correctly classifies nearly all \textit{unperturbed} cells as \textit{unperturbed}; on the other hand, the \textit{perturbed} cells that one misclassifies as \textit{unperturbed} are swamped in number by the truly \textit{unperturbed} cells, resulting in a small bias.
 
@@
 \end{center}
+
+Finally, we study the variance of the thresholding estimator. Let $v$ denote the limiting variance of scaled estimator $\sqrt{n} \hat{\beta}^m_1$, i.e.,
+$$v(\beta^g_0, \beta^g_1, c, \pi) = \lim_{n \to \infty} \V(\sqrt{n} \hat{\beta}^m_1).$$ The limiting variance of $\sqrt{n} \hat{\beta}^m_1$ grows without bound as the threshold $c$ tends to infinity: 
+$$ \lim_{c \to \infty} v(\beta^g_0, \beta^g_1, c, \pi) = \infty.$$ Thus, while selecting a large threshold is effective from a bias reduction standpoint when $\pi$ is small, it causes the variance to explode.
+
+To more precisely characterize the bias-variance tradeoff of the thresholding method, we consider a slightly simpler model: we set $\beta^m_0$ and $\beta^g_0$ to zero in (\ref{theoretical_model}) and study the no-intercept least squares estimator
+$$ \hat{\beta}^m_1 =  \frac{ \sum_{i=1}^n \hat{p}_i m_i }{ \sum_{i=1}^n \hat{p}_i^2}.$$ We uncover an exact bias-variance tradeoff as a function of the selected threshold (Figure \ref{thresholding_theoretical}c).
+The bias-variance analysis reveals that the best strategy for minimizing mean-squared error is to select a threshold that induces moderate bias.
+(A downside of this approach, of course, is that constructing valid confidence intervals becomes more challenging.)
 

### 2021-10-18 text/outline/outline.tex
finish section 9.1.1

@@
 
-% Several applied methods for linking perturbations to changes in gene expression in single-cell CRISPR screens ; however, these methods focus on hypothesis test (rather than estimation) 
-% SCEPTRE \cite{Barry2020}
-
 The closest parallels to our work in the statistical methodology literature are Gr\"{u}n \& Leisch \cite{Grun2008} and Ibrahim \cite{Ibrahim1990}.
@@
 
-We study the thresholding method from a theoretical perspective in an idealized Gaussian setting. Suppose we observe data $(g_1, m_1), \dots, (g_n, m_n)$ from the following model:
+We study the thresholding method from a theoretical perspective in an idealized Gaussian setting. Suppose that we observe data $(g_1, m_1), \dots, (g_n, m_n)$ from the following model:
 
@@
 \end{equation}
-For a given threshold $c \in \R$, let $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_0$ for $\beta^m_0$ is given by $$\hat{\beta}^m_0 = \frac{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}}) (m_i - \overline{m})}{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})}.$$ The limit in probability (as $n \to \infty$) of $\hat{\beta}^m_0$ is
-$$\hat{\beta}^m_1 \xrightarrow{P} \beta^m_1 \left(\frac{ \pi( \omega - \E[ \hat{p}_i ])}{ \E[\hat{p}_i] (1 - \E[\hat{p}_i])}\right),$$ where
+For a given threshold $c \in \R$, let $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ for $\beta^m_1$ is given by $$\hat{\beta}^m_1 = \frac{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}}) (m_i - \overline{m})}{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})}.$$ The limit in probability (as $n \to \infty$) of $\hat{\beta}^m_0$ is
+\begin{equation}\label{thresh_est_intercepts}
+\hat{\beta}^m_1 \xrightarrow{P} \beta^m_1 \left(\frac{ \pi( \omega - \E[ \hat{p}_i ])}{ \E[\hat{p}_i] (1 - \E[\hat{p}_i])}\right),
+\end{equation}
+where
 $$\begin{cases}
-\omega = \pi\left( 1 - \Phi(c - \beta_1^g - \beta_0^g) \right),\\ \zeta = (1-\pi)(1 - \Phi\left( c - \beta^g_0 \right)), \\ \E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi. 
+\omega = 1 - \Phi\left(c - \beta_1^g - \beta_0^g \right) ,\\ \zeta = 1 - \Phi\left( c - \beta^g_0 \right), \\ \E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi. 
 \end{cases}$$
@@
 
-Next, we generalize the expression for bias to arbitrary $\pi \in [0,1/2]$. The following simple relationship holds:
+Next, we generalize the expression for bias when the threshold is large to arbitrary $\pi \in [0,1/2]$. The following simple relationship holds:
 $$ \lim_{ c \to \infty } b(\beta^g_0, \beta^g_1, c, \pi) = \pi.$$ In other words, when the perturbation probability is equal to $\pi$, setting the threshold to a large number yields a bias of $\pi$ (Figure \ref{thresholding_theoretical}b).
@@
 When $\pi$ is very small (e.g., $\pi = 0.01$), almost all cells are unperturbed.
-Therefore, in selecting a large threshold, one correctly classifies nearly all \textit{unperturbed} cells as \textit{unperturbed}; on the other hand, the \textit{perturbed} cells that one misclassifies as \textit{unperturbed} are swamped in number by the truly \textit{unperturbed} cells, resulting in a small bias.
+Therefore, in selecting a large threshold, one correctly classifies nearly all unperturbed cells as unperturbed; on the other hand, the \textit{perturbed} cells that one misclassifies as \textit{unperturbed} are swamped in number by the truly unperturbed cells, resulting in a small bias.
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=1\linewidth]{../../figures/thresholding_theoretical/plot}
+	\caption{\textbf{Theoretical challenges of thresholded regression.}}
+	\label{thresholding_theoretical}
+\end{figure}
 
@@
 
-Finally, we study the variance of the thresholding estimator. Let $v$ denote the limiting variance of scaled estimator $\sqrt{n} \hat{\beta}^m_1$, i.e.,
+Finally, to shed light on limitations of the large threshold selection strategy, we study the variance of the thresholding estimator.
+Let $v$ denote the limiting variance of scaled estimator $\sqrt{n} \hat{\beta}^m_1$, i.e.,
 $$v(\beta^g_0, \beta^g_1, c, \pi) = \lim_{n \to \infty} \V(\sqrt{n} \hat{\beta}^m_1).$$ The limiting variance of $\sqrt{n} \hat{\beta}^m_1$ grows without bound as the threshold $c$ tends to infinity: 
@@
 (A downside of this approach, of course, is that constructing valid confidence intervals becomes more challenging.)
-
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=1\linewidth]{../../figures/thresholding_theoretical/plot}
-	\caption{\textbf{Theoretical challenges of thresholded regression.}}
-	\label{thresholding_theoretical}
-\end{figure}
 
@@
 
-\subsection{Theoretical results for thresholding estimator}
+\subsection{Theoretical details for thresholding estimator}
+
+This section contains proofs for results presented the section ``Theoretical analysis of thresholding estimator.''
+
+\subsubsection{Limit of $\hat{\beta}^m_1$ in Gaussian model with intercepts}
+
+Consider the thresholding estimator $\hat{\beta}^m_1$ for the Gaussian model with intercepts (\ref{theoretical_model}).
+Dividing by $n$ in (\ref{thresh_est_intercepts}), we can express $\hat{\beta}^m_1$ as
+$$ \hat{\beta}^m_1 = \frac{ \frac{1}{n} \sum_{i=1}^n ( \hat{p}_i - \overline{\hat{p}_i})(m_i - \overline{m})}{ \frac{1}{n} \sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})}.$$ By WLLN,
+$$ \hat{\beta}^m_1 \xrightarrow{P} \frac{\textrm{Cov}(\hat{p}_i, m_i)}{\V\left(\hat{p}_i\right)}.$$ To compute this quantity, we first compute several simpler quantities:
+\begin{itemize}
+\item[1.] \textbf{Expectation of $m_i$}: $\E[m_i] = \beta^m_0 + \beta^m_1\pi.$
+\item[2.] \textbf{Expectation of $\hat{p}_i$}: Let $f$ denote the $N(0,1)$ density. Then \begin{multline*}
+\E[\hat{p}_i] = \P\left[\hat{p}_i = 1\right] = \P\left[\beta^g_0 + \beta^g_1 + \tau_i \geq c \right] = \\ \textrm{(By LOTP) } \P\left[ \beta^g_0 + \tau_i \geq c \right]\P\left[p_i = 0\right] + \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \P[p_i = 1] \\ = \P\left[ \tau_i \geq c - \beta^g_0\right](1- \pi) + \P\left[ \tau_i \geq c - \beta^g_1 - \beta^g_0 \right](\pi) \\ = \left(\int_{c - \beta^g_0}^\infty f\right) (1 - \pi) + \left( \int_{c - \beta^g_1 - \beta^g_0}^\infty f \right)(\pi) = \zeta(1-\pi) + \omega \pi,
+\end{multline*}
+where $$\zeta = \int_{c - \beta_0^g}^\infty f = 1 - \Phi(c - \beta^g_0)$$ and 
+$$ \omega = \int_{c - \beta^g_1 - \beta^g_0} f = 1 - \Phi(c - \beta^g_1 - \beta^g_0).$$
+\item[3.] \textbf{Expectation of $\hat{p}_i p_i$}: 
+$$\E\left[ \hat{p}_i p_i \right] = \E\left[ \hat{p}_i | p_i = 1 \right] \P\left[ p_i =1 \right] = \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \pi = \omega \pi.$$
+\item[4.] \textbf{Expectation of $\hat{p}_i m_i$}:
+\begin{multline*}
+\E\left[\hat{p}_i m_i\right] = \E[\hat{p}_i (\beta^m_0 + \beta^m_1 p_i + \ep_i)] = \beta^m_0 \E\left[\hat{p}_i\right] + \beta^m_1 \E\left[\hat{p}_i p_i\right] + \E[\hat{p}_i \ep_i] \\ = \beta^m_0 \E[\hat{p}_i] + \beta^m_1 \omega \pi + \E[\hat{p}_i] \E[\ep_i] = \beta^m_0 \E[\hat{p}_i] + \beta^m_1 \omega \pi.
+\end{multline*}
+\item[5.] \textbf{Variance of $\hat{p}_i$}: Because $\hat{p}_i$ is binary, we have that $\V[\hat{p}_i] = \E[\hat{p}_i]\left(1 - \E[\hat{p}_i]\right) .$
+\item[6.] \textbf{Covariance of $\hat{p}_i, m_i$}:
+\begin{multline*}
+\textrm{Cov}\left(\hat{p}_i, m_i\right) = \E\left[\hat{p}_i m_i\right] - \E[\hat{p}_i] \E[m_i] = \beta^m_0 \E[\hat{p}_i] + \beta^m_1 \omega \pi - \E[\hat{p}_i]( \beta^m_0 + \beta^m_1 \pi)\\ = \beta^m_1 \omega \pi - \E[\hat{p}_i] \beta_1^m \pi = \beta^m_1 \pi \left( \omega - \E[\hat{p}_i]\right).
+\end{multline*}
+Combining these expressions, we have that
+$$ \hat{\beta}^m_1 \xrightarrow{P} \frac{\beta^m_1 \pi (\omega - \E[\hat{p}_i])}{\E[\hat{p}_i](1 - \E[\hat{p}_i])} = \beta^m_1 \gamma(\beta^g_0, \beta^g_1, c, \pi).$$
+\end{itemize}
+\subsubsection{Re-expressing $\gamma$ when $\pi = 1/2$}
+
+Suppose $\pi = 1/2$. We rewrite the attenuation fraction $\gamma$ in a way that makes it more amenable to theoretical analysis.
 

### 2021-10-18 text/outline/outline.tex
compute limit in beta

@@
 Dividing by $n$ in (\ref{thresh_est_intercepts}), we can express $\hat{\beta}^m_1$ as
-$$ \hat{\beta}^m_1 = \frac{ \frac{1}{n} \sum_{i=1}^n ( \hat{p}_i - \overline{\hat{p}_i})(m_i - \overline{m})}{ \frac{1}{n} \sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})}.$$ By WLLN,
+$$ \hat{\beta}^m_1 = \frac{ \frac{1}{n} \sum_{i=1}^n ( \hat{p}_i - \overline{\hat{p}_i})(m_i - \overline{m})}{ \frac{1}{n} \sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})}.$$ By weak LLN,
 $$ \hat{\beta}^m_1 \xrightarrow{P} \frac{\textrm{Cov}(\hat{p}_i, m_i)}{\V\left(\hat{p}_i\right)}.$$ To compute this quantity, we first compute several simpler quantities:
 \begin{itemize}
-\item[1.] \textbf{Expectation of $m_i$}: $\E[m_i] = \beta^m_0 + \beta^m_1\pi.$
-\item[2.] \textbf{Expectation of $\hat{p}_i$}: Let $f$ denote the $N(0,1)$ density. Then \begin{multline*}
+\item[1.] Expectation of $m_i$: $\E[m_i] = \beta^m_0 + \beta^m_1\pi.$
+\item[2.] Expectation of $\hat{p}_i$: Let $f$ denote the $N(0,1)$ density. Then \begin{multline*}
 \E[\hat{p}_i] = \P\left[\hat{p}_i = 1\right] = \P\left[\beta^g_0 + \beta^g_1 + \tau_i \geq c \right] = \\ \textrm{(By LOTP) } \P\left[ \beta^g_0 + \tau_i \geq c \right]\P\left[p_i = 0\right] + \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \P[p_i = 1] \\ = \P\left[ \tau_i \geq c - \beta^g_0\right](1- \pi) + \P\left[ \tau_i \geq c - \beta^g_1 - \beta^g_0 \right](\pi) \\ = \left(\int_{c - \beta^g_0}^\infty f\right) (1 - \pi) + \left( \int_{c - \beta^g_1 - \beta^g_0}^\infty f \right)(\pi) = \zeta(1-\pi) + \omega \pi,
@@
 where $$\zeta = \int_{c - \beta_0^g}^\infty f = 1 - \Phi(c - \beta^g_0)$$ and 
-$$ \omega = \int_{c - \beta^g_1 - \beta^g_0} f = 1 - \Phi(c - \beta^g_1 - \beta^g_0).$$
-\item[3.] \textbf{Expectation of $\hat{p}_i p_i$}: 
+$$ \omega = \int_{c - \beta^g_1 - \beta^g_0}^\infty f = 1 - \Phi(c - \beta^g_1 - \beta^g_0).$$
+\item[3.] Expectation of $\hat{p}_i p_i$: 
 $$\E\left[ \hat{p}_i p_i \right] = \E\left[ \hat{p}_i | p_i = 1 \right] \P\left[ p_i =1 \right] = \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \pi = \omega \pi.$$
-\item[4.] \textbf{Expectation of $\hat{p}_i m_i$}:
+\item[4.] Expectation of $\hat{p}_i m_i$:
 \begin{multline*}
@@
 \end{multline*}
-\item[5.] \textbf{Variance of $\hat{p}_i$}: Because $\hat{p}_i$ is binary, we have that $\V[\hat{p}_i] = \E[\hat{p}_i]\left(1 - \E[\hat{p}_i]\right) .$
-\item[6.] \textbf{Covariance of $\hat{p}_i, m_i$}:
+\item[5.] Variance of $\hat{p}_i$: Because $\hat{p}_i$ is binary, we have that $\V[\hat{p}_i] = \E[\hat{p}_i]\left(1 - \E[\hat{p}_i]\right) .$
+\item[6.] Covariance of $\hat{p}_i, m_i$:
 \begin{multline*}
@@
 
-Suppose $\pi = 1/2$. We rewrite the attenuation fraction $\gamma$ in a way that makes it more amenable to theoretical analysis.
+Suppose $\pi = 1/2$. We rewrite the attenuation fraction $\gamma$ in a way that makes it more amenable to theoretical analysis. We leverage the fact that $f$ integrates to unity and is even. First,
+\begin{multline*} \E\left[\hat{p}_i\right] = (1/2) \left( \int_{c - \beta^g_0}^\infty f \right) + (1/2) \left( \int_{c - \beta_0^g - \beta_1^g}^\infty f \right) \\ = (1/2) \left( \int_{-\infty}^{ \beta_0^g - c} f \right) + (1/2) \left( \int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right).
+\end{multline*} and so \begin{multline*} 1 - \E\left[\hat{p}_i\right] = (1/2) + (1/2) - \E[\hat{p}_i]  = (1/2) \left(1 - \int_{c - \beta_0^g}^\infty f \right)  + (1/2) \left( 1 - \int_{c - \beta^g_0 - \beta^g_1}^\infty f \right) \\ = (1/2) \left( \int_{-\infty}^{c - \beta^g_0} f \right) + (1/2) \left( \int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right)
+\end{multline*}
+Next,
+$$\omega = \int_{c - \beta^g_1 - \beta^g_0}^\infty f,$$ and so
+\begin{multline*}
+\omega - \E[\hat{p}_i]  = \left( \int_{c - \beta^g_1 - \beta^g_0}^\infty f \right) - (1/2) \left( \int_{c - \beta^g_0}^\infty f \right) - (1/2) \left( \int_{c - \beta_0^g - \beta_1^g}^\infty f \right) \\ = (1/2) \left(\int_{c - \beta^g_0 - \beta^g_1 }^\infty f \right) - (1/2) \left(\int_{c - \beta^g_0}^\infty f \right) =  (1/2)\left(\int_{-\infty}^{\beta_0^g + \beta_1^g - c} f \right) - (1/2) \left( \int_{-\infty}^{ \beta_0^g - c} f \right).
+\end{multline*}
+Therefore,
+\begin{multline*}
+\gamma(\beta_0^g, \beta_1^g, c, 1/2) = \frac{ (1/2) (\omega - \E[\hat{p}_i])}{ \E[\hat{p}_i] \left(1 - \E[\hat{p}_i] \right)} \\ = \frac{(1/2) \left[ (1/2) \left( \int_{ -\infty}^{\beta_0^g + \beta_1^g - c} f \right) - (1/2) \left( \int_{-\infty}^{ \beta_0^g - c} f \right) \right]}{ \left[(1/2) \left( \int_{-\infty}^{ \beta_0^g - c} f \right) + (1/2) \left( \int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \right] \left[ (1/2) \left( \int_{-\infty}^{c - \beta^g_0} f \right) + (1/2) \left( \int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \right]} \\ = \frac{\left( \int_{ -\infty}^{\beta_0^g + \beta_1^g - c} f \right) - \left( \int_{-\infty}^{ \beta_0^g - c} f \right) }{ \left[ \left( \int_{-\infty}^{ \beta_0^g - c} f \right) + \left( \int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \right] \left[\left( \int_{-\infty}^{c - \beta^g_0} f \right) + \left(\int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \right]}.
+\end{multline*}
+Treating $\beta^g_0$ as a fixed constant, define 
+$$g(\beta^g_1, c) = \left( \int_{ -\infty}^{\beta_0^g + \beta_1^g - c} f \right) - \left( \int_{-\infty}^{ \beta_0^g - c} f \right) ,$$ and
+$$h(\beta^g_1, c) = \left[ \left( \int_{-\infty}^{ \beta_0^g - c} f \right) + \left( \int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \right] \left[\left( \int_{-\infty}^{c - \beta^g_0} f \right) + \left(\int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \right].$$ We can write
+$$\gamma(\beta^g_0, \beta^g_1, c, 1/2) = \frac{ g(\beta^g_1, c) }{ h(\beta^g_1, c)}.$$ We will write $g(\beta^g_1, \cdot)$ (resp., $g(\cdot, \beta^g_1)$) to denote the univariate function that results from holding the second (resp., first) argument of $g$ fixed.
+We adopt similar notation for $h$.
+
+\subsubsection{Bayes-optimal decision boundary as a critical value}
+Let $c_\textrm{bayes} = \beta^g_0 + (1/2)\beta^g_1.$ We show that $c = c_\textrm{bayes}$ is a critical value of $\gamma$ for $\pi = 1/2$ and fixed $(\beta^g_0, \beta^g_1)$ , i.e, $$\frac{\partial  \gamma(\beta^g_0, \beta^g_1, c, 1/2)}{ \partial c} \vline_{[c = c_\textrm{bayes}]} = 0.$$ By quotient rule,
+$$ \frac{\partial  \gamma(\beta^g_0, \beta^g_1, c, 1/2)}{ \partial c} = \frac{g'(\cdot, c) h(\cdot, c) - g(\cdot, c) h'(\cdot, c)}{\left[ h(\cdot, c) \right]^2}.$$ First,
+$$g'(\cdot, c) = - f( \beta^g_0 + \beta^g_1 - c ) + f(\beta^g_0 - c).$$ Second, 
+\begin{multline*}
+h'(\cdot, c) = -[f(\beta^g_0 - c)  + f( \beta^g_0 + \beta^g_1 - c )]  \left[\left( \int_{-\infty}^{c - \beta^g_0} f \right) + \left(\int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \right] \\ +[ f(c - \beta^g_0) + f(c - \beta^g_0 - \beta^g_1) ] \left[ \left( \int_{-\infty}^{ \beta_0^g - c} f \right) + \left( \int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \right].
+\end{multline*}
+Evaluating $g'(\cdot, c)$ and $h'(\cdot, c)$ at $c = c_\textrm{bayes} $, we find that
+$$g'(\beta^g_0 + (1/2) \beta^g_1) = -f(\beta^g_1/2) + f(- \beta^g_1/2) = -f(\beta^g_1/2) + f(\beta^g_1/2) = 0,$$ where the second equality follows from the evenness of $f$. Similarly, 
+\begin{multline*}
+ h'(\beta^g_0 + (1/2) \beta^g_1) = - \left[ f( -\beta^g_1/2) + f( \beta^g_1/2 ) \right] \left[ \int_{-\infty}^{ \beta^g_1/2} f + \int_{-\infty}^{ -\beta^g_1/2} f \right] + \\ \left[ f( \beta^g_1/2) + f( -\beta^g_1/2 ) \right] \left[ \int_{-\infty}^{ -\beta^g_1/2} f + \int_{-\infty}^{ \beta^g_1/2 f} \right] = 0.
+\end{multline*}
+We conclude that
+$$ \frac{\partial \gamma(\beta^g_0, \beta^g_1, c, 1/2)}{ \partial c} \vline_{[c = c_\textrm{bayes}]} = \frac{0\cdot h(\cdot, c_\textrm{bayes}) - g(\cdot, c_\textrm{bayes}) \cdot 0}{ h(\cdot, c_\textrm{bayes})^2} = 0.$$ Finally, because asymptotic bias $b$ can be expressed in terms of $\gamma$ as $$b(\beta^g_0.
+\beta^g_1, c, \pi) = \beta^m_1 \left[ 1 - \gamma(\beta^g_0, \beta^g_1, c, \pi) \right],$$ it follows that $c = c_\textrm{bayes}$ is a critical value of $b$ for $\pi = 1/2$ and fixed $(\beta^g_0, \beta^g_1)$.
+
+\subsubsection{Limit in $\beta^g_1$}
+
+We compute the limit of $\gamma(\beta^g_0, \beta^g_1, c, 1/2)$ in $\beta^g_1.$ We have that
+
+\begin{multline*}
+ \lim_{\beta^g_1 \to \infty } \gamma(\beta^g_0, \beta^g_1, c, 1/2) = \frac{ 1 - \int_{-\infty}^{\beta^g_0 - c } f}{\left[ 1 + \int_{-\infty}^{ \beta^g_0 - c} f \right] \left[\int_{-\infty}^{ c - \beta^g_0} f \right]} \\ = \frac{\int_{-\infty}^{c - \beta^g_0} f }{ \left[1 + \int_{-\infty}^{ \beta^g_0 - c} f \right] \left[\int_{-\infty}^{ c - \beta^g_0} f \right]} = \frac{1}{1 + \int_{-\infty}^{\beta^g_0 - c} f} < 1.
+ \end{multline*}
+
+ Similarly,
+\begin{multline*}
+\lim_{\beta^g_1 \to -\infty} \gamma(\beta^g_0, \beta^g_1, c, 1/2) = \frac{ -\int_{-\infty}^{\beta^g_0 - c} f }{\left[ \int_{-\infty}^{\beta^g_0 - c} f \right] \left[ \int_{-\infty}^{c - \beta^g_0} f + 1 \right]} = \frac{-1}{1 + \int_{-\infty}^{c - \beta^g_0} f} > -1.
+\end{multline*}
+
+\subsubsection{Monotonicity in $\beta^g_1$}
 

### 2021-10-19 text/outline/outline.tex
update notation of thresholding theory section

--- 2021-10-18 text/outline/outline.tex
@@
 To investigate the basic question of ``What is a good threshold selection strategy?'', we study the relationship between the asymptotic bias $b$ of $\hat{\beta}^m_1$ and the selected threshold $c$.
-For simplicity, we set the perturbation probability $\pi$ to $1/2$.
+For simplicity, we begin by setting the perturbation probability $\pi$ to $1/2$.
 Additionally, without loss of generality, we set the target of inference $\beta^m_1$ to $1$.
@@
 
-This section contains proofs for results presented the section ``Theoretical analysis of thresholding estimator.''
+This section contains proofs for results presented the section ``Theoretical analysis of thresholding estimator.'' The subsections are arranged as follows: FILL IN...
 
-\subsubsection{Limit of $\hat{\beta}^m_1$ in Gaussian model with intercepts}
+\subsubsection{Notation}
+ All notation introduced here pertains to the Gaussian model with intercepts (\ref{theoretical_model}). Recall that the attenuation function $\gamma: \R^4 \to \R$ is defined by
+$$ \gamma(\beta^g_1, c, \pi, \beta^g_0) = \frac{\pi(\omega - \E[\hat{p}_i])}{ \E[\hat{p}_i](1 -\E[\hat{p}_i])},$$ where $$\begin{cases}
+\omega = 1 - \Phi\left(c - \beta_1^g - \beta_0^g \right) ,\\ \zeta = 1 - \Phi\left( c - \beta^g_0 \right), \\ \E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi. 
+\end{cases}$$ Additionally, recall that the bias function $b: \R^5 \to \R$ is defined by
+$$ b(\beta^g_1, c, \pi, \beta^g_0, \beta^m_1) = \beta^m_1 (1 - \gamma(\beta^g_1, c, \pi, \beta^g_0)).$$ The parameters $\beta^g_0$ and $\beta^m_1$ will be given, fixed constants throughout the proofs.
+Therefore, to minimize notation, we typically use $\gamma(\beta^g_1, c, \pi)$ (resp., $b(\beta^g_1, c, \pi)$) to refer to the function $\gamma$ (resp., $b$) evaluated at $(\beta^g_1, c, \pi, \beta^g_0)$ (resp., $(\beta^g_1, c, \pi, \beta^g_0, \beta^m_1)$).
+Next, we define the functions $g$ and $h: \R^4 \to \R$ by
+\begin{equation}\label{def_g}
+g(\beta^g_1, c, \pi, \beta^g_0) = (1-\pi)\left( \int_{ -\infty}^{\beta_0^g + \beta_1^g - c} f \right) - (1-\pi)\left( \int_{-\infty}^{ \beta_0^g - c} f \right)\end{equation}
+ and
+\begin{multline}\label{def_h}
+h(\beta^g_1, c, \pi, \beta^g_0) = \left[(1-\pi)\left( \int_{-\infty}^{ \beta_0^g - c} f \right) + \pi\left( \int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \right] \cdot \\ \left[(1-\pi)\left( \int_{-\infty}^{c - \beta^g_0} f \right) + \pi\left(\int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \right],
+\end{multline} where $f: \R \to [0,\infty)$ is the $N(0,1)$ density.
+We will refer to $g$ and $h$ repeatedly throughout the proofs.
+Similarly to above, we typically use the notation $g(\beta^g_1, c, \pi)$ (resp.\ $h(\beta^g_1, c, \pi)$) to refer to $g$ (resp.\ $h$) evaluated at $(\beta^g_1, c, \pi, \beta^g_0)$.
 
-Consider the thresholding estimator $\hat{\beta}^m_1$ for the Gaussian model with intercepts (\ref{theoretical_model}).
+Finally, for a given function $r: \R^{p} \to \R$, point $x \in \R^p$, and index $i \in \{1, \dots, p\}$, we use the symbol $D_i r(x)$ to refer to the derivative of the $i$th component of $r$ evaluated at $x$ (\textit{sensu} \cite{fitzpatrick2009}).
+For example, $D_1 \gamma(\beta^g_1, c, 1/2)$ is the derivative of the first component of $\gamma$ (the component corresponding to $\beta^g_1$) evaluated at $(\beta^g_1, c, 1/2)$.
+Likewise,  $D_2g(\beta^g_1, c, \pi)$ is the derivative of the second component of $g$ (the component corresponding to $c$) evaluated at $(\beta^g_1, c, \pi).$
+
+\subsubsection{Limit of $\hat{\beta}^m_1$}
+
+We derive the limit in probability of $\hat{\beta}^m_1$ for the Gaussian model with intercepts (\ref{theoretical_model}).
 Dividing by $n$ in (\ref{thresh_est_intercepts}), we can express $\hat{\beta}^m_1$ as
-$$ \hat{\beta}^m_1 = \frac{ \frac{1}{n} \sum_{i=1}^n ( \hat{p}_i - \overline{\hat{p}_i})(m_i - \overline{m})}{ \frac{1}{n} \sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})}.$$ By WLLN,
+$$ \hat{\beta}^m_1 = \frac{ \frac{1}{n} \sum_{i=1}^n ( \hat{p}_i - \overline{\hat{p}_i})(m_i - \overline{m})}{ \frac{1}{n} \sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})}.$$ By weak LLN,
 $$ \hat{\beta}^m_1 \xrightarrow{P} \frac{\textrm{Cov}(\hat{p}_i, m_i)}{\V\left(\hat{p}_i\right)}.$$ To compute this quantity, we first compute several simpler quantities:
 \begin{itemize}
-\item[1.] \textbf{Expectation of $m_i$}: $\E[m_i] = \beta^m_0 + \beta^m_1\pi.$
-\item[2.] \textbf{Expectation of $\hat{p}_i$}: Let $f$ denote the $N(0,1)$ density. Then \begin{multline*}
-\E[\hat{p}_i] = \P\left[\hat{p}_i = 1\right] = \P\left[\beta^g_0 + \beta^g_1 + \tau_i \geq c \right] = \\ \textrm{(By LOTP) } \P\left[ \beta^g_0 + \tau_i \geq c \right]\P\left[p_i = 0\right] + \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \P[p_i = 1] \\ = \P\left[ \tau_i \geq c - \beta^g_0\right](1- \pi) + \P\left[ \tau_i \geq c - \beta^g_1 - \beta^g_0 \right](\pi) \\ = \left(\int_{c - \beta^g_0}^\infty f\right) (1 - \pi) + \left( \int_{c - \beta^g_1 - \beta^g_0}^\infty f \right)(\pi) = \zeta(1-\pi) + \omega \pi,
+\item[1.] Expectation of $m_i$: $\E[m_i] = \beta^m_0 + \beta^m_1\pi.$
+\item[2.] Expectation of $\hat{p}_i$: Let $f$ denote the $N(0,1)$ density. Then \begin{multline*}
+\E[\hat{p}_i] = \P\left[\hat{p}_i = 1\right] = \P\left[\beta^g_0 + \beta^g_1 + \tau_i \geq c \right] = \\ \textrm{(By LOTP) } \P\left[ \beta^g_0 + \tau_i \geq c \right]\P\left[p_i = 0\right] + \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \P[p_i = 1] \\ = \P\left[ \tau_i \geq c - \beta^g_0\right](1- \pi) + \P\left[ \tau_i \geq c - \beta^g_1 - \beta^g_0 \right](\pi) \\ = \left(\int_{c - \beta^g_0}^\infty f\right) (1 - \pi) + \left( \int_{c - \beta^g_1 - \beta^g_0}^\infty f \right)(\pi) = \zeta(1-\pi) + \omega \pi.
 \end{multline*}
-where $$\zeta = \int_{c - \beta_0^g}^\infty f = 1 - \Phi(c - \beta^g_0)$$ and 
-$$ \omega = \int_{c - \beta^g_1 - \beta^g_0} f = 1 - \Phi(c - \beta^g_1 - \beta^g_0).$$
-\item[3.] \textbf{Expectation of $\hat{p}_i p_i$}: 
+\item[3.] Expectation of $\hat{p}_i p_i$: 
 $$\E\left[ \hat{p}_i p_i \right] = \E\left[ \hat{p}_i | p_i = 1 \right] \P\left[ p_i =1 \right] = \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \pi = \omega \pi.$$
-\item[4.] \textbf{Expectation of $\hat{p}_i m_i$}:
+\item[4.] Expectation of $\hat{p}_i m_i$:
 \begin{multline*}
@@
 Combining these expressions, we have that
-$$ \hat{\beta}^m_1 \xrightarrow{P} \frac{\beta^m_1 \pi (\omega - \E[\hat{p}_i])}{\E[\hat{p}_i](1 - \E[\hat{p}_i])} = \beta^m_1 \gamma(\beta^g_0, \beta^g_1, c, \pi).$$
+$$ \hat{\beta}^m_1 \xrightarrow{P} \frac{\beta^m_1 \pi (\omega - \E[\hat{p}_i])}{\E[\hat{p}_i](1 - \E[\hat{p}_i])} = \beta^m_1 \gamma(\beta^g_1, c, \pi).$$
 \end{itemize}
-\subsubsection{Re-expressing $\gamma$ when $\pi = 1/2$}
 
-Suppose $\pi = 1/2$. We rewrite the attenuation fraction $\gamma$ in a way that makes it more amenable to theoretical analysis.
+\subsubsection{Re-expressing $\gamma$ in a simpler form}
+
+We rewrite the attenuation fraction $\gamma$ in a way that makes it more amenable to theoretical analysis. We leverage the fact that $f$ integrates to unity and is even.
+First,
+\begin{multline}\label{thm:gamma_expression_1} \E\left[\hat{p}_i\right] = (1 - \pi) \left( \int_{c - \beta^g_0}^\infty f \right) + \pi \left(\int_{c - \beta_0^g - \beta_1^g}^\infty f \right) \\ = (1 - \pi) \left( \int_{-\infty}^{ \beta_0^g - c} f \right) + \pi \left( \int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right), \end{multline}
+ and so \begin{multline}\label{thm:gamma_expression_2} 1 - \E\left[\hat{p}_i\right] = (1 - \pi) + \pi - \E[\hat{p}_i]  = (1-\pi) \left(1 - \int_{c - \beta_0^g}^\infty f \right)  + \pi \left(1 - \int_{c - \beta^g_0 - \beta^g_1}^\infty f \right) \\ = (1 - \pi) \left( \int_{-\infty}^{c - \beta^g_0} f \right) + \pi \left( \int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right).
+\end{multline}
+Next,
+\begin{equation}\label{thm:gamma_expression_3}
+\omega = \int_{c - \beta^g_1 - \beta^g_0}^\infty f,\end{equation} and so
+\begin{multline}\label{thm:gamma_expression_4}
+\omega - \E[\hat{p}_i]  = \left( \int_{c - \beta^g_1 - \beta^g_0}^\infty f \right) - (1 - \pi) \left( \int_{c - \beta^g_0}^\infty f \right) - \pi \left( \int_{c - \beta_0^g - \beta_1^g}^\infty f \right) \\ = (1 - \pi) \left(\int_{c - \beta^g_0 - \beta^g_1 }^\infty f \right) - (1- \pi) \left(\int_{c - \beta^g_0}^\infty f \right) \\ =  (1 - \pi)\left(\int_{-\infty}^{\beta_0^g + \beta_1^g - c} f \right) - (1 - \pi) \left( \int_{-\infty}^{ \beta_0^g - c} f \right).
+\end{multline}
+Combining (\ref{thm:gamma_expression_1}, \ref{thm:gamma_expression_2}, \ref{thm:gamma_expression_3}, \ref{thm:gamma_expression_4}), we find that
+\begin{multline}\label{gamma_alternative}
+\gamma(\beta^g_1, c, \pi) = \frac{\pi(\omega - \E[\hat{p}_i])}{\E[\hat{p}_i](1 - \E[\hat{p}_i])} \\ = \frac{\pi \left[ (1 - \pi) \left( \int_{ -\infty}^{\beta_0^g + \beta_1^g - c} f \right) - (1 - \pi) \left( \int_{-\infty}^{ \beta_0^g - c} f \right) \right]}{\left[(1-\pi) \left( \int_{-\infty}^{ \beta_0^g - c} f \right) + \pi \left( \int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \right] \left[(1 - \pi) \left( \int_{-\infty}^{c - \beta^g_0} f \right) + \pi \left( \int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \right]}.
+\end{multline}
+As a corollary, when $\pi = 1/2$, we have that
+\begin{multline}\label{gamma_alternative_pi_half}
+\gamma(\beta^g_1, c, 1/2)  \\ = \frac{\left( \int_{ -\infty}^{\beta_0^g + \beta_1^g - c} f \right) - \left( \int_{-\infty}^{ \beta_0^g - c} f \right) }{ \left[ \left( \int_{-\infty}^{ \beta_0^g - c} f \right) + \left( \int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \right] \left[\left( \int_{-\infty}^{c - \beta^g_0} f \right) + \left(\int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \right]}.
+\end{multline}
+Recalling the definitions of $g$ (\ref{def_g}) and $h$ (\ref{def_h}), we can write $\gamma$ as
+$$ \gamma(\beta^g_1, c, \pi) = \frac{g(\beta^g_1, c, \pi)}{h(\beta^g_1, c,\pi)}.$$
+
+\subsubsection{Derivatives of  $g$ and $h$ in $c$}
+
+We compute the derivatives of $g$ and $h$ in $c$, which we will need for subsequent results.
+
+% \subsubsection{Bayes-optimal decision boundary as a critical value}
+% Let $c_\textrm{bayes} = \beta^g_0 + (1/2)\beta^g_1.$ We show that $c = c_\textrm{bayes}$ is a critical value of $\gamma$ for $\pi = 1/2$ and fixed $(\beta^g_0, \beta^g_1)$ , i.e, $$\frac{\partial  \gamma(\beta^g_0, \beta^g_1, c, 1/2)}{ \partial c} \vline_{[c = c_\textrm{bayes}]} = 0.$$ By quotient rule,
+% $$ \frac{\partial  \gamma(\beta^g_0, \beta^g_1, c, 1/2)}{ \partial c} = \frac{g'(\cdot, c) h(\cdot, c) - g(\cdot, c) h'(\cdot, c)}{\left[ h(\cdot, c) \right]^2}.$$ First,
+% $$g'(\cdot, c) = - f( \beta^g_0 + \beta^g_1 - c ) + f(\beta^g_0 - c).$$ Second, 
+% \begin{multline*}
+% h'(\cdot, c) = -[f(\beta^g_0 - c)  + f( \beta^g_0 + \beta^g_1 - c )]  \left[\left( \int_{-\infty}^{c - \beta^g_0} f \right) + \left(\int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \right] \\ +[ f(c - \beta^g_0) + f(c - \beta^g_0 - \beta^g_1) ] \left[ \left( \int_{-\infty}^{ \beta_0^g - c} f \right) + \left( \int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \right].
+% \end{multline*}
+% Evaluating $g'(\cdot, c)$ and $h'(\cdot, c)$ at $c = c_\textrm{bayes} $, we find that
+% $$g'(\beta^g_0 + (1/2) \beta^g_1) = -f(\beta^g_1/2) + f(- \beta^g_1/2) = -f(\beta^g_1/2) + f(\beta^g_1/2) = 0,$$ where the second equality follows from the evenness of $f$. Similarly, 
+% \begin{multline*}
+% h'(\beta^g_0 + (1/2) \beta^g_1) = - \left[ f( -\beta^g_1/2) + f( \beta^g_1/2 ) \right] \left[ \int_{-\infty}^{ \beta^g_1/2} f + \int_{-\infty}^{ -\beta^g_1/2} f \right] + \\ \left[ f( \beta^g_1/2) + f( -\beta^g_1/2 ) \right] \left[ \int_{-\infty}^{ -\beta^g_1/2} f + \int_{-\infty}^{ \beta^g_1/2 f} \right] = 0.
+% \end{multline*}
+% We conclude that
+% $$ \frac{\partial \gamma(\beta^g_0, \beta^g_1, c, 1/2)}{ \partial c} \vline_{[c = c_\textrm{bayes}]} = \frac{0\cdot h(\cdot, c_\textrm{bayes}) - g(\cdot, c_\textrm{bayes}) \cdot 0}{ h(\cdot, c_\textrm{bayes})^2} = 0.$$ Finally, because asymptotic bias $b$ can be expressed in terms of $\gamma$ as $$b(\beta^g_0.
+\beta^g_1, c, \pi) = \beta^m_1 \left[ 1 - \gamma(\beta^g_0, \beta^g_1, c, \pi) \right],$$ it follows that $c = c_\textrm{bayes}$ is a critical value of $b$ for $\pi = 1/2$ and fixed $(\beta^g_0, \beta^g_1)$.
+
+% \subsubsection{Limit in $\beta^g_1$}
+
+% We compute the limit of $\gamma(\beta^g_0, \beta^g_1, c, 1/2)$ in $\beta^g_1.$ We have that
+
+% \begin{multline*}
+ %\lim_{\beta^g_1 \to \infty } \gamma(\beta^g_0, \beta^g_1, c, 1/2) = \frac{ 1 - \int_{-\infty}^{\beta^g_0 - c } f}{\left[ 1 + \int_{-\infty}^{ \beta^g_0 - c} f \right] \left[\int_{-\infty}^{ c - \beta^g_0} f \right]} \\ = \frac{\int_{-\infty}^{c - \beta^g_0} f }{ \left[1 + \int_{-\infty}^{ \beta^g_0 - c} f \right] \left[\int_{-\infty}^{ c - \beta^g_0} f \right]} = \frac{1}{1 + \int_{-\infty}^{\beta^g_0 - c} f} < 1.
+% \end{multline*}
+
+% Similarly,
+% \begin{multline*}
+% \lim_{\beta^g_1 \to -\infty} \gamma(\beta^g_0, \beta^g_1, c, 1/2) = \frac{ -\int_{-\infty}^{\beta^g_0 - c} f }{\left[ \int_{-\infty}^{\beta^g_0 - c} f \right] \left[ \int_{-\infty}^{c - \beta^g_0} f + 1 \right]} = \frac{-1}{1 + \int_{-\infty}^{c - \beta^g_0} f} > -1.
+% \end{multline*}
+
+% \subsubsection{Monotonicity in $\beta^g_1$}
+% We prove that $\gamma$ is monotonic in $\beta_1^g$ for given $(\beta^g_0, c)$ and $\pi = 1/2$. Recall that
+% $$\gamma(\beta^g_0, \beta^g_1, c, 1/2) = \frac{ g(\beta^g_1, c) }{ h(\beta^g_1, c)}.$$ By the quotient rule,
+% $$D_2 \gamma (\beta^g_0, \beta^g_1, c, 1/2) = \frac{D_1 g( \beta^g_1, c) h(\beta^g_1, c) - g(\beta^1_g, c) D_1 h(\beta^g_1, c) }{ [ h(\beta^g_1, c)^2]}.$$ The derivative $D_1 g(\beta^g_1, c)$ of $g$ is
+% $$D_1 g(\beta^g_1, c) = f( \beta^g_0 + \beta^g_1 - c ).$$
 

### 2021-10-20 text/outline/outline.tex
limit in c of gamma

@@
 \end{multline}
-As a corollary, when $\pi = 1/2$, we have that
+As a corollary, when $\pi = 1/2$,
 \begin{multline}\label{gamma_alternative_pi_half}
@@
 Recalling the definitions of $g$ (\ref{def_g}) and $h$ (\ref{def_h}), we can write $\gamma$ as
-$$ \gamma(\beta^g_1, c, \pi) = \frac{g(\beta^g_1, c, \pi)}{h(\beta^g_1, c,\pi)}.$$
+$$ \gamma(\beta^g_1, c, \pi) = \frac{\pi g(\beta^g_1, c, \pi)}{h(\beta^g_1, c,\pi)}.$$
 
@@
 
-We compute the derivatives of $g$ and $h$ in $c$, which we will need for subsequent results.
+We compute the derivatives of $g$ and $h$ in $c$, which we will need to prove subsequent results. First, by FTC and the evenness of $f$, we have that
+\begin{multline}\label{dg_dc}
+D_2 g(\beta^g_1, c, \pi) = -(1-\pi)f( \beta^g_0 + \beta^g_1 - c ) + (1-\pi) f(\beta^g_0 - c) \\ = (1-\pi) f(c - \beta^g_0) - (1-\pi)f(c - \beta^g_0 - \beta^g_1).
+\end{multline}
+Second, we have that
+\begin{multline}\label{dh_dc}
+D_2 h(\beta^g_1, c, \pi) \\ = -[(1-\pi)f(\beta^g_0 - c) + \pi f( \beta^g_0 + \beta^g_1 - c )]\left[(1-\pi)\left(\int_{-\infty}^{c - \beta^g_0} f \right) + \pi \left(\int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \right] \\ + [(1-\pi) f(c - \beta^g_0) +  \pi f(c - \beta^g_0 - \beta^g_1)] \left[(1-\pi)\left( \int_{-\infty}^{\beta_0^g - c} f \right) + \pi \left(\int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \right] \\ = \left[ (1-\pi) f(c - \beta^g_0) +  \pi f(c - \beta^g_0 - \beta^g_1) \right] \cdot \\ \bigg[ (1-\pi)\left( \int_{-\infty}^{\beta_0^g - c} f \right) + \pi \left(\int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \\ - (1-\pi)\left(\int_{-\infty}^{c - \beta^g_0} f \right) - \pi \left(\int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \bigg].
+\end{multline}
+
+\subsubsection{Limit of $\gamma$ in $c$}
+
+Assume (without loss of generality) that $\beta^g_1 > 0$.
+We compute $\lim_{c \to \infty} \gamma(\beta^g_1, c, \pi)$.
+Observe that $$\lim_{c \to \infty} g(\beta^g_1, c, \pi) = \lim_{c \to \infty} h(\beta^g_1, c, \pi)  = 0.$$ Therefore, we can apply L'H\^{o}pital's rule.
+We have by (\ref{dg_dc}) and (\ref{dh_dc}) that \begin{multline}\label{c_limit_product}
+\lim_{c \to \infty} \gamma(\beta^g_1, c, \pi) = \lim_{c \to \infty} \frac{\pi D_2 g(\beta^g_1, c, \pi)}{D_2h(\beta^g_1, c, \pi)} \\ = \lim_{c \to \infty} \bigg\{ \frac{(1-\pi) f(c - \beta^g_0) + \pi f(c - \beta^g_0 - \beta^g_1)}{\pi (1-\pi) f(c - \beta^g_0) - \pi (1-\pi)f(c - \beta^g_0 - \beta^g_1)} \\ \cdot \bigg[ (1-\pi)\left( \int_{-\infty}^{\beta_0^g - c} f \right) + \pi \left(\int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \\ - (1-\pi)\left(\int_{-\infty}^{c - \beta^g_0} f \right) - \pi \left(\int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \bigg] \bigg\}^{-1}.
+ \end{multline}
+ We evaluate the two terms in the product (\ref{c_limit_product}) separately. First, observe that
+ \begin{equation}\label{c_limit_product_2}
+ \frac{(1-\pi) f(c - \beta^g_0) + \pi f(c - \beta^g_0 - \beta^g_1)}{\pi (1-\pi) f(c - \beta^g_0) - \pi (1-\pi)f(c - \beta^g_0 - \beta^g_1)} = \frac{\frac{(1-\pi) f(c - \beta^g_0)}{ f(c - \beta^g_0 - \beta^g_1)} + \pi}{\frac{ \pi(1-\pi) f(c - \beta^g_0)}{ f(c - \beta^g_0 - \beta^g_1)} - \pi(1-\pi)}.
+ \end{equation}
+ To evaluate the limit of (\ref{c_limit_product_2}), must evaluate the limit of
+ \begin{multline}\label{c_limit_product_3}
+ \frac{f(c - \beta^g_0)}{f(c - \beta^g_0 - \beta^g_1)} = \frac{\exp{[-(1/2)(c - \beta_0^g)^2]}}{\exp{[-(1/2)( c - \beta^g_0 - \beta^g_1)^2]}} \\ = \frac{\exp[ -(1/2)(c^2 - 2 c \beta^g_0 + (\beta^g_0)^2)]}{\exp\left[-(1/2)( c^2 - 2c \beta^g_0 - 2 c \beta^g_1 + (\beta^g_0)^2 + 2( \beta^g_0 \beta^g_1) + (\beta^g_1)^2)\right]} \\ = \exp\big[-c^2/2 + c \beta^g_0 - (\beta^g_0)^2/2 \\ + c^2/2 - c \beta^g_0 - c \beta^g_1 + (\beta^g_0)^2/2 + \beta^g_0 \beta^g_1 + (\beta^g_1)^2/2 \big] \\ = \exp[ -c \beta^g_1 + \beta^g_0 \beta^g_1 + (\beta^g_1)^2/2] = \exp[ \beta^g_0 \beta^g_1 + (\beta^g_1)^2/2]\exp[ -c \beta^g_1].
+\end{multline}
+Computing the limit of (\ref{c_limit_product_3}), we obtain
+$$
+\lim_{c \to \infty} \frac{f(c - \beta^g_0)}{f(c - \beta^g_0 - \beta^g_1)} = \exp[ \beta^g_0 \beta^g_1 + (\beta^g_1)^2/2] \lim_{c \to \infty} \exp[ -c \beta^g_1] = 0
+$$ for $\beta^g_1 > 0$. We now can evaluate the limit of (\ref{c_limit_product_2}) (for $\beta^g_1 > 0$):
+$$ \lim_{c \to \infty} \frac{(1-\pi) f(c - \beta^g_0) + \pi f(c - \beta^g_0 - \beta^g_1)}{\pi (1-\pi) f(c - \beta^g_0) - \pi (1-\pi)f(c - \beta^g_0 - \beta^g_1)} = \frac{-\pi}{\pi(1-\pi)} = -\frac{1}{1 -\pi}.$$ Next, we compute the limit of the other term in the product (\ref{c_limit_product}):
+\begin{multline}\label{c_limit_product_4}
+\lim_{c \to \infty} \bigg[ (1-\pi)\left( \int_{-\infty}^{\beta_0^g - c} f \right) + \pi \left(\int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \\ - (1-\pi)\left(\int_{-\infty}^{c - \beta^g_0} f \right) - \pi \left(\int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \bigg] = -(1-\pi) - \pi = -1.
+\end{multline}
+Combining (\ref{c_limit_product_2}) and (\ref{c_limit_product_4}), the limit (\ref{c_limit_product}) evaluates to
+$$ \lim_{c \to \infty} \gamma(\beta^g_1, c, \pi) = \left(  \frac{ 1 }{ 1 - \pi }\right)^{-1} = 1 - \pi.$$ It follows that the limit in $c$ of the bias $b$ is
+$$\lim_{c \to \infty} b(\beta^g_1, c, \pi) = \beta_1^m(1 - \lim_{c \to \infty} \gamma(\beta^g_1, c, \pi)) = \beta^m_1 \pi.$$
+A corollary is that
+$$\lim_{c \to \infty} b(\beta^g_1, c, 1/2) = (1/2)\beta^m_1.$$
+
+\subsubsection{Critical values of }
 

### 2021-10-20 text/outline/outline.tex
add critical value proof

@@
 
-\subsubsection{Critical values of }
+\subsubsection{Bayes-optimal decision boundary as a critical value of $\gamma$}
+Let $c_\textrm{bayes} = \beta^g_0 + (1/2)\beta^g_1.$ We show that $c = c_\textrm{bayes}$ is a critical value of $\gamma$ for $\pi = 1/2$ and given $\beta^g_1$, i.e, $$D_2 \gamma (\beta^g_1, c_\textrm{bayes}, 1/2) = 0.$$ By the quotient rule,
+\begin{multline}\label{quotient_rule}
+D_2\gamma(\beta^g_1, c, 1/2) \\ = \frac{(1/2) D_2g(\beta^g_1, c, 1/2) h(\beta^g_1, c, 1/2) - (1/2) g(\beta^g_1, c, 1/2) D_2 h(\beta^g_1, c, 1/2)}{[h(\beta^g_1, c, \pi)]^2} \\ = \frac{\left[4(1/2) D_2g(\beta^g_1, c, 1/2)\right] h(\beta^g_1, c, 1/2) - (1/2) g(\beta^g_1, c, 1/2) \left[4 D_2 h(\beta^g_1, c, 1/2)\right]}{[2h(\beta^g_1, c, \pi)]^2},
+\end{multline}
+where the second equality multiplies both numerator and denominator by $4$. We have by (\ref{dg_dc}) that
+\begin{multline}\label{dg_dc_bayes}
+4 \cdot (1/2) \cdot D_2 g(\beta^g_1, c_\textrm{bayes}, 1/2)  = f( \beta^g_1/2) - f( -\beta^g_1/2) \\ = f(\beta^g_1/2) - f(\beta^g_1/2) = 0.
+\end{multline}
+Next, we have by (\ref{dh_dc}) that
+\begin{multline}\label{dh_dc_bayes}
+4 \cdot D_2 h(\beta^g_1, c_\textrm{bayes}, \pi) = [f( \beta^g_1/2) + f( -\beta^g_1/2)] \cdot \\ \left[  \int_{-\infty}^{-\beta^g_1/2} f + \int_{-\infty}^{\beta^g_1/2} f - \int_{-\infty}^{\beta^g_1/2} f - \int_{-\infty}^{-\beta^g_1/2} f \right] = 0.
+\end{multline}
+Plugging in (\ref{dh_dc_bayes}) and (\ref{dg_dc_bayes}) to (\ref{quotient_rule}), we find that 
+$$D_2(\beta^g_1, c_\textrm{bayes}, 1/2) = 0.$$
 
-% \subsubsection{Bayes-optimal decision boundary as a critical value}
-% Let $c_\textrm{bayes} = \beta^g_0 + (1/2)\beta^g_1.$ We show that $c = c_\textrm{bayes}$ is a critical value of $\gamma$ for $\pi = 1/2$ and fixed $(\beta^g_0, \beta^g_1)$ , i.e, $$\frac{\partial  \gamma(\beta^g_0, \beta^g_1, c, 1/2)}{ \partial c} \vline_{[c = c_\textrm{bayes}]} = 0.$$ By quotient rule,
-% $$ \frac{\partial  \gamma(\beta^g_0, \beta^g_1, c, 1/2)}{ \partial c} = \frac{g'(\cdot, c) h(\cdot, c) - g(\cdot, c) h'(\cdot, c)}{\left[ h(\cdot, c) \right]^2}.$$ First,
 % $$g'(\cdot, c) = - f( \beta^g_0 + \beta^g_1 - c ) + f(\beta^g_0 - c).$$ Second, 

### 2021-10-21 text/outline/outline.tex
proof of monotonicity

@@
 \begin{multline}\label{gamma_alternative_pi_half}
-\gamma(\beta^g_1, c, 1/2)  \\ = \frac{\left( \int_{ -\infty}^{\beta_0^g + \beta_1^g - c} f \right) - \left( \int_{-\infty}^{ \beta_0^g - c} f \right) }{ \left[ \left( \int_{-\infty}^{ \beta_0^g - c} f \right) + \left( \int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \right] \left[\left( \int_{-\infty}^{c - \beta^g_0} f \right) + \left(\int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \right]}.
+\gamma(\beta^g_1, c, 1/2)  \\ = \frac{\left( \int_{ -\infty}^{\beta_0^g + \beta_1^g - c} f \right) - \left( \int_{-\infty}^{ \beta_0^g - c} f \right)}{\left[ \left( \int_{-\infty}^{ \beta_0^g - c} f \right) + \left( \int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \right] \left[\left( \int_{-\infty}^{c - \beta^g_0} f \right) + \left(\int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \right]}.
 \end{multline}
@@
 $$ \gamma(\beta^g_1, c, \pi) = \frac{\pi g(\beta^g_1, c, \pi)}{h(\beta^g_1, c,\pi)}.$$
+The special case (\ref{gamma_alternative_pi_half}) is identical to
+\begin{equation}\label{gamma_alt2_pi_half}
+\gamma(\beta^g_1, c, 1/2) = \frac{(4)(1/2)g(\beta^g_1, c, 1/2)}{4 h(\beta^g_1, c, 1/2)} = \frac{2 g(\beta^g_1, c, 1/2)}{4h(\beta^g_1, c, 1/2)},
+\end{equation}
+i.e., the numerator and denominator of  (\ref{gamma_alt2_pi_half}) coincide with those of (\ref{gamma_alternative_pi_half}).
 
 \subsubsection{Derivatives of  $g$ and $h$ in $c$}
-
 We compute the derivatives of $g$ and $h$ in $c$, which we will need to prove subsequent results. First, by FTC and the evenness of $f$, we have that
@@
 Plugging in (\ref{dh_dc_bayes}) and (\ref{dg_dc_bayes}) to (\ref{quotient_rule}), we find that 
-$$D_2(\beta^g_1, c_\textrm{bayes}, 1/2) = 0.$$
+$$D_2(\beta^g_1, c_\textrm{bayes}, 1/2) = 0.$$ It follows that $c_\textrm{bayes}$ is a critical value of the bias function $b$.
 
-% $$g'(\cdot, c) = - f( \beta^g_0 + \beta^g_1 - c ) + f(\beta^g_0 - c).$$ Second, 
-% \begin{multline*}
-% h'(\cdot, c) = -[f(\beta^g_0 - c)  + f( \beta^g_0 + \beta^g_1 - c )]  \left[\left( \int_{-\infty}^{c - \beta^g_0} f \right) + \left(\int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \right] \\ +[ f(c - \beta^g_0) + f(c - \beta^g_0 - \beta^g_1) ] \left[ \left( \int_{-\infty}^{ \beta_0^g - c} f \right) + \left( \int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \right].
-% \end{multline*}
-% Evaluating $g'(\cdot, c)$ and $h'(\cdot, c)$ at $c = c_\textrm{bayes} $, we find that
-% $$g'(\beta^g_0 + (1/2) \beta^g_1) = -f(\beta^g_1/2) + f(- \beta^g_1/2) = -f(\beta^g_1/2) + f(\beta^g_1/2) = 0,$$ where the second equality follows from the evenness of $f$. Similarly, 
-% \begin{multline*}
-% h'(\beta^g_0 + (1/2) \beta^g_1) = - \left[ f( -\beta^g_1/2) + f( \beta^g_1/2 ) \right] \left[ \int_{-\infty}^{ \beta^g_1/2} f + \int_{-\infty}^{ -\beta^g_1/2} f \right] + \\ \left[ f( \beta^g_1/2) + f( -\beta^g_1/2 ) \right] \left[ \int_{-\infty}^{ -\beta^g_1/2} f + \int_{-\infty}^{ \beta^g_1/2 f} \right] = 0.
-% \end{multline*}
-% We conclude that
-% $$ \frac{\partial \gamma(\beta^g_0, \beta^g_1, c, 1/2)}{ \partial c} \vline_{[c = c_\textrm{bayes}]} = \frac{0\cdot h(\cdot, c_\textrm{bayes}) - g(\cdot, c_\textrm{bayes}) \cdot 0}{ h(\cdot, c_\textrm{bayes})^2} = 0.$$ Finally, because asymptotic bias $b$ can be expressed in terms of $\gamma$ as $$b(\beta^g_0.
-\beta^g_1, c, \pi) = \beta^m_1 \left[ 1 - \gamma(\beta^g_0, \beta^g_1, c, \pi) \right],$$ it follows that $c = c_\textrm{bayes}$ is a critical value of $b$ for $\pi = 1/2$ and fixed $(\beta^g_0, \beta^g_1)$.
+\subsubsection{Limit in $\beta^g_1$}
 
-% \subsubsection{Limit in $\beta^g_1$}
+We compute the limit of $\gamma$ in $\beta^g_1$ given $\pi = 1/2$.  First,
+\begin{multline*}
+ \lim_{\beta^g_1 \to \infty} \gamma(\beta^g_1, c, 1/2) = \frac{ 1 - \int_{-\infty}^{\beta^g_0 - c} f}{\left[ 1 + \int_{-\infty}^{ \beta^g_0 - c} f \right] \left[\int_{-\infty}^{ c - \beta^g_0} f \right]} \\ = \frac{\int_{-\infty}^{c - \beta^g_0} f }{ \left[1 + \int_{-\infty}^{ \beta^g_0 - c} f \right] \left[\int_{-\infty}^{ c - \beta^g_0} f \right]} = \frac{1}{1 + \int_{-\infty}^{\beta^g_0 - c} f} < 1.
+\end{multline*}
+Similarly,
+\begin{multline*}
+\lim_{\beta^g_1 \to -\infty} \gamma(\beta^g_1, c, 1/2) = \frac{ -\int_{-\infty}^{\beta^g_0 - c} f }{\left[ \int_{-\infty}^{\beta^g_0 - c} f \right] \left[ \int_{-\infty}^{c - \beta^g_0} f + 1 \right]} = \frac{-1}{1 + \int_{-\infty}^{c - \beta^g_0} f} > -1.
+\end{multline*}
 
-% We compute the limit of $\gamma(\beta^g_0, \beta^g_1, c, 1/2)$ in $\beta^g_1.$ We have that
+\subsubsection{Monotonicity in $\beta^g_1$}
 
-% \begin{multline*}
- %\lim_{\beta^g_1 \to \infty } \gamma(\beta^g_0, \beta^g_1, c, 1/2) = \frac{ 1 - \int_{-\infty}^{\beta^g_0 - c } f}{\left[ 1 + \int_{-\infty}^{ \beta^g_0 - c} f \right] \left[\int_{-\infty}^{ c - \beta^g_0} f \right]} \\ = \frac{\int_{-\infty}^{c - \beta^g_0} f }{ \left[1 + \int_{-\infty}^{ \beta^g_0 - c} f \right] \left[\int_{-\infty}^{ c - \beta^g_0} f \right]} = \frac{1}{1 + \int_{-\infty}^{\beta^g_0 - c} f} < 1.
-% \end{multline*}
+We show that $\gamma$ is monotonically increasing in $\beta_1^g$ for $\pi = 1/2$.
+First, we compute the derivative of (\ref{gamma_alt2_pi_half}) with respect to $\beta^g_1$.
+The quotient rule implies that
+\begin{multline}\label{d_gamma_d_beta}
+D_1 \gamma(\beta^g_1, c, 1/2) \\ = \frac{D_1[2g(\beta^g_1, c, 1/2)] 4h(\beta^g_1, c, 1/2) - 2g(\beta^g_1, c, 1/2)D_1[4h(\beta^g_1, c, 1/2)]}{[4h(\beta^g_1, c, 1/2)]^2}.
+\end{multline}
+First,
+$$\textcolor{violet}{D_1[2g(\beta^g_1, c, 1/2)] = f(\beta^g_0 + \beta^g_1 - c)}.$$
+Next,
+\begin{multline*}
+\textcolor{teal}{D_1[4h(\beta^g_1, c, 1/2)] = f(\beta^g_0 + \beta^g_1 - c) \left[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g) \right]} \\ \textcolor{teal}{- f(\beta^g_0 + \beta^g_1 - c) \left[\Phi(\beta_0^g - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right]}.\end{multline*}
+Recall that
+$$\textcolor{blue}{2g(\beta^g_1,c,1/2) = \Phi(\beta^g_0 + \beta^g_1 - c) - \Phi(\beta^g_0 - c)}.$$
+and
+$$\textcolor{red}{4h(\beta^g_1, c, 1/2) = \left[ \Phi(\beta^g_0 - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right] \left[\Phi(c - \beta^g_0) + \Phi( c - \beta^g_0 - \beta^g_1) \right]}.$$
+To show that $\gamma$ is monotonically increasing in $\beta^g_1$, it is sufficient to show that (\ref{d_gamma_d_beta}) is nonnegative, i.e.
+\begin{equation}\label{thm:gamma_monotonic_1}
+\textcolor{violet}{D_1[2g(\beta^g_1, c, 1/2)]}\textcolor{red}{4h(\beta^g_1, c, 1/2)} \geq \textcolor{blue}{2g(\beta^g_1, c, 1/2)}\textcolor{teal}{D_1[4h(\beta^g_1, c, 1/2)]}.
+\end{equation}
+The product $\textcolor{violet}{D_1[2g(\beta^g_1, c, 1/2)]}\textcolor{red}{4h(\beta^g_1, c, 1/2)}$ is
+\begin{multline}\label{g_prime_times_h}
+\textcolor{violet}{D_1[2g(\beta^g_1, c, 1/2)]}\textcolor{red}{4h(\beta^g_1, c, 1/2)} \\ = \textcolor{violet}{f(\beta^g_0 + \beta^g_1 - c)}\textcolor{red}{\left[ \Phi(\beta_0^g - c ) + \Phi(\beta^g_0 + \beta^g_1 - c)\right]\left[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g) \right]}.
+\end{multline} 
+Next, the product $\textcolor{blue}{2g(\beta^g_1, c, 1/2)}\textcolor{teal}{D_1[4h(\beta^g_1, c, 1/2)]}$ is
+\begin{multline}\label{g_times_h_prime}
+\textcolor{blue}{2g(\beta^g_1, c, 1/2)}\textcolor{teal}{D_1[4h(\beta^g_1, c, 1/2)]} \\ = \textcolor{teal}{f(\beta^g_0 + \beta^g_1 - c)}\textcolor{blue}{\left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c)\right]}\textcolor{teal}{\left[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g) \right]} \\ - \textcolor{teal}{f(\beta^g_0 + \beta^g_1 - c)}\textcolor{teal}{\left[\Phi(\beta_0^g - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right]} \textcolor{blue}{\left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c)\right]}.
+\end{multline}
+Divide both sides by $f(\beta^g_0 - \beta^g_1 - c) > 0$. Then, add
+$$\textcolor{teal}{\left[\Phi(\beta_0^g - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right]}\textcolor{blue}{\left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c)\right]}.$$ The left-hand side becomes
+\begin{multline*}
+\textcolor{red}{\left[ \Phi(\beta_0^g - c ) + \Phi(\beta^g_0 + \beta^g_1 - c)\right]\left[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g) \right]} \\ +\textcolor{teal}{\left[\Phi(\beta_0^g - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right]}\textcolor{blue}{\left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c)\right]} \\ = \left[\Phi(\beta^g_0 - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right] \\ \cdot \left[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g) + \Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c) \right].
+\end{multline*}
+Meanwhile, the right-hand side becomes
+$$\left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c)\right]\left[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g) \right].$$
+Therefore, we want to demonstrate that
+\begin{multline}\label{basic_ineq}
+\left[\Phi(\beta^g_0 - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right] \\ \cdot \left[\textcolor{brown}{\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c)} + \textcolor{olive}{\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g)} \right] \\ \geq \left[\textcolor{brown}{\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c)}\right]\left[\textcolor{olive}{\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g)}\right].
+\end{multline}
+(We changed the color scheme to highlight like terms.)
+We take cases on the sign on $\beta^g_1$ to prove this inequality.
+\\ \\ \noindent
+\textbf{Case 1}: $\beta^1_g < 0$.
+Then $ \beta^g_1 + (\beta^g - c) < (\beta^g_0 - c),$ implying $\Phi(\beta^g_0 + \beta^g_1 - c) < \Phi(\beta^g_0 - c),$ or $[\Phi(\beta^g_0 + \beta^g_1 - c) - \Phi(\beta^g_0 - c)] < 0.$ Moreover, the quantity $[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g)]$ is positive.
+Thus, the right-hand side of (\ref{basic_ineq}) is negative.
 
-% Similarly,
-% \begin{multline*}
-% \lim_{\beta^g_1 \to -\infty} \gamma(\beta^g_0, \beta^g_1, c, 1/2) = \frac{ -\int_{-\infty}^{\beta^g_0 - c} f }{\left[ \int_{-\infty}^{\beta^g_0 - c} f \right] \left[ \int_{-\infty}^{c - \beta^g_0} f + 1 \right]} = \frac{-1}{1 + \int_{-\infty}^{c - \beta^g_0} f} > -1.
-% \end{multline*}
+Turning our attention of the left-hand side, we have that
+\begin{multline*}
+\Phi(\beta^g_0 + \beta^g_1 - c) + \Phi( c - \beta^g_0 - \beta^g_1) = \int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f + \int_{-\infty}^{c - \beta^g_0 - \beta^g_1} f \\ = \int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f  + \int_{\beta^g_0 + \beta^g_1 - c}^\infty f = 1.
+\end{multline*}
+Additionally, $\Phi(\beta^g_0 - c) < 1,$ and $ \Phi(c - \beta^g_0) > 0$. Combining these expressions, we find that
+$$ \left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c) + \Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g) \right] > 0.
+$$ Because $\left[\Phi(\beta^g_0 - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right] > 0,$ it follows that the entire left-hand side of (\ref{basic_ineq}) is positive.
+\\ \\ \noindent
+\textbf{Case 2}: $\beta^1_g \geq 0.$  We show that the first term on the LHS of (\ref{basic_ineq}) is greater than the first term on the RHS, and likewise that the second term on the LHS is greater than the second term on the RHS.
+First,
+$$ \Phi(\beta^g_0 - c) \geq - \Phi(\beta^g_0 - c),$$ implying
+$$ \Phi(\beta^g_0 - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \geq \Phi(\beta^g_0 - \beta^g_1 - c) - \Phi(\beta^g_0 - c).$$
+Second, $\beta^g_1 \geq 0$ implies that $\beta^g_1 + \beta^g_0 - c \geq \beta^g_0 - c,$ and so $$\Phi(\beta^g_1 + \beta^g_0 - c) - \Phi(\beta^g_0 - c) \geq 0.$$ Finally, 
+$$ \Phi(\beta^g_1 + \beta^g_0 - c) - \Phi(\beta^g_0 - c) + \Phi(c - \beta^g_0) + \Phi(c - \beta^g_0 - \beta^g_1) \geq \Phi(c - \beta^g_0) + \Phi(c - \beta^g_0 - \beta^g_1).$$
+Therefore, the inequality (\ref{basic_ineq}) holds.
 
-% \subsubsection{Monotonicity in $\beta^g_1$}
-% We prove that $\gamma$ is monotonic in $\beta_1^g$ for given $(\beta^g_0, c)$ and $\pi = 1/2$. Recall that
-% $$\gamma(\beta^g_0, \beta^g_1, c, 1/2) = \frac{ g(\beta^g_1, c) }{ h(\beta^g_1, c)}.$$ By the quotient rule,
-% $$D_2 \gamma (\beta^g_0, \beta^g_1, c, 1/2) = \frac{D_1 g( \beta^g_1, c) h(\beta^g_1, c) - g(\beta^1_g, c) D_1 h(\beta^g_1, c) }{ [ h(\beta^g_1, c)^2]}.$$ The derivative $D_1 g(\beta^g_1, c)$ of $g$ is
-% $$D_1 g(\beta^g_1, c) = f( \beta^g_0 + \beta^g_1 - c ).$$
+\subsubsection{Attenuation bias}
+
+For given $(\beta^g_0, \beta^g_1)$,
 

### 2021-10-21 text/outline/outline.tex
proof of monotonicity in beta

--- 2021-10-20 text/outline/outline.tex
@@
 \usepackage{xcolor}
+\newtheorem{proposition}{Proposition}
+\newtheorem{lemma}{Lemma}
 \newcommand{\blue}[1]{\textcolor{blue}{#1}}
@@
 \begin{center}
-\textbf{CRISPR genome editing, single-cell sequencing, and exponential family measurement error models}
+\textbf{CRISPR genome engineering, single-cell sequencing, and exponential family measurement error models}
 \end{center}
@@
 Our work makes two main contributions.
-First, we study the thresholding method from empirical and theoretical perspectives.
+First, we conduct an in-depth study of the thresholding method from empirical and theoretical perspectives.
 Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian model.
@@
 \end{equation}
-For a given threshold $c \in \R$, let $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ for $\beta^m_1$ is given by $$\hat{\beta}^m_1 = \frac{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}}) (m_i - \overline{m})}{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})}.$$ The limit in probability (as $n \to \infty$) of $\hat{\beta}^m_0$ is
+For a given threshold $c \in \R$, let $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ for $\beta^m_1$ is given by $$\hat{\beta}^m_1 = \frac{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}}) (m_i - \overline{m})}{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})}.$$
+
+\begin{proposition}
+ The limit in probability (as $n \to \infty$) of $\hat{\beta}^m_0$ is
 \begin{equation}\label{thresh_est_intercepts}
@@
 \end{cases}$$
+\end{proposition}
 Let $\gamma: \R^4 \to \R$ be defined by
-$$ \gamma( \beta^g_0, \beta^g_1, c, \pi ) = \frac{\pi (\omega - \E[\hat{p}_i])}{\E[\hat{p}_i] (1 - \E[\hat{p}_i]) }.$$ We call $\gamma$ the \textit{attenuation function}. Observe that \begin{itemize}
+$$ \gamma( \beta^g_0, \beta^g_1, c, \pi ) = \frac{\pi (\omega - \E[\hat{p}_i])}{\E[\hat{p}_i] (1 - \E[\hat{p}_i]) }.$$ We call $\gamma$ the ``attenuation function.'' Observe that \begin{itemize}
 \item[i.] $\gamma$ does not depend on $\beta^m_1$ or $\beta^m_0$, and
@@
 \begin{itemize}
-\item[1.] The Bayes-optimal threshold $c_\textrm{bayes}$ is a critical value of the bias function $b$ (Figure \ref{thresholding_theoretical}a, all panels), i.e. 
-$$ \frac{\partial b( \beta^g_0, \beta^g_1, c, 1/2)}{\partial c} \vline_{c = c_\textrm{bayes}} = 0.$$
-\item[2.] For certain values of $\beta^g_0$ and $\beta^g_1$, $c_\textrm{bayes}$ is a maximizer of the bias (Figure \ref{thresholding_theoretical}a, left); for other values of $\beta^g_0$ and $\beta^g_1$, $c_\textrm{bayes}$ is a minimizer of the bias (Figure \ref{thresholding_theoretical}a, right).
-\item[3.] The limit of the bias in $c$ is $1/2$, i.e. 
-$$\lim_{c \to \infty} b(\beta^g_0, \beta^g_1, c, 1/2) = 1/2.$$ In other words, we always can set the threshold to a large number and achieve a bias of $1/2$ (Figure \ref{thresholding_theoretical}a, all panels).
+\item \begin{proposition}
+The Bayes-optimal threshold $c_\textrm{bayes}$ is a critical value of the limiting bias $b$ (holding other parameters fixed; Figure \ref{thresholding_theoretical}a, all panels).
+\end{proposition}
+For certain values of $\beta^g_0$ and $\beta^g_1$, $c_\textrm{bayes}$ is a maximizer of the bias (Figure \ref{thresholding_theoretical}a, left); for other values of $\beta^g_0$ and $\beta^g_1$, $c_\textrm{bayes}$ is a minimizer of the bias (Figure \ref{thresholding_theoretical}a, right).
+\item
+\begin{proposition}
+The limit of the bias in $c$ is $1/2$, i.e. 
+$$\lim_{c \to \infty} b(\beta^g_0, \beta^g_1, c, 1/2) = 1/2.$$
+\end{proposition}
+ In other words, we always can set the threshold to a large number and achieve a bias of $1/2$ (Figure \ref{thresholding_theoretical}a, all panels).
 This fact establishes an upper bound on the bias of thresholded regression (under optimal threshold selection).
-\item[4.] For $\beta^g_1 < 2 \Phi(3/4) \approx 1.35$, $c = c_\textrm{bayes}$ yields a \textit{larger} bias than $c = \infty$ (Figure \ref{thresholding_theoretical}a, left); for $\beta^g_1 > 2 \Phi(3/4)$, $c_\textrm{bayes}$ yields a \textit{smaller} bias than $c = \infty$ (Figure \ref{thresholding_theoretical}a, right); finally, for $\beta^g_1 = 2 \Phi(3/4)$, $c = c_\textrm{bayes}$ and $c = \infty$ both yield a bias of $1/2$ (Figure \ref{thresholding_theoretical}a, middle).
-\item[5.] The bias decreases monotonically as $\beta_1^g$ increases (Figure \ref{thresholding_theoretical}a; fix threshold and scan panels from left to right).
+ \item
+ \begin{proposition}
+ For $\beta^g_1 < 2 \Phi(3/4) \approx 1.35$, $c = c_\textrm{bayes}$ yields a \textit{larger} bias than $c = \infty$ (Figure \ref{thresholding_theoretical}a, left); for $\beta^g_1 > 2 \Phi(3/4)$, $c_\textrm{bayes}$ yields a \textit{smaller} bias than $c = \infty$ (Figure \ref{thresholding_theoretical}a, right); finally, for $\beta^g_1 = 2 \Phi(3/4)$, $c = c_\textrm{bayes}$ and $c = \infty$ both yield a bias of $1/2$ (Figure \ref{thresholding_theoretical}a, middle).
+ \end{proposition}
+\item \begin{proposition}
+Limiting bias decreases monotonically as $\beta_1^g$ increases (Figure \ref{thresholding_theoretical}a; fix threshold and scan panels from left to right).
+\end{proposition}
 This is consistent with the intuition that the problem becomes easier as the gRNA mixture distribution becomes increasingly well-separated.
-\item[6.] For all $\beta^g_0, \beta^g_1,$ and $c \in \R,$ the bias is strictly greater than $0$ and less than $2$ (Figure \ref{thresholding_theoretical}a, all panels).
-Therefore, the thresholding method suffers from attenuation bias.% a common phenomenon in errors-in-variables models.
+\item 
+\begin{proposition} For all $\beta^g_0, \beta^g_1,$ and $c \in \R,$ the bias is strictly greater than $0$ and less than $2$ (Figure \ref{thresholding_theoretical}a, all panels). 
+\end{proposition}
+Therefore, the thresholding method suffers from attenuation bias.
 \end{itemize}
@@
 
-Next, we generalize the expression for bias when the threshold is large to arbitrary $\pi \in [0,1/2]$. The following simple relationship holds:
-$$ \lim_{ c \to \infty } b(\beta^g_0, \beta^g_1, c, \pi) = \pi.$$ In other words, when the perturbation probability is equal to $\pi$, setting the threshold to a large number yields a bias of $\pi$ (Figure \ref{thresholding_theoretical}b).
-We can build intuition for this result by considering an extreme example.
+Next, we generalize the expression for bias when the threshold is large to arbitrary $\pi \in [0,1/2]$:
+\begin{proposition}
+The following relationship holds:
+$$ \lim_{ c \to \infty } b(\beta^g_0, \beta^g_1, c, \pi) = \pi.$$
+\end{proposition}
+In other words, when the perturbation probability is equal to $\pi$ and the threshold large, the bias is $\pi$.
+We can understand this result intuitively by considering an extreme example.
 When $\pi$ is very small (e.g., $\pi = 0.01$), almost all cells are unperturbed.
-Therefore, in selecting a large threshold, one correctly classifies nearly all unperturbed cells as unperturbed; on the other hand, the \textit{perturbed} cells that one misclassifies as \textit{unperturbed} are swamped in number by the truly unperturbed cells, resulting in a small bias.
+Therefore, if we select a large threshold, we correctly classify nearly all unperturbed cells as unperturbed; on the other hand, the \textit{perturbed} cells that we misclassify as \textit{unperturbed} are swamped in number by the truly unperturbed cells, resulting in a small bias.
 
@@
 
-Finally, to shed light on limitations of the large threshold selection strategy, we study the variance of the thresholding estimator.
+% Finally, to shed light on limitations of the large threshold selection strategy, we study the variance of the thresholding estimator.
 Let $v$ denote the limiting variance of scaled estimator $\sqrt{n} \hat{\beta}^m_1$, i.e.,
-$$v(\beta^g_0, \beta^g_1, c, \pi) = \lim_{n \to \infty} \V(\sqrt{n} \hat{\beta}^m_1).$$ The limiting variance of $\sqrt{n} \hat{\beta}^m_1$ grows without bound as the threshold $c$ tends to infinity: 
-$$ \lim_{c \to \infty} v(\beta^g_0, \beta^g_1, c, \pi) = \infty.$$ Thus, while selecting a large threshold is effective from a bias reduction standpoint when $\pi$ is small, it causes the variance to explode.
+% $$v(\beta^g_0, \beta^g_1, c, \pi) = \lim_{n \to \infty} \V(\sqrt{n} \hat{\beta}^m_1).$$ The limiting variance of $\sqrt{n} \hat{\beta}^m_1$ grows without bound as the threshold $c$ tends to infinity:
+% $$ \lim_{c \to \infty} v(\beta^g_0, \beta^g_1, c, \pi) = \infty.$$ Thus, while selecting a large threshold is effective from a bias reduction standpoint when $\pi$ is small, it causes the variance to explode.
 
@@
 $$ \hat{\beta}^m_1 =  \frac{ \sum_{i=1}^n \hat{p}_i m_i }{ \sum_{i=1}^n \hat{p}_i^2}.$$ We uncover an exact bias-variance tradeoff as a function of the selected threshold (Figure \ref{thresholding_theoretical}c).
+
+\begin{proposition}
+The limiting distribution is:
+$$\sqrt{n}(\hat{\beta}^m - \beta^m) \xrightarrow{d} N(0, v).$$
+\end{proposition}
+
 The bias-variance analysis reveals that the best strategy for minimizing mean-squared error is to select a threshold that induces moderate bias.
@@
 Plugging in (\ref{dh_dc_bayes}) and (\ref{dg_dc_bayes}) to (\ref{quotient_rule}), we find that 
-$$D_2(\beta^g_1, c_\textrm{bayes}, 1/2) = 0.$$
+$$D_2(\beta^g_1, c_\textrm{bayes}, 1/2) = 0.$$ It follows that $c_\textrm{bayes}$ is a critical value of the bias function $b$.
 
-% $$g'(\cdot, c) = - f( \beta^g_0 + \beta^g_1 - c ) + f(\beta^g_0 - c).$$ Second, 
-% \begin{multline*}
-% h'(\cdot, c) = -[f(\beta^g_0 - c)  + f( \beta^g_0 + \beta^g_1 - c )]  \left[\left( \int_{-\infty}^{c - \beta^g_0} f \right) + \left(\int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \right] \\ +[ f(c - \beta^g_0) + f(c - \beta^g_0 - \beta^g_1) ] \left[ \left( \int_{-\infty}^{ \beta_0^g - c} f \right) + \left( \int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \right].
-% \end{multline*}
-% Evaluating $g'(\cdot, c)$ and $h'(\cdot, c)$ at $c = c_\textrm{bayes} $, we find that
-% $$g'(\beta^g_0 + (1/2) \beta^g_1) = -f(\beta^g_1/2) + f(- \beta^g_1/2) = -f(\beta^g_1/2) + f(\beta^g_1/2) = 0,$$ where the second equality follows from the evenness of $f$. Similarly, 
-% \begin{multline*}
-% h'(\beta^g_0 + (1/2) \beta^g_1) = - \left[ f( -\beta^g_1/2) + f( \beta^g_1/2 ) \right] \left[ \int_{-\infty}^{ \beta^g_1/2} f + \int_{-\infty}^{ -\beta^g_1/2} f \right] + \\ \left[ f( \beta^g_1/2) + f( -\beta^g_1/2 ) \right] \left[ \int_{-\infty}^{ -\beta^g_1/2} f + \int_{-\infty}^{ \beta^g_1/2 f} \right] = 0.
-% \end{multline*}
-% We conclude that
-% $$ \frac{\partial \gamma(\beta^g_0, \beta^g_1, c, 1/2)}{ \partial c} \vline_{[c = c_\textrm{bayes}]} = \frac{0\cdot h(\cdot, c_\textrm{bayes}) - g(\cdot, c_\textrm{bayes}) \cdot 0}{ h(\cdot, c_\textrm{bayes})^2} = 0.$$ Finally, because asymptotic bias $b$ can be expressed in terms of $\gamma$ as $$b(\beta^g_0.
-\beta^g_1, c, \pi) = \beta^m_1 \left[ 1 - \gamma(\beta^g_0, \beta^g_1, c, \pi) \right],$$ it follows that $c = c_\textrm{bayes}$ is a critical value of $b$ for $\pi = 1/2$ and fixed $(\beta^g_0, \beta^g_1)$.
+\subsubsection{Comparison of Bayes-optimal decision boundary to large threshold}
 
-% \subsubsection{Limit in $\beta^g_1$}
+We study the bias produced by setting the threshold to a large number versus setting the threshold to the Bayes-optimal decision boundary.
+Let $r: \R \to \R$ be the value of attenuation function evaluated at the Bayes-optimal decision boundary $c_\textrm{bayes} = \beta^g_0 + (1/2) \beta^g_1$:
+\begin{multline*}
+r(\beta^g_1) = \gamma(\beta^g_1, \beta^g_0 + (1/2)\beta^g_1, 1/2) = \frac{\Phi(\beta^g_1/2) - \Phi(-\beta^g_1/2)}{\left[\Phi(-\beta^g_1/2) + \Phi( \beta^g_1/2) \right] \left[\Phi(\beta^g_1/2) + \Phi( -\beta^g_1/2)\right]} \\ = \frac{\int_{-\beta^g_1/2}^{\beta^g_1/2} f}{\left[ 1 - \Phi(\beta^g_1/2) + \Phi(\beta^g_1/2) \right]\left[ \Phi(\beta^g_1/2) + 1 - \Phi(\beta^g_1/2) \right]} = 2 \int_{0}^{\beta^g_1/2} f = 2 \Phi(\beta^g_1/2) - 1.
+\end{multline*}
+We set $r$ to $1/2$ and solve for $\beta^g_1$:
+$$ 2\Phi(\beta^g_1/2) -1 = 1/2 \iff \Phi(\beta^g_1/2) = 3/4 \iff \beta^g_1 = 2 \Phi^{-1}(3/4) \approx 1.35.$$ The bias induced selecting a large threshold is $1/2$, i.e.
+$$\gamma(\beta^g_1, \infty, 1/2) = 1/2.$$
+Additionally, $r$ is a strictly increasing function.
+Therefore, $\gamma(\beta^g_1, c_\textrm{bayes},1/2) < \gamma(\beta^g_1, \infty, 1/2)$ on $\beta^g_1 \in [0, 2\Phi^{-1}(3/4))$; $\gamma(\beta^g_1, c_\textrm{bayes},1/2) = \gamma(\beta^g_1, \infty, 1/2)$ for $\beta^g_1 = 2\Phi^{-1}(3/4)$; and $\gamma(\beta^g_1, c_\textrm{bayes},1/2) > \gamma(\beta^g_1, \infty, 1/2)$ on $\beta^g_1 \in (2\Phi^{-1}(3/4), \infty)$.
+Because $b(\beta^g_1, c, 1/2) = \beta^m_1(1 - \gamma(\beta^g_1, c, 1/2))$ similar results hold for the bias (but with the inequalities flipped).
 
-% We compute the limit of $\gamma(\beta^g_0, \beta^g_1, c, 1/2)$ in $\beta^g_1.$ We have that
+\subsubsection{Limit in $\beta^g_1$}
+We compute the limit of $\gamma$ in $\beta^g_1$ given $\pi = 1/2$.  First,
+\begin{multline*}
+ \lim_{\beta^g_1 \to \infty} \gamma(\beta^g_1, c, 1/2) = \frac{ 1 - \int_{-\infty}^{\beta^g_0 - c} f}{\left[ 1 + \int_{-\infty}^{ \beta^g_0 - c} f \right] \left[\int_{-\infty}^{ c - \beta^g_0} f \right]} \\ = \frac{\int_{-\infty}^{c - \beta^g_0} f }{ \left[1 + \int_{-\infty}^{ \beta^g_0 - c} f \right] \left[\int_{-\infty}^{ c - \beta^g_0} f \right]} = \frac{1}{1 + \int_{-\infty}^{\beta^g_0 - c} f} < 1.
+\end{multline*}
+Similarly,
+\begin{multline*}
+\lim_{\beta^g_1 \to -\infty} \gamma(\beta^g_1, c, 1/2) = \frac{ -\int_{-\infty}^{\beta^g_0 - c} f }{\left[ \int_{-\infty}^{\beta^g_0 - c} f \right] \left[ \int_{-\infty}^{c - \beta^g_0} f + 1 \right]} = \frac{-1}{1 + \int_{-\infty}^{c - \beta^g_0} f} > -1.
+\end{multline*}
 
-% \begin{multline*}
- %\lim_{\beta^g_1 \to \infty } \gamma(\beta^g_0, \beta^g_1, c, 1/2) = \frac{ 1 - \int_{-\infty}^{\beta^g_0 - c } f}{\left[ 1 + \int_{-\infty}^{ \beta^g_0 - c} f \right] \left[\int_{-\infty}^{ c - \beta^g_0} f \right]} \\ = \frac{\int_{-\infty}^{c - \beta^g_0} f }{ \left[1 + \int_{-\infty}^{ \beta^g_0 - c} f \right] \left[\int_{-\infty}^{ c - \beta^g_0} f \right]} = \frac{1}{1 + \int_{-\infty}^{\beta^g_0 - c} f} < 1.
-% \end{multline*}
+\subsubsection{Monotonicity in $\beta^g_1$}
+We show that $\gamma$ is monotonically increasing in $\beta^g_1$ for $\pi = 1/2$ and given $(\beta^g_1, \beta^g_0)$.
+We begin by stating and proving two lemmas.
+The first lemma establishes an inequality that will serve as the basis for the proof.
 
-% Similarly,
-% \begin{multline*}
-% \lim_{\beta^g_1 \to -\infty} \gamma(\beta^g_0, \beta^g_1, c, 1/2) = \frac{ -\int_{-\infty}^{\beta^g_0 - c} f }{\left[ \int_{-\infty}^{\beta^g_0 - c} f \right] \left[ \int_{-\infty}^{c - \beta^g_0} f + 1 \right]} = \frac{-1}{1 + \int_{-\infty}^{c - \beta^g_0} f} > -1.
-% \end{multline*}
+\begin{lemma}
+The following inequality holds: 
+\begin{multline}\label{basic_ineq_cp}
+\left[\Phi(\beta^g_0 - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right] \\ \cdot \left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c) + \Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g) \right] \\ \geq \left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c)\right]\left[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g)\right].
+\end{multline}
+\end{lemma}
+\textbf{Proof}: We take cases on the sign on $\beta^g_1$.
 
-% \subsubsection{Monotonicity in $\beta^g_1$}
-% We prove that $\gamma$ is monotonic in $\beta_1^g$ for given $(\beta^g_0, c)$ and $\pi = 1/2$. Recall that
-% $$\gamma(\beta^g_0, \beta^g_1, c, 1/2) = \frac{ g(\beta^g_1, c) }{ h(\beta^g_1, c)}.$$ By the quotient rule,
-% $$D_2 \gamma (\beta^g_0, \beta^g_1, c, 1/2) = \frac{D_1 g( \beta^g_1, c) h(\beta^g_1, c) - g(\beta^1_g, c) D_1 h(\beta^g_1, c) }{ [ h(\beta^g_1, c)^2]}.$$ The derivative $D_1 g(\beta^g_1, c)$ of $g$ is
-% $$D_1 g(\beta^g_1, c) = f( \beta^g_0 + \beta^g_1 - c ).$$
+\noindent
+\underline{Case 1}: $\beta^1_g < 0$.
+Then $ \beta^g_1 + (\beta^g - c) < (\beta^g_0 - c),$ implying $\Phi(\beta^g_0 + \beta^g_1 - c) < \Phi(\beta^g_0 - c),$ or $[\Phi(\beta^g_0 + \beta^g_1 - c) - \Phi(\beta^g_0 - c)] < 0.$ Moreover, $[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g)]$ is positive.
+Therefore, the right-hand side of (\ref{basic_ineq_cp}) is negative.
+
+Turning our attention of the left-hand side of (\ref{basic_ineq_cp}), we see that
+\begin{equation}\label{basic_ineq_cp_2}
+\Phi(\beta^g_0 + \beta^g_1 - c) + \Phi( c - \beta^g_0 - \beta^g_1) = 1 -\Phi(\beta^g_0 + \beta^g_1 - c) + \Phi( c - \beta^g_0 - \beta^g_1) = 1.
+\end{equation}
+Additionally, $\Phi(\beta^g_0 - c) < 1$ and $ \Phi(c - \beta^g_0) > 0$. Combining these facts with (\ref{basic_ineq_cp_2}), we find that
+$$ \left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c) + \Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g) \right] > 0.
+$$ Finally, because $\left[\Phi(\beta^g_0 - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right] > 0,$ the entire left-hand side of (\ref{basic_ineq_cp}) is positive.
+
+\noindent
+\underline{Case 2}: $\beta^1_g \geq 0.$  We will show that the first term on the LHS of (\ref{basic_ineq_cp}) is greater than the first term on the RHS of (\ref{basic_ineq_cp}), and likewise that the second term on the LHS is greater than the second term on the RHS, implying the truth of the inequality.
+Focusing on the first term, the positivity of $\Phi(\beta^g_0 -c)$ implies that
+$$ \Phi(\beta^g_0 - c) \geq - \Phi(\beta^g_0 - c),$$ and so
+$$ \Phi(\beta^g_0 - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \geq \Phi(\beta^g_0 - \beta^g_1 - c) - \Phi(\beta^g_0 - c).$$
+Next, focusing on the second term, $\beta^g_1 \geq 0$ implies that 
+\begin{equation}\label{basic_ineq_cp_3}
+\beta^g_1 + \beta^g_0 - c \geq \beta^g_0 - c \implies \Phi(\beta^g_1 + \beta^g_0 - c) - \Phi(\beta^g_0 - c) \geq 0.
+\end{equation}
+Adding $\Phi(c - \beta^g_0) + \Phi(c - \beta^g_0 - \beta^g_1)$ to both sides of (\ref{basic_ineq_cp_3}) yields
+\begin{multline*}
+\Phi(\beta^g_1 + \beta^g_0 - c) - \Phi(\beta^g_0 - c) + \Phi(c - \beta^g_0) + \Phi(c - \beta^g_0 - \beta^g_1) \\ \geq \Phi(c - \beta^g_0) + \Phi(c - \beta^g_0 - \beta^g_1). \textrm{ } \square
+\end{multline*}
+The second lemma establishes the derivatives of the functions $2\cdot g$ and $4 \cdot h$.
+\begin{lemma}
+The derivatives in $\beta^g_1$ of $2\cdot g$ and $4\cdot h$ are
+\begin{equation}\label{dg_dbeta}
+\textcolor{violet}{D_1[2g(\beta^g_1, c, 1/2)] = f(\beta^g_0 + \beta^g_1 - c)}
+\end{equation}
+and
+\begin{multline}\label{dh_dbeta}
+\textcolor{teal}{D_1[4h(\beta^g_1, c, 1/2)] = f(\beta^g_0 + \beta^g_1 - c) \left[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g) \right]} \\ \textcolor{teal}{- f(\beta^g_0 + \beta^g_1 - c) \left[\Phi(\beta_0^g - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right]}.\end{multline}
+\end{lemma}
+\textbf{Proof}: Apply the FTC and product rule. $\square$
+
+We move onto the proof of the monotonicity of $\gamma$ in $\beta^g_1$.
+Subtracting $$\left[\Phi(\beta_0^g - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right]\left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c)\right]$$ from both sides of (\ref{basic_ineq_cp}) and multiplying by $f(\beta^g_0 + \beta^g_1 - c) > 0$ yields
+\begin{multline}\label{basic_ineq_cp_4}
+\textcolor{violet}{f(\beta^g_0 + \beta^g_1 - c)} \textcolor{red}{ \left[\Phi(\beta^g_0 - c) + \Phi\left(\beta^g_0 + \beta^g_1 - c \right) \right] \left[ \Phi(c - \beta^g_0) + \Phi(c - \beta^g_0 - \beta^g_1) \right]}  \\ \geq \textcolor{teal}{f(\beta^g_0 + \beta^g_1 - c) \left[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g)\right]}\textcolor{blue}{\left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c)\right]} \\ -\textcolor{teal}{f(\beta^g_0 + \beta^g_1 - c)   \left[\Phi(\beta_0^g - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right]} \textcolor{blue}{\left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c)\right]}.
+\end{multline}
+Next, recall that
+\begin{equation}\label{def_2g}
+\textcolor{blue}{2g(\beta^g_1,c,1/2) = \Phi(\beta^g_0 + \beta^g_1 - c) - \Phi(\beta^g_0 - c)}.
+\end{equation}
+and
+\begin{equation}\label{def_4h}
+\textcolor{red}{4h(\beta^g_1, c, 1/2) = \left[ \Phi(\beta^g_0 - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right] \left[\Phi(c - \beta^g_0) + \Phi( c - \beta^g_0 - \beta^g_1) \right]}.
+\end{equation}
+Substituting (\ref{dg_dbeta}, \ref{dh_dbeta}, \ref{def_2g}, \ref{def_4h}) into (\ref{basic_ineq_cp_4}) produces
+\begin{equation*}
+\textcolor{violet}{D_1[2g(\beta^g_1, c, 1/2)]}\textcolor{red}{4h(\beta^g_1, c, 1/2)} \geq \textcolor{blue}{2g(\beta^g_1, c, 1/2)}\textcolor{teal}{D_1[4h(\beta^g_1, c, 1/2)]},
+\end{equation*}
+or 
+\begin{equation}\label{basic_ineq_cp_5}
+\textcolor{violet}{D_1[2g(\beta^g_1, c, 1/2)]}\textcolor{red}{4h(\beta^g_1, c, 1/2)} - \textcolor{blue}{2g(\beta^g_1, c, 1/2)}\textcolor{teal}{D_1[4h(\beta^g_1, c, 1/2)]} \geq 0.
+\end{equation}
+The quotient rule implies that
+\begin{multline}\label{d_gamma_d_beta}
+D_1 \gamma(\beta^g_1, c, 1/2) \\ = \frac{ \textcolor{violet}{D_1[2g(\beta^g_1, c, 1/2)]}\textcolor{red}{4h(\beta^g_1, c, 1/2)} - \textcolor{blue}{2g(\beta^g_1, c, 1/2)}\textcolor{teal}{D_1[4h(\beta^g_1, c, 1/2)]} }{[4h(\beta^g_1, c, 1/2)]^2}.
+\end{multline}
+We conclude by (\ref{basic_ineq_cp_5}) and (\ref{d_gamma_d_beta}) that $\gamma(\cdot, c, 1/2)$ is monotonically increasing in its first argument.
+
+\subsubsection{Strict attenuation bias}
+
+For given $(\beta^g_0, c)$, we have that
+$$ \lim_{\beta^g_1 \to \infty} \gamma(\beta^g_1, c, 1/2, \beta^g_0) = \frac{1}{1 + \Phi(\beta^g_0 - c)},$$ and $$\lim_{\beta^g_1 \to -\infty} \gamma(\beta^g_1, c, 1/2, \beta^g_0) = -\frac{1}{1 + \Phi(c - \beta^g_0)}.$$
+Moreover, $\gamma(\beta^g_1, c, 1/2, \beta^g_0)$ is monotonically increasing in $\beta^g_1$. It follows that 
+$$-1 < -\frac{1}{1 + \Phi(c - \beta^g_0)} \leq \gamma(\beta^g_1, c, 1/2, \beta^g_0) \leq \frac{1}{1 - \Phi(\beta^g_0 - c)} < 1$$ for all $\beta^g_1 \in \R$.
+But $\beta^g_0$ and $c$ were chosen arbitrarily, and so
+$$-1 < \gamma(\beta^g_1, c, 1/2, \beta^g_0) < 1$$ for all $(\beta^g_1, c, \beta^g_0) \in \R^3$.
+
+\subsubsection{Bias-variance decomposition in no-intercept model}
 

### 2021-10-22 text/outline/outline.tex
add bv decomp

@@
 \item 
-\begin{proposition} For all $\beta^g_0, \beta^g_1,$ and $c \in \R,$ the bias is strictly greater than $0$ and less than $2$ (Figure \ref{thresholding_theoretical}a, all panels). 
+\begin{proposition} For all $\beta^g_0, \beta^g_1,$ and $c \in \R,$ the incurs strict attenuation bias (Figure \ref{thresholding_theoretical}a, all panels).
 \end{proposition}
-Therefore, the thresholding method suffers from attenuation bias.
 \end{itemize}
@@
 
-Next, we generalize the expression for bias when the threshold is large to arbitrary $\pi \in [0,1/2]$:
+Next, we generalize the expression for bias to arbitrary $\pi \in [0,1/2]$:
 \begin{proposition}
@@
 \end{proposition}
-In other words, when the perturbation probability is equal to $\pi$ and the threshold large, the bias is $\pi$.
+In other words, if the perturbation probability is $\pi$, and if we set the threshold to a large number, then the bias also is $\pi$.
 We can understand this result intuitively by considering an extreme example.
 When $\pi$ is very small (e.g., $\pi = 0.01$), almost all cells are unperturbed.
-Therefore, if we select a large threshold, we correctly classify nearly all unperturbed cells as unperturbed; on the other hand, the \textit{perturbed} cells that we misclassify as \textit{unperturbed} are swamped in number by the truly unperturbed cells, resulting in a small bias.
+Therefore, in selecting a large threshold, we correctly classify nearly all unperturbed cells as unperturbed; on the other hand, the \textit{perturbed} cells that we misclassify as \textit{unperturbed} are swamped in number by the truly unperturbed cells, resulting in a small bias.
 
@@
 
-% Finally, to shed light on limitations of the large threshold selection strategy, we study the variance of the thresholding estimator.
-Let $v$ denote the limiting variance of scaled estimator $\sqrt{n} \hat{\beta}^m_1$, i.e.,
-% $$v(\beta^g_0, \beta^g_1, c, \pi) = \lim_{n \to \infty} \V(\sqrt{n} \hat{\beta}^m_1).$$ The limiting variance of $\sqrt{n} \hat{\beta}^m_1$ grows without bound as the threshold $c$ tends to infinity:
-% $$ \lim_{c \to \infty} v(\beta^g_0, \beta^g_1, c, \pi) = \infty.$$ Thus, while selecting a large threshold is effective from a bias reduction standpoint when $\pi$ is small, it causes the variance to explode.
-
-To more precisely characterize the bias-variance tradeoff of the thresholding method, we consider a slightly simpler model: we set $\beta^m_0$ and $\beta^g_0$ to zero in (\ref{theoretical_model}) and study the no-intercept least squares estimator
-$$ \hat{\beta}^m_1 =  \frac{ \sum_{i=1}^n \hat{p}_i m_i }{ \sum_{i=1}^n \hat{p}_i^2}.$$ We uncover an exact bias-variance tradeoff as a function of the selected threshold (Figure \ref{thresholding_theoretical}c).
-
+To shed light on possible limitations of the large threshold selection strategy, we derive an exact bias-variance decomposition for the thresholding estimator.
+We consider a slightly simpler, no-intercept version of (\ref{theoretical_model}):
+\begin{equation}\label{theoretical_model_no_int}
+\begin{cases}
+m_i = \beta_m p_i + \ep_i \\
+g_i = \beta_g p_i + \tau_i \\
+p_i \sim \textrm{Bern}(\pi) \\
+\ep_i, \tau_i \sim N(0,1) \\
+p_i \indep \tau_i \indep \ep_i,
+\end{cases}
+\end{equation}
+The thresholding estimator $\hat{\beta}_m$ is given by
+\begin{equation}\label{thresh_estimator_no_int}
+ \hat{\beta}_m = \frac{ \sum_{i=1}^n \hat{p}_i m_i }{\sum_{i=1}^n \hat{p}_i^2}. 
+\end{equation}
 \begin{proposition}
-The limiting distribution is:
-$$\sqrt{n}(\hat{\beta}^m - \beta^m) \xrightarrow{d} N(0, v).$$
+The limiting distribution of $\hat{\beta}_m$ is
+$$\sqrt{n}\left(\hat{\beta}_m - l \right) \xrightarrow{d} N\left(0, v\right).$$
 \end{proposition}
-
-The bias-variance analysis reveals that the best strategy for minimizing mean-squared error is to select a threshold that induces moderate bias.
-(A downside of this approach, of course, is that constructing valid confidence intervals becomes more challenging.)
+The bias decreases and the variance increases as the threshold tends to infinity, consistent with the intuition that a large threshold yields a low misclassification rate at the cost of a small ``effective sample size.'' The best strategy for maximizing estimation accuracy (as quantified by mean squared error) is to select a threshold that induces moderate bias.
+A downside of this approach, however, is that constructing valid confidence intervals becomes more challenging.
 
@@
 \item[2.] Expectation of $\hat{p}_i$: Let $f$ denote the $N(0,1)$ density. Then \begin{multline*}
-\E[\hat{p}_i] = \P\left[\hat{p}_i = 1\right] = \P\left[\beta^g_0 + \beta^g_1 + \tau_i \geq c \right] = \\ \textrm{(By LOTP) } \P\left[ \beta^g_0 + \tau_i \geq c \right]\P\left[p_i = 0\right] + \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \P[p_i = 1] \\ = \P\left[ \tau_i \geq c - \beta^g_0\right](1- \pi) + \P\left[ \tau_i \geq c - \beta^g_1 - \beta^g_0 \right](\pi) \\ = \left(\int_{c - \beta^g_0}^\infty f\right) (1 - \pi) + \left( \int_{c - \beta^g_1 - \beta^g_0}^\infty f \right)(\pi) = \zeta(1-\pi) + \omega \pi.
+\E[\hat{p}_i] = \P\left[\hat{p}_i = 1\right] = \P\left[\beta^g_0 + \beta^g_1 p_i + \tau_i \geq c \right] = \\ \textrm{(By LOTP) } \P\left[ \beta^g_0 + \tau_i \geq c \right]\P\left[p_i = 0\right] + \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \P[p_i = 1] \\ = \P\left[ \tau_i \geq c - \beta^g_0\right](1- \pi) + \P\left[ \tau_i \geq c - \beta^g_1 - \beta^g_0 \right](\pi) \\ = \left(\int_{c - \beta^g_0}^\infty f\right) (1 - \pi) + \left( \int_{c - \beta^g_1 - \beta^g_0}^\infty f \right)(\pi) = \zeta(1-\pi) + \omega \pi.
 \end{multline*}
 \item[3.] Expectation of $\hat{p}_i p_i$: 
-$$\E\left[ \hat{p}_i p_i \right] = \E\left[ \hat{p}_i | p_i = 1 \right] \P\left[ p_i =1 \right] = \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \pi = \omega \pi.$$
+$$\E\left[ \hat{p}_i p_i \right] = \E\left[\hat{p}_i | p_i = 1 \right] \P\left[ p_i =1 \right] = \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \pi = \omega \pi.$$
 \item[4.] Expectation of $\hat{p}_i m_i$:
@@
 
+We prove the bias-variance decomposition for the no-intercept model (\ref{theoretical_model_no_int}). Define $l$ (for ``limit'') by
+$$l = \beta_m \left(\frac{\omega \pi}{\zeta(1-\pi) + \omega \pi}\right),$$ where
+$$
+\begin{cases}
+\omega = \int_{c - \beta_g}^\infty f = 1 - \Phi(\beta_g - c) \\
+\zeta = \int_{c}^\infty f = 1 - \Phi(-c).
+\end{cases}
+$$
+We have that
+\begin{equation*}
+\hat{\beta}_m - l = \frac{\sum_{i=1}^n \hat{p}_i m_i}{ \sum_{i=1}^n \hat{p}^2_i} - l = \frac{\sum_{i=1}^n \hat{p}_i m_i}{ \sum_{i=1}^n \hat{p}^2_i} - \frac{l \sum_{i=1}^n \hat{p}_i^2 }{ \sum_{i=1}^n \hat{p}_i^2} \\ = \frac{\sum_{i=1}^n \hat{p}_i(m_i - l \hat{p}_i)}{ \sum_{i=1}^n \hat{p}_i^2}.
+\end{equation*}
+Therefore,
+\begin{equation}\label{bc_decomp_1}
+\sqrt{n}(\hat{\beta}_m - l) = \frac{(1/\sqrt{n})\sum_{i=1}^n \hat{p}_i(m_i - l \hat{p}_i)}{(1/n)\sum_{i=1}^n \hat{p}_i^2}.
+\end{equation}
+Next, we compute the expectation and variance of $\hat{p}_i(m_i - l\hat{p}_i)$. To do so, we first compute several simpler quantities:
+\begin{enumerate}
+\item Expectation of $\hat{p}_i$: 
+\begin{multline*}
+\E[\hat{p}_i] = \P(p_i\beta_g + \tau_i \geq c) =  \P(\beta_g + \tau_i \geq c)\pi + \P(\tau_i \geq c)(1-\pi) \\ = \pi \omega + (1-\pi)\zeta.
+\end{multline*}
+\item Expectation of $\hat{p}_i p_i$: $$\E\left[\hat{p}_i p_i\right] = \E\left[\hat{p}_i | p_i = 1 \right]\P\left[p_i = 1\right] = \omega \pi.$$
+\item Expectation of $\hat{p}_i m_i$:
+\begin{multline*}
+\E[\hat{p}_i m_i] = \E\left[\hat{p}_i(\beta_m p_i + \ep_i)\right] = \E\left[\beta_m \hat{p}_i p_i + \hat{p}_i \ep_i \right] \\ = \beta_m \E\left[ \hat{p}_i p_i \right] + \E[\hat{p}_i]\E[\ep_i] = \beta_m \omega \pi + 0 = \beta_m \omega \pi.
+\end{multline*}
+\item Expectation of $\hat{p}_i m_i^2$: \begin{multline*}
+\E\left[\hat{p}_i m_i^2\right] = \E \left[ \hat{p}_i( \beta_m p_i + \ep_i )^2 \right] = \E\left[ \hat{p}_i \left( \beta_m^2 p_i^2 + 2 \beta_m p_i \ep_i + \ep_i^2 \right)  \right] \\ = \E\left[ \hat{p}_i p_i \beta^2_m + 2 \beta_m p_i \hat{p}_i \ep_i + \hat{p}_i \ep_i^2 \right] = \beta^2_m \E[ \hat{p}_i p_i] + 2 \beta_m \E[p_i\hat{p}_i] \E[\ep_i] + \E[\hat{p}_i] \E[ \ep^2_i ] \\ = \beta^2_m \E[ \hat{p}_i p_i] + \E[\hat{p}_i] = \beta^2_m \omega \pi + \E[ \hat{p}_i].
+\end{multline*}
+\end{enumerate}
+
+Now, we can compute the expectation and variance of $\hat{p}_i(m_i - l\hat{p}_i)$. First,
+\begin{multline}\label{bv_decomp_2}
+\E\left[\hat{p}_i(m_i - l\hat{p}_i) \right] = \E[\hat{p}_i m_i] - l \E[\hat{p}_i] \\ = \beta_m \omega \pi - \left(\frac{\beta_m \omega \pi}{\zeta (1-\pi) + \omega \pi}\right)[\zeta (1-\pi) + \omega \pi] = 0.
+\end{multline}
+Additionally,
+\begin{multline}\label{bv_decomp_3}
+\V\left[\hat{p}_i(m_i - l\hat{p}_i)\right] = \E\left[\hat{p}_i^2(m_i - l\hat{p}_i)^2\right] - (\E\left[ \hat{p}_i(m_i - l\hat{p}_i)\right])^2 \\ = \E\left[ \hat{p}_i m_i^2\right] - 2l \E[m_i\hat{p}_i] +l^2 \E[\hat{p}_i]= \beta^2_m \omega \pi + \E[ \hat{p}_i] -2l \beta_m \omega \pi + l^2 \E[\hat{p}_i] \\ = \beta_m\omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2).
+\end{multline}
+Therefore, by CLT, (\ref{bv_decomp_2}), and (\ref{bv_decomp_3}),
+\begin{equation}\label{bv_decomp_4}
+(1/\sqrt{n})\sum_{i=1}^n \hat{p}_i(m_i - l \hat{p}_i) \xrightarrow{d} N\left(0, \beta_m\omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2) \right).
+\end{equation}
+Next, by weak LLN,
+\begin{equation}\label{bv_decomp_5}
+(1/n) \sum_{i=1}^n \hat{p}_i^2 = (1/n) \sum_{i=1}^n \hat{p}_i \xrightarrow{P} \E[\hat{p}_i].
+\end{equation}
+Finally, by (\ref{bc_decomp_1}), (\ref{bv_decomp_4}), (\ref{bv_decomp_5}), and Slutsky's Theorem,
+$$ \sqrt{n}(\hat{\beta}_m - l) \xrightarrow{d} N\left(0, \frac{ \beta_m\omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2) }{\left(\E[\hat{p}_i]\right)^2} \right).$$ Thus, for large $n \in \N$, we find that
+$$ \begin{cases}
+\E [\hat{\beta}_m] \approx l, \\
+\V[\hat{\beta}_m] \approx \left[\beta_m\omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2)\right]/[n\E^2[\hat{p}_i]],
+\end{cases}$$
+completing the bias-variance decomposition.
+
 \subsection{Derivation of EM algorithm}

### 2021-10-23 text/outline/outline.tex
add more text to thresholding theory section

@@
 \begin{center}
-\textbf{CRISPR genome engineering, single-cell sequencing, and exponential family measurement error models}
+\textbf{CRISPR genome editing, single-cell sequencing, and exponential family measurement error models}
 \end{center}
@@
 \begin{abstract}
-CRISPR genome engineering and single-cell sequencing have transformed biological discovery.
+CRISPR genome editing single-cell sequencing have transformed biological discovery.
 Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases.
@@
 \end{equation}
-For a given threshold $c \in \R$, let $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ for $\beta^m_1$ is given by $$\hat{\beta}^m_1 = \frac{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}}) (m_i - \overline{m})}{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})}.$$
+For a given threshold $c \in \R$, the imputed perturbation assignment $\hat{p}_i$ is $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ for $\beta^m_1$ is given by $$\hat{\beta}^m_1 = \frac{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}}) (m_i - \overline{m})}{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})}.$$
 
 \begin{proposition}
- The limit in probability (as $n \to \infty$) of $\hat{\beta}^m_0$ is
+ The almost sure limit (as $n \to \infty$) of $\hat{\beta}^m_1$ is
 \begin{equation}\label{thresh_est_intercepts}
-\hat{\beta}^m_1 \xrightarrow{P} \beta^m_1 \left(\frac{ \pi( \omega - \E[ \hat{p}_i ])}{ \E[\hat{p}_i] (1 - \E[\hat{p}_i])}\right),
+\hat{\beta}^m_1 \xrightarrow{a.s.} \beta^m_1 \left(\frac{ \pi( \omega - \E[ \hat{p}_i ])}{ \E[\hat{p}_i] (1 - \E[\hat{p}_i])}\right),
 \end{equation}
@@
 $$\begin{cases}
-\omega = 1 - \Phi\left(c - \beta_1^g - \beta_0^g \right) ,\\ \zeta = 1 - \Phi\left( c - \beta^g_0 \right), \\ \E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi. 
+\E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi , \\
+\omega = \Phi\left(\beta_1^g + \beta_0^g -c \right) ,\\ \zeta = \Phi\left( \beta^g_0 - c \right). \\
 \end{cases}$$
@@
 Let $\gamma: \R^4 \to \R$ be defined by
-$$ \gamma( \beta^g_0, \beta^g_1, c, \pi ) = \frac{\pi (\omega - \E[\hat{p}_i])}{\E[\hat{p}_i] (1 - \E[\hat{p}_i]) }.$$ We call $\gamma$ the ``attenuation function.'' Observe that \begin{itemize}
+$$ \gamma(\beta^g_1, \pi, c, \beta^g_0) = \frac{\pi (\omega - \E[\hat{p}_i])}{\E[\hat{p}_i] (1 - \E[\hat{p}_i]) }.$$ We call $\gamma$ the ``attenuation function.'' Observe that \begin{itemize}
 \item[i.] $\gamma$ does not depend on $\beta^m_1$ or $\beta^m_0$, and
-\item[ii.] $\hat{\beta}^m_1 \xrightarrow{P} [\gamma(\beta_0^g, \beta_1^g, c, \pi)] \beta^m_1.$
+\item[ii.] $\hat{\beta}^m_1 \xrightarrow{a.s.} [\gamma(\beta_0^g, \beta_1^g, c, \pi)] \beta^m_1.$
 \end{itemize}
-Let $b: \R^4 \to \R$ be the asymptotic bias of $\hat{\beta}^m_1$.
-We can express $b$ in terms of $\gamma$ as $$b\left(\beta^g_0, \beta^g_1, c, \pi \right) = \beta^m_1 - \gamma( \beta_0^g, \beta_1^g, c, \pi) \beta^m_1 = \beta^m_1 \left[ 1- \gamma(\beta^g_0, \beta^g_1, c, \pi) \right].$$ The asymptotic bias vanishes when the attenuation function equals $1$.
+Let $b: \R^5 \to \R$ be the asymptotic bias of $\hat{\beta}^m_1$, i.e.
+$$
+b(\beta^g_1, \pi, c, \beta^g_0, \beta^m_1) = \lim_{n\to\infty} \left(\beta^m_1 - \E[\hat{\beta}^m_1]\right) = \beta^m_1 - \E \left( \lim_{a.s.} \hat{\beta}^m_1 \right) = \beta^m_1 - \lim_{a.s.} \hat{\beta}^m_1,
+$$
+where the $\lim_{a.s.}$ denotes a.s.\ convergence.
+\textcolor{red}{(HOW TO MAKE THE ABOVE RIGOROUS?)} We can express $b$ in terms of $\gamma$ as $$b(\beta^g_1, \pi, c, \beta^g_0, \beta^m_1) = \beta^m_1 - \gamma(\beta^g_1, \pi, c, \beta^g_0)  \beta^m_1 = \beta^m_1 \left[1 - \gamma(\beta^g_1, \pi, c, \beta^g_0)  \right].$$ The asymptotic bias vanishes when the attenuation function equals $1$.
 
@@
 To investigate the basic question of ``What is a good threshold selection strategy?'', we study the relationship between the asymptotic bias $b$ of $\hat{\beta}^m_1$ and the selected threshold $c$.
-For simplicity, we begin by setting the perturbation probability $\pi$ to $1/2$.
-Additionally, without loss of generality, we set the target of inference $\beta^m_1$ to $1$.
+For simplicity, we set the perturbation probability $\pi$ to $1/2$.
 Let $c_\textrm{bayes} \in \R$ be the Bayes-optimal decision boundary for classifying cells as perturbed or unperturbed, i.e.
-$$c_\textrm{bayes} = \argmin_{c \in \R} \P(\hat{p}_i \neq p_i).$$ Simple algebra shows that $c_\textrm{bayes} = \beta_0^g + (1/2) \beta^g_1.$ Below, we state several results relating the limiting bias $b$ of $\hat{\beta}^m_1$ to $c_\textrm{bayes}$ and $\beta^g_1$.
-We defer proofs to the appendix.
+$$c_\textrm{bayes} = \argmin_{c \in \R} \P(\hat{p}_i \neq p_i).$$ Simple algebra shows that $c_\textrm{bayes} = \beta_0^g + (1/2) \beta^g_1.$ Below, we state several results on the limiting bias $b$ of $\hat{\beta}^m_1$.
+We refer throughout to Figure \ref{thresholding_theoretical}a, which displays plots of asymptotic bias versus threshold for different values of $\beta^g_1$; the target of inference, $\beta^m_1$, is fixed to $1$ in these plots.
 
 \begin{itemize}
+\item 
+\begin{proposition}\label{prop:att_bias} Assume without loss of generality that $\beta^m_1 > 0.$ Fix $\pi = 1/2$. For all $(\beta^g_1, c, \beta^g_0) \in \R^3$,
+	$$b(\beta^g_1, 1/2, c, \beta^g_0, \beta^m_1) > 0.$$
+\end{proposition}
+The thresholding method suffers strict attenuation bias for all choices of the threshold and over all possible model parameters (Figure \ref{thresholding_theoretical}a).
+Attenuation bias is a common attribute of estimators that ignore measurement in errors-in-variables models \cite{Stefanski2000a}.
+
 \item \begin{proposition}
-The Bayes-optimal threshold $c_\textrm{bayes}$ is a critical value of the limiting bias $b$ (holding other parameters fixed; Figure \ref{thresholding_theoretical}a, all panels).
+For $\pi = 1/2$ and given $(\beta^g_1, \beta^g_0, \beta^m_1) \in \R^3$, the Bayes-optimal decision boundary $c_\textrm{bayes}$ is a critical value of the limiting bias function $b$, i.e.
+$$ \frac{\partial b}{\partial c}\left(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0, \beta^m_1\right) = 0.$$
 \end{proposition}
-For certain values of $\beta^g_0$ and $\beta^g_1$, $c_\textrm{bayes}$ is a maximizer of the bias (Figure \ref{thresholding_theoretical}a, left); for other values of $\beta^g_0$ and $\beta^g_1$, $c_\textrm{bayes}$ is a minimizer of the bias (Figure \ref{thresholding_theoretical}a, right).
+The Bayes-optimal decision boundary is an optimum (or possibly a saddle point) of the limiting bias function (Figure \ref{thresholding_theoretical}a, vertical blue lines).
+Interestingly, $c_\textrm{bayes}$ is in some cases a maximizer of the bias (Figure \ref{thresholding_theoretical}a, left) and in other cases a minimizer of the bias (Figure \ref{thresholding_theoretical}a, right).
 \item
-\begin{proposition}
-The limit of the bias in $c$ is $1/2$, i.e. 
-$$\lim_{c \to \infty} b(\beta^g_0, \beta^g_1, c, 1/2) = 1/2.$$
+\begin{proposition}\label{prop:limit}
+Fix $\pi = 1/2$. As the threshold $c$ tends to infinity, the asymptotic bias $b$ tends to $\beta_1^m/2$, i.e.
+$$\lim_{c \to \infty} b(\beta^g_1, 1/2, c, \beta^g_0, \beta^m_1) = \beta_1^m/2.$$
 \end{proposition}
- In other words, we always can set the threshold to a large number and achieve a bias of $1/2$ (Figure \ref{thresholding_theoretical}a, all panels).
-This fact establishes an upper bound on the bias of thresholded regression (under optimal threshold selection).
+We always can set the threshold to a large number and attain a bias of $\beta_1^m/2$ (Figure \ref{thresholding_theoretical}a, all panels).
+This result establishes an upper bound on the bias of thresholded regression (under optimal threshold selection strategy).
+
+ \item The following proposition compares the two threshold selection strategies introduced above (i.e., large number versus Bayes-optimal decision boundary) head-to-head.
+ \begin{proposition} Let $\beta^m_1 > 0$.
+For $\beta^g_1 \in [0, 2\Phi^{-1}(3/4))$, we have that $$b(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0, \beta^m_1) > b(\beta^g_1, 1/2, \infty, \beta^g_0, \beta^m_1).$$ For $\beta^g_1 = 2\Phi^{-1}(3/4)$, we have that $$ b(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0, \beta^m_1) = b(\beta^g_1, 1/2, \infty, \beta^g_0, \beta^m_1).$$ Finally, for $\beta^g_1 \in (2\Phi^{-1}(3/4), \infty)$, we have that
+ $$b(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0, \beta^m_1) < b(\beta^g_1, 1/2, \infty, \beta^g_0, \beta^m_1).$$
+ \end{proposition}
+In other words, setting the threshold to a large number yields a smaller bias when $\beta^g_1$ is small (i.e., $\beta^g_1 < 2\Phi^{-1}(3/4) \approx 1.35$; Figure \ref{thresholding_theoretical}a, left); setting the threshold to the Bayes-optimal decision boundary yields a smaller bias when $\beta^g_1$ is big (i.e., $\beta^g_1 > 2\Phi^{-1}(3/4)$; Figure \ref{thresholding_theoretical}a, right); and the two approaches coincide when $\beta^g_1$ is intermediate (i.e., $\beta^g_1 = 2\Phi^{-1}(3/4)$; Figure \ref{thresholding_theoretical}a, middle).
+
 \item
- \begin{proposition}
- For $\beta^g_1 < 2 \Phi(3/4) \approx 1.35$, $c = c_\textrm{bayes}$ yields a \textit{larger} bias than $c = \infty$ (Figure \ref{thresholding_theoretical}a, left); for $\beta^g_1 > 2 \Phi(3/4)$, $c_\textrm{bayes}$ yields a \textit{smaller} bias than $c = \infty$ (Figure \ref{thresholding_theoretical}a, right); finally, for $\beta^g_1 = 2 \Phi(3/4)$, $c = c_\textrm{bayes}$ and $c = \infty$ both yield a bias of $1/2$ (Figure \ref{thresholding_theoretical}a, middle).
+ \begin{proposition}\label{prop_monotonic} Fix $\pi = 1/2$. The limiting bias $b$ decreases monotonically in $\beta_1^g$, i.e.
+$$\frac{\partial b}{\partial(\beta^g_1)}\left(\beta^g_1, 1/2, c, \beta^g_0, \beta^m_1\right) \leq 0.$$
 \end{proposition}
-\item \begin{proposition}
-Limiting bias decreases monotonically as $\beta_1^g$ increases (Figure \ref{thresholding_theoretical}a; fix threshold and scan panels from left to right).
-\end{proposition}
-This is consistent with the intuition that the problem becomes easier as the gRNA mixture distribution becomes increasingly well-separated.
-\item 
-\begin{proposition} For all $\beta^g_0, \beta^g_1,$ and $c \in \R,$ the incurs strict attenuation bias (Figure \ref{thresholding_theoretical}a, all panels).
-\end{proposition}
+This result formalizes the intuition that the problem becomes easier as the gRNA mixture distribution becomes increasingly well-separated.
+To visualize Proposition (\ref{prop_monotonic}), one can fix a threshold (e.g., $c = 0$) and scan for bias across the panels.
 \end{itemize}
@@
 These results are subtle, but we can summarize them as follows.
-First, selecting a threshold that minimizes the bias is deceptively challenging, as there is no rule of thumb that we can apply universally (e.g., ``always choose the Bayes-optimal decision boundary'' or ``always choose a large number'').
-Second, even if we \textit{have} selected a good threshold, the thresholding method incurs a strict attenuation bias.
+First, selecting a threshold that minimizes the bias is challenging, as there is no rule of thumb that we can apply universally (e.g., ``always choose the Bayes-optimal decision boundary'' or ``always choose a large number'') due to the complexity of the bias function.
+Second, even if we \textit{have} selected a good threshold, we incur nonzero attenuation bias.
 
@@
 \end{center}
-
-Next, we generalize the expression for bias to arbitrary $\pi \in [0,1/2]$:
+We generalize the expression for bias to arbitrary $\pi \in [0,1/2]$:
 \begin{proposition}
-The following relationship holds:
-$$ \lim_{ c \to \infty } b(\beta^g_0, \beta^g_1, c, \pi) = \pi.$$
+As the threshold $c$ tends to infinity, the limiting bias $b$ tends to $\beta^m_1 \pi,$ i.e.
+$$ \lim_{ c \to \infty } b(\beta^g_1, \pi, c, \beta^g_0, \beta^m_1) = \beta^m_1 \pi.$$
 \end{proposition}
-In other words, if the perturbation probability is $\pi$, and if we set the threshold to a large number, then the bias also is $\pi$.
-We can understand this result intuitively by considering an extreme example.
-When $\pi$ is very small (e.g., $\pi = 0.01$), almost all cells are unperturbed.
+That is, if the perturbation probability is $\pi$, and if we set the threshold to a large number, then the asymptotic bias is $\beta^m_1 \pi$ (Figure \ref{thresholding_theoretical}b; $\beta^m_1$ fixed at $1$).
+We can understand this result intuitively by considering an extreme example: when $\pi$ is very small (e.g., $\pi = 0.01$), almost all cells are unperturbed.
 Therefore, in selecting a large threshold, we correctly classify nearly all unperturbed cells as unperturbed; on the other hand, the \textit{perturbed} cells that we misclassify as \textit{unperturbed} are swamped in number by the truly unperturbed cells, resulting in a small bias.
@@
 	\includegraphics[width=1\linewidth]{../../figures/thresholding_theoretical/plot}
-	\caption{\textbf{Theoretical challenges of thresholded regression.}}
+	\caption{\textbf{Theoretical challenges of thresholded regression.} \textbf{a,} (Asymptotic) bias versus threshold for different values of $\beta^g_1$.
+The bias function is highly nonconvex and strictly nonzero.
+Across all panels, $\beta^g_0 = 0, \beta^m_1 = 1$, and $\pi = 1/2$.
+\textbf{b,} Bias versus $\pi$ when the threshold is set to a large number.
+Bias is directly proportional to $\pi$ in this setting.
+\textbf{c,} Bias-variance decomposition of thresholding estimator in no-intercept model.
+Bias decreases and variance increases as the threshold tends to infinity.
+$\beta^m_1 = 1, \beta^g_1 = 1,$ and $\pi = 0.1$.}
 	\label{thresholding_theoretical}
@@
 
-To shed light on possible limitations of the large threshold selection strategy, we derive an exact bias-variance decomposition for the thresholding estimator.
-We consider a slightly simpler, no-intercept version of (\ref{theoretical_model}):
+Finally, to shed light on limitations of the large threshold selection strategy, we derive an exact bias-variance decomposition for the thresholding estimator.
+We consider a slightly simpler, no-intercept version of (\ref{theoretical_model}) for this purpose:
 \begin{equation}\label{theoretical_model_no_int}
@@
 \ep_i, \tau_i \sim N(0,1) \\
-p_i \indep \tau_i \indep \ep_i,
+p_i \indep \tau_i \indep \ep_i.
 \end{cases}
 \end{equation}
-The thresholding estimator $\hat{\beta}_m$ is given by
+The thresholding estimator $\hat{\beta}_m$ in the no-intercept case is
 \begin{equation}\label{thresh_estimator_no_int}
@@
 The limiting distribution of $\hat{\beta}_m$ is
-$$\sqrt{n}\left(\hat{\beta}_m - l \right) \xrightarrow{d} N\left(0, v\right).$$
+$$\sqrt{n}(\hat{\beta}_m - l) \xrightarrow{d} N\left(0, \frac{ \beta_m\omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2) }{\left(\E[\hat{p}_i]\right)^2} \right),$$ where
+$$\begin{cases}
+l = \beta_m\omega \pi/[\zeta(1-\pi) + \omega \pi], \\
+\E[\hat{p}_i] = \pi \omega + (1-\pi) \zeta, \\
+\omega = \Phi(\beta_g - c), \\
+\zeta = \Phi(- c).\\
+\end{cases}
+$$
 \end{proposition}
-The bias decreases and the variance increases as the threshold tends to infinity, consistent with the intuition that a large threshold yields a low misclassification rate at the cost of a small ``effective sample size.'' The best strategy for maximizing estimation accuracy (as quantified by mean squared error) is to select a threshold that induces moderate bias.
+This result yields an exact bias-variance decomposition for $\hat{\beta}_m$ for large $n$ (Figure \ref{thresholding_theoretical}c).
+As the threshold tends to infinity, the bias decreases and the variance increases, consistent with the intuition that a large threshold reduces the misclassification rate at the cost of decreasing the ``effective sample size.'' The best strategy for maximizing estimation accuracy (as quantified by mean squared error) is to select a threshold that induces moderate bias.
 A downside of this approach, however, is that constructing valid confidence intervals becomes more challenging.
@@
 $$ \gamma(\beta^g_1, c, \pi, \beta^g_0) = \frac{\pi(\omega - \E[\hat{p}_i])}{ \E[\hat{p}_i](1 -\E[\hat{p}_i])},$$ where $$\begin{cases}
-\omega = 1 - \Phi\left(c - \beta_1^g - \beta_0^g \right) ,\\ \zeta = 1 - \Phi\left( c - \beta^g_0 \right), \\ \E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi. 
+\omega = \Phi\left(c - \beta_1^g - \beta_0^g \right) ,\\ \zeta = \Phi\left( c - \beta^g_0 \right), \\ \E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi. 
 \end{cases}$$ Additionally, recall that the bias function $b: \R^5 \to \R$ is defined by
-$$ b(\beta^g_1, c, \pi, \beta^g_0, \beta^m_1) = \beta^m_1 (1 - \gamma(\beta^g_1, c, \pi, \beta^g_0)).$$ The parameters $\beta^g_0$ and $\beta^m_1$ will be given, fixed constants throughout the proofs.
+$$ b(\beta^g_1, c, \pi, \beta^g_0, \beta^m_1) = \beta^m_1 (1 - \gamma(\beta^g_1, c, \pi, \beta^g_0)).$$ The parameters $\beta^g_0$ and $\beta^m_1$ are given, fixed constants throughout the proofs.
 Therefore, to minimize notation, we typically use $\gamma(\beta^g_1, c, \pi)$ (resp., $b(\beta^g_1, c, \pi)$) to refer to the function $\gamma$ (resp., $b$) evaluated at $(\beta^g_1, c, \pi, \beta^g_0)$ (resp., $(\beta^g_1, c, \pi, \beta^g_0, \beta^m_1)$).
@@
 \subsubsection{Re-expressing $\gamma$ in a simpler form}
-
 We rewrite the attenuation fraction $\gamma$ in a way that makes it more amenable to theoretical analysis. We leverage the fact that $f$ integrates to unity and is even.
@@
 
+\textcolor{red}{DOES STRICT ATTENUATION BIAS HOLD UNIFORMLY OVER THE PARAMETER SPACE?}
+
 \subsubsection{Bias-variance decomposition in no-intercept model}

### 2021-10-23 text/outline/outline.tex
update thresholding theory plot

@@
 \begin{abstract}
-CRISPR genome editing single-cell sequencing have transformed biological discovery.
+CRISPR genome engineering and single-cell sequencing have transformed biological discovery.
 Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases.
@@
 \end{equation}
-For a given threshold $c \in \R$, the imputed perturbation assignment $\hat{p}_i$ is $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ for $\beta^m_1$ is given by $$\hat{\beta}^m_1 = \frac{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}}) (m_i - \overline{m})}{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})}.$$
+For a given threshold $c \in \R$, the imputed perturbation assignment $\hat{p}_i$ is $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ for $\beta^m_1$ is given by $$\hat{\beta}^m_1 = \frac{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}}) (m_i - \overline{m})}{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})^2 }.$$
 
@@
 \end{proposition}
-The thresholding method suffers strict attenuation bias for all choices of the threshold and over all possible model parameters (Figure \ref{thresholding_theoretical}a).
+The thresholding method incurs strict attenuation bias (i.e., the true effect size is \textit{under}estimated) for all choices of the threshold and over all possible model parameters (Figure \ref{thresholding_theoretical}a).
 Attenuation bias is a common attribute of estimators that ignore measurement in errors-in-variables models \cite{Stefanski2000a}.
@@
 \end{proposition}
-We always can set the threshold to a large number and attain a bias of $\beta_1^m/2$ (Figure \ref{thresholding_theoretical}a, all panels).
+In other words, we always can set the threshold to a large number and attain a bias of $\beta_1^m/2$ (Figure \ref{thresholding_theoretical}a, all panels).
 This result establishes an upper bound on the bias of thresholded regression (under optimal threshold selection strategy).
@@
  \end{proposition}
-In other words, setting the threshold to a large number yields a smaller bias when $\beta^g_1$ is small (i.e., $\beta^g_1 < 2\Phi^{-1}(3/4) \approx 1.35$; Figure \ref{thresholding_theoretical}a, left); setting the threshold to the Bayes-optimal decision boundary yields a smaller bias when $\beta^g_1$ is big (i.e., $\beta^g_1 > 2\Phi^{-1}(3/4)$; Figure \ref{thresholding_theoretical}a, right); and the two approaches coincide when $\beta^g_1$ is intermediate (i.e., $\beta^g_1 = 2\Phi^{-1}(3/4)$; Figure \ref{thresholding_theoretical}a, middle).
+That is, setting the threshold to a large number yields a smaller bias when $\beta^g_1$ is small (i.e., $\beta^g_1 < 2\Phi^{-1}(3/4) \approx 1.35$; Figure \ref{thresholding_theoretical}a, left); setting the threshold to the Bayes-optimal decision boundary yields a smaller bias when $\beta^g_1$ is big (i.e., $\beta^g_1 > 2\Phi^{-1}(3/4)$; Figure \ref{thresholding_theoretical}a, right); and the two approaches coincide when $\beta^g_1$ is intermediate (i.e., $\beta^g_1 = 2\Phi^{-1}(3/4)$; Figure \ref{thresholding_theoretical}a, middle).
 
@@
 \end{proposition}
-That is, if the perturbation probability is $\pi$, and if we set the threshold to a large number, then the asymptotic bias is $\beta^m_1 \pi$ (Figure \ref{thresholding_theoretical}b; $\beta^m_1$ fixed at $1$).
+Thus, if the perturbation probability is $\pi$, and if we set the threshold to a large number, then the asymptotic bias is $\beta^m_1 \pi$ (Figure \ref{thresholding_theoretical}b; $\beta^m_1$ fixed at $1$ in plot).
 We can understand this result intuitively by considering an extreme example: when $\pi$ is very small (e.g., $\pi = 0.01$), almost all cells are unperturbed.
@@
 Bias is directly proportional to $\pi$ in this setting.
-\textbf{c,} Bias-variance decomposition of thresholding estimator in no-intercept model.
+$\beta^m_1 = 1$.
+\textbf{c,} Bias-variance decomposition in no-intercept model.
 Bias decreases and variance increases as the threshold tends to infinity.

### 2021-10-23 R_scripts/plotting/thresholding_theoretical.R

@@ -1,5 +1,8 @@
 load_all("~/research_code/glmeiv/")
-my_cols <- c("firebrick3", "dodgerblue3")
+my_cols <- c("firebrick3", "dodgerblue4", "orchid4")
+my_theme <- theme(panel.border = element_blank(), panel.grid.major = element_blank(),
+                               panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))
+my_lwd <- 0.65
 library(magrittr)
 library(ggplot2)
 library(cowplot)
@@ -14,40 +17,40 @@ beta_gs <- c(1, qnorm(0.75) * 2, 1.5)
 to_plot_a <- lapply(X = beta_gs, function(curr_beta_g) {
   y <- get_tresholding_estimator_bias(m_perturbation = 1,
                                       g_perturbation = curr_beta_g,
+                                      g_intercept = 0,
                                       pi = 0.5,
                                       c = x_grid,
-                                      return_bias = FALSE)
-  data.frame(Estimate = y, Threshold = x_grid, beta_g = curr_beta_g)
+                                      return_bias = TRUE)
+  data.frame(Bias = y, Threshold = x_grid, beta_g = curr_beta_g)
 }) %>% do.call(what = "rbind", args = .) %>%
-  dplyr::mutate(beta_g_disp = paste0("beta[g] == ", round(beta_g, 2)) %>% factor())
+  dplyr::mutate(beta_g_disp = paste0("beta[1]^g == ", round(beta_g, 2)) %>% factor())
 
 # subplot a) bias in pi = 1/2 model
-p_a <- ggplot(data = to_plot_a, mapping = aes(x = Threshold, y = Estimate)) +
-  geom_hline(yintercept = 0.5, col = "darkred") +
+p_a <- ggplot(data = to_plot_a, mapping = aes(x = Threshold, y = Bias)) +
+  geom_hline(yintercept = 0.5, col = "darkred", lwd = my_lwd) +
   facet_wrap(.~beta_g_disp, labeller = label_parsed, scales = "free_y") +
-  theme_cowplot(font_size = 11) + scale_x_continuous(expand = c(NA, 0)) +
-  geom_vline(data = dplyr::filter(to_plot_a, beta_g_disp == "beta[g] == 1"),
-             mapping = aes(xintercept = beta_gs[1]/2), col = my_cols[2]) +
[…]

### 2021-10-25 text/outline/outline.tex
finish draft of thresholding theory section (+ appendix)

@@
 \usepackage{graphicx}
+\setcounter{tocdepth}{2}
 \usepackage{xcolor}
@@
 \end{abstract}
+
+\tableofcontents
 
@@
 
-\subsection{Theoretical analysis}
+\subsection{Theoretical analysis}\label{sec:thresholding_theory}
 
@@
 \end{equation}
-For a given threshold $c \in \R$, the imputed perturbation assignment $\hat{p}_i$ is $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ for $\beta^m_1$ is given by $$\hat{\beta}^m_1 = \frac{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}}) (m_i - \overline{m})}{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})^2 }.$$
+For a given threshold $c \in \R$, the imputed perturbation assignment $\hat{p}_i$ is given by $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ for $\beta^m_1$ is $$\hat{\beta}^m_1 = \frac{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}}) (m_i - \overline{m})}{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})^2 }.$$
 
-\begin{proposition}
+\begin{proposition}\label{prop:convergence}
  The almost sure limit (as $n \to \infty$) of $\hat{\beta}^m_1$ is
@@
 \end{itemize}
-Let $b: \R^5 \to \R$ be the asymptotic bias of $\hat{\beta}^m_1$, i.e.
-$$
-b(\beta^g_1, \pi, c, \beta^g_0, \beta^m_1) = \lim_{n\to\infty} \left(\beta^m_1 - \E[\hat{\beta}^m_1]\right) = \beta^m_1 - \E \left( \lim_{a.s.} \hat{\beta}^m_1 \right) = \beta^m_1 - \lim_{a.s.} \hat{\beta}^m_1,
-$$
-where the $\lim_{a.s.}$ denotes a.s.\ convergence.
-\textcolor{red}{(HOW TO MAKE THE ABOVE RIGOROUS?)} We can express $b$ in terms of $\gamma$ as $$b(\beta^g_1, \pi, c, \beta^g_0, \beta^m_1) = \beta^m_1 - \gamma(\beta^g_1, \pi, c, \beta^g_0)  \beta^m_1 = \beta^m_1 \left[1 - \gamma(\beta^g_1, \pi, c, \beta^g_0)  \right].$$ The asymptotic bias vanishes when the attenuation function equals $1$.
+Let $b: \R^4 \to \R$ be the asymptotic \textit{relative} bias of $\hat{\beta}^m_1$:
+\begin{multline*}
+b(\beta^g_1, \pi, c, \beta^g_0) = \left(\frac{1}{\beta^m_1}\right) \lim_{n\to\infty} \left(\beta^m_1 - \E[\hat{\beta}^m_1]\right) = \left( \frac{1}{\beta^m_1}\right)\left(\beta^m_1 - \E \left(\lim_{a.s.} \hat{\beta}^m_1 \right)\right) \\ = \frac{1}{\beta^m_1} \left(\beta^m_1 - \gamma(\beta^g_1, \pi, c, \beta^g_0)\beta^m_1\right) = 1 - \gamma(\beta^g_1, \pi, c, \beta^g_0),
+\end{multline*}
+where $\lim_{a.s.}$ denotes a.s.\ convergence. The asymptotic relative bias vanishes when the attenuation function equals $1$. \textcolor{red}{(HOW TO MAKE THE ABOVE RIGOROUS?)}
 
@@
 
-To investigate the basic question of ``What is a good threshold selection strategy?'', we study the relationship between the asymptotic bias $b$ of $\hat{\beta}^m_1$ and the selected threshold $c$.
+To investigate the basic question of ``What is a good threshold selection strategy?'', we study the relationship between the asymptotic relative bias $b$ of $\hat{\beta}^m_1$ and the selected threshold $c$.
 For simplicity, we set the perturbation probability $\pi$ to $1/2$.
 Let $c_\textrm{bayes} \in \R$ be the Bayes-optimal decision boundary for classifying cells as perturbed or unperturbed, i.e.
-$$c_\textrm{bayes} = \argmin_{c \in \R} \P(\hat{p}_i \neq p_i).$$ Simple algebra shows that $c_\textrm{bayes} = \beta_0^g + (1/2) \beta^g_1.$ Below, we state several results on the limiting bias $b$ of $\hat{\beta}^m_1$.
-We refer throughout to Figure \ref{thresholding_theoretical}a, which displays plots of asymptotic bias versus threshold for different values of $\beta^g_1$; the target of inference, $\beta^m_1$, is fixed to $1$ in these plots.
+$$c_\textrm{bayes} = \argmin_{c \in \R} \P(\hat{p}_i \neq p_i).$$ Simple algebra shows that $c_\textrm{bayes} = \beta_0^g + (1/2) \beta^g_1.$ Below, we give several results for the asymptotic relative bias $b$ of $\hat{\beta}^m_1$.
+We refer throughout to Figure \ref{thresholding_theoretical}a, which displays plots of asymptotic relative bias versus threshold for different values of $\beta^g_1$.
+We sometimes refer to ``asymptotic relative bias'' using the shortened term ``bias'' for succinctness.
 
@@
 \item 
-\begin{proposition}\label{prop:att_bias} Assume without loss of generality that $\beta^m_1 > 0.$ Fix $\pi = 1/2$. For all $(\beta^g_1, c, \beta^g_0) \in \R^3$,
-	$$b(\beta^g_1, 1/2, c, \beta^g_0, \beta^m_1) > 0.$$
+\begin{proposition}\label{prop:att_bias} Fix $\pi = 1/2$. For all $(\beta^g_1, c, \beta^g_0) \in \R^3$, the asymptotic relative bias is positive, i.e. 
+$$b(\beta^g_1, 1/2, c, \beta^g_0) > 0.$$
 \end{proposition}
-The thresholding method incurs strict attenuation bias (i.e., the true effect size is \textit{under}estimated) for all choices of the threshold and over all possible model parameters (Figure \ref{thresholding_theoretical}a).
+The thresholding method incurs strict attenuation bias (i.e., it \textit{under}estimates the true effect size) for all choices of the threshold and over all possible values of the model parameters (Figure \ref{thresholding_theoretical}a).
 Attenuation bias is a common attribute of estimators that ignore measurement in errors-in-variables models \cite{Stefanski2000a}.
-
-\item \begin{proposition}
-For $\pi = 1/2$ and given $(\beta^g_1, \beta^g_0, \beta^m_1) \in \R^3$, the Bayes-optimal decision boundary $c_\textrm{bayes}$ is a critical value of the limiting bias function $b$, i.e.
-$$ \frac{\partial b}{\partial c}\left(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0, \beta^m_1\right) = 0.$$
+\item
+\begin{proposition}\label{prop:monotonic} Fix $\pi = 1/2$. The asymptotic relative bias $b$ decreases monotonically in $\beta_1^g$, i.e.
+	$$\frac{\partial b}{\partial(\beta^g_1)}\left(\beta^g_1, 1/2, c, \beta^g_0\right) \leq 0.$$
 \end{proposition}
-The Bayes-optimal decision boundary is an optimum (or possibly a saddle point) of the limiting bias function (Figure \ref{thresholding_theoretical}a, vertical blue lines).
+This result formalizes the intuition that the problem becomes easier as the gRNA mixture distribution becomes increasingly well-separated.
+To visualize Proposition (\ref{prop:monotonic}), one can fix a threshold (e.g., $c = 0$) and scan for bias across the panels.
+\item \begin{proposition}\label{prop:bayes_opt}
+For $\pi = 1/2$ and given $(\beta^g_1, \beta^g_0) \in \R^2$, the Bayes-optimal decision boundary $c_\textrm{bayes}$ is a critical value of the bias function $b$, i.e.
+$$ \frac{\partial b}{\partial c}\left(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0\right) = 0.$$
+\end{proposition}
+The Bayes-optimal decision boundary is an optimum (or possibly a saddle point) of the asymptotic relative bias function (Figure \ref{thresholding_theoretical}a, vertical blue lines).
 Interestingly, $c_\textrm{bayes}$ is in some cases a maximizer of the bias (Figure \ref{thresholding_theoretical}a, left) and in other cases a minimizer of the bias (Figure \ref{thresholding_theoretical}a, right).
 \item
-\begin{proposition}\label{prop:limit}
-Fix $\pi = 1/2$. As the threshold $c$ tends to infinity, the asymptotic bias $b$ tends to $\beta_1^m/2$, i.e.
-$$\lim_{c \to \infty} b(\beta^g_1, 1/2, c, \beta^g_0, \beta^m_1) = \beta_1^m/2.$$
+\begin{proposition}\label{prop:c_limit_half}
+Assume without loss of generality that $\beta^g_1 > 0$, and fix $\pi = 1/2$. As the threshold $c$ tends to infinity, the asymptotic relative bias $b$ tends to $1/2$, i.e.
+$$\lim_{c \to \infty} b(\beta^g_1, 1/2, c, \beta^g_0) = 1/2.$$
 \end{proposition}
-In other words, we always can set the threshold to a large number and attain a bias of $\beta_1^m/2$ (Figure \ref{thresholding_theoretical}a, all panels).
+In other words, we always can set the threshold to a large number and attain a relative bias of $1/2$ (Figure \ref{thresholding_theoretical}a, all panels).
 This result establishes an upper bound on the bias of thresholded regression (under optimal threshold selection strategy).
@@
  \item The following proposition compares the two threshold selection strategies introduced above (i.e., large number versus Bayes-optimal decision boundary) head-to-head.
- \begin{proposition} Let $\beta^m_1 > 0$.
-For $\beta^g_1 \in [0, 2\Phi^{-1}(3/4))$, we have that $$b(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0, \beta^m_1) > b(\beta^g_1, 1/2, \infty, \beta^g_0, \beta^m_1).$$ For $\beta^g_1 = 2\Phi^{-1}(3/4)$, we have that $$ b(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0, \beta^m_1) = b(\beta^g_1, 1/2, \infty, \beta^g_0, \beta^m_1).$$ Finally, for $\beta^g_1 \in (2\Phi^{-1}(3/4), \infty)$, we have that
- $$b(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0, \beta^m_1) < b(\beta^g_1, 1/2, \infty, \beta^g_0, \beta^m_1).$$
+ \begin{proposition}\label{prop:comparison} Assume without loss of generality that $\beta^g_1 > 0$.
+For $\beta^g_1 \in [0, 2\Phi^{-1}(3/4))$, we have that $$b(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0) > b(\beta^g_1, 1/2, \infty, \beta^g_0).$$ For $\beta^g_1 = 2\Phi^{-1}(3/4)$, we have that $$ b(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0) = b(\beta^g_1, 1/2, \infty, \beta^g_0).$$ Finally, for $\beta^g_1 \in (2\Phi^{-1}(3/4), \infty)$, we have that
+ $$b(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0) < b(\beta^g_1, 1/2, \infty, \beta^g_0).$$
  \end{proposition}
-That is, setting the threshold to a large number yields a smaller bias when $\beta^g_1$ is small (i.e., $\beta^g_1 < 2\Phi^{-1}(3/4) \approx 1.35$; Figure \ref{thresholding_theoretical}a, left); setting the threshold to the Bayes-optimal decision boundary yields a smaller bias when $\beta^g_1$ is big (i.e., $\beta^g_1 > 2\Phi^{-1}(3/4)$; Figure \ref{thresholding_theoretical}a, right); and the two approaches coincide when $\beta^g_1$ is intermediate (i.e., $\beta^g_1 = 2\Phi^{-1}(3/4)$; Figure \ref{thresholding_theoretical}a, middle).
-
-\item
- \begin{proposition}\label{prop_monotonic} Fix $\pi = 1/2$. The limiting bias $b$ decreases monotonically in $\beta_1^g$, i.e.
-$$\frac{\partial b}{\partial(\beta^g_1)}\left(\beta^g_1, 1/2, c, \beta^g_0, \beta^m_1\right) \leq 0.$$
-\end{proposition}
-This result formalizes the intuition that the problem becomes easier as the gRNA mixture distribution becomes increasingly well-separated.
-To visualize Proposition (\ref{prop_monotonic}), one can fix a threshold (e.g., $c = 0$) and scan for bias across the panels.
+Setting the threshold to a large number yields a smaller bias when $\beta^g_1$ is small (i.e., $\beta^g_1 < 2\Phi^{-1}(3/4) \approx 1.35$; Figure \ref{thresholding_theoretical}a, left); setting the threshold to the Bayes-optimal decision boundary yields a smaller bias when $\beta^g_1$ is large (i.e., $\beta^g_1 > 2\Phi^{-1}(3/4)$; Figure \ref{thresholding_theoretical}a, right); and the two approaches coincide when $\beta^g_1$ is intermediate (i.e., $\beta^g_1 = 2\Phi^{-1}(3/4)$; Figure \ref{thresholding_theoretical}a, middle).
 \end{itemize}
@@
 \end{center}
-We generalize the expression for bias to arbitrary $\pi \in [0,1/2]$:
-\begin{proposition}
-As the threshold $c$ tends to infinity, the limiting bias $b$ tends to $\beta^m_1 \pi,$ i.e.
-$$ \lim_{ c \to \infty } b(\beta^g_1, \pi, c, \beta^g_0, \beta^m_1) = \beta^m_1 \pi.$$
+We generalize the expression for bias when the threshold is large to arbitrary $\pi \in [0,1/2]$:
+\begin{proposition}\label{prop:c_limit}
+Assume without loss of generality that $\beta^g_1 > 0$. As the threshold $c$ tends to infinity, the asymptotic relative bias $b$ tends to $\pi,$ i.e.
+$$ \lim_{ c \to \infty } b(\beta^g_1, \pi, c, \beta^g_0) = \pi.$$
 \end{proposition}
-Thus, if the perturbation probability is $\pi$, and if we set the threshold to a large number, then the asymptotic bias is $\beta^m_1 \pi$ (Figure \ref{thresholding_theoretical}b; $\beta^m_1$ fixed at $1$ in plot).
+In other words, if the perturbation probability is $\pi$, and if we set the threshold to a large number, then the asymptotic relative bias is $\pi$ (Figure \ref{thresholding_theoretical}b).
 We can understand this result intuitively by considering an extreme example: when $\pi$ is very small (e.g., $\pi = 0.01$), almost all cells are unperturbed.
@@
 	\includegraphics[width=1\linewidth]{../../figures/thresholding_theoretical/plot}
-	\caption{\textbf{Theoretical challenges of thresholded regression.} \textbf{a,} (Asymptotic) bias versus threshold for different values of $\beta^g_1$.
+	\caption{\textbf{Theoretical challenges of thresholded regression.} \textbf{a,} Asymptotic relative bias versus threshold for different values of $\beta^g_1$.
 The bias function is highly nonconvex and strictly nonzero.
-Across all panels, $\beta^g_0 = 0, \beta^m_1 = 1$, and $\pi = 1/2$.
-\textbf{b,} Bias versus $\pi$ when the threshold is set to a large number.
-Bias is directly proportional to $\pi$ in this setting.
-$\beta^m_1 = 1$.
-\textbf{c,} Bias-variance decomposition in no-intercept model.
+Vertical blue lines, Bayes-optimal decision boundaries.
+Across all panels, $\beta^g_0 = 0$ and $\pi = 1/2$.
+\textbf{b,} Asymptotic relative bias versus $\pi$ when the threshold is set to a large number.
+The two quantities coincide exactly.
+\textbf{c,} Bias-variance decomposition for thresholding method in no-intercept model.
 Bias decreases and variance increases as the threshold tends to infinity.
-$\beta^m_1 = 1, \beta^g_1 = 1,$ and $\pi = 0.1$.}
+$\beta^g_1 = 1, \beta^m_1 = 1,$ and $\pi = 0.1$.}
 	\label{thresholding_theoretical}
@@
 \end{equation}
-\begin{proposition}
+\begin{proposition}\label{prop:bv_decomp}
 The limiting distribution of $\hat{\beta}_m$ is
@@
 \subsection{Theoretical details for thresholding estimator}
+This section contains proofs of the propositions presented Section \ref{sec:thresholding_theory}, ``Theoretical analysis of thresholding estimator.'' The subsections are organized as follows.
+Section (\ref{sec:notation}) introduces some notation.
+Section (\ref{sec:convergence}) establishes almost sure convergence of the thresholding estimator in the model (\ref{theoretical_model}), proving Proposition \ref{prop:convergence}.
+Section (\ref{sec:simplication}) simplifies the expression for the attenuation function $\gamma$, and section (\ref{sec:derivatives})  computes derivatives of $\gamma$ to be used throughout the proofs.
+Section (\ref{sec:c_limit}) establishes the limit in $c$ of $\gamma$, proving Proposition \ref{prop:c_limit} and as a corollary Proposition \ref{prop:c_limit_half}.
+Section (\ref{sec:bayes_opt}) establishes that the Bayes-optimal decision boundary is a critical value of $\gamma$, proving Proposition \ref{prop:bayes_opt}, and section (\ref{sec:comparison}) compares the competing threshold selection strategies head-to-head, proving Proposition \ref{prop:comparison}.
+Section (\ref{sec:monotone}) demonstrates that $\gamma$ is monotone in $\beta^g_1$, proving Proposition \ref{prop:monotonic}, and Section (\ref{sec:att_bias}) establishes attenuation bias of the thresholding estimator, proving Proposition \ref{prop:att_bias}.
+Finally, Section (\ref{sec:bv_decomp}) derives the bias-variance decomposition of the thresholding estimator in the model (\ref{theoretical_model_no_int}), proving Proposition \ref{prop:bv_decomp}.
 
-This section contains proofs for results presented the section ``Theoretical analysis of thresholding estimator.'' The subsections are arranged as follows: FILL IN...
+% Labels of propositions and theorems
+% propositions:
+% prop:convergence
+% prop:att_bias
+% prop:bayes_opt
+% prop:c_limit_half
+% prop:comparison
+% prop:monotonic
+% prop:c_limit
+% prop:bv_decomp
 
-\subsubsection{Notation}
- All notation introduced here pertains to the Gaussian model with intercepts (\ref{theoretical_model}). Recall that the attenuation function $\gamma: \R^4 \to \R$ is defined by
-$$ \gamma(\beta^g_1, c, \pi, \beta^g_0) = \frac{\pi(\omega - \E[\hat{p}_i])}{ \E[\hat{p}_i](1 -\E[\hat{p}_i])},$$ where $$\begin{cases}
-\omega = \Phi\left(c - \beta_1^g - \beta_0^g \right) ,\\ \zeta = \Phi\left( c - \beta^g_0 \right), \\ \E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi. 
-\end{cases}$$ Additionally, recall that the bias function $b: \R^5 \to \R$ is defined by
-$$ b(\beta^g_1, c, \pi, \beta^g_0, \beta^m_1) = \beta^m_1 (1 - \gamma(\beta^g_1, c, \pi, \beta^g_0)).$$ The parameters $\beta^g_0$ and $\beta^m_1$ are given, fixed constants throughout the proofs.
-Therefore, to minimize notation, we typically use $\gamma(\beta^g_1, c, \pi)$ (resp., $b(\beta^g_1, c, \pi)$) to refer to the function $\gamma$ (resp., $b$) evaluated at $(\beta^g_1, c, \pi, \beta^g_0)$ (resp., $(\beta^g_1, c, \pi, \beta^g_0, \beta^m_1)$).
-Next, we define the functions $g$ and $h: \R^4 \to \R$ by
+% sections
+% sec:notation
+% sec:convergence
+% sec:simplication
+% sec:derivatives
+% sec:c_limit
+% sec:bayes_opt
+% sec:comparison
+% sec:beta_lim
+% sec:monotone
+% sec:att_bias
+% sec:bv_decomp
+
+\subsubsection{Notation}\label{sec:notation}
+ All notation introduced in this subsection (i.e., \ref{sec:notation}) pertains to the Gaussian model with intercepts (\ref{theoretical_model}).
+Recall that the attenuation function $\gamma: \R^4 \to \R$ is defined by
+$$ \gamma(\beta^g_1, c, \pi, \beta^g_0) = \frac{\pi(\omega - \E[\hat{p}_i])}{ \E[\hat{p}_i](1 -\E[\hat{p}_i])},$$ where $$\begin{cases} \E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi, \\
+\omega = \Phi\left(\beta_1^g + \beta_0^g - c \right) ,\\ \zeta = \Phi\left( \beta^g_0 - c \right).
+\end{cases}$$ Additionally, recall that the asymptotic relative bias function $b: \R^4 \to \R$ is
+$$ b(\beta^g_1, c, \pi, \beta^g_0) = 1 - \gamma(\beta^g_1, c, \pi, \beta^g_0).$$ Next, we define the functions $g$ and $h: \R^4 \to \R$ by
 \begin{equation}\label{def_g}
-g(\beta^g_1, c, \pi, \beta^g_0) = (1-\pi)\left( \int_{ -\infty}^{\beta_0^g + \beta_1^g - c} f \right) - (1-\pi)\left( \int_{-\infty}^{ \beta_0^g - c} f \right)\end{equation}
+g(\beta^g_1, c, \pi, \beta^g_0) = (1-\pi)\left( \Phi(\beta_0^g + \beta_1^g - c)\right) - (1-\pi)\left(\Phi(\beta_0^g - c)\right)\end{equation}
  and
 \begin{multline}\label{def_h}
-h(\beta^g_1, c, \pi, \beta^g_0) = \left[(1-\pi)\left( \int_{-\infty}^{ \beta_0^g - c} f \right) + \pi\left( \int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \right] \cdot \\ \left[(1-\pi)\left( \int_{-\infty}^{c - \beta^g_0} f \right) + \pi\left(\int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \right],
-\end{multline} where $f: \R \to [0,\infty)$ is the $N(0,1)$ density.
-We will refer to $g$ and $h$ repeatedly throughout the proofs.
-Similarly to above, we typically use the notation $g(\beta^g_1, c, \pi)$ (resp.\ $h(\beta^g_1, c, \pi)$) to refer to $g$ (resp.\ $h$) evaluated at $(\beta^g_1, c, \pi, \beta^g_0)$.
+h(\beta^g_1, c, \pi, \beta^g_0) = \left[(1-\pi)\left( \Phi(\beta_0^g - c)\right) + \pi\left(\Phi(\beta^g_0 + \beta^g_1 - c) \right) \right] \cdot \\ \left[(1-\pi)\left( \Phi(c - \beta^g_0) \right) + \pi\left(\Phi(c - \beta_0^g - \beta_1^g) \right) \right].
+\end{multline}
+We use $f:\R \to \R$ to denote the $N(0,1)$ density, and we denote the right-tail probability probability of $f$ by $\bar{\Phi}$, i.e., 
+$$\bar{\Phi}(x) = \int_{x}^{\infty} f = \Phi(-x).$$
 
+The parameter $\beta^g_0$ is a given, fixed constant throughout the proofs.
+Therefore, to minimize notation, we typically use $\gamma(\beta^g_1, c, \pi)$ (resp., $b(\beta^g_1, c, \pi),$ $g(\beta^g_1, c, \pi),$ $h(\beta^g_1, c, \pi)$) to refer to the function $\gamma$ (resp., $b, g, h$) evaluated at $(\beta^g_1, c, \pi, \beta^g_0)$.
 Finally, for a given function $r: \R^{p} \to \R$, point $x \in \R^p$, and index $i \in \{1, \dots, p\}$, we use the symbol $D_i r(x)$ to refer to the derivative of the $i$th component of $r$ evaluated at $x$ (\textit{sensu} \cite{fitzpatrick2009}).
@@
 
-\subsubsection{Limit of $\hat{\beta}^m_1$}
+\subsubsection{Almost sure limit of $\hat{\beta}^m_1$}\label{sec:convergence}
 
@@
 \item[1.] Expectation of $m_i$: $\E[m_i] = \beta^m_0 + \beta^m_1\pi.$
-\item[2.] Expectation of $\hat{p}_i$: Let $f$ denote the $N(0,1)$ density. Then \begin{multline*}
-\E[\hat{p}_i] = \P\left[\hat{p}_i = 1\right] = \P\left[\beta^g_0 + \beta^g_1 p_i + \tau_i \geq c \right] = \\ \textrm{(By LOTP) } \P\left[ \beta^g_0 + \tau_i \geq c \right]\P\left[p_i = 0\right] + \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \P[p_i = 1] \\ = \P\left[ \tau_i \geq c - \beta^g_0\right](1- \pi) + \P\left[ \tau_i \geq c - \beta^g_1 - \beta^g_0 \right](\pi) \\ = \left(\int_{c - \beta^g_0}^\infty f\right) (1 - \pi) + \left( \int_{c - \beta^g_1 - \beta^g_0}^\infty f \right)(\pi) = \zeta(1-\pi) + \omega \pi.
+\item[2.] Expectation of $\hat{p}_i$: \begin{multline*}
+\E[\hat{p}_i] = \P\left[\hat{p}_i = 1\right] = \P\left[\beta^g_0 + \beta^g_1 p_i + \tau_i \geq c \right] = \\ \textrm{(By LOTP) } \P\left[ \beta^g_0 + \tau_i \geq c \right]\P\left[p_i = 0\right] + \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \P[p_i = 1] \\ = \P\left[ \tau_i \geq c - \beta^g_0\right](1- \pi) + \P\left[ \tau_i \geq c - \beta^g_1 - \beta^g_0 \right](\pi) \\ =  \left(\bar{\Phi}(c - \beta^g_0) \right) (1 - \pi) + \left( \bar{\Phi}(c - \beta^g_1 - \beta^g_0) \right)(\pi) = \\  \Phi(\beta^g_0 - c) (1-\pi) + \Phi(\beta^g_1 + \beta^g_0 - c) \pi = \zeta(1-\pi) + \omega \pi.
 \end{multline*}
@@
 
-\subsubsection{Re-expressing $\gamma$ in a simpler form}
-We rewrite the attenuation fraction $\gamma$ in a way that makes it more amenable to theoretical analysis. We leverage the fact that $f$ integrates to unity and is even.
-First,
-\begin{multline}\label{thm:gamma_expression_1} \E\left[\hat{p}_i\right] = (1 - \pi) \left( \int_{c - \beta^g_0}^\infty f \right) + \pi \left(\int_{c - \beta_0^g - \beta_1^g}^\infty f \right) \\ = (1 - \pi) \left( \int_{-\infty}^{ \beta_0^g - c} f \right) + \pi \left( \int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right), \end{multline}
- and so \begin{multline}\label{thm:gamma_expression_2} 1 - \E\left[\hat{p}_i\right] = (1 - \pi) + \pi - \E[\hat{p}_i]  = (1-\pi) \left(1 - \int_{c - \beta_0^g}^\infty f \right)  + \pi \left(1 - \int_{c - \beta^g_0 - \beta^g_1}^\infty f \right) \\ = (1 - \pi) \left( \int_{-\infty}^{c - \beta^g_0} f \right) + \pi \left( \int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right).
+\subsubsection{Re-expressing $\gamma$ in a simpler form}\label{sec:simplication}
+We rewrite the attenuation fraction $\gamma$ in a way that makes it more amenable to theoretical analysis. We leverage the fact that $f$ integrates to unity and is even. We have that
+\begin{multline}\label{thm:gamma_expression_1} \E\left[\hat{p}_i\right] = (1 - \pi) \bar{\Phi}(c - \beta_0^g) + \pi \bar{\Phi}(c - \beta^g_0 - \beta^g_1) \\ = (1 - \pi) \Phi(\beta_0^g - c) + \pi\Phi(\beta^g_0 + \beta^g_1 - c), \end{multline}
+ and so \begin{multline}\label{thm:gamma_expression_2} 1 - \E\left[\hat{p}_i\right] = (1 - \pi) + \pi - \E[\hat{p}_i]  = (1-\pi) \left(1 - \bar{\Phi}(c - \beta_0^g)\right)  + \pi \left(1 - \bar{\Phi}(c - \beta^g_0 - \beta^g_1) \right) \\ = (1 - \pi)\Phi(c - \beta^g_0) + \pi \Phi(c - \beta_0^g - \beta_1^g).
 \end{multline}
@@
 \begin{equation}\label{thm:gamma_expression_3}
-\omega = \int_{c - \beta^g_1 - \beta^g_0}^\infty f,\end{equation} and so
+\omega = \Phi(\beta^g_1 + \beta^g_0 - c),\end{equation} and so
 \begin{multline}\label{thm:gamma_expression_4}
-\omega - \E[\hat{p}_i]  = \left( \int_{c - \beta^g_1 - \beta^g_0}^\infty f \right) - (1 - \pi) \left( \int_{c - \beta^g_0}^\infty f \right) - \pi \left( \int_{c - \beta_0^g - \beta_1^g}^\infty f \right) \\ = (1 - \pi) \left(\int_{c - \beta^g_0 - \beta^g_1 }^\infty f \right) - (1- \pi) \left(\int_{c - \beta^g_0}^\infty f \right) \\ =  (1 - \pi)\left(\int_{-\infty}^{\beta_0^g + \beta_1^g - c} f \right) - (1 - \pi) \left( \int_{-\infty}^{ \beta_0^g - c} f \right).
+\omega - \E[\hat{p}_i] = \Phi(\beta^g_1 + \beta^g_0 - c) - (1-\pi)\Phi(\beta^g_0 - c) - \pi \Phi(\beta^g_0 + \beta^g_1 - c)  \\ (1-\pi)\Phi(\beta^g_1 + \beta^g_0 - c)  - (1-\pi)\Phi(\beta^g_0 - c).
 \end{multline}
+
 Combining (\ref{thm:gamma_expression_1}, \ref{thm:gamma_expression_2}, \ref{thm:gamma_expression_3}, \ref{thm:gamma_expression_4}), we find that
 \begin{multline}\label{gamma_alternative}
-\gamma(\beta^g_1, c, \pi) = \frac{\pi(\omega - \E[\hat{p}_i])}{\E[\hat{p}_i](1 - \E[\hat{p}_i])} \\ = \frac{\pi \left[ (1 - \pi) \left( \int_{ -\infty}^{\beta_0^g + \beta_1^g - c} f \right) - (1 - \pi) \left( \int_{-\infty}^{ \beta_0^g - c} f \right) \right]}{\left[(1-\pi) \left( \int_{-\infty}^{ \beta_0^g - c} f \right) + \pi \left( \int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \right] \left[(1 - \pi) \left( \int_{-\infty}^{c - \beta^g_0} f \right) + \pi \left( \int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \right]}.
+\gamma(\beta^g_1, c, \pi) = \frac{\pi(\omega - \E[\hat{p}_i])}{\E[\hat{p}_i](1 - \E[\hat{p}_i])} \\ = \frac{\pi \left[(1 - \pi) \Phi(\beta_0^g + \beta_1^g - c) - (1 - \pi) \Phi(\beta_0^g - c)\right]}{\left[(1-\pi)\Phi(\beta_0^g - c) + \pi \Phi(\beta^g_0 + \beta^g_1 - c) \right] \left[(1 - \pi) \Phi(c - \beta^g_0) + \pi\Phi(c - \beta_0^g - \beta_1^g) \right]}.
 \end{multline}
@@
 \begin{multline}\label{gamma_alternative_pi_half}
-\gamma(\beta^g_1, c, 1/2)  \\ = \frac{\left( \int_{ -\infty}^{\beta_0^g + \beta_1^g - c} f \right) - \left( \int_{-\infty}^{ \beta_0^g - c} f \right)}{\left[ \left( \int_{-\infty}^{ \beta_0^g - c} f \right) + \left( \int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \right] \left[\left( \int_{-\infty}^{c - \beta^g_0} f \right) + \left(\int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \right]}.
+\gamma(\beta^g_1, c, 1/2)  \\ = \frac{\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c) }{\left[\Phi(\beta_0^g - c) +\Phi(\beta^g_0 + \beta^g_1 - c)\right] \left[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g) \right]}.
 \end{multline}
@@
 i.e., the numerator and denominator of  (\ref{gamma_alt2_pi_half}) coincide with those of (\ref{gamma_alternative_pi_half}).
+We sometimes will use the notation $2\cdot g$ and $4\cdot h$ to refer to the numerator and denominator of (\ref{gamma_alternative_pi_half}), respectively.
 
-\subsubsection{Derivatives of  $g$ and $h$ in $c$}
+\subsubsection{Derivatives of  $g$ and $h$ in $c$}\label{sec:derivatives}
 We compute the derivatives of $g$ and $h$ in $c$, which we will need to prove subsequent results. First, by FTC and the evenness of $f$, we have that
@@
 \begin{multline}\label{dh_dc}
-D_2 h(\beta^g_1, c, \pi) \\ = -[(1-\pi)f(\beta^g_0 - c) + \pi f( \beta^g_0 + \beta^g_1 - c )]\left[(1-\pi)\left(\int_{-\infty}^{c - \beta^g_0} f \right) + \pi \left(\int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \right] \\ + [(1-\pi) f(c - \beta^g_0) +  \pi f(c - \beta^g_0 - \beta^g_1)] \left[(1-\pi)\left( \int_{-\infty}^{\beta_0^g - c} f \right) + \pi \left(\int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \right] \\ = \left[ (1-\pi) f(c - \beta^g_0) +  \pi f(c - \beta^g_0 - \beta^g_1) \right] \cdot \\ \bigg[ (1-\pi)\left( \int_{-\infty}^{\beta_0^g - c} f \right) + \pi \left(\int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \\ - (1-\pi)\left(\int_{-\infty}^{c - \beta^g_0} f \right) - \pi \left(\int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \bigg].
+D_2 h(\beta^g_1, c, \pi) \\ = -[(1-\pi)f(\beta^g_0 - c) + \pi f( \beta^g_0 + \beta^g_1 - c )]\left[(1-\pi)\Phi(c - \beta^g_0) + \pi \Phi(c - \beta_0^g - \beta_1^g)  \right] \\ + [(1-\pi) f(c - \beta^g_0) +  \pi f(c - \beta^g_0 - \beta^g_1)] \left[(1-\pi) \Phi(\beta_0^g - c) + \pi \Phi(\beta^g_0 + \beta^g_1 - c) \right] \\ = \left[ (1-\pi) f(c - \beta^g_0) +  \pi f(c - \beta^g_0 - \beta^g_1) \right] \cdot \\ \bigg[ (1-\pi) \Phi(\beta_0^g - c) + \pi\Phi(\beta^g_0 + \beta^g_1 - c) \\ - (1-\pi) \Phi(c - \beta^g_0) - \pi \Phi(c - \beta_0^g - \beta_1^g) \bigg].
 \end{multline}
 
-\subsubsection{Limit of $\gamma$ in $c$}
+\subsubsection{Limit of $\gamma$ in $c$}\label{sec:c_limit}
 
@@
 We have by (\ref{dg_dc}) and (\ref{dh_dc}) that \begin{multline}\label{c_limit_product}
-\lim_{c \to \infty} \gamma(\beta^g_1, c, \pi) = \lim_{c \to \infty} \frac{\pi D_2 g(\beta^g_1, c, \pi)}{D_2h(\beta^g_1, c, \pi)} \\ = \lim_{c \to \infty} \bigg\{ \frac{(1-\pi) f(c - \beta^g_0) + \pi f(c - \beta^g_0 - \beta^g_1)}{\pi (1-\pi) f(c - \beta^g_0) - \pi (1-\pi)f(c - \beta^g_0 - \beta^g_1)} \\ \cdot \bigg[ (1-\pi)\left( \int_{-\infty}^{\beta_0^g - c} f \right) + \pi \left(\int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \\ - (1-\pi)\left(\int_{-\infty}^{c - \beta^g_0} f \right) - \pi \left(\int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \bigg] \bigg\}^{-1}.
+\lim_{c \to \infty} \gamma(\beta^g_1, c, \pi) = \lim_{c \to \infty} \frac{\pi D_2 g(\beta^g_1, c, \pi)}{D_2h(\beta^g_1, c, \pi)} \\ = \lim_{c \to \infty} \bigg\{ \frac{(1-\pi) f(c - \beta^g_0) + \pi f(c - \beta^g_0 - \beta^g_1)}{\pi (1-\pi) f(c - \beta^g_0) - \pi (1-\pi)f(c - \beta^g_0 - \beta^g_1)} \\ \cdot \bigg[ (1-\pi) \Phi(\beta_0^g - c) + \pi \Phi(\beta^g_0 + \beta^g_1 - c) \\ - (1-\pi) \Phi(c - \beta^g_0) - \pi \Phi(c - \beta_0^g - \beta_1^g) \bigg] \bigg\}^{-1}.
  \end{multline}
- We evaluate the two terms in the product (\ref{c_limit_product}) separately. First, observe that
+ We evaluate the two terms in the product (\ref{c_limit_product}) separately. Dividing by $f(c - \beta^g_0 - \beta^g_1) > 0$, we see that
  \begin{equation}\label{c_limit_product_2}
@@
  \end{equation}
- To evaluate the limit of (\ref{c_limit_product_2}), must evaluate the limit of
+ To evaluate the limit of (\ref{c_limit_product_2}), we first evaluate the limit of
  \begin{multline}\label{c_limit_product_3}
@@
 \end{multline}
-Computing the limit of (\ref{c_limit_product_3}), we obtain
+Taking the limit in (\ref{c_limit_product_3}), we obtain
 $$
 \lim_{c \to \infty} \frac{f(c - \beta^g_0)}{f(c - \beta^g_0 - \beta^g_1)} = \exp[ \beta^g_0 \beta^g_1 + (\beta^g_1)^2/2] \lim_{c \to \infty} \exp[ -c \beta^g_1] = 0
-$$ for $\beta^g_1 > 0$. We now can evaluate the limit of (\ref{c_limit_product_2}) (for $\beta^g_1 > 0$):
+$$ for $\beta^g_1 > 0$. We now can evaluate the limit of (\ref{c_limit_product_2}):
 $$ \lim_{c \to \infty} \frac{(1-\pi) f(c - \beta^g_0) + \pi f(c - \beta^g_0 - \beta^g_1)}{\pi (1-\pi) f(c - \beta^g_0) - \pi (1-\pi)f(c - \beta^g_0 - \beta^g_1)} = \frac{-\pi}{\pi(1-\pi)} = -\frac{1}{1 -\pi}.$$ Next, we compute the limit of the other term in the product (\ref{c_limit_product}):
 \begin{multline}\label{c_limit_product_4}
-\lim_{c \to \infty} \bigg[ (1-\pi)\left( \int_{-\infty}^{\beta_0^g - c} f \right) + \pi \left(\int_{-\infty}^{ \beta^g_0 + \beta^g_1 - c} f \right) \\ - (1-\pi)\left(\int_{-\infty}^{c - \beta^g_0} f \right) - \pi \left(\int_{-\infty}^{c - \beta_0^g - \beta_1^g} f \right) \bigg] = -(1-\pi) - \pi = -1.
+\lim_{c \to \infty} \bigg[ (1-\pi)\Phi(\beta_0^g - c) + \pi \Phi(\beta^g_0 + \beta^g_1 - c) \\ - (1-\pi)\Phi(c - \beta^g_0) - \pi \Phi(c - \beta_0^g - \beta_1^g) \bigg] = -(1-\pi) - \pi = -1.
 \end{multline}
 Combining (\ref{c_limit_product_2}) and (\ref{c_limit_product_4}), the limit (\ref{c_limit_product}) evaluates to
-$$ \lim_{c \to \infty} \gamma(\beta^g_1, c, \pi) = \left(  \frac{ 1 }{ 1 - \pi }\right)^{-1} = 1 - \pi.$$ It follows that the limit in $c$ of the bias $b$ is
-$$\lim_{c \to \infty} b(\beta^g_1, c, \pi) = \beta_1^m(1 - \lim_{c \to \infty} \gamma(\beta^g_1, c, \pi)) = \beta^m_1 \pi.$$
+$$ \lim_{c \to \infty} \gamma(\beta^g_1, c, \pi) = \left(  \frac{ 1 }{ 1 - \pi }\right)^{-1} = 1 - \pi.$$ It follows that the limit in $c$ of the asymptotic relative bias $b$ is
+$$\lim_{c \to \infty} b(\beta^g_1, c, \pi) = 1 - \lim_{c \to \infty} \gamma(\beta^g_1, c, \pi) = \pi.$$
 A corollary is that
-$$\lim_{c \to \infty} b(\beta^g_1, c, 1/2) = (1/2)\beta^m_1.$$
+$$\lim_{c \to \infty} b(\beta^g_1, c, 1/2) = 1/2.$$
 
-\subsubsection{Bayes-optimal decision boundary as a critical value of $\gamma$}
-Let $c_\textrm{bayes} = \beta^g_0 + (1/2)\beta^g_1.$ We show that $c = c_\textrm{bayes}$ is a critical value of $\gamma$ for $\pi = 1/2$ and given $\beta^g_1$, i.e, $$D_2 \gamma (\beta^g_1, c_\textrm{bayes}, 1/2) = 0.$$ By the quotient rule,
-\begin{multline}\label{quotient_rule}
-D_2\gamma(\beta^g_1, c, 1/2) \\ = \frac{(1/2) D_2g(\beta^g_1, c, 1/2) h(\beta^g_1, c, 1/2) - (1/2) g(\beta^g_1, c, 1/2) D_2 h(\beta^g_1, c, 1/2)}{[h(\beta^g_1, c, \pi)]^2} \\ = \frac{\left[4(1/2) D_2g(\beta^g_1, c, 1/2)\right] h(\beta^g_1, c, 1/2) - (1/2) g(\beta^g_1, c, 1/2) \left[4 D_2 h(\beta^g_1, c, 1/2)\right]}{[2h(\beta^g_1, c, \pi)]^2},
-\end{multline}
-where the second equality multiplies both numerator and denominator by $4$. We have by (\ref{dg_dc}) that
-\begin{multline}\label{dg_dc_bayes}
-4 \cdot (1/2) \cdot D_2 g(\beta^g_1, c_\textrm{bayes}, 1/2)  = f( \beta^g_1/2) - f( -\beta^g_1/2) \\ = f(\beta^g_1/2) - f(\beta^g_1/2) = 0.
-\end{multline}
-Next, we have by (\ref{dh_dc}) that
+\subsubsection{Bayes-optimal decision boundary as a critical value of $\gamma$}\label{sec:bayes_opt}
+Let $c_\textrm{bayes} = \beta^g_0 + (1/2)\beta^g_1.$ We show that $c = c_\textrm{bayes}$ is a critical value of $\gamma$ for $\pi = 1/2$ and given $\beta^g_1$, i.e, $$D_2 \gamma (\beta^g_1, c_\textrm{bayes}, 1/2) = 0.$$ Differentiating (\ref{gamma_alt2_pi_half}), the quotient rule implies that
+\begin{equation}\label{quotient_rule}
+D_2\gamma(\beta^g_1, c, 1/2) \\ = \frac{D_2[2g(\beta^g_1, c, 1/2)] 4h(\beta^g_1, c, 1/2) - 2g(\beta^g_1, c, 1/2) D_2[4h(\beta^g_1, c, 1/2)]}{[4h(\beta^g_1, c, \pi)]^2}.
+\end{equation}
+We have by (\ref{dg_dc}) that
+\begin{equation}\label{dg_dc_bayes}
+D_2[2g(\beta^g_1, c_\textrm{bayes}, 1/2)] = f( \beta^g_1/2) - f( -\beta^g_1/2) = f(\beta^g_1/2) - f(\beta^g_1/2) = 0.
+\end{equation}
+Similarly, we have by (\ref{dh_dc}) that
 \begin{multline}\label{dh_dc_bayes}
-4 \cdot D_2 h(\beta^g_1, c_\textrm{bayes}, \pi) = [f( \beta^g_1/2) + f( -\beta^g_1/2)] \cdot \\ \left[  \int_{-\infty}^{-\beta^g_1/2} f + \int_{-\infty}^{\beta^g_1/2} f - \int_{-\infty}^{\beta^g_1/2} f - \int_{-\infty}^{-\beta^g_1/2} f \right] = 0.
+D_2[4 h(\beta^g_1, c_\textrm{bayes}, \pi)] = [f( \beta^g_1/2) + f( -\beta^g_1/2)] \cdot \\ \left[  \Phi(-\beta^g_1/2) + \Phi(\beta^g_1/2) -  \Phi(\beta^g_1/2) - \Phi(-\beta^g_1/2) \right] = 0.
 \end{multline}
 Plugging in (\ref{dh_dc_bayes}) and (\ref{dg_dc_bayes}) to (\ref{quotient_rule}), we find that 
-$$D_2(\beta^g_1, c_\textrm{bayes}, 1/2) = 0.$$ It follows that $c_\textrm{bayes}$ is a critical value of the bias function $b$.
+$$D_2[\gamma(\beta^g_1, c_\textrm{bayes}, 1/2)] = 0.$$ Finally, because
+$$b(\beta^g_1, c, 1/2) = 1 - \gamma(\beta^g_1, c, 1/2),$$ it follows that
+$$D_2[b(\beta^g_1, c_\textrm{bayes}, 1/2)] = -D_2[\gamma(\beta^g_1, c_\textrm{bayes}, 1/2)] = 0.$$
 
-\subsubsection{Comparison of Bayes-optimal decision boundary to large threshold}
+\subsubsection{Comparing Bayes-optimal decision boundary and large threshold}\label{sec:comparison}
 
-We study the bias produced by setting the threshold to a large number versus setting the threshold to the Bayes-optimal decision boundary.
-Let $r: \R \to \R$ be the value of attenuation function evaluated at the Bayes-optimal decision boundary $c_\textrm{bayes} = \beta^g_0 + (1/2) \beta^g_1$:
+We compare the bias produced by setting the threshold to a large number to the bias produced by setting the threshold to the Bayes-optimal decision boundary.
+Let $r: \R^{\geq 0} \to \R$ be the value of attenuation function evaluated at the Bayes-optimal decision boundary $c_\textrm{bayes} = \beta^g_0 + (1/2) \beta^g_1$, i.e.
 \begin{multline*}
@@
 We set $r$ to $1/2$ and solve for $\beta^g_1$:
-$$ 2\Phi(\beta^g_1/2) -1 = 1/2 \iff \Phi(\beta^g_1/2) = 3/4 \iff \beta^g_1 = 2 \Phi^{-1}(3/4) \approx 1.35.$$ The bias induced selecting a large threshold is $1/2$, i.e.
-$$\gamma(\beta^g_1, \infty, 1/2) = 1/2.$$
-Additionally, $r$ is a strictly increasing function.
-Therefore, $\gamma(\beta^g_1, c_\textrm{bayes},1/2) < \gamma(\beta^g_1, \infty, 1/2)$ on $\beta^g_1 \in [0, 2\Phi^{-1}(3/4))$; $\gamma(\beta^g_1, c_\textrm{bayes},1/2) = \gamma(\beta^g_1, \infty, 1/2)$ for $\beta^g_1 = 2\Phi^{-1}(3/4)$; and $\gamma(\beta^g_1, c_\textrm{bayes},1/2) > \gamma(\beta^g_1, \infty, 1/2)$ on $\beta^g_1 \in (2\Phi^{-1}(3/4), \infty)$.
-Because $b(\beta^g_1, c, 1/2) = \beta^m_1(1 - \gamma(\beta^g_1, c, 1/2))$ similar results hold for the bias (but with the inequalities flipped).
+\begin{multline*}
+r(\beta^g_1) = 1/2 \iff 2\Phi(\beta^g_1/2) -1 = 1/2 \\ \iff \Phi(\beta^g_1/2) = 3/4 \iff \beta^g_1 = 2 \Phi^{-1}(3/4) \approx 1.35.
+\end{multline*}
+Because $r$ is a strictly increasing function, it follows that $r(\beta^g_1) < 1/2$ for $\beta^g_1 < 2\Phi^{-1}(3/4)$ and $r(\beta^g_1) > 1/2$ for $\beta^g_1 > 2\Phi^{-1}(3/4).$ Next, because $$b(\beta^g_1, c_\textrm{bayes}, 1/2) = 1 - \gamma(\beta^g_1, c_\textrm{bayes}, 1/2) = 1 - r(\beta^g_1),$$ we have that $b(\beta^g_1, c_\textrm{bayes}, 1/2) > 1/2$ for $\beta^g_1 < 2 \Phi^{-1}(3/4)$ and $b(\beta^g_1, c_\textrm{bayes}, 1/2) < 1/2$ for $\beta^g_1 > 2 \Phi^{-1}(3/4)$.
+Recall that the bias induced by sending the threshold to infinity (as stated in Proposition \ref{prop:c_limit_half} and proven in Section \ref{sec:c_limit}) is $1/2$, i.e.
+$$b(\beta^g_1, \infty, 1/2) = 1/2.$$ We conclude that $b(\beta^g_1, c_\textrm{bayes},1/2) > b(\beta^g_1, \infty, 1/2)$ on $\beta^g_1 \in [0, 2\Phi^{-1}(3/4))$; $b(\beta^g_1, c_\textrm{bayes},1/2) = b(\beta^g_1, \infty, 1/2)$ for $\beta^g_1 = 2\Phi^{-1}(3/4)$; and $b(\beta^g_1, c_\textrm{bayes},1/2) < b(\beta^g_1, \infty, 1/2)$ on $\beta^g_1 \in (2\Phi^{-1}(3/4), \infty)$.
 
-\subsubsection{Limit in $\beta^g_1$}
-We compute the limit of $\gamma$ in $\beta^g_1$ given $\pi = 1/2$.  First,
-\begin{multline*}
- \lim_{\beta^g_1 \to \infty} \gamma(\beta^g_1, c, 1/2) = \frac{ 1 - \int_{-\infty}^{\beta^g_0 - c} f}{\left[ 1 + \int_{-\infty}^{ \beta^g_0 - c} f \right] \left[\int_{-\infty}^{ c - \beta^g_0} f \right]} \\ = \frac{\int_{-\infty}^{c - \beta^g_0} f }{ \left[1 + \int_{-\infty}^{ \beta^g_0 - c} f \right] \left[\int_{-\infty}^{ c - \beta^g_0} f \right]} = \frac{1}{1 + \int_{-\infty}^{\beta^g_0 - c} f} < 1.
-\end{multline*}
-Similarly,
-\begin{multline*}
-\lim_{\beta^g_1 \to -\infty} \gamma(\beta^g_1, c, 1/2) = \frac{ -\int_{-\infty}^{\beta^g_0 - c} f }{\left[ \int_{-\infty}^{\beta^g_0 - c} f \right] \left[ \int_{-\infty}^{c - \beta^g_0} f + 1 \right]} = \frac{-1}{1 + \int_{-\infty}^{c - \beta^g_0} f} > -1.
-\end{multline*}
-
-\subsubsection{Monotonicity in $\beta^g_1$}
-We show that $\gamma$ is monotonically increasing in $\beta^g_1$ for $\pi = 1/2$ and given $(\beta^g_1, \beta^g_0)$.
+\subsubsection{Monotonicity in $\beta^g_1$}\label{sec:monotone}
+We show that $\gamma$ is monotonically increasing in $\beta^g_1$ for $\pi = 1/2$ and given threshold $c$.
 We begin by stating and proving two lemmas.
@@
 $$ Finally, because $\left[\Phi(\beta^g_0 - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right] > 0,$ the entire left-hand side of (\ref{basic_ineq_cp}) is positive.
+The inequality holds for $\beta^g_1 < 0$.
 
@@
 \begin{multline*}
-\Phi(\beta^g_1 + \beta^g_0 - c) - \Phi(\beta^g_0 - c) + \Phi(c - \beta^g_0) + \Phi(c - \beta^g_0 - \beta^g_1) \\ \geq \Phi(c - \beta^g_0) + \Phi(c - \beta^g_0 - \beta^g_1). \textrm{ } \square
+\Phi(\beta^g_1 + \beta^g_0 - c) - \Phi(\beta^g_0 - c) + \Phi(c - \beta^g_0) + \Phi(c - \beta^g_0 - \beta^g_1) \\ \geq \Phi(c - \beta^g_0) + \Phi(c - \beta^g_0 - \beta^g_1). \textrm{ }
 \end{multline*}
-The second lemma establishes the derivatives of the functions $2\cdot g$ and $4 \cdot h$.
+The inequality holds for $\beta^g_1 \geq 0$. Combining the cases, the inequality holds for all $\beta^g_1 \in \R$. $\square$
+
+The second lemma establishes the derivatives of the functions $2\cdot g$ and $4 \cdot h$ in $\beta^g_1$.
 \begin{lemma}
@@
 \end{lemma}
-\textbf{Proof}: Apply the FTC and product rule. $\square$
+\textbf{Proof}: Apply FTC and product rule. $\square$
 
-We move onto the proof of the monotonicity of $\gamma$ in $\beta^g_1$.
+We are ready to prove the monotonicity of $\gamma$ in $\beta^g_1$.
 Subtracting $$\left[\Phi(\beta_0^g - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right]\left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c)\right]$$ from both sides of (\ref{basic_ineq_cp}) and multiplying by $f(\beta^g_0 + \beta^g_1 - c) > 0$ yields
@@
 \end{multline}
-We conclude by (\ref{basic_ineq_cp_5}) and (\ref{d_gamma_d_beta}) that $\gamma(\cdot, c, 1/2)$ is monotonically increasing in its first argument.
+We conclude by (\ref{basic_ineq_cp_5}) and (\ref{d_gamma_d_beta}) that $\gamma$ is monotonically increasing in $\beta^g_1$.
+Finally, $b(\beta^g_1, c, \pi) = 1 - \gamma(\beta^g_1, c, \pi)$ is monotonically decreasing in $\beta^g_1$.
 
-\subsubsection{Strict attenuation bias}
+\subsubsection{Strict attenuation bias}\label{sec:att_bias}
 
-For given $(\beta^g_0, c)$, we have that
-$$ \lim_{\beta^g_1 \to \infty} \gamma(\beta^g_1, c, 1/2, \beta^g_0) = \frac{1}{1 + \Phi(\beta^g_0 - c)},$$ and $$\lim_{\beta^g_1 \to -\infty} \gamma(\beta^g_1, c, 1/2, \beta^g_0) = -\frac{1}{1 + \Phi(c - \beta^g_0)}.$$
-Moreover, $\gamma(\beta^g_1, c, 1/2, \beta^g_0)$ is monotonically increasing in $\beta^g_1$. It follows that 
+We begin by computing the limit of $\gamma$ in $\beta^g_1$ given $\pi = 1/2$.  First,
+\begin{multline*}
+\lim_{\beta^g_1 \to \infty} \gamma(\beta^g_1, c, 1/2) = \frac{1 - \Phi(\beta^g_0 - c)}{\left[1 + \Phi(\beta^g_0 - c) \right] \left[\Phi(c - \beta^g_0) \right]} \\ = \frac{\Phi(c - \beta^g_0)}{ \left[1 + \Phi(\beta^g_0 - c) \right] \left[\Phi(c - \beta^g_0) \right]} = \frac{1}{1 + \Phi(\beta^g_0 - c)} < 1.
+\end{multline*}
+Similarly,
+\begin{multline*}
+\lim_{\beta^g_1 \to -\infty} \gamma(\beta^g_1, c, 1/2) = \frac{ - \Phi(\beta^g_0 - c)}{\left[\Phi(\beta^g_0 - c)\right] \left[\Phi(c - \beta^g_0) + 1 \right]} = \frac{-1}{1 + \Phi(c - \beta^g_0)} > -1.
+\end{multline*}
+The function $\gamma(\beta^g_1, c, 1/2, \beta^g_0)$ is monotonically increasing in $\beta^g_1$ (as stated in Proposition \ref{prop:monotonic} and proven in section \ref{sec:monotone}).
+It follows that
 $$-1 < -\frac{1}{1 + \Phi(c - \beta^g_0)} \leq \gamma(\beta^g_1, c, 1/2, \beta^g_0) \leq \frac{1}{1 - \Phi(\beta^g_0 - c)} < 1$$ for all $\beta^g_1 \in \R$.
@@
 $$-1 < \gamma(\beta^g_1, c, 1/2, \beta^g_0) < 1$$ for all $(\beta^g_1, c, \beta^g_0) \in \R^3$.
+Finally, because $b(\beta^g_1, c, 1/2, \beta^g_0) = 1 - \gamma(\beta^g_1, c, 1/2, \beta^g_0)$, it follows that
+$$ 0 < b(\beta^g_1, c, 1/2, \beta^g_0) < 2$$ for all $(\beta^g_1, c, \beta^g_0) \in \R^3$
 
-\textcolor{red}{DOES STRICT ATTENUATION BIAS HOLD UNIFORMLY OVER THE PARAMETER SPACE?}
-
-\subsubsection{Bias-variance decomposition in no-intercept model}
+\subsubsection{Bias-variance decomposition in no-intercept model}\label{sec:bv_decomp}
 
@@
 \begin{cases}
-\omega = \int_{c - \beta_g}^\infty f = 1 - \Phi(\beta_g - c) \\
-\zeta = \int_{c}^\infty f = 1 - \Phi(-c).
+\omega = \bar{\Phi}(c - \beta_g) = \Phi(\beta_g - c) \\
+\zeta = \bar{\Phi}(c) = \Phi(-c).
 \end{cases}
@@
 Finally, by (\ref{bc_decomp_1}), (\ref{bv_decomp_4}), (\ref{bv_decomp_5}), and Slutsky's Theorem,
-$$ \sqrt{n}(\hat{\beta}_m - l) \xrightarrow{d} N\left(0, \frac{ \beta_m\omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2) }{\left(\E[\hat{p}_i]\right)^2} \right).$$ Thus, for large $n \in \N$, we find that
+$$ \sqrt{n}(\hat{\beta}_m - l) \xrightarrow{d} N\left(0, \frac{ \beta_m\omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2) }{\left(\E[\hat{p}_i]\right)^2} \right).$$ Thus, for large $n \in \N$, we have that
 $$ \begin{cases}

### 2021-10-26 text/outline/outline.tex
write threshold introduction

@@
 
-In this section we study the thresholding method from empirical and theoretical perspectives. First, we define some notation.
+In this section we study thresholded regression from empirical and theoretical perspectives, highlighting several limitations to the method.
+Let $n \in \N$ (typically, $n \approx 100,000 - 250,000$) be the number of cells assayed in the experiment.
+Consider a single perturbation and a single gene.
+For cell $i \in \{1, \dots, n\}$, let $m_i \in \N$ be the number of gene transcripts sequenced; $g_i \in \N$ be the number of gRNA transcripts sequenced; $l^m_i \in \N$ be the number of gene transcripts sequenced across \textit{all} genes (sometimes called the ``library size''); and $z_i \in \R^{d-1}$ be the vector of technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.).
+The thresholding method is defined as follows:
+\begin{itemize}
+\item[1.] For a given threshold $c \in \N$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be $$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i \geq c, \\ \hat{p}_i = 1 \textrm{ if } g_i < c.
+\end{cases}$$
+\item[2.] Assume that $m_i$ is related to $\hat{p}_i, l_i,$ and $z_i$ through the following GLM:
+\begin{equation*}
+\begin{cases}
+m_i | (\hat{p}_i, z_i, l_i) \sim \textrm{NB}_\theta(\mu_i) \\
+\log(\mu_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left( l_i^m\right),
+\end{cases}
+\end{equation*}
+where (i) $NB_\theta(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta$; (ii) $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma \in  \R^{d-1}$ are unknown constants; and (iii) $\log(l_i^m)$ is an offset term.
+Fit a GLM to obtain estimates of the parameters.
 
-\blue{Let $n \approx 100,000 - 250,000$ be the number of cells in the experiment.
-Consider a single perturbation-gene pair.
-For cell $i \in \{ 1, \dots, n \},$ let $p_i \in \{ 0,1 \}$ indicate whether the cell was perturbed, $m_i \in \N$ be the number of observed gene UMIs, $g_i \in \N$ be the number of observed gRNA UMIs, $l^m_i \in \N$ be the gene library size, and $z_i \in \R^{d-1}$ be a vector of technical factors (e.g., batch, percent mitochondrial reads, etc.).
-The thresholding method is defined as follows:}
-\blue{
-\begin{itemize}
-\item[1.] For given threshold $c \in \N$, calculate the imputed value $\hat{p}_i$ of $p_i$ by $$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i \geq c, \\ \hat{p}_i = 1 \textrm{ if } g_i < c.
-\end{cases}$$
-\item[2.] Fit the regression model \cite{Sarkar2021}
-$$ m_i | \left(\hat{p}_i, z_i, l^m_i \right) \sim \textrm{NB}_\theta(\mu_i),$$ where $\theta >0$ is the NB size parameter, and $$\log\left(\mu_i\right) = \beta_m^0 + \beta_m \hat{p}_i + \gamma^T_m z_i + \log\left( l_i^m\right).$$
-\item[3.] Fit a GLM to obtain an estimate $\hat{\beta}_m$ of $\beta_m$. Compute a confidence interval and $p$-value for $\beta_m$.
-\end{itemize}}
+\item[3.] Construct a $p$-value and confidence interval for the target of inference $\beta^m_1$ using, e.g., its $z$-score.
+\end{itemize}
+
+The biological interpretation of the target of inference $\beta^m_1$ as follows: \textit{$\beta^m_1$ is the log-transformed fold change in mean gene expression in response to the perturbation, removing the effect of the technical factors}.
+The exponentiated parameter $\exp(\beta^m_1)$, which represents raw fold change, is more interpretable: it is the ratio of mean expression in perturbed cells to unperturbed cells.
+For example, for a given gene, suppose that we observe an average of $20$ transcripts in perturbed cells and an average of $40$ transcripts in unperturbed cells.
+Discounting the effect of the technical factors, the fold change in expression (i.e., $\exp(\beta^m_1)$) is $0.5$.
+The equality $\exp(\beta^m_1) = 1$ indicates that no change in expression has occurred, while $\exp(\beta^m_1) > 1$ and $\exp(\beta^m_1) < 1$ indicate an increase and decrease in expression, respectively.
 

### 2021-10-26 R_scripts/plotting/thresholding_empirical.R

@@ -5,6 +5,8 @@ library(cowplot)
 my_cols <- c("firebrick3", "dodgerblue3", "orchid4", "lightskyblue3")
 fig_dir <- paste0(.get_config_path("LOCAL_CODE_DIR"), "glmeiv-manuscript/figures/thresholding_empirical")
 if (!dir.exists(fig_dir)) dir.create(fig_dir)
+my_theme <- theme(panel.border = element_blank(), panel.grid.major = element_blank(),
+                  panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))
 
 # load the results
 gasp_result_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/gasperini/results/")
@@ -21,18 +23,19 @@ thresh_res_pc_wide <- thresh_res_pc %>% dplyr::select(target, value, threshold, 
 
 to_plot <- thresh_res_pc_wide %>% dplyr::filter(target == "estimate") %>% dplyr::select(threshold_1, threshold_5)
 p1 <- ggplot(data = to_plot, mapping = aes(x = threshold_5, y = threshold_1)) + geom_point(alpha = 0.7, col = my_cols[2]) +
-  xlab("Thresh = 5 (est)") + ylab("Thresh = 1 (est)") + geom_abline(slope = 1, intercept = 0) + theme_bw(base_size = 10)
+  xlab("Thresh = 5 (est)") + ylab("Thresh = 1 (est)") + geom_abline(slope = 1, intercept = 0) + theme_bw(base_size = 10) + my_theme
+
 # threshold = 1 ests are bigger than (i.e., closer to one than) threshold = 5; this is attenuation bias due to the fact that excessively many unperturbed cells have been classified as perturbed.
 
 # a) est. vs. est.
 to_plot <- thresh_res_pc_wide %>% dplyr::filter(target == "estimate") %>% dplyr::select(threshold_5, threshold_20)
 p2 <- ggplot(data = to_plot, mapping = aes(x = threshold_5, y = threshold_20)) + geom_point(alpha = 0.9, col = my_cols[2]) +
-  xlab("Thresh = 5 (est)") + ylab("Thresh = 20 (est)") + geom_abline(slope = 1, intercept = 0, lwd = 0.75) + theme_bw(base_size = 10)
[…]

### 2021-10-27 text/outline/outline.tex
ls

--- 2021-10-25 text/outline/outline.tex
@@
 
-In this section we study the thresholding method from empirical and theoretical perspectives. First, we define some notation.
+In this section we study thresholded regression from empirical and theoretical perspectives, highlighting several fundamental challenges of the method.
+Let $n \in \N$ be the number of cells assayed in the experiment (typically, $n \approx 100,000 - 250,000$).
+Consider a single perturbation and a single gene.
+For cell $i \in \{1, \dots, n\}$, let $m_i \in \N$ be the number of gene transcripts sequenced in the cell; let $g_i \in \N$ be the number of gRNA transcripts sequenced in the cell; let $l^m_i \in \N$ be the number of gene transcripts sequenced across \textit{all} genes (the library size) in the cell; and finally, let $z_i \in \R^{d-1}$ be the cell-specific vector of technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.).
+The thresholding method is defined as follows:
+\begin{itemize}
+\item[1.] For a given threshold $c \in \N$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be $$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i \geq c, \\ \hat{p}_i = 1 \textrm{ if } g_i < c.
+\end{cases}$$
+\item[2.] Assume that $m_i$ is related to $\hat{p}_i, l_i,$ and $z_i$ through the following GLM:
+\begin{equation*}
+\begin{cases}
+m_i | (\hat{p}_i, z_i, l_i) \sim \textrm{NB}_\theta(\mu_i), \\
+\log(\mu_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left( l_i^m\right),
+\end{cases}
+\end{equation*}
+where (i) $NB_\theta(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta$; (ii) $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma \in  \R^{d-1}$ are unknown parameters; and (iii) $\log(l_i^m)$ is an offset term.
+Fit a GLM to obtain estimates of the parameters.
 
-\blue{Let $n \approx 100,000 - 250,000$ be the number of cells in the experiment.
-Consider a single perturbation-gene pair.
-For cell $i \in \{ 1, \dots, n \},$ let $p_i \in \{ 0,1 \}$ indicate whether the cell was perturbed, $m_i \in \N$ be the number of observed gene UMIs, $g_i \in \N$ be the number of observed gRNA UMIs, $l^m_i \in \N$ be the gene library size, and $z_i \in \R^{d-1}$ be a vector of technical factors (e.g., batch, percent mitochondrial reads, etc.).
-The thresholding method is defined as follows:}
-\blue{
-\begin{itemize}
-\item[1.] For given threshold $c \in \N$, calculate the imputed value $\hat{p}_i$ of $p_i$ by $$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i \geq c, \\ \hat{p}_i = 1 \textrm{ if } g_i < c.
-\end{cases}$$
-\item[2.] Fit the regression model \cite{Sarkar2021}
-$$ m_i | \left(\hat{p}_i, z_i, l^m_i \right) \sim \textrm{NB}_\theta(\mu_i),$$ where $\theta >0$ is the NB size parameter, and $$\log\left(\mu_i\right) = \beta_m^0 + \beta_m \hat{p}_i + \gamma^T_m z_i + \log\left( l_i^m\right).$$
-\item[3.] Fit a GLM to obtain an estimate $\hat{\beta}_m$ of $\beta_m$. Compute a confidence interval and $p$-value for $\beta_m$.
-\end{itemize}}
+\item[3.] Compute a $p$-value and confidence interval for the target of inference $\beta^m_1$.
+\end{itemize}
+
+The biological interpretation for the target of inference $\beta^m_1$ as follows:  $\beta^m_1$ is the log-transformed fold change in gene expression in response to the perturbation, removing the effect of the technical factors.
+Fold change (obtained by exponentiating $\beta^m_1$) is the ratio of the mean expression in cells that were perturbed to the mean expression in cells that were not perturbed.
+$\exp(\beta^m_1) = 1$ indicates no change in mean expression, while $\exp(\beta^m_1) > 1$ and $\exp(\beta^m_1) < 1$ indicate an increase and decrease in expression in response to the perturbation, respectively.
 
@@
 
-\blue{To investigate the impact of threshold selection on the thresholding method, we applied the thresholding method to the set of positive control (i.e., gene-targeting) perturbation-gene pairs in the Gasperini data using three different choices for the threshold: 1, 5, and 20.}
+We studied the behavior of the thresholding method on real data, uncovering an attenuation bias effect and a bias-variance tradeoff as a function of the selected threshold.
+We applied the thresholding method to analyze the set of 381 positive control perturbation-gene pairs in the Gasperini dataset.
+The positive control pairs consisted of perturbations that targeted gene transcription start sites (TSSs) for inhibition.
+Repressing the TSS of a given gene decreases its expression; therefore, the positive control pairs \textit{a priori} are expected to exhibit a strong, negative log fold change in expression.
 
-\blue{Description of panels a-d; threshold = 1 leads to considerable attenuation bias (a); threshold = 5 and threshold = 20 produce similar estimates, though the effect sizes are slightly greater for threshold = 20, suggesting threshold = 5 yields mild attenuation bias (b); the threshold = 20 estimates are more variable than than those of threshold = 5, as evidenced by the smaller $p$-values and wider confidence intervals.}
+To investigate the sensitivity of the thresholding method to the selected threshold, we deployed the thresholding method using three different choices for the threshold: 1, 5, and 20.
+We found that the chosen threshold substantially impacted the results (Figure \ref{thresholding_empirical}a-b).
+Estimates for log fold change produced by setting the threshold to 1 were smaller in magnitude than those produced by setting the threshold to 5.
+(Equivalently, estimates for \textit{raw} fold change were closer to the baseline of $1$ for threshold = 1; Figure \ref{thresholding_empirical}a.) Fold change estimates generated by setting the threshold to 5 and setting the threshold to 20 were more concordant; however, the threshold of 20 yielded slightly stronger effect sizes (Figure \ref{thresholding_empirical}b).
+These results are consistent with an \textit{attenuation bias} phenomenon: small thresholds tend to underestimate the true parameter value, and this effect diminishes gradually as the threshold increases.
 
-\blue{Description of panels e-f: these are the empirical gRNA count distributions for randomly selected gRNAs from the Gasperini and Xie datasets.
+We hypothesized that selecting a large threshold comes with the cost of increasing the variance of the estimator.
+
+% \blue{To investigate the impact of threshold selection on the thresholding method, we applied the thresholding method to the set of positive control (i.e., gene-targeting) perturbation-gene pairs in the Gasperini data using three different choices for the threshold: 1, 5, and 20.}
+
+% \blue{Description of panels a-d; threshold = 1 leads to considerable attenuation bias (a); threshold = 5 and threshold = 20 produce similar estimates, though the effect sizes are slightly greater for threshold = 20, suggesting threshold = 5 yields mild attenuation bias (b); the threshold = 20 estimates are more variable than than those of threshold = 5, as evidenced by the smaller $p$-values and wider confidence intervals.}
+
+% \blue{Description of panels e-f: these are the empirical gRNA count distributions for randomly selected gRNAs from the Gasperini and Xie datasets.
 There does not appear to be a clear location at which to draw the threshold; aside from the initial spike at one (zero not shown), the histograms gradually decrease.}
 
-\blue{Take-home message: (i) the threshold is a tuning parameter that substantially affects the result; (ii) as the threshold increases, bias seems to decrease and variance seems to increase; (iii) it is not clear where to draw the threshold from gRNA counts alone.}
+% \blue{Take-home message: (i) the threshold is a tuning parameter that substantially affects the result; (ii) as the threshold increases, bias seems to decrease and variance seems to increase; (iii) it is not clear where to draw the threshold from gRNA counts alone.}
 

### 2021-10-27 text/outline/outline.tex
add fig 1

@@
 Our work makes two main contributions.
-First, we conduct an in-depth study of the thresholding method from empirical and theoretical perspectives.
+First, we conduct an investigation of the thresholding method from empirical and theoretical perspectives.
 Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian model.
@@
 	\centering
-	\includegraphics[width=0.9\linewidth]{analysis_challenges}
-	\caption{\textbf{Overview of experimental design and analysis challenges}: \textbf{a,} Experimental design.
+	\includegraphics[width=1\linewidth]{../../figures/analysis_challenges/plot.pdf}
+	\caption{\textbf{Experimental design and analysis challenges}: \textbf{a,} Experimental design.
 For a given perturbation (e.g., the perturbation represented in yellow), we partition the cells into two groups: those that received the perturbation, and those that did not receive the perturbation.
@@
 
-In this section we study thresholded regression from empirical and theoretical perspectives, highlighting several fundamental challenges of the method.
+In this section we study thresholded regression from empirical and theoretical perspectives, highlighting several limitations of the method.
 Let $n \in \N$ be the number of cells assayed in the experiment (typically, $n \approx 100,000 - 250,000$).
@@
 Fold change (obtained by exponentiating $\beta^m_1$) is the ratio of the mean expression in cells that were perturbed to the mean expression in cells that were not perturbed.
-$\exp(\beta^m_1) = 1$ indicates no change in mean expression, while $\exp(\beta^m_1) > 1$ and $\exp(\beta^m_1) < 1$ indicate an increase and decrease in expression in response to the perturbation, respectively.
+$\exp(\beta^m_1) = 1$ indicates no change in mean expression, while $\exp(\beta^m_1) > 1$ and $\exp(\beta^m_1) < 1$ indicate an increase and decrease in expression in response to the perturbation, respectively (accounting for technical factors).
 
@@
 
-We studied the behavior of the thresholding method on real data, uncovering an attenuation bias effect and a bias-variance tradeoff as a function of the selected threshold.
+We examined the behavior of the thresholding method on real data and uncovered attenuation bias and bias-variance tradeoff effects.
 We applied the thresholding method to analyze the set of 381 positive control perturbation-gene pairs in the Gasperini dataset.
@@
 
-To investigate the sensitivity of the thresholding method to the selected threshold, we deployed the thresholding method using three different choices for the threshold: 1, 5, and 20.
+To investigate the sensitivity of the thresholding method to the selected threshold, we deployed the thresholding method on the positive control data using three different thresholds: 1, 5, and 20.
 We found that the chosen threshold substantially impacted the results (Figure \ref{thresholding_empirical}a-b).
-Estimates for log fold change produced by setting the threshold to 1 were smaller in magnitude than those produced by setting the threshold to 5.
-(Equivalently, estimates for \textit{raw} fold change were closer to the baseline of $1$ for threshold = 1; Figure \ref{thresholding_empirical}a.) Fold change estimates generated by setting the threshold to 5 and setting the threshold to 20 were more concordant; however, the threshold of 20 yielded slightly stronger effect sizes (Figure \ref{thresholding_empirical}b).
-These results are consistent with an \textit{attenuation bias} phenomenon: small thresholds tend to underestimate the true parameter value, and this effect diminishes gradually as the threshold increases.
+Estimates for log fold change produced by threshold = 1 were smaller in magnitude than those produced by threshold = 5.
+(Equivalently, estimates for \textit{raw} fold change were closer to the baseline of $1$ for threshold = 1; Figure \ref{thresholding_empirical}a.) Estimates produced by threshold = 5 and threshold = 20 were more concordant, but threshold = 20 yielded slightly larger effect sizes (Figure \ref{thresholding_empirical}b).
 
-We hypothesized that selecting a large threshold comes with the cost of increasing the variance of the estimator.
+We reasoned that thresholded regression underestimated effect sizes on the positive control pairs, especially for threshold = 1 (an example of \textit{attenuation bias}).
+For a given perturbation, the vast majority ($>99\%$) of cells are unperturbed.
+This imbalance leads to an asymmetry: misclassifying \textit{unperturbed} cells as \textit{perturbed} is intuitively ``worse'' than misclassifying \textit{perturbed} cells as \textit{unperturbed}.
+Misclassified \textit{unperturbed} cells contaminate the set of truly perturbed cells, leading to attenuation bias; by contrast, misclassified \textit{perturbed} cells are swamped in number and ``neutralized'' by the truly \textit{unperturbed} cells.
+Setting the threshold to a large number reduces the unperturbed-to-perturbed misclassification rate, decreasing bias.
 
-% \blue{To investigate the impact of threshold selection on the thresholding method, we applied the thresholding method to the set of positive control (i.e., gene-targeting) perturbation-gene pairs in the Gasperini data using three different choices for the threshold: 1, 5, and 20.}
+We hypothesized, however, that the reduction in bias conferred by selecting a large threshold incurs the cost of increasing the variance of the estimator.
+To investigate, we compared $p$-values and confidence intervals produced by threshold = 5 and threshold = 20 for the target of inference $\beta^m_1$.
+We found that threshold = 5 yielded smaller (i.e., more significant) $p$-values and narrower confidence intervals than did threshold = 20 (Figure \ref{thresholding_empirical}c-d).
+We concluded that thresholded regression has a bias-variance tradeoff: as the threshold increases, bias decreases and variance increases.
 
-% \blue{Description of panels a-d; threshold = 1 leads to considerable attenuation bias (a); threshold = 5 and threshold = 20 produce similar estimates, though the effect sizes are slightly greater for threshold = 20, suggesting threshold = 5 yields mild attenuation bias (b); the threshold = 20 estimates are more variable than than those of threshold = 5, as evidenced by the smaller $p$-values and wider confidence intervals.}
-
-% \blue{Description of panels e-f: these are the empirical gRNA count distributions for randomly selected gRNAs from the Gasperini and Xie datasets.
-There does not appear to be a clear location at which to draw the threshold; aside from the initial spike at one (zero not shown), the histograms gradually decrease.}
-
-% \blue{Take-home message: (i) the threshold is a tuning parameter that substantially affects the result; (ii) as the threshold increases, bias seems to decrease and variance seems to increase; (iii) it is not clear where to draw the threshold from gRNA counts alone.}
+Finally, to determine whether there is an ``obvious'' location at which to draw the threshold, we examined the empirical gRNA count distributions and checked for bimodality.
+Figures \ref{thresholding_empirical}e and \ref{thresholding_empirical}f display the empirical distribution of a randomly-selected gRNA from the Gasperini and Xie datasets, respectively (counts of $0$ omitted).
+The distributions peak at $1$ and then taper off gradually; there does not exist a sharp boundary that cleanly separates the perturbed from the unperturbed cells.
+Overall, we concluded that the thresholding method faces several challenges: (i) the threshold is a tuning parameter that significantly impacts the results; (ii) the threshold mediates an intrinsic bias-variance tradeoff; and (iii) the gRNA count distributions do not imply a clear threshold selection strategy.
 
@@
 	\includegraphics[width=1\linewidth]{../../figures/thresholding_empirical/plot}
-	\caption{\textbf{Empirical challenges of thresholded regression.}}
+	\caption{\textbf{Empirical challenges of thresholded regression.} \textbf{a-b,} Fold change estimates produced by threshold = 1 versus threshold = 5 (a) and threshold = 20 versus threshold = 5 (b).
+The selected threshold substantially impacts the results.
+\textbf{c-d,} $p$-values (c) and CI widths (d) produced by threshold = 20 versus threshold = 5.
+The latter threshold yields more confident estimates.
+\textbf{e-f}, Empirical distribution of randomly-selected gRNA from Gasperini (e) and Xie (f) data.
+The gRNA data do not appear to imply an obvious threshold selection strategy.}
 	\label{thresholding_empirical}

### 2021-10-27 R_scripts/plotting/analysis_challenges.R

@@ -1,21 +1,20 @@
-# set figure 1 dir
-fig_dir <- paste0(.get_config_path("LOCAL_CODE_DIR"), "glmeiv-manuscript/figures/")
-fig1_dir <- paste0(fig_dir, "fig1/")
-if (!dir.exists(fig1_dir)) dir.create(fig1_dir, recursive = TRUE)
+library(magrittr)
+library(ggplot2)
+library(cowplot)
 
-# load plotting objects
-plotting_objects <- paste0(.get_config_path("LOCAL_CODE_DIR"), "glmeiv-manuscript/figures/plotting_objects.R")
-source(plotting_objects)
+my_cols <- c("firebrick3", "dodgerblue3", "orchid4", "lightskyblue3")
+fig_dir <- paste0(.get_config_path("LOCAL_CODE_DIR"), "glmeiv-manuscript/figures/analysis_challenges")
+if (!dir.exists(fig_dir)) dir.create(fig_dir)
 
 # a. create mixture distributions
 x_grid <- seq(-1.2, 1.7, 0.01)
 to_plot <- rbind(data.frame(x = x_grid,
            y = dnorm(x = x_grid, mean = 0, sd = 0.25),
-           modality = "mRNA",
+           modality = "gene",
            Status = "Unperturbed"),
 data.frame(x = x_grid,
            y = dnorm(x = x_grid, mean = -0.4, sd = 0.25),
-           modality = "mRNA",
+           modality = "gene",
            Status = "Perturbed"),
 data.frame(x = x_grid,
            y = dnorm(x = x_grid, mean = 1, sd = 0.25),
@@ -24,19 +23,20 @@ data.frame(x = x_grid,
 data.frame(x = x_grid,
            y = dnorm(x = x_grid, mean = 0, sd = 0.25),
            modality = "gRNA",
-           Status = "Unperturbed")) %>% dplyr::mutate(modality = factor(x = modality, levels = c("mRNA", "gRNA")))
-my_cols <- c("royalblue4", "seagreen4")
+           Status = "Unperturbed")) %>% dplyr::mutate(modality = factor(x = modality, levels = c("gene", "gRNA")))
 p1 <- ggplot(data = to_plot, mapping = aes(x = x, y = y, col = Status)) +
   facet_grid(modality~.) + ylab("Density") + xlab("Expression") + theme_bw() +
   theme(legend.position="bottom") + theme(legend.position="bottom", legend.title=element_blank()) +
[…]

### 2021-10-27 text/outline/outline.tex
update fig 1 and background lit

@@
 Our work makes two main contributions.
-First, we conduct an investigation of the thresholding method from empirical and theoretical perspectives.
+First, we study the thresholding method from empirical and theoretical perspectives.
 Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian model.
@@
 \section{Background and analysis challenges}
+
+\subsection{Related work}
+
+Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions.
+For example, Lin, Lei, and Roeder \cite{Lin2021} developed eSVD, an extension of SVD to exponential family and curved Gaussian responses.
+Unlike SVD, eSVD models the relationship between the mean and variance of a gene's expression level, a phenomenon induced by the countedness of single-cell data \cite{Lause2021}.
+Similarly, Townes et al.\ \cite{Townes2019} proposed GLM-PCA, a generalization of PCA that directly models Poisson- or negative binomially-distributed gene expression counts.
+We see our work as a continuation of this broad effort to ``port'' common statistical methods and models to single-cell count data.
+Our focus, however, is on regression rather than dimension reduction: we extend the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.
+
+The closest parallels to our work in the statistical methodology literature are Gr\"{u}n \& Leisch \cite{Grun2008} and Ibrahim \cite{Ibrahim1990}.
+Gr\"{u}n \& Leisch derived a method for estimation and inference in a $k$-component mixture of GLMs.
+While we prefer to view GLM-EIV as a generalized errors-in-variables method,  the GLM-EIV model is equivalent to a two-component mixture of \textit{products} of GLM densities.
+Ibrahim proposed a procedure for fitting GLMs in the presence of missing-at-random covariates.
+Our method, by contrast, involves fitting two conditionally independent GLMs in the presence of a totally latent covariate.
+Thus, while Ibrahim and Gr\"{u}n \& Leisch are helpful references, our estimation and inference tasks are more complex.
+
+The genomics literature has produced several applied methods for single-cell CRISPR screen analysis.
+In a separate work we proposed SCEPTRE \cite{Barry2020}, a custom implementation of the conditional randomization test \cite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data.
+SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and providing robustness to expression model misspecification.
+Other applied methods for single-cell CRISPR screen analysis include MIMOSCA \cite{Dixit2016} and scMAGeCK \cite{Yang2019}, which, like SCEPTRE, focus on hypothesis testing, but unlike SCEPTRE, ignore the countedness of the data and do not adjust for confounders.
+Our focus in this work are on challenges complimentary to those that SCEPTRE addresses; most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression, moving beyond hypothesis testing.
+
+\subsection{Assay overview}
+
+\subsection{Analysis challenges}
 
@@
 \end{figure}
-
-\section{Related work}
-
-Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions.
-For example, Lin, Lei, and Roeder \cite{Lin2021} developed eSVD, an extension of SVD to exponential family and curved Gaussian responses.
-Unlike SVD, eSVD models the relationship between the mean and variance of a gene's expression level, a phenomenon induced by the countedness of single-cell data \cite{Lause2021}.
-Similarly, Townes et al.\ \cite{Townes2019} proposed GLM-PCA, an extension of PCA that directly models Poisson- or negative binomially-distributed gene expression counts.
-We see our work as a continuation of this broad effort to ``port'' common statistical methods and models to single-cell count data.
-Our focus, however, is on regression rather than dimension reduction: we extend the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.
-
-The closest parallels to our work in the statistical methodology literature are Gr\"{u}n \& Leisch \cite{Grun2008} and Ibrahim \cite{Ibrahim1990}.
-Gr\"{u}n \& Leisch considered estimation and inference in a $k$-component mixture of GLMs.
-While we prefer to view the GLM-EIV model as an errors-in-variables model,  the GLM-EIV model is equivalent to a two-component mixture of \textit{products} of GLM densities.
-Ibrahim derived a method for fitting GLMs in the presence of missing-at-random covariates.
-Our method, by contrast, involves fitting two conditionally independent GLMs in the presence of a totally latent covariate.
-Thus, while Ibrahim is a helpful reference, our estimation and inference tasks are more complex.
-
-The genomics literature has produced several applied methods for linking perturbations to changes in gene expression in single-cell CRISPR screens: SCEPTRE \cite{Barry2020}, MIMOSCA \cite{Dixit2016}, and scMAGeCK \cite{Yang2019}.
-These methods in general are focused on hypothesis testing rather than estimation; none, for instance, produces a confidence interval for the effect size of a perturbation on gene expression change.
-Additionally, two of these methods (MIMOSCA and scMAGeCK) use (possibly penalized) linear models to model gene expressions, thereby disregarding the countedness and sparsity of the data.
 

### 2021-10-27 R_scripts/plotting/analysis_challenges.R

@@ -29,14 +29,14 @@ p1 <- ggplot(data = to_plot, mapping = aes(x = x, y = y, col = Status)) +
   theme(legend.position="bottom") + theme(legend.position="bottom", legend.title=element_blank()) +
   geom_vline(data = dplyr::filter(to_plot, modality == "gene"), mapping = aes(xintercept = -0.4), col = "darkred") +
   geom_vline(data = dplyr::filter(to_plot, modality == "gene"), mapping = aes(xintercept = 0.0), col = "darkred") +
-  geom_hline(yintercept = 0) + scale_y_continuous(expand = c(0, 0.05)) + 
-  scale_color_manual(values = c(my_cols[2], my_cols[3])) + geom_line(lwd = 1.1) +
+  geom_hline(yintercept = 0) + scale_y_continuous(expand = c(0, 0.05)) + scale_x_continuous(expand = c(0, 0)) + 
+  scale_color_manual(values = c(my_cols[2], "black")) + geom_line(lwd = 1.1) +
   theme(legend.position = c(0.77, 0.85), legend.title = element_blank(),
         panel.border = element_blank(), panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(), axis.line.y = element_line(colour = "black"))
 
 # set up cowplot
 p_out <- plot_grid(NULL, p1, NULL, NULL, labels = c("a", "c", "b", "d"), nrow = 2, rel_heights = c(0.5, 0.5))
-f_name <- paste0(fig_dir, "/ggplot.jpg")
+f_name <- paste0(fig_dir, "/ggplot.pdf")
 
-ggsave(filename = f_name, plot = p_out, device = "jpg", scale = 1, width = 7, height = 6, dpi = 320)
+ggsave(filename = f_name, plot = p_out, device = "pdf", scale = 1, width = 7, height = 6, dpi = 320)

### 2021-10-28 text/outline/outline.tex
add assay overview text

@@
 Our method, by contrast, involves fitting two conditionally independent GLMs in the presence of a totally latent covariate.
-Thus, while Ibrahim and Gr\"{u}n \& Leisch are helpful references, our estimation and inference tasks are more complex.
+Thus, while Ibrahim and Gr\"{u}n \& Leisch are helpful references, our estimation and inference tasks are more complex than theirs.
 
 The genomics literature has produced several applied methods for single-cell CRISPR screen analysis.
-In a separate work we proposed SCEPTRE \cite{Barry2020}, a custom implementation of the conditional randomization test \cite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data.
-SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and providing robustness to expression model misspecification.
-Other applied methods for single-cell CRISPR screen analysis include MIMOSCA \cite{Dixit2016} and scMAGeCK \cite{Yang2019}, which, like SCEPTRE, focus on hypothesis testing, but unlike SCEPTRE, ignore the countedness of the data and do not adjust for confounders.
-Our focus in this work are on challenges complimentary to those that SCEPTRE addresses; most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression, moving beyond hypothesis testing.
+In a prior work we developed SCEPTRE \cite{Barry2020}, a custom implementation of the conditional randomization test \cite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data.
+SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification.
+Other applied methods for single-cell CRISPR screen analysis include MIMOSCA \cite{Dixit2016} and scMAGeCK \cite{Yang2019}.
+These methods, like SCEPTRE, focus on hypothesis testing (rather than estimation), but unlike SCEPTRE, they ignore the countedness of the data and are unable to handle confounders.
+In this work we tackle a set of analysis challenges that are complimentary to the challenges addressed by SCEPTRE; most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, moving beyond hypothesis testing.
 
-\subsection{Assay overview}
+\subsection{Assay overview and analysis challenges}
 
-\subsection{Analysis challenges}
+There are several broad classes of single-cell CRISPR screen assays, each suited to answer a different set of biological questions \cite{Gasperini2019,Datlinger2021,Mimitou2019}.
+In this work we focus on so-called high-multiplicity of infection (MOI) single-cell CRISPR screens.
+We expect the ideas that we develop for this assay to apply (with some effort) to other classes of single-cell CRISPR screens as well.
+This section motivates high MOI single-cell screens and provides some background on the experimental protocol and relevant analysis challenges.
 
-Our focus in this work is on high multiplicity-of-infection (MOI), enhancer-targeting, single-cell CRISPR screens. In this section we cover relevant biological background and motivation.
-\\ \\
-\textbf{P1}: The human genome consists of genes (segments of DNA that code for proteins), enhancers (segments of DNA that regulate the expression of one or more genes), and other genomic regions.
-Genome-wide association studies have revealed that the majority ($>95\%$) of variants associated with diseases lie outside genes and (very likely) inside enhancers.
-These noncoding variants contribute to disease by modulating the expression one or more genes, which in turn encode proteins that affect the phenotype.
-A central open challenge in genetics, therefore, is to link enhancers that harbor disease-associated variants to the genes that they target at genome-wide scale.
-\\ \\
-\textbf{P2}: High MOI single-cell CRISPR screens are the most promising biotechnology for solving this problem.
-\blue{Describe the experimental protocol here.
-Explain that we use the terms ``barcodes'' and ``gRNAs'' interchangeably, as polyadenylated gRNAs serve as barcodes in CROP-seq.
-Link to Figure \ref{analysis_challenges}.}
-\\ \\
-\textbf{P3}: Single-cell CRISPR screens pose several core analysis challenges.
+The human genome consists of genes, enhancers (segments of DNA that regulate the expression of one or more genes), and other genomic elements (that are not of importance to the current discussion).
+Genome-wide association studies (GWAS) have revealed that the majority ($>90\%$) of variants associated with diseases lie outside genes and (very likely) inside enhancers \cite{Gallagher2018}.
+These noncoding variants are thought to contribute to disease by modulating the expression one or more disease-relevant genes.
+Unfortunately, we do not know which gene (or genes) through which most noncoding variants exert their effect, limiting the interpretability of GWAS results.
+A central open challenge in genetics, therefore, is to link enhancers that harbor GWAS variants to the genes that they target at genome-wide scale \cite{Gasperini2020}.
+
+The most promising biotechnology for solving this challenge are high MOI single-cell CRISPR screens.
+High MOI single-cell CRISPR screens combine CRISPR interference (CRISPRi) -- a variant of CRISPR that \textit{represses} a targeted region of the genome, leaving the primary DNA sequence intact  -- with single-cell sequencing.
+The experimental protocol, at a high level, is as follows.
+First, the scientist designs a library of several hundred to several thousand CRISPRi perturbations, each targeting a candidate enhancer for repression.
+Next, the scientist cultures several hundred thousand cells and delivers the CRISPRi perturbations to these cells.
+The perturbations assort into the cells randomly, with each cell receiving on average 10-40 distinct perturbations.
+
+After waiting several days for CRISPRi to take effect, the scientist profiles each cell's transcriptome (i.e., its set of gene expressions) and perturbations.
+Finally, the scientist conducts perturbation-to-gene association analyses.
+Figure \ref{analysis_challenges}a depicts this process schematically, with colored bars (blue, red, and purple) representing distinct perturbations.
+For a given perturbation (e.g., the perturbation represented in blue), the scientist partitions the cells into two sets: those that received the perturbation (top) and those that did not (bottom).
+Next, for a given gene, the scientist runs a differential expression analysis across the two groups of cells, producing an estimate for the change in expression in response to the perturbation.
+If the estimated change in expression is large, the scientist can conclude that the enhancer targeted by the perturbation exerts a strong regulatory effect on the gene.
+This procedure is repeated for a large set of preselected perturbation-gene pairs.
+
+% the scientist carries out the standard single-cell sequencing procedure to assay each cell's transcriptome -- -- and
+
+% Discuss bardodes and gRNAs in the subsequent paragraph.
+% \textbf{P3}: Single-cell CRISPR screens pose several core analysis challenges.
 \blue{Describe the analysis challenges here: (i) unobserved perturbation; (ii) existence of background reads; (iii) highly discrete count data; (iv) nuisance variables}.

### 2021-10-29 text/outline/outline.tex
finish draft of part 1

@@
 As a consequence, one cannot know with certainty which cells were perturbed.
-Instead, one must leverage an indirect, noisy proxy of perturbation presence or absence -- namely, transcribed barcode counts -- to ``guess'' which cells were perturbed.
+Instead, one must leverage an indirect, noisy proxy of perturbation presence or absence -- namely, transcribed guide RNA counts -- to ``guess'' which cells were perturbed.
 Using these imputed perturbation assignments, one can attempt to estimate the effect of the perturbation on gene expression.
-The standard approach, which we call ``thresholded regression'' or the ``thresholding method,'' is to assign perturbation identities to cells by simply thresholding the barcode counts.
+The standard approach, which we call ``thresholded regression'' or the ``thresholding method,'' is to assign perturbation identities to cells by simply thresholding the guide RNA counts.
 
@@
 The goal of the analysis is to estimate the effect of $x^*$ on $y$ using the observed data $(x , y)$ only.
-In the context of the biological application, $x^*$, $y$, and $x$ are CRISPR perturbations, gene expressions, and barcode counts, respectively.
+In the context of the biological application, $x^*$, $y$, and $x$ are CRISPR perturbations, gene expressions, and guide RNA counts, respectively.
 
@@
 These methods, like SCEPTRE, focus on hypothesis testing (rather than estimation), but unlike SCEPTRE, they ignore the countedness of the data and are unable to handle confounders.
-In this work we tackle a set of analysis challenges that are complimentary to the challenges addressed by SCEPTRE; most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, moving beyond hypothesis testing.
+In this work we tackle a set of analysis challenges that are complimentary to the challenges addressed by SCEPTRE.
+Most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, moving beyond hypothesis testing.
 
-\subsection{Assay overview and analysis challenges}
+\subsection{Assay overview}
+
+\begin{figure}
+	\centering
+	\includegraphics[width=1\linewidth]{../../figures/analysis_challenges/plot.pdf}
+	\caption{\textbf{Experimental design and analysis challenges}: \textbf{a,} Experimental design.
+For a given perturbation (e.g., the perturbation indicated in blue), we partition the cells into two groups: perturbed and unperturbed.
+Next, for a given gene, we conduct a differential expression analysis across the two groups, yielding an estimate of the impact of the given perturbation on the given gene.
+\textbf{b,} DAG representing all variables in the system.
+The perturbation (latent) impacts both gene expression and gRNA expression; technical factors act as nuisance variables, also impacting gene and gRNA expression.
+The target of estimation is the effect of the perturbation on gene expression.
+\textbf{c,} Schematic illustrating the ``background read'' phenomenon.
+Due to errors in the sequencing process, unperturbed cells exhibit a nonzero gRNA count distribution (bottom).
+The target of estimation is the change in mean gene expression in response to the perturbation (top).
+\textbf{d}, Example data on four cells for a given perturbation-gene pair.
+Note that (i) the perturbation is unobserved, and (ii) the gene and gRNA data are discrete counts.}
+	\label{analysis_challenges}
+\end{figure}
 
@@
 We expect the ideas that we develop for this assay to apply (with some effort) to other classes of single-cell CRISPR screens as well.
-This section motivates high MOI single-cell screens and provides some background on the experimental protocol and relevant analysis challenges.
+In this section we motivate high MOI single-cell screens, overview the experimental protocol, and present relevant analysis challenges.
 
@@
 These noncoding variants are thought to contribute to disease by modulating the expression one or more disease-relevant genes.
-Unfortunately, we do not know which gene (or genes) through which most noncoding variants exert their effect, limiting the interpretability of GWAS results.
+We do not know the gene (or genes) through which most noncoding variants exert their effect, limiting the interpretability of GWAS results.
 A central open challenge in genetics, therefore, is to link enhancers that harbor GWAS variants to the genes that they target at genome-wide scale \cite{Gasperini2020}.
@@
 The most promising biotechnology for solving this challenge are high MOI single-cell CRISPR screens.
-High MOI single-cell CRISPR screens combine CRISPR interference (CRISPRi) -- a variant of CRISPR that \textit{represses} a targeted region of the genome, leaving the primary DNA sequence intact  -- with single-cell sequencing.
-The experimental protocol, at a high level, is as follows.
-First, the scientist designs a library of several hundred to several thousand CRISPRi perturbations, each targeting a candidate enhancer for repression.
-Next, the scientist cultures several hundred thousand cells and delivers the CRISPRi perturbations to these cells.
+High MOI single-cell CRISPR screens combine CRISPR interference (CRISPRi) -- a version of CRISPR that \textit{represses} a targeted region of the genome, leaving the primary DNA sequence intact  -- with single-cell sequencing.
+The experimental protocol is as follows.
+First, the scientist develops a library of several hundred to several thousand CRISPRi perturbations, each designed to target a candidate enhancer for repression.
+The scientist then cultures tens or hundreds of thousands of cells and delivers the CRISPRi perturbations to the cells.
 The perturbations assort into the cells randomly, with each cell receiving on average 10-40 distinct perturbations.
+Conversely, a given perturbation enters fewer than 0.1-2\% of cells.
 
-After waiting several days for CRISPRi to take effect, the scientist profiles each cell's transcriptome (i.e., its set of gene expressions) and perturbations.
+After waiting several days for CRISPRi to take effect, the scientist profiles each cell's transcriptome (i.e., its gene expressions) and the set of perturbations that it received.
 Finally, the scientist conducts perturbation-to-gene association analyses.
 Figure \ref{analysis_challenges}a depicts this process schematically, with colored bars (blue, red, and purple) representing distinct perturbations.
-For a given perturbation (e.g., the perturbation represented in blue), the scientist partitions the cells into two sets: those that received the perturbation (top) and those that did not (bottom).
-Next, for a given gene, the scientist runs a differential expression analysis across the two groups of cells, producing an estimate for the change in expression in response to the perturbation.
-If the estimated change in expression is large, the scientist can conclude that the enhancer targeted by the perturbation exerts a strong regulatory effect on the gene.
+For a given perturbation (e.g., the perturbation represented in blue), the scientist partitions the cells into two groups: those that received the perturbation (top) and those that did not (bottom).
+Next, for a given gene, the scientist runs a differential expression analysis across the two groups of cells, producing an estimate for the magnitude of gene expression change in response to the perturbation.
+If the estimated change in expression is large, the scientist can conclude that the enhancer \textit{targeted} by the perturbation exerts a strong regulatory effect on the gene.
 This procedure is repeated for a large set of preselected perturbation-gene pairs.
+% If the differential expression analyses produce $p$-values, the $p$-values are subjected to a multiple hypothesis testing correction procedure.
 
-% the scientist carries out the standard single-cell sequencing procedure to assay each cell's transcriptome -- -- and
+\subsection{Analysis challenges and data}
+High MOI single-cell CRISPR screens present several statistical challenges.
+We take a parametric approach in this work, and so we highlight challenges that are especially relevant to modeling the data generating process.
+Throughout, we consider a single perturbation-gene pair.
+First, the ``treatment'' variable -- i.e., the presence or absence of a perturbation -- is unobservable.
+Perturbed cells transcribe molecules called  \textit{guide RNAs} (or \textit{gRNAs}) that serve as indirect signals of perturbation presence.
+We can leverage the gRNA data to infer whether a given cell has been perturbed.
+Using these imputed perturbation assignments, we can estimate the effect of the perturbation on gene expression (Figure \ref{analysis_challenges}b).
+Second, ``technical factors'' -- sources of variation that are experimental rather than biological in origin -- impact both gene expression and gRNA expression (Figure \ref{analysis_challenges}b).
+We must account for these technical factors in constructing models for gene and gRNA expression.
+Third, the gene and gRNA data are sparse, discrete counts.
+Therefore, classical statistical approaches that assume Gaussianity or homoscedasticity are inapplicable.
+Finally, and most subtly, sequenced gRNAs sometimes are mapped to unperturbed cells.
+This phenomenon, which we call the ``background read'' phenomenon, results from errors in the sequencing process \cite{Replogle2020}.
+The marginal distribution of the gRNA count data is best conceptualized as a mixture model (Figure \ref{analysis_challenges}c; Gaussian distributions used for illustration purposes only).
+Unperturbed and perturbed cells both exhibit nonzero gRNA count distributions, but the gRNA count distribution for perturbed cells is greater.
+Figure \ref{analysis_challenges}d shows example data for four (of possibly tens or hundreds of thousands of) cells.
+The analysis objective is to leverage the observed data -- i.e., the gene expressions and gRNA counts -- to estimate the effect of the (latent) perturbation on gene expression, accounting for the technical factors.
 
-% Discuss bardodes and gRNAs in the subsequent paragraph.
-% \textbf{P3}: Single-cell CRISPR screens pose several core analysis challenges.
-\blue{Describe the analysis challenges here: (i) unobserved perturbation; (ii) existence of background reads; (iii) highly discrete count data; (iv) nuisance variables}.
-
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=1\linewidth]{../../figures/analysis_challenges/plot.pdf}
-	\caption{\textbf{Experimental design and analysis challenges}: \textbf{a,} Experimental design.
-For a given perturbation (e.g., the perturbation represented in yellow), we partition the cells into two groups: those that received the perturbation, and those that did not receive the perturbation.
-For a given gene, we conduct a differential expression analysis across the two groups of cells, yielding an estimate of the impact of the given perturbation on the given gene.
-\textbf{b,} DAG representing the variables in the analysis.
-The perturbation (unobserved) affects both gene expression and gRNA expression; technical factors (e.g., batch, sequencing depth, etc.) act as nuisance variables.
-The target of inference is the effect of the perturbation on gene expression (denoted with question mark).
-\textbf{c,} Schematic illustrating ``background reads.'' The gRNA modality has a nonzero, ``background read'' distribution even in the absence of a perturbation, complicating the assignment of perturbations to cells.
-\textbf{d}, Example data for a given perturbation-gene pair.
-Notice that (i) the perturbations are unobserved, and (ii) the gene and gRNA expression data take the form of discrete counts.}
-	\label{analysis_challenges}
-\end{figure}
+In this work we analyze two large-scale, high MOI, single-cell CRISPRi screen datasets published by Gasperini et al.\ and Xie et al.
+Gasperini (resp., Xie) targeted approximately 6,000 (resp., 500) candidate enhancers in a population of approximately 200,000 (resp., 100,000) cells.
+Gasperini additionally designed 381 positive control, gene-targeting perturbations and 50 non-targeting, negative control perturbations to assess method sensitivity and specificity.
 
@@
 
-In this section we study thresholded regression from empirical and theoretical perspectives, highlighting several limitations of the method.
-Let $n \in \N$ be the number of cells assayed in the experiment (typically, $n \approx 100,000 - 250,000$).
+The method that Gasperini et al.\ used in their original analysis is the so-called ``thresholding method,'' or ``thresholded regression.'' In this section we study the thresholding method from empirical and theoretical perspectives, uncovering several basic limitations of the approach.
+Let $n \in \N$ be the number of cells assayed in the experiment.
 Consider a single perturbation and a single gene.
-For cell $i \in \{1, \dots, n\}$, let $m_i \in \N$ be the number of gene transcripts sequenced in the cell; let $g_i \in \N$ be the number of gRNA transcripts sequenced in the cell; let $l^m_i \in \N$ be the number of gene transcripts sequenced across \textit{all} genes (the library size) in the cell; and finally, let $z_i \in \R^{d-1}$ be the cell-specific vector of technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.).
+For cell $i \in \{1, \dots, n\}$, let $m_i \in \N$ be the number of gene transcripts sequenced; let $g_i \in \N$ be the number of gRNA transcripts sequenced; let $l^m_i \in \N$ be the number of gene transcripts sequenced across \textit{all} genes (the library size); and finally, let $z_i \in \R^{d-1}$ be the cell-specific technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.) The letters ``m,'' ``g'', and ``l'' stand for ``mRNA,'' ``gRNA,'' and ``library,'' respectively.
 The thresholding method is defined as follows:
@@
 The biological interpretation for the target of inference $\beta^m_1$ as follows:  $\beta^m_1$ is the log-transformed fold change in gene expression in response to the perturbation, removing the effect of the technical factors.
-Fold change (obtained by exponentiating $\beta^m_1$) is the ratio of the mean expression in cells that were perturbed to the mean expression in cells that were not perturbed.
+Fold change (obtained by exponentiating $\beta^m_1$) is the ratio of the mean gene expression in cells that were perturbed to the mean gene expression in cells that were not perturbed.
 $\exp(\beta^m_1) = 1$ indicates no change in mean expression, while $\exp(\beta^m_1) > 1$ and $\exp(\beta^m_1) < 1$ indicate an increase and decrease in expression in response to the perturbation, respectively (accounting for technical factors).
 
-\subsection{Empirical analysis}
+\subsection{Empirical challenges of thresholding method}\label{sec:thresholding_empirical}
 
@@
 
-We reasoned that thresholded regression underestimated effect sizes on the positive control pairs, especially for threshold = 1 (an example of \textit{attenuation bias}).
-For a given perturbation, the vast majority ($>99\%$) of cells are unperturbed.
+We reasoned that thresholded regression systematically underestimated effect sizes on the positive control pairs, especially for small thresholds (an example of \textit{attenuation bias}).
+For a given perturbation, the vast majority ($>98\%$) of cells are unperturbed.
 This imbalance leads to an asymmetry: misclassifying \textit{unperturbed} cells as \textit{perturbed} is intuitively ``worse'' than misclassifying \textit{perturbed} cells as \textit{unperturbed}.
-Misclassified \textit{unperturbed} cells contaminate the set of truly perturbed cells, leading to attenuation bias; by contrast, misclassified \textit{perturbed} cells are swamped in number and ``neutralized'' by the truly \textit{unperturbed} cells.
+Misclassified unperturbed cells contaminate the set of truly perturbed cells, leading to attenuation bias; by contrast, misclassified perturbed cells are swamped in number and ``neutralized'' by the truly unperturbed cells.
 Setting the threshold to a large number reduces the unperturbed-to-perturbed misclassification rate, decreasing bias.
 
-We hypothesized, however, that the reduction in bias conferred by selecting a large threshold incurs the cost of increasing the variance of the estimator.
+We hypothesized, however, that the reduction in bias conferred by selecting a large threshold comes with the cost of increasing the variance of the estimator.
 To investigate, we compared $p$-values and confidence intervals produced by threshold = 5 and threshold = 20 for the target of inference $\beta^m_1$.
 We found that threshold = 5 yielded smaller (i.e., more significant) $p$-values and narrower confidence intervals than did threshold = 20 (Figure \ref{thresholding_empirical}c-d).
-We concluded that thresholded regression has a bias-variance tradeoff: as the threshold increases, bias decreases and variance increases.
+We concluded that the threshold controls a bias-variance tradeoff: as the threshold increases, bias decreases and variance increases.
 
@@
 
-\subsection{Theoretical analysis}\label{sec:thresholding_theory}
+\subsection{Theoretical challenges of thresholding method}\label{sec:thresholding_theory}
 
-We study the thresholding method from a theoretical perspective in an idealized Gaussian setting. Suppose that we observe data $(g_1, m_1), \dots, (g_n, m_n)$ from the following model:
+Next, we study the thresholding method from a theoretical perspective, recovering the attenuation bias and bias-variance tradeoff effects established in the previous section, as well as several other interesting phenomena.
+We work in an idealized Gaussian setting.
+Suppose that we observe gRNA and gene expression data $\{(g_1, m_1), \dots, (g_n, m_n)\}$ on $n \in \N$ cells from the following model:
 
@@
 \end{itemize}
-Let $b: \R^4 \to \R$ be the asymptotic \textit{relative} bias of $\hat{\beta}^m_1$:
+Let $b: \R^4 \to \R$ be the asymptotic relative bias of $\hat{\beta}^m_1$:
 \begin{multline*}
@@
 \end{multline*}
-where $\lim_{a.s.}$ denotes a.s.\ convergence. The asymptotic relative bias vanishes when the attenuation function equals $1$. \textcolor{red}{(HOW TO MAKE THE ABOVE RIGOROUS?)}
+where $\lim_{a.s.}$ denotes a.s.\ convergence. The asymptotic relative bias vanishes when the attenuation function equals $1$. \textcolor{red}{HOW TO MAKE THE ABOVE RIGOROUS?}
 
@@
 To investigate the basic question of ``What is a good threshold selection strategy?'', we study the relationship between the asymptotic relative bias $b$ of $\hat{\beta}^m_1$ and the selected threshold $c$.
-For simplicity, we set the perturbation probability $\pi$ to $1/2$.
+For simplicity, we begin by setting the perturbation probability $\pi$ to $1/2$.
 Let $c_\textrm{bayes} \in \R$ be the Bayes-optimal decision boundary for classifying cells as perturbed or unperturbed, i.e.
@@
 We sometimes refer to ``asymptotic relative bias'' using the shortened term ``bias'' for succinctness.
+\textcolor{red}{SHOULD WE MOVE PROP 4-6 	TO THE APPENDIX?
+THESE PROPS ARE LESS IMPORTANT THAN THE OTHERS AND CONTRIBUTE SOMEWHAT LESS TO THE NARRATIVE.}
 
@@
 In other words, if the perturbation probability is $\pi$, and if we set the threshold to a large number, then the asymptotic relative bias is $\pi$ (Figure \ref{thresholding_theoretical}b).
-We can understand this result intuitively by considering an extreme example: when $\pi$ is very small (e.g., $\pi = 0.01$), almost all cells are unperturbed.
-Therefore, in selecting a large threshold, we correctly classify nearly all unperturbed cells as unperturbed; on the other hand, the \textit{perturbed} cells that we misclassify as \textit{unperturbed} are swamped in number by the truly unperturbed cells, resulting in a small bias.
+We can understand this result intuitively by considering an extreme example: when $\pi$ is very small (e.g., $\pi = 0.01$), most cells are unperturbed.
+Therefore, as discussed in Section \ref{sec:thresholding_empirical}, selecting a large threshold minimizes the unperturbed-to-perturbed misclassification rate, reducing bias.
+% Therefore, in selecting a large threshold, we correctly classify nearly all unperturbed cells as unperturbed; on the other hand, the \textit{perturbed} cells that we misclassify as \textit{unperturbed} are swamped in number by the truly unperturbed cells, resulting in a small bias.
 
@@
 
-Finally, to shed light on limitations of the large threshold selection strategy, we derive an exact bias-variance decomposition for the thresholding estimator.
+Finally, to shed light on the costs of selecting a large threshold, we derive an exact bias-variance decomposition for the thresholding estimator.
 We consider a slightly simpler, no-intercept version of (\ref{theoretical_model}) for this purpose:
@@
 
-% These difficulties motivate our core research question: does bypassing thresholding altogether lead to better estimation and inference in single-cell CRISPR screens?
-To answer this question, we propose GLM-EIV, method for single-cell CRISPR screen analysis that models gRNA counts directly.
+\subsection{Thresholding method summary}
+
+Empirical and theoretical analyses reveal that the thresholding method poses several challenges: the threshold is a tuning parameter that substantially impacts the results; strict attenuation bias obtains uniformly over the parameter space and for all choices of the threshold; and there does not exist and obvious threshold selection strategy due to (i) the unimodality of the empirical gRNA count distributions and (ii) the existence of a bias-variance tradeoff mediated by the threshold.
+These difficulties motivate our core research question: \textit{Does modeling the gRNA count distribution directly, thereby circumventing the need to threshold altogether, lead to simpler, more accurate estimation and inference in single-cell CRISPR screen analysis?} To answer this question, we introduce a generalization of the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.
 

### 2021-10-29 text/outline/draft.tex
update abstract

--- 2021-10-29 text/outline/outline.tex
@@
 CRISPR genome engineering and single-cell sequencing have transformed biological discovery.
-Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases.
-In this work we study single-cell CRISPR screens from a statistical perspective.
-First, we demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter.
+Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression.
+Single-cell CRISPR screens promise to illuminate the regulatory wiring underlying diseases but present substantial statistical challenges.
+We demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter.
 We recover these phenomena in precise theoretical terms in an idealized Gaussian setting.
@@
 The perturbations assort into the cells randomly, with each cell receiving on average 10-40 distinct perturbations.
-Conversely, a given perturbation enters fewer than 0.1-2\% of cells.
+Conversely, a given perturbation enters about 0.1-2\% of cells.
 
@@
 For a given perturbation (e.g., the perturbation represented in blue), the scientist partitions the cells into two groups: those that received the perturbation (top) and those that did not (bottom).
-Next, for a given gene, the scientist runs a differential expression analysis across the two groups of cells, producing an estimate for the magnitude of gene expression change in response to the perturbation.
+Next, for a given gene, the scientist runs a differential expression analysis across the two groups of cells, producing an estimate for the extent to which the gene's expression has changed in response to the perturbation.
 If the estimated change in expression is large, the scientist can conclude that the enhancer \textit{targeted} by the perturbation exerts a strong regulatory effect on the gene.
@@
 
-\subsection{Analysis challenges and data}
+\subsection{Analysis challenges}
 High MOI single-cell CRISPR screens present several statistical challenges.

### 2021-11-01 text/outline/draft.tex
update abstract
clarify that both Gasp and Xie used thresholding

@@
 Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression.
-Single-cell CRISPR screens promise to illuminate the regulatory wiring underlying diseases but present substantial statistical challenges.
+Single-cell CRISPR screens promise to shed light on the genetic architecture underlying diseases but present substantial statistical challenges.
+%  In this work we study single-cell CRISPR screens from a statistical perspective.
 We demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter.
@@
 \end{abstract}
-
 \tableofcontents
-
+\noindent \blue{All edits in blue.}
 \section{Introduction}
@@
 Despite their promise, single-cell CRISPR screens present substantial statistical challenges.
-A major difficulty is that CRISPR perturbations are unobservable and assigned stochastically to cells.
+A major difficulty is that CRISPR perturbations \blue{are assigned stochastically to cells and cannot be observed directly}.
 As a consequence, one cannot know with certainty which cells were perturbed.
@@
 Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian model.
-Second, we introduce a new method for estimation and inference in single-cell CRISPR screens that accounts for the measurement error inherent in the experiment.
-The method, called \textit{GLM-EIV} (generalized linear model with errors-in-variables), implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic.
-Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.
+\blue{Second, we introduce a new method, GLM-EIV (generalized linear model with errors-in-variables), for single-cell CRISPR screen analysis.
+GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.
+GLM-EIV implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic.} Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.
 
@@
 In this work we tackle a set of analysis challenges that are complimentary to the challenges addressed by SCEPTRE.
-Most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, moving beyond hypothesis testing.
+Most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, \blue{a statistical objective that is unattainable within the nonparameteric hypothesis testing framework of SCEPTRE.}
 
@@
 The most promising biotechnology for solving this challenge are high MOI single-cell CRISPR screens.
-High MOI single-cell CRISPR screens combine CRISPR interference (CRISPRi) -- a version of CRISPR that \textit{represses} a targeted region of the genome, leaving the primary DNA sequence intact  -- with single-cell sequencing.
+High MOI single-cell CRISPR screens combine CRISPR interference (CRISPRi) -- a version of CRISPR that represses a targeted region of the genome -- with single-cell sequencing.
 The experimental protocol is as follows.
 First, the scientist develops a library of several hundred to several thousand CRISPRi perturbations, each designed to target a candidate enhancer for repression.
-The scientist then cultures tens or hundreds of thousands of cells and delivers the CRISPRi perturbations to the cells.
+The scientist then cultures tens or hundreds of thousands of cells and delivers the CRISPRi perturbations to these cells.
 The perturbations assort into the cells randomly, with each cell receiving on average 10-40 distinct perturbations.
@@
 For a given perturbation (e.g., the perturbation represented in blue), the scientist partitions the cells into two groups: those that received the perturbation (top) and those that did not (bottom).
-Next, for a given gene, the scientist runs a differential expression analysis across the two groups of cells, producing an estimate for the extent to which the gene's expression has changed in response to the perturbation.
+Next, for a given gene, the scientist runs a differential expression analysis across the two groups of cells, producing an estimate for the magnitude of the gene expression change in response to the perturbation.
 If the estimated change in expression is large, the scientist can conclude that the enhancer \textit{targeted} by the perturbation exerts a strong regulatory effect on the gene.
 This procedure is repeated for a large set of preselected perturbation-gene pairs.
-% If the differential expression analyses produce $p$-values, the $p$-values are subjected to a multiple hypothesis testing correction procedure.
+\blue{The enhancer-by-enhancer approach is valid because the perturbations assort into cells approximately independently of one another.}
 
 \subsection{Analysis challenges}
-High MOI single-cell CRISPR screens present several statistical challenges.
-We take a parametric approach in this work, and so we highlight challenges that are especially relevant to modeling the data generating process.
-Throughout, we consider a single perturbation-gene pair.
-First, the ``treatment'' variable -- i.e., the presence or absence of a perturbation -- is unobservable.
-Perturbed cells transcribe molecules called  \textit{guide RNAs} (or \textit{gRNAs}) that serve as indirect signals of perturbation presence.
-We can leverage the gRNA data to infer whether a given cell has been perturbed.
-Using these imputed perturbation assignments, we can estimate the effect of the perturbation on gene expression (Figure \ref{analysis_challenges}b).
-Second, ``technical factors'' -- sources of variation that are experimental rather than biological in origin -- impact both gene expression and gRNA expression (Figure \ref{analysis_challenges}b).
-We must account for these technical factors in constructing models for gene and gRNA expression.
-Third, the gene and gRNA data are sparse, discrete counts.
+\blue{High MOI single-cell CRISPR screens present several statistical challenges, four of which we highlight here.} Throughout, we consider a single perturbation-gene pair.
+First, the ``treatment'' variable -- i.e., the presence or absence of a perturbation -- cannot be directly observed.
+Instead, perturbed cells transcribe molecules called  \textit{guide RNAs} (or \textit{gRNAs}) that serve as indirect proxies of perturbation presence.
+We must leverage these gRNAs to impute (explicitly or implicitly) perturbation assignments onto the cells (Figure \ref{analysis_challenges}b).
+\blue{Second, ``technical factors'' -- sources of variation that are experimental rather than biological in origin -- impact both gene expression and gRNA expression (Figure \ref{analysis_challenges}b).
+Technical factors act as confounders in the measurement process and therefore must be accounted for in differential expression models.} Third, the gene and gRNA data are sparse, discrete counts.
 Therefore, classical statistical approaches that assume Gaussianity or homoscedasticity are inapplicable.
-Finally, and most subtly, sequenced gRNAs sometimes are mapped to unperturbed cells.
-This phenomenon, which we call the ``background read'' phenomenon, results from errors in the sequencing process \cite{Replogle2020}.
-The marginal distribution of the gRNA count data is best conceptualized as a mixture model (Figure \ref{analysis_challenges}c; Gaussian distributions used for illustration purposes only).
-Unperturbed and perturbed cells both exhibit nonzero gRNA count distributions, but the gRNA count distribution for perturbed cells is greater.
-Figure \ref{analysis_challenges}d shows example data for four (of possibly tens or hundreds of thousands of) cells.
-The analysis objective is to leverage the observed data -- i.e., the gene expressions and gRNA counts -- to estimate the effect of the (latent) perturbation on gene expression, accounting for the technical factors.
+Finally, and most subtly, sequenced gRNAs sometimes are mapped to cells that have not received a perturbation.
+This phenomenon, which we call the ``background read'' phenomenon, results from errors in the sequencing and alignment processes \cite{Replogle2020}.
+The marginal distribution of the gRNA counts is best conceptualized as a mixture model (Figure \ref{analysis_challenges}c; Gaussian distributions used for illustration purposes only).
+\blue{Unperturbed and perturbed cells both exhibit nonzero gRNA count distributions, but this distribution overall is greater for perturbed cells.} Figure \ref{analysis_challenges}d shows example data on four (of possibly tens or hundreds of thousands of) cells.
+\blue{The analysis objective is to leverage the gene expressions and gRNA counts to estimate the effect of the (latent) perturbation on gene expression, accounting for the technical factors.}
 
-In this work we analyze two large-scale, high MOI, single-cell CRISPRi screen datasets published by Gasperini et al.\ and Xie et al.
+In this work we analyze two large-scale, high MOI, single-cell CRISPR screen datasets published by Gasperini et al.\ and Xie et al.
 Gasperini (resp., Xie) targeted approximately 6,000 (resp., 500) candidate enhancers in a population of approximately 200,000 (resp., 100,000) cells.
@@
 \section{Thresholding method}
+\blue{In their original analysis of the data, Gasperini and Xie imputed perturbation identities onto the cells by thresholding the gRNA counts.
+In this section we study the thresholding method from empirical and theoretical perspectives, uncovering several limitations of the approach.
+We investigate the specific version of the thresholding method used by Gasperini, which involved a thresholding step and a follow-up negative binomial regression step.
+(Xie implemented a slightly different and, in our view, less principled version of thresholding, which we do not review here.)}
 
-The method that Gasperini et al.\ used in their original analysis is the so-called ``thresholding method,'' or ``thresholded regression.'' In this section we study the thresholding method from empirical and theoretical perspectives, uncovering several basic limitations of the approach.
 Let $n \in \N$ be the number of cells assayed in the experiment.
@@
 \begin{itemize}
-\item[1.] For a given threshold $c \in \N$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be $$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i \geq c, \\ \hat{p}_i = 1 \textrm{ if } g_i < c.
+\item[1.] For a given threshold $c \in \N$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be $$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i < c, \\ \hat{p}_i = 1 \textrm{ if } g_i \geq c.
 \end{cases}$$
 \item[2.] Assume that $m_i$ is related to $\hat{p}_i, l_i,$ and $z_i$ through the following GLM:
-\begin{equation*}
-\begin{cases}
-m_i | (\hat{p}_i, z_i, l_i) \sim \textrm{NB}_\theta(\mu_i), \\
+$$m_i | (\hat{p}_i, z_i, l_i) \sim \textrm{NB}_\theta(\mu_i),$$
+\begin{equation}\label{thresh_glm}
 \log(\mu_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left( l_i^m\right),
-\end{cases}
-\end{equation*}
+\end{equation}
 where (i) $NB_\theta(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta$; (ii) $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma \in  \R^{d-1}$ are unknown parameters; and (iii) $\log(l_i^m)$ is an offset term.
@@
 
-The biological interpretation for the target of inference $\beta^m_1$ as follows:  $\beta^m_1$ is the log-transformed fold change in gene expression in response to the perturbation, removing the effect of the technical factors.
+% Setting $\gamma_m^T$ to zero and exponentiating both sides of (\ref{thresh_glm}), we obtain
+% $$\mu_i = \left(e^{\beta^m_0 + \beta^m_1\hat{p}_i}\right) l^m_i.$$ The mean expression of unperturbed cells $\mu_i(0)$ is
+% $$\mu_i(0) = e^{\beta^m_0} l_i^m,$$ and the mean expression in perturbed cells is
+% $$ \mu_i(1) = e^{\beta^m_1} e^{\beta^m_0} l^m_i.$$ Therefore,
+% $$ \beta^m_1 = \log\left(\frac{\mu_i(1)}{\mu_i(0)}\right).$$
+% We use library size $l_i^m$
+
+The biological interpretation for the target of inference $\beta^m_1$ as follows:  $\beta^m_1$ is the log-transformed fold change in gene expression in response to the perturbation, \blue{controlling for} the technical factors.
 Fold change (obtained by exponentiating $\beta^m_1$) is the ratio of the mean gene expression in cells that were perturbed to the mean gene expression in cells that were not perturbed.
@@
 
-We hypothesized, however, that the reduction in bias conferred by selecting a large threshold comes with the cost of increasing the variance of the estimator.
+We hypothesized, however, that the reduction in bias conferred by selecting a large threshold comes at the cost of increasing the variance of the estimator.
 To investigate, we compared $p$-values and confidence intervals produced by threshold = 5 and threshold = 20 for the target of inference $\beta^m_1$.
 We found that threshold = 5 yielded smaller (i.e., more significant) $p$-values and narrower confidence intervals than did threshold = 20 (Figure \ref{thresholding_empirical}c-d).
-We concluded that the threshold controls a bias-variance tradeoff: as the threshold increases, bias decreases and variance increases.
+We concluded that the threshold controls a bias-variance tradeoff: as the threshold increases, bias of the estimator decreases and variance increases.
 
@@
 The latter threshold yields more confident estimates.
-\textbf{e-f}, Empirical distribution of randomly-selected gRNA from Gasperini (e) and Xie (f) data.
+\textbf{e-f}, Empirical distribution of randomly-selected gRNA from Gasperini (e) and Xie (f) data (0 counts not shown).
 The gRNA data do not appear to imply an obvious threshold selection strategy.}
@@
 
-Next, we study the thresholding method from a theoretical perspective, recovering the attenuation bias and bias-variance tradeoff effects established in the previous section, as well as several other interesting phenomena.
+Next, we study the thresholding method from a theoretical perspective, recovering in precise mathematical terms the attenuation bias and bias-variance tradeoff effects uncovered on real data, as well as several other interesting phenomena.
 We work in an idealized Gaussian setting.
@@
 Empirical and theoretical analyses reveal that the thresholding method poses several challenges: the threshold is a tuning parameter that substantially impacts the results; strict attenuation bias obtains uniformly over the parameter space and for all choices of the threshold; and there does not exist and obvious threshold selection strategy due to (i) the unimodality of the empirical gRNA count distributions and (ii) the existence of a bias-variance tradeoff mediated by the threshold.
-These difficulties motivate our core research question: \textit{Does modeling the gRNA count distribution directly, thereby circumventing the need to threshold altogether, lead to simpler, more accurate estimation and inference in single-cell CRISPR screen analysis?} To answer this question, we introduce a generalization of the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.
+These difficulties motivate our core research question: \textit{Does modeling the gRNA count distribution directly, thereby circumventing the need to threshold altogether, lead to simpler, more accurate estimation and inference in single-cell CRISPR screen analysis?} To answer this question, we generalize the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.
 

### 2021-11-01 R_scripts/plotting/analysis_challenges.R

@@ -33,7 +33,12 @@ p1 <- ggplot(data = to_plot, mapping = aes(x = x, y = y, col = Status)) +
   scale_color_manual(values = c(my_cols[2], "black")) + geom_line(lwd = 1.1) +
   theme(legend.position = c(0.77, 0.85), legend.title = element_blank(),
         panel.border = element_blank(), panel.grid.major = element_blank(),
-        panel.grid.minor = element_blank(), axis.line.y = element_line(colour = "black"))
+        panel.grid.minor = element_blank(),
+        axis.line.y = element_line(colour = "black"),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(),
+        axis.text.y=element_blank(),
+        axis.ticks.y=element_blank())
 
 # set up cowplot
 p_out <- plot_grid(NULL, p1, NULL, NULL, labels = c("a", "c", "b", "d"), nrow = 2, rel_heights = c(0.5, 0.5))

### 2021-11-01 R_scripts/plotting/thresholding_empirical.R

@@ -70,18 +70,18 @@ p5 <- ggplot(data = xie_gRNA %>% tibble::tibble(count = .) %>%
                dplyr::filter(count >= 1, count <= 40),
              mapping = aes(x = count)) +
   geom_histogram(binwidth = 2, col = "black", fill = my_cols[4], alpha = 0.7) +
-  scale_y_continuous(trans='log10', expand = c(0, NA)) + xlab("gRNA count") + ylab("") +
+  scale_y_continuous(trans='log10', expand = c(0, NA)) + xlab("gRNA count (Xie)") + ylab("") +
   theme_bw(base_size = 10) + my_theme
   
 p6 <- ggplot(data = gasp_gRNA %>% tibble::tibble(count = .) %>%
                dplyr::filter(count >= 1, count <= 40),
              mapping = aes(x = count)) +
   geom_histogram(binwidth = 2, col = "black", fill = my_cols[4], alpha = 0.7) +
-  scale_y_continuous(trans='log10', expand = c(0, NA)) + xlab("gRNA count") +
+  scale_y_continuous(trans='log10', expand = c(0, NA)) + xlab("gRNA count (Gasperini)") +
   ylab("") + theme_bw(base_size = 10) + my_theme
   # geom_vline(xintercept = xie_thresh, col = my_cols[1], lwd = 0.8)
 
 # combine plots
-p_out <- plot_grid(p1, p2, p3, p4, p5, p6, labels = c("a", "b", "c", "d", "e", "f"), ncol = 2, rel_heights = c(1,1,0.8), align = "vh")
+p_out <- plot_grid(p1, p2, p3, p4, p6, p5, labels = c("a", "b", "c", "d", "e", "f"), ncol = 2, rel_heights = c(1,1,0.8), align = "vh")
 
 ggsave(filename = paste0(fig_dir, "/plot.jpeg"), plot = p_out, device = "jpeg", scale = 1, width = 6, height = 6)

### 2021-11-01 text/outline/draft.tex
finish implementing gene's edits (in all sections but theory)

@@
 In this work we tackle a set of analysis challenges that are complimentary to the challenges addressed by SCEPTRE.
-Most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, \blue{a statistical objective that is unattainable within the nonparameteric hypothesis testing framework of SCEPTRE.}
+Most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, \blue{a statistical objective unattainable within the nonparameteric hypothesis testing framework of SCEPTRE.}
 
@@
 \section{Thresholding method}
-\blue{In their original analysis of the data, Gasperini and Xie imputed perturbation identities onto the cells by thresholding the gRNA counts.
-In this section we study the thresholding method from empirical and theoretical perspectives, uncovering several limitations of the approach.
-We investigate the specific version of the thresholding method used by Gasperini, which involved a thresholding step and a follow-up negative binomial regression step.
-(Xie implemented a slightly different and, in our view, less principled version of thresholding, which we do not review here.)}
+\blue{In this section we study the thresholding method from empirical and theoretical perspectives, uncovering several limitations of the approach.
+Gasperini and Xie both imputed perturbation identities onto cells via thresholding, but they carried out subsequent differential expression analyses in different ways: Gasperini ran negative binomial regressions, while Xie conducted nonparametric tests of independence.
+These two approaches pose similar challenges, but we investigate the regression-based version of the thresholding method here, as it relates most closely to GLM-EIV.}
 
@@
 \end{itemize}
-
-% Setting $\gamma_m^T$ to zero and exponentiating both sides of (\ref{thresh_glm}), we obtain
-% $$\mu_i = \left(e^{\beta^m_0 + \beta^m_1\hat{p}_i}\right) l^m_i.$$ The mean expression of unperturbed cells $\mu_i(0)$ is
-% $$\mu_i(0) = e^{\beta^m_0} l_i^m,$$ and the mean expression in perturbed cells is
-% $$ \mu_i(1) = e^{\beta^m_1} e^{\beta^m_0} l^m_i.$$ Therefore,
-% $$ \beta^m_1 = \log\left(\frac{\mu_i(1)}{\mu_i(0)}\right).$$
-% We use library size $l_i^m$
+\blue{We include the library size $l^m_i$ as an offset term in (\ref{thresh_glm}) so that $\beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i$ can be interpreted as a relative expression: exponentiating both sides of (\ref{thresh_glm}), we obtain
+$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) l_i^m.$$ We see that $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of transcripts in the cell produced by the gene under consideration.}
 

### 2021-11-06 text/outline/draft.tex
finish draft of 4.1 - 4.2

@@
 \usepackage{amsfonts}
+\usepackage{comment}
 \usepackage{amsmath}
 \usepackage{graphicx}
+\usepackage{algorithm}
+\usepackage[toc,page]{appendix}
+\usepackage{algpseudocode}
 \setcounter{tocdepth}{2}
@@
 Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression.
-Single-cell CRISPR screens promise to shed light on the genetic architecture underlying diseases but present substantial statistical challenges.
-%  In this work we study single-cell CRISPR screens from a statistical perspective.
+Single-cell CRISPR screens promise to illuminate regulatory networks underlying diseases but present substantial statistical challenges.
 We demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter.
@@
 \section{Thresholding method}
-\blue{In this section we study the thresholding method from empirical and theoretical perspectives, uncovering several limitations of the approach.
-Gasperini and Xie both imputed perturbation identities onto cells via thresholding, but they carried out subsequent differential expression analyses in different ways: Gasperini ran negative binomial regressions, while Xie conducted nonparametric tests of independence.
-These two approaches pose similar challenges, but we investigate the regression-based version of the thresholding method here, as it relates most closely to GLM-EIV.}
+\blue{In this section we study thresholded regression from empirical and theoretical perspectives, uncovering several limitations of the method.
+Gasperini and Xie both imputed perturbation identities onto the cells via thresholding, but they carried out the subsequent differential expression analysis in different ways: Gasperini used negative binomial regression, whereas Xie used nonparametric independence testing.
+These two strategies pose similar challenges, but we investigate Gasperini's variant of the thresholding method, as it relates most closely to GLM-EIV.}
 
-Let $n \in \N$ be the number of cells assayed in the experiment.
+Let $n \in \N$ be the number of cellxs assayed in the experiment.
 Consider a single perturbation and a single gene.
@@
 \end{cases}$$
-\item[2.] Assume that $m_i$ is related to $\hat{p}_i, l_i,$ and $z_i$ through the following GLM:
-$$m_i | (\hat{p}_i, z_i, l_i) \sim \textrm{NB}_\theta(\mu_i),$$
+\item[2.] Assume that $m_i$ is related to $\hat{p}_i, l^m_i,$ and $z_i$ through the following GLM:
+$$m_i | (\hat{p}_i, z_i, l^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i),$$
 \begin{equation}\label{thresh_glm}
@@
 \end{equation}
-where (i) $NB_\theta(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta$; (ii) $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma \in  \R^{d-1}$ are unknown parameters; and (iii) $\log(l_i^m)$ is an offset term.
+where (i) $NB_\theta(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta^m$; (ii) $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in  \R^{d-1}$ are unknown parameters; and (iii) $\log(l_i^m)$ is an offset term.
 Fit a GLM to obtain estimates of the parameters.
@@
 \blue{We include the library size $l^m_i$ as an offset term in (\ref{thresh_glm}) so that $\beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i$ can be interpreted as a relative expression: exponentiating both sides of (\ref{thresh_glm}), we obtain
-$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) l_i^m.$$ We see that $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of transcripts in the cell produced by the gene under consideration.}
+$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) l_i^m.$$ We see that $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration.}
 
@@
 \end{multline*}
-where $\lim_{a.s.}$ denotes a.s.\ convergence. The asymptotic relative bias vanishes when the attenuation function equals $1$. \textcolor{red}{HOW TO MAKE THE ABOVE RIGOROUS?}
+where $\lim_{a.s.}$ denotes a.s.\ convergence. The asymptotic relative bias vanishes when the attenuation function equals $1$.
 
@@
 
-\section{GLM-EIV}
+\section{Generalized linear model with errors in variables}
 
+In this section we introduce generalized linear model with errors-in-variables (GLM-EIV), derive estimation and inference procedures for the model, and propose several statistical accelerations to reduce the cost of fitting the model.
+
+\subsection{Model}
+\subsubsection*{Negative binomial model}
+Building on the work of several previous authors \cite{Townes2019,Svensson2020,Hafemeister2019}, Sarkar and Stephens \cite{Sarkar2021} proposed a simple strategy for modeling for single-cell gene expression data, which, in the framework of the negative binomial GLM, is equivalent to using the log-transformed library size as an offset term (as in (\ref{thresh_glm})).
+We generalize Sarkar and Stephens' approach to model \textit{both} gene and gRNA modalities.
+To this end, let the latent variable $p_i \in \{0,1\}$ indicate whether cell $i \in \{1, \dots, n\}$ was perturbed.
+We model the gene expression counts according to
+\begin{equation}\label{glmeiv_model_1}
+m_i |(p_i, z_i, l^m_i) \sim \textrm{NB}(\mu_i^m),
+\end{equation}
+\begin{equation}\label{glmeiv_model_2}
+\log(\mu^m_i) = \beta^m_0 + \beta^m_1 p_i + \gamma_m^T z_i + \log(l^m_i),
+\end{equation}
+where $\theta^m > 0$ is a known negative binomial size parameter, and $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in \R^{d - 2}$ are unknown constants.
+The model (\ref{glmeiv_model_1}) is identical to the thresholding model (\ref{thresh_glm}), but the imputed perturbation indicator $\hat{p}_i$ is replaced by the latent perturbation indicator $p_i$.
+Next, let $l^g_i \in \N$ be the number of gRNA transcripts sequenced across \textit{all} gRNAs in cell $i$ (i.e., the gRNA library size).
+The model for the gRNA counts is
+\begin{equation}\label{glmeiv_model_3}
+g_i | (p_i, z_i, l^g_i) \sim \textrm{NB}_{\theta^g}\left(\mu_i^g\right),
+\end{equation}
+\begin{equation}\label{glmeiv_model_4}
+\log(\mu_i^g) = \beta^g_0 + \beta^g_1p_i + \gamma^T_g z_i + \log(l^g_i),
+\end{equation}
+where, similar to above, $\theta^g > 0$ is a known negative binomial size parameter, and $\beta^g_0 \in \R, \beta^g_1 \in \R, \gamma_g \in \R^{d - 2}$ are unknown constants.
+We use a negative binomial GLM to model the gRNA counts because gRNA molecules are transcribed in the cell in the same way that gene transcripts are \cite{Datlinger2017,Hill2018}.
+Finally, we model the marginal perturbation probability as
+\begin{equation}\label{glmeiv_model_5}
+p_i \sim \textrm{Bern}(\pi),
+\end{equation} where $\pi \in (0,1/2]$. 
+ Together, (\ref{glmeiv_model_1}, \ref{glmeiv_model_2}, \ref{glmeiv_model_3}, \ref{glmeiv_model_4}, \ref{glmeiv_model_5}) define the standard GLM-EIV model.
+The terms $(\beta^m_0 + \beta^m_1p_i + \gamma^T_m z_i)$ and $(\beta^g_0 + \beta^g_1 p_i + \gamma^T_g z_i)$ can be interpreted as relative gene and gRNA expressions, similar to the analogous term in the thresholding model.
+Likewise, the target of inference $\beta^m_1$ is the log fold change in gene expression in response to the perturbation, accounting for technical factors.
+
+\subsubsection*{Full GLM-EIV model}
+To provide greater modeling flexibility, we generalize the GLM-EIV model to arbitrary exponential family response distributions and link functions.
+To increase notational compactness, let $\tilde{x_i} = [1, p_i, z_i]^T \in \R^d$ be the vector of covariates (including an intercept term) for the $i$th cell.
+(We use the tilde as a reminder that the vector is partially unobserved.) Let $\beta_m = [\beta^m_0, \beta^m_1, \gamma_m]^T \in \R^d$ and $\beta_g = [\beta^g_0, \beta^g_1, \gamma_g]^T \in \R^d$ be the unknown coefficient vectors corresponding to the gene and gRNA expression models, respectively.
+Finally, let $o^m_i$ and $o^g_i$ be the (possibly zero) offset terms for the gene and gRNA models; in practice, we typically set $o^m_i$ and $o^g_i$ to $\log(l^m_i)$ and $\log(l^g_i)$, respectively.
+
+We use a GLM approach to model the gene and gRNA expressions.
+Considering first the gene expression model, let the $i$th linear component $l^m_i$ of the model be $$l^m_i = \langle \tilde{x}_i, \beta_m \rangle + o^m_i.$$ Let the mean $\mu^m_i$ of the $i$th observation be
+$$r_m(\mu^m_i) = l^m_i,$$ where $r_m:\R \to \R$ is a strictly increasing, differentiable link function.
+Let $\psi_m: \R \to \R$ be the differentiable, cumulant-generating function of the selected exponential family distribution.
+We can express the canonical parameter $\eta^m_i$ in terms of $\psi_m$ and $r_m$ by
+$$\eta^m_i = \left([\psi'_m]^{-1} \circ r^{-1}_m\right)(l_i^m) := h_m(l_i^m).$$ Finally, let $c_m: \R \to \R$ be the carrying density of the selected exponential family distribution.
+The density $f_m$ of $m_i$ conditional on the the canonical parameter $\eta_i$ is
+$$f_m(m_i; \eta^m_i) = \exp\left\{ m_i \eta^m_i - \psi_m(\eta^m_i) + c_m(m_i) \right\}.$$ The function $c_m$ does not appear in the log-likelihood of $m_i$; therefore, the only functions relevant to inference are $\psi_m$ and $r_m$.
+
+Let the terms $l^g_i, o^g_i, \mu^g_i, \eta^g_i, \psi_g, r_g, h_g$ and $c_g$ be defined in an analogous way for the gRNA model:
+$$
+\begin{cases}
+l^g_i = \langle \tilde{x}_i, \beta_g \rangle + o^g_i,\\
+r_g(\mu^g_i) = l^g_i,\\
+\eta^g_i = \left([\psi'_g]^{-1} \circ r^{-1}_g\right)(l_i^g) := h_g(l_i^g).
+\end{cases}
+$$
+The density $f_g$ of $g_i$ given the canonical parameter is
+$$f_g(m_i; \eta^g_i) = \exp\left\{g_i \eta^g_i - \psi_g(\eta^g_i) + c_g(g_i)\right\}.$$
+Finally, the unobserved variable $p_i$ is assumed to follow a Bernoulli distribution with mean $\pi \in (0, 1/2]$. Its marginal density $f_p$ is given by
+$$f_p(p_i) = \pi^{p_i}(1-\pi)^{1 - p_i}.$$
+The unknown parameters in the model are
+$\theta = [\beta_m, \beta_g, \pi]^{T}  \in \R^{2d + 1}.$
+
+\subsubsection*{Notation}
+We briefly introduce notation that we will use throughout.
+For $k \in \{0,1\}$, let $\tilde{x}_i(k) := [1, k, z_i]^T$ denote the value of $\tilde{x}_i$ that results from setting $p_i$ to $k$.
+Next, let $l^m_i(k) := \langle \tilde{x}_i(k), \beta_m \rangle + o^m_i$ and $\eta^m_i(k) := h_m(l^m_i(k))$, and let the corresponding gRNA quantities $l^g_i(k)$ and $\eta_i^g(k)$ be defined analogously.
+Also, let the design matrix $\tilde{X} \in \R^{n \times d}$ be defined by
+$$ \tilde{X} := \begin{bmatrix}
+\tilde{x}_1^T \\ \tilde{x}_2^T \\ \vdots \\ \tilde{x}_n^T
+\end{bmatrix} = 
+\begin{bmatrix}
+1 & p_1 & z_1 \\
+1 & p_2 & z_2 \\ 
+\vdots & \vdots & \vdots \\
+1 & p_n & z_n
+\end{bmatrix}.$$ For $k \in \{0,1\}$, let $\tilde{X}(k) \in \R^{n \times d}$ be the matrix that results from setting $p_i$ to $k$ for all $i \in \{1, \dots, n\}$ in $\tilde{X}$.
+
+Next, let $[m,m]^T \in \R^{2n}$ denote the vector that results from concatenating $m := [m_1, \dots, m_n]^T \in \R^n$ to itself, i.e.
+$$ [m,m]^T := [\underbrace{m_1, m_2 \dots, m_{n-1}, m_n}_\textrm{first copy of $m$}, \underbrace{m_1, m_2, \dots, m_{n-1}, m_n}_\textrm{second copy of $m$}].$$ Define $[g,g]^T$, $[o^g,o^g]^T$, and $[o^m,o^m]^T$ similarly.
+Finally, let $\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix}$ denote the $\R^{2n \times d}$ matrix that results from vertically concatenating $\tilde{X}(0)$ and $\tilde{X}(1)$.
+
+\subsubsection*{Log likelihood and model properties}
+We derive the log-likelihood of the GLM-EIV model.
+We conduct estimation and inference \textit{conditional} on the library sizes and technical factors $l^m_i, l^g_i,$ and $z_i$; therefore, we treat these quantities as fixed constants.
+We assume that the gene expression $m_i$ and gRNA expression $g_i$ are \textit{conditionally independent} given the perturbation $p_i$.
+The joint density $f$ of $(m_i, g_i, p_i)$ given $\theta$ is
+\begin{equation}\label{full_density}
+f(m_i,g_i,p_i; \theta) = f_m(m_i | p_i) f_g(g_i | p_i) f_p(p_i) \\ = \pi^{p_i}(1-\pi)^{1-p_i} f_m(m_i; \eta^m_i) f_g(g_i; \eta^g).
+\end{equation}
+The log-likelihood is
+\begin{multline}\label{full_log_lik}
+\mathcal{L}(\theta; m_i, g_i, p_i) = \sum_{i=1}^n \log\left( \pi^{p_i}(1-\pi)^{1-p_i} \right) \\ + \sum_{i=1}^n \log\left( f_m(m_i; \eta^m_i)\right) + \sum_{i=1}^n \log\left( f_g(g_i; \eta_i^g) \right).
+\end{multline}
+Integrating over the unobserved variable $p_i$, we can write the marginal density $f$ of $(m_i, g_i)$ as
+\begin{equation}\label{marginal_density}
+f(m_i, g_i; \theta) = (1-\pi) f(m_i; \eta^m_i(0)) f(g_i; \eta^g_i(0)) + \pi f(m_i; \eta^m_i(1)) f(g_i; \eta^g_i(1)).
+\end{equation}
+We see from (\ref{marginal_density}) that the GLM-EIV model is equivalent to a two-component mixture of \textit{products} of GLM densities.
+Additionally, the GLM-EIV model is a generalization of the classical errors-in-variables model (when the predictor is binary).
+Suppose that we observe data $(x_1, y_1), \dots, (x_n, y_n)$ from the following model:
+\begin{equation}\label{classical_eiv}
+\begin{cases}
+y_i = \beta_0 + \beta_1 x^*_i + \ep_i \\
+x_i = x^*_i + \tau_i,
+\end{cases}
+\end{equation}
+where $x^*_i \sim \textrm{Bern}(\pi), \ep_i \sim N(0,1), \tau_i \sim N(0,1),$ and $\ep_i$,$\tau_i$, and $x^*_i$ are independent.
+The model (\ref{classical_eiv}) is a special case of the GLM-EIV model.
+More generally, GLM-EIV allows the use of entire regression functions (that optionally include covariates and use nonlinear links) to model $y_i$ and $x_i$.
+Moreover, $x_i$ and $y_i$ need not be Gaussian in the GLM-EIV case.
+
+\subsection{Estimation and inference}
+
+We derive an EM algorithm (Algorithm \ref{algo:em_full}) to estimate the parameters of the GLM-EIV model.
+The E step entails computing the membership probability (i.e., the probability of perturbation) of each cell.
+The membership probability $T_i(1)$ of cell $i \in \{1, \dots, n\}$ given the current parameter estimates $(\beta_m^\textrm{(t)}, \beta_g^\textrm{(t)}, \pi^\textrm{(t)})$ and observed data $(m_i, g_i)$ is
+$$T_i(1) = \P(p_i = 1| M_i = m_i, G_i = g_i, \beta^\textrm{(t)}_m, \beta^\textrm{(t)}_g, \pi^\textrm{(t)}).$$ We can calculate this quantity by applying (i) Bayes rule, (ii) the conditional independence property of $M_i$ and $G_i$, (iii) the density of $M_i$ and $G_i$, and (iv) a log-sum-exp-type trick to ensure numerical stability.
+Next, we produce updated estimates $\pi^{(t +1)}$, $\beta_g^{(t+1)}$, and $\beta_m^{(t+1)}$ of the parameters by maximizing the M step objective function.
+It turns out that maximizing the objective function is equivalent to setting $\pi^{\textrm{(t+1)}}$ to the mean of the current membership probabilities and setting $\beta_g^{(t+1)}$ and $\beta_m^{(t+1)}$ to the fitted coefficients of a GLM weighted by the current membership probabilities (Algorithm \ref{algo:em_full}).
+We iterate through the E and M steps until the marginal likelihood ($\ref{marginal_density}$) converges (see appendix for full details).
+Our EM algorithm is reminiscent of (but distinct from) that of Ibrahim \cite{Ibrahim1990}, who also leveraged weighted GLM solvers to carry out an M step.
+
+\begin{algorithm}
+	\caption{EM algorithm for GLM-EIV model.}\label{algo:em_full}
+	\begin{algorithmic}
+		\Require Pilot estimates $\beta^\textrm{curr}_m, \beta^\textrm{curr}_g,$ and $\pi^\textrm{curr}$; data $[m_1, \dots, m_n]$, $[g_1, \dots, g_n]$, $[o^m_1, \dots, o^m_n]$, $[o^g_1, \dots, o^g_n]$, and $[z_1, \dots, z_n]$.
+		\While{Not converged}
+			\For{$i \in \{1, \dots, n\}$} \Comment{E step}
+			\State $T_i(1) \gets \P\left(p_i = 1 |M_i = m_i, G_i = g_i, \beta_m^\textrm{curr}, \beta_g^\textrm{curr}, \pi^\textrm{curr} \right)$
+			\State $T_i(0) \gets 1 - T_i(1)$
+		 \EndFor
+		 \State $\pi^{\textrm{curr}} \gets (1/n) \sum_{i=1}^n T_i(1)$ \Comment{M step}
+		 \State $w \gets [T_1(0), T_2(0), \dots, T_n(0), T_1(1), T_2(1), \dots, T_n(1)]^T$
+		 \State Fit GLM with responses $[m,m]^T$, offsets $[o^m, o^m]^T$, weights $w$, and design matrix $\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix}$; set $\beta_m^\textrm{curr}$ to estimated coefficient vector.
+		\State Fit GLM with responses  $[g,g]^T$, offsets $[o^g, o^g]^T$, weights $w$, and design matrix $\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix}$; set $\beta_g^\textrm{curr}$ to estimated coefficient vector.
+		\State Compute marginal likelihood given $\beta_m^\textrm{curr}$, $\beta_g^\textrm{curr}$, and $\pi^\textrm{curr}$.
+		\EndWhile
+		\State $\hat{\beta}_m \gets \beta_m^\textrm{curr}$; $\hat{\beta}_g \gets \beta_g^\textrm{curr}$; $\hat{\pi} \gets \pi^\textrm{curr}$.
+		\State \textbf{return} $(\hat{\beta}_m, \hat{\beta}_g, \hat{\pi})$
+	\end{algorithmic}
+\end{algorithm}
+
+After fitting the model, we perform inference on the estimated parameters.
+The easiest approach, given the complexity of the log likelihood, would be to run a parametric bootstrap on the fitted model.
+This strategy, however, is prohibitively slow, as the data are large and we use an EM algorithm (that, in practice, requires multiple starts) to fit the model.
+Therefore, we derive an analytic formula for the asymptotic observed information matrix using Louis's Theorem \cite{Louis2012} (see appendix).
+Leveraging this analytic formula, we can calculate standard errors (and $p$-values and confidence intervals) quickly, enabling us to perform inference in practice on real, large-scale data.
+
+\subsection{Statistical accelerations}
+
+% We expect some of these tricks to be of independent utility, especially for designing single-cell methods.
+
+% \subsection{Computational implementation}
 \section{Simulation studies}
@@
 
-\section{Appendix}
-
-\subsection{Theoretical details for thresholding estimator}
+\begin{appendices}
+\section{Theoretical details for thresholding estimator}
 This section contains proofs of the propositions presented Section \ref{sec:thresholding_theory}, ``Theoretical analysis of thresholding estimator.'' The subsections are organized as follows.
@@
 
-\subsubsection{Notation}\label{sec:notation}
+\subsection{Notation}\label{sec:notation}
  All notation introduced in this subsection (i.e., \ref{sec:notation}) pertains to the Gaussian model with intercepts (\ref{theoretical_model}).
@@
 
-\subsubsection{Almost sure limit of $\hat{\beta}^m_1$}\label{sec:convergence}
+\subsection{Almost sure limit of $\hat{\beta}^m_1$}\label{sec:convergence}
 
@@
 
-\subsubsection{Re-expressing $\gamma$ in a simpler form}\label{sec:simplication}
+\subsection{Re-expressing $\gamma$ in a simpler form}\label{sec:simplication}
 We rewrite the attenuation fraction $\gamma$ in a way that makes it more amenable to theoretical analysis. We leverage the fact that $f$ integrates to unity and is even. We have that
@@
 
-\subsubsection{Derivatives of  $g$ and $h$ in $c$}\label{sec:derivatives}
+\subsection{Derivatives of  $g$ and $h$ in $c$}\label{sec:derivatives}
 We compute the derivatives of $g$ and $h$ in $c$, which we will need to prove subsequent results. First, by FTC and the evenness of $f$, we have that
@@
 
-\subsubsection{Limit of $\gamma$ in $c$}\label{sec:c_limit}
+\subsection{Limit of $\gamma$ in $c$}\label{sec:c_limit}
 
@@
 
-\subsubsection{Bayes-optimal decision boundary as a critical value of $\gamma$}\label{sec:bayes_opt}
+\subsection{Bayes-optimal decision boundary as a critical value of $\gamma$}\label{sec:bayes_opt}
 Let $c_\textrm{bayes} = \beta^g_0 + (1/2)\beta^g_1.$ We show that $c = c_\textrm{bayes}$ is a critical value of $\gamma$ for $\pi = 1/2$ and given $\beta^g_1$, i.e, $$D_2 \gamma (\beta^g_1, c_\textrm{bayes}, 1/2) = 0.$$ Differentiating (\ref{gamma_alt2_pi_half}), the quotient rule implies that
@@
 
-\subsubsection{Comparing Bayes-optimal decision boundary and large threshold}\label{sec:comparison}
+\subsection{Comparing Bayes-optimal decision boundary and large threshold}\label{sec:comparison}
 
@@
 
-\subsubsection{Monotonicity in $\beta^g_1$}\label{sec:monotone}
+\subsection{Monotonicity in $\beta^g_1$}\label{sec:monotone}
 We show that $\gamma$ is monotonically increasing in $\beta^g_1$ for $\pi = 1/2$ and given threshold $c$.
@@
 
-\subsubsection{Strict attenuation bias}\label{sec:att_bias}
+\subsection{Strict attenuation bias}\label{sec:att_bias}
 
@@
 
-\subsubsection{Bias-variance decomposition in no-intercept model}\label{sec:bv_decomp}
+\subsection{Bias-variance decomposition in no-intercept model}\label{sec:bv_decomp}
 
@@
 
-\subsection{Derivation of EM algorithm}
+\section{Estimation and inference in the GLM-EIV model}
+\subsection{Estimation}
 
-\subsection{Derivation of observed information matrix}
+\subsection{Inference}
 
-\subsection{Implementation using R family objects}
+\section{Statistical accelerations}
 
-\subsection{Statistical accelerations to GLM-EIV}
+\section{Additional simulation results}
 
-\subsection{Additional simulation results}
+\end{appendices}
 

### 2021-11-08 text/outline/draft.tex
add derivations for submats 1-3

@@
 \usepackage{float}
+\allowdisplaybreaks
 \usepackage[caption = false]{subfig}
@@
 CRISPR genome engineering and single-cell sequencing have transformed biological discovery.
-Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression.
-Single-cell CRISPR screens promise to illuminate regulatory networks underlying diseases but present substantial statistical challenges.
+Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and potentially illuminating regulatory networks underlying diseases.
+In this work we study single-cell CRISPR screens from a statistical perspective.
 We demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter.
@@
 We briefly introduce notation that we will use throughout.
-For $k \in \{0,1\}$, let $\tilde{x}_i(k) := [1, k, z_i]^T$ denote the value of $\tilde{x}_i$ that results from setting $p_i$ to $k$.
-Next, let $l^m_i(k) := \langle \tilde{x}_i(k), \beta_m \rangle + o^m_i$ and $\eta^m_i(k) := h_m(l^m_i(k))$, and let the corresponding gRNA quantities $l^g_i(k)$ and $\eta_i^g(k)$ be defined analogously.
-Also, let the design matrix $\tilde{X} \in \R^{n \times d}$ be defined by
+For $j \in \{0,1\}$, let $\tilde{x}_i(j) := [1, j, z_i]^T$ denote the value of $\tilde{x}_i$ that results from setting $p_i$ to $j$.
+Next, let  $l^m_i(j)$, $\eta^m_i(j),$ and $\mu^m_i(j)$ be the values of $l^m_i$, $\eta^m_i$, and $\mu^m_i$, respectively, that result from setting $p_i$ to $j$, i.e.,
+$$
+\begin{cases}
+l^m_i(j) := \langle \tilde{x}_i(j), \beta_m \rangle + o^m_i \\ \eta^m_i(j) := h_m(l^m_i(j)) \\
+\mu_i^m(j) = r_m^{-1}(l^m_i(j)).
+\end{cases}
+$$
+Let the corresponding gRNA quantities $l^g_i(j)$, $\eta_i^g(j)$, and $\mu^g_i(j)$ be defined analogously. Define the design matrix $\tilde{X} \in \R^{n \times d}$ by
 $$ \tilde{X} := \begin{bmatrix}
@@
 1 & p_n & z_n
-\end{bmatrix}.$$ For $k \in \{0,1\}$, let $\tilde{X}(k) \in \R^{n \times d}$ be the matrix that results from setting $p_i$ to $k$ for all $i \in \{1, \dots, n\}$ in $\tilde{X}$.
+\end{bmatrix}.$$ For $j \in \{0,1\}$, let $\tilde{X}(j) \in \R^{n \times d}$ be the matrix that results from setting $p_i$ to $j$ for all $i \in \{1, \dots, n\}$ in $\tilde{X}$.
 
@@
 \begin{multline}\label{full_log_lik}
-\mathcal{L}(\theta; m_i, g_i, p_i) = \sum_{i=1}^n \log\left( \pi^{p_i}(1-\pi)^{1-p_i} \right) \\ + \sum_{i=1}^n \log\left( f_m(m_i; \eta^m_i)\right) + \sum_{i=1}^n \log\left( f_g(g_i; \eta_i^g) \right).
+\mathcal{L}(\theta; m, g, p) = \sum_{i=1}^n \log\left( \pi^{p_i}(1-\pi)^{1-p_i} \right) \\ + \sum_{i=1}^n \log\left( f_m(m_i; \eta^m_i)\right) + \sum_{i=1}^n \log\left( f_g(g_i; \eta_i^g) \right).
 \end{multline}
@@
 \end{equation}
+Finally, the marginal log-likelihood is
+\begin{multline}\label{marginal_log_lik}
+\mathcal{L}(\theta; m, g) \\ = \sum_{i=1}^n \log\left[ (1-\pi) f(m_i; \eta^m_i(0)) f(g_i; \eta^g_i(0)) + \pi f(m_i; \eta^m_i(1)) f(g_i; \eta^g_i(1)) \right]
+\end{multline}
+
 We see from (\ref{marginal_density}) that the GLM-EIV model is equivalent to a two-component mixture of \textit{products} of GLM densities.
@@
 The model (\ref{classical_eiv}) is a special case of the GLM-EIV model.
-More generally, GLM-EIV allows the use of entire regression functions (that optionally include covariates and use nonlinear links) to model $y_i$ and $x_i$.
-Moreover, $x_i$ and $y_i$ need not be Gaussian in the GLM-EIV case.
+More generally, GLM-EIV allows the use of regression functions (that optionally include covariates and use nonlinear links) to model $y_i$ and $x_i$; $y_i$ and $x_i$ need not be Gaussian.
 
@@
 It turns out that maximizing the objective function is equivalent to setting $\pi^{\textrm{(t+1)}}$ to the mean of the current membership probabilities and setting $\beta_g^{(t+1)}$ and $\beta_m^{(t+1)}$ to the fitted coefficients of a GLM weighted by the current membership probabilities (Algorithm \ref{algo:em_full}).
-We iterate through the E and M steps until the marginal likelihood ($\ref{marginal_density}$) converges (see appendix for full details).
-Our EM algorithm is reminiscent of (but distinct from) that of Ibrahim \cite{Ibrahim1990}, who also leveraged weighted GLM solvers to carry out an M step.
+We iterate through the E and M steps until the marginal log likelihood (\ref{marginal_log_lik}) converges (see appendix for full details).
+Our EM algorithm is reminiscent of (but distinct from) that of Ibrahim \cite{Ibrahim1990}, who also applied weighted GLM solvers to carry out the M step.
 
@@
 \subsection{Estimation}
+We estimate the parameters of the GLM-EIV model using an EM algorithm.
+
+\subsubsection*{E step}
+The E step entails computing the membership probability of each cell.
+Let $\theta^{(t)} = (\beta_m^{(t)}, \beta_g^{(t)}, \pi^{(t)})$ be the parameter estimate at the $t$-th iteration of the algorithm.
+For $k \in \{0,1\}$, let $[\eta^m_i(k)]^{(t)}$ be the $i$th canonical parameter at the $t$-th iteration of the algorithm of the gene expression distribution that results from setting $p_i$ to $k$, i.e.
+$$
+[\eta^m_i(k)]^{(t)} := h_m\left( \langle \tilde{x}_i(k) , \beta_m^{(t)} \rangle + o^m_i \right).
+$$ Similarly, let $\left[\eta^g_i(k)\right]^{(t)}$ be defined by
+$$\left[\eta^g_i(k)\right]^{(t)} :=  h_m\left( \langle \tilde{x}_i(k) , \beta_g^{(t)} \rangle + o^g_i \right).$$
+Next, for $k \in \{0,1\},$ define $\alpha^{(t)}_i(k)$ by
+\begin{multline*}
+\alpha^{(t)}_i(k) := \P\left( M_i = m_i, G_i = g_i | P_i = k, \theta^{(t)} \right) \\ = \P\left( M_i = m_i | P_i = k, \theta^{(t)} \right) \P\left(G_i = g_i | P_i = k, \theta^{(t)} \right) \textrm{ (because $G_i \indep M_i | P_i$)} \\ = f_m\left(m_i; \left[ \eta^m_i(k) \right]^{(t)}\right) f_g\left(g_i; \left[ \eta^g_i(k) \right]^{(t)} \right).
+\end{multline*}
+Finally, let 
+$$
+\begin{cases}
+\pi^{(t)}(1) := \pi^{(t)} = \P\left(P_i = 1 | \theta^{(t)} \right) \\
+\pi^{(t)}(0) := 1 - \pi^{(t)} = \P\left(P_i = 0 | \theta^{(t)} \right).
+\end{cases}
+$$
+The $i$th membership probability $T^{(t)}_i(1)$ is
+\begin{multline*}
+T^{(t)}_i(1) = \P(P_i = 1 | M_i = m_i, G_i = g_i, \theta^{(t)}) \\ = \frac{\pi^{(t)}(1) \alpha^{(t)}_i(1)}{ \sum_{k=0}^1 \pi^{(t)}(k) \alpha^{(t)}_i(k)} \textrm{ (by Bayes rule)} = \frac{1}{\frac{ \pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)} + 1} \\ = \frac{1}{ \exp\left(\log\left(\frac{\pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)}\right)\right) + 1} = \frac{ 1 }{ \exp\left(q^{(t)}_i\right) + 1},
+\end{multline*}
+where we set $$q_i^{(t)} := \log\left(\frac{\pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)}\right).$$
+Next, we have that
+\begin{multline*}
+q^{(t)}_i = \log\left[ \pi^{(t)}(0) \right] + \log\left[ f_m\left(m_i; \left[ \eta^m_i(0) \right]^{(t)}\right) \right] + \log\left[ f_g\left(g_i; \left[ \eta^g_i(0) \right]^{(t)}\right) \right] \\ - \log\left[ \pi^{(t)}(1) \right] - \log\left[ f_m\left(m_i; \left[ \eta^m_i(1) \right]^{(t)}\right) \right] - \log\left[ f_g\left(g_i; \left[ \eta^g_i(1) \right]^{(t)}\right) \right],
+\end{multline*}
+which we can compute. We therefore conclude that
+$$ T_i^{(t)} = \frac{1}{\exp\left(q^{(t)}_i\right) + 1}.$$
+\subsection*{M step}
+Recall that the log-likelihood (\ref{full_log_lik}) of the GLM-EIV model  is
+\begin{multline*}
+\mathcal{L}(\theta; m, g, p) = \sum_{i=1}^n \left[ p_i \log(\pi) + (1-p_i) \log(1-\pi) \right] \\ + \sum_{i=1}^n \log\left( f_m(m_i; \eta^m_i)\right) + \sum_{i=1}^n \log\left( f_g(g_i; \eta_i^g) \right).
+\end{multline*}
+Define $Q(\theta | \theta^{(t)}) = \E_{\left(P |M = m, G = g, \theta^{(t)}\right)}\left[ \mathcal{L}(\theta; m, g, p) \right].$ We have that
+\begin{multline}\label{Q_funct}
+Q(\theta |\theta^{(t)}) = \sum_{i=1}^n \left[T^{(t)}_i(1)\log(\pi) + T_i^{(t)}(0) \log(1 - \pi)\right] \\ + \sum_{k=0}^1 \sum_{i=1}^n T^{(t)}_i(k) \log \left[ f_m(m_i; \eta_i^m(k)) \right] + \sum_{k=0}^1 \sum_{i=1}^n T^{(t)}_i(k) \log \left[ f_g( g_i; \eta^{g,b}_i(k)) \right].
+\end{multline}
+The three terms of (\ref{Q_funct}) are functions of different parameters: the first is a function of $\pi,$ the second is a function of $\beta_m,$ and the third is a function of $\beta_g$.
+Therefore, to find the maximizer $\theta^{(t+1)}$ of (\ref{Q_funct}), we maximize the three terms separately.
+Differentiating the first term with respect to $\pi$, we find that
+	\begin{multline*}
+\frac{ \partial }{\partial \pi } \sum_{i=1}^n \left[ T^{(t)}_i(1)\log(\pi) + T_i^{(t)}(0) \log(1 - \pi)\right]  \\ = \frac{\sum_{i=1}^n T_i^{(t)}(1)}{\pi} - \frac{ \sum_{i=1}^n T_i^{(t)}(0) }{ 1 - \pi}.
+\end{multline*} Setting the derivative equal to $0$ and solving for $\pi$,
+\begin{multline*}
+\frac{\sum_{i=1}^n T_i^{(t)}(1)}{\pi} - \frac{ \sum_{i=1}^n T_i^{(t)}(0) }{ 1 - \pi} = 0 \iff \sum_{i=1}^n T_i^{(t)}(1) - \pi \sum_{i=1}^n T^{(t)}_i(1) = \pi \sum_{i=1}^n T_i(0) \\ \iff \sum_{i=1}^n T^{(t)}_i(1) - \pi\sum_{i=1}^n T_i^{(t)}(1) = \pi n - \pi\sum_{i=1}^n T_i(1) \iff \pi = \frac{ \sum_{i=1}^n T_i^{(t)} (1) }{n}.\end{multline*}
+Thus, the maximizer $\pi^{(t+1)}$ of (\ref{Q_funct}) in $\pi$ is $\pi^{(t+1)} = (1/n)\sum_{i=1}^n T^{(t)}_i(1)$.
+Next, define $w^{(t)} = [T^{(t)}_1(0), \dots, T^{(t)}_n(0), T^{(t)}_1(1), \dots, T^{(t)}_n(1)]^T \in \R^{2n}$.
+We can view the second term of (\ref{Q_funct}) as the log-likelihood of a GLM -- call it $\textrm{GLM}^{(t)}_m$ -- that has exponential family density $f_m$, link function $r_m$, responses $[m,m]^T$, offsets $[o^m, o^m]^T$, weights $w^{(t)}$, and design matrix $\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix}.$ Therefore, the maximizer $\beta^{(t+1)}_m$ of the second term of (\ref{Q_funct}) is the maximizer of $\textrm{GLM}^{(t)}_m$, which we can compute using the iteratively reweighted least squares procedure, as implemented in R's GLM function.
+Similarly, the maximizer $\beta^{(t+1)}_g$ of the third term of (\ref{Q_funct}) is the maximizer of the GLM with exponential family density $f_g$, link function $r_g$, responses $[g,g]^T$, offsets $[o^g, o^g]^T$, weights $w^{(t)}$, and design matrix $\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix}.$
 
 \subsection{Inference}
+We derive the asymptotic observed information matrix of the GLM-EIV log likelihood, enabling us to perform inference on the parameters.
+First, we define some notation.
+For $i \in \{1, \dots, n\}$, $j \in \{0, 1\}$, and $\theta = (\pi, \beta_m, \beta_g),$ let $T^\theta_i(j)$ be defined by
+$$T^\theta_i(j) = \P_\theta\left(P_i = j | M_i = m_i, G_i = g_i\right).$$ Also, define the $\R^n$ vectors $T^\theta(0)$ and $T(1)$ by
+$$T^\theta(0) := [T^\theta_1(0), T^\theta_2(0), \dots, T^\theta_n(0)]$$ and 
+$$T^\theta(1) := [T^\theta_1(1), T^\theta_2(1), \dots, T^\theta_n(1)].$$ Moreover, let $\pi^\theta$
+
+Next, define the diagonal $n \times n$ matrices $\Delta^m$, $[\Delta']^m$, $V^m$, and $H^m$ by
+$$
+\begin{cases}
+	\Delta^m = \textrm{diag} \{ h_m'(l_1^m), \dots, h_m'( l_n^m ) \} \\
+	[\Delta']^m = \textrm{diag} \{ h_m''(l_1^m), \dots, h_m''( l_n^m) \} \\
+	V^m = \textrm{diag} \{ \psi_m( \eta^m_1), \dots, \psi_m( \eta^m_n) \} \\
+	H^m = \textrm{diag} \{ m_1 - \mu_1^m, \dots, m_n - \mu_n^m\}.
+\end{cases} 
+$$ Define the $n \times n$ matrices $\Delta^g, [\Delta']^g, V^g,$ and $H^g$ analogously. These matrices are \textit{unobserved}, as they depend on $\{p_1, \dots, p_n\}$.
+
+For $j \in \{0,1\}$, let the diagonal $n \times n$ matrices $\Delta^m(j), [\Delta']^m(j), V^m(j),$ and $H^m(j)$ be given by
+$$\begin{cases}
+\Delta^m(j) = \textrm{diag} \{ h_m'(l_1^m(j)), \dots, h_m'( l_n^m(j) ) \} \\
+[\Delta']^m(j) = \textrm{diag} \{ h_m''(l_1^m(j)), \dots, h_m''( l_n^m(j)) \} \\
+V^m(j) = \textrm{diag} \{ \psi_m( \eta^m_1(j)), \dots, \psi_m( \eta^m_n(j)) \} \\
+H^m(j) = \textrm{diag} \{ m_1 - \mu_1^m(j), \dots, m_n - \mu_n^m(j)\} .
+\end{cases}
+$$
+Define the matrices $\Delta^g(j)$, $[\Delta']^{g}(j)$, $V^g(j),$ and $H^g(j)$ analogously. Finally, define the vectors $s^m(j)$ and $w^m(j)$ in $\R^n$ by 
+$$ \begin{cases}
+s^m(j) = [m_1 - \mu_1^m(j), \dots, m_n - \mu_n^m(j) ]^T \\ w^m(j) = [ T_1(0)T_1(1)\Delta^m_1(j) H^m_1(j), \dots, T_n(0)T_n(1)\Delta_n^m(j) H_n^m(j)]^T,
+\end{cases} $$
+and let the vectors $s^g(j)$ and $w^g(j)$ be defined analogously.
+The quantities $\Delta^m(j), [\Delta']^m(j), V^m(j),$ $H^m(j),$ $s^m(j),$ $w^m(j),$ $\Delta^g(j), [\Delta']^g(j), V^g(j),$ $H^g(j),$ $s^g(j),$ and $w^g(j)$ are all \textit{observed}.
+
+The observed information matrix $J(\theta; m, g)$ evaluated at $\theta = (\pi, \beta_m, \beta_g)$ is the negative Hessian of the marginal log likelihood (\ref{marginal_log_lik}) evaluated at $\theta$, i.e.
+$$J(\theta; m, g) = - \nabla^2\mathcal{L}(\theta; m, g) .$$ This quantity, unfortunately, is hard to compute, as the log likelihood (\ref{marginal_log_lik}) is a complicated mixture.
+Louis \cite{Louis1982} showed that $J(\theta; m, g)$ is equivalent to the following quantity:
+\begin{multline}\label{zero_inf_info_mat}
+J(\theta; m, g) = -\E \left[\nabla^2 \mathcal{L}(\theta; m, g, p) | G = g, M = m \right] \\ + \E\left[\nabla \mathcal{L}(\theta; m, g, p) | G = g, M = m \right] \E\left[\nabla \mathcal{L}(\theta; m, g, p) | G = g, M = m \right]^T \\ - \E\left[ \nabla\mathcal{L}(\theta; m, g, p) \nabla \mathcal{L}(\theta; m, g, p)^T | G = g, M = m \right].
+\end{multline}
+The observed information matrix $J(\theta; m, g)$ has dimension $(2d+1) \times (2d + 1).$ Recall that the log-likelihood (\ref{full_log_lik}) is the sum of three terms.
+The first term depends only on $\pi$, the second on $\beta_m$, and the third on $\beta_g$.
+Therefore, the observed information matrix can be viewed as block matrix consisting of nine submatrices (Figure \ref{infomatrixbackground}; only six submatrices labelled).
+Submatrix I depends on $\pi$, submatrix II on $\beta_m$, submatrix III on $\beta_g$, submatrix IV on $\beta_m$ and $\beta_g$, submatrix V on $\pi$ and $\beta_g$, and submatrix VI on $\pi$ and $\beta_m$.
+We only need to compute these six submatrices to compute the entire matrix, as the matrix is symmetric.
+The following sections derive formulas for submatrices I-VI.
+All expectations are understood to be \textit{conditional} on $m$ and $g$.
+The notation $\nabla_v$  and $\nabla^2_v$  represent the gradient and Hessian, respectively, with respect to the vector $v$.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\linewidth]{../../figures/info_matrix/info_matrix_background}
+	\caption{Block structure of the observed information matrix $J(\theta; m, g) = -\nabla^2 \mathcal{L}(\theta; m, g)$.
+The matrix is symmetric, and so we only need to compute submatrices I-VI to compute the entire matrix.}
+	\label{infomatrixbackground}
+\end{figure}
+
+\subsubsection*{Submatrix I}
+Denote submatrix I by $J_{\pi}(\theta; m, g).$ The formula for $J_{\pi}(\theta; m, g)$ is 
+\begin{multline}\label{sub_mat_pi}
+J_{\pi}(\theta; m, g) = -\E\left[\nabla^2_\pi \mathcal{L}(\theta; m, g, p) \right] \\ + \left(\E\left[ \nabla_\pi \mathcal{L}(\theta; m, g, p) \right] \right)^2 - \E\left[(\nabla_\pi \mathcal{L}(\theta; m, g, p))^2 \right].
+\end{multline}
+
+We begin by calculating the first and second derivatives of the log-likelihood $\mathcal{L}$ with respect to $\pi$. The first derivative is
+\begin{multline*}
+\nabla_\pi \mathcal{L}(\theta; m, g, p) = \frac{\partial }{\partial \pi } \left( \sum_{i=1}^n p_i \log(\pi) + \sum_{i=1}^n (1 - p_i) \log(1 - \pi) \right) \\ = \frac{ \sum_{i=1}^n p_i }{\pi} - \frac{ \sum_{i=1}^n (1 - p_i) }{ 1 - \pi } = \frac{\sum_{i=1}^n p_i}{\pi} - \frac{n - \sum_{i=1}^n p_i}{1 - \pi} \\ = \left( \frac{1}{\pi} + \frac{1}{1 - \pi} \right) \sum_{i=1}^n p_i - \frac{n}{1-\pi}.
+\end{multline*}
+The second derivative is
+\begin{multline*}
+\nabla^2_\pi \mathcal{L}(\theta; m, g, p)  = \frac{\partial^2}{\partial^2\pi} \left( \frac{ \sum_{i=1}^n p_i }{ \pi } - \frac{ n - \sum_{i=1}^n p_i }{1 - \pi}  \right) = \frac{\left( \sum_{i=1}^n p_i \right) - n}{(1 - \pi)^2} - \frac{\sum_{i=1}^n p_i }{ \pi^2 }.
+\end{multline*}
+We compute the expectation of the first term of (\ref{sub_mat_pi}):
+\begin{multline}\label{submat_pi_1}
+\E \left[ -\nabla^2_{\pi} \mathcal{L}(\theta; m, g, p)\right] = - \E\left[\frac{ ( \sum_{i=1}^n p_i) - n}{(1 - \pi)^2} - \frac{\sum_{i=1}^n p_i}{\pi^2} \right] \\ = - \E\left\{\left[\frac{1}{(1-\pi)^2} - \frac{1}{\pi^2} \right] \sum_{i=1}^n p_i - \frac{n}{ (1 - \pi)^2 } \right\} \\ = - \left\{\left[ \frac{1}{(1-\pi)^2} - \frac{1}{\pi^2} \right] \sum_{i=1}^n T^\theta_i(1) - \frac{n}{ (1 - \pi)^2}  \right\} \\ = \left[ \frac{1}{\pi^2} - \frac{1}{(1 - \pi)^2} \right] \sum_{i=1}^n T^\theta_i(1) + \frac{n}{(1-\pi)^2}.
+\end{multline}
+Next, we compute the difference of the second two pieces of (\ref{sub_mat_pi}). To this end, define $$a := \frac{1}{(1-\pi)} + \frac{1}{\pi}$$ and $$b := \frac{n}{(1-\pi)}.$$ We have that
+\begin{multline*}
+\E \left[ \nabla_\pi \mathcal{L}(\theta; m, g, p)^2 \right] = \E \left[ \left( a \sum_{i=1}^n p_i - b\right)^2 \right] =  \E \left[ a^2 \left( \sum_{i=1}^n p_i \right)^2 - 2ab \sum_{i=1}^n p_i + b^2 \right] \\ = a^2 \sum_{i=1}^n \sum_{j=1}^n \E[p_i p_j] -2ab \sum_{i=1}^n \E [p_i] + b^2.
+\end{multline*}
+Next,
+\begin{multline*}
+\left( \E \left[\nabla_\pi \mathcal{L}(\theta; m, g, x) \right] \right)^2 = \left( a \sum_{i=1}^n \E [p_i] - b \right)^2 = a^2 \sum_{i=1}^n \sum_{j=1}^n \E[p_i]  E[p_j] - 2ab \sum_{i=1}^n \E[p_i] + b^2.
+\end{multline*}
+Therefore,
+\begin{multline}\label{submat_pi_2}
+(\E [\nabla_\pi \mathcal{L}(\theta; m, g, p)])^2 - \E \left[\nabla_\pi \mathcal{L}(\theta; m, g, p)^2 \right] \\ = a^2 \sum_{i=1}^n \sum_{j=1}^n \E[p_i] \E[p_j] - a^2 \sum_{i=1}^n \sum_{j=1}^n \E[p_i p_j] = a^2 \left( \sum_{i=1}^n \E[p_i]^2 - \E[p_i^2]\right) \\ = a^2 \left( \sum_{i=1}^n [T^\theta_i(1)]^2 - T^\theta_i(1) \right) = \left( \frac{1}{(1 - \pi)} + \frac{1}{\pi} \right)^2 \left(\sum_{i=1}^n [T^\theta_i(1)]^2 - T^\theta_i(1) \right).
+\end{multline}
+Stringing (\ref{sub_mat_pi}), (\ref{submat_pi_1}) and (\ref{submat_pi_2}) together, we obtain
+\begin{multline}\label{sub_mat_1_formula}
+J_\pi(\theta; m, g) = 
+\left[ \frac{1}{\pi^2} - \frac{1}{(1 - \pi )^2} \right] \sum_{i=1}^n T^\theta_i(1) + \frac{n}{(1-\pi )^2} \\ + \left( \frac{1}{(1 - \pi )} + \frac{1}{\pi} \right)^2 \left( \sum_{i=1}^n [T^\theta_i(1)]^2 - T^\theta_i(1) \right).
+\end{multline}
+
+\subsubsection*{Submatrix II}
+Denote submatrix II by $J_{\beta^m}(\theta; m, g).$ The formula for $J_{\beta^m}(\theta; m, g)$ is
+\begin{multline}\label{sub_mat_2}
+J_{\beta^m}(\theta; m, g) = -\E \left[\nabla_{\beta^m}^2 \mathcal{L}(\theta; m, g, p) \right] \\ + \E\left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] \E\left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T \\ - \E\left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T  \right].
+\end{multline}
+Standard GLM results imply that
+$$ -\nabla_{\beta^m}^2 \mathcal{L}(\theta; m, g, p) = \tilde{X}^T ( \Delta^m V^m \Delta^m - [\Delta']^m H^m ) \tilde{X}$$ and $$\nabla_{\beta^m}\mathcal{L}(\theta; m, g, p) = \tilde{X}^T \Delta^m s^m.$$
+We compute the first term of (\ref{sub_mat_2}). The $(k,l)$th entry of this matrix is
+\begin{multline*}
+\left( \E\left[-\nabla_{\beta^m}^2 \mathcal{L}(\theta; m, g, p)\right]\right)[k,l] = \E \left\{\tilde{X}[,k]^T (\Delta^m V^m \Delta^m - [\Delta']^mH^m) \tilde{X}[,l] \right\} \\ = \sum_{i=1}^n \E \left\{ \tilde{x}_{i,k} (\Delta^m_{i} V^m_{i} \Delta^m_{i} - [\Delta']^m_{i} H^m_{i}) \tilde{x}_{i,l} \right\} \\ = \sum_{i=1}^n \tilde{x}_{i,k}(0) T_i^{\theta}(0) [{\Delta}^m_i(0)  {V}^m_i(0) {\Delta}^m_i(0) - [\Delta']^m_i(0) {H}^m_i(0)] \tilde{x}_{i,l}(0) \\ + \sum_{i=1}^n \tilde{x}_{i,k}(1) T_i^{\theta}(1) [ {\Delta}^m_i(1)  {V}^m_i(1) {\Delta}^m_i(1) - [{\Delta}']^m_i(1) {H}^m_i(1)] \tilde{x}_{i,l}(1) \\ = \sum_{s = 0}^1 \tilde{X}(s)[,k]^T {T}^{\theta}(s) \left[ {\Delta}^m(s) {V}^m(s) {\Delta}^m(s) - [{\Delta}']^m(s) {H}^m(s) \right] \tilde{X}(s)[,l].
+\end{multline*}
+We therefore have that
+\begin{multline}\label{sub_mat_2_1}
+\E\left[-\nabla_{\beta^m}^2 \mathcal{L}(\theta; m, g, p)\right] \\ = \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) \left[ {\Delta}^m(s) {V}^m(s) {\Delta}^m(s) - [{\Delta}']^m(s) {H}^m(s) \right] \tilde{X}(s).
+\end{multline}
+Next, we compute the difference of the last two terms of (\ref{sub_mat_2}). The $(k,l)$th entry is
+\begin{multline*}
+\bigg[ \E \left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] \E \left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T \\ - \E \left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T \right] \bigg] [k,l] \\
+= \left[ \E \left[\tilde{X}^T \Delta^m s^m \right] \E \left[\tilde{X}^T \Delta^m s^m \right]^T\right][k,l] - \E \left[\tilde{X}^T \Delta^m s^m (s^m)^T \Delta^m \tilde{X} \right][k,l] \\
+= \E\left[ \tilde{X}[,k]^T \Delta^m s^m \right] \E \left[ \tilde{X}[,l]^T \Delta^m s^m \right] - \E \left[ \tilde{X}[,k]^T \Delta^m s^m (s^m)^T \Delta^m \tilde{X}[,l ] \right] \\ 
+=\E\left(\sum_{i=1}^n \tilde{x}_{ik} \Delta^m_i s^m_{i} \right) \E \left( \sum_{j=1}^n \tilde{x}_{jl} \Delta^m_j s^m_j \right) - \E \left( \sum_{i=1}^n \sum_{j=1}^n \tilde{x}_{ik} \Delta^m_i s^m_i s^m_j \Delta^m_j \tilde{x}_{jl} \right) \\
+= \sum_{i=1}^n \sum_{j=1}^n \E[ \tilde{x}_{ik} \Delta^m_is^m_i] \E [\tilde{x}_{jl} \Delta^m_j s^m_j]  -  \sum_{i=1}^n \sum_{j=1}^n \E [ \tilde{x}_{ik} \Delta^m_i s^m_i s^m_j \Delta^m_j \tilde{x}_{jl}] \\
+= \sum_{i=1}^n \sum_{j=1}^n \E[ \tilde{x}_{ik} \Delta^m_i s^m_i] \E \left[\tilde{x}_{jl} \Delta^m_j s^m_j \right]  - \sum_{i \neq j} \E [ \tilde{x}_{ik} \Delta^m_i s^m_i] \E[s^m_j \Delta^m_j \tilde{x}_{jl}] \\ - \sum_{i=1}^n \E[\tilde{x}_{ik} \Delta^m_i s^m_i s^m_i \Delta^m_i \tilde{x}_{il}] \\
+= \sum_{i=1}^n \E[\tilde{x}_{ik} \Delta^m_i s^m_i ] \E [ \tilde{x}_{il} \Delta^m_i s^m_i] - \sum_{i=1}^n \E[\tilde{x}_{ik} (\Delta_i^m)^2 (H_i^m)^2 \tilde{x}_{il}] \\ = \sum_{i=1}^n \left[\tilde{x}_{ik}(0) {\Delta}^m_i(0) T_i^{\theta}(0) {H}^m_i(0) + \tilde{x}_{ik}(1) {\Delta}^m_i(1) T_i^{\theta}(1) {H}^m_i(1) \right] \\ \cdot \left[ \tilde{x}_{il}(0) {\Delta}^m_i(0) T_i^{\theta}(0) {H}^m_i(0) + \tilde{x}_{il}(1) {\Delta}^m_i(1) T_i^{\theta}(1) {H}^m_i(1) \right] \\ - \sum_{i=1}^n \left[ \tilde{x}_{ik}(0) T_i^{\theta}(0) (\Delta_i^m(0))^2 ({H}_i^m(0))^2 \tilde{x}_{il}(0)  + \tilde{x}_{ik}(1) T_i^{\theta}(1) ({\Delta}^m_i(1))^2 ({H}_i^m(1))^m \tilde{x}_{il}(1) \right] \\ = \sum_{s=0}^1 \sum_{t=0}^1 \left[ \sum_{i=1}^n \tilde{x}_{ik}(s) T^{\theta}_i(s) {\Delta}^m_i(s) {H}^m_i(t) T_i^{\theta}(t){\Delta}^m_i(t) {H}^m_i(t) \tilde{x}_{il}(t) \right] \\ - \sum_{s=0}^1 \left[\sum_{i=1}^n \tilde{x}_{ik}(s) T_i^\theta(s) ({\Delta}^m_i(s))^2 ({H}_i^m(s))^2 \tilde{x}_{il}(s) \right] \\ = \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)[,k]^T {T}^\theta(s) {\Delta}^m(s) {H}^m(s) {T}^\theta(t) {\Delta}^m(t) {H}^m(t) \tilde{X}(k)[,l] \\ - \sum_{s=0}^1{X}(s)[,k]^T {T}^\theta(s) ({\Delta}^m(s))^2 ({H}^m(s))^2 \tilde{X}(s)[,l].
+\end{multline*}
+The sum of the last two terms on the right-hand side of (\ref{sub_mat_2}) is therefore
+\begin{multline}\label{sub_mat_2_2}
+ \E \left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] \E \left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T \\ - \E \left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T \right] \\ =
+\sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T {T}^\theta(s) {\Delta}^m(s) {H}^m(s) T^{\theta}(t) {\Delta}^m(t) {H}^m(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T {T}^\theta(s) ({\Delta}^m(s))^2 ({H}^m(s))^2 \tilde{X}(s).
+\end{multline}
+Combining (\ref{sub_mat_2}), (\ref{sub_mat_2_1}), (\ref{sub_mat_2_2}), we find that
+\begin{multline}\label{sub_mat_2_formula}
+J_{\beta^m}(\theta; m, g) = \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) \left[ {\Delta}^m(s) {V}^m(s) {\Delta}^m(s) - [{\Delta}']^m(s) {H}^m(s) \right] \tilde{X}(s) \\ + \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T {T}^\theta(s) {\Delta}^m(s) {H}^m(s) {T}^\theta(t) {\Delta}^m(t) {H}^m(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) ({\Delta}^m(s))^2 ({H}^m(s))^2 \tilde{X}(s).
+\end{multline}
+
+\subsubsection*{Submatrix III}
+Denote submatrix III by $J_{\beta^g}(\theta; m, g).$ The formula for sub-matrix III is similar to that of sub-matrix II (\ref{sub_mat_2_formula}). Substituting $g$ for $m$ in this equation yields
+\begin{multline}\label{sub_mat_3_formula}
+J_{\beta^g}(\theta; m, g) = \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) \left[{\Delta}^g(s) {V}^g(s) {\Delta}^g(s) - [{\Delta}']^g(s) {H}^g(s) \right] \tilde{X}(s) \\ + \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T {T}^\theta(s) {\Delta}^g(s) {H}^g(s) {T}^\theta(t) {\Delta}^g(t) {H}^g(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) ({\Delta}^g(s))^2 ({H}^g(s))^2 \tilde{X}(s).
+\end{multline}
+
+\subsubsection*{Submatrix IV}
+Denote sub-matrix IV by $J_b^{(\beta^g, \beta^m)}(\theta; m, g)$. The formula for $J_b^{(\beta^g, \beta^m)}(\theta; m, g)$ is 
+\begin{multline}\label{sub_mat_4_background}
+J_b^{(\beta^g, \beta^m)}(\theta; m,g) = \E \left[-\nabla_{\beta^{(g,b)}} \nabla_{\beta^m} \mathcal{L}_b(\theta; m, g, p) \right] \\ + \E\left[ \nabla_{\beta^{(g,b)}}\mathcal{L}_b(\theta ; m,g,p) \right] \E \left[\nabla_{\beta^m}\mathcal{L}_b (\theta ; m,g,p)  \right]^T \\ - \E \left[ \nabla_{\beta^{(g,b)}}\mathcal{L}_b (\theta; m,g,p) \nabla_{\beta^m}\mathcal{L}_b(\theta; m,g,p)^T  \right].
+\end{multline} The first term on the right-hand-side of (\ref{sub_mat_4_background}) is $0$.
+Recall that $$\nabla_{\beta^{(g,b)}} \mathcal{L}_b(\theta; m, g, p) =   \tilde{X}^T\Delta^{(g,b)} s^{(g,b)}, \textrm{		} \nabla_{\beta^m} \mathcal{L}_b(\theta; m, g, p) = \tilde{X}^T \Delta^m s^m.$$ The $(k,l)$th entry of the difference of the last two terms of (\ref{sub_mat_4_background}) is
+\begin{multline*}
+\left[ \E \left[\nabla_{\beta^{(g,b)}} \mathcal{L}(\theta; m, g, p) \right] \E \left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T - \E \left[ \nabla_{\beta^{(g,b)}} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T \right] \right][k,l] \\
+= \left[ \E \left[ \tilde{X}^T \Delta^{(g,b)} s^{(g,b)} \right]\E \left[ \tilde{X}^T \Delta^m s^m \right]^T\right][k,l] - \E \left[ \tilde{X}^T \Delta^{(g,b)} s^{(g,b)} (s^m)^T \Delta^m \tilde{X} \right][k,l] \\
+= \E\left[\tilde{X}[,k]^T \Delta^{(g,b)} s^{(g,b)} \right] \E \left[\tilde{X}[,l]^T \Delta^m s^m \right] - \E \left[\tilde{X}[,k]^T \Delta^{(g,b)} s^{(g,b)} (s^m)^T \Delta^m \tilde{X}[,l ] \right] \\
+=\E\left( \sum_{i=1}^n \tilde{x}_{ik} \Delta^{(g,b)}_i s^{(g,b)}_i \right) \E \left( \sum_{j=1}^n \tilde{x}_{jl} \Delta^m_j s^m_j \right) - \E \left( \sum_{i=1}^n \sum_{j=1}^n \tilde{x}_{ik} \Delta^{(g,b)}_i s^{(g,b)}_i s^m_j \Delta^m_j \tilde{x}_{jl} \right) \\
+= \sum_{i=1}^n \sum_{j=1}^n \E[\tilde{x}_{ik} \Delta^{(g,b)}_is^{(g,b)}_i] \E[ \tilde{x}_{jl} \Delta^m_j s^m_j ] - \sum_{i=1}^n \sum_{j=1}^n \E[ \tilde{x}_{ik} \Delta^{(g,b)}_i s^{(g,b)}_i s^m_j \Delta^m_j \tilde{x}_{jl}]  \\
+= \sum_{i=1}^n \sum_{j=1}^n \E[ \tilde{x}_{ik} \Delta^{(g,b)}_i s^{(g,b)}_i] \E \left[\tilde{x}_{jl} \Delta^m_j s^m_j \right]  - \sum_{i \neq j} \E[\tilde{x}_{ik} \Delta^{(g,b)}_i s^{(g,b)}_i] \E[\tilde{x}_{jl}\Delta^m_j  s^m_j ] \\ - \sum_{i=1}^n \E[\tilde{x}_{ik} \Delta^{(g,b)}_i s^{(g,b)}_i s^m_i \Delta^m_i \tilde{x}_{il}] \\
+= \sum_{i=1}^n \E[\tilde{x}_{ik} \Delta^{(g,b)}_i H^{(g,b)}_i] \E[\tilde{x}_{il} \Delta_i^m H^m_i] - \sum_{i=1}^n \E[\tilde{x}_{ik} H_i^{(g,b)} \Delta_i^{(g,b)} \Delta_i^m H_i^m \tilde{x}_{il}] \\ 
+= \sum_{i=1}^n \left[\tilde{x}_{ik}(0) {\Delta}^{(g,b)}_i(0) T_i(0) {H}^{(g,b)}_i(0) + \tilde{x}_{ik}(1) {\Delta}^{(g,b)}_i(1) T_i(1) {H}^{(g,b)}_i(1)\right] \\ 
+\cdot \left[\tilde{x}_{il}(0) {\Delta}^m_i(0) T_i(0) {H}^m_i(0) + \tilde{x}_{il}(1) {\Delta}^m_i(1) T_i(1) {H}^m_i(1)\right] 
+\\ - \sum_{i=1}^n [\tilde{x}_{ik}(0) T_i(0) {\Delta}^{(g,b)}_i(0) {H}^{(g,b)}_i(0) {\Delta}^m_i(0) {H}^m_i(0) \tilde{x}_{il}(0) \\ + \tilde{x}_{ik}(1) T_i(1) {\Delta}^{(g,b)}_i(1) {H}^{(g,b)}_i(1) {\Delta}^m_i(1) {H}^m_i(1) \tilde{x}_{il}(1) ]
+\\ = \sum_{s=0}^1 \sum_{t=0}^1 \left[\sum_{i=1}^n \tilde{x}_{ik}(s) T_i(s) {\Delta}^{(g,b)}_i(s) {H}^{(g,b)}_i(s) T_i(t){\Delta}^m_i(t) {H}^m_i(t) \tilde{x}_{il}(t) \right]
+\\ - \sum_{s=0}^1 \left[\sum_{i=1}^n \tilde{x}_{ik}(s) T_i(s) {\Delta}^{(g,b)}_i(s) {H}^{(g,b)}_i(s) {\Delta}^m_i(s) {H}^m_i(s) \tilde{x}_{il}(s)\right] 
+\\ = \sum_{s=0}^1 \sum_{t=0}^1 \left[ \tilde{X}(s)[,k]^T {T}(s) {\Delta}^{(g,b)}(s) {H}^{(g,b)}(s) {T}(t){\Delta}^m(t) {H}^m(t) \tilde{X}(t)[,l] \right]\\ - \sum_{s=0}^1 \left[ \tilde{X}[,k]^T {T}(s) {\Delta}^{(g,b)}(s) {H}^{(g,b)}(s) {\Delta}^m(s) {H}^m(s) \tilde{X}[,l](s)\right].
+\end{multline*}
+We conclude that
+\begin{multline}\label{sub_mat_4_formula_background}
+J^{(\beta^g, \beta^m)}_b(\theta; m, g) = \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T  {T}(s) {\Delta}^{(g,b)}(s) {H}^{(g,b)}(s) {T}(t){\Delta}^m(t) {H}^m(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T {T}(s) {\Delta}^{(g,b)}(s) {H}^{(g,b)}(s) {\Delta}^m(s) {H}^m(s) \tilde{X}(s).
+\end{multline}
+\\ \noindent
+
+\subsection{Implementation}
 

### 2021-11-08 text/outline/draft.tex
finish background matrix derivation

@@
 \usepackage{algpseudocode}
-\setcounter{tocdepth}{2}
+\setcounter{tocdepth}{1}
 \usepackage{xcolor}
@@
 \tableofcontents
-\noindent \blue{All edits in blue.}
+
 \section{Introduction}
@@
 The density $f_m$ of $m_i$ conditional on the the canonical parameter $\eta_i$ is
-$$f_m(m_i; \eta^m_i) = \exp\left\{ m_i \eta^m_i - \psi_m(\eta^m_i) + c_m(m_i) \right\}.$$ The function $c_m$ does not appear in the log-likelihood of $m_i$; therefore, the only functions relevant to inference are $\psi_m$ and $r_m$.
+$$f_m(m_i; \eta^m_i) = \exp\left\{ m_i \eta^m_i - \psi_m(\eta^m_i) + c_m(m_i) \right\}.$$ The function $c_m$ does not appear in the log likelihood of $m_i$; therefore, the only functions relevant to inference are $\psi_m$ and $r_m$.
 
@@
 Therefore, the observed information matrix can be viewed as block matrix consisting of nine submatrices (Figure \ref{infomatrixbackground}; only six submatrices labelled).
-Submatrix I depends on $\pi$, submatrix II on $\beta_m$, submatrix III on $\beta_g$, submatrix IV on $\beta_m$ and $\beta_g$, submatrix V on $\pi$ and $\beta_g$, and submatrix VI on $\pi$ and $\beta_m$.
+Submatrix I depends on $\pi$, submatrix II on $\beta_m$, submatrix III on $\beta_g$, submatrix IV on $\beta_m$ and $\beta_g$, submatrix V on $\pi$ and $\beta_m$, and submatrix VI on $\pi$ and $\beta_g$.
 We only need to compute these six submatrices to compute the entire matrix, as the matrix is symmetric.
@@
 We begin by calculating the first and second derivatives of the log-likelihood $\mathcal{L}$ with respect to $\pi$. The first derivative is
-\begin{multline*}
+\begin{multline}\label{d_L_d_pi}
 \nabla_\pi \mathcal{L}(\theta; m, g, p) = \frac{\partial }{\partial \pi } \left( \sum_{i=1}^n p_i \log(\pi) + \sum_{i=1}^n (1 - p_i) \log(1 - \pi) \right) \\ = \frac{ \sum_{i=1}^n p_i }{\pi} - \frac{ \sum_{i=1}^n (1 - p_i) }{ 1 - \pi } = \frac{\sum_{i=1}^n p_i}{\pi} - \frac{n - \sum_{i=1}^n p_i}{1 - \pi} \\ = \left( \frac{1}{\pi} + \frac{1}{1 - \pi} \right) \sum_{i=1}^n p_i - \frac{n}{1-\pi}.
-\end{multline*}
+\end{multline}
 The second derivative is
@@
 \begin{multline*}
-\E \left[ \nabla_\pi \mathcal{L}(\theta; m, g, p)^2 \right] = \E \left[ \left( a \sum_{i=1}^n p_i - b\right)^2 \right] =  \E \left[ a^2 \left( \sum_{i=1}^n p_i \right)^2 - 2ab \sum_{i=1}^n p_i + b^2 \right] \\ = a^2 \sum_{i=1}^n \sum_{j=1}^n \E[p_i p_j] -2ab \sum_{i=1}^n \E [p_i] + b^2.
+\E \left[\nabla_\pi \mathcal{L}(\theta; m, g, p)^2 \right] \\ = \E \left[ \left( a \sum_{i=1}^n p_i - b\right)^2 \right]  =  \E \left[ a^2 \left( \sum_{i=1}^n p_i \right)^2 - 2ab \sum_{i=1}^n p_i + b^2 \right] \\ = a^2 \sum_{i=1}^n \sum_{j=1}^n \E[p_i p_j] -2ab \sum_{i=1}^n \E [p_i] + b^2.
 \end{multline*}
@@
 \begin{multline*}
-\left( \E \left[\nabla_\pi \mathcal{L}(\theta; m, g, x) \right] \right)^2 = \left( a \sum_{i=1}^n \E [p_i] - b \right)^2 = a^2 \sum_{i=1}^n \sum_{j=1}^n \E[p_i]  E[p_j] - 2ab \sum_{i=1}^n \E[p_i] + b^2.
+\left( \E \left[\nabla_\pi \mathcal{L}(\theta; m, g, x) \right] \right)^2 = \left( a \sum_{i=1}^n \E [p_i] - b \right)^2 \\ = a^2 \sum_{i=1}^n \sum_{j=1}^n \E[p_i]  E[p_j] - 2ab \sum_{i=1}^n \E[p_i] + b^2.
 \end{multline*}
@@
 \subsubsection*{Submatrix IV}
-Denote sub-matrix IV by $J_b^{(\beta^g, \beta^m)}(\theta; m, g)$. The formula for $J_b^{(\beta^g, \beta^m)}(\theta; m, g)$ is 
-\begin{multline}\label{sub_mat_4_background}
-J_b^{(\beta^g, \beta^m)}(\theta; m,g) = \E \left[-\nabla_{\beta^{(g,b)}} \nabla_{\beta^m} \mathcal{L}_b(\theta; m, g, p) \right] \\ + \E\left[ \nabla_{\beta^{(g,b)}}\mathcal{L}_b(\theta ; m,g,p) \right] \E \left[\nabla_{\beta^m}\mathcal{L}_b (\theta ; m,g,p)  \right]^T \\ - \E \left[ \nabla_{\beta^{(g,b)}}\mathcal{L}_b (\theta; m,g,p) \nabla_{\beta^m}\mathcal{L}_b(\theta; m,g,p)^T  \right].
-\end{multline} The first term on the right-hand-side of (\ref{sub_mat_4_background}) is $0$.
-Recall that $$\nabla_{\beta^{(g,b)}} \mathcal{L}_b(\theta; m, g, p) =   \tilde{X}^T\Delta^{(g,b)} s^{(g,b)}, \textrm{		} \nabla_{\beta^m} \mathcal{L}_b(\theta; m, g, p) = \tilde{X}^T \Delta^m s^m.$$ The $(k,l)$th entry of the difference of the last two terms of (\ref{sub_mat_4_background}) is
-\begin{multline*}
-\left[ \E \left[\nabla_{\beta^{(g,b)}} \mathcal{L}(\theta; m, g, p) \right] \E \left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T - \E \left[ \nabla_{\beta^{(g,b)}} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T \right] \right][k,l] \\
-= \left[ \E \left[ \tilde{X}^T \Delta^{(g,b)} s^{(g,b)} \right]\E \left[ \tilde{X}^T \Delta^m s^m \right]^T\right][k,l] - \E \left[ \tilde{X}^T \Delta^{(g,b)} s^{(g,b)} (s^m)^T \Delta^m \tilde{X} \right][k,l] \\
-= \E\left[\tilde{X}[,k]^T \Delta^{(g,b)} s^{(g,b)} \right] \E \left[\tilde{X}[,l]^T \Delta^m s^m \right] - \E \left[\tilde{X}[,k]^T \Delta^{(g,b)} s^{(g,b)} (s^m)^T \Delta^m \tilde{X}[,l ] \right] \\
-=\E\left( \sum_{i=1}^n \tilde{x}_{ik} \Delta^{(g,b)}_i s^{(g,b)}_i \right) \E \left( \sum_{j=1}^n \tilde{x}_{jl} \Delta^m_j s^m_j \right) - \E \left( \sum_{i=1}^n \sum_{j=1}^n \tilde{x}_{ik} \Delta^{(g,b)}_i s^{(g,b)}_i s^m_j \Delta^m_j \tilde{x}_{jl} \right) \\
-= \sum_{i=1}^n \sum_{j=1}^n \E[\tilde{x}_{ik} \Delta^{(g,b)}_is^{(g,b)}_i] \E[ \tilde{x}_{jl} \Delta^m_j s^m_j ] - \sum_{i=1}^n \sum_{j=1}^n \E[ \tilde{x}_{ik} \Delta^{(g,b)}_i s^{(g,b)}_i s^m_j \Delta^m_j \tilde{x}_{jl}]  \\
-= \sum_{i=1}^n \sum_{j=1}^n \E[ \tilde{x}_{ik} \Delta^{(g,b)}_i s^{(g,b)}_i] \E \left[\tilde{x}_{jl} \Delta^m_j s^m_j \right]  - \sum_{i \neq j} \E[\tilde{x}_{ik} \Delta^{(g,b)}_i s^{(g,b)}_i] \E[\tilde{x}_{jl}\Delta^m_j  s^m_j ] \\ - \sum_{i=1}^n \E[\tilde{x}_{ik} \Delta^{(g,b)}_i s^{(g,b)}_i s^m_i \Delta^m_i \tilde{x}_{il}] \\
-= \sum_{i=1}^n \E[\tilde{x}_{ik} \Delta^{(g,b)}_i H^{(g,b)}_i] \E[\tilde{x}_{il} \Delta_i^m H^m_i] - \sum_{i=1}^n \E[\tilde{x}_{ik} H_i^{(g,b)} \Delta_i^{(g,b)} \Delta_i^m H_i^m \tilde{x}_{il}] \\ 
-= \sum_{i=1}^n \left[\tilde{x}_{ik}(0) {\Delta}^{(g,b)}_i(0) T_i(0) {H}^{(g,b)}_i(0) + \tilde{x}_{ik}(1) {\Delta}^{(g,b)}_i(1) T_i(1) {H}^{(g,b)}_i(1)\right] \\ 
-\cdot \left[\tilde{x}_{il}(0) {\Delta}^m_i(0) T_i(0) {H}^m_i(0) + \tilde{x}_{il}(1) {\Delta}^m_i(1) T_i(1) {H}^m_i(1)\right] 
-\\ - \sum_{i=1}^n [\tilde{x}_{ik}(0) T_i(0) {\Delta}^{(g,b)}_i(0) {H}^{(g,b)}_i(0) {\Delta}^m_i(0) {H}^m_i(0) \tilde{x}_{il}(0) \\ + \tilde{x}_{ik}(1) T_i(1) {\Delta}^{(g,b)}_i(1) {H}^{(g,b)}_i(1) {\Delta}^m_i(1) {H}^m_i(1) \tilde{x}_{il}(1) ]
-\\ = \sum_{s=0}^1 \sum_{t=0}^1 \left[\sum_{i=1}^n \tilde{x}_{ik}(s) T_i(s) {\Delta}^{(g,b)}_i(s) {H}^{(g,b)}_i(s) T_i(t){\Delta}^m_i(t) {H}^m_i(t) \tilde{x}_{il}(t) \right]
-\\ - \sum_{s=0}^1 \left[\sum_{i=1}^n \tilde{x}_{ik}(s) T_i(s) {\Delta}^{(g,b)}_i(s) {H}^{(g,b)}_i(s) {\Delta}^m_i(s) {H}^m_i(s) \tilde{x}_{il}(s)\right] 
-\\ = \sum_{s=0}^1 \sum_{t=0}^1 \left[ \tilde{X}(s)[,k]^T {T}(s) {\Delta}^{(g,b)}(s) {H}^{(g,b)}(s) {T}(t){\Delta}^m(t) {H}^m(t) \tilde{X}(t)[,l] \right]\\ - \sum_{s=0}^1 \left[ \tilde{X}[,k]^T {T}(s) {\Delta}^{(g,b)}(s) {H}^{(g,b)}(s) {\Delta}^m(s) {H}^m(s) \tilde{X}[,l](s)\right].
-\end{multline*}
-We conclude that
-\begin{multline}\label{sub_mat_4_formula_background}
-J^{(\beta^g, \beta^m)}_b(\theta; m, g) = \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T  {T}(s) {\Delta}^{(g,b)}(s) {H}^{(g,b)}(s) {T}(t){\Delta}^m(t) {H}^m(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T {T}(s) {\Delta}^{(g,b)}(s) {H}^{(g,b)}(s) {\Delta}^m(s) {H}^m(s) \tilde{X}(s).
+Denote sub-matrix IV by $J_{(\beta^g, \beta^m)}(\theta; m, g)$. The formula for $J_{(\beta^g, \beta^m)}(\theta; m, g)$ is 
+\begin{multline}\label{sub_mat_4}
+J_{(\beta^g, \beta^m)}(\theta; m,g) = \E \left[-\nabla_{\beta^g} \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] \\ + \E\left[ \nabla_{\beta^{g}}\mathcal{L}(\theta ; m,g,p) \right] \E \left[\nabla_{\beta^m}\mathcal{L} (\theta ; m,g,p)  \right]^T \\ - \E \left[ \nabla_{\beta^{g}}\mathcal{L} (\theta; m,g,p) \nabla_{\beta^m}\mathcal{L}(\theta; m,g,p)^T  \right].
+\end{multline}
+First, we have that
+\begin{equation}\label{sub_mat_4_1}
+\E\left[-\nabla_{\beta^g} \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] = 0,
+\end{equation}
+as differentiating $\mathcal{L}$ with with respect to $\beta^g$ yields a vector that is a function of $\beta^g$, and differentiating this vector with respect to $\beta^m$ yields $0$.
+Next, recall from GLM theory that $$\nabla_{\beta^g} \mathcal{L}(\theta; m, g, p) =   \tilde{X}^T\Delta^g s^g$$ and $$\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) = \tilde{X}^T \Delta^m s^m.$$ The $(k,l)$th entry of the last two terms of (\ref{sub_mat_4}) is
+\begin{multline}\label{sub_mat_4_2}
+\bigg[ \E \left[\nabla_{\beta^g} \mathcal{L}(\theta; m, g, p) \right] \E \left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T \\ - \E \left[ \nabla_{\beta^g} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T \right] \bigg][k,l] \\
+= \left[ \E \left[ \tilde{X}^T \Delta^g s^g \right]\E \left[ \tilde{X}^T \Delta^m s^m \right]^T\right][k,l] - \E \left[ \tilde{X}^T \Delta^g s^g (s^m)^T \Delta^m \tilde{X} \right][k,l] \\ 
+= \E\left[\tilde{X}[,k]^T \Delta^g s^g \right] \E \left[\tilde{X}[,l]^T \Delta^m s^m \right] - \E \left[\tilde{X}[,k]^T \Delta^g s^g (s^m)^T \Delta^m \tilde{X}[,l ] \right] \\
+=\E\left( \sum_{i=1}^n \tilde{x}_{ik} \Delta^g_i s^g_i \right) \E \left( \sum_{j=1}^n \tilde{x}_{jl} \Delta^m_j s^m_j \right) - \E \left( \sum_{i=1}^n \sum_{j=1}^n \tilde{x}_{ik} \Delta^g_i s^g_i s^m_j \Delta^m_j \tilde{x}_{jl} \right) \\
+= \sum_{i=1}^n \sum_{j=1}^n \E[\tilde{x}_{ik} \Delta^g_is^g_i] \E[ \tilde{x}_{jl} \Delta^m_j s^m_j ] - \sum_{i=1}^n \sum_{j=1}^n \E[ \tilde{x}_{ik} \Delta^g_i s^g_i s^m_j \Delta^m_j \tilde{x}_{jl}]  \\
+= \sum_{i=1}^n \sum_{j=1}^n \E[ \tilde{x}_{ik} \Delta^g_i s^g_i] \E \left[\tilde{x}_{jl} \Delta^m_j s^m_j \right]  - \sum_{i \neq j} \E[\tilde{x}_{ik} \Delta^g_i s^g_i] \E[\tilde{x}_{jl}\Delta^m_j  s^m_j ] \\ - \sum_{i=1}^n \E[\tilde{x}_{ik} \Delta^g_i s^g_i s^m_i \Delta^m_i \tilde{x}_{il}] \\
+= \sum_{i=1}^n \E[\tilde{x}_{ik} \Delta^g_i H^g_i] \E[\tilde{x}_{il} \Delta_i^m H^m_i] - \sum_{i=1}^n \E[\tilde{x}_{ik} H_i^g \Delta_i^g \Delta_i^m H_i^m \tilde{x}_{il}] \\ 
+= \sum_{i=1}^n \left[\tilde{x}_{ik}(0) {\Delta}^g_i(0) T^\theta_i(0) {H}^g_i(0) + \tilde{x}_{ik}(1) {\Delta}^g_i(1) T^\theta_i(1) {H}^g_i(1)\right] \\ 
+\cdot \left[\tilde{x}_{il}(0) {\Delta}^m_i(0) T^\theta_i(0) {H}^m_i(0) + \tilde{x}_{il}(1) {\Delta}^m_i(1) T^\theta_i(1) {H}^m_i(1)\right] 
+\\ - \sum_{i=1}^n [\tilde{x}_{ik}(0) T^\theta_i(0) {\Delta}^g_i(0) {H}^g_i(0) {\Delta}^m_i(0) {H}^m_i(0) \tilde{x}_{il}(0) \\ + \tilde{x}_{ik}(1) T^\theta_i(1) {\Delta}^g_i(1) {H}^g_i(1) {\Delta}^m_i(1) {H}^m_i(1) \tilde{x}_{il}(1) ]
+\\ = \sum_{s=0}^1 \sum_{t=0}^1 \left[\sum_{i=1}^n \tilde{x}_{ik}(s) T^\theta_i(s) {\Delta}^g_i(s) {H}^g_i(s) T^\theta_i(t){\Delta}^m_i(t) {H}^m_i(t) \tilde{x}_{il}(t) \right]
+\\ - \sum_{s=0}^1 \left[\sum_{i=1}^n \tilde{x}_{ik}(s) T^\theta_i(s) {\Delta}^g_i(s) {H}^g_i(s) {\Delta}^m_i(s) {H}^m_i(s) \tilde{x}_{il}(s)\right] 
+\\ = \sum_{s=0}^1 \sum_{t=0}^1 \left[ \tilde{X}(s)[,k]^T T^\theta(s) {\Delta}^g(s) {H}^g(s) T^\theta(t){\Delta}^m(t) {H}^m(t) \tilde{X}(t)[,l] \right]\\ - \sum_{s=0}^1 \left[ \tilde{X}[,k]^T T^\theta(s) {\Delta}^g(s) {H}^g(s) {\Delta}^m(s) {H}^m(s) \tilde{X}[,l](s)\right].
+\end{multline}
+Combining (\ref{sub_mat_4}), (\ref{sub_mat_4_1}), and (\ref{sub_mat_4_2}) produces
+\begin{multline}\label{sub_mat_4_formula}
+J_{(\beta^g, \beta^m)}(\theta; m, g) = \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T  T^\theta(s) {\Delta}^g(s) {H}^g(s) T^\theta(t){\Delta}^m(t) {H}^m(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T T^\theta(s) {\Delta}^g(s) {H}^g(s) {\Delta}^m(s) {H}^m(s) \tilde{X}(s).
 \end{multline}
 \\ \noindent
+\subsubsection*{Submatrix V}
+Denote submatrix V by $J_{(\beta^m,\pi)}(\theta; m, g).$ The formula for $J_{(\beta^m,\pi)}(\theta; m, g)$ is
+\begin{multline}\label{sub_mat_5}
+J_{(\beta^m,\pi)}(\theta; m, g) = \E \left[ - \nabla_{\beta^m} \nabla_{ \pi } \mathcal{L}(\theta; m, g, p) \right] \\ + \E\left[ \nabla_{\beta^m}\mathcal{L}(\theta ; m,g,p) \right] \E \left[ \nabla_{\pi}\mathcal{L}(\theta ; m,g,p) \right]^T  - \E \left[ \nabla_{\beta^m}\mathcal{L}(\theta; m,g,p) \nabla_{\pi}\mathcal{L}(\theta; m,g,p)^T \right].
+\end{multline}
+We have that
+\begin{equation}\label{sub_mat_5_1}
+\E \left[ - \nabla_{\beta^m} \nabla_{ \pi } \mathcal{L}(\theta; m, g, p) \right] = 0,
+\end{equation}
+as $\beta^m$ and $\pi$ separate in the log likelihood. Next, set $a := 1/\pi + 1/(1 - \pi)$ and $b := n/(1 - \pi).$ Recall from GLM theory that
+$$\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) = \tilde{X}^T \Delta^m s^m$$ and from (\ref{d_L_d_pi}) that
+$$ a \sum_{i=1}^n p_i - b.$$
+The $k$th entry of the last two terms of (\ref{sub_mat_5}) is
+\begin{multline}\label{sub_mat_5_2}
+\E \left[\nabla_\pi \mathcal{L}(\theta; m, g, p) \right] \E\left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)[k] \right] - \E \left[\nabla_{\pi}\mathcal{L}(\theta; m,g,p) \nabla_{\beta^m}\mathcal{L}(\theta; m,g,p)[k] \right] \\= \left(\E \left[ a \sum_{i=1}^n p_i - b \right] \right) \left(\E\left[ \tilde{X}[,k]^T \Delta^m s^m \right] \right) - \E \left[ \left( a \sum_{i=1}^n p_i - b \right) \tilde{X}[,k]^T \Delta^m s^m \right] \\ = \left( a \sum_{i=1}^n \E[p_i] - b \right) \left( \sum_{j=1}^n \E [ \tilde{x}_{jk}\Delta^m_js^m_j] \right) - \E \left[ \left( a \sum_{i=1}^n p_i - b \right) \left( \sum_{j=1}^n \tilde{x}_{jk} \Delta^m_j s^m_j \right) \right] \\ = a \sum_{i=1}^n \sum_{j=1}^n \E [p_i] \E[ \tilde{x}_{jk} \Delta^m_j s^m_j] - b \sum_{j=1}^n \E[\tilde{x}_{jk} \Delta^m_j s^m_j] \\ - \left[ a \sum_{i=1}^n \sum_{j=1}^n \E [ p_i \tilde{x}_{jk} \Delta^m_j s^m_j] - b \sum_{j=1}^n \E[\tilde{x}_{jk} \Delta^m_j s^m_j] \right] \\ =  a \sum_{i=1}^n \sum_{j=1}^n \E[p_i] \E[\tilde{x}_{jk} \Delta^m_j s^m_j] - a\sum_{i \neq j} \E[p_i] \E[\tilde{x}_{jk} \Delta^m_j s^m_j] - a\sum_{i=1}^n \E[ p_i \tilde{x}_{ik} \Delta^m_i s^m_i ] \\ = a \sum_{i=1}^n \E[p_i] \E[ \tilde{x}_{ik} \Delta^m_i s^m_i] - a \sum_{i=1}^n \E[p_i \tilde{x}_{ik} \Delta^m_i s^m_i] \\ = a \sum_{i=1}^n T^\theta_i(1) [T^\theta_i(0) \Delta^m_i(0) s^m_i(0) \tilde{x}_{ik}(0) + T^\theta_i(1) \Delta^m_i(1) s^m_i(1) \tilde{x}_{ik}(1)] \\ - a \sum_{i=1}^n T^\theta_i(1)\Delta^m_i(1)s^m_i(1)\tilde{x}_{ik}(1) \\ = a \sum_{i=1}^n T^\theta_i(0)T^\theta_i(1) \Delta_i^m(0)H^m_i(0)\tilde{x}_{ik}(0) \\ + a \sum_{i=1}^n \left( [T^\theta_i(1)]^2 \Delta^m_i(1)H^m_i(1) - T^\theta_i(1)\Delta^m_i(1)H^m_i(1) \right) \tilde{x}_{ik}(1)  \\ =a \left[ \sum_{i=1}^n T^\theta_i(0) T^\theta_i(1) \Delta^m_i(0) H^m_i(0) \tilde{x}_{ik}(0) + \sum_{i=1}^n T^\theta_i(1)\Delta^m_i(1)H^m_i(1)[T^\theta_i(1) - 1] \tilde{x}_{ik}(1) \right] \\ = a \left[ \sum_{i=1}^n T^\theta_i(0) T^\theta_i(1) \Delta^m_i(0) H^m_i(0) \tilde{x}_{ik}(0) - \sum_{i=1}^n T^\theta_i(0) T^\theta_i(1) \Delta^m_i(1) H^m_i(1) \tilde{x}_{ik}(1) \right] \\ = a\left(\tilde{X}(0)[,k]^T w^m(0) - \tilde{X}(1)[,k]^T w^m(1)  \right).
+\end{multline}
+Combining (\ref{sub_mat_5}), (\ref{sub_mat_5_1}), and (\ref{sub_mat_5_2}), we conclude that
+\begin{equation}\label{sub_mat_5_formula} J_{(\beta^m, \pi)}(\theta; m, g, p) = \left( \frac{1}{\pi} + \frac{1}{1 - \pi} \right) \left( \tilde{X}(0)^T w^m(0) - \tilde{X}(1)^T w^m(1)\right).
+\end{equation}
+
+\subsubsection*{Submatrix VI}
+Denote submatrix VI by $J_{(\beta^g,\pi)}(\theta; m, g).$ Calculations similar to those for submatrix V show that
+\begin{equation}\label{sub_mat_6_formula} J_{(\beta^g, \pi)}(\theta; m, g, p) = \left(\frac{1}{\pi} + \frac{1}{1 - \pi} \right) \left( \tilde{X}(0)^T w^g(0) - \tilde{X}(1)^T w^g(1)\right).
+\end{equation}
+
+\subsubsection*{Combining submatrices}
+To summarize, the formulas for submatrices I-VI are as follows:
+\begin{itemize}
+\item[I]\begin{multline*}
+J_\pi(\theta; m, g) = 
+\left[ \frac{1}{\pi^2} - \frac{1}{(1 - \pi )^2} \right] \sum_{i=1}^n T^\theta_i(1) + \frac{n}{(1-\pi )^2} \\ + \left( \frac{1}{(1 - \pi )} + \frac{1}{\pi} \right)^2 \left( \sum_{i=1}^n [T^\theta_i(1)]^2 - T^\theta_i(1) \right).
+\end{multline*}
+\item[II] \begin{multline*}
+J_{\beta^m}(\theta; m, g) = \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) \left[ {\Delta}^m(s) {V}^m(s) {\Delta}^m(s) - [{\Delta}']^m(s) {H}^m(s) \right] \tilde{X}(s) \\ + \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T {T}^\theta(s) {\Delta}^m(s) {H}^m(s) {T}^\theta(t) {\Delta}^m(t) {H}^m(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) ({\Delta}^m(s))^2 ({H}^m(s))^2 \tilde{X}(s).
+\end{multline*}
+\item[III] \begin{multline*}
+J_{\beta^g}(\theta; m, g) = \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) \left[{\Delta}^g(s) {V}^g(s) {\Delta}^g(s) - [{\Delta}']^g(s) {H}^g(s) \right] \tilde{X}(s) \\ + \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T {T}^\theta(s) {\Delta}^g(s) {H}^g(s) {T}^\theta(t) {\Delta}^g(t) {H}^g(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) ({\Delta}^g(s))^2 ({H}^g(s))^2 \tilde{X}(s).
+\end{multline*}
+\item[IV] \begin{multline*}
+J_{(\beta^g, \beta^m)}(\theta; m, g) = \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T  T^\theta(s) {\Delta}^g(s) {H}^g(s) T^\theta(t){\Delta}^m(t) {H}^m(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T T^\theta(s) {\Delta}^g(s) {H}^g(s) {\Delta}^m(s) {H}^m(s) \tilde{X}(s).
+\end{multline*}
+\item[V] $$ J_{(\beta^m, \pi)}(\theta; m, g, p) = \left( \frac{1}{\pi} + \frac{1}{1 - \pi} \right) \left( \tilde{X}(0)^T w^m(0) - \tilde{X}(1)^T w^m(1)\right).  $$
+\item[VI] $$ J_{(\beta^g, \pi)}(\theta; m, g, p) = \left(\frac{1}{\pi} + \frac{1}{1 - \pi} \right) \left( \tilde{X}(0)^T w^g(0) - \tilde{X}(1)^T w^g(1)\right).$$
+\end{itemize}
+We stitch these pieces together and transpose submatrices IV, V, and VI to produce the whole information matrix $J(\theta; m, g)$.
+Evaluating this matrix at the EM estimate $\theta^\textrm{EM}$ and inverting yields the asymptotic covariance matrix, which we can use to compute standard errors.
 

### 2021-11-09 text/outline/draft.tex
finish appendix section B

@@
 CRISPR genome engineering and single-cell sequencing have transformed biological discovery.
-Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and potentially illuminating regulatory networks underlying diseases.
+Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases.
 In this work we study single-cell CRISPR screens from a statistical perspective.
@@
 	\begin{algorithmic}
-		\Require Pilot estimates $\beta^\textrm{curr}_m, \beta^\textrm{curr}_g,$ and $\pi^\textrm{curr}$; data $[m_1, \dots, m_n]$, $[g_1, \dots, g_n]$, $[o^m_1, \dots, o^m_n]$, $[o^g_1, \dots, o^g_n]$, and $[z_1, \dots, z_n]$.
+		\Require Pilot estimates $\beta^\textrm{curr}_m, \beta^\textrm{curr}_g,$ and $\pi^\textrm{curr}$; data $[m_1, \dots, m_n]$, $[g_1, \dots, g_n]$, $[o^m_1, \dots, o^m_n]$, $[o^g_1, \dots, o^g_n]$, and $[z_1, \dots, z_n]$; gene expression distribution $f_m$ and link function $r^{-1}_m$; gRNA expression distribution $f_g$ and link function $r^{-1}_g$.
 		\While{Not converged}
@@
 		 \State $w \gets [T_1(0), T_2(0), \dots, T_n(0), T_1(1), T_2(1), \dots, T_n(1)]^T$
-		 \State Fit GLM with responses $[m,m]^T$, offsets $[o^m, o^m]^T$, weights $w$, and design matrix $\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix}$; set $\beta_m^\textrm{curr}$ to estimated coefficient vector.
-		\State Fit GLM with responses  $[g,g]^T$, offsets $[o^g, o^g]^T$, weights $w$, and design matrix $\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix}$; set $\beta_g^\textrm{curr}$ to estimated coefficient vector.
+		 \State Fit GLM with responses $[m,m]^T$, offsets $[o^m, o^m]^T$, weights $w$, and design matrix $\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix}$, distribution $f_m$, and link function $r^{-1}_g$; set $\beta_m^\textrm{curr}$ to estimated coefficient vector.
+		\State Fit GLM with responses  $[g,g]^T$, offsets $[o^g, o^g]^T$, weights $w$, design matrix $\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix}$, distribution $f_g$, and link function $r^{-1}_g$; set $\beta_g^\textrm{curr}$ to estimated coefficient vector.
 		\State Compute marginal likelihood given $\beta_m^\textrm{curr}$, $\beta_g^\textrm{curr}$, and $\pi^\textrm{curr}$.
@@
 This strategy, however, is prohibitively slow, as the data are large and we use an EM algorithm (that, in practice, requires multiple starts) to fit the model.
-Therefore, we derive an analytic formula for the asymptotic observed information matrix using Louis's Theorem \cite{Louis2012} (see appendix).
+Therefore, we derive an analytic formula for the asymptotic observed information matrix using Louis's Theorem \cite{Louis1982} (see appendix).
 Leveraging this analytic formula, we can calculate standard errors (and $p$-values and confidence intervals) quickly, enabling us to perform inference in practice on real, large-scale data.
@@
 $$T^\theta(0) := [T^\theta_1(0), T^\theta_2(0), \dots, T^\theta_n(0)]$$ and 
-$$T^\theta(1) := [T^\theta_1(1), T^\theta_2(1), \dots, T^\theta_n(1)].$$ Moreover, let $\pi^\theta$
+$$T^\theta(1) := [T^\theta_1(1), T^\theta_2(1), \dots, T^\theta_n(1)].$$
 
@@
 	[\Delta']^m = \textrm{diag} \{ h_m''(l_1^m), \dots, h_m''( l_n^m) \} \\
-	V^m = \textrm{diag} \{ \psi_m( \eta^m_1), \dots, \psi_m( \eta^m_n) \} \\
+	V^m = \textrm{diag} \{ \psi''_m( \eta^m_1), \dots, \psi''_m( \eta^m_n) \} \\
 	H^m = \textrm{diag} \{ m_1 - \mu_1^m, \dots, m_n - \mu_n^m\}.
@@
 [\Delta']^m(j) = \textrm{diag} \{ h_m''(l_1^m(j)), \dots, h_m''( l_n^m(j)) \} \\
-V^m(j) = \textrm{diag} \{ \psi_m( \eta^m_1(j)), \dots, \psi_m( \eta^m_n(j)) \} \\
-H^m(j) = \textrm{diag} \{ m_1 - \mu_1^m(j), \dots, m_n - \mu_n^m(j)\} .
+V^m(j) = \textrm{diag} \{\psi''_m( \eta^m_1(j)), \dots, \psi''_m( \eta^m_n(j))\} \\
+H^m(j) = \textrm{diag} \{m_1 - \mu_1^m(j), \dots, m_n - \mu_n^m(j)\} .
 \end{cases}
@@
 \subsection{Implementation}
+To evaluate the observed information matrix, we need to compute the matrices $\Delta^m(j),$ $[\Delta']^m(j),$ $V^m(j),$ and $H^m(j)$ and the vectors $s^m(j)$ and $w^m(j)$ for $j \in \{0,1\}$.
+We likewise need to compute the analogous gRNA quantities.
+We describe how to compute these quantities in R by extending base family objects.
+We implicitly condition on $p_i$, $z^m_i$, and $o^m_i$.
+
+An \texttt{R} family object contains several functions, including \texttt{linkinv}, \texttt{variance}, and \texttt{mu.eta}.
+\texttt{linkinv} is the inverse link function $r_m^{-1}$.
+\texttt{variance} takes as an argument the mean $\mu^m_i$ and returns the variance $[\sigma_i^m]^2$.
+\texttt{mu.eta} is the derivative of the inverse link function $[r^{-1}_m]^{'}$.
+We extend the \texttt{R} family object by adding two additional functions: \texttt{skewness} and \texttt{mu.eta.prime}.
+\texttt{skewness} returns the skewness $\gamma^m_i$ of the distribution as a function of the mean $\mu_i$, i.e.
+$$\texttt{skewness}\left(\mu_i\right) = \E \left[\left(\frac{m_i - \mu_i^m}{ \sigma_i^m}\right)^3\right] := \gamma_i^m.$$ Finally, the \texttt{mu.eta.prime} is the second derivative of the inverse link function $[r^{-1}_m]''.$ Algorithm \ref{algo:computing_info_matrices} computes the matrices $\Delta^m(j)$, $[\Delta']^m(j)$, $V^m(j)$, $H^m(j)$, and vector $s^m(j)$ for given $\beta_m$ and given family object.
+(The vector $w^m(j)$ can be computed in terms of $\Delta^m(j)$ and $H^m(j)$.) We use $\sigma^m_i(j)$ (resp.
+$\gamma^m_i(j)$) to refer to the standard deviation (resp.
+skewness) of the gene expression distribution the $i$th cell when the perturbation $p_i$ is set to $j$.
+
+\begin{algorithm}
+	\caption{Computing the matrices $\Delta^m(j)$, $[\Delta']^m(j)$, $V^m(j)$, $H^m(j)$, and $s^m(j)$ for given $\beta_m$.}\label{algo:computing_info_matrices}
+	\begin{algorithmic}[3]
+		\Require A coefficient vector $\beta_m$; data $[m_1, \dots, m_n]$, $[o^m_1, \dots, o^m_n]$, and $[z_1, \dots, z_n]$; and a family object containing functions \texttt{linkinv}, \texttt{variance}, \texttt{mu.eta}, \texttt{mu.eta.prime}, and \texttt{skewness}.
+		\For{$j \in \{0, 1\}$}
+			\For{$i \in \{1, \dots, n\}$}
+			\State $l^m_i(j) \gets \langle \beta_m, \tilde{x}_i(j) \rangle + o^m_i$
+			\State $\mu^m_i(j) \gets \texttt{linkinv}(l^m_i(j))$
+			\State $[\sigma_i^m(j)]^2 \gets \texttt{variance}(\mu_i^m(j))$
+			\State $h_m'(l_i^m(j)) \gets \texttt{mu.eta}(l_i^m(j))/[\sigma_i^m(j)]^2$
+			\State $\gamma^m_i(j) \gets \texttt{skewness}(\mu^m_i(j))$
+			\State $[r_m^{-1}]''(l_i^m(j)) \gets \texttt{mu.eta.prime}(l^m_i(j))$
+			\State $$h_m''(l_i^m(j)) \gets \frac{[r^{-1}]''(l_i^m(j)) - [([\sigma_i^m(j)]^2)^{3/2}][\gamma^m_i(j)] [h_m'(l_i^m(j))]^2}{[\sigma_i^m(j)]^2}$$
+			\Comment Assign quantities to matrices
+			\State $\Delta_{i}^m(j) \gets h_m'( l_i^m(j))$
+			\State $[\Delta']^m_{i}(j) \gets h''(l^m_i(j))$
+			\State $V^m_{i}(j) \gets [\sigma^m_i(j)]^2$
+			\State $H^m_{i}(j) \gets s^m_i(j) \gets m_i - \mu^m_i(j)$
+			\EndFor
+		\EndFor
+	\end{algorithmic}
+\end{algorithm}
+
+All steps of the algorithm are obvious except the calculation of $h'_m(l^m_i(j))$ (line 6), $h''(l^m_i(j))$ (line 9), and $V^m_i(j)$ (line 12).
+We omit the $(j)$ notation for compactness.
+First, we prove the correctness of the expression for $h'_m(l^m_i)$.
+Recall the basic GLM identities
+\begin{equation}\label{computing_info_matrix_1}
+\psi_m''(\eta_i^m) =  [\sigma^m_i]^2
+\end{equation}
+ and, for all $t \in \R$, 
+\begin{equation}\label{computing_info_matrix_2}
+ r_m^{-1}(t) = \psi_m'(h_m(t)).
+\end{equation}
+ Differentiating (\ref{computing_info_matrix_2}) in $t$, we find that
+\begin{equation}\label{computing_info_matrix_3}
+ (r_m^{-1})'(t) = \psi_m''(h_m(t))h_m'(t), 
+\end{equation}
+or  $$h_m'(t) = \frac{(r_m^{-1})'(t) }{\psi_m''(h_m(t))}.$$ Finally, plugging in $l^m_i$ for $t$,
+$$ h_m'(l_i) = \frac{(r_m^{-1})'(l^m_i)}{\psi''(h_m(l^m_i))} = \frac{(r_m^{-1})'(l^m_i)}{\psi_m''(\eta^m_i)} = \textrm{ by (\ref{computing_info_matrix_1}) } \frac{(r_m^{-1})'(l_i^m)}{[\sigma_i^m]^2}.$$
+
+Next, we prove the correctness for the expression for $h_m''(l_i^m)$. Recall the exponential family identity 
+\begin{equation}\label{computing_info_matrix_4}
+\psi'''_m(\eta^m_i) = \gamma^m_i  ([\sigma^m_i]^2)^{3/2}.
+\end{equation}
+Differentiating (\ref{computing_info_matrix_3}) in $t$, we obtain
+$$ (r_m^{-1})''(t) = \psi_m'''(h_m(t)) [h_m'(t)]^2 + \psi_m''(h_m(t)) h_m''(t),$$ or $$h_m''(t) =\frac{(r_m^{-1})''(t) - \psi'''(h_m(t))[h_m'(t)]^2}{\psi_m''(h_m(t))}.$$ Plugging in $l^m_i$ for $t$, we find that
+\begin{multline*}
+h_m''(l^m_i) = \frac{(r_m^{-1})''(l^m_i) - \psi_m'''(\eta^m_i) [h_m'(l_i^m)]^2}{[\sigma_i^m]^2} \\ = \textrm{ (by \ref{computing_info_matrix_4}) }  \frac{(r_m^{-1})''(l^m_i) - ([\sigma_i^m]^2)^{3/2} (\gamma_i^m) [h_m'(l^m_i)]^2 }{[\sigma_i^m]^2}.
+\end{multline*}
+Finally, the expression for $V^m_i$ follows from (\ref{computing_info_matrix_1}). We can apply a similar algorithm to compute the analogous matrices for the gRNA modality.
+
+% Table of family object functions vs distribution and link.
+
+\section{Zero-inflated model}
+\subsection{Relationship to background real model}
+\subsection{Estimation}
+\subsection{Inference}
 

### 2021-11-10 text/outline/draft.tex
add text for sec 4.3

@@
 \usepackage{algorithm}
+\usepackage{tabularx}
+\usepackage{array}
+\newcommand{\PreserveBackslash}[1]{\let\temp=\\#1\let\\=\temp}
+\newcolumntype{C}[1]{>{\PreserveBackslash\centering}p{#1}}
 \usepackage[toc,page]{appendix}
 \usepackage{algpseudocode}
+\renewcommand{\algorithmicrequire}{\textbf{Input:}}
+\renewcommand{\algorithmicensure}{\textbf{Output:}}
+\makeatletter
+\newcommand{\multiline}[1]{%
+	\begin{tabularx}{\dimexpr\linewidth-\ALG@thistlm}[t]{@{}X@{}}
+		#1
+	\end{tabularx}
+}
+\makeatother
 \setcounter{tocdepth}{1}
@@
 GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.
-GLM-EIV implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic.} Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.
+In doing so, GLM-EIV implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic.} Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.
 
@@
 
-Let $n \in \N$ be the number of cellxs assayed in the experiment.
+Let $n \in \N$ be the number of cells assayed in the experiment.
 Consider a single perturbation and a single gene.
-For cell $i \in \{1, \dots, n\}$, let $m_i \in \N$ be the number of gene transcripts sequenced; let $g_i \in \N$ be the number of gRNA transcripts sequenced; let $l^m_i \in \N$ be the number of gene transcripts sequenced across \textit{all} genes (the library size); and finally, let $z_i \in \R^{d-1}$ be the cell-specific technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.) The letters ``m,'' ``g'', and ``l'' stand for ``mRNA,'' ``gRNA,'' and ``library,'' respectively.
+For cell $i \in \{1, \dots, n\}$, let $m_i \in \N$ be the number of gene transcripts sequenced; let $g_i \in \N$ be the number of gRNA transcripts sequenced; let $d^m_i \in \N$ be the number of gene transcripts sequenced across \textit{all} genes (the library size or sequencing depth); and finally, let $z_i \in \R^{d-1}$ be the cell-specific technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.) The letters ``m,'' ``g'', and ``d'' stand for ``mRNA,'' ``gRNA,'' and ``depth,'' respectively.
 The thresholding method is defined as follows:
@@
 \end{cases}$$
-\item[2.] Assume that $m_i$ is related to $\hat{p}_i, l^m_i,$ and $z_i$ through the following GLM:
-$$m_i | (\hat{p}_i, z_i, l^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i),$$
+\item[2.] Assume that $m_i$ is related to $\hat{p}_i, d^m_i,$ and $z_i$ through the following GLM:
+$$m_i | (\hat{p}_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i),$$
 \begin{equation}\label{thresh_glm}
-\log(\mu_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left( l_i^m\right),
-\end{equation}
-where (i) $NB_\theta(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta^m$; (ii) $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in  \R^{d-1}$ are unknown parameters; and (iii) $\log(l_i^m)$ is an offset term.
+\log(\mu_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left(d_i^m\right),
+\end{equation}
+where (i) $NB_\theta(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta^m$; (ii) $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in  \R^{d-1}$ are unknown parameters; and (iii) $\log(d_i^m)$ is an offset term.
 Fit a GLM to obtain estimates of the parameters.
@@
 \end{itemize}
-\blue{We include the library size $l^m_i$ as an offset term in (\ref{thresh_glm}) so that $\beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i$ can be interpreted as a relative expression: exponentiating both sides of (\ref{thresh_glm}), we obtain
-$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) l_i^m.$$ We see that $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration.}
+\blue{We include the library size $d^m_i$ as an offset term in (\ref{thresh_glm}) so that $\beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i$ can be interpreted as a relative expression: exponentiating both sides of (\ref{thresh_glm}), we obtain
+$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m.$$ We see that $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration.}
 
@@
 \begin{equation}\label{glmeiv_model_1}
-m_i |(p_i, z_i, l^m_i) \sim \textrm{NB}(\mu_i^m),
+m_i |(p_i, z_i, d^m_i) \sim \textrm{NB}(\mu_i^m),
 \end{equation}
 \begin{equation}\label{glmeiv_model_2}
-\log(\mu^m_i) = \beta^m_0 + \beta^m_1 p_i + \gamma_m^T z_i + \log(l^m_i),
+\log(\mu^m_i) = \beta^m_0 + \beta^m_1 p_i + \gamma_m^T z_i + \log(d^m_i),
 \end{equation}
@@
 The model (\ref{glmeiv_model_1}) is identical to the thresholding model (\ref{thresh_glm}), but the imputed perturbation indicator $\hat{p}_i$ is replaced by the latent perturbation indicator $p_i$.
-Next, let $l^g_i \in \N$ be the number of gRNA transcripts sequenced across \textit{all} gRNAs in cell $i$ (i.e., the gRNA library size).
+Next, let $d^g_i \in \N$ be the number of gRNA transcripts sequenced across \textit{all} gRNAs in cell $i$ (i.e., the gRNA library size).
 The model for the gRNA counts is
 \begin{equation}\label{glmeiv_model_3}
-g_i | (p_i, z_i, l^g_i) \sim \textrm{NB}_{\theta^g}\left(\mu_i^g\right),
+g_i | (p_i, z_i, d^g_i) \sim \textrm{NB}_{\theta^g}\left(\mu_i^g\right),
 \end{equation}
 \begin{equation}\label{glmeiv_model_4}
-\log(\mu_i^g) = \beta^g_0 + \beta^g_1p_i + \gamma^T_g z_i + \log(l^g_i),
+\log(\mu_i^g) = \beta^g_0 + \beta^g_1p_i + \gamma^T_g z_i + \log(d^g_i),
 \end{equation}
@@
 (We use the tilde as a reminder that the vector is partially unobserved.) Let $\beta_m = [\beta^m_0, \beta^m_1, \gamma_m]^T \in \R^d$ and $\beta_g = [\beta^g_0, \beta^g_1, \gamma_g]^T \in \R^d$ be the unknown coefficient vectors corresponding to the gene and gRNA expression models, respectively.
-Finally, let $o^m_i$ and $o^g_i$ be the (possibly zero) offset terms for the gene and gRNA models; in practice, we typically set $o^m_i$ and $o^g_i$ to $\log(l^m_i)$ and $\log(l^g_i)$, respectively.
+Finally, let $o^m_i$ and $o^g_i$ be the (possibly zero) offset terms for the gene and gRNA models; in practice, we typically set $o^m_i$ and $o^g_i$ to $\log(d^m_i)$ and $\log(d^g_i)$, respectively.
 
@@
 $$
-Let the corresponding gRNA quantities $l^g_i(j)$, $\eta_i^g(j)$, and $\mu^g_i(j)$ be defined analogously. Define the design matrix $\tilde{X} \in \R^{n \times d}$ by
-$$ \tilde{X} := \begin{bmatrix}
-\tilde{x}_1^T \\ \tilde{x}_2^T \\ \vdots \\ \tilde{x}_n^T
-\end{bmatrix} = 
+Let the corresponding gRNA quantities $l^g_i(j)$, $\eta_i^g(j)$, and $\mu^g_i(j)$ be defined analogously. Define the observed design matrix $X \in  \R^{n \times {d-1}}$ by 
+$$X := \begin{bmatrix} 
+1 & z_1 \\
+1 & z_2 \\
+\vdots & \vdots \\
+1 & z_n
+\end{bmatrix}.$$ 
+Let $\tilde{X} \in \R^{n \times d}$ be the augmented design matrix that results from concatenating the column of (unobserved) $p_i$s to $X$, i.e.
+$$ \tilde{X}  := 
 \begin{bmatrix}
@@
 1 & p_n & z_n
-\end{bmatrix}.$$ For $j \in \{0,1\}$, let $\tilde{X}(j) \in \R^{n \times d}$ be the matrix that results from setting $p_i$ to $j$ for all $i \in \{1, \dots, n\}$ in $\tilde{X}$.
-
-Next, let $[m,m]^T \in \R^{2n}$ denote the vector that results from concatenating $m := [m_1, \dots, m_n]^T \in \R^n$ to itself, i.e.
-$$ [m,m]^T := [\underbrace{m_1, m_2 \dots, m_{n-1}, m_n}_\textrm{first copy of $m$}, \underbrace{m_1, m_2, \dots, m_{n-1}, m_n}_\textrm{second copy of $m$}].$$ Define $[g,g]^T$, $[o^g,o^g]^T$, and $[o^m,o^m]^T$ similarly.
-Finally, let $\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix}$ denote the $\R^{2n \times d}$ matrix that results from vertically concatenating $\tilde{X}(0)$ and $\tilde{X}(1)$.
+\end{bmatrix} = \begin{bmatrix}
+\tilde{x}_1^T \\ \tilde{x}_2^T \\ \vdots \\ \tilde{x}_n^T
+\end{bmatrix}.$$ Furthermore, for $j \in \{0,1\}$, let $\tilde{X}(j) \in \R^{n \times d}$ be the matrix that results from setting $p_i$ to $j$ for all $i \in \{1, \dots, n\}$ in $\tilde{X}$.
+Finally, let $$\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix} = [\tilde{X}(0)^T, \tilde{X}(1)^T]^T$$ be the $\R^{2n \times d}$ matrix that results from vertically concatenating $\tilde{X}(0)$ and $\tilde{X}(1)$.
+
+Next, define $m := [m_1, \dots, m_n]$, and let $g$, $p$, $o^m$, and $o^g$ be defined analogously. Also, let $[m,m]^T \in \R^{2n}$ be the vector that results from concatenating $m$ to itself, i.e.
+$$ [m,m]^T := [\underbrace{m_1, m_2 \dots, m_{n-1}, m_n}_\textrm{first copy of $m$}, \underbrace{m_1, m_2, \dots, m_{n-1}, m_n}_\textrm{second copy of $m$}],$$ and let $[g,g]^T$, $[o^g,o^g]^T$, and $[o^m,o^m]^T$ be defined similarly.
 
@@
 	\begin{algorithmic}
-		\Require Pilot estimates $\beta^\textrm{curr}_m, \beta^\textrm{curr}_g,$ and $\pi^\textrm{curr}$; data $[m_1, \dots, m_n]$, $[g_1, \dots, g_n]$, $[o^m_1, \dots, o^m_n]$, $[o^g_1, \dots, o^g_n]$, and $[z_1, \dots, z_n]$; gene expression distribution $f_m$ and link function $r^{-1}_m$; gRNA expression distribution $f_g$ and link function $r^{-1}_g$.
+		\Require Pilot estimates $\beta^\textrm{curr}_m, \beta^\textrm{curr}_g,$ and $\pi^\textrm{curr}$; data $m$, $g$, $o^m$, $o^g$, and $X$; gene expression distribution $f_m$ and link function $r^{-1}_m$; gRNA expression distribution $f_g$ and link function $r^{-1}_g$.
 		\While{Not converged}
@@
 		 \State $w \gets [T_1(0), T_2(0), \dots, T_n(0), T_1(1), T_2(1), \dots, T_n(1)]^T$
-		 \State Fit GLM with responses $[m,m]^T$, offsets $[o^m, o^m]^T$, weights $w$, and design matrix $\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix}$, distribution $f_m$, and link function $r^{-1}_g$; set $\beta_m^\textrm{curr}$ to estimated coefficient vector.
-		\State Fit GLM with responses  $[g,g]^T$, offsets $[o^g, o^g]^T$, weights $w$, design matrix $\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix}$, distribution $f_g$, and link function $r^{-1}_g$; set $\beta_g^\textrm{curr}$ to estimated coefficient vector.
-		\State Compute marginal likelihood given $\beta_m^\textrm{curr}$, $\beta_g^\textrm{curr}$, and $\pi^\textrm{curr}$.
+		 \For{$k \in \{g,m\}$}
+		 \State  \multiline{ 
+		 	Fit a GLM $GLM_k$ with responses $[k,k]^T$, offsets $[o^k, o^k]^T$, weights $w$, design matrix $[\tilde{X}(0)^T, \tilde{X}(1)^T]^T$, distribution $f_k$, and link function $r^{-1}_k$.
+		}
+		 \State Set $\beta_k^\textrm{curr}$ to the estimated coefficients of $GLM_k$.
+		 \EndFor
+		\State Compute marginal likelihood using $\beta_m^\textrm{curr}$, $\beta_g^\textrm{curr}$, and $\pi^\textrm{curr}$.
 		\EndWhile
@@
 \subsection{Statistical accelerations}
-
-% We expect some of these tricks to be of independent utility, especially for designing single-cell methods.
+A downside of the the EM algorithm (Algorithm \ref{algo:em_full}) is that it requires fitting many GLMs.
+Assuming that we run the algorithm 15 times using randomly-generated pilot estimates (to ensure convergence to the global maximum), and assuming that the algorithm iterates through E and M steps about 10 times per run, we must fit approximately 300 GLMs.
+(These numbers are based on an exploratory application of the method to real and simulated data.) We devised a strategy (Algorithm \ref{algo:pilot_estimates}) for producing a highly accurate pilot estimate $(\pi^\textrm{pilot}, \beta_m^\textrm{pilot}, \beta_g^\textrm{pilot})$, thereby enabling us to run the algorithm once and converge upon the MLE within a few iterations.
+The strategy involves layering four statistical ``tricks'' (some new, some old) on top of one another; some of these tricks could be independently useful for accelerating other single-cell methods (e.g., SCEPTRE \cite{Barry2020}).
+
+The first step (Algorithm \ref{algo:pilot_estimates}, lines 2-4) is to obtain good parameter estimates for $[\beta^m_0, \gamma_m]^T$ and $[\beta^g_0, \gamma_g]^T$.
+Recall that the underlying gene expression parameter vector $\beta_m$ is $\beta_m = [\beta^m_0, \beta^m_1, \gamma_m]^T \in \R^d$, where $\beta^m_0$ is the intercept, $\beta^m_1$ is the effect of the perturbation, and $\gamma_m^T$ is the effect of the technical factors.
+To produce estimates $[\beta^m_0]^\textrm{pilot}$ and $[\gamma_m^T]^\textrm{pilot}$ of $\beta^m_0$ and $\gamma_m^T$, we regress the gene expressions $m$ onto the technical factors $X$.
+The intuition for this is as follows: the probability of perturbation $\pi$ is very small.
+Therefore, the true log likelihood is approximately equal to the log likelihood that results from omitting $p_i$ from the model:
+\begin{multline*}
+\sum_{i=1}^n f_m(m_i; \eta^m_i) = \sum_{i=1}^n f_m(m_i; h_m(\beta_0 + \beta_1 p_i + \gamma^T z_i + o^m_i)) \\ = \underbrace{\sum_{p_i =1}^n f_m(m_i; h_m(\beta_0 + \beta_1 + \gamma^T z_i + o^m_i))}_\textrm{few terms}  + \underbrace{\sum_{p_i = 0}^n f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i))}_\textrm{many terms} \\ \approx \sum_{i = 1}^n f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i)).
+\end{multline*}
+We similarly obtain estimates $[\beta^g_0]^\textrm{pilot}$ and $[\gamma^T_g]^\textrm{pilot}$ for $\beta^g_0$ and $\gamma^T_g$ by regressing the gRNA counts $g$ onto the technical factors $X$.
+
+Next, we obtain estimates $[\beta_1^m]^\textrm{pilot},$ $[\beta_1^g]^\textrm{pilot},$ and $\pi^\textrm{pilot},$ for $\beta^m_1$, $\beta^g_1$, and $\pi$ by fitting a ``reduced'' GLM-EIV model (Algorithm \ref{algo:pilot_estimates}, lines 4 - 29).
+First, we extract the fitted values (on the scale of the linear component) of the above-fitted GLMs:
+$$ \hat{f}^k_i = [\beta^k_0]^\textrm{pilot} + \langle [\gamma^T_k]^\textrm{pilot}, z_i \rangle + o^k_i,$$ for $k \in \{m,g\}.$ The log likelihood of the no-intercept, univariate GLM with predictor $p_i$ and offset $\hat{f}^m_i$ is approximately equal to the true log likelihood:
+\begin{multline*}
+\sum_{i=1}^n f_m(m_i; h_m(\beta_0 + \beta_1 p_i + \gamma^T z_i + o^m_i)) \approx \sum_{i=1}^n f_m(m_i; h_m( \beta_1 p_i + \hat{f}^m_i)),
+\end{multline*}
+and similarly for the gRNA modality.
+Therefore, to estimate $\beta^m_1$ $\beta^g_1,$ and $\pi$, we fit a ``reduced'' GLM-EIV, a special case of the full GLM-EIV with gene expressions $m$, gRNA counts $g$, gene offsets $\hat{f}^m := [\hat{f}^m_1, \dots, \hat{f}^m_n]^T$, gRNA offsets $\hat{f}^g := [\hat{f}^g_1, \dots, \hat{f}^g_n]^T$, and \textit{no} intercept term or covariates.
+Intuitively, we ``encode'' all information about technical factors, library sizes, and baseline expression levels into $\hat{f}^m$ and $\hat{f}^g$.
+
+Finally, the M step of the reduced GLM-EIV algorithm involves fitting two no-intercept, univariate GLMs with offsets (Algorithm \ref{algo:pilot_estimates}, line 20).
+We derived analytic formulas for the MLEs of these GLMs in the three most important cases: Gaussian response with identity link, Poisson response with log link, and negative binomial response with log link (see appendix; the latter formula is asymptotically exact).
+Because we do not need to use an iterative procedure (e.g., IRLS) to run the M step of the ``reduced'' GLM-EIV algorithm, this algorithm is fast.
+Overall, our procedure for obtaining pilot parameter estimates (Algorithm \ref{algo:pilot_estimates}) requires fitting two GLMs (via IRLS).
+
+\begin{algorithm}
+	\caption{Computing pilot parameter estimates.}\label{algo:pilot_estimates}
+	\begin{algorithmic}[2]
+		\Require Data $m$, $g$, $o^m$, $o^g$, and $X$; gene expression distribution $f_m$ and link function $r^{-1}_m$; gRNA expression distribution $f_g$ and link function $r^{-1}_g$; number of EM starts $B$.
+		\For {$k \in \{m,g\}$}
+		\State Fit a GLM $GLM_k$ with responses $k$, offsets $o^k$, design matrix $X$, distribution $f_k$, and link function $r^{-1}_k$.
+		\State Set $[\beta_0^k]^\textrm{pilot}$ and $[\gamma_k^T]^\textrm{pilot}$ to the fitted coefficients of $GLM_k$.
+		\For{$i \in \{1, \dots, n\}$}
+			\State $\hat{f}^k_i \gets [\beta_0^k]^\textrm{pilot} + \langle [\gamma_k^T]^\textrm{pilot} , z_i \rangle + o_i^k$ \Comment{untransformed fitted values}
+		\EndFor
+		\EndFor
+		\State \texttt{bestLik} $\gets -\infty$ \Comment{Reduced GLM-EIV}
+		\State $p \gets [\underbrace{0, \dots, 0}_\textrm{n}, \underbrace{1, \dots, 1}_\textrm{n}]$
+		\For{$i \in \{1,\dots,B\}$}
+		\State Randomly generate starting parameters $\pi^\textrm{curr}, [\beta_1^m]^\textrm{curr}, [\beta^g_1]^\textrm{curr}.$
+		\While{Not converged}
+		\For{$i \in \{1, \dots, n\}$} \Comment{E step}
+		\State $T_i(1) \gets \P(P_i = 1 | M_i = m_i, G_i = g_i, \pi^\textrm{curr}, [\beta^g_1]^\textrm{curr}, [\beta_1^m]^\textrm{curr})$
+		\State $T_i(0) \gets 1 - T_i(1)$
+		\EndFor
+		\State $\pi^{\textrm{curr}} \gets (1/n) \sum_{i=1}^n T_i(1)$ \Comment{M step}
+		 \State $w \gets [T_1(0), T_2(0), \dots, T_n(0), T_1(1), T_2(1), \dots, T_n(1)]^T$
+		 \For{$k \in \{g,m\}$}
+		 \State \multiline{Fit no-intercept, univariate GLM $GLM_k$ with predictors $p$, responses $[k,k]^T$, offsets $[\hat{f}^k, \hat{f}^k]^T,$ and weights $w$.}
+		 \State Set $[\beta^k_1]^\textrm{curr}$ to fitted coefficient of $GLM_k$.
+		 \EndFor
+		 \State \multiline{
+		 Compute log likelihood \texttt{currLik} using $\pi^\textrm{curr}$,$ [\beta^m_1]^\textrm{curr}$, and $[\beta^g_1]^\textrm{curr}.$}
+		\EndWhile
+		\If{\texttt{currLik} $>$ \texttt{bestLik}}
+		\State bestLik $\gets$ currLik
+		\State $\pi^\textrm{pilot} \gets \pi^\textrm{curr};$ $[\beta^m_1]^\textrm{pilot} \gets [\beta^m_1]^\textrm{curr}; [\beta^g_1]^\textrm{pilot} \gets [\beta^g_1]^\textrm{curr}$
+		\EndIf
+		\EndFor
+		\State \textbf{return} $(\pi^\textrm{pilot}, [\beta^m_0]^\textrm{pilot}, [\beta^m_1]^\textrm{pilot}, [\gamma^T_m]^\textrm{pilot}, [\beta^g_0]^\textrm{pilot}, [\beta^g_1]^\textrm{pilot}, [\gamma^T_g]^\textrm{pilot})$
+	\end{algorithmic}
+\end{algorithm}
+
+% The underlying parameter vector is $\beta_0, \beta_1, \gamma,$ where $\beta_0$ is an intercept term, $\beta_1$ is the perturbation effect, and $\gamma$ are the technical factor effects.
+The first step is to obtain pilot paramter estimates for $\beta_0$ and $\gamma^m$.
+We present algo X.
+This works because the marginal perturbation probability $\pi$ is small; therefore, the model for $l_i^m$ is approximately correct, and we can get good estimates.
+
+% Next, we obtain pilot estimates for $\beta_1^m, \pi, \beta_1^g$.
+To do this, we fit a reduced GLM-EIV.
+The ``reduced'' GLM-EIV is a special case in which we model gene expressions and gRNA counts using an intercept and offsets only.
+This is approximately correct, because our pilot estimates for the other parameters are nearly correct.
+
+% The final trick is to fit the M step quickly.
+The GLMs are ``mean + offset'' models; we can fit these mean + offset models using IRLS.
+However, this iterative procedure is slow.
+When we use Poisson or NB models with log link (the most common case), we derived analytic formulas that are exactly or asymptotically correct.
+(Combines two tricks).
 
@@
 Next, define $w^{(t)} = [T^{(t)}_1(0), \dots, T^{(t)}_n(0), T^{(t)}_1(1), \dots, T^{(t)}_n(1)]^T \in \R^{2n}$.
-We can view the second term of (\ref{Q_funct}) as the log-likelihood of a GLM -- call it $\textrm{GLM}^{(t)}_m$ -- that has exponential family density $f_m$, link function $r_m$, responses $[m,m]^T$, offsets $[o^m, o^m]^T$, weights $w^{(t)}$, and design matrix $\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix}.$ Therefore, the maximizer $\beta^{(t+1)}_m$ of the second term of (\ref{Q_funct}) is the maximizer of $\textrm{GLM}^{(t)}_m$, which we can compute using the iteratively reweighted least squares procedure, as implemented in R's GLM function.
+We can view the second term of (\ref{Q_funct}) as the log-likelihood of a GLM -- call it $\textrm{GLM}^{(t)}_m$ -- that has exponential family density $f_m$, link function $r_m$, responses $[m,m]^T$, offsets $[o^m, o^m]^T$, weights $w^{(t)}$, and design matrix $\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix}.$ Therefore, the maximizer $\beta^{(t+1)}_m$ of the second term of (\ref{Q_funct}) is the maximizer of $\textrm{GLM}^{(t)}_m$, which we can compute using the iteratively reweighted least squares (IRLS) procedure, as implemented in R's GLM function.
 Similarly, the maximizer $\beta^{(t+1)}_g$ of the third term of (\ref{Q_funct}) is the maximizer of the GLM with exponential family density $f_g$, link function $r_g$, responses $[g,g]^T$, offsets $[o^g, o^g]^T$, weights $w^{(t)}$, and design matrix $\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix}.$
@@
 \end{multline*}
-Finally, the expression for $V^m_i$ follows from (\ref{computing_info_matrix_1}). We can apply a similar algorithm to compute the analogous matrices for the gRNA modality.
-
-% Table of family object functions vs distribution and link.
+Finally, the expression for $V^m_i$ follows from (\ref{computing_info_matrix_1}).
+We can apply a similar algorithm to compute the analogous matrices for the gRNA modality.
+Table \ref{family_object_functions} shows the \texttt{linkinv}, \texttt{variance}, \texttt{mu.eta}, \texttt{skewness}, and \texttt{mu.eta.prime} functions for several common family objects (which are defined by a distribution and link function).
+\begin{table}
+\caption{\texttt{linkinv}, \texttt{variance}, \texttt{mu.eta}, \texttt{skewness}, \texttt{mu.eta.prime} for common family objects (i.e., pairs of distributions and link functions).}\label{family_object_functions}
+\begin{tabular}{|C{2.5cm}|C{3.5cm}|C{3.4cm}|C{2.5cm}|}
+	\hline 
+	& Gaussian response, identity link & Poisson response, log link & NB response ($\theta > 0$ fixed), log link \\ 
+	\hline 
+	\texttt{linkinv} & $x$ & $\exp(x)$ & $\exp(x)$  \\ 
+	\hline 
+	\texttt{variance} & $x$ & $x$ & $x + x^2/\theta$ \\ 
+	\hline 
+	\texttt{mu.eta} & $1$ & $x$  & $\exp(x)$ \\ 
+	\hline 
+	\texttt{skewness} & $0$ & $x^{-1/2}$ & $\frac{2 x + \theta}{\sqrt{\theta x} \sqrt{x + \theta}}$ \\ 
+	\hline 
+	\texttt{mu.eta.prime} & $0$ & $\exp(x)$ & $\exp(x)$ \\ 
+	\hline 
+\end{tabular}
+\end{table}
 

### 2021-11-12 text/outline/draft.tex
add statistical acceleration section in appendix

@@
 \documentclass[12pt]{article}
+ \usepackage[vmargin=1.18in,hmargin=1.1in]{geometry}
 \usepackage{amsfonts}
@@
 In this work we tackle a set of analysis challenges that are complimentary to the challenges addressed by SCEPTRE.
-Most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, \blue{a statistical objective unattainable within the nonparameteric hypothesis testing framework of SCEPTRE.}
+Most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, \blue{an unattainable objective within the nonparameteric hypothesis testing framework of SCEPTRE.}
 
@@
 The easiest approach, given the complexity of the log likelihood, would be to run a parametric bootstrap on the fitted model.
-This strategy, however, is prohibitively slow, as the data are large and we use an EM algorithm (that, in practice, requires multiple starts) to fit the model.
+This strategy, however, is prohibitively slow, as the data are large and we use an EM algorithm to fit the model.
 Therefore, we derive an analytic formula for the asymptotic observed information matrix using Louis's Theorem \cite{Louis1982} (see appendix).
@@
 Assuming that we run the algorithm 15 times using randomly-generated pilot estimates (to ensure convergence to the global maximum), and assuming that the algorithm iterates through E and M steps about 10 times per run, we must fit approximately 300 GLMs.
-(These numbers are based on an exploratory application of the method to real and simulated data.) We devised a strategy (Algorithm \ref{algo:pilot_estimates}) for producing a highly accurate pilot estimate $(\pi^\textrm{pilot}, \beta_m^\textrm{pilot}, \beta_g^\textrm{pilot})$, thereby enabling us to run the algorithm once and converge upon the MLE within a few iterations.
-The strategy involves layering four statistical ``tricks'' (some new, some old) on top of one another; some of these tricks could be independently useful for accelerating other single-cell methods (e.g., SCEPTRE \cite{Barry2020}).
-
-The first step (Algorithm \ref{algo:pilot_estimates}, lines 2-4) is to obtain good parameter estimates for $[\beta^m_0, \gamma_m]^T$ and $[\beta^g_0, \gamma_g]^T$.
-Recall that the underlying gene expression parameter vector $\beta_m$ is $\beta_m = [\beta^m_0, \beta^m_1, \gamma_m]^T \in \R^d$, where $\beta^m_0$ is the intercept, $\beta^m_1$ is the effect of the perturbation, and $\gamma_m^T$ is the effect of the technical factors.
-To produce estimates $[\beta^m_0]^\textrm{pilot}$ and $[\gamma_m^T]^\textrm{pilot}$ of $\beta^m_0$ and $\gamma_m^T$, we regress the gene expressions $m$ onto the technical factors $X$.
-The intuition for this is as follows: the probability of perturbation $\pi$ is very small.
-Therefore, the true log likelihood is approximately equal to the log likelihood that results from omitting $p_i$ from the model:
-\begin{multline*}
-\sum_{i=1}^n f_m(m_i; \eta^m_i) = \sum_{i=1}^n f_m(m_i; h_m(\beta_0 + \beta_1 p_i + \gamma^T z_i + o^m_i)) \\ = \underbrace{\sum_{p_i =1}^n f_m(m_i; h_m(\beta_0 + \beta_1 + \gamma^T z_i + o^m_i))}_\textrm{few terms}  + \underbrace{\sum_{p_i = 0}^n f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i))}_\textrm{many terms} \\ \approx \sum_{i = 1}^n f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i)).
-\end{multline*}
-We similarly obtain estimates $[\beta^g_0]^\textrm{pilot}$ and $[\gamma^T_g]^\textrm{pilot}$ for $\beta^g_0$ and $\gamma^T_g$ by regressing the gRNA counts $g$ onto the technical factors $X$.
-
-Next, we obtain estimates $[\beta_1^m]^\textrm{pilot},$ $[\beta_1^g]^\textrm{pilot},$ and $\pi^\textrm{pilot},$ for $\beta^m_1$, $\beta^g_1$, and $\pi$ by fitting a ``reduced'' GLM-EIV model (Algorithm \ref{algo:pilot_estimates}, lines 4 - 29).
-First, we extract the fitted values (on the scale of the linear component) of the above-fitted GLMs:
-$$ \hat{f}^k_i = [\beta^k_0]^\textrm{pilot} + \langle [\gamma^T_k]^\textrm{pilot}, z_i \rangle + o^k_i,$$ for $k \in \{m,g\}.$ The log likelihood of the no-intercept, univariate GLM with predictor $p_i$ and offset $\hat{f}^m_i$ is approximately equal to the true log likelihood:
-\begin{multline*}
-\sum_{i=1}^n f_m(m_i; h_m(\beta_0 + \beta_1 p_i + \gamma^T z_i + o^m_i)) \approx \sum_{i=1}^n f_m(m_i; h_m( \beta_1 p_i + \hat{f}^m_i)),
-\end{multline*}
-and similarly for the gRNA modality.
-Therefore, to estimate $\beta^m_1$ $\beta^g_1,$ and $\pi$, we fit a ``reduced'' GLM-EIV, a special case of the full GLM-EIV with gene expressions $m$, gRNA counts $g$, gene offsets $\hat{f}^m := [\hat{f}^m_1, \dots, \hat{f}^m_n]^T$, gRNA offsets $\hat{f}^g := [\hat{f}^g_1, \dots, \hat{f}^g_n]^T$, and \textit{no} intercept term or covariates.
-Intuitively, we ``encode'' all information about technical factors, library sizes, and baseline expression levels into $\hat{f}^m$ and $\hat{f}^g$.
-
-Finally, the M step of the reduced GLM-EIV algorithm involves fitting two no-intercept, univariate GLMs with offsets (Algorithm \ref{algo:pilot_estimates}, line 20).
-We derived analytic formulas for the MLEs of these GLMs in the three most important cases: Gaussian response with identity link, Poisson response with log link, and negative binomial response with log link (see appendix; the latter formula is asymptotically exact).
-Because we do not need to use an iterative procedure (e.g., IRLS) to run the M step of the ``reduced'' GLM-EIV algorithm, this algorithm is fast.
-Overall, our procedure for obtaining pilot parameter estimates (Algorithm \ref{algo:pilot_estimates}) requires fitting two GLMs (via IRLS).
+(These numbers are based on an exploratory application of the method to real and simulated data.) We devised a strategy (Algorithm \ref{algo:pilot_estimates}) to produce a highly accurate pilot estimate $(\pi^\textrm{pilot}, \beta_m^\textrm{pilot}, \beta_g^\textrm{pilot})$ of the true parameters, enabling us to run the algorithm once and converge upon the MLE within a few iterations.
+The strategy involves layering several statistical ``tricks'' (some new, some old) on top of one another.
+We expect these tricks to be independently useful for accelerating other single-cell methods (e.g., SCEPTRE \cite{Barry2020}).
 
@@
 		\For {$k \in \{m,g\}$}
-		\State Fit a GLM $GLM_k$ with responses $k$, offsets $o^k$, design matrix $X$, distribution $f_k$, and link function $r^{-1}_k$.
+		\State \multiline{Fit a GLM $GLM_k$ with responses $k$, offsets $o^k$, design matrix $X$, distribution $f_k$, and link function $r^{-1}_k$.}
 		\State Set $[\beta_0^k]^\textrm{pilot}$ and $[\gamma_k^T]^\textrm{pilot}$ to the fitted coefficients of $GLM_k$.
@@
 		\State \texttt{bestLik} $\gets -\infty$ \Comment{Reduced GLM-EIV}
-		\State $p \gets [\underbrace{0, \dots, 0}_\textrm{n}, \underbrace{1, \dots, 1}_\textrm{n}]$
 		\For{$i \in \{1,\dots,B\}$}
@@
 		\For{$k \in \{g,m\}$}
-		 \State \multiline{Fit no-intercept, univariate GLM $GLM_k$ with predictors $p$, responses $[k,k]^T$, offsets $[\hat{f}^k, \hat{f}^k]^T,$ and weights $w$.}
+		\State \multiline{Fit no-intercept, univariate GLM $GLM_k$ with predictors $[\underbrace{0, \dots, 0}_\textrm{n}, \underbrace{1, \dots, 1}_\textrm{n}]$, responses $[k,k]^T$, offsets $[\hat{f}^k, \hat{f}^k]^T,$ and weights $w$.}
 		\State Set $[\beta^k_1]^\textrm{curr}$ to fitted coefficient of $GLM_k$.
@@
 
-% The underlying parameter vector is $\beta_0, \beta_1, \gamma,$ where $\beta_0$ is an intercept term, $\beta_1$ is the perturbation effect, and $\gamma$ are the technical factor effects.
-The first step is to obtain pilot paramter estimates for $\beta_0$ and $\gamma^m$.
-We present algo X.
-This works because the marginal perturbation probability $\pi$ is small; therefore, the model for $l_i^m$ is approximately correct, and we can get good estimates.
-
-% Next, we obtain pilot estimates for $\beta_1^m, \pi, \beta_1^g$.
-To do this, we fit a reduced GLM-EIV.
-The ``reduced'' GLM-EIV is a special case in which we model gene expressions and gRNA counts using an intercept and offsets only.
-This is approximately correct, because our pilot estimates for the other parameters are nearly correct.
-
-% The final trick is to fit the M step quickly.
-The GLMs are ``mean + offset'' models; we can fit these mean + offset models using IRLS.
-However, this iterative procedure is slow.
-When we use Poisson or NB models with log link (the most common case), we derived analytic formulas that are exactly or asymptotically correct.
-(Combines two tricks).
-
-% \subsection{Computational implementation}
+The first step (Algorithm \ref{algo:pilot_estimates}, lines 2-7) is to obtain good parameter estimates for $[\beta^m_0, \gamma_m]^T$ and $[\beta^g_0, \gamma_g]^T$ using regression.
+Recall that the underlying gene expression parameter vector $\beta_m$ is $\beta_m = [\beta^m_0, \beta^m_1, \gamma_m]^T \in \R^d$, where $\beta^m_0$ is the intercept, $\beta^m_1$ is the effect of the perturbation, and $\gamma_m^T$ is the effect of the technical factors.
+To produce estimates $[\beta^m_0]^\textrm{pilot}$ and $[\gamma_m^T]^\textrm{pilot}$, we regress the gene expressions $m$ onto the technical factors $X$.
+The intuition for why this works is as follows: the probability of perturbation $\pi$ is very small.
+Therefore, the true log likelihood is approximately equal to the log likelihood that results from omitting $p_i$ from the model:
+\begin{multline*}
+\sum_{i=1}^n f_m(m_i; \eta^m_i) = \sum_{i=1}^n f_m(m_i; h_m(\beta_0 + \beta_1 p_i + \gamma^T z_i + o^m_i)) \\ = \underbrace{\sum_{p_i =1}^n f_m(m_i; h_m(\beta_0 + \beta_1 + \gamma^T z_i + o^m_i))}_\textrm{few terms}  + \underbrace{\sum_{p_i = 0}^n f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i))}_\textrm{many terms} \\ \approx \sum_{i = 1}^n f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i)).
+\end{multline*}
+We similarly can obtain pilot estimates $[\beta^g_0]^\textrm{pilot}$ and $[\gamma^T_g]^\textrm{pilot}$ by regressing the gRNA counts $g$ onto the technical factors $X$.
+We extract the fitted values (on the scale of the linear component) for use in a subsequent step: $$\hat{f}^k_i = [\beta^k_0]^\textrm{pilot} + \langle [\gamma^T_k]^\textrm{pilot}, z_i \rangle + o^k_i,$$
+for $k \in \{m,g\}.$
+
+Next, we obtain estimates $[\beta_1^m]^\textrm{pilot},$ $[\beta_1^g]^\textrm{pilot},$ and $\pi^\textrm{pilot}$ for $\beta^m_1$, $\beta^g_1$, and $\pi$ by fitting a ``reduced'' GLM-EIV (Algorithm \ref{algo:pilot_estimates}, lines 8 onward).
+The log likelihood of the no-intercept, univariate GLM with predictor $p_i$ and offset $\hat{f}^m_i$ is approximately equal to the true log likelihood:
+$$ \sum_{i=1}^n f_m(m_i; \eta^m_i) =
+\sum_{i=1}^n f_m(m_i; h_m(\beta_0 + \beta_1 p_i + \gamma^T z_i + o^m_i)) \approx \sum_{i=1}^n f_m(m_i; h_m( \beta_1 p_i + \hat{f}^m_i)).
+$$
+Therefore, to estimate $\beta^m_1$, $\beta^g_1$, and $\pi$, we fit a GLM-EIV model with gene expressions $m$, gRNA counts $g$, gene offsets $\hat{f}^m := [\hat{f}^m_1, \dots, \hat{f}^m_n]^T$, gRNA offsets $\hat{f}^g := [\hat{f}^g_1, \dots, \hat{f}^g_n]^T$, and \textit{no} intercept or covariate terms.
+Intuitively, we ``encode'' all information about technical factors, library sizes, and baseline expression levels into $\hat{f}^m$ and $\hat{f}^g$.
+We run the algorithm $B \approx 15$ times over randomly-selected starting values for $\beta^m$, $\beta^g$, and $\pi$ and select the solution with greatest the log likelihood.
+
+The M step of the ``reduced'' GLM-EIV algorithm requires fitting two no-intercept, univariate GLMs with offsets (Algorithm \ref{algo:pilot_estimates}, line 20).
+We derive analytic formulas for the MLEs of these GLMs in the three most important cases: Gaussian response with identity link, Poisson response with log link, and negative binomial response with log link (see appendix; the latter formula is asymptotically exact, and, to the best of our knowledge, new).
+Consequently, we do not need to run the relatively slow IRLS procedure to carry out the M step of the ``reduced'' GLM-EIV algorithm.
+Overall, the proposed algorithm for obtaining the pilot parameter estimates (Algorithm \ref{algo:pilot_estimates}) requires fitting two GLMs (via IRLS).
+
+\subsection{Computation}
+We develop a computational infrastructure to apply GLM-EIV to large-scale, single-cell CRISPR screen data.\footnote{Pipeline available at \texttt{github.com/timothy-barry/glmeiv-pipeline}.} The infrastructure leverages Nextflow, a programming language that facilitates building data-intensive pipelines \cite{DITommaso2017}, and ondisc\footnote{Package available at \texttt{github.com/timothy-barry/ondisc}; preprint forthcoming.}, an R package that we developed (in a separate project) to facilitate large-scale computing on single-cell data.
+Nextflow and ondisc together enable the construction of highly portable single-cell pipelines: one can analyze data \textit{out-of-memory} on a laptop or in a \textit{distributed} fashion across tens or hundreds of nodes on a cloud (e.g., Microsoft Azure) or high-performance cluster.
+
+Leveraging these technologies, we develop a scalable and efficient pipeline for deploying GLM-EIV (Algorithm \ref{algo:at_scale}).
+First, we run a round of ``precomputations'' on all $d_g$ genes and $d_p$ perturbations.
+The precomputations involve regressing the gene expressions (or gRNA counts) onto the technical factors, thereby ``factoring out'' lines 2-7 from Algorithm \ref{algo:pilot_estimates}.
+Next, we run differential expression analyses on the full set of gene-perturbation pairs; for a given pair, this amounts to obtaining the set of pilot parameters (Algorithm \ref{algo:pilot_estimates}, lines 8 onward), fitting the GLM-EIV model (Algorithm \ref{algo:em_full}), and performing inference.
+The three loops in Algorithm \ref{algo:at_scale} are embarassingly parallel and therefore can be massively parallelized.
+
+\begin{algorithm}
+	\caption{Applying GLM-EIV at scale.}\label{algo:at_scale}
+	\begin{algorithmic}
+		\State $G \gets \{\textrm{gene}_1, \dots, \textrm{gene}_{d_g}\}; P \gets \{\textrm{perturbation}_1, \dots, \textrm{perturbation}_{d_p}\}$
+
+		\For{gene $\in G$}
+			\State Run precomputation (Algorithm \ref{algo:pilot_estimates}, lines 2-7) on gene. \State Save $\hat{f}^m$, $[\beta^m_0]^\textrm{pilot}$ and $[\gamma^T_m]^\textrm{pilot}$.
+		\EndFor
+		\For{perturbation $\in P$}
+		\State Run precomputation  (Algorithm \ref{algo:pilot_estimates}, lines 2-7) on perturbation.
+		\State Save $\hat{f}^g$, $[\beta^g_0]^\textrm{pilot}$ and $[\gamma^T_g]^\textrm{pilot}$.
+		\EndFor
+		\For{(gene, perturbation) $\in G \times P$}
+		\State Load $\hat{f}^m, \hat{f}^g,$ $[\beta^m_0]^\textrm{pilot}$ $[\gamma^T_m]^\textrm{pilot}$, $[\beta^g_0]^\textrm{pilot}$ and $[\gamma^T_g]^\textrm{pilot}$.
+		\State Compute $[\beta^m_1]^\textrm{pilot}, [\beta^g_1]^\textrm{pilot}, \pi^\textrm{pilot}$ (Algorithm \ref{algo:pilot_estimates}, lines 8 onward).
+		\State Run GLM-EIV using the pilot parameters (Algorithm \ref{algo:em_full}).
+		\EndFor
+	\end{algorithmic}
+\end{algorithm}
+
+\subsection{Zero-inflated model}
+
 \section{Simulation studies}
@@
 
-An \texttt{R} family object contains several functions, including \texttt{linkinv}, \texttt{variance}, and \texttt{mu.eta}.
-\texttt{linkinv} is the inverse link function $r_m^{-1}$.
-\texttt{variance} takes as an argument the mean $\mu^m_i$ and returns the variance $[\sigma_i^m]^2$.
-\texttt{mu.eta} is the derivative of the inverse link function $[r^{-1}_m]^{'}$.
-We extend the \texttt{R} family object by adding two additional functions: \texttt{skewness} and \texttt{mu.eta.prime}.
-\texttt{skewness} returns the skewness $\gamma^m_i$ of the distribution as a function of the mean $\mu_i$, i.e.
-$$\texttt{skewness}\left(\mu_i\right) = \E \left[\left(\frac{m_i - \mu_i^m}{ \sigma_i^m}\right)^3\right] := \gamma_i^m.$$ Finally, the \texttt{mu.eta.prime} is the second derivative of the inverse link function $[r^{-1}_m]''.$ Algorithm \ref{algo:computing_info_matrices} computes the matrices $\Delta^m(j)$, $[\Delta']^m(j)$, $V^m(j)$, $H^m(j)$, and vector $s^m(j)$ for given $\beta_m$ and given family object.
-(The vector $w^m(j)$ can be computed in terms of $\Delta^m(j)$ and $H^m(j)$.) We use $\sigma^m_i(j)$ (resp.
-$\gamma^m_i(j)$) to refer to the standard deviation (resp.
-skewness) of the gene expression distribution the $i$th cell when the perturbation $p_i$ is set to $j$.
-
 \begin{algorithm}
-	\caption{Computing the matrices $\Delta^m(j)$, $[\Delta']^m(j)$, $V^m(j)$, $H^m(j)$, and $s^m(j)$ for given $\beta_m$.}\label{algo:computing_info_matrices}
+	\caption{Computing the matrices $\Delta^m(j)$, $[\Delta']^m(j)$, $V^m(j)$, $H^m(j)$, and $s^m(j)$ given given $\beta_m$.}\label{algo:computing_info_matrices}
 	\begin{algorithmic}[3]
@@
 
+An \texttt{R} family object contains several functions, including \texttt{linkinv}, \texttt{variance}, and \texttt{mu.eta}.
+\texttt{linkinv} is the inverse link function $r_m^{-1}$.
+\texttt{variance} takes as an argument the mean $\mu^m_i$ and returns the variance $[\sigma_i^m]^2$.
+\texttt{mu.eta} is the derivative of the inverse link function $[r^{-1}_m]^{'}$.
+We extend the \texttt{R} family object by adding two additional functions: \texttt{skewness} and \texttt{mu.eta.prime}.
+\texttt{skewness} returns the skewness $\gamma^m_i$ of the distribution as a function of the mean $\mu_i$, i.e.
+$$\texttt{skewness}\left(\mu_i\right) = \E \left[\left(\frac{m_i - \mu_i^m}{ \sigma_i^m}\right)^3\right] := \gamma_i^m.$$ Finally, the \texttt{mu.eta.prime} is the second derivative of the inverse link function $[r^{-1}_m]''.$ Algorithm \ref{algo:computing_info_matrices} computes the matrices $\Delta^m(j)$, $[\Delta']^m(j)$, $V^m(j)$, $H^m(j)$, and vector $s^m(j)$ for given $\beta_m$ and given family object.
+(The vector $w^m(j)$ can be computed in terms of $\Delta^m(j)$ and $H^m(j)$.) We use $\sigma^m_i(j)$ (resp.
+$\gamma^m_i(j)$) to refer to the standard deviation (resp.
+skewness) of the gene expression distribution the $i$th cell when the perturbation $p_i$ is set to $j$.
+
 All steps of the algorithm are obvious except the calculation of $h'_m(l^m_i(j))$ (line 6), $h''(l^m_i(j))$ (line 9), and $V^m_i(j)$ (line 12).
@@
 \begin{table}
+\centering
 \caption{\texttt{linkinv}, \texttt{variance}, \texttt{mu.eta}, \texttt{skewness}, \texttt{mu.eta.prime} for common family objects (i.e., pairs of distributions and link functions).}\label{family_object_functions}
@@
 \section{Statistical accelerations}
+A key step in the algorithm for computing the pilot parameter estimates (Algorithm \ref{algo:pilot_estimates}) is to fit a weighted, no-intercept, univariate GLM with nonzero offset terms and a binary predictor variable (line 20).
+We derive an analytic formula for the MLE of this GLM for three important pairs of response distributions and link functions: Gaussian response with identity link, Poisson response with log link, and negative binomial response with log link.
+The GLM that we seek to estimate has responses $[m,m]^T$, predictors $[\underbrace{0, \dots, 0}_\textrm{n}, \underbrace{1, \dots, 1}_\textrm{n}]$, offsets $[\hat{f}^m, \hat{f}^m],$ and weights $w = [T_1(0), \dots, T_n(0), T_1(1), \dots, T_n(1)]^T.$ Throughout, $C$ denotes a universal constant.
+The log likelihood of this GLM is
+\begin{multline*}
+$$\mathcal{L}(\beta_1; m) = \sum_{i=1}^n T_i(0) f_m(m_i; h_m(\beta_1 + \hat{f}^m_i )) + \sum_{i=1}^n T_i(1) f_m(m_i; h_m(\hat{f}^m_i)) \\ = \sum_{i=1}^n T_i(1) f_m(m_i; h_m(\beta_1 + \hat{f}^m_i )) + C.$$
+\end{multline*}
+Thus, finding the MLE $\hat{\beta_1}$ is equivalent to estimating a GLM with intercept $\beta_1$, offsets $\hat{f}^m$, weights $T_i(1)$, and \textit{no} covariate terms.
+We term such a GLM a \textit{intercept-plus-offset} model.
+Below, we study intercept-plus-offset models in generality.
+
+\subsection*{Intercept-plus-offset models}
+Let $\beta \in \R$ be an unknown constant.
+Let $o_1, \dots, o_n \sim \mathcal{P}_1$, where $\mathcal{P}_1$ is a distribution.
+Let $Y_i|o_i, \dots, Y_n|o_i$ be exponential family-distributed random variables with identity sufficient statistic.
+Suppose the mean $\mu_i$ of $Y_i|o_i$ is given by $$r(\mu_i) = \beta + o_i,$$ where $r: \R \to \R$ is a strictly increasing, differentiable link function.
+We call this model the \textit{intercept-plus-offset} model.
+
+Next, let $w_1, \dots, w_n \sim \mathcal{P}_2$ be weights, where $\mathcal{P}_2$ is a distribution bounded above by $1$ and below by $0$.
+(A special case is $w_i = 1$ for all $i \in \{1, \dots, n\}$.) Throughout, we assume that $y_iw_i$ and $\exp(o_i)w_i$ have finite first moment.
+Suppose the cumulant-generating function and carrying density of the exponential family distribution are $\psi:\R \to \R$ and $c: \R \to \R$, respectively.
+The canonical parameter $\eta_i$ of the $i$th observation is
+\begin{equation}\label{can_param}
+\eta_i = ([\psi']^{-1} \circ r^{-1})(\beta + o_i) := h(\beta + o_i),
+\end{equation}
+ and the density $f$ of $Y_i | \eta_i$ is
+$$f(y_i; \eta_i) = \exp\{y_i \eta_i - \psi(\eta_i) + c(y_i)\}.$$ The weighted log likelihood is
+\begin{equation}\label{m_plus_o_mle}
+\mathcal{L}(\beta;y_i) = \sum_{i=1}^n w_i\log\left[f(y_i;\eta_i)\right] = C + \sum_{i=1}^n w_i(y_i \eta _i - \psi(\eta_i)).
+\end{equation}
+Our goal is to find the weighted MLE $\hat{\beta}$ of $\beta$.
+We consider three important choices for the exponential family distribution and link function, which amounts to maximizing (\ref{m_plus_o_mle}) for three specific choices of $r$ and $\psi$.
+In the first two cases -- Gaussian distribution with identity link and Poisson distribution with log link -- we find the \textit{finite-sample} maximizer of (\ref{m_plus_o_mle}); by contrast, in the third case -- negative binomial distribution with log link -- we find the \textit{asymptotic} maximizer.
+
+\subsubsection*{Gaussian}
+First, consider a Gaussian response distribution and identity link function $r(\mu) = \mu$. The cumulant-generating function $\psi$ is $\psi(\eta) = \eta^2/2$, and so, by (\ref{can_param}),
+$$h(t) = [\psi']^{-1}( r^{-1}(t)) = [\psi']^{-1}(t) = t.$$
+Plugging $\eta_i = h(\beta + o_i) = \beta + o_i$ and $\psi(\eta_i) = (1/2)(\beta + o_i)^2$ into (\ref{m_plus_o_mle}), we obtain
+$$\mathcal{L}(\beta; y) = \sum_{i=1}^n w_i (y_i(\beta + o_i) - (\beta + o_i)^2/2).$$ The derivative of this expression in $\beta$ is
+$$\frac{\partial \mathcal{L}(\beta;y)}{\partial\beta} = \sum_{i=1}^n w_i (y_i - \beta - o_i) = \sum_{i=1}^n w_i(y_i - o_i) - \beta \sum_{i=1}^n w_i.$$ Setting this quantity to 0 and solving for $\beta$, we find that the MLE $\hat{\beta}^\textrm{gauss}$ is
+$$\hat{\beta}^\textrm{gauss} = \frac{\sum_{i=1}^n w_i (y_i - o_i)}{\sum_{i=1}^n w_i}.$$
+
+\subsubsection*{Poisson}
+Next, consider a Poisson response distribution and log link function $r(\mu) = \log(\mu).$ The cumulant-generating function $\psi$ is $\psi(\eta) = e^\eta.$ Therefore, by (\ref{can_param}),
+$$h(t) = [\psi']^{-1}(r^{-1}(t)) = [\psi']^{-1} \left(\exp(t) \right) = \log(\exp(t)) = t.$$ Plugging $\eta_i = h(\beta + o_i) = \beta + o_i$ and $\psi(\eta_i) = \exp(\beta + o_i)$ into (\ref{m_plus_o_mle}), we obtain
+$$ \mathcal{L}(\beta; y) = \sum_{i=1}^n w_i \left( y_i(\beta + o_i) - \exp(\beta + o_i) \right).$$ The derivative of this function in $\beta$ is 
+$$\frac{\partial \mathcal{L}(\beta; y)}{\partial \beta} = \sum_{i=1}^n w_iy_i - w_i \exp(\beta + o_i) = \sum_{i=1}^n w_i y_i - \exp(\beta) \sum_{i=1}^n w_i \exp(o_i).$$
+Setting to zero and solving for $\beta$, we find that the MLE $\hat{\beta}^\textrm{pois}$ is
+\begin{equation}\label{pois_mle}
+\hat{\beta}^\textrm{pois} = \log\left(\frac{\sum_{i=1}^n w_i y_i}{\sum_{i=1}^n w_i e^{o_i}}\right).
+\end{equation}
+
+\subsubsection*{Negative binomial}
+Finally, we consider a negative binomial response distribution (with fixed size parameter $\theta > 0$) and log link function $r(\mu) = \log(\mu)$. The cumulant-generating function $\psi$ is
+$\psi(\eta) = -\theta \log(1 - e^\eta).$ The derivative $\psi'$ of $\psi$ is
+$$ \psi'(t) = \theta \left(\frac{e^t}{1 - e^t}\right) = \frac{\theta}{e^{-t} - 1}.$$ Define the function $\delta: \R \to \R$ by $\delta(t) = -\log\left(\theta/t + 1 \right).$ We see that
+$$\psi'(\delta(t)) = \frac{\theta}{\exp\left(\log(\theta/t + 1 )\right) - 1} = t,$$ and so $\delta = [\psi']^{-1}.$ By (\ref{can_param}), we have that
+$$
+h(t) = [\psi']^{-1}(r^{-1}(t)) = -\log\left(\frac{\theta}{\exp(t)} + 1 \right) = \log\left(\frac{\exp(t)}{\theta + \exp(t)}\right).
+$$
+Therefore,
+\begin{multline}\label{nb_mo_1}
+\eta_i = h(\beta + o_i) = \log\left(\frac{\exp(\beta+o_i)}{\theta + \exp(\beta + o_i)} \right) = \beta + o_i - \log\left(\theta + e^{\beta}e^{o_i}\right) \\ = \beta - \log\left(\theta + e^{\beta}e^{o_i} \right) + C,
+\end{multline}
+ and
+ \begin{multline}\label{nb_mo_2}
+ \psi(\eta_i) = -\theta\log\left(1 - \frac{\exp(\beta+o_i)}{\theta + \exp(\beta + o_i)} \right) = -\theta \log \left(\frac{\theta}{\theta + \exp(\beta + o_i)} \right) \\ = -\theta \log (\theta) + \theta \log[\theta + \exp(\beta + o_i)] = \theta \log(\theta + e^{\beta}e^{o_i}) + C.
+ \end{multline}
+ Plugging (\ref{nb_mo_1}) and (\ref{nb_mo_2}) into (\ref{m_plus_o_mle}), the log-likelihood (up to a constant) is
+\begin{multline*}
+\mathcal{L}(\beta; y) = \beta \sum_{i=1}^n w_i y_i - \sum_{i=1}^n w_i y_i \log(\theta + e^\beta e^{o_i}) - \theta \sum_{i=1}^n w_i \log(\theta + e^\beta  e^{o_i}) \\ = \beta \sum_{i=1}^n w_i y_i - \sum_{i=1}^n (y_i + \theta)w_i\log(\theta + e^\beta e^{o_i}).
+\end{multline*}
+The derivative of $\mathcal{L}$ in $\beta$ is
+$$\frac{\partial \mathcal{L}(\beta;y)}{\partial \beta} = \sum_{i=1}^n w_i y_i  - \sum_{i=1}^n \frac{w_i(y_i + \theta) e^{\beta} e^{o_i}}{ \theta + e^{\beta} e^{o_i}}.$$
+Setting the derivative to zero, the MLE equation is
+\begin{equation}\label{nb_mle}
+e^\beta \sum_{i=1}^n \frac{w_i e^{o_i} (y_i + \theta)}{e^\beta e^{o_i} + \theta} = \sum_{i=1}^n w_i y_i.
+\end{equation}
+We cannot solve for $\beta$ in (\ref{nb_mle}) analytically. However, we can derive an asymptotically exact solution. By the law of total expectation,
+$$ \E \left[\frac{w_i  e^{o_i} (y_i + \theta)}{e^{\beta + o_i} + \theta} \right] = \E \left[\E\left[\frac{w_i e^{o_i} (y_i + \theta) }{e^{\beta + o_i} + \theta} \bigg| (o_i, w_i) \right] \right] = \E \left[\frac{w_i e^{o_i} (e^{\beta + o_i} + \theta)}{ e^{\beta + o_i} + \theta} \right] = \E [w_i e^{o_i}];
+$$
+the second equality holds because $\E[y_i | o_i ] = \mu_i = e^{\beta + o_i}.$ Dividing by $n$ on both sides of (\ref{nb_mle}) and rearranging,
+\begin{equation}\label{nb_mo_3}
+\beta = \log\left( \frac{ (1/n) \sum_{i=1}^n w_i e^{o_i} (y_i + \theta)/(e^\beta e^{o_i} + \theta)}{ (1/n) \sum_{i=1}^n w_i y_i. } \right).
+\end{equation}
+By weak LLN, the limit (in probability) of the MLE $\hat{\beta}^\textrm{NB}$ is
+\begin{equation}
+\hat{\beta}^{(\textrm{NB})} \xrightarrow{P} \log\left(\frac{\E[w_i y_i]}{\E[w_i e^{o_i}]} \right).
+\end{equation}
+But the Poisson MLE $\hat{\beta}^{(\textrm{Pois})}$ (\ref{pois_mle}) converges in probability to the same limit:
+$$ \hat{\beta}^{(\textrm{pois})} =  \log \left(\frac{ (1/n) \sum_{i=1}^n w_i y_i}{(1/n)\sum_{i=1}^n w_i e^{o_i}} \right) \xrightarrow{P} \log \left(\frac{\E[w_i y_i]}{ \E[w_i e^{o_i}]} \right).$$ Therefore, for large $n$, we can approximate $\hat{\beta}^{\textrm{(NB)}}$ by $\hat{\beta}^{(\textrm{pois})}$.
 

### 2021-11-14 text/outline/draft.tex
complete draft of all math content

@@
 \makeatother
-\setcounter{tocdepth}{1}
+\setcounter{tocdepth}{2}
 \usepackage{xcolor}
@@
 \begin{center} 
-\textbf{CRISPR genome editing, single-cell sequencing, and exponential family measurement error models}
+\textbf{A new class of exponential family measurement error models, with application to single-cell CRISPR screens}
 \end{center}
@@
 \tableofcontents
-
+\newpage
 \section{Introduction}
@@
 Our work makes two main contributions.
-First, we study the thresholding method from empirical and theoretical perspectives.
+First, we conduct a careful study of the thresholding method.
 Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian model.
@@
 SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification.
-Other applied methods for single-cell CRISPR screen analysis include MIMOSCA \cite{Dixit2016} and scMAGeCK \cite{Yang2019}.
+% Other applied methods for single-cell CRISPR screen analysis include MIMOSCA \cite{Dixit2016} and scMAGeCK \cite{Yang2019}.
 These methods, like SCEPTRE, focus on hypothesis testing (rather than estimation), but unlike SCEPTRE, they ignore the countedness of the data and are unable to handle confounders.
-In this work we tackle a set of analysis challenges that are complimentary to the challenges addressed by SCEPTRE.
-Most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, \blue{an unattainable objective within the nonparameteric hypothesis testing framework of SCEPTRE.}
+In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE.
+Most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, \blue{an objective that is unattainable within the nonparameteric hypothesis testing framework of SCEPTRE.}
 
@@
 \textbf{b,} DAG representing all variables in the system.
-The perturbation (latent) impacts both gene expression and gRNA expression; technical factors act as nuisance variables, also impacting gene and gRNA expression.
+The perturbation (latent) impacts both gene expression and gRNA expression; technical factors act as confounders, also impacting gene and gRNA expression.
 The target of estimation is the effect of the perturbation on gene expression.
@@
 \begin{equation}\label{marginal_density}
-f(m_i, g_i; \theta) = (1-\pi) f(m_i; \eta^m_i(0)) f(g_i; \eta^g_i(0)) + \pi f(m_i; \eta^m_i(1)) f(g_i; \eta^g_i(1)).
+f(m_i, g_i; \theta) = (1-\pi) f_m(m_i; \eta^m_i(0)) f_g(g_i; \eta^g_i(0)) + \pi f_m(m_i; \eta^m_i(1)) f_g(g_i; \eta^g_i(1)).
 \end{equation}
@@
 \begin{multline}\label{marginal_log_lik}
-\mathcal{L}(\theta; m, g) \\ = \sum_{i=1}^n \log\left[ (1-\pi) f(m_i; \eta^m_i(0)) f(g_i; \eta^g_i(0)) + \pi f(m_i; \eta^m_i(1)) f(g_i; \eta^g_i(1)) \right]
+\mathcal{L}(\theta; m, g) \\ = \sum_{i=1}^n \log\left[(1-\pi) f_m(m_i; \eta^m_i(0)) f_g(g_i; \eta^g_i(0)) + \pi f_m(m_i; \eta^m_i(1)) f_g(g_i; \eta^g_i(1)) \right]
 \end{multline}
@@
 Leveraging this analytic formula, we can calculate standard errors (and $p$-values and confidence intervals) quickly, enabling us to perform inference in practice on real, large-scale data.
+The derivation is fairly involved; simulation studies (subsequent section) confirm the correctness of the results.
 
@@
 
-Leveraging these technologies, we develop a scalable and efficient pipeline for deploying GLM-EIV (Algorithm \ref{algo:at_scale}).
+Leveraging these technologies, we develop a scalable and efficient pipeline for GLM-EIV (Algorithm \ref{algo:at_scale}).
 First, we run a round of ``precomputations'' on all $d_g$ genes and $d_p$ perturbations.
 The precomputations involve regressing the gene expressions (or gRNA counts) onto the technical factors, thereby ``factoring out'' lines 2-7 from Algorithm \ref{algo:pilot_estimates}.
-Next, we run differential expression analyses on the full set of gene-perturbation pairs; for a given pair, this amounts to obtaining the set of pilot parameters (Algorithm \ref{algo:pilot_estimates}, lines 8 onward), fitting the GLM-EIV model (Algorithm \ref{algo:em_full}), and performing inference.
+Next, we run differential expression analyses on the full set of gene-perturbation pairs; for a given pair, this amounts to obtaining the complete set of pilot parameters (Algorithm \ref{algo:pilot_estimates}, lines 8 onward), fitting the GLM-EIV model (Algorithm \ref{algo:em_full}), and performing inference.
 The three loops in Algorithm \ref{algo:at_scale} are embarassingly parallel and therefore can be massively parallelized.
@@
 \subsection{Zero-inflated model}
+The GLM-EIV model assumes that unperturbed cells exhibit a nonzero ``background read'' gRNA count distribution due to errors in the sequencing and alignment processes.
+This assumption very likely is appropriate for most current single-cell CRISPR screen datasets \cite{Replogle2020}.
+However, background contamination may be negligible for some datasets, in which case the gRNA counts of unperturbed cells should be modeled as point masses at zero.
+We introduce such an extension -- which we call the ``zero-inflated GLM-EIV model'' -- in Appendix \ref{sec:zero_inf_model}.
+We derive efficient methods for estimation and inference in this model similar to those outlined above.
+However, given the ubiquity of background contamination in current single-cell CRISPR screen datasets, we focus our attention on the standard (or ``background read'') GLM-EIV model for the remainder of the main text.
 
@@
 \section{Discussion}
+
+\bibliographystyle{unsrt}
+\newpage
+\bibliography{/Users/timbarry/optionFiles/glmeiv.bib}
 
@@
 $$ Similarly, let $\left[\eta^g_i(k)\right]^{(t)}$ be defined by
-$$\left[\eta^g_i(k)\right]^{(t)} :=  h_m\left( \langle \tilde{x}_i(k) , \beta_g^{(t)} \rangle + o^g_i \right).$$
+$$\left[\eta^g_i(k)\right]^{(t)} :=  h_g\left( \langle \tilde{x}_i(k) , \beta_g^{(t)} \rangle + o^g_i \right).$$
 Next, for $k \in \{0,1\},$ define $\alpha^{(t)}_i(k)$ by
@@
 The $i$th membership probability $T^{(t)}_i(1)$ is
-\begin{multline*}
-T^{(t)}_i(1) = \P(P_i = 1 | M_i = m_i, G_i = g_i, \theta^{(t)}) \\ = \frac{\pi^{(t)}(1) \alpha^{(t)}_i(1)}{ \sum_{k=0}^1 \pi^{(t)}(k) \alpha^{(t)}_i(k)} \textrm{ (by Bayes rule)} = \frac{1}{\frac{ \pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)} + 1} \\ = \frac{1}{ \exp\left(\log\left(\frac{\pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)}\right)\right) + 1} = \frac{ 1 }{ \exp\left(q^{(t)}_i\right) + 1},
-\end{multline*}
-where we set $$q_i^{(t)} := \log\left(\frac{\pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)}\right).$$
+\begin{multline}\label{e_step_1}
+T^{(t)}_i(1) = \P(P_i = 1 | M_i = m_i, G_i = g_i, \theta^{(t)})  = \frac{\pi^{(t)}(1) \alpha^{(t)}_i(1)}{ \sum_{k=0}^1 \pi^{(t)}(k) \alpha^{(t)}_i(k)} \textrm{ (by Bayes rule)} \\ = \frac{1}{\frac{ \pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)} + 1} = \frac{1}{ \exp\left(\log\left(\frac{\pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)}\right)\right) + 1} = \frac{ 1 }{ \exp\left(q^{(t)}_i\right) + 1},
+\end{multline}
+where we set 
+\begin{equation}\label{e_step_2}
+q_i^{(t)} := \log\left(\frac{\pi^{(t)}(0) \alpha_i^{(t)}(0)}{\pi^{(t)}(1) \alpha_i^{(t)}(1)}\right).
+\end{equation}
 Next, we have that
@@
 For $i \in \{1, \dots, n\}$, $j \in \{0, 1\}$, and $\theta = (\pi, \beta_m, \beta_g),$ let $T^\theta_i(j)$ be defined by
-$$T^\theta_i(j) = \P_\theta\left(P_i = j | M_i = m_i, G_i = g_i\right).$$ Also, define the $\R^n$ vectors $T^\theta(0)$ and $T(1)$ by
-$$T^\theta(0) := [T^\theta_1(0), T^\theta_2(0), \dots, T^\theta_n(0)]$$ and 
-$$T^\theta(1) := [T^\theta_1(1), T^\theta_2(1), \dots, T^\theta_n(1)].$$
-
+$$T^\theta_i(j) = \P_\theta\left(P_i = j | M_i = m_i, G_i = g_i\right).$$ Let the $n \times n$ matrix $T^\theta(j)$ be given by $$T^\theta(j) = \textrm{diag}\left\{T^\theta_1(j), \dots, T^\theta_n(j)\right\}.$$
 Next, define the diagonal $n \times n$ matrices $\Delta^m$, $[\Delta']^m$, $V^m$, and $H^m$ by
@@
 \begin{cases}
-	\Delta^m = \textrm{diag} \{ h_m'(l_1^m), \dots, h_m'( l_n^m ) \} \\
-	[\Delta']^m = \textrm{diag} \{ h_m''(l_1^m), \dots, h_m''( l_n^m) \} \\
+	\Delta^m = \textrm{diag} \{h_m'(l_1^m), \dots, h_m'(l_n^m)\} \\
+	[\Delta']^m = \textrm{diag} \{h_m''(l_1^m), \dots, h_m''(l_n^m) \} \\
 	V^m = \textrm{diag} \{ \psi''_m( \eta^m_1), \dots, \psi''_m( \eta^m_n) \} \\
@@
 
-\section{Zero-inflated model}
-\subsection{Relationship to background real model}
+\section{Zero-inflated model}\label{sec:zero_inf_model}
+In this section we introduce the ``zero-inflated'' GLM-EIV model.
+The zero-inflated GLM-EIV model is appropriate to use when the unperturbed cells do not transcribe \textit{any} gRNA molecules (i.e., when there are no background reads).
+Let $x_i = [1, z_i]^T \in \R^{d-1}$ be the vector of observed covariates, including an intercept term.
+($x_i$ is the same as $\tilde{x}_i$, but with the perturbation indicator $p_i$ removed.) Let $\beta_{g,z} = [\beta^g_0, \gamma_g] \in \R^{d-1}$ be an unknown coefficient vector.
+($\beta_{g,z}$ is the same as $\beta_g$, but with the perturbation effect $\beta^g_1$ removed).
+Let the linear component $l^{g,z}_i$, mean $\mu^{g,z}_i$, and canonical parameter $\eta^{g,z}_i$ of gRNA count distribution of the $i$th cell be given by
+$$\begin{cases}
+l^{g,z}_i = \langle x_i, \beta_{g,z} \rangle + o^g_i \\
+r_g(\mu^{g,z}_i) = l^{g,z}_i \\
+\eta^{g,z}_i = ([\psi'_g]^{-1} \circ r^{-1}_g)(l^{g,z}_i) := h_g(l^{g,z}_i).
+\end{cases}
+$$
+The density $f_{g,z}$ of gRNA counts in the zero-inflated model is as follows:
+$$f_{g,z}(g_i; \eta^{g,z}_i, p_i) = [f_g(g_i;\eta^{g,z}_i)]^{p_i} \mathbb{I}(g_i = 0)^{1-p_i}.$$
+In other words, when the cell is \textit{perturbed} (i.e., $p_i = 1$), the zero-inflated density $f_{g,z}$ coincides with the background-read density $f_g$; by contrast, when the cell is \textit{unperturbed} (i.e., $p_i = 0$), the zero-inflated density $f_{g,z}$ is a point mass at zero.
+The gene expression density $f_m$ and perturbation indicator density $f_p$ are the same across the background read and zero-inflated models.
+We assume that the gene expression $m_i$ and gRNA count $g_i$ are conditionally independent given the perturbation indicator $p_i$.
+The joint density $f_z$ of $(m_i, p_i, z_i)$ is
+\begin{multline*}
+f_z(m_i, g_i, p_i) = f_m(m_i | p_i) f_{g,z}(g_i | p_i) f_p(p_i) \\ = \pi^{p_i} (1-\pi)^{1-p_i} f_m(m_i; \eta^m_i)[f_g(g_i; \eta_i^{g,z})]^{p_i} \mathbb{I}(g_i = 0)^{1-p_i}.
+\end{multline*}
+The log-likelihood $\mathcal{L}_z$ is
+\begin{multline*}
+\mathcal{L}_z(\theta; m, g, p) = \sum_{i=1}^n \log \left[ \pi^{p_i} (1-\pi)^{1-p_i} \right] + \sum_{i=1}^n \log\left[f_m(m_i; \eta^m_i) \right] \\ + \sum_{i=1}^n p_i \log \left[f_{g}(g_i; \eta_i^{g,z}) \right] + \sum_{i=1}^n (1 - p_i)\log\left[ \mathbb{I}(g_i = 0) \right],
+\end{multline*}
+where $\theta = [\pi, \beta_m, \beta_{g,z}]$ is the vector of unknown parameters.
+Integrating over the unobserved variable $p_i$, the marginal density $f_z$ of $(m_i, g_i)$ is
+$$f_z(m_i, g_i; \theta) = (1-\pi) f_m(m_i;\eta^m_i(0)) \mathbb{I}(g_i = 0) + \pi f_m(m_i; \eta^m_i(1)) f_g(g_i;\eta^{g,z}_i).$$ Finally, the marginal log-likelihood is
+$$
+\mathcal{L}_z(\theta; m_i, g_i) = \sum_{i=1}^n \log\left[ (1-\pi) f_m(m_i;\eta^m_i(0)) \mathbb{I}(g_i = 0) + \pi f_m(m_i; \eta^m_i(1)) f_g(g_i;\eta^{g,z}_i) \right].
+$$
+
 \subsection{Estimation}
+To estimate the parameters of the zero-inflated GLM-EIV model, we use an EM algorithm similar to Algorithm \ref{algo:em_full} but with two changes.
+First, we use a different formula for the $i$th membership probability at the $t$-th step of the algorithm $T^{(t)}_i(1)$.
+(We use $T^{(t)}_i(1)$ to denote the $i$th membership probability in \textit{both} the background read and zero inflated cases; the difference should be clear from context.) Let $\theta^{(t)} = (\pi^{(t)}, \beta^{(t)}_m, \beta^{(t)}_{g,z})$ be the parameter estimate at the $t$-th iteration of the algorithm.
+Arguing in a manner similar to the background read case, we have that $$T^{(t)}_i(1) = \frac{1}{ \exp(q_i^{\left(t,z\right)}) + 1},$$ where
+$$q_i^{(t,z)} = \log \left(\frac{(1 - \pi^{(t)}) \P(M_i = m_i | P_i = 0, \theta^{(t)}) \P(G_i = g_i | P_i = 0, \theta^{(t)})}{(\pi^{(t)}) \P(M_i = m_i | P_i = 1, \theta^{(t)}) \P(G_i = g_i | P_i = 1, \theta^{(t)})}\right).$$
+The expression for $q^{(t,z)}_i$ is
+\begin{multline*}
+q^{(t,z)}_i = \log\left[ 1 - \pi^{(t)} \right] + \log\left[ f_m\left(m_i; \left[ \eta^m_i(0) \right]^{(t)}\right) \right] + \log\left[ \mathbb{I}(g_i = 0) \right] \\ - \log\left[ \pi^{(t)} \right] - \log\left[ f_m\left(m_i; \left[ \eta^m_i(1) \right]^{(t)}\right) \right] - \log\left[ f_g\left(g_i; \left[\eta^{g,z}_i \right]^{(t)}\right) \right],
+\end{multline*}
+where $[\eta^{g,z}_i]^{(t)} = h_g( \langle x_i, \beta^{(t)}_{g,z} \rangle + o^g_i).$ Notice that if $g_i \geq 1$, then $T^{(t)}_i(1) = 1.$ This comports with our intuition that a nonzero gRNA count indicates the presence of a perturbation.
+
+Next, we consider the M step of the EM algorithm, which is similar to the background read case but slightly different.
+Define $Q_z(\theta | \theta^{(t)}) = \E_{(P | M = m, G = g, \theta^{(t)})} \left[ \mathcal{L}_z(\theta; m, g, p) \right].$ We have that
+\begin{multline}\label{q_funct_zero_inf}
+Q_z(\theta | \theta^{(t)}) = \sum_{i=1}^n \left[ T^{(t)}_i(1)\log(\pi) + T_i^{(t)}(0) \log(1 - \pi)\right] + \sum_{i=1}^n \sum_{j=0}^1 T^{(t)}_i(j) \log \left[ f_m( m_i; \eta_i^m(j)) \right] \\ + \sum_{i=1}^n T_i^{(t)}(1) \left[ \log( f_g(g_i; \eta_i^{g, z}))\right] + C.
+\end{multline}
+The three terms of (\ref{q_funct_zero_inf}) are functions of $\pi$, $\beta_m$, and $\beta_{g,z}$, respectively.
+The maximizer $\pi^{(t)}$ and $\beta_m^{(t+1)}$ of the first and second term are the same as in the background read case.
+The maximizer $\beta^{(t+1)}_{g,z}$ of the third term is the maximizer of the GLM with exponential family density $f_g$, link function $r_g$, responses $g$, weights $T^{(t)}(1)$, design matrix $X$, offsets $o^g$.
+
 \subsection{Inference}
+Next, we derive the asymptotic observed information matrix for the zero-inflated model, allowing us to perform inference.
+Again, let $T^{\theta}(1) := \textrm{diag}\{T^{\theta}_1(1), \dots, T^{\theta}_n(1)\},$ but note that $T^\theta_i(1) = \P(P_i = 1 | G_i = g_i, M_i = m_i, \theta)$ is computed differently than in the background read case.
+Define the $n \times n$ matrices $\Delta^{(g,z)}, [\Delta']^{(g,z)}, V^{(g,z)},$ and $H^{(g,z)}$ by
+$$
+\begin{cases}
+\Delta^{(g,z)} = \textrm{diag} \{ h_g'(l_1^{g,z} ), \dots, h_g'( l_n^{g,z}  ) \} \\
+[\Delta']^{(g,z)}  = \textrm{diag} \{ h_g''(l_1^{g,z} ), \dots, h_g''( l_n^{g,z} ) \} \\
+V^{(g,z)}  = \textrm{diag} \{ \psi_g( \eta^{g,z} _1), \dots, \psi_g( \eta^{g,z}_n) \} \\
+H^{(g,z)}  = \textrm{diag} \{ m_1 - \mu_1^{g,z} , \dots, m_n - \mu_n^{g,z} \} .
+\end{cases}
+$$
+Also, define the $\R^n$ vectors $s^{(g,z)}$ and $w^{(g,z)}$ by
+$$s^{(g,z)} = [g_1 - \mu_1^{g,z}, \dots, g_n - \mu_n^{g,z}]^T,$$ and $$w^{(g,z)} = [T^\theta_1(0)T^\theta_1(1) \Delta_1^{(g,z)}H^{(g,z)}_1,\dots, T^\theta_n(0)T^\theta_n(1) \Delta_n^{(g,z)}H^{(g,z)}_n].$$ These quantities are computable, as they do not depend on the unobserved variables $p_1, \dots, p_n$.
+Finally, let the unobserved, $n\times n$ matrix $P$ be defined by $P = \textrm{diag}\{ p_1, \dots, p_n \}$.
+
+The observed information matrix $J_z(\theta; m, g)$ is given by
+$$ J_z(\theta; m, g) = -\nabla^2 \mathcal{L}_z(\theta; m, g).$$ Louis's theorem implies that
+\begin{multline*}
+J_z(\theta; m, g) = -\E \left[\nabla^2 \mathcal{L}_z(\theta; m, g, p) | G = g, M = m \right] \\ + \E\left[\nabla \mathcal{L}_z(\theta; m, g, p) | G = g, M = m \right] \E\left[\nabla \mathcal{L}_z(\theta; m, g, p) | G = g, M = m \right]^T \\ - \E\left[\nabla\mathcal{L}_z(\theta; m, g, p) \nabla \mathcal{L}_z(\theta; m, g, p)^T | G = g, M = m \right].
+\end{multline*}
+The matrix $J_z(\theta; m, g)$ has dimension $d \times d$ and consists of nine submatrices (Figure \ref{infomatrixzeroinf}).
+Three of these submatrices (i.e., I, II, and V) are the same as the corresponding submatrices in the background read case.
+We therefore must compute the remaining submatrices (i.e., III, IV, and VI) to compute the entire matrix $J_z(\theta; m, g)$.
+Again, in the following, all expectations are understood to be conditional on $m$ and $g$.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\linewidth]{../../figures/info_matrix/info_matrix_zero_inf}
+	\caption{Block structure of the observed information matrix $J_z(\theta; m, g) = -\nabla^2 \mathcal{L}_z(\theta; m, g)$ for the zero-inflated model.
+Submatrices I, II, and VI are the same as in the background read model; therefore, we only need to compute submatrices III, VI, and V.}
+	\label{infomatrixzeroinf}
+\end{figure}
+
+\subsubsection*{Submatrix III (zero-inflated)}
+Denote submatrix III by $J_{\beta_{(g,z)}}(\theta; m, g)$ The formula for $J_{\beta_{(g,z)}}(\theta; m, g)$ is
+\begin{multline}\label{sub_mat_3_zeroinf}
+J_{\beta_{(g,z)}}(\theta; m, g) = -\E \left[\nabla_{\beta_{(g,z)}}^2 \mathcal{L}_z(\theta; m, g, p) \right] \\ + \E\left[ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right] \E\left[\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right]^T \\ - \E\left[ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p)^T \right].
+\end{multline}
+GLM theory indicates that
+$$ -\nabla_{\beta_{(g,z)}}^2 \mathcal{L}_z(\theta; m, g, p) =  X^T P ( \Delta^{(g,z)}V^{(g,z)}\Delta^{(g,z)} - (\Delta')^{(g,z)} H^{(g,z)}) X$$ and
+$$ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) = X^T P \Delta^{(g,z)} s^{(g,z)}.$$
+We begin by computing the first term of (\ref{sub_mat_3_zeroinf}).
+The only random matrix among $X,$ $P$, $\Delta^{(g,z)}$, $V^{(g,z)}$, $(\Delta')^{(g,z)},$ and $H^{(g,z)}$ is $P$.
+Therefore, by the linearity of expectation,
+\begin{multline}\label{sub_mat_3_zeroinf_1}
+-\E \left[\nabla_{\beta_{(g,z)}}^2 \mathcal{L}_z(\theta; m, g, p) \right]   = \E \left[ X^T P (\Delta^{(g,z)}V^{(g,z)}\Delta^{(g,z)} - (\Delta')^{(g,z)} H^{(g,z)} \right] \\ = X^T T^{\theta}(1) ( \Delta^{(g,z)}V^{(g,z)}\Delta^{(g,z)} - (\Delta')^{(g,z)} H^{(g,z)}) X.
+\end{multline}
+Next, we compute the difference of the last two terms of (\ref{sub_mat_3_zeroinf}). The $(k,l)$th entry of this matrix is
+\begin{multline*}
+\bigg[ \E \left[\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right] \E \left[\nabla_{\beta{(g,z)}} \mathcal{L}_z(\theta; m, g, p)\right]^T \\ - \E \left[ \nabla_{{\beta_{(g,z)}}} \mathcal{L}_z(\theta; m, g, p) \nabla_{{\beta_{(g,z)}}} \mathcal{L}_z(\theta; m, g, p)^T \right] \bigg] [k,l] \\
+= \left[ \E \left[ {X}^T P \Delta^{(g,z)} s^{(g,z)} \right]\E \left[ {X}^T P \Delta^{(g,z)} s^{(g,z)} \right]^T\right][k,l] - \E \left[ {X}^T P \Delta s^{(g,z)} (s^{(g,z)})^T \Delta^{(g,z)} P {X}^T \right][k,l] \\
+= \E\left[{X}[,k]^T P \Delta^{(g,z)} s^{(g,z)} \right] \E \left[ {X}[,l]^T P \Delta^{(g,z)} s^{(g,z)} \right] - \E \left[ {X}[,k]^T P \Delta^{(g,z)} s^{(g,z)} (s^{(g,z)})^T \Delta^{(g,z)} P {X}[,l ] \right] \\
+=\E\left(\sum_{i=1}^n {x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_{i} \right) \E \left( \sum_{j=1}^n {x}_{jl} P_j \Delta^{(g,z)}_j s^{(g,z)}_j \right) \\ - \E \left( \sum_{i=1}^n \sum_{j=1}^n {x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i s^{(g,z)}_j \Delta^{(g,z)}_j P_j{x}_{jl} \right) \\
+= \sum_{i=1}^n \sum_{j=1}^n \E[ {x}_{ik} P_i \Delta^{(g,z)}_is^{(g,z)}_i] \E [ {x}_{jl} P_j \Delta^{(g,z)}_j s^{(g,z)}_j]  -  \sum_{i=1}^n \sum_{j=1}^n \E [ {x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i s^{(g,z)}_j \Delta^{(g,z)}_jP_j {x}_{jl}  ]  \\
+= \sum_{i=1}^n \sum_{j=1}^n \E[ {x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i ] \E \left[{x}_{jl} P_j \Delta^{(g,z)}_j s^{(g,z)}_j \right]  - \sum_{i \neq j} \E [{x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i] \E[s^{(g,z)}_j P_j \Delta^{(g,z)}_j {x}_{jl}] \\ - \sum_{i=1}^n \E[ {x}_{ik}P_i \Delta^{(g,z)}_i s^{(g,z)}_i s^{(g,z)}_i \Delta^{(g,z)}_i P_i{x}_{il}] \\
+= \sum_{i=1}^n \E[ {x}_{ik} P_i \Delta^{(g,z)}_i H^{(g,z)}_i] \E[{x}_{il}P_i\Delta^{(g,z)}_i H^{(g,z)}_i] - \sum_{i=1}^n \E[{x}_{ik}P_i^2(\Delta^{(g,z)}_i)^2 (H^{(g,z)}_i)^2 {x}_{il}] \\ = \sum_{i=1}^n {x}_{ik} T_i^{\theta}(1)^2(\Delta^{(g,z)}_i)^2 (H_i^{(g,z)})^2{x}_{il} - \sum_{i=1}^n {x}_{ik} T_i^{\theta}(1) (\Delta_i^{(g,z)})^2 (H_i^{(g,z)})^2 {x}_{il} \\ = {X}[,k]^T T^{\theta}(1)^2 (\Delta^{(g,z)})^2 (H^{(g,z)})^2 {X}[,l] - {X}[,k]^T T^{\theta}(1) (\Delta^{(g,z)})^2 (H^{(g,z)})^2 {X}[,l]
+\end{multline*}
+Therefore, we have that
+\begin{multline}\label{sub_mat_3_zeroinf_2}
+\E\left[ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right] \E\left[\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right]^T - \E\left[ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p)^T \right] \\ =
+X^T T^{\theta}(1)^2 (\Delta^{(g,z)})^2(H^{(g,z)})^2{X} - {X}^T T^{\theta}(1) (\Delta^{(g,z)})^2 (H^{(g,z)})^2 {X} \\ = - X^T T^{\theta}(1) \left(\Delta^{(g,z)}\right)^2 \left( H^{(g,z)} \right)^2 \left( I - T^{\theta}(1)  \right) X.
+\end{multline}
+Combining (\ref{sub_mat_3_zeroinf}), (\ref{sub_mat_3_zeroinf_1}), and (\ref{sub_mat_3_zeroinf_2}), we conclude that
+\begin{multline}\label{sub_mat_3_zeroinf_formula}
+J_{\beta_{(g, z)}} = \left( \theta; m, g \right) =  X^T T^{\theta}(1) ( \Delta^{(g,z)}V^{(g,z)}\Delta^{(g,z)} - (\Delta')^{(g,z)} H^{(g,z)}) X \\ - X^T T^{\theta}(1) \left(\Delta^{(g,z)}\right)^2 \left( H^{(g,z)} \right)^2 \left( I - T^{\theta}(1) \right) X.
+\end{multline}
+
+\subsubsection*{Submatrix IV (zero-inflated)}
+Denote submatrix IV by $J_{(\beta_{(g,z)}, \beta_m)}(\theta; m, g)$. The formula for submatrix IV is
+\begin{multline}\label{sub_mat_4_zeroinf}
+J_{(\beta_{(g,z)}, \beta_m)}(\theta; m, g) = - \E \left[\nabla_{\beta_{(g,z)}} \nabla_{\beta_m} \mathcal{L}_z (\theta; m, g, p) \right] \\ + \E\left[ \nabla_{\beta_{(g,z)} }\mathcal{L}_z(\theta ; m,g,p) \right]\E\left[\nabla_{\beta_m}\mathcal{L}_z(\theta; m,g,p)\right]^T \\ - \E \left[\nabla_{\beta_{(g,z)}}\mathcal{L}_z (\theta; m,g,p) \nabla_{\beta_m}\mathcal{L}_z(\theta; m,g,p)]^T \right].
+\end{multline} 
+First, we have that
+\begin{equation}\label{sub_mat_4_zeroinf_1}
+- \E \left[\nabla_{\beta_{(g,z)}} \nabla_{\beta_m} \mathcal{L}_z (\theta; m, g, p) \right] = 0,
+\end{equation}
+ as the derivative in $\beta_m$ of $\mathcal{L}_z(\theta; m, g, p)$ is a function of $\beta_m$, and the derivative in $\beta_{(g,z)}$ of this term is $0$.
+Next, we compute the difference of the last two terms of (\ref{sub_mat_4_zeroinf}).
+Entry $(k,l)$ of this matrix is
+\begin{multline}\label{sub_mat_4_zeroinf_2}
+[\E[\nabla_{\beta_{(g,z)}}\mathcal{L}_z(\theta; m, g, p) ] \E [\nabla_{\beta_m} \mathcal{L}_z(\theta; m, g, p) ]^T \\ - \E [\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \nabla_{\beta_m}\mathcal{L}_z(\theta; m, g, p)^T ] ][k,l] \\
+= \left[ \E \left[ {X}^T P \Delta^{(g,z)} s^{(g,z)} \right] \E\left[\tilde{X}^T \Delta^m s^m\right]^T\right][k,l] - \E \left[{X}^T P \Delta^{(g,z)} s^{(g,z)} (s^m)^T \Delta^m \tilde{X} \right][k,l] \\ = \left[\E\left[{X}[,k]^T P \Delta^{(g,z)} s^{(g,z)}\right]\E\left[\tilde{X}[,l]^T \Delta^m s^m\right]^T\right] - \E \left[{X}[,k]^T P \Delta^{(g,z)} s^{(g,z)} (s^m)^T \Delta^m \tilde{X}[,l] \right] \\ = \E\left(\sum_{i=1}^n x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i\right) \E \left(\sum_{j=1}^n \tilde{x}_{jl} \Delta^m_j s^m_j \right) - \E \left(\sum_{i=1}^n \sum_{j=1}^n x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i \Delta^m_j s^m_j \tilde{x}_{jl}  \right) \\
+= \sum_{i=1}^n \sum_{j=1}^n \E [x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i] \E[\Delta^m_js^m_j \tilde{x}_{jl}] - \sum_{i=1}^n \sum_{j=1}^n \E[x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i \Delta^m_j s^m_j \tilde{x}_{jl}] \\
+= \sum_{i=1}^n \sum_{j=1}^n \E [x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i ] \E[\Delta^m_js^m_j \tilde{x}_{jl}] - \sum_{i \neq j} \E [x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i] \E[\Delta^m_js^m_j \tilde{x}_{jl}] \\ - \sum_{i=1}^n \E[x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i \Delta^m_j s^m_j \tilde{x}_{jl}] \\
+= \sum_{i=1}^n \E[x_{ik} P_i \Delta_i^{(g,z)} H^{(g,z)}_i] \E[\tilde{x}_{il} \Delta^m_i H^m_i] - \sum_{i=1}^n \E[x_{ik} P_i \Delta_i^{(g,z)} H_i^{(g,z)} \Delta_i^m H_i^m \tilde{x}_{il}] \\ 
+= \sum_{i=1}^n \left[x_{ik} T_i^\theta(1) \Delta^{(g,z)}_i H^{(g,z)}_i \right] \cdot \left[{\Delta}^m_i(0) T^\theta_i(0) {H}^m_i(0) \tilde{x}_{il}(0) + {\Delta}^m_i(1) T_i^\theta(1) {H}^m_i(1) \tilde{x}_{il}(1)\right]
+\\ - \sum_{i=1}^n \left[x_{ik} T_i^\theta(1) \Delta^{(g,z)}_i H^{(g,z)}_i \Delta^m_i(1) H^m_i(1) \tilde{x}_{il}(1)\right] 
+\\ = \sum_{s=0}^1 \sum_{i=1}^n x_{ik} T_i^\theta(1) H_i^{(g,z)} \Delta_i^{(g,z)} T^\theta_i(s) \Delta_i^m(s) H^m(s) \tilde{x}_{il}(s) \\ - \sum_{i=1}^n \left[x_{il}T_i^\theta(1) \Delta^{(g,z)}_i H^{(g,z)}_i \Delta^m_i(1) H^m_i(1) \tilde{x}_{ik}(1)\right] \\ = \sum_{s=0}^1 X[,k]^T T^\theta(1) H^{(g,z)} \Delta^{(g,z)} T^\theta(s)\Delta^m(s)H^m(s) \tilde{X}(s)[,l] \\ - X[,k]^T \Delta^{(g,z)} H^{(g,z)} T^\theta(1)\Delta^m(1) H^m(1) \tilde{X}[,l].
+\end{multline}
+Combining (\ref{sub_mat_3_zeroinf}), (\ref{sub_mat_3_zeroinf_1}), and (\ref{sub_mat_3_zeroinf_2}) yields
+\begin{multline}\label{sub_mat_4_zeroinf_formula} J_{(\beta_{(g,z)}, \beta_m)}(\theta; m, g) = \left(\sum_{s=0}^1 X^T  T^\theta(1) H^{(g,z)} \Delta^{(g,z)} T^\theta(s) \Delta^m(s) H^m(s) \tilde{X}(s) \right) \\ - X^T \Delta^{(g,z)} H^{(g,z)} T^\theta(1) \Delta^m(1)H^m(1) \tilde{X}(1).
+\end{multline}
+
+\subsubsection*{Submatrix VI (zero-inflated)}
+Denote submatrix VI by $J_{(\beta_{(g,z)},\pi)}(\theta; m, g).$ The formula for $J_{(\beta_{(g,z)},\pi)}(\theta; m, g)$ is
+\begin{multline}\label{sub_mat_6_zeroinf}
+J_{(\beta_{(g,z)},\pi)}(\theta; m, g) = \E \left[-\nabla_{\beta_{(g,z)}} \nabla_{\pi} \mathcal{L}_z(\theta; m, g, p) \right] \\ + \E\left[ \nabla_{\beta_{(g,z)}}\mathcal{L}_z(\theta; m, g ,p)\right] \E \left[ \nabla_{\pi}\mathcal{L}_z(\theta; m,g,p) \right] \\ - \E \left[ \nabla_{\beta_{(g,z)}}\mathcal{L}_z(\theta; m,g,p) \nabla_{\pi}\mathcal{L}_z(\theta; m,g,p) \right].
+\end{multline} Recall that $\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) = X^T P \Delta^{(g,z)}s^{(g,z)}$ and $\nabla_\pi \mathcal{L}_z(\theta; m, g, p) = a \left(\sum_{i=1}^n p_i\right) - b,$ where $$a = \frac{1}{\pi} + \frac{1}{1 - \pi}, \textrm{    } b = \frac{n}{1 - \pi}.$$
+We have that
+\begin{equation}\label{sub_mat_6_zeroinf_1}
+\E \left[-\nabla_{\beta_{(g,z)}} \nabla_{\pi} \mathcal{L}_z(\theta; m, g, p) \right] = 0,
+\end{equation}
+as the derivative in $\pi$ of $\mathcal{L}_z(\theta; m, g, p)$ is a function of $\pi$, and the derivative in $\beta_{(g,z)}$ of this term is $0$.
+Next, we compute the difference of the second two terms of (\ref{sub_mat_6_zeroinf}).
+The $k$th entry of this vector is
+\begin{multline}\label{sub_mat_6_zeroinf_2}
+\E \left[ \nabla_\pi \mathcal{L}_z(\theta; m, g, p) \right] \E\left[\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, x)[k] \right] - \E \left[\nabla_{\pi}\mathcal{L}(\theta; m,g,p) \nabla_{\beta_{(g,z)}} \mathcal{L} (\theta; m,g,p)[k] \right] \\= \left(\E \left[ a \sum_{i=1}^n p_i - b \right] \right)\left( \E \left[{X}[,k]^T P \Delta^{(g,z)} s^{(g,z)} \right]\right) - \E \left[ \left( a \sum_{i=1}^n p_i - b \right) {X}[,k]^T P \Delta^{(g,z)} s^{(g,z)}\right] \\ = \left( a \sum_{i=1}^n \E[p_i] - b \right) \left( \sum_{j=1}^n \E [x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] \right) \\ - \E \left[ \left( a \sum_{i=1}^n p_i - b \right) \left( \sum_{j=1}^n \tilde{x}_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j \right) \right] \\= a \sum_{i=1}^n \sum_{j=1}^n \E [p_i] \E[x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] - b \sum_{j=1}^n \E[x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] \\ - \left[a \sum_{i=1}^n \sum_{j=1}^n \E [ p_i x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] - b \sum_{j=1}^n \E[x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] \right] \\ = a \sum_{i=1}^n \sum_{j=1}^n \E[p_i] \E[{x}_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] - a \sum_{i \neq j} \E[p_i] \E[x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] - a \sum_{i=1}^n \E[ x_{ik} p_i^2 \Delta^{(g,z)}_i s^{(g,z)}_i] \\ = a \sum_{i=1}^n \E[p_i] \E[x_{ik} p_i \Delta^{(g,z)}_i s^{(g,z)}_i] - a \sum_{i=1}^n \E[x_{ik} p_i^2 \Delta^{(g,z)}_i s^{(g,z)}_i] \\ = a \sum_{i=1}^n T^\theta_i(1) x_{ik} T^\theta_i(1) \Delta^{(g,z)}_i s^{(g,z)}_i - a \sum_{i=1}^n x_{ik} T^\theta_i(1) \Delta^{(g,z)}_i s^{(g,z)}_i \\ = a \sum_{i=1}^n \left( {x}_{ik} T^\theta_i(1)^2 \Delta^{(g,z)}_i s^{(g,z)}_i - x_{ik} T^\theta_i(1) \Delta^{(g,z)}_i s^{(g,z)}_i \right) = a \sum_{i=1}^n {x}_{ik}T^\theta_i(1)\Delta^{(g,z)}_i s^{(g,z)}_i \left( T^\theta_i(1) - 1\right) \\ = -a\sum_{i=1}^n x_{ik} T_i(0) T^\theta_i(1)\Delta^{(g,z)}_i H^{(g,z)}_i = -a X[,k]^T w^{(g,z)}.
+\end{multline}
+Combining (\ref{sub_mat_6_zeroinf}), (\ref{sub_mat_6_zeroinf_1}), and (\ref{sub_mat_6_zeroinf_2}), we conclude that
+\begin{equation}\label{sub_mat_6_zeroinf_formula} J_{(\beta_{(g,z)},\pi)}(\theta; m, g) = -\left( \frac{1}{\pi} + \frac{1}{1-\pi} \right)X^T w^{(g,z)}.\end{equation}
 
@@
 The log likelihood of this GLM is
-\begin{multline*}
+\begin{multline}\label{stat_acc_1}
 $$\mathcal{L}(\beta_1; m) = \sum_{i=1}^n T_i(0) f_m(m_i; h_m(\beta_1 + \hat{f}^m_i )) + \sum_{i=1}^n T_i(1) f_m(m_i; h_m(\hat{f}^m_i)) \\ = \sum_{i=1}^n T_i(1) f_m(m_i; h_m(\beta_1 + \hat{f}^m_i )) + C.$$
-\end{multline*}
+\end{multline}
 Thus, finding the MLE $\hat{\beta_1}$ is equivalent to estimating a GLM with intercept $\beta_1$, offsets $\hat{f}^m$, weights $T_i(1)$, and \textit{no} covariate terms.
@@
 
-Next, let $w_1, \dots, w_n \sim \mathcal{P}_2$ be weights, where $\mathcal{P}_2$ is a distribution bounded above by $1$ and below by $0$.
-(A special case is $w_i = 1$ for all $i \in \{1, \dots, n\}$.) Throughout, we assume that $y_iw_i$ and $\exp(o_i)w_i$ have finite first moment.
+We derive the (weighted) log likelihood of this model.
+Let $w_1, \dots, w_n \sim \mathcal{P}_2$ be weights, where $\mathcal{P}_2$ is a distribution bounded above by $1$ and below by $0$.
+(A special case, which corresponds to no weights, is $w_i = 1$ for all $i \in \{1, \dots, n\}$.) Throughout, we assume that $y_iw_i$ and $\exp(o_i)w_i$ have finite first moment.
 Suppose the cumulant-generating function and carrying density of the exponential family distribution are $\psi:\R \to \R$ and $c: \R \to \R$, respectively.
@@
 Our goal is to find the weighted MLE $\hat{\beta}$ of $\beta$.
-We consider three important choices for the exponential family distribution and link function, which amounts to maximizing (\ref{m_plus_o_mle}) for three specific choices of $r$ and $\psi$.
-In the first two cases -- Gaussian distribution with identity link and Poisson distribution with log link -- we find the \textit{finite-sample} maximizer of (\ref{m_plus_o_mle}); by contrast, in the third case -- negative binomial distribution with log link -- we find the \textit{asymptotic} maximizer.
+We consider three important choices for the exponential family distribution and link function.
+% This amounts to maximizing (\ref{m_plus_o_mle}) for three specific choices of $r$ and $\psi$.
+In the first two cases -- Gaussian distribution with identity link and Poisson distribution with log link -- we find the \textit{finite-sample} maximizer of (\ref{m_plus_o_mle}); by contrast, in the third case -- negative binomial distribution with log link -- we find an \textit{asymptotically exact} maximizer.
 
@@
 $$ \psi'(t) = \theta \left(\frac{e^t}{1 - e^t}\right) = \frac{\theta}{e^{-t} - 1}.$$ Define the function $\delta: \R \to \R$ by $\delta(t) = -\log\left(\theta/t + 1 \right).$ We see that
-$$\psi'(\delta(t)) = \frac{\theta}{\exp\left(\log(\theta/t + 1 )\right) - 1} = t,$$ and so $\delta = [\psi']^{-1}.$ By (\ref{can_param}), we have that
+$$\psi'(\delta(t)) = \frac{\theta}{\exp\left(\log(\theta/t + 1 )\right) - 1} = t,$$ implying $\delta = [\psi']^{-1}.$ By (\ref{can_param}), we have that
 $$
@@
 $$\frac{\partial \mathcal{L}(\beta;y)}{\partial \beta} = \sum_{i=1}^n w_i y_i  - \sum_{i=1}^n \frac{w_i(y_i + \theta) e^{\beta} e^{o_i}}{ \theta + e^{\beta} e^{o_i}}.$$
-Setting the derivative to zero, the MLE equation is
+Setting the derivative to zero, the equation defining the MLE is
 \begin{equation}\label{nb_mle}
@@
 \begin{equation}
-\hat{\beta}^{(\textrm{NB})} \xrightarrow{P} \log\left(\frac{\E[w_i y_i]}{\E[w_i e^{o_i}]} \right).
-\end{equation}
-But the Poisson MLE $\hat{\beta}^{(\textrm{Pois})}$ (\ref{pois_mle}) converges in probability to the same limit:
-$$ \hat{\beta}^{(\textrm{pois})} =  \log \left(\frac{ (1/n) \sum_{i=1}^n w_i y_i}{(1/n)\sum_{i=1}^n w_i e^{o_i}} \right) \xrightarrow{P} \log \left(\frac{\E[w_i y_i]}{ \E[w_i e^{o_i}]} \right).$$ Therefore, for large $n$, we can approximate $\hat{\beta}^{\textrm{(NB)}}$ by $\hat{\beta}^{(\textrm{pois})}$.
+\hat{\beta}^{\textrm{NB}} \xrightarrow{P} \log\left(\frac{\E[w_i y_i]}{\E[w_i e^{o_i}]} \right).
+\end{equation}
+But the Poisson MLE $\hat{\beta}^{\textrm{Pois}}$ (\ref{pois_mle}) converges in probability to the same limit:
+$$ \hat{\beta}^{\textrm{pois}} =  \log \left(\frac{ (1/n) \sum_{i=1}^n w_i y_i}{(1/n)\sum_{i=1}^n w_i e^{o_i}} \right) \xrightarrow{P} \log \left(\frac{\E[w_i y_i]}{ \E[w_i e^{o_i}]} \right).$$ Therefore, for large $n$, we can approximate $\hat{\beta}^{\textrm{NB}}$ by $\hat{\beta}^{\textrm{pois}}$.
+
+Suppose the log likelihood is unweighted (i.e., $w_i = 1$ for all $i$).
+We present an alternate (and simpler, more general) derivation of the approximate formula for the MLE.
+Let $y_i$ be a random variable with finite first moment and conditional mean $\E[y_i | o_i] = \exp(\beta + o_i).$ By the law of total expectation,
+$$ \E[y_i] = \E[\E[ y_i | o_i]] = \E[\exp(\beta + o_i)] = \exp(\beta) \E(e^{o_i}),$$ or
+\begin{equation}\label{nb_mo_4}
+\beta = \log\left(\frac{\E[y_i]}{E[e^{o_i}]} \right).
+\end{equation}
+Assuming the MLE is well-behaved, we have that $\hat{\beta} \xrightarrow{P} \beta.$ On the other hand, by WLLN and (\ref{nb_mo_4}),
+\begin{equation}\label{nb_mo_5}
+\log\left( \frac{(1/n) \sum_{i=1}^n y_i }{(1/n) \sum_{i=1}^n e^{o_i}} \right) \xrightarrow{P}  \log\left(\frac{ \E[y_i] }{ \E[e^{o_i}] } \right) = \beta.
+\end{equation}
+ Therefore, the LHS of (\ref{nb_mo_5}) approximates the MLE $\hat{\beta}$ for large $n$.
+
+\subsection*{Application to GLM-EIV}
+The GLM that we seek to estimate (\ref{stat_acc_1}) is an approximate intercept-plus-offset model: $T_1(1), \dots, T_n(1)$ are the weights $w_1,\dots, w_n$, and $\hat{f}^m_1, \dots, \hat{f}^m_n$ are the offsets $o_1, \dots, o_m$.
+Of course, $T_1(1), \dots, T_1(n)$ are in general dependent random variables, as are $\hat{f}^m_1, \dots, \hat{f}^m_n.$ $T_i(1)$ depends on $m_i$ and $g_i$, as well as the final parameter estimate $(\hat{\pi}, \hat{\beta}_m, \hat{\beta}_g),$ which itself is a function of $m$ and $g$; the situation is similar for the $\hat{f}^m_i$s.
+In practice, we find that the intercept-plus-offset model is very good approximation to the GLM (\ref{stat_acc_1}), especially when the number of cells $n$ is large.
+Additionally, we note that the GLM (\ref{stat_acc_1}) is fit as a subroutine of the algorithm for producing pilot parameter estimates (Algorithm \ref{algo:pilot_estimates}).
+The quality of the pilot parameter estimates does not affect the validity of the estimation and inference procedures (Algorithm \ref{algo:em_full}), barring issues related to convergence to local optima (which are outside the scope fo the current work).
 
@@
 \end{appendices}
-
-\bibliographystyle{unsrt}
-\newpage
-\bibliography{/Users/timbarry/optionFiles/glmeiv.bib}
-
 \end{document}

### 2021-11-17 text/outline/draft.tex
update text; set up simulations

@@
 \begin{center} 
-\textbf{A new class of exponential family measurement error models, with application to single-cell CRISPR screens}
+\textbf{A class of exponential family measurement error models, with application to single-cell CRISPR screens}
 \end{center}
@@
 Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases.
-In this work we study single-cell CRISPR screens from a statistical perspective.
+Despite their promise, single-cell CRISPR screens present substantial statistical challenges.
 We demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter.
@@
 Unlike SVD, eSVD models the relationship between the mean and variance of a gene's expression level, a phenomenon induced by the countedness of single-cell data \cite{Lause2021}.
-Similarly, Townes et al.\ \cite{Townes2019} proposed GLM-PCA, a generalization of PCA that directly models Poisson- or negative binomially-distributed gene expression counts.
+Similarly, Townes et al.\ \cite{Townes2019} proposed GLM-PCA, a generalization of PCA that directly models negative binomially-distributed gene expression counts while regressing out technical factors.
 We see our work as a continuation of this broad effort to ``port'' common statistical methods and models to single-cell count data.
@@
 There are several broad classes of single-cell CRISPR screen assays, each suited to answer a different set of biological questions \cite{Gasperini2019,Datlinger2021,Mimitou2019}.
-In this work we focus on so-called high-multiplicity of infection (MOI) single-cell CRISPR screens.
+In this work we focus on high-multiplicity of infection (MOI) single-cell CRISPR screens.
 We expect the ideas that we develop for this assay to apply (with some effort) to other classes of single-cell CRISPR screens as well.
@@
 These noncoding variants are thought to contribute to disease by modulating the expression one or more disease-relevant genes.
-We do not know the gene (or genes) through which most noncoding variants exert their effect, limiting the interpretability of GWAS results.
+Scientists do not know the gene (or genes) through which most noncoding variants exert their effect, limiting the interpretability of GWAS results.
 A central open challenge in genetics, therefore, is to link enhancers that harbor GWAS variants to the genes that they target at genome-wide scale \cite{Gasperini2020}.
@@
 We introduce such an extension -- which we call the ``zero-inflated GLM-EIV model'' -- in Appendix \ref{sec:zero_inf_model}.
-We derive efficient methods for estimation and inference in this model similar to those outlined above.
-However, given the ubiquity of background contamination in current single-cell CRISPR screen datasets, we focus our attention on the standard (or ``background read'') GLM-EIV model for the remainder of the main text.
+We derive efficient methods for estimation and inference in this model.
+Given the ubiquity of background contamination in current single-cell CRISPR screen datasets, we focus our attention on the standard (or ``background read'') GLM-EIV model for the remainder of the main text.
 

### 2021-11-17 R_scripts/simulations/create_simspec_objects.R

@@ -1,0 +1,75 @@
+library(magrittr)
+library(glmeiv)
+library(simulatr)
+
+args <- commandArgs(trailingOnly = TRUE)
+overwrite <- if (is.na(args[1])) TRUE else as.logical(args[1])
+save_obj <- function(obj, file_path, overwrite) {
+  if (!file.exists(file_path)) { # if file does not exist, save
+    saveRDS(obj, file_path)
+  } else { # if file does exist, save only if overwrite true
+    if (overwrite) {
+      saveRDS(obj, file_path)
+    }
+  }
+}
+
+####################################################################
+# Experiment 1
+# Vary g_pert, fix all other parameters
+# Vary distribution (Poisson, NB (known theta), NB (estimated theta)
+# One covariate: batch (bernoulli variable)
+####################################################################
+set.seed(4)
+theta <- 20
+n <- 150000
+sim_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/simulations/spec_objects")
+to_save_sim_1 <- paste0(sim_dir, "/sim_spec_1.rds")
+g_perturbation_grid <- seq(0, 3.5, 0.5)
+param_grid <- expand.grid(g_perturbation = g_perturbation_grid,
+                          fam_str = c("poisson", "nb_theta_unknown", "nb_theta_known"))
+param_grid$grid_id <- seq(1, nrow(param_grid))
+fam_obj <- lapply(as.character(param_grid$fam_str), function(str) {
+  switch(EXPR = str,
+         "poisson" = poisson(),
+         "nb_theta_unknown" = MASS::negative.binomial(theta),
+         "nb_theta_known" = MASS::negative.binomial(theta)) %>% glmeiv::augment_family_object()
+})
+param_grid$m_fam <- param_grid$g_fam <- fam_obj
+param_grid$run_unknown_theta_precomputation <- as.character(param_grid$fam_str) == "nb_theta_unknown"
+
+fixed_params <- list(
+  seed = 4,
+  n = n,
+  B = 500,
+  m_intercept = log(0.01),
+  m_perturbation = log(0.25),
+  g_intercept = log(0.005),
+  covariate_matrix = data.frame(batch = rbinom(n = n, size = 1, prob = 0.5)),
+  m_covariate_coefs = log(0.9),
+  g_covariate_coefs = log(1.1),
+  # n_processors = 20,
[…]

### 2021-11-17 bash_scripts/run_all.sh

@@ -4,14 +4,18 @@
 ##########
 # 0. setup
 ##########
+# source file paths
 bash setup.sh
 source ~/.research_config
+
+# set a couple variables
+sim_dir=$LOCAL_GLMEIV_DATA_DIR"public/simulations/"
 
 #######################
 # 1. Gasperini analysis
 #######################
 # QC Gasperini data
-Rscript ../R_scripts/qc_gasperini.R
+Rscript ../R_scripts/analysis/qc_gasperini.R
 # run glmeiv at scale on Gasperini data
 bash run_gasp_glmeiv.sh
 # run thresholding method at scale on Gasperini data
@@ -23,5 +27,11 @@ bash run_gasp_thresh.sh
 # QC Xie data
 
 # run glmeiv at scale on Xie data
-Rscript ../R_scripts/qc_xie.R
+Rscript ../R_scripts/analysis/qc_xie.R
 Rscript run_xie_glmeiv.sh
+
+################
+# 3. Simulations
+################
+Rscript ../R_scripts/simulations/create_simspec_objects.R
+$SIMULATR -f $sim_dir"spec_objects/sim_spec_1.rds" -r $sim_dir"results/raw_result_1.rds" -b 2

### 2021-11-17 R_scripts/simulations/create_simspec_objects.R
set n processors to 20

@@ -48,8 +48,7 @@ fixed_params <- list(
   covariate_matrix = data.frame(batch = rbinom(n = n, size = 1, prob = 0.5)),
   m_covariate_coefs = log(0.9),
   g_covariate_coefs = log(1.1),
-  # n_processors = 20,
-  n_processors = 2,
+  n_processors = 20,
   alpha = 0.95,
   n_em_rep = 15,
   save_membership_probs_mult = 500,
@@ -67,7 +66,7 @@ fixed_params <- list(
 one_rep_times <- list(generate_data_function = 10,
                       thresholding = 5,
                       glmeiv_fast = 20,
-                      glmeiv_slow = 250)
+                      glmeiv_slow = 500)
 
 sim_spec_1 <- create_simulatr_specifier_object(param_grid = param_grid, fixed_params = fixed_params,
                                                one_rep_times = one_rep_times, methods = c("glmeiv_fast", "glmeiv_slow", "thresholding"))

### 2021-11-17 R_scripts/simulations/create_simspec_objects.R
increase time request

@@ -66,7 +66,7 @@ fixed_params <- list(
 one_rep_times <- list(generate_data_function = 10,
                       thresholding = 5,
                       glmeiv_fast = 20,
-                      glmeiv_slow = 500)
+                      glmeiv_slow = 600)
 
 sim_spec_1 <- create_simulatr_specifier_object(param_grid = param_grid, fixed_params = fixed_params,
                                                one_rep_times = one_rep_times, methods = c("glmeiv_fast", "glmeiv_slow", "thresholding"))

### 2021-11-17 R_scripts/simulations/create_simspec_objects.R
increase wall time for glmeiv_slow

@@ -66,7 +66,7 @@ fixed_params <- list(
 one_rep_times <- list(generate_data_function = 10,
                       thresholding = 5,
                       glmeiv_fast = 20,
-                      glmeiv_slow = 600)
+                      glmeiv_slow = 1000)
 
 sim_spec_1 <- create_simulatr_specifier_object(param_grid = param_grid, fixed_params = fixed_params,
                                                one_rep_times = one_rep_times, methods = c("glmeiv_fast", "glmeiv_slow", "thresholding"))

### 2021-11-17 R_scripts/simulations/create_simspec_objects.R
fix log issue

@@ -13,6 +13,8 @@ save_obj <- function(obj, file_path, overwrite) {
     }
   }
 }
+sim_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/simulations/spec_objects")
+
 
 ####################################################################
 # Experiment 1
@@ -23,9 +25,8 @@ save_obj <- function(obj, file_path, overwrite) {
 set.seed(4)
 theta <- 20
 n <- 150000
-sim_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/simulations/spec_objects")
 to_save_sim_1 <- paste0(sim_dir, "/sim_spec_1.rds")
-g_perturbation_grid <- seq(0, 3.5, 0.5)
+g_perturbation_grid <- log(seq(1, 3.5, 0.5))
 param_grid <- expand.grid(g_perturbation = g_perturbation_grid,
                           fam_str = c("poisson", "nb_theta_unknown", "nb_theta_known"))
 param_grid$grid_id <- seq(1, nrow(param_grid))
@@ -63,12 +64,24 @@ fixed_params <- list(
   m_covariate_coefs_guess_range = log(c(0.25, 2)),
   g_covariate_coefs_guess_range = log(c(0.25, 2)))
 
-one_rep_times <- list(generate_data_function = 10,
-                      thresholding = 5,
-                      glmeiv_fast = 20,
-                      glmeiv_slow = 1000)
+one_rep_times <- list(generate_data_function = NA_integer_,
+                      thresholding = NA_integer_,
+                      glmeiv_fast = NA_integer_,
+                      glmeiv_slow = NA_integer_)
 
-sim_spec_1 <- create_simulatr_specifier_object(param_grid = param_grid, fixed_params = fixed_params,
-                                               one_rep_times = one_rep_times, methods = c("glmeiv_fast", "glmeiv_slow", "thresholding"))
+sim_spec_1 <- create_simulatr_specifier_object(param_grid = param_grid,
+                                               fixed_params = fixed_params,
+                                               one_rep_times = one_rep_times,
+                                               methods = c("glmeiv_fast", "glmeiv_slow", "thresholding"))
 # check <- simulatr::check_simulatr_specifier_object(simulatr_spec = sim_spec_1, B_in = 2, parallel = TRUE)
 save_obj(obj = sim_spec_1, file_path = paste0(sim_dir, "/sim_spec_1.rds"), overwrite = overwrite)
[…]

### 2021-11-18 bash_scripts/run_all.sh
update run_all.sh to run sim_spec_0

@@ -34,4 +34,4 @@ Rscript run_xie_glmeiv.sh
 # 3. Simulations
 ################
 Rscript ../R_scripts/simulations/create_simspec_objects.R
-$SIMULATR -f $sim_dir"spec_objects/sim_spec_1.rds" -r $sim_dir"results/raw_result_1.rds" -b 2
+$SIMULATR -f $sim_dir"spec_objects/sim_spec_0.rds" -r $sim_dir"results/raw_result_1.rds" -b 2

### 2021-11-18 R_scripts/simulations/create_simspec_objects.R
create sim spec 0

@@ -1,5 +1,5 @@
 library(magrittr)
-library(glmeiv)
+load_all("~/research_code/glmeiv/")
 library(simulatr)
 
 args <- commandArgs(trailingOnly = TRUE)
@@ -15,7 +15,64 @@ save_obj <- function(obj, file_path, overwrite) {
 }
 sim_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/simulations/spec_objects")
 
+one_rep_times <- list(generate_data_function = NA_integer_,
+                      thresholding = NA_integer_,
+                      glmeiv_fast = NA_integer_,
+                      glmeiv_slow = NA_integer_)
 
+##############################
+# Experiment 0
+# Vary g_pert
+# Fix distribution to Poisson
+# All three methods
+# One covariate (batch)
+##############################
+set.seed(4)
+theta <- 20
+n <- 150000
+to_save_sim_1 <- paste0(sim_dir, "/sim_spec_0.rds")
+g_perturbation_grid <- log(seq(1, 3.5, 0.5))
+param_grid <- expand.grid(g_perturbation = g_perturbation_grid)
+param_grid$grid_id <- seq(1, nrow(param_grid))
+
+fixed_params <- list(
+  m_fam = poisson() %>% augment_family_object(),
+  g_fam = poisson() %>% augment_family_object(),
+  seed = 4,
+  n = n,
+  B = 500,
+  m_intercept = log(0.01),
+  m_perturbation = log(0.25),
+  g_intercept = log(0.005),
+  covariate_matrix = data.frame(batch = rbinom(n = n, size = 1, prob = 0.5)),
+  m_covariate_coefs = log(0.9),
+  g_covariate_coefs = log(1.1),
+  n_processors = 20,
+  alpha = 0.95,
+  n_em_rep = 15,
+  save_membership_probs_mult = 500,
+  pi = 0.02,
+  m_offset = log(rpois(n = n, lambda = 10000)),
+  g_offset = log(rpois(n = n, lambda = 5000)),
+  pi_guess_range = c(1e-5, 0.03),
+  m_perturbation_guess_range = log(c(0.1, 1.5)),
+  g_perturbation_guess_range = log(c(0.5, 10)),
+  m_intercept_guess_range = log(c(1e-4, 1e-1)),
+  g_intercept_guess_range = log(c(1e-4, 1e-1)),
[…]

### 2021-11-18 text/outline/draft.tex
create sim spec 0
call library(glmeiv)

@@
 We recover these phenomena in precise theoretical terms in an idealized Gaussian setting.
-Next, we introduce GLM-EIV (“generalized linear model with errors-in-variables”), a new method for single-cell CRISPR screen analysis.
+Next, we introduce GLM-EIV (``exponential family errors-in-variables model''), a new method for single-cell CRISPR screen analysis.
 GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed, overcoming limitations of thresholded regression.
@@
 GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.
-In doing so, GLM-EIV implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic.} Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.
+In doing so, GLM-EIV is able to implicitly estimate the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic.} Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.
 

### 2021-11-18 R_scripts/simulations/create_simspec_objects.R

@@ -1,5 +1,5 @@
 library(magrittr)
-load_all("~/research_code/glmeiv/")
+library(glmeiv)
 library(simulatr)
 
 args <- commandArgs(trailingOnly = TRUE)

### 2021-11-18 text/outline/draft.tex
add sim spec 1

@@
 First, we conduct a careful study of the thresholding method.
-Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian model.
+Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian setting.
 \blue{Second, we introduce a new method, GLM-EIV (generalized linear model with errors-in-variables), for single-cell CRISPR screen analysis.
 GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.
-In doing so, GLM-EIV is able to implicitly estimate the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic.} Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.
+GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic.} Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.
 
@@
 \textbf{c,} Schematic illustrating the ``background read'' phenomenon.
-Due to errors in the sequencing process, unperturbed cells exhibit a nonzero gRNA count distribution (bottom).
+Due to errors in the sequencing and alignment processes, unperturbed cells exhibit a nonzero gRNA count distribution (bottom).
 The target of estimation is the change in mean gene expression in response to the perturbation (top).

### 2021-11-18 R_scripts/simulations/create_simspec_objects.R

@@ -68,11 +68,9 @@ sim_spec_0 <- create_simulatr_specifier_object(param_grid = param_grid,
                                                one_rep_times = one_rep_times,
                                                methods = c("glmeiv_fast", "glmeiv_slow", "thresholding"))
 # check <- simulatr::check_simulatr_specifier_object(simulatr_spec = sim_spec_0, B_in = 2, parallel = TRUE)
-save_obj(obj = sim_spec_0, file_path = paste0(sim_dir, "/sim_spec_0.rds"), overwrite = overwrite)
+save_obj(obj = sim_spec_0, file_path = paste0(sim_dir, "/sim_spec_0.rds"), overwrite = FALSE)
 
 
-# TURN OFF EVERYTHING BELOW:
-if (FALSE) {
 ####################################################################
 # Experiment 1
 # Vary g_pert, fix all other parameters
@@ -83,13 +81,12 @@ set.seed(4)
 theta <- 20
 n <- 150000
 to_save_sim_1 <- paste0(sim_dir, "/sim_spec_1.rds")
-g_perturbation_grid <- log(seq(1, 3.5, 0.5))
+g_perturbation_grid <- log(seq(1, 4, 0.75))
 param_grid <- expand.grid(g_perturbation = g_perturbation_grid,
-                          fam_str = c("poisson", "nb_theta_unknown", "nb_theta_known"))
+                          fam_str = c("nb_theta_unknown", "nb_theta_known"))
 param_grid$grid_id <- seq(1, nrow(param_grid))
 fam_obj <- lapply(as.character(param_grid$fam_str), function(str) {
   switch(EXPR = str,
-         "poisson" = poisson(),
          "nb_theta_unknown" = MASS::negative.binomial(theta),
          "nb_theta_known" = MASS::negative.binomial(theta)) %>% glmeiv::augment_family_object()
 })
@@ -125,20 +122,21 @@ sim_spec_1 <- create_simulatr_specifier_object(param_grid = param_grid,
                                                fixed_params = fixed_params,
                                                one_rep_times = one_rep_times,
                                                methods = c("glmeiv_fast", "glmeiv_slow", "thresholding"))
-# check <- simulatr::check_simulatr_specifier_object(simulatr_spec = sim_spec_1, B_in = 2, parallel = TRUE)
[…]

### 2021-11-18 bash_scripts/run_all.sh
update run_all.sh to run sim spec 1

@@ -34,4 +34,5 @@ Rscript run_xie_glmeiv.sh
 # 3. Simulations
 ################
 Rscript ../R_scripts/simulations/create_simspec_objects.R
-$SIMULATR -f $sim_dir"spec_objects/sim_spec_0.rds" -r $sim_dir"results/raw_result_1.rds" -b 2
+$SIMULATR -f $sim_dir"spec_objects/sim_spec_0.rds" -r $sim_dir"results/raw_result_0.rds"
+$SIMULATR -f $sim_dir"spec_objects/sim_spec_1.rds" -r $sim_dir"results/raw_result_1.rds"

### 2021-11-18 R_scripts/simulations/create_simspec_objects.R
remove sim spec check

@@ -122,7 +122,7 @@ sim_spec_1 <- create_simulatr_specifier_object(param_grid = param_grid,
                                                fixed_params = fixed_params,
                                                one_rep_times = one_rep_times,
                                                methods = c("glmeiv_fast", "glmeiv_slow", "thresholding"))
-check <- simulatr::check_simulatr_specifier_object(simulatr_spec = sim_spec_1, B_in = 2, parallel = TRUE)
+# check <- simulatr::check_simulatr_specifier_object(simulatr_spec = sim_spec_1, B_in = 2, parallel = TRUE)
 save_obj(obj = sim_spec_1, file_path = paste0(sim_dir, "/sim_spec_1.rds"), overwrite = overwrite)
 
 
@@ -172,6 +172,6 @@ sim_spec_2 <- create_simulatr_specifier_object(param_grid = param_grid,
                                                one_rep_times = one_rep_times,
                                                methods = c("glmeiv_slow", "glmeiv_fast", "thresholding"))
 
-check <- simulatr::check_simulatr_specifier_object(simulatr_spec = sim_spec_2, B_in = 2, parallel = TRUE)
+# check <- simulatr::check_simulatr_specifier_object(simulatr_spec = sim_spec_2, B_in = 2, parallel = TRUE)
 check$results$thresholding %>% filter(parameter == "m_perturbation", run_id == 1)
 }

### 2021-11-19 R_scripts/plotting/main_text_sim.R
start on main_text_sim plot

@@ -1,0 +1,45 @@
+library(dplyr)
+library(ggplot2)
+load_all("~/research_code/simulatr/")
+
+my_cols <- c("firebrick3", "dodgerblue3", "orchid4")
+fig_dir <- paste0(.get_config_path("LOCAL_CODE_DIR"), "glmeiv-manuscript/figures/main_text_sim")
+if (!dir.exists(fig_dir)) dir.create(fig_dir)
+my_theme <- theme(panel.border = element_blank(), panel.grid.major = element_blank(),
+                  panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))
+
+# load the results and specifier objects
+sim_result_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/simulations/results")
+sim_spec_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/simulations/spec_objects")
+
+############
+# 1. Poisson
+############
+pois_res <- readRDS(paste0(sim_result_dir, "/raw_result_0.rds")) %>% tibble()
+pois_res$target[pois_res$target == "confint_higher"] <- "confint_upper"
+
+pois_spec <- readRDS(paste0(sim_spec_dir, "/sim_spec_0.rds"))
+
+# summarize the results
+summarized_pois_results <- summarize_results(sim_spec = pois_spec, sim_res = pois_res,
+                                             metrics = c("bias", "mse", "coverage", "count", "time"),
+                                             parameters = "m_perturbation") %>% mutate(distribution = "Poisson")
+
+##################
+# 3. Make the plot
+##################
+
+to_plot_all <- summarized_pois_results %>% filter(metric != "count") %>%
+  mutate(metric_fct = factor(metric, levels = c("bias", "mse", "coverage", "time"),
+                             labels = c("Bias", "MSE", "Coverage", "Time (s)")),
+         Method = factor(method, levels = c("glmeiv_slow", "glmeiv_fast", "thresholding"),
+                         labels = c("GLM-EIV", "GLM-EIV (accelerated)", "Thresholding"))) %>%
+  arrange(Method) %>% mutate(exp_g_perturbation = exp(g_perturbation))
+
+ggplot(data = to_plot_all, mapping = aes(x = exp_g_perturbation, y = value, col = Method)) + 
[…]

### 2021-11-19 R_scripts/plotting/old/figure3.R

@@ -25,7 +25,7 @@ summarized_results_gaus <- simulatr::summarize_results(sim_spec = sim_spec, sim_
                                                        metrics = c("bias", "mse", "coverage"),
                                                        parameters = c("m_perturbation"),
                                                        threshold = 0.1) %>% tibble::as_tibble() %>% dplyr::filter(arm_arm_pi_big)
-pi <- summarized_results_gaus$pi[1]  
+pi <- summarized_results_gaus$pi[1]
 m_pert <- sim_spec@fixed_parameters$m_perturbation
 my_funct <- function(x) glmeiv::get_tresholding_estimator_bias(m_perturbation = m_pert, g_perturbation = x, pi = pi)
 
@@ -56,8 +56,6 @@ summarized_results_pois <- simulatr::summarize_results(sim_spec = sim_spec, sim_
 ####################
 sim_spec <- readRDS(paste0(sim_dir, "/sim_spec_5.rds")) # simulatr specifier object
 sim_res <- readRDS(paste0(sim_dir, "/raw_result_5.rds")) # raw results; note: PSC failed for certain settings in
-
-
 
 id_classifications <- glmeiv::obtain_valid_ids(sim_res = sim_res, pi_upper = 0.4)
 valid_ids <- id_classifications$valid_ids
@@ -91,8 +89,8 @@ p1 <- ggplot(to_plot_all, aes(x = g_perturbation, value, col = Method)) +
   geom_hline(data = dplyr::filter(to_plot_all, metric_fct == "Coverage"), mapping = aes(yintercept = 0.95), colour = "black") +
   geom_line() +
   geom_function(data = dplyr::filter(to_plot_all, metric_fct == "Bias", distribution == "Gaussian"), fun = my_funct, col = "black", lwd = 0.8) +
-  geom_vline(data = dplyr::filter(to_plot_all, distribution == "Poisson"), mapping = aes(xintercept = 1.5), col = "gray60") + 
+  geom_vline(data = dplyr::filter(to_plot_all, distribution == "Poisson"), mapping = aes(xintercept = 1.5), col = "gray60") +
   geom_vline(data = dplyr::filter(to_plot_all, distribution == "Poisson"), mapping = aes(xintercept = 2.5), col = "gray60") + geom_point() +
-  theme_bw() + theme(legend.position = "bottom", panel.grid.major = element_blank(), panel.grid.minor = element_blank())
[…]

### 2021-11-19 R_scripts/plotting/main_text_sim.R
add sim_spec 2

@@ -24,19 +24,22 @@ pois_spec <- readRDS(paste0(sim_spec_dir, "/sim_spec_0.rds"))
 summarized_pois_results <- summarize_results(sim_spec = pois_spec, sim_res = pois_res,
                                              metrics = c("bias", "mse", "coverage", "count", "time"),
                                              parameters = "m_perturbation") %>% mutate(distribution = "Poisson")
+# remove times with 0 counts
+thresh_grid_row_exclude <- summarized_pois_results %>% filter(metric == "count", value == 0) %>% dplyr::pull(grid_row_id)
+summarized_pois_results_trimmed <- summarized_pois_results %>%
+   filter(!(grid_id %in% thresh_grid_row_exclude & method == "thresholding"))
 
 ##################
 # 3. Make the plot
 ##################
-
-to_plot_all <- summarized_pois_results %>% filter(metric != "count") %>%
+to_plot_all <- summarized_pois_results_trimmed %>% filter(metric != "count") %>%
   mutate(metric_fct = factor(metric, levels = c("bias", "mse", "coverage", "time"),
                              labels = c("Bias", "MSE", "Coverage", "Time (s)")),
          Method = factor(method, levels = c("glmeiv_slow", "glmeiv_fast", "thresholding"),
                          labels = c("GLM-EIV", "GLM-EIV (accelerated)", "Thresholding"))) %>%
   arrange(Method) %>% mutate(exp_g_perturbation = exp(g_perturbation))
 
-ggplot(data = to_plot_all, mapping = aes(x = exp_g_perturbation, y = value, col = Method)) + 
+p <- ggplot(data = to_plot_all, mapping = aes(x = exp_g_perturbation, y = value, col = Method)) + 
   facet_grid(metric_fct ~ distribution, scales = "free_y") + xlab(expression(exp(beta[g]))) + scale_color_manual(values = my_cols) +
   geom_hline(data = dplyr::filter(to_plot_all, metric_fct == "Bias"), mapping = aes(yintercept = 0), colour = "black") +
   geom_hline(data = dplyr::filter(to_plot_all, metric_fct == "MSE"), mapping = aes(yintercept = 0), colour = "black") +

### 2021-11-19 R_scripts/simulations/create_simspec_objects.R

@@ -30,7 +30,6 @@ one_rep_times <- list(generate_data_function = NA_integer_,
 set.seed(4)
 theta <- 20
 n <- 150000
-to_save_sim_1 <- paste0(sim_dir, "/sim_spec_0.rds")
 g_perturbation_grid <- log(seq(1, 3.5, 0.5))
 param_grid <- expand.grid(g_perturbation = g_perturbation_grid)
 param_grid$grid_id <- seq(1, nrow(param_grid))
@@ -80,7 +79,6 @@ save_obj(obj = sim_spec_0, file_path = paste0(sim_dir, "/sim_spec_0.rds"), overw
 set.seed(4)
 theta <- 20
 n <- 150000
-to_save_sim_1 <- paste0(sim_dir, "/sim_spec_1.rds")
 g_perturbation_grid <- log(seq(1, 4, 0.75))
 param_grid <- expand.grid(g_perturbation = g_perturbation_grid,
                           fam_str = c("nb_theta_unknown", "nb_theta_known"))
@@ -126,16 +124,14 @@ sim_spec_1 <- create_simulatr_specifier_object(param_grid = param_grid,
 save_obj(obj = sim_spec_1, file_path = paste0(sim_dir, "/sim_spec_1.rds"), overwrite = overwrite)
 
 
-if (FALSE) {
 ########################################
 # Experiment 2
 # Gaussian response distribution
 # Vary g_perturbation
-# Two covariates: batch and library size
+# Two covariates: batch, library size
 ########################################
 set.seed(4)
 n <- 150000
-to_save_sim_2 <- paste0(sim_dir, "/sim_spec_2.rds")
 g_perturbation_grid <- seq(0, 7)
 param_grid <- data.frame(g_perturbation = g_perturbation_grid,
                          grid_id = seq(1, length(g_perturbation_grid)))
@@ -144,13 +140,14 @@ fixed_params <- list(
   g_fam = gaussian() %>% augment_family_object(),
   seed = 4,
   n = n,
-  B = 500,
+  B = 1000,
   m_intercept = 3,
   m_perturbation = -4,
   g_intercept = 1,
-  covariate_matrix = data.frame(lib_size = rpois(n = n, lambda = 10000)),
-  m_covariate_coefs = 0.0025,
-  g_covariate_coefs = -0.005,
+  covariate_matrix = data.frame(lib_size = rpois(n = n, lambda = 10000),
[…]

### 2021-11-19 bash_scripts/run_all.sh

@@ -36,3 +36,4 @@ Rscript run_xie_glmeiv.sh
 Rscript ../R_scripts/simulations/create_simspec_objects.R
 $SIMULATR -f $sim_dir"spec_objects/sim_spec_0.rds" -r $sim_dir"results/raw_result_0.rds"
 $SIMULATR -f $sim_dir"spec_objects/sim_spec_1.rds" -r $sim_dir"results/raw_result_1.rds"
+$SIMULATR -f $sim_dir"spec_objects/sim_spec_2.rds" -r $sim_dir"results/raw_result_2.rds"

### 2021-11-20 text/outline/draft.tex
update simpsec objects: poisson grid [1,4]; ep_tol 1e-5 for glmeiv_slow.

@@
 Despite their promise, single-cell CRISPR screens present substantial statistical challenges.
-A major difficulty is that CRISPR perturbations \blue{are assigned stochastically to cells and cannot be observed directly}.
+A major difficulty is that CRISPR perturbations are assigned stochastically to cells and cannot be observed directly.
 As a consequence, one cannot know with certainty which cells were perturbed.
@@
 Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian setting.
-\blue{Second, we introduce a new method, GLM-EIV (generalized linear model with errors-in-variables), for single-cell CRISPR screen analysis.
+Second, we introduce a new method, GLM-EIV (generalized linear model with errors-in-variables), for single-cell CRISPR screen analysis.
 GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.
-GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic.} Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.
+GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic.
+Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.
 
@@
 SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification.
-% Other applied methods for single-cell CRISPR screen analysis include MIMOSCA \cite{Dixit2016} and scMAGeCK \cite{Yang2019}.
+Other applied methods for single-cell CRISPR screen analysis include MIMOSCA \cite{Dixit2016} and scMAGeCK \cite{Yang2019}.
 These methods, like SCEPTRE, focus on hypothesis testing (rather than estimation), but unlike SCEPTRE, they ignore the countedness of the data and are unable to handle confounders.
 In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE.
-Most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, \blue{an objective that is unattainable within the nonparameteric hypothesis testing framework of SCEPTRE.}
+Most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is unattainable within the nonparameteric hypothesis testing framework of SCEPTRE.
 
@@
 This procedure is repeated for a large set of preselected perturbation-gene pairs.
-\blue{The enhancer-by-enhancer approach is valid because the perturbations assort into cells approximately independently of one another.}
+The enhancer-by-enhancer approach is valid because the perturbations assort into cells approximately independently of one another.
 
 \subsection{Analysis challenges}
-\blue{High MOI single-cell CRISPR screens present several statistical challenges, four of which we highlight here.} Throughout, we consider a single perturbation-gene pair.
+High MOI single-cell CRISPR screens present several statistical challenges, four of which we highlight here.
+Throughout, we consider a single perturbation-gene pair.
 First, the ``treatment'' variable -- i.e., the presence or absence of a perturbation -- cannot be directly observed.
@@
 We must leverage these gRNAs to impute (explicitly or implicitly) perturbation assignments onto the cells (Figure \ref{analysis_challenges}b).
-\blue{Second, ``technical factors'' -- sources of variation that are experimental rather than biological in origin -- impact both gene expression and gRNA expression (Figure \ref{analysis_challenges}b).
-Technical factors act as confounders in the measurement process and therefore must be accounted for in differential expression models.} Third, the gene and gRNA data are sparse, discrete counts.
-Therefore, classical statistical approaches that assume Gaussianity or homoscedasticity are inapplicable.
-Finally, and most subtly, sequenced gRNAs sometimes are mapped to cells that have not received a perturbation.
-This phenomenon, which we call the ``background read'' phenomenon, results from errors in the sequencing and alignment processes \cite{Replogle2020}.
+Second, ``technical factors'' -- sources of variation that are experimental rather than biological in origin -- impact the measurement of both gene and gRNA expressions and therefore act as confounders (Figure \ref{analysis_challenges}b).
+Third, the gene and gRNA data are sparse, discrete counts.
+Consequently, classical statistical approaches that assume Gaussianity or homoscedasticity are inapplicable.
+Finally, and most subtly, sequenced gRNAs sometimes map to cells that have not received a perturbation.
+This phenomenon, which we ``background contamination,'' results from errors in the sequencing and alignment processes \cite{Replogle2020}.
 The marginal distribution of the gRNA counts is best conceptualized as a mixture model (Figure \ref{analysis_challenges}c; Gaussian distributions used for illustration purposes only).
-\blue{Unperturbed and perturbed cells both exhibit nonzero gRNA count distributions, but this distribution overall is greater for perturbed cells.} Figure \ref{analysis_challenges}d shows example data on four (of possibly tens or hundreds of thousands of) cells.
-\blue{The analysis objective is to leverage the gene expressions and gRNA counts to estimate the effect of the (latent) perturbation on gene expression, accounting for the technical factors.}
-
-In this work we analyze two large-scale, high MOI, single-cell CRISPR screen datasets published by Gasperini et al.\ and Xie et al.
+Unperturbed and perturbed cells both exhibit nonzero gRNA count distributions, but this distribution is shifted upward for perturbed cells.
+Figure \ref{analysis_challenges}d shows example data on four (of possibly tens or hundreds of thousands of) cells.
+The analysis objective is to leverage the gene expressions and gRNA counts to estimate the effect of the (latent) perturbation on gene expression, accounting for the technical factors.
+
+In this work we analyze two large-scale, high MOI, single-cell CRISPR screen datasets published by Gasperini et al.\ and Xie et al.\ in 2019.
 Gasperini (resp., Xie) targeted approximately 6,000 (resp., 500) candidate enhancers in a population of approximately 200,000 (resp., 100,000) cells.
@@
 \section{Thresholding method}
-\blue{In this section we study thresholded regression from empirical and theoretical perspectives, uncovering several limitations of the method.
-Gasperini and Xie both imputed perturbation identities onto the cells via thresholding, but they carried out the subsequent differential expression analysis in different ways: Gasperini used negative binomial regression, whereas Xie used nonparametric independence testing.
-These two strategies pose similar challenges, but we investigate Gasperini's variant of the thresholding method, as it relates most closely to GLM-EIV.}
+We study the thresholding method from empirical and theoretical perspectives, highlighting several limitations of the approach.
+Gasperini and Xie both imputed perturbation assignments onto the cells via thresholding, but they carried out subsequent differential expression analyses in different ways: Gasperini used negative binomial regression, whereas Xie applied nonparametric tests of independence.
+We study Gasperini's variant of the thresholding method, as it relates more closely to GLM-EIV and (in our view) is the more natural approach.
 
@@
 \end{equation}
-where (i) $NB_\theta(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta^m$; (ii) $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in  \R^{d-1}$ are unknown parameters; and (iii) $\log(d_i^m)$ is an offset term.
-Fit a GLM to obtain estimates of the parameters.
-
-\item[3.] Compute a $p$-value and confidence interval for the target of inference $\beta^m_1$.
+where (i) $NB_{\theta^m}(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta^m$; (ii) $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in  \R^{d-1}$ are unknown parameters; and (iii) $\log(d_i^m)$ is an offset term.
+
+\item[3.]  Fit a GLM to obtain estimates of the parameters. Compute a $p$-value and confidence interval for the target of inference $\beta^m_1$.
 \end{itemize}
-\blue{We include the library size $d^m_i$ as an offset term in (\ref{thresh_glm}) so that $\beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i$ can be interpreted as a relative expression: exponentiating both sides of (\ref{thresh_glm}), we obtain
-$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m.$$ We see that $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration.}
-
-The biological interpretation for the target of inference $\beta^m_1$ as follows:  $\beta^m_1$ is the log-transformed fold change in gene expression in response to the perturbation, \blue{controlling for} the technical factors.
-Fold change (obtained by exponentiating $\beta^m_1$) is the ratio of the mean gene expression in cells that were perturbed to the mean gene expression in cells that were not perturbed.
-$\exp(\beta^m_1) = 1$ indicates no change in mean expression, while $\exp(\beta^m_1) > 1$ and $\exp(\beta^m_1) < 1$ indicate an increase and decrease in expression in response to the perturbation, respectively (accounting for technical factors).
+The sequencing depth $d^m_i$ is included as an offset term in (\ref{thresh_glm}) so that $\beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i$ can be interpreted as a relative expression: exponentiating both sides of (\ref{thresh_glm}) yields
+$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m.$$ The term $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration.
+The target of inference, $\beta^m_1$, is the log-transformed fold change in gene expression in response to the perturbation, \blue{controlling for} the technical factors.
+Fold change (i.e., $\exp(\beta^m_1)$) is the ratio of mean gene expression in perturbed cells to unperturbed cells: $\exp(\beta^m_1) = 1$ indicates no change in expression, while $\exp(\beta^m_1) > 1$ and $\exp(\beta^m_1) < 1$ indicate an increase and decrease in expression in response to the perturbation, respectively (accounting for technical factors).
 
@@
 The positive control pairs consisted of perturbations that targeted gene transcription start sites (TSSs) for inhibition.
-Repressing the TSS of a given gene decreases its expression; therefore, the positive control pairs \textit{a priori} are expected to exhibit a strong, negative log fold change in expression.
-
-To investigate the sensitivity of the thresholding method to the selected threshold, we deployed the thresholding method on the positive control data using three different thresholds: 1, 5, and 20.
+Repressing the TSS of a given gene decreases its expression; therefore, the positive control pairs \textit{a priori} are expected to exhibit a strong decrease in expression.
+
+To investigate the sensitivity of the thresholding method to the selected threshold, we deployed the thresholding method on the positive control data using three different choices for the threshold: 1, 5, and 20.
 We found that the chosen threshold substantially impacted the results (Figure \ref{thresholding_empirical}a-b).
@@
 
-We reasoned that thresholded regression systematically underestimated effect sizes on the positive control pairs, especially for small thresholds (an example of \textit{attenuation bias}).
+We reasoned that the thresholding method systematically underestimates effect sizes on the positive control pairs, especially when the threshold is small.
 For a given perturbation, the vast majority ($>98\%$) of cells are unperturbed.
@@
 
-We hypothesized, however, that the reduction in bias conferred by selecting a large threshold comes at the cost of increasing the variance of the estimator.
-To investigate, we compared $p$-values and confidence intervals produced by threshold = 5 and threshold = 20 for the target of inference $\beta^m_1$.
+We hypothesized, however, that the reduction in bias obtained by selecting a large threshold comes at the cost of increasing the variance of the estimator.
+To investigate, we compared $p$-values and confidence intervals produced by threshold = 5 versus threshold = 20 for the target of inference $\beta^m_1$.
 We found that threshold = 5 yielded smaller (i.e., more significant) $p$-values and narrower confidence intervals than did threshold = 20 (Figure \ref{thresholding_empirical}c-d).
@@
 \subsubsection*{Negative binomial model}
-Building on the work of several previous authors \cite{Townes2019,Svensson2020,Hafemeister2019}, Sarkar and Stephens \cite{Sarkar2021} proposed a simple strategy for modeling for single-cell gene expression data, which, in the framework of the negative binomial GLM, is equivalent to using the log-transformed library size as an offset term (as in (\ref{thresh_glm})).
+Building on the work of several previous authors \cite{Townes2019,Svensson2020,Hafemeister2019}, Sarkar and Stephens \cite{Sarkar2021} proposed a simple strategy for modeling for single-cell gene expression data, which, in the framework of negative binomial GLMs, is equivalent to using the log-transformed library size as an offset term (as in (\ref{thresh_glm})).
 We generalize Sarkar and Stephens' approach to model \textit{both} gene and gRNA modalities.
@@
 where, similar to above, $\theta^g > 0$ is a known negative binomial size parameter, and $\beta^g_0 \in \R, \beta^g_1 \in \R, \gamma_g \in \R^{d - 2}$ are unknown constants.
-We use a negative binomial GLM to model the gRNA counts because gRNA molecules are transcribed in the cell in the same way that gene transcripts are \cite{Datlinger2017,Hill2018}.
+We use a negative binomial GLM to model the gRNA counts because the gRNA molecules are transcribed in the cell in the same way as gene transcripts \cite{Datlinger2017,Hill2018}.
 Finally, we model the marginal perturbation probability as
@@
 The easiest approach, given the complexity of the log likelihood, would be to run a parametric bootstrap on the fitted model.
-This strategy, however, is prohibitively slow, as the data are large and we use an EM algorithm to fit the model.
-Therefore, we derive an analytic formula for the asymptotic observed information matrix using Louis's Theorem \cite{Louis1982} (see appendix).
+This strategy, however, is prohibitively slow, as the data are large and the EM algorithm is iterative.
+Therefore, we derive an analytic formula for the asymptotic observed information matrix using Louis's Theorem \cite{Louis1982}.
 Leveraging this analytic formula, we can calculate standard errors (and $p$-values and confidence intervals) quickly, enabling us to perform inference in practice on real, large-scale data.
-The derivation is fairly involved; simulation studies (subsequent section) confirm the correctness of the results.
+The derivation is fairly involved (see appendix); simulation studies confirm the correctness of the results (see Section \ref{sec:simulation}) .
 
@@
 Assuming that we run the algorithm 15 times using randomly-generated pilot estimates (to ensure convergence to the global maximum), and assuming that the algorithm iterates through E and M steps about 10 times per run, we must fit approximately 300 GLMs.
-(These numbers are based on an exploratory application of the method to real and simulated data.) We devised a strategy (Algorithm \ref{algo:pilot_estimates}) to produce a highly accurate pilot estimate $(\pi^\textrm{pilot}, \beta_m^\textrm{pilot}, \beta_g^\textrm{pilot})$ of the true parameters, enabling us to run the algorithm once and converge upon the MLE within a few iterations.
+(These numbers are based on exploratory applications of the method to real and simulated data.) We devised a strategy (Algorithm \ref{algo:pilot_estimates}) to produce a highly accurate pilot estimate $(\pi^\textrm{pilot}, \beta_m^\textrm{pilot}, \beta_g^\textrm{pilot})$ of the true parameters, enabling us to run the algorithm once and converge upon the MLE within a few iterations.
 The strategy involves layering several statistical ``tricks'' (some new, some old) on top of one another.
@@
 To produce estimates $[\beta^m_0]^\textrm{pilot}$ and $[\gamma_m^T]^\textrm{pilot}$, we regress the gene expressions $m$ onto the technical factors $X$.
-The intuition for why this works is as follows: the probability of perturbation $\pi$ is very small.
+The intuition for this procedure is as follows: the probability of perturbation $\pi$ is very small.
 Therefore, the true log likelihood is approximately equal to the log likelihood that results from omitting $p_i$ from the model:
@@
 Consequently, we do not need to run the relatively slow IRLS procedure to carry out the M step of the ``reduced'' GLM-EIV algorithm.
-Overall, the proposed algorithm for obtaining the pilot parameter estimates (Algorithm \ref{algo:pilot_estimates}) requires fitting two GLMs (via IRLS).
+Overall, the proposed algorithm for obtaining the pilot parameter estimates (Algorithm \ref{algo:pilot_estimates}) requires fitting only two GLMs (via IRLS).
 
@@
 We introduce such an extension -- which we call the ``zero-inflated GLM-EIV model'' -- in Appendix \ref{sec:zero_inf_model}.
-We derive efficient methods for estimation and inference in this model.
-Given the ubiquity of background contamination in current single-cell CRISPR screen datasets, we focus our attention on the standard (or ``background read'') GLM-EIV model for the remainder of the main text.
-
-\section{Simulation studies}
+We additionally derive efficient methods for estimation and inference in the zero-inflated model.
+Given the ubiquity of background contamination in current single-cell CRISPR screen datasets, we focus our attention on the standard GLM-EIV model for the remainder of the main text.
+
+\section{Simulation studies}\label{sec:simulation}
 

### 2021-11-20 R_scripts/plotting/main_text_sim.R

@@ -17,7 +17,6 @@ sim_spec_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/simula
 ############
 pois_res <- readRDS(paste0(sim_result_dir, "/raw_result_0.rds")) %>% tibble()
 pois_res$target[pois_res$target == "confint_higher"] <- "confint_upper"
-
 pois_spec <- readRDS(paste0(sim_spec_dir, "/sim_spec_0.rds"))
 
 # summarize the results
@@ -29,18 +28,37 @@ thresh_grid_row_exclude <- summarized_pois_results %>% filter(metric == "count",
 summarized_pois_results_trimmed <- summarized_pois_results %>%
    filter(!(grid_id %in% thresh_grid_row_exclude & method == "thresholding"))
 
+#######
+# 2. NB
+#######
+nb_res <- readRDS(paste0(sim_result_dir, "/raw_result_1.rds")) %>% tibble()
+nb_res$target[nb_res$target == "confint_higher"] <- "confint_upper"
+nb_spec <- readRDS(paste0(sim_spec_dir, "/sim_spec_1.rds"))
+
+# summarize the results
+summarized_nb_results <- summarize_results(sim_spec = nb_spec, sim_res = nb_res,
+                                           metrics = c("bias", "mse", "coverage", "count", "time"),
+                                           parameters = "m_perturbation")
+thresh_grid_row_exclude <- summarized_nb_results %>% filter(metric == "count", value == 0) %>% dplyr::pull(grid_row_id)
+summarized_nb_results_trimmed <- summarized_nb_results %>% 
+  filter(!(grid_id %in% thresh_grid_row_exclude & method == "thresholding")) %>%
+  mutate(distribution = ifelse(fam_str == "nb_theta_known", "NB (theta known)", "NB (theta est.)")) %>%
+  select(-g_fam, -m_fam, -run_unknown_theta_precomputation, -fam_str)
+
 ##################
 # 3. Make the plot
 ##################
-to_plot_all <- summarized_pois_results_trimmed %>% filter(metric != "count") %>%
+to_plot_all <- rbind(summarized_pois_results_trimmed, summarized_nb_results_trimmed) %>% filter(metric != "count") %>%
   mutate(metric_fct = factor(metric, levels = c("bias", "mse", "coverage", "time"),
                              labels = c("Bias", "MSE", "Coverage", "Time (s)")),
[…]

### 2021-11-20 R_scripts/simulations/create_simspec_objects.R

@@ -30,7 +30,7 @@ one_rep_times <- list(generate_data_function = NA_integer_,
 set.seed(4)
 theta <- 20
 n <- 150000
-g_perturbation_grid <- log(seq(1, 3.5, 0.5))
+g_perturbation_grid <- log(seq(1, 4, 0.5))
 param_grid <- expand.grid(g_perturbation = g_perturbation_grid)
 param_grid$grid_id <- seq(1, nrow(param_grid))
 
@@ -79,7 +79,7 @@ save_obj(obj = sim_spec_0, file_path = paste0(sim_dir, "/sim_spec_0.rds"), overw
 set.seed(4)
 theta <- 20
 n <- 150000
-g_perturbation_grid <- log(seq(1, 4, 0.75))
+g_perturbation_grid <- log(seq(1, 4, 0.5))
 param_grid <- expand.grid(g_perturbation = g_perturbation_grid,
                           fam_str = c("nb_theta_unknown", "nb_theta_known"))
 param_grid$grid_id <- seq(1, nrow(param_grid))

### 2021-11-22 R_scripts/plotting/main_text_sim.R
add decrease ep_tol in Gaussian expt to 1e-7

@@ -5,8 +5,6 @@ load_all("~/research_code/simulatr/")
 my_cols <- c("firebrick3", "dodgerblue3", "orchid4")
 fig_dir <- paste0(.get_config_path("LOCAL_CODE_DIR"), "glmeiv-manuscript/figures/main_text_sim")
 if (!dir.exists(fig_dir)) dir.create(fig_dir)
-my_theme <- theme(panel.border = element_blank(), panel.grid.major = element_blank(),
-                  panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))
 
 # load the results and specifier objects
 sim_result_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/simulations/results")
@@ -21,7 +19,7 @@ pois_spec <- readRDS(paste0(sim_spec_dir, "/sim_spec_0.rds"))
 
 # summarize the results
 summarized_pois_results <- summarize_results(sim_spec = pois_spec, sim_res = pois_res,
-                                             metrics = c("bias", "mse", "coverage", "count", "time"),
+                                             metrics = c("bias", "mse", "coverage", "ci_width", "count", "time"),
                                              parameters = "m_perturbation") %>% mutate(distribution = "Poisson")
 # remove times with 0 counts
 thresh_grid_row_exclude <- summarized_pois_results %>% filter(metric == "count", value == 0) %>% dplyr::pull(grid_row_id)
@@ -37,7 +35,7 @@ nb_spec <- readRDS(paste0(sim_spec_dir, "/sim_spec_1.rds"))
 
 # summarize the results
 summarized_nb_results <- summarize_results(sim_spec = nb_spec, sim_res = nb_res,
-                                           metrics = c("bias", "mse", "coverage", "count", "time"),
+                                           metrics = c("bias", "mse", "ci_width", "coverage", "count", "time"),
                                            parameters = "m_perturbation")
 thresh_grid_row_exclude <- summarized_nb_results %>% filter(metric == "count", value == 0) %>% dplyr::pull(grid_row_id)
 summarized_nb_results_trimmed <- summarized_nb_results %>% 
@@ -49,8 +47,8 @@ summarized_nb_results_trimmed <- summarized_nb_results %>%
 # 3. Make the plot
 ##################
[…]

### 2021-11-22 R_scripts/plotting/supplement_sim.R

@@ -1,0 +1,41 @@
+library(dplyr)
+library(ggplot2)
+load_all("~/research_code/simulatr/")
+
+my_cols <- c("dodgerblue3", "orchid4")
+fig_dir <- paste0(.get_config_path("LOCAL_CODE_DIR"), "glmeiv-manuscript/figures/supplement_sim")
+if (!dir.exists(fig_dir)) dir.create(fig_dir)
+
+# load the results and specifier objects
+sim_result_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/simulations/results")
+sim_spec_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/simulations/spec_objects")
+
+#############
+# 1. Gaussian
+#############
+gaus_res <- readRDS(paste0(sim_result_dir, "/raw_result_2.rds")) %>% tibble()
+gaus_res$target[gaus_res$target == "confint_higher"] <- "confint_upper"
+gaus_spec <- readRDS(paste0(sim_spec_dir, "/sim_spec_2.rds"))
+
+# summarize results
+summarized_gaus_results <- summarize_results(sim_spec = gaus_spec, sim_res = gaus_res,
+                                             metrics = c("bias", "mse", "coverage", "ci_width"),
+                                             parameters = "m_perturbation")
+
+to_plot_all <- summarized_gaus_results %>% filter(metric != "count") %>%
+  mutate(metric_fct = factor(metric, levels = c("bias", "mse", "coverage", "ci_width", "time"),
+                             labels = c("Bias", "MSE", "Coverage", "CI width", "Time (s)")),
+         Method = factor(method, levels = c("glmeiv_slow", "glmeiv_fast", "thresholding"),
+                         labels = c("GLM-EIV", "GLM-EIV (accelerated)", "Thresholding"))) %>%
+  arrange(Method)
+
+p <- ggplot(data = to_plot_all, mapping = aes(x = g_perturbation, y = value, col = Method)) + 
+  facet_wrap(. ~ metric_fct, scales = "free_y", nrow = 1) + xlab(expression(beta[g])) + scale_color_manual(values = my_cols) +
+  geom_hline(data = dplyr::filter(to_plot_all, metric_fct == "Bias"), mapping = aes(yintercept = 0), colour = "black") +
[…]

### 2021-11-22 R_scripts/simulations/create_simspec_objects.R

@@ -162,7 +162,9 @@ fixed_params <- list(
   g_intercept_guess_range = c(1, 1),
   m_covariate_coefs_guess_range = c(0.0025, 0.0025),
   g_covariate_coefs_guess_range = c(-0.005, -0.005),
-  run_unknown_theta_precomputation = FALSE)
+  run_unknown_theta_precomputation = FALSE,
+  exponentiate_coefs = FALSE,
+  ep_tol = 1e-7)
 
 sim_spec_2 <- create_simulatr_specifier_object(param_grid = param_grid,
                                                fixed_params = fixed_params,

### 2021-11-22 bash_scripts/run_gasp_glmeiv.sh

@@ -20,4 +20,4 @@ nextflow run https://github.com/timothy-barry/glmeiv-pipeline -r main \
 --gRNA_metadata $gRNA_metadata \
 --g_offsets $g_offsets \
 --g_fam_str $g_fam \
--bg > $PWD/log -ansi-log false -with-trace -profile local
+-profile az -bg > $PWD/log -ansi-log false -with-trace

### 2021-11-22 bash_scripts/run_xie_glmeiv.sh

@@ -4,22 +4,41 @@ source set_xie_args.sh
 # Run Nextflow pipeline
 #######################
 rm -f trace.txt
-# nextflow run $glmeiv_nf_pipeline \
+
+# echo args
+clear
+echo covariate_matrix: $covariate_matrix
+echo result_dir: $result_dir
+echo gene_odm: $gene_odm
+echo gene_metadata: $gene_metadata
+echo m lib size: $m_offsets
+echo m family: $m_fam
+echo gRNA odm: $gRNA_odm
+echo gRNA metadata: $gRNA_metadata
+echo g lib size: $g_offsets
+echo g fam: $g_fam
+echo work dir: $work_dir
+echo profile: $prof
+echo pairs: $gRNA_gene_pairs
+echo gene pod size: $gene_pod_size
+echo gRNA pod size: $gRNA_pod_size
+echo pair pod size: $pair_pod_size
+
+# run the pipeline
 nextflow run https://github.com/timothy-barry/glmeiv-pipeline -r main \
---pairs $gRNA_gene_pairs \
---covariate_matrix $covariate_matrix \
---pair_pod_size $pair_pod_size \
---result_dir $result_dir \
---gene_pod_size $gene_pod_size \
---gene_odm $gene_odm \
---gene_metadata $gene_metadata \
---m_offsets $m_offsets \
---m_fam_str $m_fam \
---gRNA_pod_size $gRNA_pod_size \
---gRNA_odm $gRNA_odm \
---gRNA_metadata $gRNA_metadata \
---g_offsets $g_offsets \
---g_fam_str $g_fam \
---g_theta $g_theta \
---result_file_name $result_file_name \
--bg > $PWD/log -ansi-log false -with-trace
+ --pairs $gRNA_gene_pairs \
+ --covariate_matrix $covariate_matrix \
+ --pair_pod_size $pair_pod_size \
+ --result_dir $result_dir \
+ --gene_pod_size $gene_pod_size \
+ --gene_odm $gene_odm \
+ --gene_metadata $gene_metadata \
+ --m_offsets $m_offsets \
+ --m_fam_str $m_fam \
+ --gRNA_pod_size $gRNA_pod_size \
+ --gRNA_odm $gRNA_odm \
+ --gRNA_metadata $gRNA_metadata \
+ --g_offsets $g_offsets \
+ --g_fam_str $g_fam \
+ --result_file_name $result_file_name \
+ -bg > $PWD/log -ansi-log false -with-trace -profile $prof -w $work_dir

### 2021-11-22 bash_scripts/set_xie_args.sh

@@ -2,10 +2,12 @@ source ~/.research_config
 
 # use -t flag to set "trial" parameters; otherwise, sets "at-scale" parameters
 trial=false
-while getopts t OPT
+azure=false
+while getopts ta OPT
 do
     case "$OPT" in
         t) trial=true ;;
+        a) azure=true ;;
     esac
 done
 
@@ -18,12 +20,25 @@ thresholding_nf_pipeline=$PWD"/../nextflow_scripts/thresholding.nf"
 ######################
 # 2. Set all arguments
 ######################
-# i. Location of backing .odm files
-backing_files_dir=$LOCAL_XIE_2019_DATA_DIR"processed/"
+# i. Local vs Azure
+if [ $azure = true ]
+then
+  glmeiv_dir="az://main/glmeiv/"
+  xie_2019_dir="az://main/xie-2019/"
+  work_dir="az://main/work"
+  prof="az"
+else
+  glmeiv_dir=$LOCAL_GLMEIV_DATA_DIR
+  xie_2019_dir=$LOCAL_XIE_2019_DATA_DIR
+  work_dir=$LOCAL_GLMEIV_DATA_DIR"work"
+  prof="local"
+fi
+backing_files_dir=$xie_2019_dir"processed/"
+processed_data_dir=$glmeiv_dir"public/xie/data/"
+
+# ii. Location of backing .odm files
 gene_odm=$backing_files_dir"gene/expression_matrix.odm"
 gRNA_odm=$backing_files_dir"gRNA/raw_grouped.odm"
-# ii. Location of processed data
-processed_data_dir=$LOCAL_GLMEIV_DATA_DIR"public/xie/data/"
 # iii. metadata files
 gene_metadata=$processed_data_dir"gene_metadata.rds"
 gRNA_metadata=$processed_data_dir"gRNA_metadata.rds"

### 2021-11-23 text/outline/draft.tex
add decrease ep_tol in Gaussian expt to 1e-7
add some simulation text

@@
 \begin{center} 
-\textbf{A class of exponential family measurement error models, with application to single-cell CRISPR screens}
+\textbf{A class of exponential family measurement error models applied to single-cell CRISPR screens}
 \end{center}
@@
 We recover these phenomena in precise theoretical terms in an idealized Gaussian setting.
-Next, we introduce GLM-EIV (``exponential family errors-in-variables model''), a new method for single-cell CRISPR screen analysis.
+Next, we introduce GLM-EIV (``GLM-based errors-in-variables''), a new method for single-cell CRISPR screen analysis.
 GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed, overcoming limitations of thresholded regression.
-We develop a computational infrastructure to deploy GLM-EIV across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters.
+We develop computational infrastructure to deploy GLM-EIV across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters.
 Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, yielding new biological insights.
@@
 Our work makes two main contributions.
-First, we conduct a careful study of the thresholding method.
+First, we conduct a detailed study of the thresholding method.
 Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian setting.
-Second, we introduce a new method, GLM-EIV (generalized linear model with errors-in-variables), for single-cell CRISPR screen analysis.
+Second, we introduce a new method, GLM-EIV (``GLM-based errors-in-variables''), for single-cell CRISPR screen analysis.
 GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.
@@
 
-We implement several statistical accelerations (that likely are of independent utility) to bring the cost of GLM-EIV down to within an order of magnitude of the thresholding method.
-Finally, we develop a computational infrastructure to deploy GLM-EIV at-scale across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters.
-Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, high multiplicity-of-infection single-cell CRISPR screen datasets, yielding new biological and statistical insights.
+We implement several statistical accelerations (that likely are of independent utility) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method.
+Finally, we develop a Docker-containerized application to deploy GLM-EIV at-scale across tens or hundreds of nodes on clouds (e.g., Microsoft Azure, Amazon Web Services) and high-performance clusters.
+Leveraging this application, we apply GLM-EIV to analyze two recent, large-scale, high multiplicity-of-infection single-cell CRISPR screen datasets, yielding new biological and statistical insights.
 
@@
 $$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m.$$ The term $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration.
-The target of inference, $\beta^m_1$, is the log-transformed fold change in gene expression in response to the perturbation, \blue{controlling for} the technical factors.
+The target of inference, $\beta^m_1$, is the log-transformed fold change in gene expression in response to the perturbation, controlling for the technical factors.
 Fold change (i.e., $\exp(\beta^m_1)$) is the ratio of mean gene expression in perturbed cells to unperturbed cells: $\exp(\beta^m_1) = 1$ indicates no change in expression, while $\exp(\beta^m_1) > 1$ and $\exp(\beta^m_1) < 1$ indicate an increase and decrease in expression in response to the perturbation, respectively (accounting for technical factors).
@@
 
-\subsection{Model}
+\subsection{Model}\label{sec:model}
 \subsubsection*{Negative binomial model}
@@
 We develop a computational infrastructure to apply GLM-EIV to large-scale, single-cell CRISPR screen data.\footnote{Pipeline available at \texttt{github.com/timothy-barry/glmeiv-pipeline}.} The infrastructure leverages Nextflow, a programming language that facilitates building data-intensive pipelines \cite{DITommaso2017}, and ondisc\footnote{Package available at \texttt{github.com/timothy-barry/ondisc}; preprint forthcoming.}, an R package that we developed (in a separate project) to facilitate large-scale computing on single-cell data.
-Nextflow and ondisc together enable the construction of highly portable single-cell pipelines: one can analyze data \textit{out-of-memory} on a laptop or in a \textit{distributed} fashion across tens or hundreds of nodes on a cloud (e.g., Microsoft Azure) or high-performance cluster.
+Nextflow and ondisc together enable the construction of highly portable single-cell pipelines: one can analyze data \textit{out-of-memory} on a laptop or in a \textit{distributed} fashion across tens or hundreds of nodes on a cloud (e.g., Microsoft Azure, Google Cloud) or high-performance cluster.
 
@@
 \section{Simulation studies}\label{sec:simulation}
+
+We conducted a simulation study to compare the empirical performance of GLM-EIV and the thresholding method.
+We generated data on $n = 150,000$ cells from the GLM-EIV model (Section \ref{sec:model}), setting the the target of inference $\beta^m_1$ to $\log(0.25)$ and the probability of perturbation $\pi$ to $0.02$.
+Note that $\beta^m_1 = \log(0.25)$ denotes a reduction in gene expression by a factor of $4$; this is a fairly large effect size on the order of a positive control pair.
+We included ``sequencing batch'' (Bernoulli-distributed) as a covariate and sequencing depth (Poisson-distributed) as an offset.
+We varied the log-fold change in gRNA expression $\beta^g_1$ over a grid defined on $[\log(1), \log(4)];$ $\beta^g_1$ controls problem difficulty, with higher values corresponding to easier problem settings.
+Finally, we generated the gene expression and gRNA count data using two response distributions: Poisson and negative binomial (with fixed size parameter $\theta = 20$).
+For each parameter setting (defined by a $\beta^g_1$-distribution pair), we synthesized $n_\textrm{sim} = 500$ datasets.
+Section \ref{sec:extra_sims} presents additional simulation results on Gaussian response distributions.
+
+We applied three methods to the simulated data: ``standard'' GLM-EIV (Algorithm \ref{algo:em_full}), accelerated GLM-EIV (Algorithm \ref{algo:pilot_estimates}), and thresholded regression.
+We applied the methods to the negative binomial data twice: once taking the size parameter $\theta$ as known, and once taking $\theta$ as unknown.
+In the latter case we used the \texttt{glm.nb} function from the \texttt{MASS} package to estimate $\theta$ \cite{Ripley2013}.
+We display the results of the simulation study in Figure \ref{main_text_sim}.
+Columns correspond to distributions (i.e., Poisson, negative binomial with known $\theta$, and negative binomial with unknown $\theta$), and rows correspond to performance metrics (i.e., bias, mean-squared error, CI coverage rate (nominal rate $95\%$), CI width, and method execution time).
+The parameter $\beta^g_1$ is plotted on the $x$-axis.
+The methods are depicted in different colors (GLM-EIV = red, accelerated GLM-EIV = blue, and thresholding = purple; GLM-EIV masked by accelerated GLM-EIV in several panels).
+
+First, we observed that GLM-EIV uniformly dominated thresholded regression on statistical performance metrics.
+GLM-EIV exhibited lower bias (row 1) and mean-squared error (row 2) than thresholded regression; additionally, GLM-EIV had superior confidence interval coverage (row 3) despite having produced generally narrower confidence intervals (row 4).
+Intuitively, GLM-EIV outperformed the thresholding method because (i) GLM-EIV leveraged information from \textit{both} modalities (rather than the gRNA modality alone) to assign perturbation identities to cells, and (ii) GLM-EIV produced soft rather than hard assignments.
+Additionally, we observed that accelerated GLM-EIV exhibited identical performance to (standard) GLM-EIV on statistical performance metrics (rows 1-4) despite having a substantially reduced computational cost (bottom row).
+In fact, the execution time of accelerated GLM-EIV was approximately within an order of magnitude of that of the thresholding method (bottom row).
+
+% Interestingly, thresholded regression exhibited better CI coverage for theta est.
+rather than theta known (row 3).
+Estimating theta leads to slight \textit{inflation} bias (i.e., \textit{overestimating} the effect size), whereas thresholdeding leads to slight \textit{attenuation} bias (as demonstrated previously).
+These two phenomena cancel, causing the thresholding method to exhibit better CI coverage when estimating theta.
+Our focus on this work is \textit{not} on negative binomial size parameter estimation; GLM-EIV and thresholded regression are (in principle) compatible with \textit{any} negative binomial size parameter estimation procedure, including those based on more sophisticated techniques, such as regularization.
+An interesting future direction would be to assess the performance of the methods under different size parameter estimation approaches.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.9\linewidth]{../../figures/main_text_sim/plot}
+	\caption{\textbf{Simulation study}.
+Columns correspond to distributions (Poisson, NB with known $\theta$, NB with estimated $\theta$), and rows correspond to metrics (bias, MSE, coverage, CI width, and time).
+Methods are shown in different colors; GLM-EIV (red) is masked by accelerated GLM-EIV (blue) in several panels.
+GLM-EIV demonstrated superior statistical performance to the thresholding method across all metrics (rows 1-4).
+Accelerated GLM-EIV had substantially lower computational cost than standard GLM-EIV (bottom row) despite having exhibited identical statistical performance (rows 1-4).}
+	\label{main_text_sim}
+\end{figure}
 
@@
 
-\section{Additional simulation results}
+\section{Additional simulation results}\label{sec:extra_sims}
+\begin{figure}
+	\centering
+	\includegraphics[width=1.0\linewidth]{../../figures/supplement_sim/gaussian}
+	\caption{}
+	\label{fig:gaussian_sim}
+\end{figure}
 

### 2021-11-23 R_scripts/plotting/main_text_sim.R

@@ -65,4 +65,4 @@ p <- ggplot(data = to_plot_all, mapping = aes(x = exp_g_perturbation, y = value,
 
 # save
 fp <- paste0(fig_dir, "/plot.pdf")
-ggsave(filename = fp, plot = p, device = "pdf", scale = 1.1, width = 5, height = 5.5)
+ggsave(filename = fp, plot = p, device = "pdf", scale = 1.4, width = 4.25, height = 5.25)

### 2021-11-23 R_scripts/plotting/supplement_sim.R

@@ -25,9 +25,9 @@ summarized_gaus_results <- summarize_results(sim_spec = gaus_spec, sim_res = gau
 to_plot_all <- summarized_gaus_results %>% filter(metric != "count") %>%
   mutate(metric_fct = factor(metric, levels = c("bias", "mse", "coverage", "ci_width", "time"),
                              labels = c("Bias", "MSE", "Coverage", "CI width", "Time (s)")),
-         Method = factor(method, levels = c("glmeiv_slow", "glmeiv_fast", "thresholding"),
-                         labels = c("GLM-EIV", "GLM-EIV (accelerated)", "Thresholding"))) %>%
-  arrange(Method)
+         Method = factor(method, levels = c("glmeiv_fast", "thresholding"),
+                         labels = c("GLM-EIV (accelerated)", "Thresholding"))) %>%
+  arrange(desc(Method))
 
 p <- ggplot(data = to_plot_all, mapping = aes(x = g_perturbation, y = value, col = Method)) + 
   facet_wrap(. ~ metric_fct, scales = "free_y", nrow = 1) + xlab(expression(beta[g])) + scale_color_manual(values = my_cols) +
@@ -38,4 +38,4 @@ p <- ggplot(data = to_plot_all, mapping = aes(x = g_perturbation, y = value, col
   theme_bw() + theme(legend.position = "bottom", panel.grid.major = element_blank(), panel.grid.minor = element_blank()) + ylab("")
 
 fp <- paste0(fig_dir, "/gaussian.pdf")
-ggsave(filename = fp, plot = p, device = "pdf", scale = 1.1, width = 7, height = 2.5)
+ggsave(filename = fp, plot = p, device = "pdf", scale = 1.0, width = 7, height = 2.5)

### 2021-11-23 text/outline/draft.tex
finish draft of simulations

--- 2021-11-20 text/outline/draft.tex
@@
 We recover these phenomena in precise theoretical terms in an idealized Gaussian setting.
-Next, we introduce GLM-EIV (``exponential family errors-in-variables model''), a new method for single-cell CRISPR screen analysis.
+Next, we introduce GLM-EIV (``GLM-based errors-in-variables''), a new method for single-cell CRISPR screen analysis.
 GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed, overcoming limitations of thresholded regression.
-We develop a computational infrastructure to deploy GLM-EIV across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters.
+We develop a computational infrastructure to deploy GLM-EIV across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters.
 Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, yielding new biological insights.
@@
 
-We implement several statistical accelerations (that likely are of independent utility) to bring the cost of GLM-EIV down to within an order of magnitude of the thresholding method.
-Finally, we develop a computational infrastructure to deploy GLM-EIV at-scale across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters.
-Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, high multiplicity-of-infection single-cell CRISPR screen datasets, yielding new biological and statistical insights.
+We implement several statistical accelerations (that likely are of independent utility) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method.
+Finally, we develop a Docker-containerized application to deploy GLM-EIV at-scale across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters.
+Leveraging this application, we apply GLM-EIV to analyze two recent, large-scale, high multiplicity-of-infection single-cell CRISPR screen datasets, yielding new biological and statistical insights.
 
@@
 Unlike SVD, eSVD models the relationship between the mean and variance of a gene's expression level, a phenomenon induced by the countedness of single-cell data \cite{Lause2021}.
-Similarly, Townes et al.\ \cite{Townes2019} proposed GLM-PCA, a generalization of PCA that directly models negative binomially-distributed gene expression counts while regressing out technical factors.
+Similarly, Townes et al.\ \cite{Townes2019} proposed GLM-PCA, a generalization of PCA that directly models negative binomially-distributed gene expression counts while controlling for technical factors.
 We see our work as a continuation of this broad effort to ``port'' common statistical methods and models to single-cell count data.
@@
 \begin{multline*}
-\sum_{i=1}^n f_m(m_i; \eta^m_i) = \sum_{i=1}^n f_m(m_i; h_m(\beta_0 + \beta_1 p_i + \gamma^T z_i + o^m_i)) \\ = \underbrace{\sum_{p_i =1}^n f_m(m_i; h_m(\beta_0 + \beta_1 + \gamma^T z_i + o^m_i))}_\textrm{few terms}  + \underbrace{\sum_{p_i = 0}^n f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i))}_\textrm{many terms} \\ \approx \sum_{i = 1}^n f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i)).
+\sum_{i=1}^n f_m(m_i; \eta^m_i) = \sum_{i=1}^n f_m(m_i; h_m(\beta_0 + \beta_1 p_i + \gamma^T z_i + o^m_i)) \\ = \underbrace{\sum_{i : p_i =1} f_m(m_i; h_m(\beta_0 + \beta_1 + \gamma^T z_i + o^m_i))}_\textrm{few terms}  + \underbrace{\sum_{i : p_i = 0} f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i))}_\textrm{many terms} \\ \approx \sum_{i = 1}^n f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i)).
 \end{multline*}
@@
 \section{Simulation studies}\label{sec:simulation}
+
+We conducted a simulation study to compare the empirical performance of GLM-EIV to that of the thresholding method.
+We generated data on $n = 150,000$ cells from the GLM-EIV model (Section \ref{sec:model}), setting the the target of inference $\beta^m_1$ to $\log(0.25)$ and the probability of perturbation $\pi$ to $0.02$.
+$\beta^m_1 = \log(0.25)$ represents a decrease in gene expression by a factor of $4$, which is a fairly large effect size on the order of what we might observe for a positive control pair.
+We included ``sequencing batch'' (modeled as a Bernoulli-distributed variable) as a covariate and sequencing depth (modeled as a Poisson-distributed variable) as an offset.
+We varied the log-fold change in gRNA expression, $\beta^g_1,$ over a grid on the interval $[\log(1), \log(4)];$ $\beta^g_1$ controls problem difficulty, with higher values corresponding to easier problems.
+Finally, we generated the gene expression and gRNA count data from two response distributions: Poisson and negative binomial (size parameter fixed at $\theta = 20$ for the latter).
+For each parameter setting (defined by a $\beta^g_1$-distribution pair), we synthesized $n_\textrm{sim} = 500$ i.i.d.\ datasets.
+Section \ref{sec:extra_sims} presents additional simulation results on Gaussian response distributions.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.9\linewidth]{../../figures/main_text_sim/plot}
+	\caption{\textbf{Simulation study}.
+Columns correspond to distributions (Poisson, NB with known $\theta$, NB with estimated $\theta$), and rows correspond to metrics (bias, MSE, coverage, CI width, and time).
+Methods are shown in different colors; GLM-EIV (red) is masked by accelerated GLM-EIV (blue) in several panels.
+GLM-EIV demonstrated superior statistical performance to the thresholding method on all metrics (rows 1-4).
+Accelerated GLM-EIV had substantially lower computational cost than ``vanilla'' GLM-EIV (bottom row) despite demonstrating identical statistical performance (rows 1-4).}
+	\label{main_text_sim}
+\end{figure}
+
+We applied three methods to the simulated data: ``vanilla'' GLM-EIV (Algorithm \ref{algo:em_full}), accelerated GLM-EIV (Algorithm \ref{algo:pilot_estimates}), and thresholded regression.
+We used the Bayes-optimal decision boundary for classification as the threshold for the thresholding method.
+We ran all methods on the negative binomial data twice: once treating the size parameter $\theta$ as a known constant and once treating $\theta$ as unknown.
+In the latter case we used the \texttt{glm.nb} function from the \texttt{MASS} package to estimate $\theta$ \cite{Ripley2013}.
+We display the results of the simulation study in Figure \ref{main_text_sim}.
+Columns correspond to distributions (i.e., Poisson, negative binomial with known $\theta$, and negative binomial with unknown $\theta$), rows correspond to performance metrics (i.e., bias, mean squared error, CI coverage rate (nominal rate $95\%$), CI width, and method execution time).
+The problem difficulty parameter $\beta^g_1$ is plotted on the horizontal axis, and the methods are depicted in different colors (GLM-EIV = red, accelerated GLM-EIV = blue, and thresholding = purple; GLM-EIV masked by accelerated GLM-EIV in several panels).
+
+First, we observed that GLM-EIV dominated thresholded regression on all statistical metrics: GLM-EIV exhibited lower bias (row 1) and mean squared error (row 2) than thresholded regression; additionally, GLM-EIV had superior confidence interval coverage (row 3) despite having produced generally narrower confidence intervals (row 4).
+Intuitively, GLM-EIV outperformed the thresholding method because (i) GLM-EIV leveraged information from \textit{both} modalities (rather than the gRNA modality alone) to assign perturbation identities to cells, and (ii) GLM-EIV produced soft rather than hard assignments, capturing the inherent uncertainty in whether a perturbation occurred.
+Next, we found that accelerated GLM-EIV performed as well as ``vanilla'' GLM-EIV on the statistical metrics (rows 1-4) while substantially reducing computational cost (bottom row).
+In fact, the execution time of accelerated GLM-EIV was approximately within an order of magnitude of that of the thresholding method (bottom row).
+
+Interestingly, thresholded regression exhibited better confidence interval coverage under estimated $\theta$ than under known $\theta$ (row 3).
+Estimating $\theta$ leads to slight inflation bias (i.e., overestimating the true effect size), whereas, as we showed previously, thresholding leads to attenuation bias (i.e., underestimating the true effect size).
+These phenomena partially cancel, yielding less biased estimates.
+We note that GLM-EIV and the thresholding method in principal are compatible with \textit{any} $\theta$ estimation procedure, including those based on more sophisticated techniques, such as regularization \cite{Hafemeister2019}.
+We defer rigorous investigation of the impact of different $\theta$ estimation strategies on these methods to future work.
 
@@
 
-\section{Additional simulation results}
+\section{Additional simulation study}\label{sec:extra_sims}
+We ran an additional simulation study in which we modeled the gene and gRNA expressions using a Gaussian distribution with identity link.
+We generated data on $n = 150,000$ cells, fixing the target of inference $\beta^m_1$ to $-4$ and the probability of perturbation $\pi$ to $0.05$.
+We included ``sequencing batch'' (modeled as a Bernoulli-distributed variable) and ``sequencing depth'' (modeled as a Poisson-distributed variable) as covariates in the model.
+We did not include sequencing depth as an offset because use of the identity link renders offsets meaningless.
+We varied $\beta^g_1$ over a grid on the interval $[0,7].$ We generated $n_\textrm{sim} = 1,000$ synthetic datasets for each value of $\beta^g_1$.
+We applied accelerated GLM-EIV and thresholded regression to the simulated data.
+We assessed these methods on the metrics of bias, mean squared error, confidence interval coverage rate, and confidence interval width.
+We found that accelerated GLM-EIV outperformed the thresholding method: the former method exhibited smaller bias, smaller mean squared error, higher confidence interval coverage rate, and smaller confidence interval width than the latter method (Figure \ref{fig:gaussian_sim}).
+
+\begin{figure}
+	\centering
+	\includegraphics[width=1.0\linewidth]{../../figures/supplement_sim/gaussian}
+	\caption{Additional simulation results on Gaussian data.
+GLM-EIV (accelerated) outperformed the thresholding method on bias, mean squared error, confidence interval coverage rate, and confidence interval width metrics.}\label{fig:gaussian_sim}
+\end{figure}
 

### 2021-11-25 bash_scripts/run_gasp_glmeiv.sh
update passing of workdir to nextflow script in Gasperini data analysis

@@ -20,4 +20,4 @@ nextflow run https://github.com/timothy-barry/glmeiv-pipeline -r main \
 --gRNA_metadata $gRNA_metadata \
 --g_offsets $g_offsets \
 --g_fam_str $g_fam \
--profile az -bg > $PWD/log -ansi-log false -with-trace
+-profile local -bg > $PWD/log -ansi-log false -with-trace -w $work_dir

### 2021-11-25 bash_scripts/run_gasp_thresholding.sh

@@ -14,4 +14,4 @@ nextflow run $thresholding_nf_pipeline --pairs $gRNA_gene_pairs \
 --m_offsets $m_offsets \
 --gRNA_odm $gRNA_odm \
 --gRNA_metadata $gRNA_metadata \
--bg > $PWD/log -ansi-log false -with-trace
+-bg > $PWD/log -ansi-log false -with-trace -w $work_dir

### 2021-11-25 bash_scripts/set_gasp_args.sh

@@ -19,6 +19,7 @@ thresholding_nf_pipeline=$PWD"/../nextflow_scripts/thresholding.nf"
 ######################
 # 2. Set all arguments
 ######################
+work_dir=$LOCAL_GLMEIV_DATA_DIR"work"
 # i. Location of backing .odm files
 backing_files_dir=$LOCAL_GASPERINI_2019_DATA_DIR"at-scale/processed/"
 gene_odm=$backing_files_dir"gene/gasp_scale_gene_expressions.odm"
@@ -33,15 +34,15 @@ covariate_matrix=$processed_data_dir"covariate_matrix.rds"
 m_offsets=$processed_data_dir"m_offsets.rds"
 g_offsets=$processed_data_dir"g_offsets.rds"
 # v. family strings
-m_fam="poisson"
+m_fam="nb"
 g_fam="poisson"
 # vi. pairs to analyze and pod sizes
 if [ $trial = true ]
 then
-  gRNA_gene_pairs=$processed_data_dir"gRNA_gene_pairs_sample.rds"
-  gene_pod_size=3 #500
-  gRNA_pod_size=3 #500
-  pair_pod_size=3 #500
+  gRNA_gene_pairs=$processed_data_dir"gene_gRNA_pairs_sample_problem.rds"
+  gene_pod_size=5 #500
+  gRNA_pod_size=5 #500
+  pair_pod_size=5 #500
 else
   gRNA_gene_pairs=$processed_data_dir"gRNA_gene_pairs.rds"
   gene_pod_size=200

### 2021-11-25 R_scripts/analysis/xie_analysis.R
change pair pod size to 250

@@ -3,7 +3,7 @@ library(ggplot2)
 
 # load the results
 xie_result_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/xie/results/")
-glmeiv_res <- readRDS(paste0(xie_result_dir, "result_glmeiv.rds")) %>% dplyr::distinct() # ensure ahead of time rows are distinct; check gRNA_gene_pairs data frame.
+glmeiv_res <- readRDS(paste0(xie_result_dir, "glmeiv_result.rds")) %>% dplyr::distinct() # ensure ahead of time rows are distinct; check gRNA_gene_pairs data frame.
 thresh_res <- readRDS(paste0(xie_result_dir, "result_thresholding.rds")) %>% dplyr::distinct()
 
 # examine g_perturbation for glmeiv in the different categories

### 2021-11-25 bash_scripts/set_gasp_args.sh

@@ -47,7 +47,7 @@ else
   gRNA_gene_pairs=$processed_data_dir"gRNA_gene_pairs.rds"
   gene_pod_size=200
   gRNA_pod_size=500
-  pair_pod_size=500
+  pair_pod_size=250
 fi
 
 # vii. results directory

### 2021-11-30 R_scripts/processing/qc_xie.R
run resampling analysis on full set of positive control pairs

@@ -11,9 +11,9 @@ gene_odm <- read_odm(odm_fp = paste0(xie_offsite, "gene/expression_matrix.odm"),
 gRNA_odm <- read_odm(odm_fp = paste0(xie_offsite, "gRNA/raw_grouped.odm"),
                      metadata_fp = paste0(xie_offsite, "gRNA/raw_grouped_metadata.rds"))
 
-#######################
-# 2. QC on genes, gRNAs
-#######################
+#############################
+# 2. QC on genes (no gRNA qc)
+#############################
 highly_expressed_genes <- gene_odm %>% get_feature_covariates() %>%
   dplyr::mutate(p_exp = n_nonzero / ncol(gene_odm)) %>%
   dplyr::filter(p_exp >= 0.08) %>% row.names()
@@ -31,13 +31,13 @@ g_offset <- multimodal_odm %>% get_cell_covariates() %>% dplyr::pull(gRNA_n_umis
 ##################
 # 4. Get the pairs
 ##################
-gRNA_gene_pairs <- readRDS(paste0(xie_offsite, "aux/pairs_grouped.rds")) %>% dplyr::filter(protein_coding)
-# sample 5000 cis pairs and 50000 negative control
+gRNA_gene_pairs <- readRDS(paste0(xie_offsite, "aux/pairs_grouped.rds"))
 set.seed(4)
 gRNA_gene_pairs_sub <- dplyr::filter(gRNA_gene_pairs, gene_id %in% get_feature_ids(gene_odm))
 # check for duplication of pair id
 gRNA_gene_pairs_sub %>% dplyr::summarize(pair_id = paste0(gene_id, gRNA_id)) %>%
   dplyr::pull() %>% duplicated() %>% any()
+# combine the cis pairs with 50,000, randomly-selected negative control pairs
 gRNA_gene_sample <- rbind(gRNA_gene_pairs_sub %>% dplyr::filter(type == "cis"),
                           gRNA_gene_pairs_sub %>% dplyr::filter(type == "neg_control") %>% dplyr::slice_sample(n = 50000))
 

### 2021-11-30 text/outline/draft.tex
update passing of workdir to nextflow script in Gasperini data analysis
update Xie qc to make it more similar to gasperini

@@
 
-\section{Generalized linear model with errors in variables}
+\section{GLM-based errors-in-variables}
 
@@
 
+\section{Real data analysis details}
+
+We performed quality control.
+We kept cells with percent mitochondrial reads less than 8$\%$.
+For the Gasperini dataset, we furthermore kept cells with UMI and gRNA counts in the 5th to 95th percentiles to reduce the effect of outliers.
+(We did not perform this latter filter on the Xie data, as the latter dataset appeared to be less noisy and contained fewer cells.) Next, we kept genes expressed in at least $10\%$ of cells with a mean expression level of at least $1$.
+We did not QC the gRNAs.
+
+The quality-controlled Xie dataset contained $n = 101,508$ cells, $1,030$ genes, and $516$ gRNAs.
+There were $50,000$ \textit{in silico} negative control pairs and $681$ candidate \textit{cis} pairs.
+The quality-controlled Gasperini dataset contained $n = 170,645$ cells, $2,079$ genes, $6,598$ gRNAs.
+There were $17,028$ candidate \textit{cis} pairs, $97,818$ negative control pairs, and $322$ positive control pairs.
+
 \end{appendices}

### 2021-11-30 R_scripts/analysis/xie_analysis.R

@@ -3,8 +3,8 @@ library(ggplot2)
 
 # load the results
 xie_result_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/xie/results/")
-glmeiv_res <- readRDS(paste0(xie_result_dir, "glmeiv_result.rds")) %>% dplyr::distinct() # ensure ahead of time rows are distinct; check gRNA_gene_pairs data frame.
-thresh_res <- readRDS(paste0(xie_result_dir, "result_thresholding.rds")) %>% dplyr::distinct()
+# glmeiv_res <- readRDS(paste0(xie_result_dir, "glmeiv_result.rds")) %>% dplyr::distinct()
+# thresh_res <- readRDS(paste0(xie_result_dir, "result_thresholding.rds")) %>% dplyr::distinct()
 
 # examine g_perturbation for glmeiv in the different categories
 glmeiv_res %>% dplyr::filter(parameter == "g_perturbation",

### 2021-11-30 R_scripts/processing/qc_xie.R

@@ -4,36 +4,61 @@ glmeiv_offsite_dir_xie_data <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"),
 xie_offsite <- paste0(.get_config_path("LOCAL_XIE_2019_DATA_DIR"), "processed/")
 
 ##############
-# 1. Load data
+# 0. Load data
 ##############
 gene_odm <- read_odm(odm_fp = paste0(xie_offsite, "gene/expression_matrix.odm"),
                      metadata_fp = paste0(xie_offsite, "gene/metadata.rds"))
 gRNA_odm <- read_odm(odm_fp = paste0(xie_offsite, "gRNA/raw_grouped.odm"),
                      metadata_fp = paste0(xie_offsite, "gRNA/raw_grouped_metadata.rds"))
 
-#############################
-# 2. QC on genes (no gRNA qc)
-#############################
-highly_expressed_genes <- gene_odm %>% get_feature_covariates() %>%
-  dplyr::mutate(p_exp = n_nonzero / ncol(gene_odm)) %>%
-  dplyr::filter(p_exp >= 0.08) %>% row.names()
-gene_odm <- gene_odm[highly_expressed_genes,]
-
-##########################################
-# 3. Global cell covariate matrix, offsets
-##########################################
+##################
+# 1. QC on cells
+##################
 multimodal_odm <- multimodal_ondisc_matrix(list(gene = gene_odm, gRNA = gRNA_odm))
-global_covariate_matrix <- multimodal_odm %>% get_cell_covariates() %>%
-  dplyr::select(batch = gene_batch)
-m_offset <- multimodal_odm %>% get_cell_covariates() %>% dplyr::pull(gene_n_umis) %>% log()
-g_offset <- multimodal_odm %>% get_cell_covariates() %>% dplyr::pull(gRNA_n_umis) %>% log()
 
-##################
-# 4. Get the pairs
-##################
+# determine which cells contain fewer than 0.08 p_mito
+ok_cells <- multimodal_odm %>% get_cell_covariates() %>%
+  dplyr::summarize(ok = gene_p_mito < 0.08) %>% dplyr::pull(ok)
+
+# subset the multimodal ODM according to QC metrics
+multimodal_odm_sub <- multimodal_odm[,ok_cells]
+
+# obtain global covariate matrix
[…]

### 2021-11-30 bash_scripts/run_xie_glmeiv.sh

@@ -41,4 +41,4 @@ nextflow run https://github.com/timothy-barry/glmeiv-pipeline -r main \
  --g_offsets $g_offsets \
  --g_fam_str $g_fam \
  --result_file_name $result_file_name \
- -bg > $PWD/log -ansi-log false -with-trace -profile $prof -w $work_dir
+ -bg > $PWD/log -ansi-log false -with-trace -w $work_dir

### 2021-11-30 bash_scripts/run_all.sh
update run_all

@@ -25,10 +25,11 @@ bash run_gasp_thresh.sh
 # 2. Xie analysis
 #################
 # QC Xie data
-
-# run glmeiv at scale on Xie data
 Rscript ../R_scripts/analysis/qc_xie.R
-Rscript run_xie_glmeiv.sh
+# run glmeiv at scale on Xie data
+bash run_xie_glmeiv.sh
+# run thresholding method at scale on Xie data
+bash run_xie_thresh.sh
 
 ################
 # 3. Simulations
@@ -37,3 +38,7 @@ Rscript ../R_scripts/simulations/create_simspec_objects.R
 $SIMULATR -f $sim_dir"spec_objects/sim_spec_0.rds" -r $sim_dir"results/raw_result_0.rds"
 $SIMULATR -f $sim_dir"spec_objects/sim_spec_1.rds" -r $sim_dir"results/raw_result_1.rds"
 $SIMULATR -f $sim_dir"spec_objects/sim_spec_2.rds" -r $sim_dir"results/raw_result_2.rds"
+
+#############
+# 4. Plotting
+#############

### 2021-11-30 bash_scripts/run_all.sh
update run_all so that file paths are correct

@@ -15,7 +15,7 @@ sim_dir=$LOCAL_GLMEIV_DATA_DIR"public/simulations/"
 # 1. Gasperini analysis
 #######################
 # QC Gasperini data
-Rscript ../R_scripts/analysis/qc_gasperini.R
+Rscript ../R_scripts/processing/qc_gasperini.R
 # run glmeiv at scale on Gasperini data
 bash run_gasp_glmeiv.sh
 # run thresholding method at scale on Gasperini data
@@ -25,7 +25,7 @@ bash run_gasp_thresh.sh
 # 2. Xie analysis
 #################
 # QC Xie data
-Rscript ../R_scripts/analysis/qc_xie.R
+Rscript ../R_scripts/processing/qc_xie.R
 # run glmeiv at scale on Xie data
 bash run_xie_glmeiv.sh
 # run thresholding method at scale on Xie data

### 2021-12-01 R_scripts/plotting/data_analysis.R
add code for real data plots

@@ -1,0 +1,55 @@
+library(dplyr)
+library(ggplot2)
+
+my_cols <- c("firebrick3", "dodgerblue3", "orchid4")
+fig_dir <- paste0(.get_config_path("LOCAL_CODE_DIR"), "glmeiv-manuscript/figures/data_analysis")
+if (!dir.exists(fig_dir)) dir.create(fig_dir)
+result_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/gasperini/results")
+
+########################
+# 1. Resampling analysis
+########################
+resampling_df <- readRDS(paste0(result_dir, "/resampling_result.rds"))
+# a function to plot the result given a given pair
+f <- function(resampling_df, pair_id, x_max = 0.5) {
+  df <- dplyr::filter(resampling_df, pair_id == !!pair_id, contam_level <= x_max) %>%
+    dplyr::filter(parameter == "m_perturbation", target %in% c("estimate", "confint_lower", "confint_upper")) %>%
+    dplyr::select(-parameter, -pair_id) %>%
+    tidyr::pivot_wider(data = ., names_from = "target", values_from = "value") %>%
+    dplyr::mutate(method = factor(method, c("glmeiv", "thresholding"), c("GLM-EIV", "Thresholding")))
+  df_synth <- df %>% dplyr::filter(contam_level >= 0) %>% dplyr::group_by(method, contam_level) %>%
+    dplyr::summarize(m_est = mean(estimate, na.rm = TRUE), m_ci_upper = mean(confint_upper, na.rm = TRUE), m_ci_lower = mean(confint_lower, na.rm = TRUE))
+  df_gt <- df %>% dplyr::filter(contam_level == -1)
+  p <- ggplot(data = df_synth, mapping = aes(x = contam_level, y = m_est)) + facet_grid(.~method) +
+    geom_hline(data = df_gt, mapping = aes(yintercept = estimate), color = my_cols[2], lwd = 0.75) +
+    theme_bw() + xlab("Excess background contamination") + 
+    scale_color_manual(values=c("grey70", "grey70", "black")) + ylab("Estiamte") + geom_ribbon(aes(ymin = m_ci_lower, ymax = m_ci_upper), fill = "grey70", alpha = 0.5) + geom_line(lwd = 0.75) +
[…]

### 2021-12-02 bash_scripts/run_xie_glmeiv.sh
add resume to pipeline

@@ -41,4 +41,4 @@ nextflow run https://github.com/timothy-barry/glmeiv-pipeline -r main \
  --g_offsets $g_offsets \
  --g_fam_str $g_fam \
  --result_file_name $result_file_name \
- -bg > $PWD/log -ansi-log false -with-trace -w $work_dir
+ -bg > $PWD/log -ansi-log false -with-trace -w $work_dir -resume

### 2021-12-03 text/outline/draft.tex
ensure up-to-date
change pair pod size to 200

@@
 \begin{center} 
-\textbf{A class of exponential family measurement error models applied to single-cell CRISPR screens}
+\textbf{A class of exponential family measurement error models applied to CRISPR genome editing and single-cell sequencing}
 \end{center}
@@
 
-We implement several statistical accelerations (that likely are of independent utility) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method.
+We implement several statistical accelerations (that possibly are independently useful) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method.
 Finally, we develop a Docker-containerized application to deploy GLM-EIV at-scale across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters.
@@
 
+% We anticipate three practical uses of the GLM-EIV in single-cell CRISPR screen analysis: first, determining the extent of background contamination, which will inform subsequent analysis decisions.
+Second, if the amount of background contamination is sufficiently high, thresholding will fail, and so GLM-EIV should be used to get a good answer.
+On the other hand, if the amount of background contamination is low, we would use the two methods in conjunction: first, fit GLM-EIV to some subset of the data (e.g., the positive control pairs) to get the approximate Bayes-optimal decision boundary; and second, using this decision boundary in the thresholding method.
+As we have shown, the threshold exerts a substantial impact on the results.
+In this way, GLM-EIV becomes a threshold selection procedure, enabling statistically rigorous and scientifically reproducible threshold selection procedure, eliminating the aribrariness of selecting a threshold by, for example, eye.
+
 \section{Discussion}

### 2021-12-03 R_scripts/plotting/data_analysis.R

@@ -15,6 +15,7 @@ result_dir_gasp <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/gas
 ########################
 # 1. Resampling analysis
 ########################
+n_pairs <- resampling_df$pair_id %>% unique() %>% length()
 resampling_df <- readRDS(paste0(result_dir_gasp, "/resampling_result.rds"))
 # a function to plot the result given a given pair
 f <- function(resampling_df, pair_id, x_max = 0.5) {
@@ -35,20 +36,23 @@ f <- function(resampling_df, pair_id, x_max = 0.5) {
   return(p)
 }
 pair_ids <- as.character(unique(resampling_df$pair_id))
-p1 <- f(resampling_df, pair_ids[10], x_max = 0.4)
+p1 <- f(resampling_df, pair_ids[37], x_max = 0.4) # good choices: 37
 
 # an "aggregate" version of the above analysis
-aggregate_df <- resampling_df %>% filter(contam_level >= 0, parameter == "m_perturbation", target == "estimate") %>%
-  select(-parameter, -target) %>% group_by(method, pair_id, contam_level) %>%
-  summarize(m = mean(value)) %>% ungroup() %>% group_by(method, pair_id) %>%
+aggregate_df <- resampling_df %>% filter(contam_level >= 0, parameter == "m_perturbation", target %in% c("estimate", "confint_lower", "confint_upper")) %>%
+  select(-parameter) %>% tidyr::pivot_wider(names_from = "target", values_from = "value") %>% group_by(method, pair_id, contam_level) %>%
+  summarize(m = mean(estimate), m_lower_ci = mean(confint_lower, na.rm = TRUE), m_upper_ci = mean(confint_upper, na.rm = TRUE)) %>% ungroup() %>% group_by(method, pair_id) %>%
   group_modify(.f = function(tbl, key) {
     baseline_est <- tbl$m[tbl$contam_level == 0]
     p_change <- abs(tbl$m - baseline_est)/baseline_est
-    mutate(tbl, p_change = p_change)
+    ci_cover <- (tbl$m_lower_ci < baseline_est) & (tbl$m_upper_ci > baseline_est)
+    mutate(tbl, p_change = p_change, ci_cover = ci_cover)
   }) %>% ungroup() %>% group_by(contam_level, method) %>% summarize(median_p_change = median(p_change),
-                                                                    mean_p_change = mean(p_change),
[…]

### 2021-12-03 bash_scripts/set_gasp_args.sh

@@ -50,7 +50,7 @@ else
   gRNA_gene_pairs=$processed_data_dir"gRNA_gene_pairs.rds"
   gene_pod_size=200
   gRNA_pod_size=500
-  pair_pod_size=250
+  pair_pod_size=200
   B=50
   result_file_name="glmeiv_result.rds"
 fi

### 2021-12-03 bash_scripts/set_xie_args.sh

@@ -61,7 +61,7 @@ else
   gRNA_gene_pairs=$processed_data_dir"gRNA_gene_pairs.rds"
   gene_pod_size=200
   gRNA_pod_size=500
-  pair_pod_size=500
+  pair_pod_size=200
   result_file_name="glmeiv_result.rds"
 fi
 

### 2021-12-03 text/outline/draft.tex
update real data plotting code

@@
 \begin{center} 
-\textbf{A class of exponential family measurement error models applied to CRISPR genome editing and single-cell sequencing}
+\textbf{A class of exponential family measurement error models, with application to CRISPR genome editing and single-cell sequencing}
 \end{center}
@@
 Next, we introduce GLM-EIV (``GLM-based errors-in-variables''), a new method for single-cell CRISPR screen analysis.
-GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed, overcoming limitations of thresholded regression.
+GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed, overcoming limitations of thresholded regression in challenging problem settings.
 We develop a computational infrastructure to deploy GLM-EIV across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters.
@@
 
-\subsection{Computation}
-We develop a computational infrastructure to apply GLM-EIV to large-scale, single-cell CRISPR screen data.\footnote{Pipeline available at \texttt{github.com/timothy-barry/glmeiv-pipeline}.} The infrastructure leverages Nextflow, a programming language that facilitates building data-intensive pipelines \cite{DITommaso2017}, and ondisc\footnote{Package available at \texttt{github.com/timothy-barry/ondisc}; preprint forthcoming.}, an R package that we developed (in a separate project) to facilitate large-scale computing on single-cell data.
+\subsection{Computing}
+We developed a computational infrastructure to apply GLM-EIV to large-scale, single-cell CRISPR screen data.\footnote{Pipeline available at \texttt{github.com/timothy-barry/glmeiv-pipeline}.} The infrastructure leverages Nextflow, a programming language that facilitates building data-intensive pipelines \cite{DITommaso2017}, and ondisc\footnote{Package available at \texttt{github.com/timothy-barry/ondisc}; preprint forthcoming.}, an R package that we developed (in a separate project) to facilitate large-scale computing on single-cell data.
 Nextflow and ondisc together enable the construction of highly portable single-cell pipelines: one can analyze data \textit{out-of-memory} on a laptop or in a \textit{distributed} fashion across tens or hundreds of nodes on a cloud (e.g., Microsoft Azure, Google Cloud) or high-performance cluster.
@@
 Intuitively, GLM-EIV outperformed the thresholding method because (i) GLM-EIV leveraged information from \textit{both} modalities (rather than the gRNA modality alone) to assign perturbation identities to cells, and (ii) GLM-EIV produced soft rather than hard assignments, capturing the inherent uncertainty in whether a perturbation occurred.
-Next, we found that accelerated GLM-EIV performed as well as ``vanilla'' GLM-EIV on the statistical metrics (rows 1-4) while substantially reducing computational cost (bottom row).
+Next, we found that accelerated GLM-EIV performed as well as ``vanilla'' GLM-EIV on the statistical metrics (rows 1-4) despite have substantially lower computational cost (bottom row).
 In fact, the execution time of accelerated GLM-EIV was approximately within an order of magnitude of that of the thresholding method (bottom row).
@@
 \section{Discussion}
+
+We see the practical use of this method as follows.
+First, deploy GLM-EIV to some random subset of pairs to assess the extent of background contamination.
+If the background contamination sufficiently high, apply GLM-EIV to analyze the data directly.
+On the other hand, if the background contamination is low, the thresholding method is an appealing method due to its speed and stability.
+A challenge, as we have shown, is that the selected threshold exerts a substantial impact on the results.
+GLM-EIV approximates the Bayes-optimal decision boundary, which, as we showed in an idealized Gaussian setting, minimizes attenuation bias when the gRNA count distribution is sufficiently-well separated.
+Therefore, in problem settings where thresholding is tenable, GLM-EIV provides a statistically principled, scientifically reproducible, empirically effective threshold selection procedure.
+
+To our knowledge, this is the first statistical work to study single-cell CRISPR screens.
+One of our objectives is to bring to the attention of the statistical community an emerging class of functional genomics assays that likely will exert a major impact on biology research.
+Additionally, this is the first work to leverage the ondisc-Nextflow-cloud/HPC technology stack, a powerful and user-friendly set of tools for large-scale single-cell analysis that other researchers (especially statisticians) likely will find useful.
 

### 2021-12-03 R_scripts/plotting/data_analysis.R

@@ -2,10 +2,12 @@ library(dplyr)
 library(ggplot2)
 library(ggpmisc)
 library(tidyr)
+library(cowplot)
+gasp_res <- readr::read_tsv(paste0(file = .get_config_path("LOCAL_GASPERINI_2019_DATA_DIR"), "at-scale/raw/GSE120861_all_deg_results.at_scale.txt"))
+pair_info <- gasp_res %>% select(pairs4merge, ENSG, gene_short_name, gRNA_group, target_site.chr, target_site.start, target_site.stop)
 
 my_theme <- theme(panel.border = element_blank(), panel.grid.major = element_blank(),
                   panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"), panel.background = element_rect("white"))
-
 
 my_cols <- c("firebrick3", "dodgerblue3", "orchid4")
 fig_dir <- paste0(.get_config_path("LOCAL_CODE_DIR"), "glmeiv-manuscript/figures/data_analysis")
@@ -17,8 +19,13 @@ result_dir_gasp <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/gas
 ########################
 n_pairs <- resampling_df$pair_id %>% unique() %>% length()
 resampling_df <- readRDS(paste0(result_dir_gasp, "/resampling_result.rds"))
-# a function to plot the result given a given pair
-f <- function(resampling_df, pair_id, x_max = 0.5) {
+
+# a function that plots the result given a given pair on the same plot
+f2 <- function(resampling_df, pair_id, x_max = 0.4, pair_info) {
+  split_pair <- strsplit(pair_id, ":")[[1]]
+  my_info <- pair_info %>% filter(ENSG == split_pair[1], gRNA_group == split_pair[2])
+  title <- paste0("Gene ", my_info$gene_short_name)
+  
   df <- dplyr::filter(resampling_df, pair_id == !!pair_id, contam_level <= x_max) %>%
     dplyr::filter(parameter == "m_perturbation", target %in% c("estimate", "confint_lower", "confint_upper")) %>%
     dplyr::select(-parameter, -pair_id) %>%
@@ -27,16 +34,19 @@ f <- function(resampling_df, pair_id, x_max = 0.5) {
   df_synth <- df %>% dplyr::filter(contam_level >= 0) %>% dplyr::group_by(method, contam_level) %>%
[…]

### 2021-12-03 bash_scripts/run_xie_glmeiv.sh
add azure option back to pipeline invoke script

@@ -41,4 +41,4 @@ nextflow run https://github.com/timothy-barry/glmeiv-pipeline -r main \
  --g_offsets $g_offsets \
  --g_fam_str $g_fam \
  --result_file_name $result_file_name \
- -bg > $PWD/log -ansi-log false -with-trace -w $work_dir -resume
+ -bg > $PWD/log -ansi-log false -with-trace -w $work_dir -resume $prof

### 2021-12-03 bash_scripts/set_xie_args.sh

@@ -26,12 +26,12 @@ then
   glmeiv_dir="az://main/glmeiv/"
   xie_2019_dir="az://main/xie-2019/"
   work_dir="az://main/work"
-  prof="az"
+  prof="-profile az"
 else
   glmeiv_dir=$LOCAL_GLMEIV_DATA_DIR
   xie_2019_dir=$LOCAL_XIE_2019_DATA_DIR
   work_dir=$LOCAL_GLMEIV_DATA_DIR"work"
-  prof="local"
+  prof=""
 fi
 backing_files_dir=$xie_2019_dir"processed/"
 processed_data_dir=$glmeiv_dir"public/xie/data/"

### 2021-12-04 bash_scripts/run_gasp_glmeiv.sh
remove -resume optoin

@@ -20,4 +20,4 @@ nextflow run https://github.com/timothy-barry/glmeiv-pipeline -r main \
 --gRNA_metadata $gRNA_metadata \
 --g_offsets $g_offsets \
 --g_fam_str $g_fam \
--profile local -bg > $PWD/log -ansi-log false -with-trace -w $work_dir
+-bg > $PWD/log -ansi-log false -with-trace -w $work_dir

### 2021-12-04 bash_scripts/run_xie_glmeiv.sh

@@ -41,4 +41,4 @@ nextflow run https://github.com/timothy-barry/glmeiv-pipeline -r main \
  --g_offsets $g_offsets \
  --g_fam_str $g_fam \
  --result_file_name $result_file_name \
- -bg > $PWD/log -ansi-log false -with-trace -w $work_dir -resume $prof
+ -bg > $PWD/log -ansi-log false -with-trace -w $work_dir $prof

### 2021-12-04 text/outline/draft.tex
enable submission via sbatch

@@
 
-We implement several statistical accelerations (that possibly are independently useful) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method.
+We implement several statistical accelerations (that possibly are of independent utility) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method.
 Finally, we develop a Docker-containerized application to deploy GLM-EIV at-scale across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters.
-Leveraging this application, we apply GLM-EIV to analyze two recent, large-scale, high multiplicity-of-infection single-cell CRISPR screen datasets, yielding new biological and statistical insights.
+Leveraging this application, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets.
+% In highly noisy settings ; by contrast, in less noisy settings, GLM-EIV serves as a principled and reproducible threshold selection procedure.
 
@@
 One of our objectives is to bring to the attention of the statistical community an emerging class of functional genomics assays that likely will exert a major impact on biology research.
-Additionally, this is the first work to leverage the ondisc-Nextflow-cloud/HPC technology stack, a powerful and user-friendly set of tools for large-scale single-cell analysis that other researchers (especially statisticians) likely will find useful.
+Additionally, this is the first work to leverage the ondisc-Nextflow-cloud/HPC technology stack, a set of programming tools for large-scale single-cell analysis that we view as powerful and user-friendly.
 

### 2021-12-04 bash_scripts/run_gasp_glmeiv.sh

@@ -1,9 +1,16 @@
+#!/bin/bash
+#SBATCH -p RM-shared
+#SBATCH -t 48:00:00
+#SBATCH --ntasks-per-node=2
+
 source set_gasp_args.sh
+export NXF_OPTS="-Xms500M -Xmx2G" # limit NF to 2 GB of memory
 
 #######################
 # Run Nextflow pipeline
 #######################
 rm -f trace.txt
+
 # nextflow run $glmeiv_nf_pipeline \
 nextflow run https://github.com/timothy-barry/glmeiv-pipeline -r main \
 --pairs $gRNA_gene_pairs \

### 2021-12-04 bash_scripts/run_xie_glmeiv.sh

@@ -1,4 +1,10 @@
+#!/bin/bash
+#SBATCH -p RM-shared
+#SBATCH -t 48:00:00
+#SBATCH --ntasks-per-node=2
+
 source set_xie_args.sh
+export NXF_OPTS="-Xms500M -Xmx2G" # limit NF to 2 GB of memory
 
 #######################
 # Run Nextflow pipeline

### 2021-12-06 bash_scripts/run_gasp_glmeiv.sh
decrease n pairs to 150 in gasp analysis

@@ -27,4 +27,4 @@ nextflow run https://github.com/timothy-barry/glmeiv-pipeline -r main \
 --gRNA_metadata $gRNA_metadata \
 --g_offsets $g_offsets \
 --g_fam_str $g_fam \
--bg > $PWD/log -ansi-log false -with-trace -w $work_dir
+-with-trace -w $work_dir

### 2021-12-06 bash_scripts/run_xie_glmeiv.sh

@@ -47,4 +47,4 @@ nextflow run https://github.com/timothy-barry/glmeiv-pipeline -r main \
  --g_offsets $g_offsets \
  --g_fam_str $g_fam \
  --result_file_name $result_file_name \
- -bg > $PWD/log -ansi-log false -with-trace -w $work_dir $prof
+ -with-trace -w $work_dir $prof

### 2021-12-06 bash_scripts/set_gasp_args.sh

@@ -50,7 +50,7 @@ else
   gRNA_gene_pairs=$processed_data_dir"gRNA_gene_pairs.rds"
   gene_pod_size=200
   gRNA_pod_size=500
-  pair_pod_size=200
+  pair_pod_size=150
   B=50
   result_file_name="glmeiv_result.rds"
 fi

### 2021-12-06 R_scripts/analysis/approx_bayes_bdy.R
enable use of custom, command-line threshold in thresholding pipeline

@@ -1,0 +1,39 @@
+library(magrittr)
+load_all("~/research_code/glmeiv/")
+library(dplyr)
+
+########
+# 1. XIE
+########
+# load the results
+xie_result_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/xie/results/")
+glmeiv_res <- paste0(xie_result_dir, "glmeiv_result.rds") %>% readRDS()
+med_est <- glmeiv_res %>% filter(type == "cis", target == "estimate") %>% select(parameter, value, pair_id) %>%
+  tidyr::pivot_wider(data = ., id_cols = "pair_id", names_from = "parameter", values_from = "value") %>% 
+  select(-pair_id) %>% summarize_all(median) %>% select(., starts_with("g"), "pi")
+
+# Load the data
+xie_data_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/xie/data/")
+covariate_matrix <- paste0(xie_data_dir, "covariate_matrix.rds") %>% readRDS()
+g_offset <- paste0(xie_data_dir, "g_offset.rds") %>% readRDS()
+g_fam <- poisson() %>% augment_family_object()
+
+# set arguments to Bayes-opt function
+g_intercept <- med_est$g_intercept %>% log()
+g_pert <- med_est$g_perturbation %>% log()
+g_coef <- med_est[, c("g_batchbatch_2", "g_batchbatch_3", "g_batchbatch_4", "g_batchbatch_5", "g_p_mito")] %>% as.numeric()
+pi <- med_est$pi
+
+# compute median bayes-optimal decision boundary
+bdy <- get_optimal_threshold(g_intercept = g_intercept, g_perturbation = g_pert, g_fam = g_fam, pi = pi,
+                      covariate_matrix = NULL, g_covariate_coefs = NULL, g_offset = median(g_offset)) %>% round()
+f_name <- paste0(xie_data_dir, "dec_bdy.txt")
+if (file.exists(f_name)) file.remove(f_name)
+file_con <- file(f_name)
+writeLines(text = as.character(bdy), con = file_con)
+close(file_con)
+
+##############
+# 2. GASPERINI
+##############
+

### 2021-12-06 bash_scripts/run_xie_thresholding.sh

@@ -14,4 +14,5 @@ nextflow run $thresholding_nf_pipeline --pairs $gRNA_gene_pairs \
 --m_offsets $m_offsets \
 --gRNA_odm $gRNA_odm \
 --gRNA_metadata $gRNA_metadata \
--bg > $PWD/log -ansi-log false -with-trace
+--threshold $thresh \
+-bg > $PWD/log -ansi-log false -with-trace -w $work_dir

### 2021-12-06 bash_scripts/set_xie_args.sh

@@ -67,3 +67,6 @@ fi
 
 # vii. results directory
 result_dir=$LOCAL_GLMEIV_DATA_DIR"public/xie/results"
+
+# viii. threshold
+thresh=$(cat $processed_data_dir"dec_bdy.txt")

### 2021-12-06 nextflow_scripts/bin/run_analysis.R

@@ -1,8 +1,3 @@
-##############################
-# 0. Set a few hyperparameters
-##############################
-threshold <- "adaptive"
-
 ########################################
 # 1. Load packages and command-line args
 ########################################
@@ -16,7 +11,8 @@ gene_metadata_fp <- args[3L]
 m_offsets_fp <- args[4L]
 gRNA_odm_fp <- args[5L]
 gRNA_metadata_fp <- args[6L]
-other_args <- args[seq(7L, n_args)]
+threshold <- as.numeric(args[7L])
+other_args <- args[seq(8L, n_args)]
 
 #########################################
 # 2. Load ODMs, covariate matrix, offsets
@@ -47,15 +43,16 @@ for (i in seq(1L, n_pairs)) {
   if (i == 1 || gRNA_ids[i] != gRNA_ids[i - 1]) { # likewise for gRNAs
     g <- as.numeric(gRNA_odm[[gRNA,]])
   }
+  
   # loop through different thresholds, fitting model to each
-  # perform threshold
-  if (is.numeric(threshold)) {
-    curr_threshold <- threshold
-  } else {
-    g_prime <- g[g >= 1]
-    curr_threshold <- mean(g_prime)
-  }
-  phat <- as.integer(g >= curr_threshold)
+  # if (is.numeric(threshold)) {
+  #  curr_threshold <- threshold
+  # } else {
+  #  g_prime <- g[g >= 1]
+  #  curr_threshold <- mean(g_prime)
+  # }
+  
+  phat <- as.integer(g >= threshold)
   # run thresholding method
   fit <- glmeiv::run_thresholding_method(phat = phat, m = m, m_fam = m_fam,
                                          m_offset = m_offset, covariate_matrix = covariate_matrix,

### 2021-12-07 text/outline/draft.tex
decrease n pairs to 150 in gasp analysis
update text

@@
 \usepackage{tabularx}
+\usepackage{amssymb}
 \usepackage{array}
@@
 
-\subsubsection*{Full GLM-EIV model}
+\subsubsection*{General model}
 To provide greater modeling flexibility, we generalize the GLM-EIV model to arbitrary exponential family response distributions and link functions.
@@
 \end{equation}
-The log-likelihood is
+The complete-data log-likelihood is
 \begin{multline}\label{full_log_lik}
@@
 \end{multline}
-Integrating over the unobserved variable $p_i$, we can write the marginal density $f$ of $(m_i, g_i)$ as
+Integrating over the unobserved variable $p_i$, we can write the density $f$ of $(m_i, g_i)$ as
 \begin{equation}\label{marginal_density}
@@
 \end{equation}
-Finally, the marginal log-likelihood is
+Finally, the log-likelihood is
 \begin{multline}\label{marginal_log_lik}
@@
 where $x^*_i \sim \textrm{Bern}(\pi), \ep_i \sim N(0,1), \tau_i \sim N(0,1),$ and $\ep_i$,$\tau_i$, and $x^*_i$ are independent.
-The model (\ref{classical_eiv}) is a special case of the GLM-EIV model.
-More generally, GLM-EIV allows the use of regression functions (that optionally include covariates and use nonlinear links) to model $y_i$ and $x_i$; $y_i$ and $x_i$ need not be Gaussian.
+The model (\ref{classical_eiv}) is a GLM-EIV model with identity link, Gaussian response, and no covariate terms.
 
@@
 
-\subsection{Computing}
+\subsection{Computing}\label{sec:computing}
 We developed a computational infrastructure to apply GLM-EIV to large-scale, single-cell CRISPR screen data.\footnote{Pipeline available at \texttt{github.com/timothy-barry/glmeiv-pipeline}.} The infrastructure leverages Nextflow, a programming language that facilitates building data-intensive pipelines \cite{DITommaso2017}, and ondisc\footnote{Package available at \texttt{github.com/timothy-barry/ondisc}; preprint forthcoming.}, an R package that we developed (in a separate project) to facilitate large-scale computing on single-cell data.
@@
 
-\section{Real data analysis}
-
-% We anticipate three practical uses of the GLM-EIV in single-cell CRISPR screen analysis: first, determining the extent of background contamination, which will inform subsequent analysis decisions.
-Second, if the amount of background contamination is sufficiently high, thresholding will fail, and so GLM-EIV should be used to get a good answer.
-On the other hand, if the amount of background contamination is low, we would use the two methods in conjunction: first, fit GLM-EIV to some subset of the data (e.g., the positive control pairs) to get the approximate Bayes-optimal decision boundary; and second, using this decision boundary in the thresholding method.
-As we have shown, the threshold exerts a substantial impact on the results.
-In this way, GLM-EIV becomes a threshold selection procedure, enabling statistically rigorous and scientifically reproducible threshold selection procedure, eliminating the aribrariness of selecting a threshold by, for example, eye.
+\section{Data analysis}
+
+Leveraging our computational infrastructure (Section \ref{sec:computing}), we applied GLM-EIV to analyze the entire Gasperini and Xie datasets (Figure \ref{fig:real_data}).
+Our theoretical analyses (Section \ref{sec:thresholding_theory}) and simulation studies (Section \ref{sec:simulation}) revealed that the Bayes-optimal decision boundary is a good choice for the threshold when the gRNA count distribution is well separated.
+Therefore, under the assumption that the effect of the perturbation on gRNA expression is similar across pairs, we used the fitted GLM-EIV models to approximate the Bayes-optimal decision boundary for both datasets, yielding a boundary of $3$ for Gasperini and $7$ for Xie.
+We then applied the thresholding method to analyze both datasets, setting the threshold to the Bayes-optimal decision boundary estimated using GLM-EIV.
+
+% We compared GLM-EIV to thresholded regression on the real data and surprisingly found that the two methods produced similar results.
+Figures \ref{fig:real_data}a and \ref{fig:real_data}b show the estimate of the target of inference $\exp(\beta^m_1)$
+
+To better understand this result, we examined the estimated effect of the perturbation on gRNA expression, $\exp(\hat{\beta}_1^g)$; this parameter plays an important role in determining problem difficulty.
+The value of $\exp(\hat{\beta}_1^g)$ was greater than expected: the 95\% CI for this parameter averaged across pairs was $[XX,XX]$ and $[300, 316]$ on the Gasperini and Xie data, respectively.
+We concluded that both datasets lay in an ``easy'' region of the parameter space, making thresholding a tenable strategy (provided the threshold is selected well).
+
+To illuminate the settings in which GLM-EIV and the thresholding method diverge, we artificially increased the problem difficulty by generating partially-synthetic datasets.
+To this end, we sampled gRNA counts from a realistic model, holding fixed the raw gene expressions, library sizes, covariates, and fitted perturbation probabilities.
+We defined a normalized parameter called ``excess background contamination'' that controlled the problem difficulty.
+When excess background contamination was equal to zero, we sampled gRNA counts directly from the fitted model.
+As excess background contamination increased, we sampled gRNA counts from a slightly modified version of the fitted model in which the mean gRNA expression of \textit{unperturbed} cells was increased while the mean gRNA expression of \textit{perturbed} cells was held constant.
+Finally, when excess background contamination was equal to $1$ (the maximum value), we set the mean gRNA expression of unperturbed cells equal to that of perturbed cells and sampled from the resulting model.
+
+We generated partially-synthetic data for each of the 322 positive control pairs in the Gasperini dataset, varying excess background contamination over the interval $[0,0.4].$ We applied GLM-EIV and the thresholding method to analyze these data.
+We present results on two representative pairs (the pair containing gene \textit{LRIF1} and the pair containing gene \textit{NDUFA2}) in Figures \ref{fig:real_data}c-d.
+First, we observed that the estimate produced by the methods on the raw data (depicted as a horizontal black line) coincided almost exactly with the estimate produced by the methods on the partially-synthetic data generated by setting excess background contamination to zero (trend present across all genes; average relative difference $0.003$).
+This indicates that the fitted model accurately tracked the gRNA count distribution.
+Next, we observed that, as excess background contamination increased, the thresholding method exhibited severe attenuation bias, whereas GLM-EIV remained roughly stable.
+
+We conducted a similar analysis on the entire set of positive control pairs.
+First, for each pair, we computed the ``relative estimate change'' (REC), defined as the relative difference between the estimate at some level of excess contamination and zero excess contamination (Figure \ref{fig:real_data}d).
+% We plotted median
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.9\linewidth]{../../figures/data_analysis/annotated_plot_crop}
+	\caption{\textbf{Data analysis}.
+\textbf{a-b}, Fold change estimates produced by GLM-EIV and thresholded regression on Gasperini (\textbf{a}) and Xie (\textbf{b}) negative control pairs.
+\textbf{c-d}, Estimates produced by GLM-EIV and thresholded regression on two positive control pairs -- \textit{LRIF1} (\textbf{a}) and \textit{NDUFA2} (\textbf{b}) -- plotted as a function of excess background contamination.
+Grey bands, 95\% CIs outputted by the methods.
+\textbf{e-f}, Median relative estimate change (REC; \textbf{e}) and confidence interval coverage rate (\textbf{f}) aggregated across \textit{all} 322 positive control pairs, plotted as a function of excess background contamination.
+Colored bands, pointwise 95\% CIs for population quantities.
+Panels (\textbf{c-f}) together illustrate that GLM-EIV was more stable than thresholded regression as background contamination increased.}
+	\label{fig:real_data}
+\end{figure}
 
@@
 
-We see the practical use of this method as follows.
+% We see the practical use of this method as follows.
 First, deploy GLM-EIV to some random subset of pairs to assess the extent of background contamination.
@@
 
-To our knowledge, this is the first statistical work to study single-cell CRISPR screens.
+\begin{figure}
+\centering
+\includegraphics[width=0.9\linewidth]{../../figures/dec_tree/dec_tree_crop}
+\caption{\textbf{Use of GLM-EIV in practice}.
+The decision tree above illustrates how we anticipate GLM-EIV could be used in practice.
+First, apply GLM-EIV to a set of randomly-sampled gene-perturbation pairs to assess background contamination level (positive control pairs work best for this purpose).
+If GLM-EIV indicates that background contamination is high (e.g., $\exp(\beta^g_1) \lessapprox $ 10), apply GLM-EIV to analyze the entire dataset; otherwise, approximate the Bayes-optimal decision boundary using the fitted GLM-EIV models.
+Next, apply a thresholding method (e.g., SCEPTRE or thresholded negative binomial regression) to analyze the data, setting the threshold to the estimated Bayes-optimal decision boundary.}
+\end{figure}
+
+% To our knowledge, this is the first statistical work to study single-cell CRISPR screens.
 One of our objectives is to bring to the attention of the statistical community an emerging class of functional genomics assays that likely will exert a major impact on biology research.

### 2021-12-07 R_scripts/plotting/data_analysis.R

@@ -38,7 +38,7 @@ f2 <- function(resampling_df, pair_id, x_max = 0.4, pair_info) {
   p <- ggplot(data = df_synth, mapping = aes(x = contam_level, y = m_est, col = method)) + 
     theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), legend.position = "blank", legend.title=element_blank(), plot.title = element_text(hjust = 0.5, size = 11)) +
     scale_color_manual(values = c(my_cols[1], my_cols[3])) + geom_hline(yintercept = mean(df_gt$estimate), col = "black", lwd = 0.85) + 
-    xlab("Excess background contamination") + scale_x_continuous(expand = expansion(mult = c(0.05, 0.03))) +  geom_ribbon(aes(ymin = m_ci_lower, ymax = m_ci_upper), fill = "grey80", alpha = 0.6, linetype=0) + geom_line(lwd = 0.85) +
+    xlab("Excess background contamination") + scale_x_continuous(expand = expansion(mult = c(0.05, 0.03))) +  geom_ribbon(aes(ymin = m_ci_lower, ymax = m_ci_upper), fill = "grey60", alpha = 0.8, linetype = 0) + geom_line(lwd = 0.85) +
     ylab("Estimate") + scale_fill_manual(values = c(my_cols[1], my_cols[3])) + my_theme +
     ggtitle(label = title)
   return(p)
@@ -94,7 +94,6 @@ raw_est_vs_fitted_model_est
 # Raw data analyses
 ####################
 # Define functions
-
 # 1. join data frames for comparison
 join_results <- function(glmeiv_res, thresh_res, site_type) {
   left_join(x = filter(.data = glmeiv_res, target == "estimate", parameter == "m_perturbation", type == !!site_type) %>% select(pair_id, value),
@@ -175,7 +174,6 @@ comparison_df_sub <- comparison_df %>% filter(glmeiv_ok)
 p2 <- ggplot(data = comparison_df_sub, mapping = aes(x = value_glmeiv, y = value_thresh)) + geom_point(alpha = 0.25, color = my_cols[2], cex = 0.9) + 
   geom_abline(slope = 1, intercept = 0, lwd = 0.85) + xlab("GLM-EIV estimate") + ylab("Thresholding estimate") + my_theme +
[…]

### 2021-12-07 nextflow_scripts/bin/run_analysis.R

@@ -43,15 +43,8 @@ for (i in seq(1L, n_pairs)) {
   if (i == 1 || gRNA_ids[i] != gRNA_ids[i - 1]) { # likewise for gRNAs
     g <- as.numeric(gRNA_odm[[gRNA,]])
   }
-  
+
   # loop through different thresholds, fitting model to each
-  # if (is.numeric(threshold)) {
-  #  curr_threshold <- threshold
-  # } else {
-  #  g_prime <- g[g >= 1]
-  #  curr_threshold <- mean(g_prime)
-  # }
-  
   phat <- as.integer(g >= threshold)
   # run thresholding method
   fit <- glmeiv::run_thresholding_method(phat = phat, m = m, m_fam = m_fam,
@@ -62,3 +55,8 @@ for (i in seq(1L, n_pairs)) {
 
 out <- do.call(rbind, out_l) %>% dplyr::mutate_at(c("parameter", "target", "gene_id", "gRNA_id"), factor)
 saveRDS(object = out, file = "raw_result.rds")
+
+
+# leftover code
+# g_prime <- g[g >= 1]
+# curr_threshold <- mean(g_prime)

### 2021-12-07 text/outline/draft.tex
finish draft of data analysis section

@@
 Our theoretical analyses (Section \ref{sec:thresholding_theory}) and simulation studies (Section \ref{sec:simulation}) revealed that the Bayes-optimal decision boundary is a good choice for the threshold when the gRNA count distribution is well separated.
-Therefore, under the assumption that the effect of the perturbation on gRNA expression is similar across pairs, we used the fitted GLM-EIV models to approximate the Bayes-optimal decision boundary for both datasets, yielding a boundary of $3$ for Gasperini and $7$ for Xie.
-We then applied the thresholding method to analyze both datasets, setting the threshold to the Bayes-optimal decision boundary estimated using GLM-EIV.
-
-% We compared GLM-EIV to thresholded regression on the real data and surprisingly found that the two methods produced similar results.
-Figures \ref{fig:real_data}a and \ref{fig:real_data}b show the estimate of the target of inference $\exp(\beta^m_1)$
-
-To better understand this result, we examined the estimated effect of the perturbation on gRNA expression, $\exp(\hat{\beta}_1^g)$; this parameter plays an important role in determining problem difficulty.
-The value of $\exp(\hat{\beta}_1^g)$ was greater than expected: the 95\% CI for this parameter averaged across pairs was $[XX,XX]$ and $[300, 316]$ on the Gasperini and Xie data, respectively.
-We concluded that both datasets lay in an ``easy'' region of the parameter space, making thresholding a tenable strategy (provided the threshold is selected well).
-
-To illuminate the settings in which GLM-EIV and the thresholding method diverge, we artificially increased the problem difficulty by generating partially-synthetic datasets.
-To this end, we sampled gRNA counts from a realistic model, holding fixed the raw gene expressions, library sizes, covariates, and fitted perturbation probabilities.
-We defined a normalized parameter called ``excess background contamination'' that controlled the problem difficulty.
-When excess background contamination was equal to zero, we sampled gRNA counts directly from the fitted model.
-As excess background contamination increased, we sampled gRNA counts from a slightly modified version of the fitted model in which the mean gRNA expression of \textit{unperturbed} cells was increased while the mean gRNA expression of \textit{perturbed} cells was held constant.
-Finally, when excess background contamination was equal to $1$ (the maximum value), we set the mean gRNA expression of unperturbed cells equal to that of perturbed cells and sampled from the resulting model.
-
-We generated partially-synthetic data for each of the 322 positive control pairs in the Gasperini dataset, varying excess background contamination over the interval $[0,0.4].$ We applied GLM-EIV and the thresholding method to analyze these data.
-We present results on two representative pairs (the pair containing gene \textit{LRIF1} and the pair containing gene \textit{NDUFA2}) in Figures \ref{fig:real_data}c-d.
-First, we observed that the estimate produced by the methods on the raw data (depicted as a horizontal black line) coincided almost exactly with the estimate produced by the methods on the partially-synthetic data generated by setting excess background contamination to zero (trend present across all genes; average relative difference $0.003$).
-This indicates that the fitted model accurately tracked the gRNA count distribution.
-Next, we observed that, as excess background contamination increased, the thresholding method exhibited severe attenuation bias, whereas GLM-EIV remained roughly stable.
-
-We conducted a similar analysis on the entire set of positive control pairs.
-First, for each pair, we computed the ``relative estimate change'' (REC), defined as the relative difference between the estimate at some level of excess contamination and zero excess contamination (Figure \ref{fig:real_data}d).
-% We plotted median
+Therefore, under the assumption that the effect of the perturbation on gRNA expression is similar across pairs, we used the fitted GLM-EIV models to approximate the Bayes-optimal decision boundary on both datasets, yielding a boundary of $3$ for Gasperini and $7$ for Xie.
+We then applied the thresholding method to the data, setting the threshold to the Bayes-optimal decision boundary estimated using GLM-EIV.
+
+We compared GLM-EIV to thresholded regression on the real data, focusing specifically on the negative control pairs (i.e., gene-perturbation pairs that, by design, are expected to exhibit a fold change of $1$, or no association).
+We found that GLM-EIV and the thresholding method produced similar results (Figure \ref{fig:real_data}a-b): estimates, CI coverage rates, and CI widths were concordant.
+We note that GLM-EIV produced outlier estimates (likely due to non-global EM convergence) on a small ($< 2\%$ on Gasperini, $<0.05\%$ on Xie) number of pairs consisting of a handful of genes (not plotted).
+The estimated effect of the perturbation on gene expression $\exp(\hat{\beta}_1^g)$ was unexpectedly large (95\% CI $[XX,XX]$ and $[300,316]$ on Gasperini and Xie data, respectively).
+We concluded that the datasets lay in an ``easy'' region of the parameter space, making thresholding a tenable strategy (provided the threshold is selected well).
+
+To explore the settings in which GLM-EIV and the thresholding method diverge, we artificially increased the problem difficulty by generating partially-synthetic datasets.
+To this end, we defined a parameter called ``excess background contamination'' that controls the extent to which the gRNA count distribution is separated.
+When excess background contamination is zero, we sample gRNA counts directly from the fitted GLM-EIV model.
+As excess background contamination increases, we sample gRNA counts from a slightly modified version of the fitted model in which the mean gRNA expression of \textit{unperturbed} cells is increased while the mean gRNA expression of \textit{perturbed} cells is held constant.
+Finally, when excess background contamination is $1$ (the maximum value), we set the mean gRNA expression of unperturbed cells equal to that of perturbed cells and sample gRNA counts from the resulting model.
+We hold fixed the real-data gene expressions, library sizes, covariates, and fitted perturbation probabilities in all settings.
+
+We generated partially-synthetic data in the above manner for each of the 322 positive control pairs in the Gasperini dataset, varying excess background contamination over the interval $[0,0.4].$ We then applied GLM-EIV and the thresholding method to analyze the data.
+We present results on two example pairs (the pair containing gene \textit{LRIF1} and the pair containing gene \textit{NDUFA2}) in Figures \ref{fig:real_data}c-d.
+We observed that the estimate produced by the methods on the raw data (depicted as a horizontal black line) coincided almost exactly with the estimate produced by the methods on the partially-synthetic data generated by setting excess background contamination to zero (result replicated across nearly all pairs; average relative difference $0.003$).
+This suggested that the fitted models accurately tracked the gRNA count distributions, giving us confidence in our procedure for generating realistic, partially-synthetic datasets.
+Next, we observed that as excess background contamination increased, the performance of thresholded regression degraded considerably while that of GLM-EIV remained stable on the genes \textit{LRIF1} and \textit{NDUFA2}.
+
+We generalized the above analysis to the entire set of positive control pairs.
+First, for each pair, we computed the ``relative estimate change'' (REC) as a function of excess background contamination, defined as the relative difference between the estimate at a given level of excess contamination and zero excess contamination (Figure \ref{fig:real_data}d).
+Next, we computed the median REC across all positive control pairs (Figure \ref{fig:real_data}e; bands indicate pointwise 95\% CIs for the population median REC).
+As excess background contamination increased, thresholded regression exhibited severe attenuation bias (as reflected by large median REC values); GLM-EIV, by contrast, remained mostly stable.
+Finally, taking the estimate obtained on the raw data as ``ground truth,'' we computed the CI coverage rate across pairs as a function of excess contamination (Figure \ref{fig:real_data}f; bands indicate $95\%$ pointwise CIs for population coverage rate).
+GLM-EIV exhibited significantly higher CI coverage rates than thresholded regression as the data became increasingly contaminated.
 
@@
 	\caption{\textbf{Data analysis}.
-\textbf{a-b}, Fold change estimates produced by GLM-EIV and thresholded regression on Gasperini (\textbf{a}) and Xie (\textbf{b}) negative control pairs.
+\textbf{a-b}, Estimates for fold change produced by GLM-EIV and thresholded regression on Gasperini (\textbf{a}) and Xie (\textbf{b}) negative control pairs.
 \textbf{c-d}, Estimates produced by GLM-EIV and thresholded regression on two positive control pairs -- \textit{LRIF1} (\textbf{a}) and \textit{NDUFA2} (\textbf{b}) -- plotted as a function of excess background contamination.
-Grey bands, 95\% CIs outputted by the methods.
-\textbf{e-f}, Median relative estimate change (REC; \textbf{e}) and confidence interval coverage rate (\textbf{f}) aggregated across \textit{all} 322 positive control pairs, plotted as a function of excess background contamination.
+Grey bands, 95\% CIs for the target of inference outputted by the methods.
+\textbf{e-f}, Median relative estimate change (REC; \textbf{e}) and confidence interval coverage rate (\textbf{f}) across \textit{all} 322 positive control pairs, plotted as a function of excess background contamination.
 Colored bands, pointwise 95\% CIs for population quantities.
-Panels (\textbf{c-f}) together illustrate that GLM-EIV was more stable than thresholded regression as background contamination increased.}
+Panels (\textbf{c-f}) together illustrate that GLM-EIV demonstrated greater stability than thresholded regression as background contamination increased.}
 	\label{fig:real_data}

### 2021-12-08 text/outline/draft.tex
complete draft

@@
 
-We compared GLM-EIV to thresholded regression on the real data, focusing specifically on the negative control pairs (i.e., gene-perturbation pairs that, by design, are expected to exhibit a fold change of $1$, or no association).
-We found that GLM-EIV and the thresholding method produced similar results (Figure \ref{fig:real_data}a-b): estimates, CI coverage rates, and CI widths were concordant.
-We note that GLM-EIV produced outlier estimates (likely due to non-global EM convergence) on a small ($< 2\%$ on Gasperini, $<0.05\%$ on Xie) number of pairs consisting of a handful of genes (not plotted).
-The estimated effect of the perturbation on gene expression $\exp(\hat{\beta}_1^g)$ was unexpectedly large (95\% CI $[XX,XX]$ and $[300,316]$ on Gasperini and Xie data, respectively).
-We concluded that the datasets lay in an ``easy'' region of the parameter space, making thresholding a tenable strategy (provided the threshold is selected well).
-
-To explore the settings in which GLM-EIV and the thresholding method diverge, we artificially increased the problem difficulty by generating partially-synthetic datasets.
-To this end, we defined a parameter called ``excess background contamination'' that controls the extent to which the gRNA count distribution is separated.
-When excess background contamination is zero, we sample gRNA counts directly from the fitted GLM-EIV model.
-As excess background contamination increases, we sample gRNA counts from a slightly modified version of the fitted model in which the mean gRNA expression of \textit{unperturbed} cells is increased while the mean gRNA expression of \textit{perturbed} cells is held constant.
-Finally, when excess background contamination is $1$ (the maximum value), we set the mean gRNA expression of unperturbed cells equal to that of perturbed cells and sample gRNA counts from the resulting model.
-We hold fixed the real-data gene expressions, library sizes, covariates, and fitted perturbation probabilities in all settings.
-
-We generated partially-synthetic data in the above manner for each of the 322 positive control pairs in the Gasperini dataset, varying excess background contamination over the interval $[0,0.4].$ We then applied GLM-EIV and the thresholding method to analyze the data.
-We present results on two example pairs (the pair containing gene \textit{LRIF1} and the pair containing gene \textit{NDUFA2}) in Figures \ref{fig:real_data}c-d.
-We observed that the estimate produced by the methods on the raw data (depicted as a horizontal black line) coincided almost exactly with the estimate produced by the methods on the partially-synthetic data generated by setting excess background contamination to zero (result replicated across nearly all pairs; average relative difference $0.003$).
-This suggested that the fitted models accurately tracked the gRNA count distributions, giving us confidence in our procedure for generating realistic, partially-synthetic datasets.
-Next, we observed that as excess background contamination increased, the performance of thresholded regression degraded considerably while that of GLM-EIV remained stable on the genes \textit{LRIF1} and \textit{NDUFA2}.
-
-We generalized the above analysis to the entire set of positive control pairs.
-First, for each pair, we computed the ``relative estimate change'' (REC) as a function of excess background contamination, defined as the relative difference between the estimate at a given level of excess contamination and zero excess contamination (Figure \ref{fig:real_data}d).
-Next, we computed the median REC across all positive control pairs (Figure \ref{fig:real_data}e; bands indicate pointwise 95\% CIs for the population median REC).
-As excess background contamination increased, thresholded regression exhibited severe attenuation bias (as reflected by large median REC values); GLM-EIV, by contrast, remained mostly stable.
-Finally, taking the estimate obtained on the raw data as ``ground truth,'' we computed the CI coverage rate across pairs as a function of excess contamination (Figure \ref{fig:real_data}f; bands indicate $95\%$ pointwise CIs for population coverage rate).
-GLM-EIV exhibited significantly higher CI coverage rates than thresholded regression as the data became increasingly contaminated.
-
 \begin{figure}
@@
 Colored bands, pointwise 95\% CIs for population quantities.
-Panels (\textbf{c-f}) together illustrate that GLM-EIV demonstrated greater stability than thresholded regression as background contamination increased.}
-	\label{fig:real_data}
+Panels (\textbf{c-f}) together illustrate that GLM-EIV demonstrated greater stability than thresholded regression as background contamination increased.}\label{fig:real_data}
 \end{figure}
 
+We compared GLM-EIV to thresholded regression on the real data, focusing specifically on the negative control pairs (i.e., gene-perturbation pairs that, by design, are expected to exhibit a fold change of $1$, or no association).
+We found that GLM-EIV and the thresholding method produced similar results (Figure \ref{fig:real_data}a-b): estimates, CI coverage rates, and CI widths were concordant.
+We note that GLM-EIV produced outlier estimates (likely due to non-global EM convergence) on a small ($< 2\%$ on Gasperini, $<0.05\%$ on Xie) number of pairs consisting of a handful of genes (not plotted).
+The estimated effect of the perturbation on gene expression $\exp(\hat{\beta}_1^g)$ was unexpectedly large (95\% CI $[XX,XX]$ and $[300,316]$ on Gasperini and Xie data, respectively).
+We concluded that the datasets lay in an ``easy'' region of the parameter space, making thresholding a tenable strategy (provided the threshold is selected well).
+
+To explore the settings in which GLM-EIV and the thresholding method diverge, we artificially increased the problem difficulty by generating partially-synthetic datasets.
+To this end, we defined a parameter called ``excess background contamination'' that controls the extent to which the gRNA count distribution is separated.
+When excess background contamination is zero, we sample gRNA counts directly from the fitted GLM-EIV model.
+As excess background contamination increases, we sample gRNA counts from a slightly modified version of the fitted model in which the mean gRNA expression of \textit{unperturbed} cells is increased while the mean gRNA expression of \textit{perturbed} cells is held constant.
+Finally, when excess background contamination is $1$ (the maximum value), we set the mean gRNA expression of unperturbed cells equal to that of perturbed cells and sample gRNA counts from the resulting model.
+We hold fixed the real-data gene expressions, library sizes, covariates, and fitted perturbation probabilities in all settings.
+
+We generated partially-synthetic data in the above manner for each of the 322 positive control pairs in the Gasperini dataset, varying excess background contamination over the interval $[0,0.4].$ We then applied GLM-EIV and the thresholding method to analyze the data.
+We present results on two example pairs (the pair containing gene \textit{LRIF1} and the pair containing gene \textit{NDUFA2}) in Figures \ref{fig:real_data}c-d.
+We observed that the estimate produced by the methods on the raw data (depicted as a horizontal black line) coincided almost exactly with the estimate produced by the methods on the partially-synthetic data generated by setting excess background contamination to zero (This result replicated across nearly all pairs; average relative difference $0.003$.) We concluded that the fitted models accurately tracked the gRNA count distributions, giving us confidence in our procedure for generating realistic, partially-synthetic datasets.
+We found that as excess background contamination increased, the performance of thresholded regression degraded considerably while that of GLM-EIV remained stable.
+
+We generalized the above analysis to the entire set of positive control pairs.
+First, for each pair we computed its ``relative estimate change'' (REC) as a function of excess background contamination, defined as the relative difference between the estimate at a given level of excess contamination and zero excess contamination (Figure \ref{fig:real_data}d).
+Next, we computed the median REC across all positive control pairs (Figure \ref{fig:real_data}e; bands indicate pointwise 95\% CIs for the population median REC).
+As excess background contamination increased, thresholded regression exhibited severe attenuation bias (as reflected by large median REC values); GLM-EIV, by contrast, remained mostly stable.
+Finally, taking the estimate obtained on the raw data as ``ground truth,'' we computed the CI coverage rate across pairs as a function of excess contamination (Figure \ref{fig:real_data}f; bands indicate $95\%$ pointwise CIs for population coverage rate).
+GLM-EIV exhibited a significantly higher CI coverage rate than thresholded regression as the data became increasingly contaminated.
+
 \section{Discussion}
 
-% We see the practical use of this method as follows.
-First, deploy GLM-EIV to some random subset of pairs to assess the extent of background contamination.
-If the background contamination sufficiently high, apply GLM-EIV to analyze the data directly.
-On the other hand, if the background contamination is low, the thresholding method is an appealing method due to its speed and stability.
-A challenge, as we have shown, is that the selected threshold exerts a substantial impact on the results.
-GLM-EIV approximates the Bayes-optimal decision boundary, which, as we showed in an idealized Gaussian setting, minimizes attenuation bias when the gRNA count distribution is sufficiently-well separated.
-Therefore, in problem settings where thresholding is tenable, GLM-EIV provides a statistically principled, scientifically reproducible, empirically effective threshold selection procedure.
+In this work we introduced GLM-EIV (``GLM-based errors in variables''), a new method for single-cell CRISPR screen analysis.
+GLM-EIV generalizes the classical errors-in-variables model (when the latent predictor is binary) to arbitrary exponential distributions and link functions, facilitating the analysis of single-cell CRISPR screen data.
+We demonstrated through theoretical analyses, simulation studies, and real data analyses that GLM-EIV outperforms thresholded regression by a considerable margin in high background contamination settings.
+On the other hand, in low background contamination settings, GLM-EIV provides a statistically principled and empirically effective procedure for selecting a threshold for use in a thresholding method (e.g., thresholded negative binomial regression, as studied in this work, or even SCEPTRE \cite{Barry2020}).
+GLM-EIV thereby neutralizes a tuning parameter that exerts a significant impact on the results.
+Figure \ref{fig:dec_tree} summarizes how we anticipate GLM-EIV being used in practice.
+
+To our knowledge this is the first paper to examine single-cell CRISPR screens from a rigorous statistical perspective.
+We hope that this work can serve as an introduction to the broader statistics community to an emerging class functional genomics assays that likely will exert a major impact on biology research in the coming years \cite{Przybyla2021}.
+Additionally, this is the first work to leverage the ondisc-Nextflow-cloud/HPC technology stack, an (in our view) tightly-integrated, user-friendly, and powerful set of computational tools for out-of-memory and massively distributed single-cell analysis.
+
+We anticipate that GLM-EIV could be applied to other types of single-cell CRISPR screen and multimodal single-cell data.
+For example, we likely (with some effort) could extend GLM-EIV to ``low-multiplicity of infenction'' screens \cite{Schraivogel2020} in which each cell receives one or two perturbations rather than dozens (as is the case in ``high multiplicity screens,'' studied in this work).
+We also likely could apply GLM-EIV to analyze multimodal single-cell chromatin accessibility data.
+A question of interest in such experiments is whether chromatin state (i.e., closed or open) is associated with the expression of a gene (or abundance of a protein) \cite{Mimitou2021}.
+We do not directly observe chromatin state; instead, we observe transposase-tagged DNA fragments, counts that serve as proxies for whether a given region of chromatin is open or closed.
+GLM-EIV might be applied in such experiments to aid in the selection of thresholds or to analyze whole datasets.
 
@@
 If GLM-EIV indicates that background contamination is high (e.g., $\exp(\beta^g_1) \lessapprox $ 10), apply GLM-EIV to analyze the entire dataset; otherwise, approximate the Bayes-optimal decision boundary using the fitted GLM-EIV models.
-Next, apply a thresholding method (e.g., SCEPTRE or thresholded negative binomial regression) to analyze the data, setting the threshold to the estimated Bayes-optimal decision boundary.}
+Next, apply a thresholding method (e.g., SCEPTRE or thresholded negative binomial regression) to analyze the data, setting the threshold to the estimated Bayes-optimal decision boundary.}\label{fig:dec_tree}
 \end{figure}
-
-% To our knowledge, this is the first statistical work to study single-cell CRISPR screens.
-One of our objectives is to bring to the attention of the statistical community an emerging class of functional genomics assays that likely will exert a major impact on biology research.
-Additionally, this is the first work to leverage the ondisc-Nextflow-cloud/HPC technology stack, a set of programming tools for large-scale single-cell analysis that we view as powerful and user-friendly.
 

### 2021-12-08 text/outline/draft.tex
update gasp optimal bdy; setup thresholding method

@@
 To our knowledge this is the first paper to examine single-cell CRISPR screens from a rigorous statistical perspective.
-We hope that this work can serve as an introduction to the broader statistics community to an emerging class functional genomics assays that likely will exert a major impact on biology research in the coming years \cite{Przybyla2021}.
+We hope that this work can help introduce the broader statistics community to an emerging class functional genomics assays that likely will exert a major impact on biology research in the coming years \cite{Przybyla2021}.
 Additionally, this is the first work to leverage the ondisc-Nextflow-cloud/HPC technology stack, an (in our view) tightly-integrated, user-friendly, and powerful set of computational tools for out-of-memory and massively distributed single-cell analysis.

### 2021-12-08 R_scripts/analysis/approx_bayes_bdy.R

@@ -36,4 +36,29 @@ close(file_con)
 ##############
 # 2. GASPERINI
 ##############
+gasp_result_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/gasperini/results/")
+glmeiv_res <- paste0(gasp_result_dir, "result_glmeiv.rds") %>% readRDS()
+med_est <- glmeiv_res %>% filter(site_type == "selfTSS", target == "estimate") %>% select(parameter, value, pair_id) %>%
+  tidyr::pivot_wider(data = ., id_cols = "pair_id", names_from = "parameter", values_from = "value") %>% 
+  select(-pair_id) %>% summarize_all(median) %>% select(., starts_with("g"), "pi")
 
+# Load the data
+gasp_data_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/gasperini/data/")
+covariate_matrix <- paste0(gasp_data_dir, "covariate_matrix.rds") %>% readRDS()
+g_offset <- paste0(gasp_data_dir, "g_offsets.rds") %>% readRDS()
+g_fam <- poisson() %>% augment_family_object()
+
+# set arguments to Bayes-opt function
+g_intercept <- med_est$g_intercept %>% log()
+g_pert <- med_est$g_perturbation %>% log()
+pi <- med_est$pi
+
+# compute bayes-opt dec bdy and save
+bdy <- get_optimal_threshold(g_intercept = g_intercept, g_perturbation = g_pert, g_fam = g_fam, pi = pi,
+                             covariate_matrix = NULL, g_covariate_coefs = NULL, g_offset = median(g_offset)) %>% round()
+
+f_name <- paste0(gasp_data_dir, "dec_bdy.txt")
+if (file.exists(f_name)) file.remove(f_name)
+file_con <- file(f_name)
+writeLines(text = as.character(bdy), con = file_con)
+close(file_con)

### 2021-12-08 bash_scripts/run_gasp_thresholding.sh

@@ -14,4 +14,5 @@ nextflow run $thresholding_nf_pipeline --pairs $gRNA_gene_pairs \
 --m_offsets $m_offsets \
 --gRNA_odm $gRNA_odm \
 --gRNA_metadata $gRNA_metadata \
+--threshold $thresh \
 -bg > $PWD/log -ansi-log false -with-trace -w $work_dir

### 2021-12-08 bash_scripts/set_gasp_args.sh

@@ -57,3 +57,6 @@ fi
 
 # vii. results directory
 result_dir=$LOCAL_GLMEIV_DATA_DIR"public/gasperini/results"
+
+# viii. threshold
+thresh=$(cat $processed_data_dir"dec_bdy.txt")

### 2021-12-08 text/outline/draft.tex
add Fig 5a

@@
 We found that GLM-EIV and the thresholding method produced similar results (Figure \ref{fig:real_data}a-b): estimates, CI coverage rates, and CI widths were concordant.
-We note that GLM-EIV produced outlier estimates (likely due to non-global EM convergence) on a small ($< 2\%$ on Gasperini, $<0.05\%$ on Xie) number of pairs consisting of a handful of genes (not plotted).
-The estimated effect of the perturbation on gene expression $\exp(\hat{\beta}_1^g)$ was unexpectedly large (95\% CI $[XX,XX]$ and $[300,316]$ on Gasperini and Xie data, respectively).
+We note that GLM-EIV produced outlier estimates (likely due to non-global EM convergence) on a small ($< 2.5 \%$ on Gasperini, $<0.05\%$ on Xie) number of pairs consisting of a handful of genes (not plotted).
+The estimated effect of the perturbation on gene expression $\exp(\hat{\beta}_1^g)$ was unexpectedly large (95\% CI $[4306, 5186]$ and $[300, 316]$ on Gasperini and Xie data, respectively).
 We concluded that the datasets lay in an ``easy'' region of the parameter space, making thresholding a tenable strategy (provided the threshold is selected well).

### 2021-12-08 R_scripts/plotting/data_analysis.R

@@ -17,8 +17,8 @@ result_dir_gasp <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/gas
 ########################
 # 1. Resampling analysis
 ########################
-n_pairs <- resampling_df$pair_id %>% unique() %>% length()
 resampling_df <- readRDS(paste0(result_dir_gasp, "/resampling_result.rds"))
+n_pairs <- resampling_df$pair_id %>% unique() %>% length()
 
 # a function that plots the result given a given pair on the same plot
 f2 <- function(resampling_df, pair_id, x_max = 0.4, pair_info) {
@@ -118,32 +118,31 @@ get_ci_coverage_rate <- function(df, site_type = NULL) {
 ####################
 # Gasperini analysis
 ####################
-if (FALSE) {
 #  1. load the results
 glmeiv_res <- paste0(result_dir_gasp, "/result_glmeiv.rds") %>% readRDS() %>% mutate(type = site_type)
 thresh_res <- paste0(result_dir_gasp, "/result_thresholding.rds") %>% readRDS() %>% mutate(type = site_type)
 
-# 2. Ensure we are examining the same pairs (Gasp missing ~100 due to probable node failure)
+# 2. Ensure we are examining the same pairs
 glmeiv_pairs <- glmeiv_res %>% filter(parameter == "m_perturbation", target == "estimate") %>% pull(pair_id) %>% as.character() %>% sort()
 thresh_pairs <- thresh_res %>% filter(parameter == "m_perturbation", target == "estimate") %>% pull(pair_id) %>% as.character() %>% sort()
+identical(glmeiv_pairs, thresh_pairs) # pairs are the same across glmeiv and thresholding
 
-glmeiv_res <- glmeiv_res %>% filter(pair_id %in% ok_glmeiv_pairs)
-thresh_res <- thresh_res %>% filter(pair_id %in% ok_glmeiv_pairs)
+# 3. mean g_perturbation
+glmeiv_res %>% filter(parameter == "g_perturbation") %>% pivot_wider(id_cols = "pair_id", names_from = "target") %>%
+  summarize(m_est = mean(estimate), m_confint_lower = mean(confint_lower, na.rm = TRUE), m_confint_upper = mean(confint_upper[confint_upper < 10000], na.rm = TRUE))
[…]

### 2021-12-09 text/jsm/jsm_paper.tex
add jsm text

--- 2021-12-08 text/outline/draft.tex
@@
-\documentclass[12pt]{article}
- \usepackage[vmargin=1.18in,hmargin=1.1in]{geometry}
+\documentclass[11pt]{article}
+\usepackage{setspace}
+\doublespacing
+\usepackage[margin=1in]{geometry}
 \usepackage{amsfonts}
@@
 \usepackage{amssymb}
+\usepackage{amsthm}
 \usepackage{array}
+
+\usepackage[
+backend=biber,
+style=numeric-comp,
+maxcitenames=1,
+maxbibnames=3,
+backref=true,
+doi=false,
+isbn=false,
+url=false,
+eprint=false
+]{biblatex}
+
+\addbibresource{/Users/timbarry/optionFiles/glmeiv.bib}
 \newcommand{\PreserveBackslash}[1]{\let\temp=\\#1\let\\=\temp}
@@
 \noindent
-Tim, Gene, Kathryn
 \begin{center} 
-\textbf{A class of exponential family measurement error models, with application to CRISPR genome editing and single-cell sequencing}
+\textbf{A new class of exponential family measurement error models applied to single-cell CRISPR screens} \\
+Timothy Barry, Kathryn Roeder, Eugene Katsevich
 \end{center}
@@
 \end{abstract}
-\tableofcontents
-\newpage
+
 \section{Introduction}
@@
 
-We study estimation and inference in single-cell CRISPR screens from a statistical perspective, formulating the data generating mechanism using a new class of errors-in-variables (or measurement error) models.
-We assume that the response variable $y$ is a GLM of an underlying predictor variable $x^*$.
-We do not observe $x^*$ directly; rather, we observe a noisy version $x$ of $x^*$ that itself is a GLM of $x^*$.
-The goal of the analysis is to estimate the effect of $x^*$ on $y$ using the observed data $(x , y)$ only.
-In the context of the biological application, $x^*$, $y$, and $x$ are CRISPR perturbations, gene expressions, and guide RNA counts, respectively.
+We study estimation and inference in single-cell CRISPR screens from a statistical perspective, formulating the data generating mechanism using a new class of measurement error models.
+We assume that the response variable $y$ is a GLM of an underlying predictor variable $x^*$ and vector of confounders $z$.
+We do not observe $x^*$ directly; rather, we observe a noisy version $x$ of $x^*$ that itself is a GLM of $x^*$ and $z$.
+The goal of the analysis is to estimate the effect of $x^*$ on $y$ using the observed data $(x, y, z)$ only.
+In the context of the biological application, $x^*$, $x$, $y$, and $z$ are CRISPR perturbations, guide RNA counts, gene expressions, and technical confounders, respectively.
 
@@
 GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic.
-Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.
-
+% Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.
 We implement several statistical accelerations (that possibly are of independent utility) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method.
+
 Finally, we develop a Docker-containerized application to deploy GLM-EIV at-scale across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters.
 Leveraging this application, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets.
-% In highly noisy settings ; by contrast, in less noisy settings, GLM-EIV serves as a principled and reproducible threshold selection procedure.
+We find that in some settings, GLM-EIV outperforms thresholded regression by a considerable margin; in other settings the two methods work best in conjunction, with GLM-EIV providing a statistically principled and empirically effective procedure for selecting the threshold.
 
@@
 \subsection{Related work}
-
 Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions.
-For example, Lin, Lei, and Roeder \cite{Lin2021} developed eSVD, an extension of SVD to exponential family and curved Gaussian responses.
-Unlike SVD, eSVD models the relationship between the mean and variance of a gene's expression level, a phenomenon induced by the countedness of single-cell data \cite{Lause2021}.
-Similarly, Townes et al.\ \cite{Townes2019} proposed GLM-PCA, a generalization of PCA that directly models negative binomially-distributed gene expression counts while controlling for technical factors.
-We see our work as a continuation of this broad effort to ``port'' common statistical methods and models to single-cell count data.
-Our focus, however, is on regression rather than dimension reduction: we extend the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.
+For example, Lin et al.\ \cite{Lin2021} and Townes et al.\ \cite{Townes2019}  (separately) developed eSVD and GLM-PCA, generalizations of SVD and PCA, respectively, to exponential family response distributions.
+Unlike their vanilla counterparts, eSVD and GLM-PCA can model gene expression counts directly, thereby capturing the relationship between the mean and variance of a gene's expression level \cite{Lause2021}, improving performance on dimension reduction tasks.
+We see our work as a continuation of this effort to ``port'' common statistical methods and models to single-cell count data.
+Our focus, however, is on regression rather than dimension reduction.
+We extend the classical errors-in-variables model in several directions, most importantly generalizing to response distributions and sources of measurement error that are exponential family-distributed.
 
@@
 Thus, while Ibrahim and Gr\"{u}n \& Leisch are helpful references, our estimation and inference tasks are more complex than theirs.
+Other relevant references include Aigner \cite{Aigner1973} and Savoca \cite{Savoca2000}.
 
@@
 SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification.
-Other applied methods for single-cell CRISPR screen analysis include MIMOSCA \cite{Dixit2016} and scMAGeCK \cite{Yang2019}.
-These methods, like SCEPTRE, focus on hypothesis testing (rather than estimation), but unlike SCEPTRE, they ignore the countedness of the data and are unable to handle confounders.
 In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE.
-Most importantly, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is unattainable within the nonparameteric hypothesis testing framework of SCEPTRE.
+Most importantly, we aim to account for the fact that the perturbation is measured with noise, an issue that SCEPTRE circumvents via thresholding.
+Another difference is that we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is more challenging to attain within the nonparametric hypothesis testing framework of SCEPTRE.
+
+ %Other applied methods for single-cell CRISPR screen analysis include MIMOSCA \cite{Dixit2016} and scMAGeCK \cite{Yang2019}.
+% These methods, like SCEPTRE, focus on hypothesis testing (rather than estimation), but unlike SCEPTRE, they ignore the countedness of the data and are unable to handle confounders.
 
 \subsection{Assay overview}
-
-\begin{figure}
-	\centering
-	\includegraphics[width=1\linewidth]{../../figures/analysis_challenges/plot.pdf}
-	\caption{\textbf{Experimental design and analysis challenges}: \textbf{a,} Experimental design.
-For a given perturbation (e.g., the perturbation indicated in blue), we partition the cells into two groups: perturbed and unperturbed.
-Next, for a given gene, we conduct a differential expression analysis across the two groups, yielding an estimate of the impact of the given perturbation on the given gene.
-\textbf{b,} DAG representing all variables in the system.
-The perturbation (latent) impacts both gene expression and gRNA expression; technical factors act as confounders, also impacting gene and gRNA expression.
-The target of estimation is the effect of the perturbation on gene expression.
-\textbf{c,} Schematic illustrating the ``background read'' phenomenon.
-Due to errors in the sequencing and alignment processes, unperturbed cells exhibit a nonzero gRNA count distribution (bottom).
-The target of estimation is the change in mean gene expression in response to the perturbation (top).
-\textbf{d}, Example data on four cells for a given perturbation-gene pair.
-Note that (i) the perturbation is unobserved, and (ii) the gene and gRNA data are discrete counts.}
-	\label{analysis_challenges}
-\end{figure}
 
@@
 
-The human genome consists of genes, enhancers (segments of DNA that regulate the expression of one or more genes), and other genomic elements (that are not of importance to the current discussion).
+The human genome consists of genes, enhancers (segments of DNA that regulate the expression of one or more genes), and other genomic elements (that are not of importance to the current work).
 Genome-wide association studies (GWAS) have revealed that the majority ($>90\%$) of variants associated with diseases lie outside genes and (very likely) inside enhancers \cite{Gallagher2018}.
-These noncoding variants are thought to contribute to disease by modulating the expression one or more disease-relevant genes.
+These noncoding variants are thought to contribute to disease by modulating the expression of one or more disease-relevant genes.
 Scientists do not know the gene (or genes) through which most noncoding variants exert their effect, limiting the interpretability of GWAS results.
@@
 
-The most promising biotechnology for solving this challenge are high MOI single-cell CRISPR screens.
+High MOI single-cell CRISPR screens are the most promising biotechnology for solving this challenge.
 High MOI single-cell CRISPR screens combine CRISPR interference (CRISPRi) -- a version of CRISPR that represses a targeted region of the genome -- with single-cell sequencing.
@@
 
-In this work we analyze two large-scale, high MOI, single-cell CRISPR screen datasets published by Gasperini et al.\ and Xie et al.\ in 2019.
+In this work we analyze two large-scale, high MOI, single-cell CRISPR screen datasets published by Gasperini et al.\ \cite{Gasperini2019} and Xie et al.\ \cite{Xie2019a}.
 Gasperini (resp., Xie) targeted approximately 6,000 (resp., 500) candidate enhancers in a population of approximately 200,000 (resp., 100,000) cells.
-Gasperini additionally designed 381 positive control, gene-targeting perturbations and 50 non-targeting, negative control perturbations to assess method sensitivity and specificity.
+Gasperini additionally designed several hundred positive control, gene-targeting perturbations and 50 non-targeting, negative control perturbations to assess method sensitivity and specificity.
 
 \section{Thresholding method}
-We study the thresholding method from empirical and theoretical perspectives, highlighting several limitations of the approach.
-Gasperini and Xie both imputed perturbation assignments onto the cells via thresholding, but they carried out subsequent differential expression analyses in different ways: Gasperini used negative binomial regression, whereas Xie applied nonparametric tests of independence.
-We study Gasperini's variant of the thresholding method, as it relates more closely to GLM-EIV and (in our view) is the more natural approach.
+We study thresholding from empirical and theoretical perspectives, highlighting several inherent difficulties of the approach.
+Gasperini and Xie imputed perturbation assignments onto the cells via thresholding, but they carried out the subsequent differential expression analysis in different ways: Gasperini fitted negative binomial regression models to the data, whereas Xie applied nonparametric tests of independence.
+We study Gasperini's variant of the thresholding method, as it relates more closely to GLM-EIV and (in our view) is the more natural method.
 
@@
 Consider a single perturbation and a single gene.
-For cell $i \in \{1, \dots, n\}$, let $m_i \in \N$ be the number of gene transcripts sequenced; let $g_i \in \N$ be the number of gRNA transcripts sequenced; let $d^m_i \in \N$ be the number of gene transcripts sequenced across \textit{all} genes (the library size or sequencing depth); and finally, let $z_i \in \R^{d-1}$ be the cell-specific technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.) The letters ``m,'' ``g'', and ``d'' stand for ``mRNA,'' ``gRNA,'' and ``depth,'' respectively.
+For cell $i \in \{1, \dots, n\}$, let $m_i \in \N$ be the number of gene transcripts sequenced; let $g_i \in \N$ be the number of gRNA transcripts sequenced; let $d^m_i \in \N$ be the number of gene transcripts sequenced across \textit{all} genes (i.e., the library size or sequencing depth); and finally, let $z_i \in \R^{d-1}$ be the cell-specific technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.) The letters ``m,'' ``g'', and ``d'' stand for ``mRNA,'' ``gRNA,'' and ``depth,'' respectively.
 The thresholding method is defined as follows:
 \begin{itemize}
-\item[1.] For a given threshold $c \in \N$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be $$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i < c, \\ \hat{p}_i = 1 \textrm{ if } g_i \geq c.
-\end{cases}$$
+\item[1.] For a given threshold $c \in \N$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be $$ \hat{p}_i = 0 \textrm{ if } g_i < c; \quad \hat{p}_i = 1 \textrm{ if } g_i \geq c.$$
+%$$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i < c, \\ \hat{p}_i = 1 \textrm{ if } g_i \geq c. \end{cases}$$
 \item[2.] Assume that $m_i$ is related to $\hat{p}_i, d^m_i,$ and $z_i$ through the following GLM:
-$$m_i | (\hat{p}_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i),$$
 \begin{equation}\label{thresh_glm}
+m_i | (\hat{p}_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i); \quad
 \log(\mu_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left(d_i^m\right),
@@
 
-\item[3.]  Fit a GLM to obtain estimates of the parameters. Compute a $p$-value and confidence interval for the target of inference $\beta^m_1$.
+\item[3.]  Fit a GLM to obtain an estimate and confidence interval for the target of inference $\beta^m_1$.
 \end{itemize}
-The sequencing depth $d^m_i$ is included as an offset term in (\ref{thresh_glm}) so that $\beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i$ can be interpreted as a relative expression: exponentiating both sides of (\ref{thresh_glm}) yields
-$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m.$$ The term $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration.
-The target of inference, $\beta^m_1$, is the log-transformed fold change in gene expression in response to the perturbation, controlling for the technical factors.
-Fold change (i.e., $\exp(\beta^m_1)$) is the ratio of mean gene expression in perturbed cells to unperturbed cells: $\exp(\beta^m_1) = 1$ indicates no change in expression, while $\exp(\beta^m_1) > 1$ and $\exp(\beta^m_1) < 1$ indicate an increase and decrease in expression in response to the perturbation, respectively (accounting for technical factors).
+The sequencing depth $d^m_i$ is included as an offset term in (\ref{thresh_glm}) so that $\beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i$ can be interpreted as a relative expression.
+Exponentiating both sides of (\ref{thresh_glm}) yields
+$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m,$$ where $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration.
+
+Fold change in this context is defined as the ratio of the mean expression of perturbed cells to the mean expression of unperturbed cells.
+The target of inference $\beta^m_1$ is the \textit{log} fold change in expression in response to the perturbation, controlling for the technical factors.
+Therefore, $\exp(\beta^m_1) = 1$ (eq., $\beta^m_1 = 0$) indicates no change in expression, while $\exp(\beta^m_1) > 1$ (eq., $\beta^m_1 > 0$) and $\exp(\beta^m_1) < 1$ (eq., $\beta^m_1 < 0$) indicate an increase and decrease in expression, respectively.
 
@@
 We examined the behavior of the thresholding method on real data and uncovered attenuation bias and bias-variance tradeoff effects.
-We applied the thresholding method to analyze the set of 381 positive control perturbation-gene pairs in the Gasperini dataset.
+We applied the thresholding method to analyze the set of 322 positive control perturbation-gene pairs in the Gasperini dataset.
 The positive control pairs consisted of perturbations that targeted gene transcription start sites (TSSs) for inhibition.
@@
 Overall, we concluded that the thresholding method faces several challenges: (i) the threshold is a tuning parameter that significantly impacts the results; (ii) the threshold mediates an intrinsic bias-variance tradeoff; and (iii) the gRNA count distributions do not imply a clear threshold selection strategy.
-
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=1\linewidth]{../../figures/thresholding_empirical/plot}
-	\caption{\textbf{Empirical challenges of thresholded regression.} \textbf{a-b,} Fold change estimates produced by threshold = 1 versus threshold = 5 (a) and threshold = 20 versus threshold = 5 (b).
-The selected threshold substantially impacts the results.
-\textbf{c-d,} $p$-values (c) and CI widths (d) produced by threshold = 20 versus threshold = 5.
-The latter threshold yields more confident estimates.
-\textbf{e-f}, Empirical distribution of randomly-selected gRNA from Gasperini (e) and Xie (f) data (0 counts not shown).
-The gRNA data do not appear to imply an obvious threshold selection strategy.}
-	\label{thresholding_empirical}
-\end{figure}
-\newpage
 
@@
 
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=1\linewidth]{../../figures/thresholding_theoretical/plot}
-	\caption{\textbf{Theoretical challenges of thresholded regression.} \textbf{a,} Asymptotic relative bias versus threshold for different values of $\beta^g_1$.
-The bias function is highly nonconvex and strictly nonzero.
-Vertical blue lines, Bayes-optimal decision boundaries.
-Across all panels, $\beta^g_0 = 0$ and $\pi = 1/2$.
-\textbf{b,} Asymptotic relative bias versus $\pi$ when the threshold is set to a large number.
-The two quantities coincide exactly.
-\textbf{c,} Bias-variance decomposition for thresholding method in no-intercept model.
-Bias decreases and variance increases as the threshold tends to infinity.
-$\beta^g_1 = 1, \beta^m_1 = 1,$ and $\pi = 0.1$.}
-	\label{thresholding_theoretical}
-\end{figure}
-
 \begin{center}
@@
 We use a negative binomial GLM to model the gRNA counts because the gRNA molecules are transcribed in the cell in the same way as gene transcripts \cite{Datlinger2017,Hill2018}.
-Finally, we model the marginal perturbation probability as
-\begin{equation}\label{glmeiv_model_5}
+Finally, we model the marginal perturbation probability as \begin{equation}\label{glmeiv_model_5}
 p_i \sim \textrm{Bern}(\pi),
@@
 
-\begin{figure}
+We applied three methods to the simulated data: ``vanilla'' GLM-EIV (Algorithm \ref{algo:em_full}), accelerated GLM-EIV (Algorithm \ref{algo:pilot_estimates}), and thresholded regression.
+We used the Bayes-optimal decision boundary for classification as the threshold for the thresholding method.
+We ran all methods on the negative binomial data twice: once treating the size parameter $\theta$ as a known constant and once treating $\theta$ as unknown.
+In the latter case we used the \texttt{glm.nb} function from the \texttt{MASS} package to estimate $\theta$ \cite{Ripley2013}.
+We display the results of the simulation study in Figure \ref{main_text_sim}.
+Columns correspond to distributions (i.e., Poisson, negative binomial with known $\theta$, and negative binomial with unknown $\theta$), rows correspond to performance metrics (i.e., bias, mean squared error, CI coverage rate (nominal rate $95\%$), CI width, and method execution time).
+The problem difficulty parameter $\beta^g_1$ is plotted on the horizontal axis, and the methods are depicted in different colors (GLM-EIV = red, accelerated GLM-EIV = blue, and thresholding = purple; GLM-EIV masked by accelerated GLM-EIV in several panels).
+
+First, we observed that GLM-EIV dominated thresholded regression on all statistical metrics: GLM-EIV exhibited lower bias (row 1) and mean squared error (row 2) than thresholded regression; additionally, GLM-EIV had superior confidence interval coverage (row 3) despite having produced generally narrower confidence intervals (row 4).
+Intuitively, GLM-EIV outperformed the thresholding method because (i) GLM-EIV leveraged information from \textit{both} modalities (rather than the gRNA modality alone) to assign perturbation identities to cells, and (ii) GLM-EIV produced soft rather than hard assignments, capturing the inherent uncertainty in whether a perturbation occurred.
+Next, we found that accelerated GLM-EIV performed as well as ``vanilla'' GLM-EIV on the statistical metrics (rows 1-4) despite have substantially lower computational cost (bottom row).
+In fact, the execution time of accelerated GLM-EIV was approximately within an order of magnitude of that of the thresholding method (bottom row).
+
+Interestingly, thresholded regression exhibited better confidence interval coverage under estimated $\theta$ than under known $\theta$ (row 3).
+Estimating $\theta$ leads to slight inflation bias (i.e., overestimating the true effect size), whereas, as we showed previously, thresholding leads to attenuation bias (i.e., underestimating the true effect size).
+These phenomena partially cancel, yielding less biased estimates.
+We note that GLM-EIV and the thresholding method in principal are compatible with \textit{any} $\theta$ estimation procedure, including those based on more sophisticated techniques, such as regularization \cite{Hafemeister2019}.
+We defer rigorous investigation of the impact of different $\theta$ estimation strategies on these methods to future work.
+
+\section{Data analysis}
+
+Leveraging our computational infrastructure (Section \ref{sec:computing}), we applied GLM-EIV to analyze the entire Gasperini and Xie datasets (Figure \ref{fig:real_data}).
+Our theoretical analyses (Section \ref{sec:thresholding_theory}) and simulation studies (Section \ref{sec:simulation}) revealed that the Bayes-optimal decision boundary is a good choice for the threshold when the gRNA count distribution is well separated.
+Therefore, under the assumption that the effect of the perturbation on gRNA expression is similar across pairs, we used the fitted GLM-EIV models to approximate the Bayes-optimal decision boundary on both datasets, yielding a boundary of $3$ for Gasperini and $7$ for Xie.
+We then applied the thresholding method to the data, setting the threshold to the Bayes-optimal decision boundary estimated using GLM-EIV.
+
+We compared GLM-EIV to thresholded regression on the real data, focusing specifically on the negative control pairs (i.e., gene-perturbation pairs that, by design, are expected to exhibit a fold change of $1$, or no association).
+We found that GLM-EIV and the thresholding method produced similar results (Figure \ref{fig:real_data}a-b): estimates, CI coverage rates, and CI widths were concordant.
+We note that GLM-EIV produced outlier estimates (likely due to non-global EM convergence) on a small ($< 2.5 \%$ on Gasperini, $<0.05\%$ on Xie) number of pairs consisting of a handful of genes (not plotted).
+The estimated effect of the perturbation on gene expression $\exp(\hat{\beta}_1^g)$ was unexpectedly large (95\% CI $[4306, 5186]$ and $[300, 316]$ on Gasperini and Xie data, respectively).
+We concluded that the datasets lay in an ``easy'' region of the parameter space, making thresholding a tenable strategy (provided the threshold is selected well).
+
+To explore the settings in which GLM-EIV and the thresholding method diverge, we artificially increased the problem difficulty by generating partially-synthetic datasets.
+To this end, we defined a parameter called ``excess background contamination'' that controls the extent to which the gRNA count distribution is separated.
+When excess background contamination is zero, we sample gRNA counts directly from the fitted GLM-EIV model.
+As excess background contamination increases, we sample gRNA counts from a slightly modified version of the fitted model in which the mean gRNA expression of \textit{unperturbed} cells is increased while the mean gRNA expression of \textit{perturbed} cells is held constant.
+Finally, when excess background contamination is $1$ (the maximum value), we set the mean gRNA expression of unperturbed cells equal to that of perturbed cells and sample gRNA counts from the resulting model.
+We hold fixed the real-data gene expressions, library sizes, covariates, and fitted perturbation probabilities in all settings.
+
+We generated partially-synthetic data in the above manner for each of the 322 positive control pairs in the Gasperini dataset, varying excess background contamination over the interval $[0,0.4].$ We then applied GLM-EIV and the thresholding method to analyze the data.
+We present results on two example pairs (the pair containing gene \textit{LRIF1} and the pair containing gene \textit{NDUFA2}) in Figures \ref{fig:real_data}c-d.
+We observed that the estimate produced by the methods on the raw data (depicted as a horizontal black line) coincided almost exactly with the estimate produced by the methods on the partially-synthetic data generated by setting excess background contamination to zero (This result replicated across nearly all pairs; average relative difference $0.003$.) We concluded that the fitted models accurately tracked the gRNA count distributions, giving us confidence in our procedure for generating realistic, partially-synthetic datasets.
+We found that as excess background contamination increased, the performance of thresholded regression degraded considerably while that of GLM-EIV remained stable.
+
+We generalized the above analysis to the entire set of positive control pairs.
+First, for each pair we computed its ``relative estimate change'' (REC) as a function of excess background contamination, defined as the relative difference between the estimate at a given level of excess contamination and zero excess contamination (Figure \ref{fig:real_data}d).
+Next, we computed the median REC across all positive control pairs (Figure \ref{fig:real_data}e; bands indicate pointwise 95\% CIs for the population median REC).
+As excess background contamination increased, thresholded regression exhibited severe attenuation bias (as reflected by large median REC values); GLM-EIV, by contrast, remained mostly stable.
+Finally, taking the estimate obtained on the raw data as ``ground truth,'' we computed the CI coverage rate across pairs as a function of excess contamination (Figure \ref{fig:real_data}f; bands indicate $95\%$ pointwise CIs for population coverage rate).
+GLM-EIV exhibited a significantly higher CI coverage rate than thresholded regression as the data became increasingly contaminated.
+
+\section{Discussion}
+
+In this work we introduced GLM-EIV (``GLM-based errors in variables''), a new method for single-cell CRISPR screen analysis.
+GLM-EIV generalizes the classical errors-in-variables model (when the latent predictor is binary) to arbitrary exponential distributions and link functions, facilitating the analysis of single-cell CRISPR screen data.
+We demonstrated through theoretical analyses, simulation studies, and real data analyses that GLM-EIV outperforms thresholded regression by a considerable margin in high background contamination settings.
+On the other hand, in low background contamination settings, GLM-EIV provides a statistically principled and empirically effective procedure for selecting a threshold for use in a thresholding method (e.g., thresholded negative binomial regression, as studied in this work, or even SCEPTRE \cite{Barry2020}).
+GLM-EIV thereby neutralizes a tuning parameter that exerts a significant impact on the results.
+Figure \ref{fig:dec_tree} summarizes how we anticipate GLM-EIV being used in practice.
+
+To our knowledge this is the first single-cell CRISPR screen paper oriented toward a statistical audience.
+We hope that this work helps to introduce the broader statistics community to an emerging class functional genomics assays that likely will exert a major impact on biology research in the coming years \cite{Przybyla2021}.
+Additionally, this is the first work to leverage the ondisc-Nextflow-cloud/HPC technology stack, a tightly-integrated, user-friendly, and powerful set of computational tools for large-scale single-cell analysis likely of interest to other researchers.
+
+We anticipate that GLM-EIV could be applied to other types of single-cell CRISPR screen and multimodal single-cell data.
+For example, we likely (with some effort) could extend GLM-EIV to ``low-multiplicity of infection'' screens \cite{Schraivogel2020} in which each cell receives one or two perturbations rather than dozens (as is the case in ``high multiplicity screens,'' studied in this work).
+We also likely could apply GLM-EIV to analyze multimodal single-cell chromatin accessibility assays.
+A question of interest in such experiments is whether chromatin state (i.e., closed or open) is associated with the expression of a gene (or abundance of a protein) \cite{Mimitou2021}.
+We do not directly observe chromatin state; instead, we observe tagged DNA fragments, counts that serve as proxies for whether a given region of chromatin is open or closed.
+GLM-EIV might be applied in such experiments to aid in the selection of thresholds or to analyze whole datasets.
+\printbibliography
+
+\clearpage
+\section*{Figures}
+% 1. Challenges
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=1\linewidth]{../../figures/analysis_challenges/plot.pdf}
+	\caption{\textbf{Experimental design and analysis challenges}: \textbf{a,} Experimental design.
+For a given perturbation (e.g., the perturbation indicated in blue), we partition the cells into two groups: perturbed and unperturbed.
+Next, for a given gene, we conduct a differential expression analysis across the two groups, yielding an estimate of the impact of the given perturbation on the given gene.
+\textbf{b,} DAG representing all variables in the system.
+The perturbation (latent) impacts both gene expression and gRNA expression; technical factors act as confounders, also impacting gene and gRNA expression.
+The target of estimation is the effect of the perturbation on gene expression.
+\textbf{c,} Schematic illustrating the ``background read'' phenomenon.
+Due to errors in the sequencing and alignment processes, unperturbed cells exhibit a nonzero gRNA count distribution (bottom).
+The target of estimation is the change in mean gene expression in response to the perturbation (top).
+\textbf{d}, Example data on four cells for a given perturbation-gene pair.
+Note that (i) the perturbation is unobserved, and (ii) the gene and gRNA data are discrete counts.}
+	\label{analysis_challenges}
+\end{figure}
+
+\clearpage
+% 2. Emprical challenges
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=1\linewidth]{../../figures/thresholding_empirical/plot}
+	\caption{\textbf{Empirical challenges of thresholded regression.} \textbf{a-b,} Fold change estimates produced by threshold = 1 versus threshold = 5 (a) and threshold = 20 versus threshold = 5 (b).
+The selected threshold substantially impacts the results.
+\textbf{c-d,} $p$-values (c) and CI widths (d) produced by threshold = 20 versus threshold = 5.
+The latter threshold yields more confident estimates.
+\textbf{e-f}, Empirical distribution of randomly-selected gRNA from Gasperini (e) and Xie (f) data (0 counts not shown).
+The gRNA data do not appear to imply an obvious threshold selection strategy.}
+	\label{thresholding_empirical}
+\end{figure}
+
+\clearpage
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=1\linewidth]{../../figures/thresholding_theoretical/plot}
+	\caption{\textbf{Theoretical challenges of thresholded regression.} \textbf{a,} Asymptotic relative bias versus threshold for different values of $\beta^g_1$.
+The bias function is highly nonconvex and strictly nonzero.
+Vertical blue lines, Bayes-optimal decision boundaries.
+Across all panels, $\beta^g_0 = 0$ and $\pi = 1/2$.
+\textbf{b,} Asymptotic relative bias versus $\pi$ when the threshold is set to a large number.
+The two quantities coincide exactly.
+\textbf{c,} Bias-variance decomposition for thresholding method in no-intercept model.
+Bias decreases and variance increases as the threshold tends to infinity.
+$\beta^g_1 = 1, \beta^m_1 = 1,$ and $\pi = 0.1$.}
+	\label{thresholding_theoretical}
+\end{figure}
+
+\clearpage
+% Theory
+\begin{figure}[h!]
 	\centering
@@
 
-We applied three methods to the simulated data: ``vanilla'' GLM-EIV (Algorithm \ref{algo:em_full}), accelerated GLM-EIV (Algorithm \ref{algo:pilot_estimates}), and thresholded regression.
-We used the Bayes-optimal decision boundary for classification as the threshold for the thresholding method.
-We ran all methods on the negative binomial data twice: once treating the size parameter $\theta$ as a known constant and once treating $\theta$ as unknown.
-In the latter case we used the \texttt{glm.nb} function from the \texttt{MASS} package to estimate $\theta$ \cite{Ripley2013}.
-We display the results of the simulation study in Figure \ref{main_text_sim}.
-Columns correspond to distributions (i.e., Poisson, negative binomial with known $\theta$, and negative binomial with unknown $\theta$), rows correspond to performance metrics (i.e., bias, mean squared error, CI coverage rate (nominal rate $95\%$), CI width, and method execution time).
-The problem difficulty parameter $\beta^g_1$ is plotted on the horizontal axis, and the methods are depicted in different colors (GLM-EIV = red, accelerated GLM-EIV = blue, and thresholding = purple; GLM-EIV masked by accelerated GLM-EIV in several panels).
-
-First, we observed that GLM-EIV dominated thresholded regression on all statistical metrics: GLM-EIV exhibited lower bias (row 1) and mean squared error (row 2) than thresholded regression; additionally, GLM-EIV had superior confidence interval coverage (row 3) despite having produced generally narrower confidence intervals (row 4).
-Intuitively, GLM-EIV outperformed the thresholding method because (i) GLM-EIV leveraged information from \textit{both} modalities (rather than the gRNA modality alone) to assign perturbation identities to cells, and (ii) GLM-EIV produced soft rather than hard assignments, capturing the inherent uncertainty in whether a perturbation occurred.
-Next, we found that accelerated GLM-EIV performed as well as ``vanilla'' GLM-EIV on the statistical metrics (rows 1-4) despite have substantially lower computational cost (bottom row).
-In fact, the execution time of accelerated GLM-EIV was approximately within an order of magnitude of that of the thresholding method (bottom row).
-
-Interestingly, thresholded regression exhibited better confidence interval coverage under estimated $\theta$ than under known $\theta$ (row 3).
-Estimating $\theta$ leads to slight inflation bias (i.e., overestimating the true effect size), whereas, as we showed previously, thresholding leads to attenuation bias (i.e., underestimating the true effect size).
-These phenomena partially cancel, yielding less biased estimates.
-We note that GLM-EIV and the thresholding method in principal are compatible with \textit{any} $\theta$ estimation procedure, including those based on more sophisticated techniques, such as regularization \cite{Hafemeister2019}.
-We defer rigorous investigation of the impact of different $\theta$ estimation strategies on these methods to future work.
-
-\section{Data analysis}
-
-Leveraging our computational infrastructure (Section \ref{sec:computing}), we applied GLM-EIV to analyze the entire Gasperini and Xie datasets (Figure \ref{fig:real_data}).
-Our theoretical analyses (Section \ref{sec:thresholding_theory}) and simulation studies (Section \ref{sec:simulation}) revealed that the Bayes-optimal decision boundary is a good choice for the threshold when the gRNA count distribution is well separated.
-Therefore, under the assumption that the effect of the perturbation on gRNA expression is similar across pairs, we used the fitted GLM-EIV models to approximate the Bayes-optimal decision boundary on both datasets, yielding a boundary of $3$ for Gasperini and $7$ for Xie.
-We then applied the thresholding method to the data, setting the threshold to the Bayes-optimal decision boundary estimated using GLM-EIV.
-
+\clearpage
 \begin{figure}
@@
 
-We compared GLM-EIV to thresholded regression on the real data, focusing specifically on the negative control pairs (i.e., gene-perturbation pairs that, by design, are expected to exhibit a fold change of $1$, or no association).
-We found that GLM-EIV and the thresholding method produced similar results (Figure \ref{fig:real_data}a-b): estimates, CI coverage rates, and CI widths were concordant.
-We note that GLM-EIV produced outlier estimates (likely due to non-global EM convergence) on a small ($< 2.5 \%$ on Gasperini, $<0.05\%$ on Xie) number of pairs consisting of a handful of genes (not plotted).
-The estimated effect of the perturbation on gene expression $\exp(\hat{\beta}_1^g)$ was unexpectedly large (95\% CI $[4306, 5186]$ and $[300, 316]$ on Gasperini and Xie data, respectively).
-We concluded that the datasets lay in an ``easy'' region of the parameter space, making thresholding a tenable strategy (provided the threshold is selected well).
-
-To explore the settings in which GLM-EIV and the thresholding method diverge, we artificially increased the problem difficulty by generating partially-synthetic datasets.
-To this end, we defined a parameter called ``excess background contamination'' that controls the extent to which the gRNA count distribution is separated.
-When excess background contamination is zero, we sample gRNA counts directly from the fitted GLM-EIV model.
-As excess background contamination increases, we sample gRNA counts from a slightly modified version of the fitted model in which the mean gRNA expression of \textit{unperturbed} cells is increased while the mean gRNA expression of \textit{perturbed} cells is held constant.
-Finally, when excess background contamination is $1$ (the maximum value), we set the mean gRNA expression of unperturbed cells equal to that of perturbed cells and sample gRNA counts from the resulting model.
-We hold fixed the real-data gene expressions, library sizes, covariates, and fitted perturbation probabilities in all settings.
-
-We generated partially-synthetic data in the above manner for each of the 322 positive control pairs in the Gasperini dataset, varying excess background contamination over the interval $[0,0.4].$ We then applied GLM-EIV and the thresholding method to analyze the data.
-We present results on two example pairs (the pair containing gene \textit{LRIF1} and the pair containing gene \textit{NDUFA2}) in Figures \ref{fig:real_data}c-d.
-We observed that the estimate produced by the methods on the raw data (depicted as a horizontal black line) coincided almost exactly with the estimate produced by the methods on the partially-synthetic data generated by setting excess background contamination to zero (This result replicated across nearly all pairs; average relative difference $0.003$.) We concluded that the fitted models accurately tracked the gRNA count distributions, giving us confidence in our procedure for generating realistic, partially-synthetic datasets.
-We found that as excess background contamination increased, the performance of thresholded regression degraded considerably while that of GLM-EIV remained stable.
-
-We generalized the above analysis to the entire set of positive control pairs.
-First, for each pair we computed its ``relative estimate change'' (REC) as a function of excess background contamination, defined as the relative difference between the estimate at a given level of excess contamination and zero excess contamination (Figure \ref{fig:real_data}d).
-Next, we computed the median REC across all positive control pairs (Figure \ref{fig:real_data}e; bands indicate pointwise 95\% CIs for the population median REC).
-As excess background contamination increased, thresholded regression exhibited severe attenuation bias (as reflected by large median REC values); GLM-EIV, by contrast, remained mostly stable.
-Finally, taking the estimate obtained on the raw data as ``ground truth,'' we computed the CI coverage rate across pairs as a function of excess contamination (Figure \ref{fig:real_data}f; bands indicate $95\%$ pointwise CIs for population coverage rate).
-GLM-EIV exhibited a significantly higher CI coverage rate than thresholded regression as the data became increasingly contaminated.
-
-\section{Discussion}
-
-In this work we introduced GLM-EIV (``GLM-based errors in variables''), a new method for single-cell CRISPR screen analysis.
-GLM-EIV generalizes the classical errors-in-variables model (when the latent predictor is binary) to arbitrary exponential distributions and link functions, facilitating the analysis of single-cell CRISPR screen data.
-We demonstrated through theoretical analyses, simulation studies, and real data analyses that GLM-EIV outperforms thresholded regression by a considerable margin in high background contamination settings.
-On the other hand, in low background contamination settings, GLM-EIV provides a statistically principled and empirically effective procedure for selecting a threshold for use in a thresholding method (e.g., thresholded negative binomial regression, as studied in this work, or even SCEPTRE \cite{Barry2020}).
-GLM-EIV thereby neutralizes a tuning parameter that exerts a significant impact on the results.
-Figure \ref{fig:dec_tree} summarizes how we anticipate GLM-EIV being used in practice.
-
-To our knowledge this is the first paper to examine single-cell CRISPR screens from a rigorous statistical perspective.
-We hope that this work can help introduce the broader statistics community to an emerging class functional genomics assays that likely will exert a major impact on biology research in the coming years \cite{Przybyla2021}.
-Additionally, this is the first work to leverage the ondisc-Nextflow-cloud/HPC technology stack, an (in our view) tightly-integrated, user-friendly, and powerful set of computational tools for out-of-memory and massively distributed single-cell analysis.
-
-We anticipate that GLM-EIV could be applied to other types of single-cell CRISPR screen and multimodal single-cell data.
-For example, we likely (with some effort) could extend GLM-EIV to ``low-multiplicity of infenction'' screens \cite{Schraivogel2020} in which each cell receives one or two perturbations rather than dozens (as is the case in ``high multiplicity screens,'' studied in this work).
-We also likely could apply GLM-EIV to analyze multimodal single-cell chromatin accessibility data.
-A question of interest in such experiments is whether chromatin state (i.e., closed or open) is associated with the expression of a gene (or abundance of a protein) \cite{Mimitou2021}.
-We do not directly observe chromatin state; instead, we observe transposase-tagged DNA fragments, counts that serve as proxies for whether a given region of chromatin is open or closed.
-GLM-EIV might be applied in such experiments to aid in the selection of thresholds or to analyze whole datasets.
-
-\begin{figure}
+\clearpage
+% Dec tree
+\begin{figure}[h!]
 	\centering
@@
 
-\bibliographystyle{unsrt}
-\newpage
-\bibliography{/Users/timbarry/optionFiles/glmeiv.bib}
-
+\clearpage
 \begin{appendices}
+\begin{refsection}
 \section{Theoretical details for thresholding estimator}
@@
 There were $17,028$ candidate \textit{cis} pairs, $97,818$ negative control pairs, and $322$ positive control pairs.
-
+\printbibliography
+\end{refsection}
 \end{appendices}

### 2021-12-14 text/jsm/jsm_paper.tex
update appendix

@@
+% set document class
 \documentclass[11pt]{article}
+
+% load packages
 \usepackage{setspace}
-\doublespacing
 \usepackage[margin=1in]{geometry}
@@
 \usepackage{array}
-
+\usepackage{titlesec} 
+\usepackage{xcolor}
+\usepackage{float}
+\usepackage[toc,page]{appendix}
+\usepackage{algpseudocode}
+\usepackage{afterpage}
 \usepackage[
@@
 
+% set options
+\doublespacing
 \addbibresource{/Users/timbarry/optionFiles/glmeiv.bib}
@@
 \newcolumntype{C}[1]{>{\PreserveBackslash\centering}p{#1}}
-\usepackage[toc,page]{appendix}
-\usepackage{algpseudocode}
 \renewcommand{\algorithmicrequire}{\textbf{Input:}}
@@
 \setcounter{tocdepth}{2}
-\usepackage{xcolor}
 \newtheorem{proposition}{Proposition}
@@
 \newcommand{\blue}[1]{\textcolor{blue}{#1}}
-\usepackage{float}
 \allowdisplaybreaks
@@
 \DeclareMathOperator*{\argmin}{arg\,min}
+\titleformat{\subsection}[runin]
+{\normalfont\large\bfseries}{\thesubsection}{1em}{}
+\titleformat{\subsubsection}[runin]
+{\normalfont\normalsize\bfseries}{\thesubsubsection}{1em}{}
+\newcommand\blankpage{%
+	\null
+	\thispagestyle{empty}%
+	\addtocounter{page}{-1}%
+	\newpage}
 
@@
 \begin{center} 
-\textbf{A new class of exponential family measurement error models applied to single-cell CRISPR screens} \\
-Timothy Barry, Kathryn Roeder, Eugene Katsevich
+\textbf{A class of exponential family measurement error models for CRISPR genome engineering and single-cell sequencing} \\
+Timothy Barry, Eugene Katsevich, Kathryn Roeder
 \end{center}
@@
 Despite their promise, single-cell CRISPR screens present substantial statistical challenges.
-We demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter.
-We recover these phenomena in precise theoretical terms in an idealized Gaussian setting.
-Next, we introduce GLM-EIV (``GLM-based errors-in-variables''), a new method for single-cell CRISPR screen analysis.
-GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed, overcoming limitations of thresholded regression in challenging problem settings.
+We demonstrate through theoretical and real data analyses that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter.
+To overcome these limitations, we introduce GLM-EIV (``GLM-based errors-in-variables''), a new method for single-cell CRISPR screen analysis.
+GLM-EIV extends the classical errors-in-variables model to response distributions and sources of measurement error that are (i) exponential family-distributed and (ii) potentially impacted by the same set of confounding variables.
 We develop a computational infrastructure to deploy GLM-EIV across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters.
-Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, yielding new biological insights.
+Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, demonstrating improved performance in challenging problem settings.
 \end{abstract}
@@
 First, we conduct a detailed study of the thresholding method.
-Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian setting.
+Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in a simplified Gaussian setting.
 Second, we introduce a new method, GLM-EIV (``GLM-based errors-in-variables''), for single-cell CRISPR screen analysis.
-GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.
+GLM-EIV extends the classical errors-in-variables model to response distributions and sources of measurement error that are (i) exponential family-distributed and (ii) potentially impacted by the same set of confounding variables.
 GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic.
@@
 For example, Lin et al.\ \cite{Lin2021} and Townes et al.\ \cite{Townes2019}  (separately) developed eSVD and GLM-PCA, generalizations of SVD and PCA, respectively, to exponential family response distributions.
-Unlike their vanilla counterparts, eSVD and GLM-PCA can model gene expression counts directly, thereby capturing the relationship between the mean and variance of a gene's expression level \cite{Lause2021}, improving performance on dimension reduction tasks.
+Unlike their vanilla counterparts, eSVD and GLM-PCA can model gene expression counts directly, thereby capturing the relationship between the mean and variance of a gene's expression level \cite{Lause2021} and improving performance on dimension reduction tasks.
 We see our work as a continuation of this effort to ``port'' common statistical methods and models to single-cell count data.
 Our focus, however, is on regression rather than dimension reduction.
-We extend the classical errors-in-variables model in several directions, most importantly generalizing to response distributions and sources of measurement error that are exponential family-distributed.
+We extend the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.
 
@@
 In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE.
-Most importantly, we aim to account for the fact that the perturbation is measured with noise, an issue that SCEPTRE circumvents via thresholding.
-Another difference is that we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is more challenging to attain within the nonparametric hypothesis testing framework of SCEPTRE.
+Most importantly, we account for the fact that the perturbation is measured with noise, an issue that SCEPTRE circumvents via thresholding.
+Additionally, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is challenging to attain within the nonparametric hypothesis testing framework of SCEPTRE.
 
@@
 \section{Thresholding method}
-We study thresholding from empirical and theoretical perspectives, highlighting several inherent difficulties of the approach.
+We study thresholding from empirical and theoretical perspectives, highlighting several inherent limitations of the approach.
 Gasperini and Xie imputed perturbation assignments onto the cells via thresholding, but they carried out the subsequent differential expression analysis in different ways: Gasperini fitted negative binomial regression models to the data, whereas Xie applied nonparametric tests of independence.
@@
 Exponentiating both sides of (\ref{thresh_glm}) yields
-$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m,$$ where $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration.
-
-Fold change in this context is defined as the ratio of the mean expression of perturbed cells to the mean expression of unperturbed cells.
+$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m.$$ We see that $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration.
+
+Fold change in this context is defined as the ratio of mean gene expression in perturbed cells to mean gene expression in unperturbed cells.
 The target of inference $\beta^m_1$ is the \textit{log} fold change in expression in response to the perturbation, controlling for the technical factors.
-Therefore, $\exp(\beta^m_1) = 1$ (eq., $\beta^m_1 = 0$) indicates no change in expression, while $\exp(\beta^m_1) > 1$ (eq., $\beta^m_1 > 0$) and $\exp(\beta^m_1) < 1$ (eq., $\beta^m_1 < 0$) indicate an increase and decrease in expression, respectively.
-
-\subsection{Empirical challenges of thresholding method}\label{sec:thresholding_empirical}
+Therefore, $\exp(\beta^m_1) = 1$ (eq., $\beta^m_1 = 0$) indicates no change in expression, whereas $\exp(\beta^m_1) > 1$ (eq., $\beta^m_1 > 0$) and $\exp(\beta^m_1) < 1$ (eq., $\beta^m_1 < 0$) indicate an increase and decrease in expression, respectively.
+
+\subsection{Empirical challenges of the thresholding method}\label{sec:thresholding_empirical}
 
@@
 
-To investigate the sensitivity of the thresholding method to the selected threshold, we deployed the thresholding method on the positive control data using three different choices for the threshold: 1, 5, and 20.
+To investigate the sensitivity of the thresholding method to the selected threshold, we deployed the thresholding method to the positive control data using three different choices for the threshold: 1, 5, and 20.
 We found that the chosen threshold substantially impacted the results (Figure \ref{thresholding_empirical}a-b).
-Estimates for log fold change produced by threshold = 1 were smaller in magnitude than those produced by threshold = 5.
-(Equivalently, estimates for \textit{raw} fold change were closer to the baseline of $1$ for threshold = 1; Figure \ref{thresholding_empirical}a.) Estimates produced by threshold = 5 and threshold = 20 were more concordant, but threshold = 20 yielded slightly larger effect sizes (Figure \ref{thresholding_empirical}b).
-
-We reasoned that the thresholding method systematically underestimates effect sizes on the positive control pairs, especially when the threshold is small.
-For a given perturbation, the vast majority ($>98\%$) of cells are unperturbed.
+Estimates for fold change produced by threshold = 1 were smaller in magnitude (i.e., closer to the baseline of $1$) than those produced by threshold = 5.
+(Figure \ref{thresholding_empirical}a.) Estimates produced by threshold = 5 and threshold = 20 were more concordant (Figure \ref{thresholding_empirical}b).
+
+We reasoned that thresholded regression systematically underestimated true effect sizes, especially for small thresholds.
+For a given perturbation, the majority ($>98\%$) of cells are unperturbed.
 This imbalance leads to an asymmetry: misclassifying \textit{unperturbed} cells as \textit{perturbed} is intuitively ``worse'' than misclassifying \textit{perturbed} cells as \textit{unperturbed}.
@@
 
-We hypothesized, however, that the reduction in bias obtained by selecting a large threshold comes at the cost of increasing the variance of the estimator.
-To investigate, we compared $p$-values and confidence intervals produced by threshold = 5 versus threshold = 20 for the target of inference $\beta^m_1$.
-We found that threshold = 5 yielded smaller (i.e., more significant) $p$-values and narrower confidence intervals than did threshold = 20 (Figure \ref{thresholding_empirical}c-d).
-We concluded that the threshold controls a bias-variance tradeoff: as the threshold increases, bias of the estimator decreases and variance increases.
+We hypothesized, however, that the reduction in bias obtained by selecting a large threshold causes the variance of the estimator to increase.
+To investigate, we compared $p$-values and confidence intervals produced by threshold = 5 and threshold = 20 for the target of inference $\beta^m_1$.
+We found that threshold = 5 yielded smaller (i.e., more significant) $p$-values and narrower confidence intervals than did threshold = 20 (Figures \ref{thresholding_empirical}c-d).
+We concluded that the threshold controls a bias-variance tradeoff: as the threshold increases, the bias of the estimator decreases and the variance increases.
 
@@
 
-\subsection{Theoretical challenges of thresholding method}\label{sec:thresholding_theory}
-
-Next, we study the thresholding method from a theoretical perspective, recovering in precise mathematical terms the attenuation bias and bias-variance tradeoff effects uncovered on real data, as well as several other interesting phenomena.
-We work in an idealized Gaussian setting.
-Suppose that we observe gRNA and gene expression data $\{(g_1, m_1), \dots, (g_n, m_n)\}$ on $n \in \N$ cells from the following model:
-
+\subsection{Theoretical challenges of the thresholding method}\label{sec:thresholding_theory}
+
+Next, we studied the thresholding method from a theoretical perspective, recovering in precise mathematical terms phenomena revealed in the empirical analysis (plus several others).
+We worked in a simplified Gaussian setting.
+Suppose we observe gRNA expression and gene expression data $(g_1, m_1), \dots, (g_n, m_n)$ on $n$ cells from the following linear model:
 \begin{equation}\label{theoretical_model}
-\begin{cases}
-m_i = \beta^m_0 + \beta^m_1 p_i + \ep_i \\
-g_i = \beta^g_0 + \beta^g_1 p_i + \tau_i \\
-p_i \sim \textrm{Bern}(\pi) \\
-\ep_i, \tau_i \sim N(0,1) \\
-p_i \indep \tau_i \indep \ep_i.
-\end{cases}
-\end{equation}
-For a given threshold $c \in \R$, the imputed perturbation assignment $\hat{p}_i$ is given by $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ for $\beta^m_1$ is $$\hat{\beta}^m_1 = \frac{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}}) (m_i - \overline{m})}{\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})^2 }.$$
-
+m_i = \beta^m_0 + \beta^m_1 p_i + \ep_i; \quad
+g_i = \beta^g_0 + \beta^g_1 p_i + \tau_i; \quad
+p_i \sim \textrm{Bern}(\pi); \quad
+\ep_i, \tau_i \sim N(0,1),
+\end{equation}
+where $p_i, \tau_i,$ and $\ep_i$ are independent.
+For a given threshold $c \in \R$, the imputed perturbation assignment $\hat{p}_i$ is $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ is the OLS solution, i.e.
+$\hat{\beta}^m_1 = \left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})(m_i - \overline{m})\right]\left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})^2\right]^{-1}.$ We begin by deriving the limit of $\hat{\beta}^m_1$:
 \begin{proposition}\label{prop:convergence}
@@
 	\begin{equation}\label{thresh_est_intercepts}
-\hat{\beta}^m_1 \xrightarrow{a.s.} \beta^m_1 \left(\frac{ \pi( \omega - \E[ \hat{p}_i ])}{ \E[\hat{p}_i] (1 - \E[\hat{p}_i])}\right),
-\end{equation}
-where
-$$\begin{cases}
-\E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi , \\
-\omega = \Phi\left(\beta_1^g + \beta_0^g -c \right) ,\\ \zeta = \Phi\left( \beta^g_0 - c \right). \\
-\end{cases}$$
+	\hat{\beta}^m_1 \xrightarrow{a.s.} \beta^m_1 \left(\frac{ \pi( \omega - \E[ \hat{p}_i ])}{ \E[\hat{p}_i] (1 - \E[\hat{p}_i])}\right) \equiv \beta^m_1 \gamma(\beta^g_1, \pi, c, \beta^g_0), 
+	\end{equation} where 
+	$$
+	\E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi; \quad \omega = \Phi\left(\beta_1^g + \beta_0^g -c \right); \quad \zeta = \Phi\left( \beta^g_0 - c \right).$$
 \end{proposition}
-Let $\gamma: \R^4 \to \R$ be defined by
-$$ \gamma(\beta^g_1, \pi, c, \beta^g_0) = \frac{\pi (\omega - \E[\hat{p}_i])}{\E[\hat{p}_i] (1 - \E[\hat{p}_i]) }.$$ We call $\gamma$ the ``attenuation function.'' Observe that \begin{itemize}
-\item[i.] $\gamma$ does not depend on $\beta^m_1$ or $\beta^m_0$, and
-\item[ii.] $\hat{\beta}^m_1 \xrightarrow{a.s.} [\gamma(\beta_0^g, \beta_1^g, c, \pi)] \beta^m_1.$
-\end{itemize}
-Let $b: \R^4 \to \R$ be the asymptotic relative bias of $\hat{\beta}^m_1$:
-\begin{multline*}
-b(\beta^g_1, \pi, c, \beta^g_0) = \left(\frac{1}{\beta^m_1}\right) \lim_{n\to\infty} \left(\beta^m_1 - \E[\hat{\beta}^m_1]\right) = \left( \frac{1}{\beta^m_1}\right)\left(\beta^m_1 - \E \left(\lim_{a.s.} \hat{\beta}^m_1 \right)\right) \\ = \frac{1}{\beta^m_1} \left(\beta^m_1 - \gamma(\beta^g_1, \pi, c, \beta^g_0)\beta^m_1\right) = 1 - \gamma(\beta^g_1, \pi, c, \beta^g_0),
-\end{multline*}
-where $\lim_{a.s.}$ denotes a.s.\ convergence. The asymptotic relative bias vanishes when the attenuation function equals $1$.
-
-\begin{center}
-\textbf{Bias as a function of threshold (Panel a)}
-\end{center}
-
-To investigate the basic question of ``What is a good threshold selection strategy?'', we study the relationship between the asymptotic relative bias $b$ of $\hat{\beta}^m_1$ and the selected threshold $c$.
-For simplicity, we begin by setting the perturbation probability $\pi$ to $1/2$.
-Let $c_\textrm{bayes} \in \R$ be the Bayes-optimal decision boundary for classifying cells as perturbed or unperturbed, i.e.
-$$c_\textrm{bayes} = \argmin_{c \in \R} \P(\hat{p}_i \neq p_i).$$ Simple algebra shows that $c_\textrm{bayes} = \beta_0^g + (1/2) \beta^g_1.$ Below, we give several results for the asymptotic relative bias $b$ of $\hat{\beta}^m_1$.
-We refer throughout to Figure \ref{thresholding_theoretical}a, which displays plots of asymptotic relative bias versus threshold for different values of $\beta^g_1$.
-We sometimes refer to ``asymptotic relative bias'' using the shortened term ``bias'' for succinctness.
-\textcolor{red}{SHOULD WE MOVE PROP 4-6 	TO THE APPENDIX?
-THESE PROPS ARE LESS IMPORTANT THAN THE OTHERS AND CONTRIBUTE SOMEWHAT LESS TO THE NARRATIVE.}
-
-\begin{itemize}
-\item 
-\begin{proposition}\label{prop:att_bias} Fix $\pi = 1/2$. For all $(\beta^g_1, c, \beta^g_0) \in \R^3$, the asymptotic relative bias is positive, i.e. 
-$$b(\beta^g_1, 1/2, c, \beta^g_0) > 0.$$
-\end{proposition}
-The thresholding method incurs strict attenuation bias (i.e., it \textit{under}estimates the true effect size) for all choices of the threshold and over all possible values of the model parameters (Figure \ref{thresholding_theoretical}a).
-Attenuation bias is a common attribute of estimators that ignore measurement in errors-in-variables models \cite{Stefanski2000a}.
-\item
-\begin{proposition}\label{prop:monotonic} Fix $\pi = 1/2$. The asymptotic relative bias $b$ decreases monotonically in $\beta_1^g$, i.e.
-	$$\frac{\partial b}{\partial(\beta^g_1)}\left(\beta^g_1, 1/2, c, \beta^g_0\right) \leq 0.$$
-\end{proposition}
-This result formalizes the intuition that the problem becomes easier as the gRNA mixture distribution becomes increasingly well-separated.
-To visualize Proposition (\ref{prop:monotonic}), one can fix a threshold (e.g., $c = 0$) and scan for bias across the panels.
-\item \begin{proposition}\label{prop:bayes_opt}
-For $\pi = 1/2$ and given $(\beta^g_1, \beta^g_0) \in \R^2$, the Bayes-optimal decision boundary $c_\textrm{bayes}$ is a critical value of the bias function $b$, i.e.
-$$ \frac{\partial b}{\partial c}\left(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0\right) = 0.$$
-\end{proposition}
-The Bayes-optimal decision boundary is an optimum (or possibly a saddle point) of the asymptotic relative bias function (Figure \ref{thresholding_theoretical}a, vertical blue lines).
-Interestingly, $c_\textrm{bayes}$ is in some cases a maximizer of the bias (Figure \ref{thresholding_theoretical}a, left) and in other cases a minimizer of the bias (Figure \ref{thresholding_theoretical}a, right).
-\item
-\begin{proposition}\label{prop:c_limit_half}
-Assume without loss of generality that $\beta^g_1 > 0$, and fix $\pi = 1/2$. As the threshold $c$ tends to infinity, the asymptotic relative bias $b$ tends to $1/2$, i.e.
-$$\lim_{c \to \infty} b(\beta^g_1, 1/2, c, \beta^g_0) = 1/2.$$
-\end{proposition}
-In other words, we always can set the threshold to a large number and attain a relative bias of $1/2$ (Figure \ref{thresholding_theoretical}a, all panels).
-This result establishes an upper bound on the bias of thresholded regression (under optimal threshold selection strategy).
-
- \item The following proposition compares the two threshold selection strategies introduced above (i.e., large number versus Bayes-optimal decision boundary) head-to-head.
- \begin{proposition}\label{prop:comparison} Assume without loss of generality that $\beta^g_1 > 0$.
-For $\beta^g_1 \in [0, 2\Phi^{-1}(3/4))$, we have that $$b(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0) > b(\beta^g_1, 1/2, \infty, \beta^g_0).$$ For $\beta^g_1 = 2\Phi^{-1}(3/4)$, we have that $$ b(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0) = b(\beta^g_1, 1/2, \infty, \beta^g_0).$$ Finally, for $\beta^g_1 \in (2\Phi^{-1}(3/4), \infty)$, we have that
- $$b(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0) < b(\beta^g_1, 1/2, \infty, \beta^g_0).$$
- \end{proposition}
-Setting the threshold to a large number yields a smaller bias when $\beta^g_1$ is small (i.e., $\beta^g_1 < 2\Phi^{-1}(3/4) \approx 1.35$; Figure \ref{thresholding_theoretical}a, left); setting the threshold to the Bayes-optimal decision boundary yields a smaller bias when $\beta^g_1$ is large (i.e., $\beta^g_1 > 2\Phi^{-1}(3/4)$; Figure \ref{thresholding_theoretical}a, right); and the two approaches coincide when $\beta^g_1$ is intermediate (i.e., $\beta^g_1 = 2\Phi^{-1}(3/4)$; Figure \ref{thresholding_theoretical}a, middle).
-\end{itemize}
-
-These results are subtle, but we can summarize them as follows.
-First, selecting a threshold that minimizes the bias is challenging, as there is no rule of thumb that we can apply universally (e.g., ``always choose the Bayes-optimal decision boundary'' or ``always choose a large number'') due to the complexity of the bias function.
-Second, even if we \textit{have} selected a good threshold, we incur nonzero attenuation bias.
-
-\begin{center}
-\textbf{Generalizing to $\pi \in [0,1/2]$ (Panel b)}
-\end{center}
-We generalize the expression for bias when the threshold is large to arbitrary $\pi \in [0,1/2]$:
-\begin{proposition}\label{prop:c_limit}
-Assume without loss of generality that $\beta^g_1 > 0$. As the threshold $c$ tends to infinity, the asymptotic relative bias $b$ tends to $\pi,$ i.e.
-$$ \lim_{ c \to \infty } b(\beta^g_1, \pi, c, \beta^g_0) = \pi.$$
-\end{proposition}
-In other words, if the perturbation probability is $\pi$, and if we set the threshold to a large number, then the asymptotic relative bias is $\pi$ (Figure \ref{thresholding_theoretical}b).
-We can understand this result intuitively by considering an extreme example: when $\pi$ is very small (e.g., $\pi = 0.01$), most cells are unperturbed.
-Therefore, as discussed in Section \ref{sec:thresholding_empirical}, selecting a large threshold minimizes the unperturbed-to-perturbed misclassification rate, reducing bias.
-% Therefore, in selecting a large threshold, we correctly classify nearly all unperturbed cells as unperturbed; on the other hand, the \textit{perturbed} cells that we misclassify as \textit{unperturbed} are swamped in number by the truly unperturbed cells, resulting in a small bias.
-
-\begin{center}
-	\textbf{Bias-variance tradeoff (Panel c)}
-\end{center}
-
-Finally, to shed light on the costs of selecting a large threshold, we derive an exact bias-variance decomposition for the thresholding estimator.
-We consider a slightly simpler, no-intercept version of (\ref{theoretical_model}) for this purpose:
-\begin{equation}\label{theoretical_model_no_int}
-\begin{cases}
-m_i = \beta_m p_i + \ep_i \\
-g_i = \beta_g p_i + \tau_i \\
-p_i \sim \textrm{Bern}(\pi) \\
-\ep_i, \tau_i \sim N(0,1) \\
-p_i \indep \tau_i \indep \ep_i.
-\end{cases}
-\end{equation}
-The thresholding estimator $\hat{\beta}_m$ in the no-intercept case is
-\begin{equation}\label{thresh_estimator_no_int}
- \hat{\beta}_m = \frac{ \sum_{i=1}^n \hat{p}_i m_i }{\sum_{i=1}^n \hat{p}_i^2}. 
-\end{equation}
+The function $\gamma: \R^4 \to \R$ does not depend on the gene expression parameters $\beta^m_1$ or $\beta^m_0$. The asymptotic relative bias $b: \R^4 \to \R$ of $\hat{\beta}^m_1$ is given by
+$$b(\beta^g_1, \pi, c, \beta^g_0)  \equiv \frac{1}{\beta^m_1} \left(\beta^m_1 - \lim_\textrm{a.s.} \hat{\beta}^m_1 \right) = 1 - \gamma(\beta^g_1, \pi, c, \beta^g_0).$$
+Having derived an exact expression for  the asymptotic relative bias of  $\hat{\beta}^m_1$, we can prove several results about this quantity.
+For the sake of simplicity, we consider a slightly specialized setting in which we set $\pi$ to to a specific value (see Appendix \ref{sec:appendix_theory} for details; proofs and detailed statements likewise deferred to Appendix \ref{sec:appendix_theory}).
+
+First, the thresholding estimator strictly \textit{underestimates} (in absolute value) the true value of $\beta^m_1$ over all choices of the threshold and over all possible values of the model parameters.
+This phenomenon, called attenuation bias, is a common attribute of estimators that ignore measurement in errors-in-variables models \cite{Stefanski2000a}.
+Second, the magnitude of the bias decreases monotonically in $\beta^g_1$, comporting with the intuition that the problem becomes easier as the gRNA mixture distribution becomes increasingly well-separated.
+Third, the Bayes-optimal decision boundary $c_\textrm{bayes} \in \R$ (i.e., the most accurate decision boundary for classifying cells as perturbed or unperturbed) is a critical value of the bias function.
+Finally, and most subtly, there is no universally applicable rule for selecting a threshold that produces minimal bias: when $\beta^g_1$ is small, setting the threshold to an arbitrarily large number yields smaller bias than setting the threshold to the Bayes decision boundary; when $\beta^g_1$ is large, the reverse is true.
+
+Next, we studied the variance of the thresholding estimator, considering a slightly simpler model for this purpose.
+Suppose the intercepts in (\ref{theoretical_model}) are fixed at $0$ (i.e., $\beta^m_0$ = $\beta^g_0$ = 0).
+For notational simplicity write $\beta_m = \beta^m_1$ and $\beta_g = \beta^g_1.$ The thresholding estimator $\hat{\beta}_m$ is the no-intercept OLS solution $\hat{\beta}_m = \left[\sum_{i=1}^n \hat{p}_i m_i \right]\left[\sum_{i=1}^n \hat{p}_i^2 \right]^{-1}.$ We have the following limiting distributional result.
 \begin{proposition}\label{prop:bv_decomp}
 	The limiting distribution of $\hat{\beta}_m$ is
-$$\sqrt{n}(\hat{\beta}_m - l) \xrightarrow{d} N\left(0, \frac{ \beta_m\omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2) }{\left(\E[\hat{p}_i]\right)^2} \right),$$ where
-$$\begin{cases}
-l = \beta_m\omega \pi/[\zeta(1-\pi) + \omega \pi], \\
-\E[\hat{p}_i] = \pi \omega + (1-\pi) \zeta, \\
-\omega = \Phi(\beta_g - c), \\
-\zeta = \Phi(- c).\\
-\end{cases}
-$$
+	$$\sqrt{n}(\hat{\beta}_m - l) \xrightarrow{d} N\left(0, \frac{ \beta_m \omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2) }{\left(\E[\hat{p}_i]\right)^2} \right),$$ where $$l = \beta_m \omega \pi/[\zeta(1-\pi) + \omega \pi]; \quad
+	\E[\hat{p}_i] = \pi \omega + (1-\pi) \zeta; \quad
+	\omega = \Phi(\beta_g - c); \quad
+	\zeta = \Phi(-c).$$
 \end{proposition}
-This result yields an exact bias-variance decomposition for $\hat{\beta}_m$ for large $n$ (Figure \ref{thresholding_theoretical}c).
-As the threshold tends to infinity, the bias decreases and the variance increases, consistent with the intuition that a large threshold reduces the misclassification rate at the cost of decreasing the ``effective sample size.'' The best strategy for maximizing estimation accuracy (as quantified by mean squared error) is to select a threshold that induces moderate bias.
-A downside of this approach, however, is that constructing valid confidence intervals becomes more challenging.
-
-\subsection{Thresholding method summary}
-
-Empirical and theoretical analyses reveal that the thresholding method poses several challenges: the threshold is a tuning parameter that substantially impacts the results; strict attenuation bias obtains uniformly over the parameter space and for all choices of the threshold; and there does not exist and obvious threshold selection strategy due to (i) the unimodality of the empirical gRNA count distributions and (ii) the existence of a bias-variance tradeoff mediated by the threshold.
-These difficulties motivate our core research question: \textit{Does modeling the gRNA count distribution directly, thereby circumventing the need to threshold altogether, lead to simpler, more accurate estimation and inference in single-cell CRISPR screen analysis?} To answer this question, we generalize the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.
-
-\section{GLM-based errors-in-variables}
-
-In this section we introduce generalized linear model with errors-in-variables (GLM-EIV), derive estimation and inference procedures for the model, and propose several statistical accelerations to reduce the cost of fitting the model.
+This proposition yields an asymptotically exact bias-variance decomposition for $\hat{\beta}_m$ (Figure \ref{thresholding_theoretical}): as the threshold tends to infinity, the bias decreases and the variance increases, consistent with empirical observations that we made in Section \ref{sec:thresholding_empirical}.
+Overall, our empirical and theoretical analyses reveal that thresholded regression poses several difficulties: thresholding incurs strict attenuation bias (that in some settings is large), and selecting a good threshold is challenging.
+These considerations motivate our core research question: \textit{does modeling the gRNA counts directly, thereby circumventing thresholding, facilitate estimation and inference in single-cell CRISPR screen analysis}?
+
+\section{GLM-EIV}
+
+To answer this question, we introduce GLM-EIV (GLM-based errors-in-variables), a method that models the entire data-generating process.
 
 \subsection{Model}\label{sec:model}
-\subsubsection*{Negative binomial model}
-Building on the work of several previous authors \cite{Townes2019,Svensson2020,Hafemeister2019}, Sarkar and Stephens \cite{Sarkar2021} proposed a simple strategy for modeling for single-cell gene expression data, which, in the framework of negative binomial GLMs, is equivalent to using the log-transformed library size as an offset term (as in (\ref{thresh_glm})).
+Building on the work of several previous authors \cite{Townes2019,Svensson2020,Hafemeister2019}, Sarkar and Stephens \cite{Sarkar2021} proposed a simple strategy for modeling single-cell gene expression data, which, in the framework of negative binomial GLMs, is equivalent to using the log-transformed library size as an offset term (as in (\ref{thresh_glm})).
 We generalize Sarkar and Stephens' approach to model \textit{both} gene and gRNA modalities.
@@
 \begin{equation}\label{glmeiv_model_1}
-m_i |(p_i, z_i, d^m_i) \sim \textrm{NB}(\mu_i^m),
-\end{equation}
-\begin{equation}\label{glmeiv_model_2}
-\log(\mu^m_i) = \beta^m_0 + \beta^m_1 p_i + \gamma_m^T z_i + \log(d^m_i),
+m_i |(p_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i^m); \quad \log(\mu^m_i) = \beta^m_0 + \beta^m_1 p_i + \gamma_m^T z_i + \log(d^m_i),
 \end{equation}
@@
 The model for the gRNA counts is
-\begin{equation}\label{glmeiv_model_3}
-g_i | (p_i, z_i, d^g_i) \sim \textrm{NB}_{\theta^g}\left(\mu_i^g\right),
-\end{equation}
-\begin{equation}\label{glmeiv_model_4}
-\log(\mu_i^g) = \beta^g_0 + \beta^g_1p_i + \gamma^T_g z_i + \log(d^g_i),
+\begin{equation}\label{glmeiv_model_2}
+g_i | (p_i, z_i, d^g_i) \sim \textrm{NB}_{\theta^g}\left(\mu_i^g\right); \quad \log(\mu_i^g) = \beta^g_0 + \beta^g_1p_i + \gamma^T_g z_i + \log(d^g_i),
 \end{equation}
@@
 We use a negative binomial GLM to model the gRNA counts because the gRNA molecules are transcribed in the cell in the same way as gene transcripts \cite{Datlinger2017,Hill2018}.
-Finally, we model the marginal perturbation probability as \begin{equation}\label{glmeiv_model_5}
-p_i \sim \textrm{Bern}(\pi),
-\end{equation} where $\pi \in (0,1/2]$. 
- Together, (\ref{glmeiv_model_1}, \ref{glmeiv_model_2}, \ref{glmeiv_model_3}, \ref{glmeiv_model_4}, \ref{glmeiv_model_5}) define the standard GLM-EIV model.
+Finally, we model the marginal perturbation probability as $
+p_i \sim \textrm{Bern}(\pi),$ where $\pi \in (0,1/2]$. 
+ Together, (\ref{glmeiv_model_1}), (\ref{glmeiv_model_2}), and the marginal distribution of $p_i$ define the negative binomial GLM-EIV model.
 The terms $(\beta^m_0 + \beta^m_1p_i + \gamma^T_m z_i)$ and $(\beta^g_0 + \beta^g_1 p_i + \gamma^T_g z_i)$ can be interpreted as relative gene and gRNA expressions, similar to the analogous term in the thresholding model.
 Likewise, the target of inference $\beta^m_1$ is the log fold change in gene expression in response to the perturbation, accounting for technical factors.
+See Appendix \ref{sec:zero_inf_model} for a parallel methodology that we developed in which we model the gRNA distribution as (in a certain sense) zero-inflated counts.
 
@@
 (We use the tilde as a reminder that the vector is partially unobserved.) Let $\beta_m = [\beta^m_0, \beta^m_1, \gamma_m]^T \in \R^d$ and $\beta_g = [\beta^g_0, \beta^g_1, \gamma_g]^T \in \R^d$ be the unknown coefficient vectors corresponding to the gene and gRNA expression models, respectively.
-Finally, let $o^m_i$ and $o^g_i$ be the (possibly zero) offset terms for the gene and gRNA models; in practice, we typically set $o^m_i$ and $o^g_i$ to $\log(d^m_i)$ and $\log(d^g_i)$, respectively.
-
-We use a GLM approach to model the gene and gRNA expressions.
-Considering first the gene expression model, let the $i$th linear component $l^m_i$ of the model be $$l^m_i = \langle \tilde{x}_i, \beta_m \rangle + o^m_i.$$ Let the mean $\mu^m_i$ of the $i$th observation be
-$$r_m(\mu^m_i) = l^m_i,$$ where $r_m:\R \to \R$ is a strictly increasing, differentiable link function.
+Finally, let $o^m_i$ and $o^g_i$ be the (possibly zero) offset terms for the gene and gRNA models; in practice, we typically set $o^m_i$ and $o^g_i$ to the log-transformed library sizes (i.e., $\log(d^m_i)$ and $\log(d^g_i)$, respectively).
+
+We use GLMs to model the gene and gRNA expressions.
+Considering first the gene expression model, let the $i$th linear component $l^m_i$ of the model be $l^m_i = \langle \tilde{x}_i, \beta_m \rangle + o^m_i.$ Next, let the mean $\mu^m_i$ of the $i$th observation be $r_m(\mu^m_i) = l^m_i,$ where $r_m:\R \to \R$ is a strictly increasing, differentiable link function.
 Let $\psi_m: \R \to \R$ be the differentiable, cumulant-generating function of the selected exponential family distribution.
 We can express the canonical parameter $\eta^m_i$ in terms of $\psi_m$ and $r_m$ by
-$$\eta^m_i = \left([\psi'_m]^{-1} \circ r^{-1}_m\right)(l_i^m) := h_m(l_i^m).$$ Finally, let $c_m: \R \to \R$ be the carrying density of the selected exponential family distribution.
+$\eta^m_i = \left([\psi'_m]^{-1} \circ r^{-1}_m\right)(l_i^m) := h_m(l_i^m).$ Finally, let $c_m: \R \to \R$ be the carrying density of the selected exponential family distribution.
 The density $f_m$ of $m_i$ conditional on the the canonical parameter $\eta_i$ is
-$$f_m(m_i; \eta^m_i) = \exp\left\{ m_i \eta^m_i - \psi_m(\eta^m_i) + c_m(m_i) \right\}.$$ The function $c_m$ does not appear in the log likelihood of $m_i$; therefore, the only functions relevant to inference are $\psi_m$ and $r_m$.
+$f_m(m_i; \eta^m_i) = \exp\left\{m_i \eta^m_i - \psi_m(\eta^m_i) + c_m(m_i) \right\}.$ The function $c_m$ does not appear in the log likelihood of $m_i$; therefore, the only functions relevant to inference are $\psi_m$ and $r_m$.
 
@@
 $$
-\begin{cases}
-l^g_i = \langle \tilde{x}_i, \beta_g \rangle + o^g_i,\\
-r_g(\mu^g_i) = l^g_i,\\
+l^g_i = \langle \tilde{x}_i, \beta_g \rangle + o^g_i; \quad
+r_g(\mu^g_i) = l^g_i; \quad
 \eta^g_i = \left([\psi'_g]^{-1} \circ r^{-1}_g\right)(l_i^g) := h_g(l_i^g).
-\end{cases}
-$$
-The density $f_g$ of $g_i$ given the canonical parameter is
-$$f_g(m_i; \eta^g_i) = \exp\left\{g_i \eta^g_i - \psi_g(\eta^g_i) + c_g(g_i)\right\}.$$
-Finally, the unobserved variable $p_i$ is assumed to follow a Bernoulli distribution with mean $\pi \in (0, 1/2]$. Its marginal density $f_p$ is given by
-$$f_p(p_i) = \pi^{p_i}(1-\pi)^{1 - p_i}.$$
+$$
+The density $f_g$ of $g_i$ given the canonical parameter is $f_g(m_i; \eta^g_i) = \exp\left\{g_i \eta^g_i - \psi_g(\eta^g_i) + c_g(g_i)\right\}.$
+Finally, the unobserved variable $p_i$ is assumed to follow a Bernoulli distribution with mean $\pi \in (0, 1/2]$. Its marginal density $f_p$ is given by $f_p(p_i) = \pi^{p_i}(1-\pi)^{1 - p_i}.$
 The unknown parameters in the model are
@@
 $$
-\begin{cases}
-l^m_i(j) := \langle \tilde{x}_i(j), \beta_m \rangle + o^m_i \\ \eta^m_i(j) := h_m(l^m_i(j)) \\
+l^m_i(j) := \langle \tilde{x}_i(j), \beta_m \rangle + o^m_i; \quad \eta^m_i(j) := h_m(l^m_i(j)); \quad
 \mu_i^m(j) = r_m^{-1}(l^m_i(j)).
-\end{cases}
-$$
-Let the corresponding gRNA quantities $l^g_i(j)$, $\eta_i^g(j)$, and $\mu^g_i(j)$ be defined analogously. Define the observed design matrix $X \in  \R^{n \times {d-1}}$ by 
+$$
+Let the corresponding gRNA quantities $l^g_i(j)$, $\eta_i^g(j)$, and $\mu^g_i(j)$ be defined analogously.
+Let $X \in \R^{n \times {d-1}}$ be the observed design matrix, and let $\tilde{X} \in \R^{n \times d}$ be the augmented design matrix that results from concatenating the column of (unobserved) $p_i$s to $X$, i.e.
 $$X := \begin{bmatrix} 
@@
 1 & z_n
-\end{bmatrix}.$$ 
-Let $\tilde{X} \in \R^{n \times d}$ be the augmented design matrix that results from concatenating the column of (unobserved) $p_i$s to $X$, i.e.
-$$ \tilde{X}  := 
+\end{bmatrix}; \quad 
+\tilde{X}  := 
 \begin{bmatrix}
@@
 \tilde{x}_1^T \\ \tilde{x}_2^T \\ \vdots \\ \tilde{x}_n^T
-\end{bmatrix}.$$ Furthermore, for $j \in \{0,1\}$, let $\tilde{X}(j) \in \R^{n \times d}$ be the matrix that results from setting $p_i$ to $j$ for all $i \in \{1, \dots, n\}$ in $\tilde{X}$.
-Finally, let $$\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix} = [\tilde{X}(0)^T, \tilde{X}(1)^T]^T$$ be the $\R^{2n \times d}$ matrix that results from vertically concatenating $\tilde{X}(0)$ and $\tilde{X}(1)$.
-
-Next, define $m := [m_1, \dots, m_n]$, and let $g$, $p$, $o^m$, and $o^g$ be defined analogously. Also, let $[m,m]^T \in \R^{2n}$ be the vector that results from concatenating $m$ to itself, i.e.
+\end{bmatrix}.
+$$ 
+Furthermore, for $j \in \{0,1\}$, let $\tilde{X}(j) \in \R^{n \times d}$ be the matrix that results from setting $p_i$ to $j$ for all $i \in \{1, \dots, n\}$ in $\tilde{X}$, and let  $[\tilde{X}(0)^T, \tilde{X}(1)^T]^T$ denote the $\R^{2n \times d}$ matrix that results from vertically concatenating $\tilde{X}(0)$ and $\tilde{X}(1)$.
+Furthermore, define $m := [m_1, \dots, m_n]$, and let $g$, $p$, $o^m$, and $o^g$ be defined analogously.
+Finally, let $[m,m]^T \in \R^{2n}$ be the vector that results from concatenating $m$ to itself, i.e.
 $$ [m,m]^T := [\underbrace{m_1, m_2 \dots, m_{n-1}, m_n}_\textrm{first copy of $m$}, \underbrace{m_1, m_2, \dots, m_{n-1}, m_n}_\textrm{second copy of $m$}],$$ and let $[g,g]^T$, $[o^g,o^g]^T$, and $[o^m,o^m]^T$ be defined similarly.
@@
 We derive the log-likelihood of the GLM-EIV model.
-We conduct estimation and inference \textit{conditional} on the library sizes and technical factors $l^m_i, l^g_i,$ and $z_i$; therefore, we treat these quantities as fixed constants.
-We assume that the gene expression $m_i$ and gRNA expression $g_i$ are \textit{conditionally independent} given the perturbation $p_i$.
-The joint density $f$ of $(m_i, g_i, p_i)$ given $\theta$ is
+We conduct estimation and inference conditional on the library sizes and technical factors $l^m_i, l^g_i,$ and $z_i$; therefore, we treat these quantities as fixed constants.
+We assume that the gene expression $m_i$ and gRNA expression $g_i$ are conditionally independent given the perturbation $p_i$.
+The the joint density $f$ of $(m_i, g_i, p_i)$ given $\theta$ is $ $
 \begin{equation}\label{full_density}
@@
 \end{equation}
-The complete-data log-likelihood is
-\begin{multline}\label{full_log_lik}
-\mathcal{L}(\theta; m, g, p) = \sum_{i=1}^n \log\left( \pi^{p_i}(1-\pi)^{1-p_i} \right) \\ + \sum_{i=1}^n \log\left( f_m(m_i; \eta^m_i)\right) + \sum_{i=1}^n \log\left( f_g(g_i; \eta_i^g) \right).
-\end{multline}
 Integrating over the unobserved variable $p_i$, we can write the density $f$ of $(m_i, g_i)$ as
@@
 \end{equation}
-Finally, the log-likelihood is
-\begin{multline}\label{marginal_log_lik}
-\mathcal{L}(\theta; m, g) \\ = \sum_{i=1}^n \log\left[(1-\pi) f_m(m_i; \eta^m_i(0)) f_g(g_i; \eta^g_i(0)) + \pi f_m(m_i; \eta^m_i(1)) f_g(g_i; \eta^g_i(1)) \right]
-\end{multline}
+Therefore, the log-likelihood is
+\begin{equation}\label{marginal_log_lik}
+\mathcal{L}(\theta; m, g) = \sum_{i=1}^n \log\left[(1-\pi) f_m(m_i; \eta^m_i(0)) f_g(g_i; \eta^g_i(0)) + \pi f_m(m_i; \eta^m_i(1)) f_g(g_i; \eta^g_i(1)) \right].
+\end{equation}
 
 We see from (\ref{marginal_density}) that the GLM-EIV model is equivalent to a two-component mixture of \textit{products} of GLM densities.
-Additionally, the GLM-EIV model is a generalization of the classical errors-in-variables model (when the predictor is binary).
-Suppose that we observe data $(x_1, y_1), \dots, (x_n, y_n)$ from the following model:
+Additionally, the GLM-EIV model is a generalization the simple errors-in-variables model (when the predictor is binary); the latter is defined as follows:
 \begin{equation}\label{classical_eiv}
-\begin{cases}
-y_i = \beta_0 + \beta_1 x^*_i + \ep_i \\
+y_i = \beta_0 + \beta_1 x^*_i + \ep_i; \quad
 x_i = x^*_i + \tau_i,
-\end{cases}
-\end{equation}
-where $x^*_i \sim \textrm{Bern}(\pi), \ep_i \sim N(0,1), \tau_i \sim N(0,1),$ and $\ep_i$,$\tau_i$, and $x^*_i$ are independent.
-The model (\ref{classical_eiv}) is a GLM-EIV model with identity link, Gaussian response, and no covariate terms.
+\end{equation}
+where, $x^*_i \sim \textrm{Bern}(\pi), \ep_i, \tau_i \sim N(0,1),$ and $\ep_i$,$\tau_i$, and $x^*_i$ are independent.
+GLM-EIV extends (\ref{classical_eiv}) in at least three directions: first, GLM-EIV allows $y_i$ and $x_i$ to follow (possibly different) exponential family distributions; second, GLM-EIV allows $y_i$ and $x_i$ to be related to $x^*_i$ through (possibly different) nonlinear link functions; and finally, GLM-EIV allows confounders $z_i$ to impact both $x_i$ and $y_i$.
+In such settings $x_i$ and $y_i$ are dependent \textit{even after conditioning on $x^*_i$}, enabling GLM-EIV to capture more complex dependence relationships between $x_i$ and $y_i$ than is possible in (\ref{classical_eiv}) or other standard measurement error models.
 
@@
 We derive an EM algorithm (Algorithm \ref{algo:em_full}) to estimate the parameters of the GLM-EIV model.
-The E step entails computing the membership probability (i.e., the probability of perturbation) of each cell.
+The E step entails computing the membership probability (i.e., the probability of perturbation) in each cell.
 The membership probability $T_i(1)$ of cell $i \in \{1, \dots, n\}$ given the current parameter estimates $(\beta_m^\textrm{(t)}, \beta_g^\textrm{(t)}, \pi^\textrm{(t)})$ and observed data $(m_i, g_i)$ is
 $$T_i(1) = \P(p_i = 1| M_i = m_i, G_i = g_i, \beta^\textrm{(t)}_m, \beta^\textrm{(t)}_g, \pi^\textrm{(t)}).$$ We can calculate this quantity by applying (i) Bayes rule, (ii) the conditional independence property of $M_i$ and $G_i$, (iii) the density of $M_i$ and $G_i$, and (iv) a log-sum-exp-type trick to ensure numerical stability.
-Next, we produce updated estimates $\pi^{(t +1)}$, $\beta_g^{(t+1)}$, and $\beta_m^{(t+1)}$ of the parameters by maximizing the M step objective function.
-It turns out that maximizing the objective function is equivalent to setting $\pi^{\textrm{(t+1)}}$ to the mean of the current membership probabilities and setting $\beta_g^{(t+1)}$ and $\beta_m^{(t+1)}$ to the fitted coefficients of a GLM weighted by the current membership probabilities (Algorithm \ref{algo:em_full}).
-We iterate through the E and M steps until the marginal log likelihood (\ref{marginal_log_lik}) converges (see appendix for full details).
+Next, we produce updated estimates $\pi^{(\textrm{t} +1)}$, $\beta_g^{(\textrm{t}+1)}$, and $\beta_m^{(\textrm{t}+1)}$ of the parameters by maximizing the M step objective function.
+It turns out that maximizing this objective function is equivalent to setting $\pi^{\textrm{(t+1)}}$ to the mean of the current membership probabilities and setting $\beta_g^{(\textrm{t}+1)}$ and $\beta_m^{(\textrm{t}+1)}$ to the fitted coefficients of a GLM weighted by the current membership probabilities (Algorithm \ref{algo:em_full}).
+We iterate through the E and M steps until the log likelihood (\ref{marginal_log_lik}) converges (see Appendix \ref{sec:glmeiv_details}).
 Our EM algorithm is reminiscent of (but distinct from) that of Ibrahim \cite{Ibrahim1990}, who also applied weighted GLM solvers to carry out the M step.
@@
 		 \EndFor
-		\State Compute marginal likelihood using $\beta_m^\textrm{curr}$, $\beta_g^\textrm{curr}$, and $\pi^\textrm{curr}$.
+		\State Compute log likelihood using $\beta_m^\textrm{curr}$, $\beta_g^\textrm{curr}$, and $\pi^\textrm{curr}$.
 		\EndWhile
@@
 After fitting the model, we perform inference on the estimated parameters.
-The easiest approach, given the complexity of the log likelihood, would be to run a parametric bootstrap on the fitted model.
+The easiest approach, given the complexity of the log likelihood, would be to run a bootstrap.
 This strategy, however, is prohibitively slow, as the data are large and the EM algorithm is iterative.
-Therefore, we derive an analytic formula for the asymptotic observed information matrix using Louis's Theorem \cite{Louis1982}.
-Leveraging this analytic formula, we can calculate standard errors (and $p$-values and confidence intervals) quickly, enabling us to perform inference in practice on real, large-scale data.
-The derivation is fairly involved (see appendix); simulation studies confirm the correctness of the results (see Section \ref{sec:simulation}) .
+Therefore, we derive an analytic formula for the asymptotic observed information matrix using Louis's Theorem \cite{Louis1982} (Appendix \ref{sec:glmeiv_details}).
+Leveraging this analytic formula, we can calculate standard errors quickly, enabling us to perform inference in practice on real, large-scale data.
 
@@
 A downside of the the EM algorithm (Algorithm \ref{algo:em_full}) is that it requires fitting many GLMs.
-Assuming that we run the algorithm 15 times using randomly-generated pilot estimates (to ensure convergence to the global maximum), and assuming that the algorithm iterates through E and M steps about 10 times per run, we must fit approximately 300 GLMs.
-(These numbers are based on exploratory applications of the method to real and simulated data.) We devised a strategy (Algorithm \ref{algo:pilot_estimates}) to produce a highly accurate pilot estimate $(\pi^\textrm{pilot}, \beta_m^\textrm{pilot}, \beta_g^\textrm{pilot})$ of the true parameters, enabling us to run the algorithm once and converge upon the MLE within a few iterations.
+Assuming that we run the algorithm 15 times using randomly-generated pilot estimates, and assuming that the algorithm iterates through E and M steps about 10 times per run, we must fit approximately 300 GLMs.
+(These numbers are based on exploratory applications of the method to real and simulated data.) We devised a strategy to produce a highly accurate pilot estimate $(\pi^\textrm{pilot}, \beta_m^\textrm{pilot}, \beta_g^\textrm{pilot})$ of the true parameters, enabling us to run the algorithm once and converge upon the MLE within a few iterations.
 The strategy involves layering several statistical ``tricks'' (some new, some old) on top of one another.
-We expect these tricks to be independently useful for accelerating other single-cell methods (e.g., SCEPTRE \cite{Barry2020}).
+We expect these tricks to be independently useful for accelerating other single-cell methods.
 
 \begin{algorithm}
-	\caption{Computing pilot parameter estimates.}\label{algo:pilot_estimates}
+	\caption{Computing $[\beta_0^m]^\textrm{pilot}$, $[\gamma^T_m]^\textrm{pilot}$, $[\beta_0^g]^\textrm{pilot}$, and $[\gamma^T_g]^\textrm{pilot}$ .}\label{algo:pilot_estimates_1}
 	\begin{algorithmic}[2]
@@
 		\EndFor
+		\State \textbf{return} $([\beta_0^m]^\textrm{pilot}$, $\hat{f}^m$, $[\gamma^T_m]^\textrm{pilot}$, $[\beta_0^g]^\textrm{pilot}$, $[\gamma^T_g]^\textrm{pilot}$, $\hat{f}^g)$
+	\end{algorithmic}
+\end{algorithm}
+
+The first step (Algorithm \ref{algo:pilot_estimates_1}) is to obtain good parameter estimates for $[\beta^m_0, \gamma_m]^T$ and $[\beta^g_0, \gamma_g]^T$ via regression.
+Recall that the underlying gene expression parameter vector $\beta_m$ is $\beta_m = [\beta^m_0, \beta^m_1, \gamma_m]^T \in \R^d$, where $\beta^m_0$ is the intercept, $\beta^m_1$ is the effect of the perturbation, and $\gamma_m^T$ is the effect of the technical factors.
+To produce estimates $[\beta^m_0]^\textrm{pilot}$ and $[\gamma_m^T]^\textrm{pilot}$, we regress the gene expressions $m$ onto the technical factors $X$.
+The intuition for this procedure is as follows: the probability of perturbation $\pi$ is very small.
+Therefore, the true log likelihood is approximately equal to the log likelihood that results from omitting $p_i$ from the model:
+\begin{multline*}
+\sum_{i=1}^n f_m(m_i; \eta^m_i) = \underbrace{\sum_{i : p_i =1} f_m(m_i; h_m(\beta_0 + \beta_1 + \gamma^T z_i + o^m_i))}_\textrm{few terms}  + \underbrace{\sum_{i : p_i = 0} f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i))}_\textrm{many terms} \\ \approx \sum_{i = 1}^n f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i)).
+\end{multline*}
+We similarly can obtain pilot estimates $[\beta^g_0]^\textrm{pilot}$ and $[\gamma^T_g]^\textrm{pilot}$ by regressing the gRNA counts $g$ onto the technical factors $X$.
+We extract the fitted values (on the scale of the linear component) for use in a subsequent step: $\hat{f}^k_i = [\beta^k_0]^\textrm{pilot} + \langle [\gamma^T_k]^\textrm{pilot}, z_i \rangle + o^k_i,$ for $k \in \{m,g\}$.
+
+Next, we obtain estimates $[\beta_1^m]^\textrm{pilot},$ $[\beta_1^g]^\textrm{pilot},$ and $\pi^\textrm{pilot}$ for $\beta^m_1$, $\beta^g_1$, and $\pi$ by fitting a ``reduced'' GLM-EIV.
+We describe the reduced GLM-EIV procedure in text; the algorithm box (Algorithm \ref{algo:pilot_estimates_2}) is deferred to Appendix \ref{sec:statistical_accelerations}.
+The log likelihood of the no-intercept, univariate GLM with predictor $p_i$ and offset $\hat{f}^m_i$ is approximately equal to the true log likelihood:
+$$ \sum_{i=1}^n f_m(m_i; \eta^m_i) =
+\sum_{i=1}^n f_m(m_i; h_m(\beta_0 + \beta_1 p_i + \gamma^T z_i + o^m_i)) \approx \sum_{i=1}^n f_m(m_i; h_m( \beta_1 p_i + \hat{f}^m_i)).
+$$
+Therefore, to estimate $\beta^m_1$, $\beta^g_1$, and $\pi$, we fit a GLM-EIV model with gene expressions $m$, gRNA counts $g$, gene offsets $\hat{f}^m := [\hat{f}^m_1, \dots, \hat{f}^m_n]^T$, gRNA offsets $\hat{f}^g := [\hat{f}^g_1, \dots, \hat{f}^g_n]^T$, and \textit{no} intercept or covariate terms.
+Intuitively, we ``encode'' all information about technical factors, library sizes, and baseline expression levels into $\hat{f}^m$ and $\hat{f}^g$.
+We run the algorithm $B \approx 15$ times over randomly-selected starting values for $\beta^m$, $\beta^g$, and $\pi$ and select the solution with greatest the log likelihood.
+
+The M step of the reduced GLM-EIV algorithm requires fitting two no-intercept, univariate GLMs with offsets.
+We derive analytic formulas for the MLEs of these GLMs in the three most important cases: Gaussian response with identity link, Poisson response with log link, and negative binomial response with log link (see Appendix \ref{sec:statistical_accelerations}; the latter formula is asymptotically exact).
+Consequently, we do not need to run the relatively slow IRLS procedure to carry out the M step of the reduced GLM-EIV algorithm.
+Overall, the proposed method for obtaining the full set of pilot parameter estimates requires fitting only two GLMs (via IRLS).
+
+\subsection{Computing}\label{sec:computing}
+
+We developed a computational infrastructure to apply GLM-EIV to large-scale, single-cell CRISPR screen data.
+The infrastructure leverages Nextflow, a programming language that facilitates building data-intensive pipelines \cite{DITommaso2017}, and ondisc, an R package that we developed (in a separate project) to facilitate large-scale computing on single-cell data.
+Nextflow and ondisc together enable the construction of highly portable single-cell pipelines: one can analyze data \textit{out-of-memory} on a laptop or in a \textit{distributed} fashion across tens or hundreds of nodes on a cloud (e.g., Microsoft Azure, Google Cloud) or high-performance cluster.
+
+\begin{algorithm}
+	\caption{Applying GLM-EIV at scale.}\label{algo:at_scale}
+	\begin{algorithmic}
+		\State $G \gets \{\textrm{gene}_1, \dots, \textrm{gene}_{d_g}\}; P \gets \{\textrm{perturbation}_1, \dots, \textrm{perturbation}_{d_p}\}$
+
+		\For{gene $\in G$}
+		\State Run precomputation (Algorithm \ref{algo:pilot_estimates_1}) on gene; save $\hat{f}^m$, $[\beta^m_0]^\textrm{pilot}$ and $[\gamma^T_m]^\textrm{pilot}$.
+		\EndFor
+		\For{perturbation $\in P$}
+		\State Run precomputation  (Algorithm \ref{algo:pilot_estimates_1}) on perturbation; save $\hat{f}^g$, $[\beta^g_0]^\textrm{pilot}$ and $[\gamma^T_g]^\textrm{pilot}$.
+		\EndFor
+		\For{(gene, perturbation) $\in G \times P$}
+		\State Load $\hat{f}^m, \hat{f}^g,$ $[\beta^m_0]^\textrm{pilot}$ $[\gamma^T_m]^\textrm{pilot}$, $[\beta^g_0]^\textrm{pilot}$ and $[\gamma^T_g]^\textrm{pilot}$.
+		\State Compute $[\beta^m_1]^\textrm{pilot}, [\beta^g_1]^\textrm{pilot}, \pi^\textrm{pilot}$ by fitting a reduced GLM-EIV (Algorithm \ref{algo:pilot_estimates_2}).
+		\State Run GLM-EIV using the pilot parameters (Algorithm \ref{algo:em_full}).
+		\EndFor
+	\end{algorithmic}
+\end{algorithm}
+
+Leveraging these technologies, we develop a scalable and efficient pipeline for GLM-EIV (Algorithm \ref{algo:at_scale}).
+First, we run a round of ``precomputations'' on all $d_g$ genes and $d_p$ perturbations.
+The precomputations involve regressing the gene expressions (or gRNA counts) onto the technical factors, thereby ``factoring out'' Algorithm \ref{algo:pilot_estimates_1}.
+Next, we run differential expression analyses on the full set of gene-perturbation pairs; for a given pair, this amounts to obtaining the complete set of pilot parameters (by running a reduced GLM-EIV), fitting the GLM-EIV model (Algorithm \ref{algo:em_full}), and performing inference.
+The three loops in Algorithm \ref{algo:at_scale} are embarassingly parallel and therefore can be massively parallelized.
+
+% \subsection{Zero-inflated model}
+%The GLM-EIV model assumes that unperturbed cells exhibit a nonzero ``background read'' gRNA count distribution due to errors in the sequencing and alignment processes.
+This assumption very likely is appropriate for most current single-cell CRISPR screen datasets \cite{Replogle2020}.
+However, background contamination may be negligible for some datasets, in which case the gRNA counts of unperturbed cells should be modeled as point masses at zero.
+We introduce such an extension -- which we call the ``zero-inflated GLM-EIV model'' -- in Appendix \ref{sec:zero_inf_model}.
+We additionally derive efficient methods for estimation and inference in the zero-inflated model.
+Given the ubiquity of background contamination in current single-cell CRISPR screen datasets, we focus our attention on the standard GLM-EIV model for the remainder of the main text.
+\section{Simulation studies}\label{sec:simulation}
+
+We conducted a simulation study to compare the empirical performance of GLM-EIV to that of the thresholding method.
+We generated data on $n = 150,000$ cells from the GLM-EIV model using realistic parameter values, setting the the target of inference $\beta^m_1$ to $\log(0.25)$ and the probability of perturbation $\pi$ to $0.02$.
+$\beta^m_1 = \log(0.25)$ represents a decrease in gene expression by a factor of $4$, which is a fairly large effect size on the order of what we might observe for a positive control pair.
+We included ``sequencing batch'' (modeled as a Bernoulli-distributed variable) as a covariate and sequencing depth (modeled as a Poisson-distributed variable) as an offset.
+We varied the log-fold change in gRNA expression, $\beta^g_1,$ over a grid on the interval $[\log(1), \log(4)];$ $\beta^g_1$ controls problem difficulty, with higher values corresponding to easier problems.
+Finally, we generated the gene expression and gRNA count data from two response distributions: Poisson and negative binomial (size parameter fixed at $\theta = 20$ for the latter).
+For each parameter setting (defined by a $\beta^g_1$-distribution pair), we synthesized $n_\textrm{sim} = 500$ i.i.d.\ datasets.
+Section \ref{sec:extra_sims} presents additional simulation results on Gaussian response distributions.
+
+We applied three methods to the simulated data: ``vanilla'' GLM-EIV, accelerated GLM-EIV, and thresholded regression.
+We used the Bayes-optimal decision boundary for classification as the threshold for the thresholding method.
+We ran all methods on the negative binomial data twice: once treating the size parameter $\theta$ as a known constant and once treating $\theta$ as unknown.
+In the latter case we used the \texttt{glm.nb} function from the \texttt{MASS} package to estimate $\theta$ before applying the methods \cite{Ripley2013}.
+We display the results of the simulation study in Figure \ref{main_text_sim}.
+Columns correspond to distributions (i.e., Poisson, NB with known $\theta$, and NB with unknown $\theta$), rows correspond to performance metrics (i.e., bias, mean squared error, CI coverage rate (nominal rate $95\%$), CI width, and method execution time).
+The problem difficulty parameter $\beta^g_1$ is plotted on the horizontal axis, and the methods are depicted in different colors (GLM-EIV masked by accelerated GLM-EIV in several panels).
+
+First, we observed that GLM-EIV dominated thresholded regression on all statistical metrics: GLM-EIV exhibited lower bias (row 1) and mean squared error (row 2) than thresholded regression; additionally, GLM-EIV had superior confidence interval coverage (row 3) despite having produced generally narrower confidence intervals (row 4).
+Intuitively, GLM-EIV outperformed the thresholding method because (i) GLM-EIV leveraged information from \textit{both} modalities (rather than the gRNA modality alone) to assign perturbation identities to cells, and (ii) GLM-EIV produced soft rather than hard assignments, capturing the inherent uncertainty in whether a perturbation occurred.
+We additionally found that accelerated GLM-EIV performed as well as vanilla GLM-EIV on all statistical metrics (rows 1-4) despite having substantially lower computational cost (bottom row).
+In fact, the execution time of accelerated GLM-EIV was almost within an order of magnitude of that of the thresholding method (bottom row).
+
+Interestingly, thresholded regression exhibited better confidence interval coverage under estimated $\theta$ than under known $\theta$ (row 3).
+Estimating $\theta$ leads to slight inflation bias (i.e., overestimating the true effect size), whereas, as we showed previously, thresholding leads to attenuation bias (i.e., underestimating the true effect size).
+These phenomena partially cancel, yielding less biased estimates.
+GLM-EIV exhibited worse performance under unknown $\theta$ than known $\theta$, likely due to poor $\theta$ estimation.
+We note that GLM-EIV and the thresholding method in principal are compatible with \textit{any} $\theta$ estimation procedure, including those based on more sophisticated techniques, such as regularization \cite{Hafemeister2019}.
+We defer rigorous investigation of the impact of different $\theta$ estimation strategies on these methods to future work.
+
+\section{Data analysis}
+
+Leveraging our computational infrastructure (Section \ref{sec:computing}), we applied GLM-EIV and the thresholding method to analyze the entire Gasperini and Xie datasets.
+(We report only the most important aspects of the analysis and results in the main text; full details are available in Appendix \ref{sec:data_analysis_details}.) We set the threshold to the approximate Bayes-optimal decision boundary, as our theoretical analyses and simulation studies revealed that the Bayes-optimal decision boundary is a good threshold when the gRNA count distribution is well-separated.
+Operating under the assumption that the effect of the perturbation on gRNA expression is similar across pairs, we leveraged the fitted GLM-EIV models to approximate the Bayes boundary in the following way: we (i) sampled several hundred gene-perturbation pairs, (ii) extracted the fitted values $\hat{\beta}_g$ and $\hat{\pi}$ from the GLM-EIV models fitted to these pairs, (iii) computed the median $\overline{\hat{\beta}_g}$ and $\overline{\hat{\pi}}$ across the $\hat{\beta}_g$s and $\hat{\pi}$s, and (iv) used $\overline{\hat{\beta}_g}$ and $\overline{\hat{\pi}}$ to estimate a dataset-wide Bayes-optimal decision boundary.
+We repeated this procedure on both datasets, yielding a threshold of $3$ for Gasperini and $7$ for Xie.
+These thresholds were close to the original thresholds, which were selected in a more heuristic way.
+
+We compared GLM-EIV to thresholded regression on the real data, focusing specifically on the negative control pairs (i.e., gene-perturbation pairs that, by design, are expected to exhibit a fold change of $1$, or no association).
+We found that GLM-EIV and the thresholding method produced similar results (Figure \ref{fig:real_data}a-b): estimates, CI coverage rates, and CI widths were concordant.
+The estimated effect of the perturbation on gene expression $\exp(\hat{\beta}_1^g)$ was unexpectedly large: the 95\% CI for this parameter was $[4306, 5186]$ and $[300, 316]$ on the Gasperini and Xie data, respectively.
+We reasoned that the datasets lay in an ``easy'' region of the parameter space, making thresholding a tenable strategy (provided the threshold is selected well).
+However, this was not obvious \textit{a priori} and may not be the case for other datasets.
+We note that GLM-EIV produced outlier estimates (likely due to non-global EM convergence) on a small ($< 2.5 \%$ on Gasperini, $<0.05\%$ on Xie) number of pairs consisting of a handful of genes (not plotted).
+
+We artificially increased the difficulty of the perturbation assignment problem by generating partially-synthetic datasets.
+First, for a given pair, we sampled gRNA counts directly from the fitted GLM-EIV model.
+Next, to simulate elevated background contamination, we sampled gRNA counts from a slightly modified version of the fitted model in which we increased the mean gRNA expression of \textit{unperturbed} while holding constant the mean gRNA expression of \textit{perturbed} cells.
+We defined a parameter called ``excess background contamination'' (normed to take values in $[0,1]$) to quantify the relative distance between the unperturbed and perturbed gRNA count distributions.
+We held fixed the real-data gene expressions, library sizes, covariates, and fitted perturbation probabilities in all settings.
+
+We generated partially-synthetic data in the above manner for each of the 322 positive control pairs in the Gasperini dataset, varying excess background contamination over the interval $[0,0.4].$ We then applied GLM-EIV and the thresholding method to analyze the data.
+We present results on two example pairs (the pair containing gene \textit{LRIF1} and the pair containing gene \textit{NDUFA2}) in Figures \ref{fig:real_data}c-d.
+We observed that the estimate produced by the methods on the raw data (depicted as a horizontal black line) coincided almost exactly with the estimate produced by the methods on the partially-synthetic data generated by setting excess background contamination to zero (This result replicated across nearly all pairs; average relative difference $0.003$.) We additionally observed that as excess background contamination increased, the performance of thresholded regression degraded considerably while that of GLM-EIV remained stable.
+
+We generalized the above analysis to the entire set of positive control pairs.
+First, for each pair we computed the ``relative estimate change'' (REC) as a function of excess background contamination, defined as the magnitude of the relative difference between the estimate at a given level of excess contamination and zero excess contamination (Figure \ref{fig:real_data}d).
+Next, we computed the median REC across all positive control pairs (Figure \ref{fig:real_data}e; bands indicate pointwise 95\% CIs for the population median REC).
+As excess background contamination increased, thresholded regression exhibited severe attenuation bias (as reflected by large median REC values); GLM-EIV, by contrast, remained mostly stable.
+Finally, taking the estimate obtained on the raw data as ``ground truth,'' we computed CI coverage across pairs as a function of excess contamination.
+GLM-EIV exhibited significantly higher CI coverage than thresholded regression as the data became increasingly contaminated (Figure \ref{fig:real_data}f; bands indicate $95\%$ pointwise CIs for population coverage).
+
+\section{Discussion}
+
+In this work we introduced GLM-EIV (``GLM-based errors in variables''), a new method for single-cell CRISPR screen analysis.
+GLM-EIV extends the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed and (optionally) impacted by the same set of confounding variables.
+These extensions enable GLM-EIV to resolve novel analysis challenges posed by single-cell CRISPR screens.
+We demonstrated through simulation studies, real data analyses, and theory that GLM-EIV outperforms thresholded regression by a considerable margin in high background contamination settings.
+GLM-EIV intuitively achieves this performance gain by leveraging information from \textit{both} modalities (rather than the gRNA modality alone) to assign perturbation identities to cells.
+In low background contamination settings GLM-EIV and thresholded regression work best in conjunction, with GLM-EIV providing a statistically principled and empirically effective procedure for selecting the threshold.
+Figure \ref{fig:dec_tree} summarizes how we anticipate GLM-EIV being used in practice.
+
+To our knowledge this is the first single-cell CRISPR screen paper oriented toward a statistical audience.
+We hope that this work helps to introduce the broader statistics community to an emerging class functional genomics assays that likely will exert a major impact on biology research in the coming years \cite{Przybyla2021}.
+Additionally, this is the first work to leverage the ondisc-Nextflow-cloud/HPC technology stack, a tightly-integrated, user-friendly, and powerful set of computational tools for large-scale single-cell analysis that we expect to be of interest to other researchers.
+
+We anticipate that GLM-EIV could be applied to other types of single-cell CRISPR screen and multimodal single-cell data.
+For example, we likely (with some effort) could extend GLM-EIV to ``low-multiplicity of infection'' screens \cite{Schraivogel2020} in which each cell receives one or two perturbations rather than dozens (as is the case in ``high multiplicity screens,'' studied in this work).
+We also likely could apply GLM-EIV to analyze multimodal single-cell chromatin accessibility assays.
+A question of interest in such experiments is whether chromatin state (i.e., closed or open) is associated with the expression of a gene (or abundance of a protein) \cite{Mimitou2021}.
+We do not directly observe the chromatin state of a cell; instead, we observe tagged DNA fragments that serve as proxies for whether a given region of chromatin is open or closed.
+These tagged DNA fragments come in the form of discrete counts; therefore, GLM-EIV might be applied in such experiments to aid in the selection of thresholds or to analyze whole datasets.
+
+\section{Acknowledgments and code availability}
+We thank Xuran Wang helping to process the Xie dataset.
+We also thank Songcheng Dai for helping to deploy the GLM-EIV pipeline on Azure.
+This work used the Extreme Science and Engineering Discovery Environment (XSEDE; NSF grant ACI-1548562) and the Bridges-2 system (NSF grant ACI-1928147) at the Pittsburgh Supercomputing Center.
+Code to replicate the analyses is available at \texttt{/timothy-barry/glmeiv-manuscript}.
+
+\printbibliography
+
+\clearpage
+\section*{Figures}
+% 1. Challenges
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=1\linewidth]{../../figures/analysis_challenges/plot.pdf}
+	\caption{\textbf{Experimental design and analysis challenges}: \textbf{a,} Experimental design.
+For a given perturbation (e.g., the perturbation indicated in blue), we partition the cells into two groups: perturbed and unperturbed.
+Next, for a given gene, we conduct a differential expression analysis across the two groups, yielding an estimate of the impact of the given perturbation on the given gene.
+\textbf{b,} DAG representing all variables in the system.
+The perturbation (latent) impacts both gene expression and gRNA expression; technical factors act as confounders, also impacting gene and gRNA expression.
+The target of estimation is the effect of the perturbation on gene expression.
+\textbf{c,} Schematic illustrating the ``background read'' phenomenon.
+Due to errors in the sequencing and alignment processes, unperturbed cells exhibit a nonzero gRNA count distribution (bottom).
+The target of estimation is the change in mean gene expression in response to the perturbation (top).
+\textbf{d}, Example data on four cells for a given perturbation-gene pair.
+Note that (i) the perturbation is unobserved, and (ii) the gene and gRNA data are discrete counts.}
+	\label{analysis_challenges}
+\end{figure}
+
+\clearpage
+% 2. Emprical challenges
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=1\linewidth]{../../figures/thresholding_empirical/plot}
+	\caption{\textbf{Empirical challenges of thresholded regression.} \textbf{a-b,} Fold change estimates produced by threshold = 1 versus threshold = 5 (a) and threshold = 20 versus threshold = 5 (b).
+The selected threshold substantially impacts the results.
+\textbf{c-d,} $p$-values (c) and CI widths (d) produced by threshold = 20 versus threshold = 5.
+The latter threshold yields more confident estimates.
+\textbf{e-f}, Empirical distribution of randomly-selected gRNA from Gasperini (e) and Xie (f) data (0 counts not shown).
+The gRNA data do not appear to imply an obvious threshold selection strategy.}
+	\label{thresholding_empirical}
+\end{figure}
+
+\clearpage
+\begin{figure}[h!]
+\centering
+\includegraphics[width=0.7\linewidth]{../../figures/thresholding_theoretical/plot}
+\caption{\textbf{Thresholding method bias-variance decomposition.} Bias decreases and variance increases as the threshold tends to infinity.
+$\beta^g_1 = 1, \beta^m_1 = 1,$ and $\pi = 0.1$ in this plot.}
+\label{thresholding_theoretical}
+\end{figure}
+
+\clearpage
+% Theory
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=0.9\linewidth]{../../figures/main_text_sim/plot}
+	\caption{\textbf{Simulation study}.
+Columns correspond to distributions (Poisson, NB with known $\theta$, NB with estimated $\theta$), and rows correspond to metrics (bias, MSE, coverage, CI width, and time).
+Methods are shown in different colors; GLM-EIV (red) is masked by accelerated GLM-EIV (blue) in several panels.
+GLM-EIV demonstrated superior statistical performance to the thresholding method on all metrics (rows 1-4).
+Accelerated GLM-EIV had substantially lower computational cost than ``vanilla'' GLM-EIV (bottom row) despite demonstrating identical statistical performance (rows 1-4).}
+	\label{main_text_sim}
+\end{figure}
+
+\clearpage
+\begin{figure}
+	\centering
+	\includegraphics[width=0.9\linewidth]{../../figures/data_analysis/annotated_plot_crop}
+	\caption{\textbf{Data analysis}.
+\textbf{a-b}, Estimates for fold change produced by GLM-EIV and thresholded regression on Gasperini (\textbf{a}) and Xie (\textbf{b}) negative control pairs.
+\textbf{c-d}, Estimates produced by GLM-EIV and thresholded regression on two positive control pairs -- \textit{LRIF1} (\textbf{a}) and \textit{NDUFA2} (\textbf{b}) -- plotted as a function of excess background contamination.
+Grey bands, 95\% CIs for the target of inference outputted by the methods.
+\textbf{e-f}, Median relative estimate change (REC; \textbf{e}) and confidence interval coverage rate (\textbf{f}) across \textit{all} 322 positive control pairs, plotted as a function of excess background contamination.
+Colored bands, pointwise 95\% CIs for population quantities.
+Panels (\textbf{c-f}) together illustrate that GLM-EIV demonstrated greater stability than thresholded regression as background contamination increased.}\label{fig:real_data}
+\end{figure}
+
+\clearpage
+% Dec tree
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=0.9\linewidth]{../../figures/dec_tree/dec_tree_crop}
+	\caption{\textbf{Use of GLM-EIV in practice}.
+The decision tree above illustrates how we anticipate GLM-EIV could be used in practice.
+First, apply GLM-EIV to a set of randomly-sampled gene-perturbation pairs to assess background contamination level (positive control pairs work best for this purpose).
+If GLM-EIV indicates that background contamination is high (e.g., $\exp(\beta^g_1) \lessapprox $ 10), apply GLM-EIV to analyze the entire dataset; otherwise, approximate the Bayes-optimal decision boundary using the fitted GLM-EIV models.
+Next, apply a thresholding method (e.g., SCEPTRE or thresholded negative binomial regression) to analyze the data, setting the threshold to the estimated Bayes-optimal decision boundary.}\label{fig:dec_tree}
+\end{figure}
+\clearpage
+\begin{appendices}
+\begin{refsection}
+\section{Theoretical details for thresholding estimator}\label{sec:appendix_theory}
+
+\subsection{Proposition statements}
+This subsection contains detailed proposition statements for the informal claims made in Section \ref{sec:thresholding_theory}.
+For simplicity we derive results in the case $\pi = 1/2$.
+In reality $\pi$ is much closer to zero; however, setting $\pi$ to $1/2$ simplifies the proofs substantially and enables us to recover the most interesting and important statistical phenomena (e.g., attenuation bias, monotonic impact of $\beta^g_1$ on problem difficulty, etc.).
+We state state five propositions labeled \ref{prop:att_bias} -- \ref{prop:comparison} corresponding to the informal claims made in Section \ref{sec:thresholding_theory}; these propositions visually are depicted in Figure \ref{fig:bias_plot}.
+
+First, the thresholding method incurs strict attenuation bias (i.e., it \textit{under}estimates the true effect size) for all choices of the threshold and over all possible values of the model parameters:
+\begin{proposition}\label{prop:att_bias} Fix $\pi = 1/2$. For all $(\beta^g_1, c, \beta^g_0) \in \R^3$, the asymptotic relative bias is positive, i.e. 
+	$$b(\beta^g_1, 1/2, c, \beta^g_0) > 0.$$
+\end{proposition}
+
+Next, the asymptotic relative bias $b$ decreases monotonically in $\beta^g_1:$
+\begin{proposition}\label{prop:monotonic} Fix $\pi = 1/2$. The asymptotic relative bias $b$ decreases monotonically in $\beta_1^g$, i.e.
+		$$\frac{\partial b}{\partial(\beta^g_1)}\left(\beta^g_1, 1/2, c, \beta^g_0\right) \leq 0.$$
+\end{proposition}
+Let $c_\textrm{bayes}$ denote the Bayes-optimal decision boundary for classifying cells as perturbed or unperturbed, i.e.
+$c_\textrm{bayes} = (1/2)(\beta^g_0 + \beta^g_1)$ for $\pi = 1/2$.
+We have that $c_\textrm{bayes}$ is a critical value of the bias function.
+\begin{proposition}\label{prop:bayes_opt}
+		For $\pi = 1/2$ and given $(\beta^g_1, \beta^g_0) \in \R^2$, the Bayes-optimal decision boundary $c_\textrm{bayes}$ is a critical value of the bias function $b$, i.e.
+		$$ \frac{\partial b}{\partial c}\left(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0\right) = 0.$$
+\end{proposition} Furthermore, as the threshold tends to infinity, the asymptotic relative bias $b$ tends to $\pi$.
+\begin{proposition}\label{prop:c_limit}
+	Assume without loss of generality that $\beta^g_1 > 0$. As the threshold $c$ tends to infinity, the asymptotic relative bias $b$ tends to $\pi,$ i.e.
+	$$ \lim_{ c \to \infty } b(\beta^g_1, \pi, c, \beta^g_0) = \pi.$$
+\end{proposition}
+As a corollary, when $\pi = 1/2$, asymptotic relative bias tends to $1/2$ as $c$ tends to infinity.
+Finally, we compare two threshold selection strategies head-to-head: setting the threshold to an arbitrarily large number, and setting the threshold to the Bayes-optimal decision boundary.
+\begin{proposition}\label{prop:comparison} Assume without loss of generality that $\beta^g_1 > 0$.
+For $\beta^g_1 \in [0, 2\Phi^{-1}(3/4))$, we have that $$b(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0) > b(\beta^g_1, 1/2, \infty, \beta^g_0).$$ For $\beta^g_1 = 2\Phi^{-1}(3/4)$, we have that $$b(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0) = b(\beta^g_1, 1/2, \infty, \beta^g_0).$$ Finally, for $\beta^g_1 \in (2\Phi^{-1}(3/4), \infty)$, we have that $$b(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0) < b(\beta^g_1, 1/2, \infty, \beta^g_0).$$
+\end{proposition}
+In other words, setting the threshold to a large number yields a smaller bias when $\beta^g_1$ is small (i.e., $\beta^g_1 < 2\Phi^{-1}(3/4) \approx 1.35$; Figure \ref{thresholding_theoretical}a, left); setting the threshold to the Bayes-optimal decision boundary yields a smaller bias when $\beta^g_1$ is large (i.e., $\beta^g_1 > 2\Phi^{-1}(3/4)$; Figure \ref{thresholding_theoretical}a, right); and the two approaches coincide when $\beta^g_1$ is intermediate (i.e., $\beta^g_1 = 2\Phi^{-1}(3/4)$; Figure \ref{thresholding_theoretical}a, middle).
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=1\linewidth]{../../figures/thresholding_theoretical/app_plot}
+	\caption{\textbf{Bias as a function of threshold}.
+This figure visually depicts Propositions \ref{prop:att_bias}-\ref{prop:comparison}, which were stated informally in the main text.
+Asymptotic relative bias is plotted on the vertical axis, and the threshold is plotted on the horizontal axis.
+Panels correspond to different values of $\beta^g_1$.
+Vertical blue lines indicate the Bayes-optimal decision boundary.
+Observe that (a) bias is strictly nonzero (proposition \ref{prop:att_bias}); (b) bias decreases monotonically in $\beta^g_1$ (Proposition \ref{prop:monotonic}); (c) the Bayes-optimal decision boundary is a critical value of the bias function (Proposition \ref{prop:bayes_opt}), in some cases a maximum and in other cases a minumum; (d) as the threshold tends to infinity, the bias converges to $1/2$ (Proposition \ref{prop:c_limit}).; and (e) when $\beta^g_1 < 1.35$, an arbitrarily large number yields a smaller bias; by contrast, when $\beta^g_1 > 1.35$, the Bayes-optimal decision boundary yields a smaller bias (Proposition \ref{prop:comparison}).
+	%Across all panels, $\beta^g_0 = 0$ and $\pi = 1/2$.
+When $\beta^g_1$ is small, setting the threshold to an arbitrarily large number yields a smaller bias than setting the threshold to the Bayes-optimal decision boundary; when $\beta^g_1$ is large, the opposite is true.
+}
+	\label{fig:bias_plot}
+\end{figure}
+
+\subsection{Organization} The following subsections prove all propositions.
+Section (\ref{sec:notation}) introduces some notation.
+Section (\ref{sec:convergence}) establishes almost sure convergence of the thresholding estimator in the model (\ref{theoretical_model}), proving Proposition \ref{prop:convergence}.
+Section (\ref{sec:simplication}) simplifies the expression for the attenuation function $\gamma$, and section (\ref{sec:derivatives})  computes derivatives of $\gamma$ to be used throughout the proofs.
+Section (\ref{sec:c_limit}) establishes the limit in $c$ of $\gamma$, proving Proposition \ref{prop:c_limit}.
+Section (\ref{sec:bayes_opt}) establishes that the Bayes-optimal decision boundary is a critical value of $\gamma$, proving Proposition \ref{prop:bayes_opt}, and section (\ref{sec:comparison}) compares the competing threshold selection strategies head-to-head, proving Proposition \ref{prop:comparison}.
+Section (\ref{sec:monotone}) demonstrates that $\gamma$ is monotone in $\beta^g_1$, proving Proposition \ref{prop:monotonic}, and Section (\ref{sec:att_bias}) establishes attenuation bias of the thresholding estimator, proving Proposition \ref{prop:att_bias}.
+Finally, Section (\ref{sec:bv_decomp}) derives the bias-variance decomposition of the thresholding estimator in the no-intercept version of (\ref{theoretical_model}), proving Proposition \ref{prop:bv_decomp}.
+
+% Labels of propositions and theorems
+% propositions:
+% prop:convergence
+% prop:att_bias
+% prop:bayes_opt
+% prop:c_limit_half
+% prop:comparison
+% prop:monotonic
+% prop:c_limit
+% prop:bv_decomp
+
+% sections
+% sec:notation
+% sec:convergence
+% sec:simplication
+% sec:derivatives
+% sec:c_limit
+% sec:bayes_opt
+% sec:comparison
+% sec:beta_lim
+% sec:monotone
+% sec:att_bias
+% sec:bv_decomp
+\subsection{Notation}\label{sec:notation}
+ All notation introduced in this subsection (i.e., \ref{sec:notation}) pertains to the Gaussian model with intercepts (\ref{theoretical_model}).
+Recall that the attenuation function $\gamma: \R^4 \to \R$ is defined by
+$$ \gamma(\beta^g_1, c, \pi, \beta^g_0) = \frac{\pi(\omega - \E[\hat{p}_i])}{ \E[\hat{p}_i](1 -\E[\hat{p}_i])},$$ where $$\E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi; \quad \omega = \Phi\left(\beta_1^g + \beta_0^g - c \right); \quad \zeta = \Phi\left( \beta^g_0 - c \right).$$ Additionally, recall that the asymptotic relative bias function $b: \R^4 \to \R$ is
+$$ b(\beta^g_1, c, \pi, \beta^g_0) = 1 - \gamma(\beta^g_1, c, \pi, \beta^g_0).$$ Next, we define the functions $g$ and $h: \R^4 \to \R$ by
+\begin{equation}\label{def_g}
+g(\beta^g_1, c, \pi, \beta^g_0) = (1-\pi)\left( \Phi(\beta_0^g + \beta_1^g - c)\right) - (1-\pi)\left(\Phi(\beta_0^g - c)\right)\end{equation}
+ and
+\begin{multline}\label{def_h}
+h(\beta^g_1, c, \pi, \beta^g_0) = \left[(1-\pi)\left( \Phi(\beta_0^g - c)\right) + \pi\left(\Phi(\beta^g_0 + \beta^g_1 - c) \right) \right] \cdot \\ \left[(1-\pi)\left( \Phi(c - \beta^g_0) \right) + \pi\left(\Phi(c - \beta_0^g - \beta_1^g) \right) \right].
+\end{multline}
+We use $f:\R \to \R$ to denote the $N(0,1)$ density, and we denote the right-tail probability probability of $f$ by $\bar{\Phi}$, i.e.,
+$$\bar{\Phi}(x) = \int_{x}^{\infty} f = \Phi(-x).$$
+
+The parameter $\beta^g_0$ is a given, fixed constant throughout the proofs.
+Therefore, to minimize notation, we typically use $\gamma(\beta^g_1, c, \pi)$ (resp., $b(\beta^g_1, c, \pi),$ $g(\beta^g_1, c, \pi),$ $h(\beta^g_1, c, \pi)$) to refer to the function $\gamma$ (resp., $b, g, h$) evaluated at $(\beta^g_1, c, \pi, \beta^g_0)$.
+Finally, for a given function $r: \R^{p} \to \R$, point $x \in \R^p$, and index $i \in \{1, \dots, p\}$, we use the symbol $D_i r(x)$ to refer to the derivative of the $i$th component of $r$ evaluated at $x$ (\textit{sensu} \cite{fitzpatrick2009}).
+For example, $D_1 \gamma(\beta^g_1, c, 1/2)$ is the derivative of the first component of $\gamma$ (the component corresponding to $\beta^g_1$) evaluated at $(\beta^g_1, c, 1/2)$.
+Likewise,  $D_2g(\beta^g_1, c, \pi)$ is the derivative of the second component of $g$ (the component corresponding to $c$) evaluated at $(\beta^g_1, c, \pi).$
+
+\subsection{Almost sure limit of $\hat{\beta}^m_1$}\label{sec:convergence}
+
+We derive the limit in probability of $\hat{\beta}^m_1$ for the Gaussian model with intercepts (\ref{theoretical_model}).
+Dividing by $n$ in (\ref{thresh_est_intercepts}), we can express $\hat{\beta}^m_1$ as
+$$ \hat{\beta}^m_1 = \frac{ \frac{1}{n} \sum_{i=1}^n ( \hat{p}_i - \overline{\hat{p}_i})(m_i - \overline{m})}{ \frac{1}{n} \sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})}.$$ By weak LLN,
+$$ \hat{\beta}^m_1 \xrightarrow{P} \frac{\textrm{Cov}(\hat{p}_i, m_i)}{\V\left(\hat{p}_i\right)}.$$ To compute this quantity, we first compute several simpler quantities:
+\begin{itemize}
+\item[1.] Expectation of $m_i$: $\E[m_i] = \beta^m_0 + \beta^m_1\pi.$
+\item[2.] Expectation of $\hat{p}_i$: \begin{multline*}
+\E[\hat{p}_i] = \P\left[\hat{p}_i = 1\right] = \P\left[\beta^g_0 + \beta^g_1 p_i + \tau_i \geq c \right] = \\ \textrm{(By LOTP) } \P\left[ \beta^g_0 + \tau_i \geq c \right]\P\left[p_i = 0\right] + \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \P[p_i = 1] \\ = \P\left[ \tau_i \geq c - \beta^g_0\right](1- \pi) + \P\left[ \tau_i \geq c - \beta^g_1 - \beta^g_0 \right](\pi) \\ =  \left(\bar{\Phi}(c - \beta^g_0) \right) (1 - \pi) + \left( \bar{\Phi}(c - \beta^g_1 - \beta^g_0) \right)(\pi) = \\  \Phi(\beta^g_0 - c) (1-\pi) + \Phi(\beta^g_1 + \beta^g_0 - c) \pi = \zeta(1-\pi) + \omega \pi.
+\end{multline*}
+\item[3.] Expectation of $\hat{p}_i p_i$: 
+$\E\left[ \hat{p}_i p_i \right] = \E\left[\hat{p}_i | p_i = 1 \right] \P\left[ p_i =1 \right] = \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \pi = \omega \pi.$
+\item[4.] Expectation of $\hat{p}_i m_i$:
+\begin{multline*}
+\E\left[\hat{p}_i m_i\right] = \E[\hat{p}_i (\beta^m_0 + \beta^m_1 p_i + \ep_i)] = \beta^m_0 \E\left[\hat{p}_i\right] + \beta^m_1 \E\left[\hat{p}_i p_i\right] + \E[\hat{p}_i \ep_i] \\ = \beta^m_0 \E[\hat{p}_i] + \beta^m_1 \omega \pi + \E[\hat{p}_i] \E[\ep_i] = \beta^m_0 \E[\hat{p}_i] + \beta^m_1 \omega \pi.
+\end{multline*}
+\item[5.] Variance of $\hat{p}_i$: Because $\hat{p}_i$ is binary, we have that $\V[\hat{p}_i] = \E[\hat{p}_i]\left(1 - \E[\hat{p}_i]\right) .$
+\item[6.] Covariance of $\hat{p}_i, m_i$:
+\begin{multline*}
+\textrm{Cov}\left(\hat{p}_i, m_i\right) = \E\left[\hat{p}_i m_i\right] - \E[\hat{p}_i] \E[m_i] = \beta^m_0 \E[\hat{p}_i] + \beta^m_1 \omega \pi - \E[\hat{p}_i]( \beta^m_0 + \beta^m_1 \pi)\\ = \beta^m_1 \omega \pi - \E[\hat{p}_i] \beta_1^m \pi = \beta^m_1 \pi \left( \omega - \E[\hat{p}_i]\right).
+\end{multline*}
+Combining these expressions, we have that
+$$ \hat{\beta}^m_1 \xrightarrow{P} \frac{\beta^m_1 \pi (\omega - \E[\hat{p}_i])}{\E[\hat{p}_i](1 - \E[\hat{p}_i])} = \beta^m_1 \gamma(\beta^g_1, c, \pi).$$
+\end{itemize}
+
+\subsection{Re-expressing $\gamma$ in a simpler form}\label{sec:simplication}
+We rewrite the attenuation fraction $\gamma$ in a way that makes it more amenable to theoretical analysis. We leverage the fact that $f$ integrates to unity and is even. We have that
+\begin{equation}\label{thm:gamma_expression_1} \E\left[\hat{p}_i\right] = (1 - \pi) \bar{\Phi}(c - \beta_0^g) + \pi \bar{\Phi}(c - \beta^g_0 - \beta^g_1) = (1 - \pi) \Phi(\beta_0^g - c) + \pi\Phi(\beta^g_0 + \beta^g_1 - c), \end{equation}
+ and so \begin{multline}\label{thm:gamma_expression_2} 1 - \E\left[\hat{p}_i\right] = (1 - \pi) + \pi - \E[\hat{p}_i]  = (1-\pi) \left(1 - \bar{\Phi}(c - \beta_0^g)\right)  + \pi \left(1 - \bar{\Phi}(c - \beta^g_0 - \beta^g_1) \right) \\ = (1 - \pi)\Phi(c - \beta^g_0) + \pi \Phi(c - \beta_0^g - \beta_1^g).
+\end{multline}
+Next,
+\begin{equation}\label{thm:gamma_expression_3}
+\omega = \Phi(\beta^g_1 + \beta^g_0 - c),\end{equation} and so
+\begin{multline}\label{thm:gamma_expression_4}
+\omega - \E[\hat{p}_i] = \Phi(\beta^g_1 + \beta^g_0 - c) - (1-\pi)\Phi(\beta^g_0 - c) - \pi \Phi(\beta^g_0 + \beta^g_1 - c)  \\ (1-\pi)\Phi(\beta^g_1 + \beta^g_0 - c)  - (1-\pi)\Phi(\beta^g_0 - c).
+\end{multline}
+Combining (\ref{thm:gamma_expression_1}, \ref{thm:gamma_expression_2}, \ref{thm:gamma_expression_3}, \ref{thm:gamma_expression_4}), we find that
+\begin{multline}\label{gamma_alternative}
+\gamma(\beta^g_1, c, \pi) = \frac{\pi(\omega - \E[\hat{p}_i])}{\E[\hat{p}_i](1 - \E[\hat{p}_i])} \\ = \frac{\pi \left[(1 - \pi) \Phi(\beta_0^g + \beta_1^g - c) - (1 - \pi) \Phi(\beta_0^g - c)\right]}{\left[(1-\pi)\Phi(\beta_0^g - c) + \pi \Phi(\beta^g_0 + \beta^g_1 - c) \right] \left[(1 - \pi) \Phi(c - \beta^g_0) + \pi\Phi(c - \beta_0^g - \beta_1^g) \right]}.
+\end{multline}
+As a corollary, when $\pi = 1/2$,
+\begin{equation}\label{gamma_alternative_pi_half}
+\gamma(\beta^g_1, c, 1/2)  \\ = \frac{\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c) }{\left[\Phi(\beta_0^g - c) +\Phi(\beta^g_0 + \beta^g_1 - c)\right] \left[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g) \right]}.
+\end{equation}
+Recalling the definitions of $g$ (\ref{def_g}) and $h$ (\ref{def_h}), we can write $\gamma$ as
+$$ \gamma(\beta^g_1, c, \pi) = \frac{\pi g(\beta^g_1, c, \pi)}{h(\beta^g_1, c,\pi)}.$$
+The special case (\ref{gamma_alternative_pi_half}) is identical to
+\begin{equation}\label{gamma_alt2_pi_half}
+\gamma(\beta^g_1, c, 1/2) = \frac{(4)(1/2)g(\beta^g_1, c, 1/2)}{4 h(\beta^g_1, c, 1/2)} = \frac{2 g(\beta^g_1, c, 1/2)}{4h(\beta^g_1, c, 1/2)},
+\end{equation}
+i.e., the numerator and denominator of  (\ref{gamma_alt2_pi_half}) coincide with those of (\ref{gamma_alternative_pi_half}).
+We sometimes will use the notation $2\cdot g$ and $4\cdot h$ to refer to the numerator and denominator of (\ref{gamma_alternative_pi_half}), respectively.
+
+\subsection{Derivatives of  $g$ and $h$ in $c$}\label{sec:derivatives}
+We compute the derivatives of $g$ and $h$ in $c$, which we will need to prove subsequent results. First, by FTC and the evenness of $f$, we have that
+\begin{multline}\label{dg_dc}
+D_2 g(\beta^g_1, c, \pi) = -(1-\pi)f( \beta^g_0 + \beta^g_1 - c ) + (1-\pi) f(\beta^g_0 - c) \\ = (1-\pi) f(c - \beta^g_0) - (1-\pi)f(c - \beta^g_0 - \beta^g_1).
+\end{multline}
+Second, we have that
+\begin{multline}\label{dh_dc}
+D_2 h(\beta^g_1, c, \pi) = -[(1-\pi)f(\beta^g_0 - c) + \pi f( \beta^g_0 + \beta^g_1 - c )]\left[(1-\pi)\Phi(c - \beta^g_0) + \pi \Phi(c - \beta_0^g - \beta_1^g)  \right] \\ + [(1-\pi) f(c - \beta^g_0) +  \pi f(c - \beta^g_0 - \beta^g_1)] \left[(1-\pi) \Phi(\beta_0^g - c) + \pi \Phi(\beta^g_0 + \beta^g_1 - c) \right] \\ = \left[ (1-\pi) f(c - \beta^g_0) +  \pi f(c - \beta^g_0 - \beta^g_1) \right] \cdot \\ \bigg[ (1-\pi) \Phi(\beta_0^g - c) + \pi\Phi(\beta^g_0 + \beta^g_1 - c) - (1-\pi) \Phi(c - \beta^g_0) - \pi \Phi(c - \beta_0^g - \beta_1^g) \bigg].
+\end{multline}
+
+\subsection{Limit of $\gamma$ in $c$}\label{sec:c_limit}
+
+Assume (without loss of generality) that $\beta^g_1 > 0$.
+We compute $\lim_{c \to \infty} \gamma(\beta^g_1, c, \pi)$.
+Observe that $$\lim_{c \to \infty} g(\beta^g_1, c, \pi) = \lim_{c \to \infty} h(\beta^g_1, c, \pi)  = 0.$$ Therefore, we can apply L'H\^{o}pital's rule.
+We have by (\ref{dg_dc}) and (\ref{dh_dc}) that \begin{multline}\label{c_limit_product}
+\lim_{c \to \infty} \gamma(\beta^g_1, c, \pi) = \lim_{c \to \infty} \frac{\pi D_2 g(\beta^g_1, c, \pi)}{D_2h(\beta^g_1, c, \pi)} \\ = \lim_{c \to \infty} \bigg\{ \frac{(1-\pi) f(c - \beta^g_0) + \pi f(c - \beta^g_0 - \beta^g_1)}{\pi (1-\pi) f(c - \beta^g_0) - \pi (1-\pi)f(c - \beta^g_0 - \beta^g_1)} \\ \cdot \bigg[ (1-\pi) \Phi(\beta_0^g - c) + \pi \Phi(\beta^g_0 + \beta^g_1 - c) - (1-\pi) \Phi(c - \beta^g_0) - \pi \Phi(c - \beta_0^g - \beta_1^g) \bigg] \bigg\}^{-1}.
+ \end{multline}
+ We evaluate the two terms in the product (\ref{c_limit_product}) separately. Dividing by $f(c - \beta^g_0 - \beta^g_1) > 0$, we see that
+ \begin{equation}\label{c_limit_product_2}
+ \frac{(1-\pi) f(c - \beta^g_0) + \pi f(c - \beta^g_0 - \beta^g_1)}{\pi (1-\pi) f(c - \beta^g_0) - \pi (1-\pi)f(c - \beta^g_0 - \beta^g_1)} = \frac{\frac{(1-\pi) f(c - \beta^g_0)}{ f(c - \beta^g_0 - \beta^g_1)} + \pi}{\frac{ \pi(1-\pi) f(c - \beta^g_0)}{ f(c - \beta^g_0 - \beta^g_1)} - \pi(1-\pi)}.
+ \end{equation}
+ To evaluate the limit of (\ref{c_limit_product_2}), we first evaluate the limit of
+ \begin{multline}\label{c_limit_product_3}
+ \frac{f(c - \beta^g_0)}{f(c - \beta^g_0 - \beta^g_1)} = \frac{\exp{[-(1/2)(c - \beta_0^g)^2]}}{\exp{[-(1/2)( c - \beta^g_0 - \beta^g_1)^2]}} \\ = \frac{\exp[ -(1/2)(c^2 - 2 c \beta^g_0 + (\beta^g_0)^2)]}{\exp\left[-(1/2)( c^2 - 2c \beta^g_0 - 2 c \beta^g_1 + (\beta^g_0)^2 + 2( \beta^g_0 \beta^g_1) + (\beta^g_1)^2)\right]} \\ = \exp\big[-c^2/2 + c \beta^g_0 - (\beta^g_0)^2/2 \\ + c^2/2 - c \beta^g_0 - c \beta^g_1 + (\beta^g_0)^2/2 + \beta^g_0 \beta^g_1 + (\beta^g_1)^2/2 \big] \\ = \exp[ -c \beta^g_1 + \beta^g_0 \beta^g_1 + (\beta^g_1)^2/2] = \exp[ \beta^g_0 \beta^g_1 + (\beta^g_1)^2/2]\exp[ -c \beta^g_1].
+\end{multline}
+Taking the limit in (\ref{c_limit_product_3}), we obtain
+$$
+\lim_{c \to \infty} \frac{f(c - \beta^g_0)}{f(c - \beta^g_0 - \beta^g_1)} = \exp[ \beta^g_0 \beta^g_1 + (\beta^g_1)^2/2] \lim_{c \to \infty} \exp[ -c \beta^g_1] = 0
+$$ for $\beta^g_1 > 0$. We now can evaluate the limit of (\ref{c_limit_product_2}):
+$$ \lim_{c \to \infty} \frac{(1-\pi) f(c - \beta^g_0) + \pi f(c - \beta^g_0 - \beta^g_1)}{\pi (1-\pi) f(c - \beta^g_0) - \pi (1-\pi)f(c - \beta^g_0 - \beta^g_1)} = \frac{-\pi}{\pi(1-\pi)} = -\frac{1}{1 -\pi}.$$ Next, we compute the limit of the other term in the product (\ref{c_limit_product}):
+\begin{multline}\label{c_limit_product_4}
+\lim_{c \to \infty} \bigg[ (1-\pi)\Phi(\beta_0^g - c) + \pi \Phi(\beta^g_0 + \beta^g_1 - c) \\ - (1-\pi)\Phi(c - \beta^g_0) - \pi \Phi(c - \beta_0^g - \beta_1^g) \bigg] = -(1-\pi) - \pi = -1.
+\end{multline}
+Combining (\ref{c_limit_product_2}) and (\ref{c_limit_product_4}), the limit (\ref{c_limit_product}) evaluates to
+$$ \lim_{c \to \infty} \gamma(\beta^g_1, c, \pi) = \left(  \frac{ 1 }{ 1 - \pi }\right)^{-1} = 1 - \pi.$$ It follows that the limit in $c$ of the asymptotic relative bias $b$ is
+$$\lim_{c \to \infty} b(\beta^g_1, c, \pi) = 1 - \lim_{c \to \infty} \gamma(\beta^g_1, c, \pi) = \pi.$$
+A corollary is that
+$\lim_{c \to \infty} b(\beta^g_1, c, 1/2) = 1/2.$
+
+\subsection{Bayes-optimal decision boundary as a critical value of $\gamma$}\label{sec:bayes_opt}
+Let $c_\textrm{bayes} = \beta^g_0 + (1/2)\beta^g_1.$ We show that $c = c_\textrm{bayes}$ is a critical value of $\gamma$ for $\pi = 1/2$ and given $\beta^g_1$, i.e, $D_2 \gamma (\beta^g_1, c_\textrm{bayes}, 1/2) = 0.$ Differentiating (\ref{gamma_alt2_pi_half}), the quotient rule implies that
+\begin{equation}\label{quotient_rule}
+D_2\gamma(\beta^g_1, c, 1/2) \\ = \frac{D_2[2g(\beta^g_1, c, 1/2)] 4h(\beta^g_1, c, 1/2) - 2g(\beta^g_1, c, 1/2) D_2[4h(\beta^g_1, c, 1/2)]}{[4h(\beta^g_1, c, \pi)]^2}.
+\end{equation}
+We have by (\ref{dg_dc}) that
+\begin{equation}\label{dg_dc_bayes}
+D_2[2g(\beta^g_1, c_\textrm{bayes}, 1/2)] = f( \beta^g_1/2) - f( -\beta^g_1/2) = f(\beta^g_1/2) - f(\beta^g_1/2) = 0.
+\end{equation}
+Similarly, we have by (\ref{dh_dc}) that
+\begin{equation}\label{dh_dc_bayes}
+D_2[4 h(\beta^g_1, c_\textrm{bayes}, \pi)] = [f( \beta^g_1/2) + f( -\beta^g_1/2)] \left[  \Phi(-\beta^g_1/2) + \Phi(\beta^g_1/2) -  \Phi(\beta^g_1/2) - \Phi(-\beta^g_1/2) \right] = 0.
+\end{equation}
+Plugging in (\ref{dh_dc_bayes}) and (\ref{dg_dc_bayes}) to (\ref{quotient_rule}), we find that 
+$D_2[\gamma(\beta^g_1, c_\textrm{bayes}, 1/2)] = 0.$ Finally, because
+$$b(\beta^g_1, c, 1/2) = 1 - \gamma(\beta^g_1, c, 1/2),$$ it follows that
+$$D_2[b(\beta^g_1, c_\textrm{bayes}, 1/2)] = -D_2[\gamma(\beta^g_1, c_\textrm{bayes}, 1/2)] = 0.$$
+
+\subsection{Comparing Bayes-optimal decision boundary and large threshold}\label{sec:comparison}
+
+We compare the bias produced by setting the threshold to a large number to the bias produced by setting the threshold to the Bayes-optimal decision boundary.
+Let $r: \R^{\geq 0} \to \R$ be the value of attenuation function evaluated at the Bayes-optimal decision boundary $c_\textrm{bayes} = \beta^g_0 + (1/2) \beta^g_1$, i.e.
+\begin{multline*}
+r(\beta^g_1) = \gamma(\beta^g_1, \beta^g_0 + (1/2)\beta^g_1, 1/2) = \frac{\Phi(\beta^g_1/2) - \Phi(-\beta^g_1/2)}{\left[\Phi(-\beta^g_1/2) + \Phi( \beta^g_1/2) \right] \left[\Phi(\beta^g_1/2) + \Phi( -\beta^g_1/2)\right]} \\ = \frac{\int_{-\beta^g_1/2}^{\beta^g_1/2} f}{\left[ 1 - \Phi(\beta^g_1/2) + \Phi(\beta^g_1/2) \right]\left[ \Phi(\beta^g_1/2) + 1 - \Phi(\beta^g_1/2) \right]} = 2 \int_{0}^{\beta^g_1/2} f = 2 \Phi(\beta^g_1/2) - 1.
+\end{multline*}
+We set $r$ to $1/2$ and solve for $\beta^g_1$:
+\begin{multline*}
+r(\beta^g_1) = 1/2 \iff 2\Phi(\beta^g_1/2) -1 = 1/2 \iff \Phi(\beta^g_1/2) = 3/4 \iff \beta^g_1 = 2 \Phi^{-1}(3/4) \approx 1.35.
+\end{multline*}
+Because $r$ is a strictly increasing function, it follows that $r(\beta^g_1) < 1/2$ for $\beta^g_1 < 2\Phi^{-1}(3/4)$ and $r(\beta^g_1) > 1/2$ for $\beta^g_1 > 2\Phi^{-1}(3/4).$ Next, because $$b(\beta^g_1, c_\textrm{bayes}, 1/2) = 1 - \gamma(\beta^g_1, c_\textrm{bayes}, 1/2) = 1 - r(\beta^g_1),$$ we have that $b(\beta^g_1, c_\textrm{bayes}, 1/2) > 1/2$ for $\beta^g_1 < 2 \Phi^{-1}(3/4)$ and $b(\beta^g_1, c_\textrm{bayes}, 1/2) < 1/2$ for $\beta^g_1 > 2 \Phi^{-1}(3/4)$.
+Recall that the bias induced by sending the threshold to infinity (as stated in Proposition \ref{prop:c_limit} and proven in Section \ref{sec:c_limit}) is $1/2$, i.e.
+$$b(\beta^g_1, \infty, 1/2) = 1/2.$$ We conclude that $b(\beta^g_1, c_\textrm{bayes},1/2) > b(\beta^g_1, \infty, 1/2)$ on $\beta^g_1 \in [0, 2\Phi^{-1}(3/4))$; $b(\beta^g_1, c_\textrm{bayes},1/2) = b(\beta^g_1, \infty, 1/2)$ for $\beta^g_1 = 2\Phi^{-1}(3/4)$; and $b(\beta^g_1, c_\textrm{bayes},1/2) < b(\beta^g_1, \infty, 1/2)$ on $\beta^g_1 \in (2\Phi^{-1}(3/4), \infty)$.
+
+\subsection{Monotonicity in $\beta^g_1$}\label{sec:monotone}
+We show that $\gamma$ is monotonically increasing in $\beta^g_1$ for $\pi = 1/2$ and given threshold $c$.
+We begin by stating and proving two lemmas.
+The first lemma establishes an inequality that will serve as the basis for the proof.
+
+\begin{lemma}
+The following inequality holds: 
+\begin{multline}\label{basic_ineq_cp}
+\left[\Phi(\beta^g_0 - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right] \cdot \left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c) + \Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g) \right] \\ \geq \left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c)\right]\left[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g)\right].
+\end{multline}
+\end{lemma}
+\textbf{Proof}: We take cases on the sign on $\beta^g_1$.
+
+\noindent
+\underline{Case 1}: $\beta^1_g < 0$.
+Then $ \beta^g_1 + (\beta^g - c) < (\beta^g_0 - c),$ implying $\Phi(\beta^g_0 + \beta^g_1 - c) < \Phi(\beta^g_0 - c),$ or $[\Phi(\beta^g_0 + \beta^g_1 - c) - \Phi(\beta^g_0 - c)] < 0.$ Moreover, $[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g)]$ is positive.
+Therefore, the right-hand side of (\ref{basic_ineq_cp}) is negative.
+
+Turning our attention of the left-hand side of (\ref{basic_ineq_cp}), we see that
+\begin{equation}\label{basic_ineq_cp_2}
+\Phi(\beta^g_0 + \beta^g_1 - c) + \Phi( c - \beta^g_0 - \beta^g_1) = 1 -\Phi(\beta^g_0 + \beta^g_1 - c) + \Phi( c - \beta^g_0 - \beta^g_1) = 1.
+\end{equation}
+Additionally, $\Phi(\beta^g_0 - c) < 1$ and $ \Phi(c - \beta^g_0) > 0$. Combining these facts with (\ref{basic_ineq_cp_2}), we find that
+$$ \left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c) + \Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g) \right] > 0.
+$$ Finally, because $\left[\Phi(\beta^g_0 - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right] > 0,$ the entire left-hand side of (\ref{basic_ineq_cp}) is positive.
+The inequality holds for $\beta^g_1 < 0$.
+
+\noindent
+\underline{Case 2}: $\beta^1_g \geq 0.$  We will show that the first term on the LHS of (\ref{basic_ineq_cp}) is greater than the first term on the RHS of (\ref{basic_ineq_cp}), and likewise that the second term on the LHS is greater than the second term on the RHS, implying the truth of the inequality.
+Focusing on the first term, the positivity of $\Phi(\beta^g_0 -c)$ implies that
+$\Phi(\beta^g_0 - c) \geq - \Phi(\beta^g_0 - c),$ and so
+$$\Phi(\beta^g_0 - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \geq \Phi(\beta^g_0 - \beta^g_1 - c) - \Phi(\beta^g_0 - c).$$
+Next, focusing on the second term, $\beta^g_1 \geq 0$ implies that 
+\begin{equation}\label{basic_ineq_cp_3}
+\beta^g_1 + \beta^g_0 - c \geq \beta^g_0 - c \implies \Phi(\beta^g_1 + \beta^g_0 - c) - \Phi(\beta^g_0 - c) \geq 0.
+\end{equation}
+Adding $\Phi(c - \beta^g_0) + \Phi(c - \beta^g_0 - \beta^g_1)$ to both sides of (\ref{basic_ineq_cp_3}) yields
+\begin{equation*}
+\Phi(\beta^g_1 + \beta^g_0 - c) - \Phi(\beta^g_0 - c) + \Phi(c - \beta^g_0) + \Phi(c - \beta^g_0 - \beta^g_1) \geq \Phi(c - \beta^g_0) + \Phi(c - \beta^g_0 - \beta^g_1). \textrm{ }
+\end{equation*}
+The inequality holds for $\beta^g_1 \geq 0$. Combining the cases, the inequality holds for all $\beta^g_1 \in \R$. $\square$
+
+The second lemma establishes the derivatives of the functions $2\cdot g$ and $4 \cdot h$ in $\beta^g_1$.
+\begin{lemma}
+The derivatives in $\beta^g_1$ of $2\cdot g$ and $4\cdot h$ are
+\begin{equation}\label{dg_dbeta}
+\textcolor{violet}{D_1[2g(\beta^g_1, c, 1/2)] = f(\beta^g_0 + \beta^g_1 - c)}
+\end{equation}
+and
+\begin{multline}\label{dh_dbeta}
+\textcolor{teal}{D_1[4h(\beta^g_1, c, 1/2)] = f(\beta^g_0 + \beta^g_1 - c) \left[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g) \right]} \\ \textcolor{teal}{- f(\beta^g_0 + \beta^g_1 - c) \left[\Phi(\beta_0^g - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right]}.\end{multline}
+\end{lemma}
+\textbf{Proof}: Apply FTC and product rule. $\square$
+
+We are ready to prove the monotonicity of $\gamma$ in $\beta^g_1$.
+Subtracting $$\left[\Phi(\beta_0^g - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right]\left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c)\right]$$ from both sides of (\ref{basic_ineq_cp}) and multiplying by $f(\beta^g_0 + \beta^g_1 - c) > 0$ yields
+\begin{multline}\label{basic_ineq_cp_4}
+\textcolor{violet}{f(\beta^g_0 + \beta^g_1 - c)} \textcolor{red}{ \left[\Phi(\beta^g_0 - c) + \Phi\left(\beta^g_0 + \beta^g_1 - c \right) \right] \left[ \Phi(c - \beta^g_0) + \Phi(c - \beta^g_0 - \beta^g_1) \right]}  \\ \geq \textcolor{teal}{f(\beta^g_0 + \beta^g_1 - c) \left[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g)\right]}\textcolor{blue}{\left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c)\right]} \\ -\textcolor{teal}{f(\beta^g_0 + \beta^g_1 - c)   \left[\Phi(\beta_0^g - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right]} \textcolor{blue}{\left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c)\right]}.
+\end{multline}
+Next, recall that
+\begin{equation}\label{def_2g}
+\textcolor{blue}{2g(\beta^g_1,c,1/2) = \Phi(\beta^g_0 + \beta^g_1 - c) - \Phi(\beta^g_0 - c)}.
+\end{equation}
+and
+\begin{equation}\label{def_4h}
+\textcolor{red}{4h(\beta^g_1, c, 1/2) = \left[ \Phi(\beta^g_0 - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right] \left[\Phi(c - \beta^g_0) + \Phi( c - \beta^g_0 - \beta^g_1) \right]}.
+\end{equation}
+Substituting (\ref{dg_dbeta}, \ref{dh_dbeta}, \ref{def_2g}, \ref{def_4h}) into (\ref{basic_ineq_cp_4}) produces
+\begin{equation*}
+\textcolor{violet}{D_1[2g(\beta^g_1, c, 1/2)]}\textcolor{red}{4h(\beta^g_1, c, 1/2)} \geq \textcolor{blue}{2g(\beta^g_1, c, 1/2)}\textcolor{teal}{D_1[4h(\beta^g_1, c, 1/2)]},
+\end{equation*}
+or 
+\begin{equation}\label{basic_ineq_cp_5}
+\textcolor{violet}{D_1[2g(\beta^g_1, c, 1/2)]}\textcolor{red}{4h(\beta^g_1, c, 1/2)} - \textcolor{blue}{2g(\beta^g_1, c, 1/2)}\textcolor{teal}{D_1[4h(\beta^g_1, c, 1/2)]} \geq 0.
+\end{equation}
+The quotient rule implies that
+\begin{equation}\label{d_gamma_d_beta}
+D_1 \gamma(\beta^g_1, c, 1/2) = \frac{ \textcolor{violet}{D_1[2g(\beta^g_1, c, 1/2)]}\textcolor{red}{4h(\beta^g_1, c, 1/2)} - \textcolor{blue}{2g(\beta^g_1, c, 1/2)}\textcolor{teal}{D_1[4h(\beta^g_1, c, 1/2)]} }{[4h(\beta^g_1, c, 1/2)]^2}.
+\end{equation}
+We conclude by (\ref{basic_ineq_cp_5}) and (\ref{d_gamma_d_beta}) that $\gamma$ is monotonically increasing in $\beta^g_1$.
+Finally, $b(\beta^g_1, c, \pi) = 1 - \gamma(\beta^g_1, c, \pi)$ is monotonically decreasing in $\beta^g_1$.
+
+\subsection{Strict attenuation bias}\label{sec:att_bias}
+
+We begin by computing the limit of $\gamma$ in $\beta^g_1$ given $\pi = 1/2$.  First,
+\begin{multline*}
+\lim_{\beta^g_1 \to \infty} \gamma(\beta^g_1, c, 1/2) = \frac{1 - \Phi(\beta^g_0 - c)}{\left[1 + \Phi(\beta^g_0 - c) \right] \left[\Phi(c - \beta^g_0) \right]} \\ = \frac{\Phi(c - \beta^g_0)}{ \left[1 + \Phi(\beta^g_0 - c) \right] \left[\Phi(c - \beta^g_0) \right]} = \frac{1}{1 + \Phi(\beta^g_0 - c)} < 1.
+\end{multline*}
+Similarly,
+\begin{equation*}
+\lim_{\beta^g_1 \to -\infty} \gamma(\beta^g_1, c, 1/2) = \frac{ - \Phi(\beta^g_0 - c)}{\left[\Phi(\beta^g_0 - c)\right] \left[\Phi(c - \beta^g_0) + 1 \right]} = \frac{-1}{1 + \Phi(c - \beta^g_0)} > -1.
+\end{equation*}
+The function $\gamma(\beta^g_1, c, 1/2, \beta^g_0)$ is monotonically increasing in $\beta^g_1$ (as stated in Proposition \ref{prop:monotonic} and proven in section \ref{sec:monotone}).
+It follows that
+$$-1 < -\frac{1}{1 + \Phi(c - \beta^g_0)} \leq \gamma(\beta^g_1, c, 1/2, \beta^g_0) \leq \frac{1}{1 - \Phi(\beta^g_0 - c)} < 1$$ for all $\beta^g_1 \in \R$.
+But $\beta^g_0$ and $c$ were chosen arbitrarily, and so
+$$-1 < \gamma(\beta^g_1, c, 1/2, \beta^g_0) < 1$$ for all $(\beta^g_1, c, \beta^g_0) \in \R^3$.
+Finally, because $b(\beta^g_1, c, 1/2, \beta^g_0) = 1 - \gamma(\beta^g_1, c, 1/2, \beta^g_0)$, it follows that
+$$ 0 < b(\beta^g_1, c, 1/2, \beta^g_0) < 2$$ for all $(\beta^g_1, c, \beta^g_0) \in \R^3$
+
+\subsection{Bias-variance decomposition in no-intercept model}\label{sec:bv_decomp}
+
+We prove the bias-variance decomposition for the no-intercept version of (\ref{theoretical_model}). Define $l$ (for ``limit'') by
+$$l = \beta_m \left(\frac{\omega \pi}{\zeta(1-\pi) + \omega \pi}\right),$$ where
+$$
+\omega = \bar{\Phi}(c - \beta_g) = \Phi(\beta_g - c); \quad
+\zeta = \bar{\Phi}(c) = \Phi(-c).
+$$
+We have that
+\begin{equation*}
+\hat{\beta}_m - l = \frac{\sum_{i=1}^n \hat{p}_i m_i}{ \sum_{i=1}^n \hat{p}^2_i} - l = \frac{\sum_{i=1}^n \hat{p}_i m_i}{ \sum_{i=1}^n \hat{p}^2_i} - \frac{l \sum_{i=1}^n \hat{p}_i^2 }{ \sum_{i=1}^n \hat{p}_i^2} \\ = \frac{\sum_{i=1}^n \hat{p}_i(m_i - l \hat{p}_i)}{ \sum_{i=1}^n \hat{p}_i^2}.
+\end{equation*}
+Therefore,
+\begin{equation}\label{bc_decomp_1}
+\sqrt{n}(\hat{\beta}_m - l) = \frac{(1/\sqrt{n})\sum_{i=1}^n \hat{p}_i(m_i - l \hat{p}_i)}{(1/n)\sum_{i=1}^n \hat{p}_i^2}.
+\end{equation}
+Next, we compute the expectation and variance of $\hat{p}_i(m_i - l\hat{p}_i)$. To do so, we first compute several simpler quantities:
+\begin{enumerate}
+\item Expectation of $\hat{p}_i$: $
+\E[\hat{p}_i] = \P(p_i\beta_g + \tau_i \geq c) =  \P(\beta_g + \tau_i \geq c)\pi + \P(\tau_i \geq c)(1-\pi) = \pi \omega + (1-\pi)\zeta.$
+
+\item Expectation of $\hat{p}_i p_i$: $\E\left[\hat{p}_i p_i\right] = \E\left[\hat{p}_i | p_i = 1 \right]\P\left[p_i = 1\right] = \omega \pi.$
+\item Expectation of $\hat{p}_i m_i$:
+\begin{multline*}
+\E[\hat{p}_i m_i] = \E\left[\hat{p}_i(\beta_m p_i + \ep_i)\right] = \E\left[\beta_m \hat{p}_i p_i + \hat{p}_i \ep_i \right] \\ = \beta_m \E\left[ \hat{p}_i p_i \right] + \E[\hat{p}_i]\E[\ep_i] = \beta_m \omega \pi + 0 = \beta_m \omega \pi.
+\end{multline*}
+\item Expectation of $\hat{p}_i m_i^2$: \begin{multline*}
+\E\left[\hat{p}_i m_i^2\right] = \E \left[ \hat{p}_i( \beta_m p_i + \ep_i )^2 \right] = \E\left[ \hat{p}_i \left( \beta_m^2 p_i^2 + 2 \beta_m p_i \ep_i + \ep_i^2 \right)  \right] \\ = \E\left[ \hat{p}_i p_i \beta^2_m + 2 \beta_m p_i \hat{p}_i \ep_i + \hat{p}_i \ep_i^2 \right] = \beta^2_m \E[ \hat{p}_i p_i] + 2 \beta_m \E[p_i\hat{p}_i] \E[\ep_i] + \E[\hat{p}_i] \E[ \ep^2_i ] \\ = \beta^2_m \E[ \hat{p}_i p_i] + \E[\hat{p}_i] = \beta^2_m \omega \pi + \E[ \hat{p}_i].
+\end{multline*}
+\end{enumerate}
+
+Now, we can compute the expectation and variance of $\hat{p}_i(m_i - l\hat{p}_i)$. First,
+\begin{equation}\label{bv_decomp_2}
+\E\left[\hat{p}_i(m_i - l\hat{p}_i) \right] = \E[\hat{p}_i m_i] - l \E[\hat{p}_i] = \beta_m \omega \pi - \left(\frac{\beta_m \omega \pi}{\zeta (1-\pi) + \omega \pi}\right)[\zeta (1-\pi) + \omega \pi] = 0.
+\end{equation}
+Additionally,
+\begin{multline}\label{bv_decomp_3}
+\V\left[\hat{p}_i(m_i - l\hat{p}_i)\right] = \E\left[\hat{p}_i^2(m_i - l\hat{p}_i)^2\right] - (\E\left[ \hat{p}_i(m_i - l\hat{p}_i)\right])^2 \\ = \E\left[ \hat{p}_i m_i^2\right] - 2l \E[m_i\hat{p}_i] +l^2 \E[\hat{p}_i]= \beta^2_m \omega \pi + \E[ \hat{p}_i] -2l \beta_m \omega \pi + l^2 \E[\hat{p}_i] \\ = \beta_m\omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2).
+\end{multline}
+Therefore, by CLT, (\ref{bv_decomp_2}), and (\ref{bv_decomp_3}),
+\begin{equation}\label{bv_decomp_4}
+(1/\sqrt{n})\sum_{i=1}^n \hat{p}_i(m_i - l \hat{p}_i) \xrightarrow{d} N\left(0, \beta_m\omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2) \right).
+\end{equation}
+Next, by weak LLN,
+\begin{equation}\label{bv_decomp_5}
+(1/n) \sum_{i=1}^n \hat{p}_i^2 = (1/n) \sum_{i=1}^n \hat{p}_i \xrightarrow{P} \E[\hat{p}_i].
+\end{equation}
+Finally, by (\ref{bc_decomp_1}), (\ref{bv_decomp_4}), (\ref{bv_decomp_5}), and Slutsky's Theorem,
+$$ \sqrt{n}(\hat{\beta}_m - l) \xrightarrow{d} N\left(0, \frac{ \beta_m\omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2) }{\left(\E[\hat{p}_i]\right)^2} \right).$$ Thus, for large $n \in \N$, we have that
+$$
+\E [\hat{\beta}_m] \approx l; \quad
+\V[\hat{\beta}_m] \approx \left[\beta_m\omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2)\right]/[n\E^2[\hat{p}_i]],$$
+completing the bias-variance decomposition.
+
+\section{Estimation and inference in the GLM-EIV model}\label{sec:glmeiv_details}
+\subsection{Estimation}
+We estimate the parameters of the GLM-EIV model using an EM algorithm.
+\subsubsection*{E step}
+The E step entails computing the membership probability of each cell.
+Let $\theta^{(t)} = (\beta_m^{(t)}, \beta_g^{(t)}, \pi^{(t)})$ be the parameter estimate at the $t$-th iteration of the algorithm.
+For $k \in \{0,1\}$, let $[\eta^m_i(k)]^{(t)}$ be the $i$th canonical parameter at the $t$-th iteration of the algorithm of the gene expression distribution that results from setting $p_i$ to $k$, i.e.
+$$
+[\eta^m_i(k)]^{(t)} := h_m\left( \langle \tilde{x}_i(k) , \beta_m^{(t)} \rangle + o^m_i \right).
+$$ Similarly, let $\left[\eta^g_i(k)\right]^{(t)}$ be defined by
+$$\left[\eta^g_i(k)\right]^{(t)} :=  h_g\left( \langle \tilde{x}_i(k) , \beta_g^{(t)} \rangle + o^g_i \right).$$
+Next, for $k \in \{0,1\},$ define $\alpha^{(t)}_i(k)$ by
+\begin{multline*}
+\alpha^{(t)}_i(k) := \P\left( M_i = m_i, G_i = g_i | P_i = k, \theta^{(t)} \right) \\ = \P\left( M_i = m_i | P_i = k, \theta^{(t)} \right) \P\left(G_i = g_i | P_i = k, \theta^{(t)} \right) \textrm{ (because $G_i \indep M_i | P_i$)} \\ = f_m\left(m_i; \left[ \eta^m_i(k) \right]^{(t)}\right) f_g\left(g_i; \left[ \eta^g_i(k) \right]^{(t)} \right).
+\end{multline*}
+Finally, let 
+$$
+\begin{cases}
+\pi^{(t)}(1) := \pi^{(t)} = \P\left(P_i = 1 | \theta^{(t)} \right) \\
+\pi^{(t)}(0) := 1 - \pi^{(t)} = \P\left(P_i = 0 | \theta^{(t)} \right).
+\end{cases}
+$$
+The $i$th membership probability $T^{(t)}_i(1)$ is
+\begin{multline}\label{e_step_1}
+T^{(t)}_i(1) = \P(P_i = 1 | M_i = m_i, G_i = g_i, \theta^{(t)})  = \frac{\pi^{(t)}(1) \alpha^{(t)}_i(1)}{ \sum_{k=0}^1 \pi^{(t)}(k) \alpha^{(t)}_i(k)} \textrm{ (by Bayes rule)} \\ = \frac{1}{\frac{ \pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)} + 1} = \frac{1}{ \exp\left(\log\left(\frac{\pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)}\right)\right) + 1} = \frac{ 1 }{ \exp\left(q^{(t)}_i\right) + 1},
+\end{multline}
+where we set 
+\begin{equation}\label{e_step_2}
+q_i^{(t)} := \log\left(\frac{\pi^{(t)}(0) \alpha_i^{(t)}(0)}{\pi^{(t)}(1) \alpha_i^{(t)}(1)}\right).
+\end{equation}
+Next, we have that
+\begin{multline*}
+q^{(t)}_i = \log\left[ \pi^{(t)}(0) \right] + \log\left[ f_m\left(m_i; \left[ \eta^m_i(0) \right]^{(t)}\right) \right] + \log\left[ f_g\left(g_i; \left[ \eta^g_i(0) \right]^{(t)}\right) \right] \\ - \log\left[ \pi^{(t)}(1) \right] - \log\left[ f_m\left(m_i; \left[ \eta^m_i(1) \right]^{(t)}\right) \right] - \log\left[ f_g\left(g_i; \left[ \eta^g_i(1) \right]^{(t)}\right) \right],
+\end{multline*}
+which we can compute. We therefore conclude that
+$$ T_i^{(t)} = \frac{1}{\exp\left(q^{(t)}_i\right) + 1}.$$
+\subsection*{M step}
+The complete-data log-likelihood of the GLM-EIV model  is
+\begin{equation}\label{full_log_lik}
+\mathcal{L}(\theta; m, g, p) = \sum_{i=1}^n \left[ p_i \log(\pi) + (1-p_i) \log(1-\pi) \right] + \sum_{i=1}^n \log\left( f_m(m_i; \eta^m_i)\right) + \sum_{i=1}^n \log\left( f_g(g_i; \eta_i^g) \right).
+\end{equation}
+Define $Q(\theta | \theta^{(t)}) = \E_{\left(P |M = m, G = g, \theta^{(t)}\right)}\left[ \mathcal{L}(\theta; m, g, p) \right].$ We have that
+\begin{multline}\label{Q_funct}
+Q(\theta |\theta^{(t)}) = \sum_{i=1}^n \left[T^{(t)}_i(1)\log(\pi) + T_i^{(t)}(0) \log(1 - \pi)\right] \\ + \sum_{k=0}^1 \sum_{i=1}^n T^{(t)}_i(k) \log \left[ f_m(m_i; \eta_i^m(k)) \right] + \sum_{k=0}^1 \sum_{i=1}^n T^{(t)}_i(k) \log \left[ f_g( g_i; \eta^{g,b}_i(k)) \right].
+\end{multline}
+The three terms of (\ref{Q_funct}) are functions of different parameters: the first is a function of $\pi,$ the second is a function of $\beta_m,$ and the third is a function of $\beta_g$.
+Therefore, to find the maximizer $\theta^{(t+1)}$ of (\ref{Q_funct}), we maximize the three terms separately.
+Differentiating the first term with respect to $\pi$, we find that
+\begin{equation*}
+\frac{ \partial }{\partial \pi } \sum_{i=1}^n \left[ T^{(t)}_i(1)\log(\pi) + T_i^{(t)}(0) \log(1 - \pi)\right]  = \frac{\sum_{i=1}^n T_i^{(t)}(1)}{\pi} - \frac{ \sum_{i=1}^n T_i^{(t)}(0) }{ 1 - \pi}.
+\end{equation*} Setting the derivative equal to $0$ and solving for $\pi$,
+\begin{multline*}
+\frac{\sum_{i=1}^n T_i^{(t)}(1)}{\pi} - \frac{ \sum_{i=1}^n T_i^{(t)}(0) }{ 1 - \pi} = 0 \iff \sum_{i=1}^n T_i^{(t)}(1) - \pi \sum_{i=1}^n T^{(t)}_i(1) = \pi \sum_{i=1}^n T_i(0) \\ \iff \sum_{i=1}^n T^{(t)}_i(1) - \pi\sum_{i=1}^n T_i^{(t)}(1) = \pi n - \pi\sum_{i=1}^n T_i(1) \iff \pi = \frac{ \sum_{i=1}^n T_i^{(t)} (1) }{n}.\end{multline*}
+Thus, the maximizer $\pi^{(t+1)}$ of (\ref{Q_funct}) in $\pi$ is $\pi^{(t+1)} = (1/n)\sum_{i=1}^n T^{(t)}_i(1)$.
+Next, define $w^{(t)} = [T^{(t)}_1(0), \dots, T^{(t)}_n(0), T^{(t)}_1(1), \dots, T^{(t)}_n(1)]^T \in \R^{2n}$.
+We can view the second term of (\ref{Q_funct}) as the log-likelihood of a GLM -- call it $\textrm{GLM}^{(t)}_m$ -- that has exponential family density $f_m$, link function $r_m$, responses $[m,m]^T$, offsets $[o^m, o^m]^T$, weights $w^{(t)}$, and design matrix $\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix}.$ Therefore, the maximizer $\beta^{(t+1)}_m$ of the second term of (\ref{Q_funct}) is the maximizer of $\textrm{GLM}^{(t)}_m$, which we can compute using the iteratively reweighted least squares (IRLS) procedure, as implemented in R's GLM function.
+Similarly, the maximizer $\beta^{(t+1)}_g$ of the third term of (\ref{Q_funct}) is the maximizer of the GLM with exponential family density $f_g$, link function $r_g$, responses $[g,g]^T$, offsets $[o^g, o^g]^T$, weights $w^{(t)}$, and design matrix $\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix}.$
+
+\subsection{Inference}
+We derive the asymptotic observed information matrix of the GLM-EIV log likelihood, enabling us to perform inference on the parameters.
+First, we define some notation.
+For $i \in \{1, \dots, n\}$, $j \in \{0, 1\}$, and $\theta = (\pi, \beta_m, \beta_g),$ let $T^\theta_i(j)$ be defined by
+$$T^\theta_i(j) = \P_\theta\left(P_i = j | M_i = m_i, G_i = g_i\right).$$ Let the $n \times n$ matrix $T^\theta(j)$ be given by $$T^\theta(j) = \textrm{diag}\left\{T^\theta_1(j), \dots, T^\theta_n(j)\right\}.$$
+Next, define the diagonal $n \times n$ matrices $\Delta^m$, $[\Delta']^m$, $V^m$, and $H^m$ by
+$$
+\begin{cases}
+	\Delta^m = \textrm{diag} \{h_m'(l_1^m), \dots, h_m'(l_n^m)\} \\
+	[\Delta']^m = \textrm{diag} \{h_m''(l_1^m), \dots, h_m''(l_n^m) \} \\
+	V^m = \textrm{diag} \{ \psi''_m( \eta^m_1), \dots, \psi''_m( \eta^m_n) \} \\
+	H^m = \textrm{diag} \{ m_1 - \mu_1^m, \dots, m_n - \mu_n^m\}.
+\end{cases} 
+$$ Define the $n \times n$ matrices $\Delta^g, [\Delta']^g, V^g,$ and $H^g$ analogously. These matrices are \textit{unobserved}, as they depend on $\{p_1, \dots, p_n\}$.
+
+For $j \in \{0,1\}$, let the diagonal $n \times n$ matrices $\Delta^m(j), [\Delta']^m(j), V^m(j),$ and $H^m(j)$ be given by
+$$\begin{cases}
+\Delta^m(j) = \textrm{diag} \{ h_m'(l_1^m(j)), \dots, h_m'( l_n^m(j) ) \} \\
+[\Delta']^m(j) = \textrm{diag} \{ h_m''(l_1^m(j)), \dots, h_m''( l_n^m(j)) \} \\
+V^m(j) = \textrm{diag} \{\psi''_m( \eta^m_1(j)), \dots, \psi''_m( \eta^m_n(j))\} \\
+H^m(j) = \textrm{diag} \{m_1 - \mu_1^m(j), \dots, m_n - \mu_n^m(j)\} .
+\end{cases}
+$$
+Define the matrices $\Delta^g(j)$, $[\Delta']^{g}(j)$, $V^g(j),$ and $H^g(j)$ analogously. Finally, define the vectors $s^m(j)$ and $w^m(j)$ in $\R^n$ by 
+$$ \begin{cases}
+s^m(j) = [m_1 - \mu_1^m(j), \dots, m_n - \mu_n^m(j) ]^T \\ w^m(j) = [ T_1(0)T_1(1)\Delta^m_1(j) H^m_1(j), \dots, T_n(0)T_n(1)\Delta_n^m(j) H_n^m(j)]^T,
+\end{cases} $$
+and let the vectors $s^g(j)$ and $w^g(j)$ be defined analogously.
+The quantities $\Delta^m(j), [\Delta']^m(j), V^m(j),$ $H^m(j),$ $s^m(j),$ $w^m(j),$ $\Delta^g(j), [\Delta']^g(j), V^g(j),$ $H^g(j),$ $s^g(j),$ and $w^g(j)$ are all \textit{observed}.
+
+The observed information matrix $J(\theta; m, g)$ evaluated at $\theta = (\pi, \beta_m, \beta_g)$ is the negative Hessian of the log likelihood (\ref{marginal_log_lik}) evaluated at $\theta$, i.e.
+$J(\theta; m, g) = - \nabla^2\mathcal{L}(\theta; m, g) .$ This quantity, unfortunately, is hard to compute, as the log likelihood (\ref{marginal_log_lik}) is a complicated mixture.
+Louis \cite{Louis1982} showed that $J(\theta; m, g)$ is equivalent to the following quantity:
+\begin{multline}\label{zero_inf_info_mat}
+J(\theta; m, g) = -\E \left[\nabla^2 \mathcal{L}(\theta; m, g, p) | G = g, M = m \right] \\ + \E\left[\nabla \mathcal{L}(\theta; m, g, p) | G = g, M = m \right] \E\left[\nabla \mathcal{L}(\theta; m, g, p) | G = g, M = m \right]^T \\ - \E\left[ \nabla\mathcal{L}(\theta; m, g, p) \nabla \mathcal{L}(\theta; m, g, p)^T | G = g, M = m \right].
+\end{multline}
+The observed information matrix $J(\theta; m, g)$ has dimension $(2d+1) \times (2d + 1).$ Recall that the log-likelihood (\ref{full_log_lik}) is the sum of three terms.
+The first term depends only on $\pi$, the second on $\beta_m$, and the third on $\beta_g$.
+Therefore, the observed information matrix can be viewed as block matrix consisting of nine submatrices (Figure \ref{infomatrixbackground}; only six submatrices labelled).
+Submatrix I depends on $\pi$, submatrix II on $\beta_m$, submatrix III on $\beta_g$, submatrix IV on $\beta_m$ and $\beta_g$, submatrix V on $\pi$ and $\beta_m$, and submatrix VI on $\pi$ and $\beta_g$.
+We only need to compute these six submatrices to compute the entire matrix, as the matrix is symmetric.
+The following sections derive formulas for submatrices I-VI.
+All expectations are understood to be \textit{conditional} on $m$ and $g$.
+The notation $\nabla_v$  and $\nabla^2_v$  represent the gradient and Hessian, respectively, with respect to the vector $v$.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\linewidth]{../../figures/info_matrix/info_matrix_background}
+	\caption{Block structure of the observed information matrix $J(\theta; m, g) = -\nabla^2 \mathcal{L}(\theta; m, g)$.
+The matrix is symmetric, and so we only need to compute submatrices I-VI to compute the entire matrix.}
+	\label{infomatrixbackground}
+\end{figure}
+
+\subsubsection*{Submatrix I}
+Denote submatrix I by $J_{\pi}(\theta; m, g).$ The formula for $J_{\pi}(\theta; m, g)$ is 
+\begin{equation}\label{sub_mat_pi}
+J_{\pi}(\theta; m, g) = -\E\left[\nabla^2_\pi \mathcal{L}(\theta; m, g, p) \right] + \left(\E\left[ \nabla_\pi \mathcal{L}(\theta; m, g, p) \right] \right)^2 - \E\left[(\nabla_\pi \mathcal{L}(\theta; m, g, p))^2 \right].
+\end{equation}
+
+We begin by calculating the first and second derivatives of the log-likelihood $\mathcal{L}$ with respect to $\pi$. The first derivative is
+\begin{multline}\label{d_L_d_pi}
+\nabla_\pi \mathcal{L}(\theta; m, g, p) = \frac{\partial }{\partial \pi } \left( \sum_{i=1}^n p_i \log(\pi) + \sum_{i=1}^n (1 - p_i) \log(1 - \pi) \right) \\ = \frac{ \sum_{i=1}^n p_i }{\pi} - \frac{ \sum_{i=1}^n (1 - p_i) }{ 1 - \pi } = \frac{\sum_{i=1}^n p_i}{\pi} - \frac{n - \sum_{i=1}^n p_i}{1 - \pi} = \left( \frac{1}{\pi} + \frac{1}{1 - \pi} \right) \sum_{i=1}^n p_i - \frac{n}{1-\pi}.
+\end{multline}
+The second derivative is
+\begin{multline*}
+\nabla^2_\pi \mathcal{L}(\theta; m, g, p)  = \frac{\partial^2}{\partial^2\pi} \left( \frac{ \sum_{i=1}^n p_i }{ \pi } - \frac{ n - \sum_{i=1}^n p_i }{1 - \pi}  \right) = \frac{\left( \sum_{i=1}^n p_i \right) - n}{(1 - \pi)^2} - \frac{\sum_{i=1}^n p_i }{ \pi^2 }.
+\end{multline*}
+We compute the expectation of the first term of (\ref{sub_mat_pi}):
+\begin{multline}\label{submat_pi_1}
+\E \left[ -\nabla^2_{\pi} \mathcal{L}(\theta; m, g, p)\right] = - \E\left[\frac{ ( \sum_{i=1}^n p_i) - n}{(1 - \pi)^2} - \frac{\sum_{i=1}^n p_i}{\pi^2} \right] \\ = - \E\left\{\left[\frac{1}{(1-\pi)^2} - \frac{1}{\pi^2} \right] \sum_{i=1}^n p_i - \frac{n}{ (1 - \pi)^2 } \right\} \\ = - \left\{\left[ \frac{1}{(1-\pi)^2} - \frac{1}{\pi^2} \right] \sum_{i=1}^n T^\theta_i(1) - \frac{n}{ (1 - \pi)^2}  \right\} \\ = \left[ \frac{1}{\pi^2} - \frac{1}{(1 - \pi)^2} \right] \sum_{i=1}^n T^\theta_i(1) + \frac{n}{(1-\pi)^2}.
+\end{multline}
+Next, we compute the difference of the second two pieces of (\ref{sub_mat_pi}). To this end, define $$a := \frac{1}{(1-\pi)} + \frac{1}{\pi}$$ and $$b := \frac{n}{(1-\pi)}.$$ We have that
+\begin{multline*}
+\E \left[\nabla_\pi \mathcal{L}(\theta; m, g, p)^2 \right] \\ = \E \left[ \left( a \sum_{i=1}^n p_i - b\right)^2 \right]  =  \E \left[ a^2 \left( \sum_{i=1}^n p_i \right)^2 - 2ab \sum_{i=1}^n p_i + b^2 \right] \\ = a^2 \sum_{i=1}^n \sum_{j=1}^n \E[p_i p_j] -2ab \sum_{i=1}^n \E [p_i] + b^2.
+\end{multline*}
+Next,
+\begin{equation*}
+\left( \E \left[\nabla_\pi \mathcal{L}(\theta; m, g, x) \right] \right)^2 = \left( a \sum_{i=1}^n \E [p_i] - b \right)^2 = a^2 \sum_{i=1}^n \sum_{j=1}^n \E[p_i]  E[p_j] - 2ab \sum_{i=1}^n \E[p_i] + b^2.
+\end{equation*}
+Therefore,
+\begin{multline}\label{submat_pi_2}
+(\E [\nabla_\pi \mathcal{L}(\theta; m, g, p)])^2 - \E \left[\nabla_\pi \mathcal{L}(\theta; m, g, p)^2 \right] \\ = a^2 \sum_{i=1}^n \sum_{j=1}^n \E[p_i] \E[p_j] - a^2 \sum_{i=1}^n \sum_{j=1}^n \E[p_i p_j] = a^2 \left( \sum_{i=1}^n \E[p_i]^2 - \E[p_i^2]\right) \\ = a^2 \left( \sum_{i=1}^n [T^\theta_i(1)]^2 - T^\theta_i(1) \right) = \left( \frac{1}{(1 - \pi)} + \frac{1}{\pi} \right)^2 \left(\sum_{i=1}^n [T^\theta_i(1)]^2 - T^\theta_i(1) \right).
+\end{multline}
+Stringing (\ref{sub_mat_pi}), (\ref{submat_pi_1}) and (\ref{submat_pi_2}) together, we obtain
+\begin{multline}\label{sub_mat_1_formula}
+J_\pi(\theta; m, g) = 
+\left[ \frac{1}{\pi^2} - \frac{1}{(1 - \pi )^2} \right] \sum_{i=1}^n T^\theta_i(1) + \frac{n}{(1-\pi )^2} \\ + \left( \frac{1}{(1 - \pi )} + \frac{1}{\pi} \right)^2 \left( \sum_{i=1}^n [T^\theta_i(1)]^2 - T^\theta_i(1) \right).
+\end{multline}
+
+\subsubsection*{Submatrix II}
+Denote submatrix II by $J_{\beta^m}(\theta; m, g).$ The formula for $J_{\beta^m}(\theta; m, g)$ is
+\begin{multline}\label{sub_mat_2}
+J_{\beta^m}(\theta; m, g) = -\E \left[\nabla_{\beta^m}^2 \mathcal{L}(\theta; m, g, p) \right] \\ + \E\left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] \E\left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T - \E\left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T  \right].
+\end{multline}
+Standard GLM results imply that
+$$ -\nabla_{\beta^m}^2 \mathcal{L}(\theta; m, g, p) = \tilde{X}^T ( \Delta^m V^m \Delta^m - [\Delta']^m H^m ) \tilde{X}$$ and $$\nabla_{\beta^m}\mathcal{L}(\theta; m, g, p) = \tilde{X}^T \Delta^m s^m.$$
+We compute the first term of (\ref{sub_mat_2}). The $(k,l)$th entry of this matrix is
+\begin{multline*}
+\left( \E\left[-\nabla_{\beta^m}^2 \mathcal{L}(\theta; m, g, p)\right]\right)[k,l] = \E \left\{\tilde{X}[,k]^T (\Delta^m V^m \Delta^m - [\Delta']^mH^m) \tilde{X}[,l] \right\} \\ = \sum_{i=1}^n \E \left\{ \tilde{x}_{i,k} (\Delta^m_{i} V^m_{i} \Delta^m_{i} - [\Delta']^m_{i} H^m_{i}) \tilde{x}_{i,l} \right\} \\ = \sum_{i=1}^n \tilde{x}_{i,k}(0) T_i^{\theta}(0) [{\Delta}^m_i(0)  {V}^m_i(0) {\Delta}^m_i(0) - [\Delta']^m_i(0) {H}^m_i(0)] \tilde{x}_{i,l}(0) \\ + \sum_{i=1}^n \tilde{x}_{i,k}(1) T_i^{\theta}(1) [ {\Delta}^m_i(1)  {V}^m_i(1) {\Delta}^m_i(1) - [{\Delta}']^m_i(1) {H}^m_i(1)] \tilde{x}_{i,l}(1) \\ = \sum_{s = 0}^1 \tilde{X}(s)[,k]^T {T}^{\theta}(s) \left[ {\Delta}^m(s) {V}^m(s) {\Delta}^m(s) - [{\Delta}']^m(s) {H}^m(s) \right] \tilde{X}(s)[,l].
+\end{multline*}
+We therefore have that
+\begin{multline}\label{sub_mat_2_1}
+\E\left[-\nabla_{\beta^m}^2 \mathcal{L}(\theta; m, g, p)\right] = \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) \left[ {\Delta}^m(s) {V}^m(s) {\Delta}^m(s) - [{\Delta}']^m(s) {H}^m(s) \right] \tilde{X}(s).
+\end{multline}
+Next, we compute the difference of the last two terms of (\ref{sub_mat_2}). The $(k,l)$th entry is
+\begin{multline*}
+\bigg[ \E \left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] \E \left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T \\ - \E \left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T \right] \bigg] [k,l] \\
+= \left[ \E \left[\tilde{X}^T \Delta^m s^m \right] \E \left[\tilde{X}^T \Delta^m s^m \right]^T\right][k,l] - \E \left[\tilde{X}^T \Delta^m s^m (s^m)^T \Delta^m \tilde{X} \right][k,l] \\
+= \E\left[ \tilde{X}[,k]^T \Delta^m s^m \right] \E \left[ \tilde{X}[,l]^T \Delta^m s^m \right] - \E \left[ \tilde{X}[,k]^T \Delta^m s^m (s^m)^T \Delta^m \tilde{X}[,l ] \right] \\ 
+=\E\left(\sum_{i=1}^n \tilde{x}_{ik} \Delta^m_i s^m_{i} \right) \E \left( \sum_{j=1}^n \tilde{x}_{jl} \Delta^m_j s^m_j \right) - \E \left( \sum_{i=1}^n \sum_{j=1}^n \tilde{x}_{ik} \Delta^m_i s^m_i s^m_j \Delta^m_j \tilde{x}_{jl} \right) \\
+= \sum_{i=1}^n \sum_{j=1}^n \E[ \tilde{x}_{ik} \Delta^m_is^m_i] \E [\tilde{x}_{jl} \Delta^m_j s^m_j]  -  \sum_{i=1}^n \sum_{j=1}^n \E [ \tilde{x}_{ik} \Delta^m_i s^m_i s^m_j \Delta^m_j \tilde{x}_{jl}] \\
+= \sum_{i=1}^n \sum_{j=1}^n \E[ \tilde{x}_{ik} \Delta^m_i s^m_i] \E \left[\tilde{x}_{jl} \Delta^m_j s^m_j \right]  - \sum_{i \neq j} \E [ \tilde{x}_{ik} \Delta^m_i s^m_i] \E[s^m_j \Delta^m_j \tilde{x}_{jl}] \\ - \sum_{i=1}^n \E[\tilde{x}_{ik} \Delta^m_i s^m_i s^m_i \Delta^m_i \tilde{x}_{il}] \\
+= \sum_{i=1}^n \E[\tilde{x}_{ik} \Delta^m_i s^m_i ] \E [ \tilde{x}_{il} \Delta^m_i s^m_i] - \sum_{i=1}^n \E[\tilde{x}_{ik} (\Delta_i^m)^2 (H_i^m)^2 \tilde{x}_{il}] \\ = \sum_{i=1}^n \left[\tilde{x}_{ik}(0) {\Delta}^m_i(0) T_i^{\theta}(0) {H}^m_i(0) + \tilde{x}_{ik}(1) {\Delta}^m_i(1) T_i^{\theta}(1) {H}^m_i(1) \right] \\ \cdot \left[ \tilde{x}_{il}(0) {\Delta}^m_i(0) T_i^{\theta}(0) {H}^m_i(0) + \tilde{x}_{il}(1) {\Delta}^m_i(1) T_i^{\theta}(1) {H}^m_i(1) \right] \\ - \sum_{i=1}^n \left[ \tilde{x}_{ik}(0) T_i^{\theta}(0) (\Delta_i^m(0))^2 ({H}_i^m(0))^2 \tilde{x}_{il}(0)  + \tilde{x}_{ik}(1) T_i^{\theta}(1) ({\Delta}^m_i(1))^2 ({H}_i^m(1))^m \tilde{x}_{il}(1) \right] \\ = \sum_{s=0}^1 \sum_{t=0}^1 \left[ \sum_{i=1}^n \tilde{x}_{ik}(s) T^{\theta}_i(s) {\Delta}^m_i(s) {H}^m_i(t) T_i^{\theta}(t){\Delta}^m_i(t) {H}^m_i(t) \tilde{x}_{il}(t) \right] \\ - \sum_{s=0}^1 \left[\sum_{i=1}^n \tilde{x}_{ik}(s) T_i^\theta(s) ({\Delta}^m_i(s))^2 ({H}_i^m(s))^2 \tilde{x}_{il}(s) \right] \\ = \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)[,k]^T {T}^\theta(s) {\Delta}^m(s) {H}^m(s) {T}^\theta(t) {\Delta}^m(t) {H}^m(t) \tilde{X}(k)[,l] \\ - \sum_{s=0}^1{X}(s)[,k]^T {T}^\theta(s) ({\Delta}^m(s))^2 ({H}^m(s))^2 \tilde{X}(s)[,l].
+\end{multline*}
+The sum of the last two terms on the right-hand side of (\ref{sub_mat_2}) is therefore
+\begin{multline}\label{sub_mat_2_2}
+ \E \left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] \E \left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T - \E \left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T \right] \\ =
+\sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T {T}^\theta(s) {\Delta}^m(s) {H}^m(s) T^{\theta}(t) {\Delta}^m(t) {H}^m(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T {T}^\theta(s) ({\Delta}^m(s))^2 ({H}^m(s))^2 \tilde{X}(s).
+\end{multline}
+Combining (\ref{sub_mat_2}), (\ref{sub_mat_2_1}), (\ref{sub_mat_2_2}), we find that
+\begin{multline}\label{sub_mat_2_formula}
+J_{\beta^m}(\theta; m, g) = \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) \left[ {\Delta}^m(s) {V}^m(s) {\Delta}^m(s) - [{\Delta}']^m(s) {H}^m(s) \right] \tilde{X}(s) \\ + \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T {T}^\theta(s) {\Delta}^m(s) {H}^m(s) {T}^\theta(t) {\Delta}^m(t) {H}^m(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) ({\Delta}^m(s))^2 ({H}^m(s))^2 \tilde{X}(s).
+\end{multline}
+
+\subsubsection*{Submatrix III}
+Denote submatrix III by $J_{\beta^g}(\theta; m, g).$ The formula for sub-matrix III is similar to that of sub-matrix II (\ref{sub_mat_2_formula}). Substituting $g$ for $m$ in this equation yields
+\begin{multline}\label{sub_mat_3_formula}
+J_{\beta^g}(\theta; m, g) = \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) \left[{\Delta}^g(s) {V}^g(s) {\Delta}^g(s) - [{\Delta}']^g(s) {H}^g(s) \right] \tilde{X}(s) \\ + \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T {T}^\theta(s) {\Delta}^g(s) {H}^g(s) {T}^\theta(t) {\Delta}^g(t) {H}^g(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) ({\Delta}^g(s))^2 ({H}^g(s))^2 \tilde{X}(s).
+\end{multline}
+
+\subsubsection*{Submatrix IV}
+Denote sub-matrix IV by $J_{(\beta^g, \beta^m)}(\theta; m, g)$. The formula for $J_{(\beta^g, \beta^m)}(\theta; m, g)$ is 
+\begin{multline}\label{sub_mat_4}
+J_{(\beta^g, \beta^m)}(\theta; m,g) = \E \left[-\nabla_{\beta^g} \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] \\ + \E\left[ \nabla_{\beta^{g}}\mathcal{L}(\theta ; m,g,p) \right] \E \left[\nabla_{\beta^m}\mathcal{L} (\theta ; m,g,p)  \right]^T - \E \left[ \nabla_{\beta^{g}}\mathcal{L} (\theta; m,g,p) \nabla_{\beta^m}\mathcal{L}(\theta; m,g,p)^T  \right].
+\end{multline}
+First, we have that
+\begin{equation}\label{sub_mat_4_1}
+\E\left[-\nabla_{\beta^g} \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] = 0,
+\end{equation}
+as differentiating $\mathcal{L}$ with with respect to $\beta^g$ yields a vector that is a function of $\beta^g$, and differentiating this vector with respect to $\beta^m$ yields $0$.
+Next, recall from GLM theory that $\nabla_{\beta^g} \mathcal{L}(\theta; m, g, p) =   \tilde{X}^T\Delta^g s^g$ and $\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) = \tilde{X}^T \Delta^m s^m.$ The $(k,l)$th entry of the last two terms of (\ref{sub_mat_4}) is
+\begin{multline}\label{sub_mat_4_2}
+\bigg[ \E \left[\nabla_{\beta^g} \mathcal{L}(\theta; m, g, p) \right] \E \left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T \\ - \E \left[ \nabla_{\beta^g} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T \right] \bigg][k,l] \\
+= \left[ \E \left[ \tilde{X}^T \Delta^g s^g \right]\E \left[ \tilde{X}^T \Delta^m s^m \right]^T\right][k,l] - \E \left[ \tilde{X}^T \Delta^g s^g (s^m)^T \Delta^m \tilde{X} \right][k,l] \\ 
+= \E\left[\tilde{X}[,k]^T \Delta^g s^g \right] \E \left[\tilde{X}[,l]^T \Delta^m s^m \right] - \E \left[\tilde{X}[,k]^T \Delta^g s^g (s^m)^T \Delta^m \tilde{X}[,l ] \right] \\
+=\E\left( \sum_{i=1}^n \tilde{x}_{ik} \Delta^g_i s^g_i \right) \E \left( \sum_{j=1}^n \tilde{x}_{jl} \Delta^m_j s^m_j \right) - \E \left( \sum_{i=1}^n \sum_{j=1}^n \tilde{x}_{ik} \Delta^g_i s^g_i s^m_j \Delta^m_j \tilde{x}_{jl} \right) \\
+= \sum_{i=1}^n \sum_{j=1}^n \E[\tilde{x}_{ik} \Delta^g_is^g_i] \E[ \tilde{x}_{jl} \Delta^m_j s^m_j ] - \sum_{i=1}^n \sum_{j=1}^n \E[ \tilde{x}_{ik} \Delta^g_i s^g_i s^m_j \Delta^m_j \tilde{x}_{jl}]  \\
+= \sum_{i=1}^n \sum_{j=1}^n \E[ \tilde{x}_{ik} \Delta^g_i s^g_i] \E \left[\tilde{x}_{jl} \Delta^m_j s^m_j \right]  - \sum_{i \neq j} \E[\tilde{x}_{ik} \Delta^g_i s^g_i] \E[\tilde{x}_{jl}\Delta^m_j  s^m_j ] \\ - \sum_{i=1}^n \E[\tilde{x}_{ik} \Delta^g_i s^g_i s^m_i \Delta^m_i \tilde{x}_{il}] \\
+= \sum_{i=1}^n \E[\tilde{x}_{ik} \Delta^g_i H^g_i] \E[\tilde{x}_{il} \Delta_i^m H^m_i] - \sum_{i=1}^n \E[\tilde{x}_{ik} H_i^g \Delta_i^g \Delta_i^m H_i^m \tilde{x}_{il}] \\ 
+= \sum_{i=1}^n \left[\tilde{x}_{ik}(0) {\Delta}^g_i(0) T^\theta_i(0) {H}^g_i(0) + \tilde{x}_{ik}(1) {\Delta}^g_i(1) T^\theta_i(1) {H}^g_i(1)\right] \\ 
+\cdot \left[\tilde{x}_{il}(0) {\Delta}^m_i(0) T^\theta_i(0) {H}^m_i(0) + \tilde{x}_{il}(1) {\Delta}^m_i(1) T^\theta_i(1) {H}^m_i(1)\right] 
+\\ - \sum_{i=1}^n [\tilde{x}_{ik}(0) T^\theta_i(0) {\Delta}^g_i(0) {H}^g_i(0) {\Delta}^m_i(0) {H}^m_i(0) \tilde{x}_{il}(0) \\ + \tilde{x}_{ik}(1) T^\theta_i(1) {\Delta}^g_i(1) {H}^g_i(1) {\Delta}^m_i(1) {H}^m_i(1) \tilde{x}_{il}(1) ]
+\\ = \sum_{s=0}^1 \sum_{t=0}^1 \left[\sum_{i=1}^n \tilde{x}_{ik}(s) T^\theta_i(s) {\Delta}^g_i(s) {H}^g_i(s) T^\theta_i(t){\Delta}^m_i(t) {H}^m_i(t) \tilde{x}_{il}(t) \right]
+\\ - \sum_{s=0}^1 \left[\sum_{i=1}^n \tilde{x}_{ik}(s) T^\theta_i(s) {\Delta}^g_i(s) {H}^g_i(s) {\Delta}^m_i(s) {H}^m_i(s) \tilde{x}_{il}(s)\right] 
+\\ = \sum_{s=0}^1 \sum_{t=0}^1 \left[ \tilde{X}(s)[,k]^T T^\theta(s) {\Delta}^g(s) {H}^g(s) T^\theta(t){\Delta}^m(t) {H}^m(t) \tilde{X}(t)[,l] \right]\\ - \sum_{s=0}^1 \left[ \tilde{X}[,k]^T T^\theta(s) {\Delta}^g(s) {H}^g(s) {\Delta}^m(s) {H}^m(s) \tilde{X}[,l](s)\right].
+\end{multline}
+Combining (\ref{sub_mat_4}), (\ref{sub_mat_4_1}), and (\ref{sub_mat_4_2}) produces
+\begin{multline}\label{sub_mat_4_formula}
+J_{(\beta^g, \beta^m)}(\theta; m, g) = \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T  T^\theta(s) {\Delta}^g(s) {H}^g(s) T^\theta(t){\Delta}^m(t) {H}^m(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T T^\theta(s) {\Delta}^g(s) {H}^g(s) {\Delta}^m(s) {H}^m(s) \tilde{X}(s).
+\end{multline}
+\subsubsection*{Submatrix V}
+Denote submatrix V by $J_{(\beta^m,\pi)}(\theta; m, g).$ The formula for $J_{(\beta^m,\pi)}(\theta; m, g)$ is
+\begin{multline}\label{sub_mat_5}
+J_{(\beta^m,\pi)}(\theta; m, g) = \E \left[ - \nabla_{\beta^m} \nabla_{ \pi } \mathcal{L}(\theta; m, g, p) \right] \\ + \E\left[ \nabla_{\beta^m}\mathcal{L}(\theta ; m,g,p) \right] \E \left[ \nabla_{\pi}\mathcal{L}(\theta ; m,g,p) \right]^T  - \E \left[ \nabla_{\beta^m}\mathcal{L}(\theta; m,g,p) \nabla_{\pi}\mathcal{L}(\theta; m,g,p)^T \right].
+\end{multline}
+We have that
+\begin{equation}\label{sub_mat_5_1}
+\E \left[ - \nabla_{\beta^m} \nabla_{ \pi } \mathcal{L}(\theta; m, g, p) \right] = 0,
+\end{equation}
+as $\beta^m$ and $\pi$ separate in the log likelihood. Next, set $a := 1/\pi + 1/(1 - \pi)$ and $b := n/(1 - \pi).$ Recall from GLM theory that
+$\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) = \tilde{X}^T \Delta^m s^m$ and from (\ref{d_L_d_pi}) that
+$a \sum_{i=1}^n p_i - b.$
+The $k$th entry of the last two terms of (\ref{sub_mat_5}) is
+\begin{multline}\label{sub_mat_5_2}
+\E \left[\nabla_\pi \mathcal{L}(\theta; m, g, p) \right] \E\left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)[k] \right] - \E \left[\nabla_{\pi}\mathcal{L}(\theta; m,g,p) \nabla_{\beta^m}\mathcal{L}(\theta; m,g,p)[k] \right] \\= \left(\E \left[ a \sum_{i=1}^n p_i - b \right] \right) \left(\E\left[ \tilde{X}[,k]^T \Delta^m s^m \right] \right) - \E \left[ \left( a \sum_{i=1}^n p_i - b \right) \tilde{X}[,k]^T \Delta^m s^m \right] \\ = \left( a \sum_{i=1}^n \E[p_i] - b \right) \left( \sum_{j=1}^n \E [ \tilde{x}_{jk}\Delta^m_js^m_j] \right) - \E \left[ \left( a \sum_{i=1}^n p_i - b \right) \left( \sum_{j=1}^n \tilde{x}_{jk} \Delta^m_j s^m_j \right) \right] \\ = a \sum_{i=1}^n \sum_{j=1}^n \E [p_i] \E[ \tilde{x}_{jk} \Delta^m_j s^m_j] - b \sum_{j=1}^n \E[\tilde{x}_{jk} \Delta^m_j s^m_j] \\ - \left[ a \sum_{i=1}^n \sum_{j=1}^n \E [ p_i \tilde{x}_{jk} \Delta^m_j s^m_j] - b \sum_{j=1}^n \E[\tilde{x}_{jk} \Delta^m_j s^m_j] \right] \\ =  a \sum_{i=1}^n \sum_{j=1}^n \E[p_i] \E[\tilde{x}_{jk} \Delta^m_j s^m_j] - a\sum_{i \neq j} \E[p_i] \E[\tilde{x}_{jk} \Delta^m_j s^m_j] - a\sum_{i=1}^n \E[ p_i \tilde{x}_{ik} \Delta^m_i s^m_i ] \\ = a \sum_{i=1}^n \E[p_i] \E[ \tilde{x}_{ik} \Delta^m_i s^m_i] - a \sum_{i=1}^n \E[p_i \tilde{x}_{ik} \Delta^m_i s^m_i] \\ = a \sum_{i=1}^n T^\theta_i(1) [T^\theta_i(0) \Delta^m_i(0) s^m_i(0) \tilde{x}_{ik}(0) + T^\theta_i(1) \Delta^m_i(1) s^m_i(1) \tilde{x}_{ik}(1)] \\ - a \sum_{i=1}^n T^\theta_i(1)\Delta^m_i(1)s^m_i(1)\tilde{x}_{ik}(1) \\ = a \sum_{i=1}^n T^\theta_i(0)T^\theta_i(1) \Delta_i^m(0)H^m_i(0)\tilde{x}_{ik}(0) \\ + a \sum_{i=1}^n \left( [T^\theta_i(1)]^2 \Delta^m_i(1)H^m_i(1) - T^\theta_i(1)\Delta^m_i(1)H^m_i(1) \right) \tilde{x}_{ik}(1)  \\ =a \left[ \sum_{i=1}^n T^\theta_i(0) T^\theta_i(1) \Delta^m_i(0) H^m_i(0) \tilde{x}_{ik}(0) + \sum_{i=1}^n T^\theta_i(1)\Delta^m_i(1)H^m_i(1)[T^\theta_i(1) - 1] \tilde{x}_{ik}(1) \right] \\ = a \left[ \sum_{i=1}^n T^\theta_i(0) T^\theta_i(1) \Delta^m_i(0) H^m_i(0) \tilde{x}_{ik}(0) - \sum_{i=1}^n T^\theta_i(0) T^\theta_i(1) \Delta^m_i(1) H^m_i(1) \tilde{x}_{ik}(1) \right] \\ = a\left(\tilde{X}(0)[,k]^T w^m(0) - \tilde{X}(1)[,k]^T w^m(1)  \right).
+\end{multline}
+Combining (\ref{sub_mat_5}), (\ref{sub_mat_5_1}), and (\ref{sub_mat_5_2}), we conclude that
+\begin{equation}\label{sub_mat_5_formula} J_{(\beta^m, \pi)}(\theta; m, g, p) = \left( \frac{1}{\pi} + \frac{1}{1 - \pi} \right) \left( \tilde{X}(0)^T w^m(0) - \tilde{X}(1)^T w^m(1)\right).
+\end{equation}
+
+\subsubsection*{Submatrix VI}
+Denote submatrix VI by $J_{(\beta^g,\pi)}(\theta; m, g).$ Calculations similar to those for submatrix V show that
+\begin{equation}\label{sub_mat_6_formula} J_{(\beta^g, \pi)}(\theta; m, g, p) = \left(\frac{1}{\pi} + \frac{1}{1 - \pi} \right) \left( \tilde{X}(0)^T w^g(0) - \tilde{X}(1)^T w^g(1)\right).
+\end{equation}
+
+\subsubsection*{Combining submatrices}
+To summarize, the formulas for submatrices I-VI are as follows:
+\begin{itemize}
+\item[I]\begin{multline*}
+J_\pi(\theta; m, g) = 
+\left[ \frac{1}{\pi^2} - \frac{1}{(1 - \pi )^2} \right] \sum_{i=1}^n T^\theta_i(1) + \frac{n}{(1-\pi )^2} \\ + \left( \frac{1}{(1 - \pi )} + \frac{1}{\pi} \right)^2 \left( \sum_{i=1}^n [T^\theta_i(1)]^2 - T^\theta_i(1) \right).
+\end{multline*}
+\item[II] \begin{multline*}
+J_{\beta^m}(\theta; m, g) = \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) \left[ {\Delta}^m(s) {V}^m(s) {\Delta}^m(s) - [{\Delta}']^m(s) {H}^m(s) \right] \tilde{X}(s) \\ + \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T {T}^\theta(s) {\Delta}^m(s) {H}^m(s) {T}^\theta(t) {\Delta}^m(t) {H}^m(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) ({\Delta}^m(s))^2 ({H}^m(s))^2 \tilde{X}(s).
+\end{multline*}
+\item[III] \begin{multline*}
+J_{\beta^g}(\theta; m, g) = \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) \left[{\Delta}^g(s) {V}^g(s) {\Delta}^g(s) - [{\Delta}']^g(s) {H}^g(s) \right] \tilde{X}(s) \\ + \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T {T}^\theta(s) {\Delta}^g(s) {H}^g(s) {T}^\theta(t) {\Delta}^g(t) {H}^g(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) ({\Delta}^g(s))^2 ({H}^g(s))^2 \tilde{X}(s).
+\end{multline*}
+\item[IV] \begin{multline*}
+J_{(\beta^g, \beta^m)}(\theta; m, g) = \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T  T^\theta(s) {\Delta}^g(s) {H}^g(s) T^\theta(t){\Delta}^m(t) {H}^m(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T T^\theta(s) {\Delta}^g(s) {H}^g(s) {\Delta}^m(s) {H}^m(s) \tilde{X}(s).
+\end{multline*}
+\item[V] $$ J_{(\beta^m, \pi)}(\theta; m, g, p) = \left( \frac{1}{\pi} + \frac{1}{1 - \pi} \right) \left( \tilde{X}(0)^T w^m(0) - \tilde{X}(1)^T w^m(1)\right).  $$
+\item[VI] $$ J_{(\beta^g, \pi)}(\theta; m, g, p) = \left(\frac{1}{\pi} + \frac{1}{1 - \pi} \right) \left( \tilde{X}(0)^T w^g(0) - \tilde{X}(1)^T w^g(1)\right).$$
+\end{itemize}
+We stitch these pieces together and transpose submatrices IV, V, and VI to produce the whole information matrix $J(\theta; m, g)$.
+Evaluating this matrix at the EM estimate $\theta^\textrm{EM}$ and inverting yields the asymptotic covariance matrix, which we can use to compute standard errors.
+
+\subsection{Implementation}
+To evaluate the observed information matrix, we need to compute the matrices $\Delta^m(j),$ $[\Delta']^m(j),$ $V^m(j),$ and $H^m(j)$ and the vectors $s^m(j)$ and $w^m(j)$ for $j \in \{0,1\}$.
+We likewise need to compute the analogous gRNA quantities.
+We describe how to compute these quantities in R by extending base family objects.
+We implicitly condition on $p_i$, $z^m_i$, and $o^m_i$.
+
+\begin{algorithm}
+	\caption{Computing the matrices $\Delta^m(j)$, $[\Delta']^m(j)$, $V^m(j)$, $H^m(j)$, and $s^m(j)$ given given $\beta_m$.}\label{algo:computing_info_matrices}
+	\begin{algorithmic}[3]
+		\Require A coefficient vector $\beta_m$; data $[m_1, \dots, m_n]$, $[o^m_1, \dots, o^m_n]$, and $[z_1, \dots, z_n]$; and a family object containing functions \texttt{linkinv}, \texttt{variance}, \texttt{mu.eta}, \texttt{mu.eta.prime}, and \texttt{skewness}.
+		\For{$j \in \{0, 1\}$}
+		\For{$i \in \{1, \dots, n\}$}
+		\State $l^m_i(j) \gets \langle \beta_m, \tilde{x}_i(j) \rangle + o^m_i$
+		\State $\mu^m_i(j) \gets \texttt{linkinv}(l^m_i(j))$
+		\State $[\sigma_i^m(j)]^2 \gets \texttt{variance}(\mu_i^m(j))$
+		\State $h_m'(l_i^m(j)) \gets \texttt{mu.eta}(l_i^m(j))/[\sigma_i^m(j)]^2$
+		\State $\gamma^m_i(j) \gets \texttt{skewness}(\mu^m_i(j))$
+		\State $[r_m^{-1}]''(l_i^m(j)) \gets \texttt{mu.eta.prime}(l^m_i(j))$
+		\State $$h_m''(l_i^m(j)) \gets \frac{[r^{-1}]''(l_i^m(j)) - [([\sigma_i^m(j)]^2)^{3/2}][\gamma^m_i(j)] [h_m'(l_i^m(j))]^2}{[\sigma_i^m(j)]^2}$$
+		\Comment Assign quantities to matrices
+		\State $\Delta_{i}^m(j) \gets h_m'( l_i^m(j))$
+		\State $[\Delta']^m_{i}(j) \gets h''(l^m_i(j))$
+		\State $V^m_{i}(j) \gets [\sigma^m_i(j)]^2$
+		\State $H^m_{i}(j) \gets s^m_i(j) \gets m_i - \mu^m_i(j)$
+		\EndFor
+		\EndFor
+	\end{algorithmic}
+\end{algorithm}
+
+An \texttt{R} family object contains several functions, including \texttt{linkinv}, \texttt{variance}, and \texttt{mu.eta}.
+\texttt{linkinv} is the inverse link function $r_m^{-1}$.
+\texttt{variance} takes as an argument the mean $\mu^m_i$ and returns the variance $[\sigma_i^m]^2$.
+\texttt{mu.eta} is the derivative of the inverse link function $[r^{-1}_m]^{'}$.
+We extend the \texttt{R} family object by adding two additional functions: \texttt{skewness} and \texttt{mu.eta.prime}.
+\texttt{skewness} returns the skewness $\gamma^m_i$ of the distribution as a function of the mean $\mu_i$, i.e.
+$$\texttt{skewness}\left(\mu_i\right) = \E \left[\left(\frac{m_i - \mu_i^m}{ \sigma_i^m}\right)^3\right] := \gamma_i^m.$$ Finally, the \texttt{mu.eta.prime} is the second derivative of the inverse link function $[r^{-1}_m]''.$ Algorithm \ref{algo:computing_info_matrices} computes the matrices $\Delta^m(j)$, $[\Delta']^m(j)$, $V^m(j)$, $H^m(j)$, and vector $s^m(j)$ for given $\beta_m$ and given family object.
+(The vector $w^m(j)$ can be computed in terms of $\Delta^m(j)$ and $H^m(j)$.) We use $\sigma^m_i(j)$ (resp.
+$\gamma^m_i(j)$) to refer to the standard deviation (resp.
+skewness) of the gene expression distribution the $i$th cell when the perturbation $p_i$ is set to $j$.
+
+All steps of the algorithm are obvious except the calculation of $h'_m(l^m_i(j))$ (line 6), $h''(l^m_i(j))$ (line 9), and $V^m_i(j)$ (line 12).
+We omit the $(j)$ notation for compactness.
+First, we prove the correctness of the expression for $h'_m(l^m_i)$.
+Recall the basic GLM identities
+\begin{equation}\label{computing_info_matrix_1}
+\psi_m''(\eta_i^m) =  [\sigma^m_i]^2
+\end{equation}
+ and, for all $t \in \R$, 
+\begin{equation}\label{computing_info_matrix_2}
+ r_m^{-1}(t) = \psi_m'(h_m(t)).
+\end{equation}
+ Differentiating (\ref{computing_info_matrix_2}) in $t$, we find that
+\begin{equation}\label{computing_info_matrix_3}
+ (r_m^{-1})'(t) = \psi_m''(h_m(t))h_m'(t), 
+\end{equation}
+or  $$h_m'(t) = \frac{(r_m^{-1})'(t) }{\psi_m''(h_m(t))}.$$ Finally, plugging in $l^m_i$ for $t$,
+$$ h_m'(l_i) = \frac{(r_m^{-1})'(l^m_i)}{\psi''(h_m(l^m_i))} = \frac{(r_m^{-1})'(l^m_i)}{\psi_m''(\eta^m_i)} = \textrm{ by (\ref{computing_info_matrix_1}) } \frac{(r_m^{-1})'(l_i^m)}{[\sigma_i^m]^2}.$$
+
+Next, we prove the correctness for the expression for $h_m''(l_i^m)$. Recall the exponential family identity 
+\begin{equation}\label{computing_info_matrix_4}
+\psi'''_m(\eta^m_i) = \gamma^m_i  ([\sigma^m_i]^2)^{3/2}.
+\end{equation}
+Differentiating (\ref{computing_info_matrix_3}) in $t$, we obtain
+$$ (r_m^{-1})''(t) = \psi_m'''(h_m(t)) [h_m'(t)]^2 + \psi_m''(h_m(t)) h_m''(t),$$ or $$h_m''(t) =\frac{(r_m^{-1})''(t) - \psi'''(h_m(t))[h_m'(t)]^2}{\psi_m''(h_m(t))}.$$ Plugging in $l^m_i$ for $t$, we find that
+\begin{equation*}
+h_m''(l^m_i) = \frac{(r_m^{-1})''(l^m_i) - \psi_m'''(\eta^m_i) [h_m'(l_i^m)]^2}{[\sigma_i^m]^2} = \textrm{ (by \ref{computing_info_matrix_4}) }  \frac{(r_m^{-1})''(l^m_i) - ([\sigma_i^m]^2)^{3/2} (\gamma_i^m) [h_m'(l^m_i)]^2 }{[\sigma_i^m]^2}.
+\end{equation*}
+Finally, the expression for $V^m_i$ follows from (\ref{computing_info_matrix_1}).
+We can apply a similar algorithm to compute the analogous matrices for the gRNA modality.
+Table \ref{family_object_functions} shows the \texttt{linkinv}, \texttt{variance}, \texttt{mu.eta}, \texttt{skewness}, and \texttt{mu.eta.prime} functions for several common family objects (which are defined by a distribution and link function).
+\begin{table}
+\centering
+\caption{\texttt{linkinv}, \texttt{variance}, \texttt{mu.eta}, \texttt{skewness}, \texttt{mu.eta.prime} for common family objects (i.e., pairs of distributions and link functions).}\label{family_object_functions}
+\begin{tabular}{|C{2.5cm}|C{3.5cm}|C{3.4cm}|C{2.5cm}|}
+	\hline 
+	& Gaussian response, identity link & Poisson response, log link & NB response ($\theta > 0$ fixed), log link \\ 
+	\hline 
+	\texttt{linkinv} & $x$ & $\exp(x)$ & $\exp(x)$  \\ 
+	\hline 
+	\texttt{variance} & $x$ & $x$ & $x + x^2/\theta$ \\ 
+	\hline 
+	\texttt{mu.eta} & $1$ & $x$  & $\exp(x)$ \\ 
+	\hline 
+	\texttt{skewness} & $0$ & $x^{-1/2}$ & $\frac{2 x + \theta}{\sqrt{\theta x} \sqrt{x + \theta}}$ \\ 
+	\hline 
+	\texttt{mu.eta.prime} & $0$ & $\exp(x)$ & $\exp(x)$ \\ 
+	\hline 
+\end{tabular}
+\end{table}
+
+\section{Zero-inflated model}\label{sec:zero_inf_model}
+In this section we introduce the ``zero-inflated'' GLM-EIV model.
+The zero-inflated GLM-EIV model is appropriate to use when the unperturbed cells do not transcribe \textit{any} gRNA molecules (i.e., when there are no background reads).
+Let $x_i = [1, z_i]^T \in \R^{d-1}$ be the vector of observed covariates, including an intercept term.
+($x_i$ is the same as $\tilde{x}_i$, but with the perturbation indicator $p_i$ removed.) Let $\beta_{g,z} = [\beta^g_0, \gamma_g] \in \R^{d-1}$ be an unknown coefficient vector.
+($\beta_{g,z}$ is the same as $\beta_g$, but with the perturbation effect $\beta^g_1$ removed).
+Let the linear component $l^{g,z}_i$, mean $\mu^{g,z}_i$, and canonical parameter $\eta^{g,z}_i$ of gRNA count distribution of the $i$th cell be given by
+$$
+l^{g,z}_i = \langle x_i, \beta_{g,z} \rangle + o^g_i; \quad
+r_g(\mu^{g,z}_i) = l^{g,z}_i; \quad
+\eta^{g,z}_i = ([\psi'_g]^{-1} \circ r^{-1}_g)(l^{g,z}_i) := h_g(l^{g,z}_i).
+$$
+The density $f_{g,z}$ of gRNA counts in the zero-inflated model is as follows:
+$$f_{g,z}(g_i; \eta^{g,z}_i, p_i) = [f_g(g_i;\eta^{g,z}_i)]^{p_i} \mathbb{I}(g_i = 0)^{1-p_i}.$$
+In other words, when the cell is \textit{perturbed} (i.e., $p_i = 1$), the zero-inflated density $f_{g,z}$ coincides with the background-read density $f_g$; by contrast, when the cell is \textit{unperturbed} (i.e., $p_i = 0$), the zero-inflated density $f_{g,z}$ is a point mass at zero.
+The gene expression density $f_m$ and perturbation indicator density $f_p$ are the same across the background read and zero-inflated models.
+We assume that the gene expression $m_i$ and gRNA count $g_i$ are conditionally independent given the perturbation indicator $p_i$.
+The joint density $f_z$ of $(m_i, p_i, z_i)$ is
+\begin{equation*}
+f_z(m_i, g_i, p_i) = f_m(m_i | p_i) f_{g,z}(g_i | p_i) f_p(p_i) = \pi^{p_i} (1-\pi)^{1-p_i} f_m(m_i; \eta^m_i)[f_g(g_i; \eta_i^{g,z})]^{p_i} \mathbb{I}(g_i = 0)^{1-p_i}.
+\end{equation*}
+The complete-data log-likelihood $\mathcal{L}_z$ is
+\begin{multline*}
+\mathcal{L}_z(\theta; m, g, p) = \sum_{i=1}^n \log \left[ \pi^{p_i} (1-\pi)^{1-p_i} \right] + \sum_{i=1}^n \log\left[f_m(m_i; \eta^m_i) \right] \\ + \sum_{i=1}^n p_i \log \left[f_{g}(g_i; \eta_i^{g,z}) \right] + \sum_{i=1}^n (1 - p_i)\log\left[ \mathbb{I}(g_i = 0) \right],
+\end{multline*}
+where $\theta = [\pi, \beta_m, \beta_{g,z}]$ is the vector of unknown parameters.
+Integrating over the unobserved variable $p_i$, the marginal $f_z$ of $(m_i, g_i)$ is
+$$f_z(m_i, g_i; \theta) = (1-\pi) f_m(m_i;\eta^m_i(0)) \mathbb{I}(g_i = 0) + \pi f_m(m_i; \eta^m_i(1)) f_g(g_i;\eta^{g,z}_i).$$ Finally, the log-likelihood is
+$$
+\mathcal{L}_z(\theta; m_i, g_i) = \sum_{i=1}^n \log\left[ (1-\pi) f_m(m_i;\eta^m_i(0)) \mathbb{I}(g_i = 0) + \pi f_m(m_i; \eta^m_i(1)) f_g(g_i;\eta^{g,z}_i) \right].
+$$
+
+\subsection{Estimation}
+To estimate the parameters of the zero-inflated GLM-EIV model, we use an EM algorithm similar to Algorithm \ref{algo:em_full} but with two changes.
+First, we use a different formula for the $i$th membership probability at the $t$-th step of the algorithm $T^{(t)}_i(1)$.
+(We use $T^{(t)}_i(1)$ to denote the $i$th membership probability in \textit{both} the background read and zero inflated cases; the difference should be clear from context.) Let $\theta^{(t)} = (\pi^{(t)}, \beta^{(t)}_m, \beta^{(t)}_{g,z})$ be the parameter estimate at the $t$-th iteration of the algorithm.
+Arguing in a manner similar to the background read case, we have that $$T^{(t)}_i(1) = \frac{1}{ \exp(q_i^{\left(t,z\right)}) + 1},$$ where
+$$q_i^{(t,z)} = \log \left(\frac{(1 - \pi^{(t)}) \P(M_i = m_i | P_i = 0, \theta^{(t)}) \P(G_i = g_i | P_i = 0, \theta^{(t)})}{(\pi^{(t)}) \P(M_i = m_i | P_i = 1, \theta^{(t)}) \P(G_i = g_i | P_i = 1, \theta^{(t)})}\right).$$
+The expression for $q^{(t,z)}_i$ is
+\begin{multline*}
+q^{(t,z)}_i = \log\left[ 1 - \pi^{(t)} \right] + \log\left[ f_m\left(m_i; \left[ \eta^m_i(0) \right]^{(t)}\right) \right] + \log\left[ \mathbb{I}(g_i = 0) \right] \\ - \log\left[ \pi^{(t)} \right] - \log\left[ f_m\left(m_i; \left[ \eta^m_i(1) \right]^{(t)}\right) \right] - \log\left[ f_g\left(g_i; \left[\eta^{g,z}_i \right]^{(t)}\right) \right],
+\end{multline*}
+where $[\eta^{g,z}_i]^{(t)} = h_g( \langle x_i, \beta^{(t)}_{g,z} \rangle + o^g_i).$ Notice that if $g_i \geq 1$, then $T^{(t)}_i(1) = 1.$ This comports with our intuition that a nonzero gRNA count indicates the presence of a perturbation.
+
+Next, we consider the M step of the EM algorithm, which is similar to the background read case but slightly different.
+Define $Q_z(\theta | \theta^{(t)}) = \E_{(P | M = m, G = g, \theta^{(t)})} \left[ \mathcal{L}_z(\theta; m, g, p) \right].$ We have that
+\begin{multline}\label{q_funct_zero_inf}
+Q_z(\theta | \theta^{(t)}) = \sum_{i=1}^n \left[ T^{(t)}_i(1)\log(\pi) + T_i^{(t)}(0) \log(1 - \pi)\right] + \sum_{i=1}^n \sum_{j=0}^1 T^{(t)}_i(j) \log \left[ f_m( m_i; \eta_i^m(j)) \right] \\ + \sum_{i=1}^n T_i^{(t)}(1) \left[ \log( f_g(g_i; \eta_i^{g, z}))\right] + C.
+\end{multline}
+The three terms of (\ref{q_funct_zero_inf}) are functions of $\pi$, $\beta_m$, and $\beta_{g,z}$, respectively.
+The maximizer $\pi^{(t)}$ and $\beta_m^{(t+1)}$ of the first and second term are the same as in the background read case.
+The maximizer $\beta^{(t+1)}_{g,z}$ of the third term is the maximizer of the GLM with exponential family density $f_g$, link function $r_g$, responses $g$, weights $T^{(t)}(1)$, design matrix $X$, offsets $o^g$.
+
+\subsection{Inference}
+Next, we derive the asymptotic observed information matrix for the zero-inflated model, allowing us to perform inference.
+Again, let $T^{\theta}(1) := \textrm{diag}\{T^{\theta}_1(1), \dots, T^{\theta}_n(1)\},$ but note that $T^\theta_i(1) = \P(P_i = 1 | G_i = g_i, M_i = m_i, \theta)$ is computed differently than in the background read case.
+Define the $n \times n$ matrices $\Delta^{(g,z)}, [\Delta']^{(g,z)}, V^{(g,z)},$ and $H^{(g,z)}$ by
+$$
+\begin{cases}
+\Delta^{(g,z)} = \textrm{diag} \{ h_g'(l_1^{g,z} ), \dots, h_g'( l_n^{g,z}  ) \} \\
+[\Delta']^{(g,z)}  = \textrm{diag} \{ h_g''(l_1^{g,z} ), \dots, h_g''( l_n^{g,z} ) \} \\
+V^{(g,z)}  = \textrm{diag} \{ \psi_g( \eta^{g,z} _1), \dots, \psi_g( \eta^{g,z}_n) \} \\
+H^{(g,z)}  = \textrm{diag} \{ m_1 - \mu_1^{g,z} , \dots, m_n - \mu_n^{g,z} \} .
+\end{cases}
+$$
+Also, define the $\R^n$ vectors $s^{(g,z)}$ and $w^{(g,z)}$ by
+$$s^{(g,z)} = [g_1 - \mu_1^{g,z}, \dots, g_n - \mu_n^{g,z}]^T,$$ and $$w^{(g,z)} = [T^\theta_1(0)T^\theta_1(1) \Delta_1^{(g,z)}H^{(g,z)}_1,\dots, T^\theta_n(0)T^\theta_n(1) \Delta_n^{(g,z)}H^{(g,z)}_n].$$ These quantities are computable, as they do not depend on the unobserved variables $p_1, \dots, p_n$.
+Finally, let the unobserved, $n\times n$ matrix $P$ be defined by $P = \textrm{diag}\{ p_1, \dots, p_n \}$.
+
+The observed information matrix $J_z(\theta; m, g)$ is given by
+$J_z(\theta; m, g) = -\nabla^2 \mathcal{L}_z(\theta; m, g).$ Louis's theorem implies that
+\begin{multline*}
+J_z(\theta; m, g) = -\E \left[\nabla^2 \mathcal{L}_z(\theta; m, g, p) | G = g, M = m \right] \\ + \E\left[\nabla \mathcal{L}_z(\theta; m, g, p) | G = g, M = m \right] \E\left[\nabla \mathcal{L}_z(\theta; m, g, p) | G = g, M = m \right]^T \\ - \E\left[\nabla\mathcal{L}_z(\theta; m, g, p) \nabla \mathcal{L}_z(\theta; m, g, p)^T | G = g, M = m \right].
+\end{multline*}
+The matrix $J_z(\theta; m, g)$ has dimension $d \times d$ and consists of nine submatrices (Figure \ref{infomatrixzeroinf}).
+Three of these submatrices (i.e., I, II, and V) are the same as the corresponding submatrices in the background read case.
+We therefore must compute the remaining submatrices (i.e., III, IV, and VI) to compute the entire matrix $J_z(\theta; m, g)$.
+Again, in the following, all expectations are understood to be conditional on $m$ and $g$.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.45\linewidth]{../../figures/info_matrix/info_matrix_zero_inf}
+	\caption{Block structure of the observed information matrix $J_z(\theta; m, g) = -\nabla^2 \mathcal{L}_z(\theta; m, g)$ for the zero-inflated model.
+Submatrices I, II, and VI are the same as in the background read model; therefore, we only need to compute submatrices III, VI, and V.}
+	\label{infomatrixzeroinf}
+\end{figure}
+
+\subsubsection*{Submatrix III (zero-inflated)}
+Denote submatrix III by $J_{\beta_{(g,z)}}(\theta; m, g)$ The formula for $J_{\beta_{(g,z)}}(\theta; m, g)$ is
+\begin{multline}\label{sub_mat_3_zeroinf}
+J_{\beta_{(g,z)}}(\theta; m, g) = -\E \left[\nabla_{\beta_{(g,z)}}^2 \mathcal{L}_z(\theta; m, g, p) \right] + \E\left[ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right] \E\left[\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right]^T \\ - \E\left[ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p)^T \right].
+\end{multline}
+GLM theory indicates that
+$$ -\nabla_{\beta_{(g,z)}}^2 \mathcal{L}_z(\theta; m, g, p) =  X^T P ( \Delta^{(g,z)}V^{(g,z)}\Delta^{(g,z)} - (\Delta')^{(g,z)} H^{(g,z)}) X$$ and
+$$ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) = X^T P \Delta^{(g,z)} s^{(g,z)}.$$
+We begin by computing the first term of (\ref{sub_mat_3_zeroinf}).
+The only random matrix among $X,$ $P$, $\Delta^{(g,z)}$, $V^{(g,z)}$, $(\Delta')^{(g,z)},$ and $H^{(g,z)}$ is $P$.
+Therefore, by the linearity of expectation,
+\begin{multline}\label{sub_mat_3_zeroinf_1}
+-\E \left[\nabla_{\beta_{(g,z)}}^2 \mathcal{L}_z(\theta; m, g, p) \right]   = \E \left[ X^T P (\Delta^{(g,z)}V^{(g,z)}\Delta^{(g,z)} - (\Delta')^{(g,z)} H^{(g,z)} \right] \\ = X^T T^{\theta}(1) ( \Delta^{(g,z)}V^{(g,z)}\Delta^{(g,z)} - (\Delta')^{(g,z)} H^{(g,z)}) X.
+\end{multline}
+Next, we compute the difference of the last two terms of (\ref{sub_mat_3_zeroinf}). The $(k,l)$th entry of this matrix is
+\begin{multline*}
+\bigg[ \E \left[\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right] \E \left[\nabla_{\beta{(g,z)}} \mathcal{L}_z(\theta; m, g, p)\right]^T \\ - \E \left[ \nabla_{{\beta_{(g,z)}}} \mathcal{L}_z(\theta; m, g, p) \nabla_{{\beta_{(g,z)}}} \mathcal{L}_z(\theta; m, g, p)^T \right] \bigg] [k,l] \\
+= \left[ \E \left[ {X}^T P \Delta^{(g,z)} s^{(g,z)} \right]\E \left[ {X}^T P \Delta^{(g,z)} s^{(g,z)} \right]^T\right][k,l] - \E \left[ {X}^T P \Delta s^{(g,z)} (s^{(g,z)})^T \Delta^{(g,z)} P {X}^T \right][k,l] \\
+= \E\left[{X}[,k]^T P \Delta^{(g,z)} s^{(g,z)} \right] \E \left[ {X}[,l]^T P \Delta^{(g,z)} s^{(g,z)} \right] - \E \left[ {X}[,k]^T P \Delta^{(g,z)} s^{(g,z)} (s^{(g,z)})^T \Delta^{(g,z)} P {X}[,l ] \right] \\
+=\E\left(\sum_{i=1}^n {x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_{i} \right) \E \left( \sum_{j=1}^n {x}_{jl} P_j \Delta^{(g,z)}_j s^{(g,z)}_j \right) \\ - \E \left( \sum_{i=1}^n \sum_{j=1}^n {x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i s^{(g,z)}_j \Delta^{(g,z)}_j P_j{x}_{jl} \right) \\
+= \sum_{i=1}^n \sum_{j=1}^n \E[ {x}_{ik} P_i \Delta^{(g,z)}_is^{(g,z)}_i] \E [ {x}_{jl} P_j \Delta^{(g,z)}_j s^{(g,z)}_j]  -  \sum_{i=1}^n \sum_{j=1}^n \E [ {x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i s^{(g,z)}_j \Delta^{(g,z)}_jP_j {x}_{jl}  ]  \\
+= \sum_{i=1}^n \sum_{j=1}^n \E[ {x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i ] \E \left[{x}_{jl} P_j \Delta^{(g,z)}_j s^{(g,z)}_j \right]  - \sum_{i \neq j} \E [{x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i] \E[s^{(g,z)}_j P_j \Delta^{(g,z)}_j {x}_{jl}] \\ - \sum_{i=1}^n \E[ {x}_{ik}P_i \Delta^{(g,z)}_i s^{(g,z)}_i s^{(g,z)}_i \Delta^{(g,z)}_i P_i{x}_{il}] \\
+= \sum_{i=1}^n \E[ {x}_{ik} P_i \Delta^{(g,z)}_i H^{(g,z)}_i] \E[{x}_{il}P_i\Delta^{(g,z)}_i H^{(g,z)}_i] - \sum_{i=1}^n \E[{x}_{ik}P_i^2(\Delta^{(g,z)}_i)^2 (H^{(g,z)}_i)^2 {x}_{il}] \\ = \sum_{i=1}^n {x}_{ik} T_i^{\theta}(1)^2(\Delta^{(g,z)}_i)^2 (H_i^{(g,z)})^2{x}_{il} - \sum_{i=1}^n {x}_{ik} T_i^{\theta}(1) (\Delta_i^{(g,z)})^2 (H_i^{(g,z)})^2 {x}_{il} \\ = {X}[,k]^T T^{\theta}(1)^2 (\Delta^{(g,z)})^2 (H^{(g,z)})^2 {X}[,l] - {X}[,k]^T T^{\theta}(1) (\Delta^{(g,z)})^2 (H^{(g,z)})^2 {X}[,l]
+\end{multline*}
+Therefore, we have that
+\begin{multline}\label{sub_mat_3_zeroinf_2}
+\E\left[ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right] \E\left[\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right]^T - \E\left[ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p)^T \right] \\ =
+X^T T^{\theta}(1)^2 (\Delta^{(g,z)})^2(H^{(g,z)})^2{X} - {X}^T T^{\theta}(1) (\Delta^{(g,z)})^2 (H^{(g,z)})^2 {X} \\ = - X^T T^{\theta}(1) \left(\Delta^{(g,z)}\right)^2 \left( H^{(g,z)} \right)^2 \left( I - T^{\theta}(1)  \right) X.
+\end{multline}
+Combining (\ref{sub_mat_3_zeroinf}), (\ref{sub_mat_3_zeroinf_1}), and (\ref{sub_mat_3_zeroinf_2}), we conclude that
+\begin{multline}\label{sub_mat_3_zeroinf_formula}
+J_{\beta_{(g, z)}} = \left( \theta; m, g \right) =  X^T T^{\theta}(1) ( \Delta^{(g,z)}V^{(g,z)}\Delta^{(g,z)} - (\Delta')^{(g,z)} H^{(g,z)}) X \\ - X^T T^{\theta}(1) \left(\Delta^{(g,z)}\right)^2 \left( H^{(g,z)} \right)^2 \left( I - T^{\theta}(1) \right) X.
+\end{multline}
+
+\subsubsection*{Submatrix IV (zero-inflated)}
+Denote submatrix IV by $J_{(\beta_{(g,z)}, \beta_m)}(\theta; m, g)$. The formula for submatrix IV is
+\begin{multline}\label{sub_mat_4_zeroinf}
+J_{(\beta_{(g,z)}, \beta_m)}(\theta; m, g) = - \E \left[\nabla_{\beta_{(g,z)}} \nabla_{\beta_m} \mathcal{L}_z (\theta; m, g, p) \right] \\ + \E\left[ \nabla_{\beta_{(g,z)} }\mathcal{L}_z(\theta ; m,g,p) \right]\E\left[\nabla_{\beta_m}\mathcal{L}_z(\theta; m,g,p)\right]^T - \E \left[\nabla_{\beta_{(g,z)}}\mathcal{L}_z (\theta; m,g,p) \nabla_{\beta_m}\mathcal{L}_z(\theta; m,g,p)]^T \right].
+\end{multline} 
+First, we have that
+\begin{equation}\label{sub_mat_4_zeroinf_1}
+- \E \left[\nabla_{\beta_{(g,z)}} \nabla_{\beta_m} \mathcal{L}_z (\theta; m, g, p) \right] = 0,
+\end{equation}
+ as the derivative in $\beta_m$ of $\mathcal{L}_z(\theta; m, g, p)$ is a function of $\beta_m$, and the derivative in $\beta_{(g,z)}$ of this term is $0$.
+Next, we compute the difference of the last two terms of (\ref{sub_mat_4_zeroinf}).
+Entry $(k,l)$ of this matrix is
+\begin{multline}\label{sub_mat_4_zeroinf_2}
+[\E[\nabla_{\beta_{(g,z)}}\mathcal{L}_z(\theta; m, g, p) ] \E [\nabla_{\beta_m} \mathcal{L}_z(\theta; m, g, p) ]^T \\ - \E [\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \nabla_{\beta_m}\mathcal{L}_z(\theta; m, g, p)^T ] ][k,l] \\
+= \left[ \E \left[ {X}^T P \Delta^{(g,z)} s^{(g,z)} \right] \E\left[\tilde{X}^T \Delta^m s^m\right]^T\right][k,l] - \E \left[{X}^T P \Delta^{(g,z)} s^{(g,z)} (s^m)^T \Delta^m \tilde{X} \right][k,l] \\ = \left[\E\left[{X}[,k]^T P \Delta^{(g,z)} s^{(g,z)}\right]\E\left[\tilde{X}[,l]^T \Delta^m s^m\right]^T\right] - \E \left[{X}[,k]^T P \Delta^{(g,z)} s^{(g,z)} (s^m)^T \Delta^m \tilde{X}[,l] \right] \\ = \E\left(\sum_{i=1}^n x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i\right) \E \left(\sum_{j=1}^n \tilde{x}_{jl} \Delta^m_j s^m_j \right) - \E \left(\sum_{i=1}^n \sum_{j=1}^n x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i \Delta^m_j s^m_j \tilde{x}_{jl}  \right) \\
+= \sum_{i=1}^n \sum_{j=1}^n \E [x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i] \E[\Delta^m_js^m_j \tilde{x}_{jl}] - \sum_{i=1}^n \sum_{j=1}^n \E[x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i \Delta^m_j s^m_j \tilde{x}_{jl}] \\
+= \sum_{i=1}^n \sum_{j=1}^n \E [x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i ] \E[\Delta^m_js^m_j \tilde{x}_{jl}] - \sum_{i \neq j} \E [x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i] \E[\Delta^m_js^m_j \tilde{x}_{jl}] \\ - \sum_{i=1}^n \E[x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i \Delta^m_j s^m_j \tilde{x}_{jl}] \\
+= \sum_{i=1}^n \E[x_{ik} P_i \Delta_i^{(g,z)} H^{(g,z)}_i] \E[\tilde{x}_{il} \Delta^m_i H^m_i] - \sum_{i=1}^n \E[x_{ik} P_i \Delta_i^{(g,z)} H_i^{(g,z)} \Delta_i^m H_i^m \tilde{x}_{il}] \\ 
+= \sum_{i=1}^n \left[x_{ik} T_i^\theta(1) \Delta^{(g,z)}_i H^{(g,z)}_i \right] \cdot \left[{\Delta}^m_i(0) T^\theta_i(0) {H}^m_i(0) \tilde{x}_{il}(0) + {\Delta}^m_i(1) T_i^\theta(1) {H}^m_i(1) \tilde{x}_{il}(1)\right]
+\\ - \sum_{i=1}^n \left[x_{ik} T_i^\theta(1) \Delta^{(g,z)}_i H^{(g,z)}_i \Delta^m_i(1) H^m_i(1) \tilde{x}_{il}(1)\right] 
+\\ = \sum_{s=0}^1 \sum_{i=1}^n x_{ik} T_i^\theta(1) H_i^{(g,z)} \Delta_i^{(g,z)} T^\theta_i(s) \Delta_i^m(s) H^m(s) \tilde{x}_{il}(s) \\ - \sum_{i=1}^n \left[x_{il}T_i^\theta(1) \Delta^{(g,z)}_i H^{(g,z)}_i \Delta^m_i(1) H^m_i(1) \tilde{x}_{ik}(1)\right] \\ = \sum_{s=0}^1 X[,k]^T T^\theta(1) H^{(g,z)} \Delta^{(g,z)} T^\theta(s)\Delta^m(s)H^m(s) \tilde{X}(s)[,l] \\ - X[,k]^T \Delta^{(g,z)} H^{(g,z)} T^\theta(1)\Delta^m(1) H^m(1) \tilde{X}[,l].
+\end{multline}
+Combining (\ref{sub_mat_3_zeroinf}), (\ref{sub_mat_3_zeroinf_1}), and (\ref{sub_mat_3_zeroinf_2}) yields
+\begin{multline}\label{sub_mat_4_zeroinf_formula} J_{(\beta_{(g,z)}, \beta_m)}(\theta; m, g) = \left(\sum_{s=0}^1 X^T  T^\theta(1) H^{(g,z)} \Delta^{(g,z)} T^\theta(s) \Delta^m(s) H^m(s) \tilde{X}(s) \right) \\ - X^T \Delta^{(g,z)} H^{(g,z)} T^\theta(1) \Delta^m(1)H^m(1) \tilde{X}(1).
+\end{multline}
+
+\subsubsection*{Submatrix VI (zero-inflated)}
+Denote submatrix VI by $J_{(\beta_{(g,z)},\pi)}(\theta; m, g).$ The formula for $J_{(\beta_{(g,z)},\pi)}(\theta; m, g)$ is
+\begin{multline}\label{sub_mat_6_zeroinf}
+J_{(\beta_{(g,z)},\pi)}(\theta; m, g) = \E \left[-\nabla_{\beta_{(g,z)}} \nabla_{\pi} \mathcal{L}_z(\theta; m, g, p) \right] + \E\left[ \nabla_{\beta_{(g,z)}}\mathcal{L}_z(\theta; m, g ,p)\right] \E \left[ \nabla_{\pi}\mathcal{L}_z(\theta; m,g,p) \right] \\ - \E \left[ \nabla_{\beta_{(g,z)}}\mathcal{L}_z(\theta; m,g,p) \nabla_{\pi}\mathcal{L}_z(\theta; m,g,p) \right].
+\end{multline} Recall that $\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) = X^T P \Delta^{(g,z)}s^{(g,z)}$ and $\nabla_\pi \mathcal{L}_z(\theta; m, g, p) = a \left(\sum_{i=1}^n p_i\right) - b,$ where $$a = \frac{1}{\pi} + \frac{1}{1 - \pi}, \textrm{    } b = \frac{n}{1 - \pi}.$$
+We have that
+\begin{equation}\label{sub_mat_6_zeroinf_1}
+\E \left[-\nabla_{\beta_{(g,z)}} \nabla_{\pi} \mathcal{L}_z(\theta; m, g, p) \right] = 0,
+\end{equation}
+as the derivative in $\pi$ of $\mathcal{L}_z(\theta; m, g, p)$ is a function of $\pi$, and the derivative in $\beta_{(g,z)}$ of this term is $0$.
+Next, we compute the difference of the second two terms of (\ref{sub_mat_6_zeroinf}).
+The $k$th entry of this vector is
+\begin{multline}\label{sub_mat_6_zeroinf_2}
+\E \left[ \nabla_\pi \mathcal{L}_z(\theta; m, g, p) \right] \E\left[\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, x)[k] \right] - \E \left[\nabla_{\pi}\mathcal{L}(\theta; m,g,p) \nabla_{\beta_{(g,z)}} \mathcal{L} (\theta; m,g,p)[k] \right] \\= \left(\E \left[ a \sum_{i=1}^n p_i - b \right] \right)\left( \E \left[{X}[,k]^T P \Delta^{(g,z)} s^{(g,z)} \right]\right) - \E \left[ \left( a \sum_{i=1}^n p_i - b \right) {X}[,k]^T P \Delta^{(g,z)} s^{(g,z)}\right] \\ = \left( a \sum_{i=1}^n \E[p_i] - b \right) \left( \sum_{j=1}^n \E [x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] \right) \\ - \E \left[ \left( a \sum_{i=1}^n p_i - b \right) \left( \sum_{j=1}^n \tilde{x}_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j \right) \right] \\= a \sum_{i=1}^n \sum_{j=1}^n \E [p_i] \E[x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] - b \sum_{j=1}^n \E[x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] \\ - \left[a \sum_{i=1}^n \sum_{j=1}^n \E [ p_i x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] - b \sum_{j=1}^n \E[x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] \right] \\ = a \sum_{i=1}^n \sum_{j=1}^n \E[p_i] \E[{x}_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] - a \sum_{i \neq j} \E[p_i] \E[x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] - a \sum_{i=1}^n \E[ x_{ik} p_i^2 \Delta^{(g,z)}_i s^{(g,z)}_i] \\ = a \sum_{i=1}^n \E[p_i] \E[x_{ik} p_i \Delta^{(g,z)}_i s^{(g,z)}_i] - a \sum_{i=1}^n \E[x_{ik} p_i^2 \Delta^{(g,z)}_i s^{(g,z)}_i] \\ = a \sum_{i=1}^n T^\theta_i(1) x_{ik} T^\theta_i(1) \Delta^{(g,z)}_i s^{(g,z)}_i - a \sum_{i=1}^n x_{ik} T^\theta_i(1) \Delta^{(g,z)}_i s^{(g,z)}_i \\ = a \sum_{i=1}^n \left( {x}_{ik} T^\theta_i(1)^2 \Delta^{(g,z)}_i s^{(g,z)}_i - x_{ik} T^\theta_i(1) \Delta^{(g,z)}_i s^{(g,z)}_i \right) = a \sum_{i=1}^n {x}_{ik}T^\theta_i(1)\Delta^{(g,z)}_i s^{(g,z)}_i \left( T^\theta_i(1) - 1\right) \\ = -a\sum_{i=1}^n x_{ik} T_i(0) T^\theta_i(1)\Delta^{(g,z)}_i H^{(g,z)}_i = -a X[,k]^T w^{(g,z)}.
+\end{multline}
+Combining (\ref{sub_mat_6_zeroinf}), (\ref{sub_mat_6_zeroinf_1}), and (\ref{sub_mat_6_zeroinf_2}), we conclude that
+\begin{equation}\label{sub_mat_6_zeroinf_formula} J_{(\beta_{(g,z)},\pi)}(\theta; m, g) = -\left( \frac{1}{\pi} + \frac{1}{1-\pi} \right)X^T w^{(g,z)}.\end{equation}
+
+\section{Statistical accelerations}\label{sec:statistical_accelerations}
+
+\subsection*{Computing $\pi^\textrm{pilot}$, $[\beta^m_1]^\textrm{pilot}$, $[\beta^m_1]^\textrm{pilot}$}
+
+We present the algorithm (Algorithm \ref{algo:pilot_estimates_2}) for computing the pilot estimates $\pi^\textrm{pilot}$, $[\beta^m_1]^\textrm{pilot}$, $[\beta^m_1]^\textrm{pilot}$.
+This algorithm, alongside Algorithm \ref{algo:pilot_estimates_1}, is used to compute pilot estimates for use in Algorithm \ref{algo:em_full}.
+
+\begin{algorithm}
+	\caption{Computing $\pi^\textrm{pilot}$, $[\beta^m_1]^\textrm{pilot}$, $[\beta^m_1]^\textrm{pilot}$.}\label{algo:pilot_estimates_2}
+	\begin{algorithmic}[2]
+		\Require Data $m$, $g$; fitted offsets $\hat{f}^m,$ $\hat{f}^g$.
 		\State \texttt{bestLik} $\gets -\infty$ \Comment{Reduced GLM-EIV}
@@
 		\EndFor
-		\State \textbf{return} $(\pi^\textrm{pilot}, [\beta^m_0]^\textrm{pilot}, [\beta^m_1]^\textrm{pilot}, [\gamma^T_m]^\textrm{pilot}, [\beta^g_0]^\textrm{pilot}, [\beta^g_1]^\textrm{pilot}, [\gamma^T_g]^\textrm{pilot})$
+		\State \textbf{return} $(\pi^\textrm{pilot}, [\beta^m_1]^\textrm{pilot}, [\beta^g_1]^\textrm{pilot})$
 	\end{algorithmic}
@@
 
-The first step (Algorithm \ref{algo:pilot_estimates}, lines 2-7) is to obtain good parameter estimates for $[\beta^m_0, \gamma_m]^T$ and $[\beta^g_0, \gamma_g]^T$ using regression.
-Recall that the underlying gene expression parameter vector $\beta_m$ is $\beta_m = [\beta^m_0, \beta^m_1, \gamma_m]^T \in \R^d$, where $\beta^m_0$ is the intercept, $\beta^m_1$ is the effect of the perturbation, and $\gamma_m^T$ is the effect of the technical factors.
-To produce estimates $[\beta^m_0]^\textrm{pilot}$ and $[\gamma_m^T]^\textrm{pilot}$, we regress the gene expressions $m$ onto the technical factors $X$.
-The intuition for this procedure is as follows: the probability of perturbation $\pi$ is very small.
-Therefore, the true log likelihood is approximately equal to the log likelihood that results from omitting $p_i$ from the model:
-\begin{multline*}
-\sum_{i=1}^n f_m(m_i; \eta^m_i) = \sum_{i=1}^n f_m(m_i; h_m(\beta_0 + \beta_1 p_i + \gamma^T z_i + o^m_i)) \\ = \underbrace{\sum_{i : p_i =1} f_m(m_i; h_m(\beta_0 + \beta_1 + \gamma^T z_i + o^m_i))}_\textrm{few terms}  + \underbrace{\sum_{i : p_i = 0} f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i))}_\textrm{many terms} \\ \approx \sum_{i = 1}^n f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i)).
-\end{multline*}
-We similarly can obtain pilot estimates $[\beta^g_0]^\textrm{pilot}$ and $[\gamma^T_g]^\textrm{pilot}$ by regressing the gRNA counts $g$ onto the technical factors $X$.
-We extract the fitted values (on the scale of the linear component) for use in a subsequent step: $$\hat{f}^k_i = [\beta^k_0]^\textrm{pilot} + \langle [\gamma^T_k]^\textrm{pilot}, z_i \rangle + o^k_i,$$
-for $k \in \{m,g\}.$
-
-Next, we obtain estimates $[\beta_1^m]^\textrm{pilot},$ $[\beta_1^g]^\textrm{pilot},$ and $\pi^\textrm{pilot}$ for $\beta^m_1$, $\beta^g_1$, and $\pi$ by fitting a ``reduced'' GLM-EIV (Algorithm \ref{algo:pilot_estimates}, lines 8 onward).
-The log likelihood of the no-intercept, univariate GLM with predictor $p_i$ and offset $\hat{f}^m_i$ is approximately equal to the true log likelihood:
-$$ \sum_{i=1}^n f_m(m_i; \eta^m_i) =
-\sum_{i=1}^n f_m(m_i; h_m(\beta_0 + \beta_1 p_i + \gamma^T z_i + o^m_i)) \approx \sum_{i=1}^n f_m(m_i; h_m( \beta_1 p_i + \hat{f}^m_i)).
-$$
-Therefore, to estimate $\beta^m_1$, $\beta^g_1$, and $\pi$, we fit a GLM-EIV model with gene expressions $m$, gRNA counts $g$, gene offsets $\hat{f}^m := [\hat{f}^m_1, \dots, \hat{f}^m_n]^T$, gRNA offsets $\hat{f}^g := [\hat{f}^g_1, \dots, \hat{f}^g_n]^T$, and \textit{no} intercept or covariate terms.
-Intuitively, we ``encode'' all information about technical factors, library sizes, and baseline expression levels into $\hat{f}^m$ and $\hat{f}^g$.
-We run the algorithm $B \approx 15$ times over randomly-selected starting values for $\beta^m$, $\beta^g$, and $\pi$ and select the solution with greatest the log likelihood.
-
-The M step of the ``reduced'' GLM-EIV algorithm requires fitting two no-intercept, univariate GLMs with offsets (Algorithm \ref{algo:pilot_estimates}, line 20).
-We derive analytic formulas for the MLEs of these GLMs in the three most important cases: Gaussian response with identity link, Poisson response with log link, and negative binomial response with log link (see appendix; the latter formula is asymptotically exact, and, to the best of our knowledge, new).
-Consequently, we do not need to run the relatively slow IRLS procedure to carry out the M step of the ``reduced'' GLM-EIV algorithm.
-Overall, the proposed algorithm for obtaining the pilot parameter estimates (Algorithm \ref{algo:pilot_estimates}) requires fitting only two GLMs (via IRLS).
-
-\subsection{Computing}\label{sec:computing}
-We developed a computational infrastructure to apply GLM-EIV to large-scale, single-cell CRISPR screen data.\footnote{Pipeline available at \texttt{github.com/timothy-barry/glmeiv-pipeline}.} The infrastructure leverages Nextflow, a programming language that facilitates building data-intensive pipelines \cite{DITommaso2017}, and ondisc\footnote{Package available at \texttt{github.com/timothy-barry/ondisc}; preprint forthcoming.}, an R package that we developed (in a separate project) to facilitate large-scale computing on single-cell data.
-Nextflow and ondisc together enable the construction of highly portable single-cell pipelines: one can analyze data \textit{out-of-memory} on a laptop or in a \textit{distributed} fashion across tens or hundreds of nodes on a cloud (e.g., Microsoft Azure, Google Cloud) or high-performance cluster.
-
-Leveraging these technologies, we develop a scalable and efficient pipeline for GLM-EIV (Algorithm \ref{algo:at_scale}).
-First, we run a round of ``precomputations'' on all $d_g$ genes and $d_p$ perturbations.
-The precomputations involve regressing the gene expressions (or gRNA counts) onto the technical factors, thereby ``factoring out'' lines 2-7 from Algorithm \ref{algo:pilot_estimates}.
-Next, we run differential expression analyses on the full set of gene-perturbation pairs; for a given pair, this amounts to obtaining the complete set of pilot parameters (Algorithm \ref{algo:pilot_estimates}, lines 8 onward), fitting the GLM-EIV model (Algorithm \ref{algo:em_full}), and performing inference.
-The three loops in Algorithm \ref{algo:at_scale} are embarassingly parallel and therefore can be massively parallelized.
-
-\begin{algorithm}
-	\caption{Applying GLM-EIV at scale.}\label{algo:at_scale}
-	\begin{algorithmic}
-		\State $G \gets \{\textrm{gene}_1, \dots, \textrm{gene}_{d_g}\}; P \gets \{\textrm{perturbation}_1, \dots, \textrm{perturbation}_{d_p}\}$
-
-		\For{gene $\in G$}
-			\State Run precomputation (Algorithm \ref{algo:pilot_estimates}, lines 2-7) on gene. \State Save $\hat{f}^m$, $[\beta^m_0]^\textrm{pilot}$ and $[\gamma^T_m]^\textrm{pilot}$.
-		\EndFor
-		\For{perturbation $\in P$}
-		\State Run precomputation  (Algorithm \ref{algo:pilot_estimates}, lines 2-7) on perturbation.
-		\State Save $\hat{f}^g$, $[\beta^g_0]^\textrm{pilot}$ and $[\gamma^T_g]^\textrm{pilot}$.
-		\EndFor
-		\For{(gene, perturbation) $\in G \times P$}
-		\State Load $\hat{f}^m, \hat{f}^g,$ $[\beta^m_0]^\textrm{pilot}$ $[\gamma^T_m]^\textrm{pilot}$, $[\beta^g_0]^\textrm{pilot}$ and $[\gamma^T_g]^\textrm{pilot}$.
-		\State Compute $[\beta^m_1]^\textrm{pilot}, [\beta^g_1]^\textrm{pilot}, \pi^\textrm{pilot}$ (Algorithm \ref{algo:pilot_estimates}, lines 8 onward).
-		\State Run GLM-EIV using the pilot parameters (Algorithm \ref{algo:em_full}).
-		\EndFor
-	\end{algorithmic}
-\end{algorithm}
-
-\subsection{Zero-inflated model}
-The GLM-EIV model assumes that unperturbed cells exhibit a nonzero ``background read'' gRNA count distribution due to errors in the sequencing and alignment processes.
-This assumption very likely is appropriate for most current single-cell CRISPR screen datasets \cite{Replogle2020}.
-However, background contamination may be negligible for some datasets, in which case the gRNA counts of unperturbed cells should be modeled as point masses at zero.
-We introduce such an extension -- which we call the ``zero-inflated GLM-EIV model'' -- in Appendix \ref{sec:zero_inf_model}.
-We additionally derive efficient methods for estimation and inference in the zero-inflated model.
-Given the ubiquity of background contamination in current single-cell CRISPR screen datasets, we focus our attention on the standard GLM-EIV model for the remainder of the main text.
-
-\section{Simulation studies}\label{sec:simulation}
-
-We conducted a simulation study to compare the empirical performance of GLM-EIV to that of the thresholding method.
-We generated data on $n = 150,000$ cells from the GLM-EIV model (Section \ref{sec:model}), setting the the target of inference $\beta^m_1$ to $\log(0.25)$ and the probability of perturbation $\pi$ to $0.02$.
-$\beta^m_1 = \log(0.25)$ represents a decrease in gene expression by a factor of $4$, which is a fairly large effect size on the order of what we might observe for a positive control pair.
-We included ``sequencing batch'' (modeled as a Bernoulli-distributed variable) as a covariate and sequencing depth (modeled as a Poisson-distributed variable) as an offset.
-We varied the log-fold change in gRNA expression, $\beta^g_1,$ over a grid on the interval $[\log(1), \log(4)];$ $\beta^g_1$ controls problem difficulty, with higher values corresponding to easier problems.
-Finally, we generated the gene expression and gRNA count data from two response distributions: Poisson and negative binomial (size parameter fixed at $\theta = 20$ for the latter).
-For each parameter setting (defined by a $\beta^g_1$-distribution pair), we synthesized $n_\textrm{sim} = 500$ i.i.d.\ datasets.
-Section \ref{sec:extra_sims} presents additional simulation results on Gaussian response distributions.
-
-We applied three methods to the simulated data: ``vanilla'' GLM-EIV (Algorithm \ref{algo:em_full}), accelerated GLM-EIV (Algorithm \ref{algo:pilot_estimates}), and thresholded regression.
-We used the Bayes-optimal decision boundary for classification as the threshold for the thresholding method.
-We ran all methods on the negative binomial data twice: once treating the size parameter $\theta$ as a known constant and once treating $\theta$ as unknown.
-In the latter case we used the \texttt{glm.nb} function from the \texttt{MASS} package to estimate $\theta$ \cite{Ripley2013}.
-We display the results of the simulation study in Figure \ref{main_text_sim}.
-Columns correspond to distributions (i.e., Poisson, negative binomial with known $\theta$, and negative binomial with unknown $\theta$), rows correspond to performance metrics (i.e., bias, mean squared error, CI coverage rate (nominal rate $95\%$), CI width, and method execution time).
-The problem difficulty parameter $\beta^g_1$ is plotted on the horizontal axis, and the methods are depicted in different colors (GLM-EIV = red, accelerated GLM-EIV = blue, and thresholding = purple; GLM-EIV masked by accelerated GLM-EIV in several panels).
-
-First, we observed that GLM-EIV dominated thresholded regression on all statistical metrics: GLM-EIV exhibited lower bias (row 1) and mean squared error (row 2) than thresholded regression; additionally, GLM-EIV had superior confidence interval coverage (row 3) despite having produced generally narrower confidence intervals (row 4).
-Intuitively, GLM-EIV outperformed the thresholding method because (i) GLM-EIV leveraged information from \textit{both} modalities (rather than the gRNA modality alone) to assign perturbation identities to cells, and (ii) GLM-EIV produced soft rather than hard assignments, capturing the inherent uncertainty in whether a perturbation occurred.
-Next, we found that accelerated GLM-EIV performed as well as ``vanilla'' GLM-EIV on the statistical metrics (rows 1-4) despite have substantially lower computational cost (bottom row).
-In fact, the execution time of accelerated GLM-EIV was approximately within an order of magnitude of that of the thresholding method (bottom row).
-
-Interestingly, thresholded regression exhibited better confidence interval coverage under estimated $\theta$ than under known $\theta$ (row 3).
-Estimating $\theta$ leads to slight inflation bias (i.e., overestimating the true effect size), whereas, as we showed previously, thresholding leads to attenuation bias (i.e., underestimating the true effect size).
-These phenomena partially cancel, yielding less biased estimates.
-We note that GLM-EIV and the thresholding method in principal are compatible with \textit{any} $\theta$ estimation procedure, including those based on more sophisticated techniques, such as regularization \cite{Hafemeister2019}.
-We defer rigorous investigation of the impact of different $\theta$ estimation strategies on these methods to future work.
-
-\section{Data analysis}
-
-Leveraging our computational infrastructure (Section \ref{sec:computing}), we applied GLM-EIV to analyze the entire Gasperini and Xie datasets (Figure \ref{fig:real_data}).
-Our theoretical analyses (Section \ref{sec:thresholding_theory}) and simulation studies (Section \ref{sec:simulation}) revealed that the Bayes-optimal decision boundary is a good choice for the threshold when the gRNA count distribution is well separated.
-Therefore, under the assumption that the effect of the perturbation on gRNA expression is similar across pairs, we used the fitted GLM-EIV models to approximate the Bayes-optimal decision boundary on both datasets, yielding a boundary of $3$ for Gasperini and $7$ for Xie.
-We then applied the thresholding method to the data, setting the threshold to the Bayes-optimal decision boundary estimated using GLM-EIV.
-
-We compared GLM-EIV to thresholded regression on the real data, focusing specifically on the negative control pairs (i.e., gene-perturbation pairs that, by design, are expected to exhibit a fold change of $1$, or no association).
-We found that GLM-EIV and the thresholding method produced similar results (Figure \ref{fig:real_data}a-b): estimates, CI coverage rates, and CI widths were concordant.
-We note that GLM-EIV produced outlier estimates (likely due to non-global EM convergence) on a small ($< 2.5 \%$ on Gasperini, $<0.05\%$ on Xie) number of pairs consisting of a handful of genes (not plotted).
-The estimated effect of the perturbation on gene expression $\exp(\hat{\beta}_1^g)$ was unexpectedly large (95\% CI $[4306, 5186]$ and $[300, 316]$ on Gasperini and Xie data, respectively).
-We concluded that the datasets lay in an ``easy'' region of the parameter space, making thresholding a tenable strategy (provided the threshold is selected well).
-
-To explore the settings in which GLM-EIV and the thresholding method diverge, we artificially increased the problem difficulty by generating partially-synthetic datasets.
-To this end, we defined a parameter called ``excess background contamination'' that controls the extent to which the gRNA count distribution is separated.
-When excess background contamination is zero, we sample gRNA counts directly from the fitted GLM-EIV model.
-As excess background contamination increases, we sample gRNA counts from a slightly modified version of the fitted model in which the mean gRNA expression of \textit{unperturbed} cells is increased while the mean gRNA expression of \textit{perturbed} cells is held constant.
-Finally, when excess background contamination is $1$ (the maximum value), we set the mean gRNA expression of unperturbed cells equal to that of perturbed cells and sample gRNA counts from the resulting model.
-We hold fixed the real-data gene expressions, library sizes, covariates, and fitted perturbation probabilities in all settings.
-
-We generated partially-synthetic data in the above manner for each of the 322 positive control pairs in the Gasperini dataset, varying excess background contamination over the interval $[0,0.4].$ We then applied GLM-EIV and the thresholding method to analyze the data.
-We present results on two example pairs (the pair containing gene \textit{LRIF1} and the pair containing gene \textit{NDUFA2}) in Figures \ref{fig:real_data}c-d.
-We observed that the estimate produced by the methods on the raw data (depicted as a horizontal black line) coincided almost exactly with the estimate produced by the methods on the partially-synthetic data generated by setting excess background contamination to zero (This result replicated across nearly all pairs; average relative difference $0.003$.) We concluded that the fitted models accurately tracked the gRNA count distributions, giving us confidence in our procedure for generating realistic, partially-synthetic datasets.
-We found that as excess background contamination increased, the performance of thresholded regression degraded considerably while that of GLM-EIV remained stable.
-
-We generalized the above analysis to the entire set of positive control pairs.
-First, for each pair we computed its ``relative estimate change'' (REC) as a function of excess background contamination, defined as the relative difference between the estimate at a given level of excess contamination and zero excess contamination (Figure \ref{fig:real_data}d).
-Next, we computed the median REC across all positive control pairs (Figure \ref{fig:real_data}e; bands indicate pointwise 95\% CIs for the population median REC).
-As excess background contamination increased, thresholded regression exhibited severe attenuation bias (as reflected by large median REC values); GLM-EIV, by contrast, remained mostly stable.
-Finally, taking the estimate obtained on the raw data as ``ground truth,'' we computed the CI coverage rate across pairs as a function of excess contamination (Figure \ref{fig:real_data}f; bands indicate $95\%$ pointwise CIs for population coverage rate).
-GLM-EIV exhibited a significantly higher CI coverage rate than thresholded regression as the data became increasingly contaminated.
-
-\section{Discussion}
-
-In this work we introduced GLM-EIV (``GLM-based errors in variables''), a new method for single-cell CRISPR screen analysis.
-GLM-EIV generalizes the classical errors-in-variables model (when the latent predictor is binary) to arbitrary exponential distributions and link functions, facilitating the analysis of single-cell CRISPR screen data.
-We demonstrated through theoretical analyses, simulation studies, and real data analyses that GLM-EIV outperforms thresholded regression by a considerable margin in high background contamination settings.
-On the other hand, in low background contamination settings, GLM-EIV provides a statistically principled and empirically effective procedure for selecting a threshold for use in a thresholding method (e.g., thresholded negative binomial regression, as studied in this work, or even SCEPTRE \cite{Barry2020}).
-GLM-EIV thereby neutralizes a tuning parameter that exerts a significant impact on the results.
-Figure \ref{fig:dec_tree} summarizes how we anticipate GLM-EIV being used in practice.
-
-To our knowledge this is the first single-cell CRISPR screen paper oriented toward a statistical audience.
-We hope that this work helps to introduce the broader statistics community to an emerging class functional genomics assays that likely will exert a major impact on biology research in the coming years \cite{Przybyla2021}.
-Additionally, this is the first work to leverage the ondisc-Nextflow-cloud/HPC technology stack, a tightly-integrated, user-friendly, and powerful set of computational tools for large-scale single-cell analysis likely of interest to other researchers.
-
-We anticipate that GLM-EIV could be applied to other types of single-cell CRISPR screen and multimodal single-cell data.
-For example, we likely (with some effort) could extend GLM-EIV to ``low-multiplicity of infection'' screens \cite{Schraivogel2020} in which each cell receives one or two perturbations rather than dozens (as is the case in ``high multiplicity screens,'' studied in this work).
-We also likely could apply GLM-EIV to analyze multimodal single-cell chromatin accessibility assays.
-A question of interest in such experiments is whether chromatin state (i.e., closed or open) is associated with the expression of a gene (or abundance of a protein) \cite{Mimitou2021}.
-We do not directly observe chromatin state; instead, we observe tagged DNA fragments, counts that serve as proxies for whether a given region of chromatin is open or closed.
-GLM-EIV might be applied in such experiments to aid in the selection of thresholds or to analyze whole datasets.
-\printbibliography
-
-\clearpage
-\section*{Figures}
-% 1. Challenges
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=1\linewidth]{../../figures/analysis_challenges/plot.pdf}
-	\caption{\textbf{Experimental design and analysis challenges}: \textbf{a,} Experimental design.
-For a given perturbation (e.g., the perturbation indicated in blue), we partition the cells into two groups: perturbed and unperturbed.
-Next, for a given gene, we conduct a differential expression analysis across the two groups, yielding an estimate of the impact of the given perturbation on the given gene.
-\textbf{b,} DAG representing all variables in the system.
-The perturbation (latent) impacts both gene expression and gRNA expression; technical factors act as confounders, also impacting gene and gRNA expression.
-The target of estimation is the effect of the perturbation on gene expression.
-\textbf{c,} Schematic illustrating the ``background read'' phenomenon.
-Due to errors in the sequencing and alignment processes, unperturbed cells exhibit a nonzero gRNA count distribution (bottom).
-The target of estimation is the change in mean gene expression in response to the perturbation (top).
-\textbf{d}, Example data on four cells for a given perturbation-gene pair.
-Note that (i) the perturbation is unobserved, and (ii) the gene and gRNA data are discrete counts.}
-	\label{analysis_challenges}
-\end{figure}
-
-\clearpage
-% 2. Emprical challenges
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=1\linewidth]{../../figures/thresholding_empirical/plot}
-	\caption{\textbf{Empirical challenges of thresholded regression.} \textbf{a-b,} Fold change estimates produced by threshold = 1 versus threshold = 5 (a) and threshold = 20 versus threshold = 5 (b).
-The selected threshold substantially impacts the results.
-\textbf{c-d,} $p$-values (c) and CI widths (d) produced by threshold = 20 versus threshold = 5.
-The latter threshold yields more confident estimates.
-\textbf{e-f}, Empirical distribution of randomly-selected gRNA from Gasperini (e) and Xie (f) data (0 counts not shown).
-The gRNA data do not appear to imply an obvious threshold selection strategy.}
-	\label{thresholding_empirical}
-\end{figure}
-
-\clearpage
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=1\linewidth]{../../figures/thresholding_theoretical/plot}
-	\caption{\textbf{Theoretical challenges of thresholded regression.} \textbf{a,} Asymptotic relative bias versus threshold for different values of $\beta^g_1$.
-The bias function is highly nonconvex and strictly nonzero.
-Vertical blue lines, Bayes-optimal decision boundaries.
-Across all panels, $\beta^g_0 = 0$ and $\pi = 1/2$.
-\textbf{b,} Asymptotic relative bias versus $\pi$ when the threshold is set to a large number.
-The two quantities coincide exactly.
-\textbf{c,} Bias-variance decomposition for thresholding method in no-intercept model.
-Bias decreases and variance increases as the threshold tends to infinity.
-$\beta^g_1 = 1, \beta^m_1 = 1,$ and $\pi = 0.1$.}
-	\label{thresholding_theoretical}
-\end{figure}
-
-\clearpage
-% Theory
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=0.9\linewidth]{../../figures/main_text_sim/plot}
-	\caption{\textbf{Simulation study}.
-Columns correspond to distributions (Poisson, NB with known $\theta$, NB with estimated $\theta$), and rows correspond to metrics (bias, MSE, coverage, CI width, and time).
-Methods are shown in different colors; GLM-EIV (red) is masked by accelerated GLM-EIV (blue) in several panels.
-GLM-EIV demonstrated superior statistical performance to the thresholding method on all metrics (rows 1-4).
-Accelerated GLM-EIV had substantially lower computational cost than ``vanilla'' GLM-EIV (bottom row) despite demonstrating identical statistical performance (rows 1-4).}
-	\label{main_text_sim}
-\end{figure}
-
-\clearpage
-\begin{figure}
-	\centering
-	\includegraphics[width=0.9\linewidth]{../../figures/data_analysis/annotated_plot_crop}
-	\caption{\textbf{Data analysis}.
-\textbf{a-b}, Estimates for fold change produced by GLM-EIV and thresholded regression on Gasperini (\textbf{a}) and Xie (\textbf{b}) negative control pairs.
-\textbf{c-d}, Estimates produced by GLM-EIV and thresholded regression on two positive control pairs -- \textit{LRIF1} (\textbf{a}) and \textit{NDUFA2} (\textbf{b}) -- plotted as a function of excess background contamination.
-Grey bands, 95\% CIs for the target of inference outputted by the methods.
-\textbf{e-f}, Median relative estimate change (REC; \textbf{e}) and confidence interval coverage rate (\textbf{f}) across \textit{all} 322 positive control pairs, plotted as a function of excess background contamination.
-Colored bands, pointwise 95\% CIs for population quantities.
-Panels (\textbf{c-f}) together illustrate that GLM-EIV demonstrated greater stability than thresholded regression as background contamination increased.}\label{fig:real_data}
-\end{figure}
-
-\clearpage
-% Dec tree
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=0.9\linewidth]{../../figures/dec_tree/dec_tree_crop}
-	\caption{\textbf{Use of GLM-EIV in practice}.
-The decision tree above illustrates how we anticipate GLM-EIV could be used in practice.
-First, apply GLM-EIV to a set of randomly-sampled gene-perturbation pairs to assess background contamination level (positive control pairs work best for this purpose).
-If GLM-EIV indicates that background contamination is high (e.g., $\exp(\beta^g_1) \lessapprox $ 10), apply GLM-EIV to analyze the entire dataset; otherwise, approximate the Bayes-optimal decision boundary using the fitted GLM-EIV models.
-Next, apply a thresholding method (e.g., SCEPTRE or thresholded negative binomial regression) to analyze the data, setting the threshold to the estimated Bayes-optimal decision boundary.}\label{fig:dec_tree}
-\end{figure}
-
-\clearpage
-\begin{appendices}
-\begin{refsection}
-\section{Theoretical details for thresholding estimator}
-This section contains proofs of the propositions presented Section \ref{sec:thresholding_theory}, ``Theoretical analysis of thresholding estimator.'' The subsections are organized as follows.
-Section (\ref{sec:notation}) introduces some notation.
-Section (\ref{sec:convergence}) establishes almost sure convergence of the thresholding estimator in the model (\ref{theoretical_model}), proving Proposition \ref{prop:convergence}.
-Section (\ref{sec:simplication}) simplifies the expression for the attenuation function $\gamma$, and section (\ref{sec:derivatives})  computes derivatives of $\gamma$ to be used throughout the proofs.
-Section (\ref{sec:c_limit}) establishes the limit in $c$ of $\gamma$, proving Proposition \ref{prop:c_limit} and as a corollary Proposition \ref{prop:c_limit_half}.
-Section (\ref{sec:bayes_opt}) establishes that the Bayes-optimal decision boundary is a critical value of $\gamma$, proving Proposition \ref{prop:bayes_opt}, and section (\ref{sec:comparison}) compares the competing threshold selection strategies head-to-head, proving Proposition \ref{prop:comparison}.
-Section (\ref{sec:monotone}) demonstrates that $\gamma$ is monotone in $\beta^g_1$, proving Proposition \ref{prop:monotonic}, and Section (\ref{sec:att_bias}) establishes attenuation bias of the thresholding estimator, proving Proposition \ref{prop:att_bias}.
-Finally, Section (\ref{sec:bv_decomp}) derives the bias-variance decomposition of the thresholding estimator in the model (\ref{theoretical_model_no_int}), proving Proposition \ref{prop:bv_decomp}.
-
-% Labels of propositions and theorems
-% propositions:
-% prop:convergence
-% prop:att_bias
-% prop:bayes_opt
-% prop:c_limit_half
-% prop:comparison
-% prop:monotonic
-% prop:c_limit
-% prop:bv_decomp
-
-% sections
-% sec:notation
-% sec:convergence
-% sec:simplication
-% sec:derivatives
-% sec:c_limit
-% sec:bayes_opt
-% sec:comparison
-% sec:beta_lim
-% sec:monotone
-% sec:att_bias
-% sec:bv_decomp
-
-\subsection{Notation}\label{sec:notation}
- All notation introduced in this subsection (i.e., \ref{sec:notation}) pertains to the Gaussian model with intercepts (\ref{theoretical_model}).
-Recall that the attenuation function $\gamma: \R^4 \to \R$ is defined by
-$$ \gamma(\beta^g_1, c, \pi, \beta^g_0) = \frac{\pi(\omega - \E[\hat{p}_i])}{ \E[\hat{p}_i](1 -\E[\hat{p}_i])},$$ where $$\begin{cases} \E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi, \\
-\omega = \Phi\left(\beta_1^g + \beta_0^g - c \right) ,\\ \zeta = \Phi\left( \beta^g_0 - c \right).
-\end{cases}$$ Additionally, recall that the asymptotic relative bias function $b: \R^4 \to \R$ is
-$$ b(\beta^g_1, c, \pi, \beta^g_0) = 1 - \gamma(\beta^g_1, c, \pi, \beta^g_0).$$ Next, we define the functions $g$ and $h: \R^4 \to \R$ by
-\begin{equation}\label{def_g}
-g(\beta^g_1, c, \pi, \beta^g_0) = (1-\pi)\left( \Phi(\beta_0^g + \beta_1^g - c)\right) - (1-\pi)\left(\Phi(\beta_0^g - c)\right)\end{equation}
- and
-\begin{multline}\label{def_h}
-h(\beta^g_1, c, \pi, \beta^g_0) = \left[(1-\pi)\left( \Phi(\beta_0^g - c)\right) + \pi\left(\Phi(\beta^g_0 + \beta^g_1 - c) \right) \right] \cdot \\ \left[(1-\pi)\left( \Phi(c - \beta^g_0) \right) + \pi\left(\Phi(c - \beta_0^g - \beta_1^g) \right) \right].
-\end{multline}
-We use $f:\R \to \R$ to denote the $N(0,1)$ density, and we denote the right-tail probability probability of $f$ by $\bar{\Phi}$, i.e., 
-$$\bar{\Phi}(x) = \int_{x}^{\infty} f = \Phi(-x).$$
-
-The parameter $\beta^g_0$ is a given, fixed constant throughout the proofs.
-Therefore, to minimize notation, we typically use $\gamma(\beta^g_1, c, \pi)$ (resp., $b(\beta^g_1, c, \pi),$ $g(\beta^g_1, c, \pi),$ $h(\beta^g_1, c, \pi)$) to refer to the function $\gamma$ (resp., $b, g, h$) evaluated at $(\beta^g_1, c, \pi, \beta^g_0)$.
-Finally, for a given function $r: \R^{p} \to \R$, point $x \in \R^p$, and index $i \in \{1, \dots, p\}$, we use the symbol $D_i r(x)$ to refer to the derivative of the $i$th component of $r$ evaluated at $x$ (\textit{sensu} \cite{fitzpatrick2009}).
-For example, $D_1 \gamma(\beta^g_1, c, 1/2)$ is the derivative of the first component of $\gamma$ (the component corresponding to $\beta^g_1$) evaluated at $(\beta^g_1, c, 1/2)$.
-Likewise,  $D_2g(\beta^g_1, c, \pi)$ is the derivative of the second component of $g$ (the component corresponding to $c$) evaluated at $(\beta^g_1, c, \pi).$
-
-\subsection{Almost sure limit of $\hat{\beta}^m_1$}\label{sec:convergence}
-
-We derive the limit in probability of $\hat{\beta}^m_1$ for the Gaussian model with intercepts (\ref{theoretical_model}).
-Dividing by $n$ in (\ref{thresh_est_intercepts}), we can express $\hat{\beta}^m_1$ as
-$$ \hat{\beta}^m_1 = \frac{ \frac{1}{n} \sum_{i=1}^n ( \hat{p}_i - \overline{\hat{p}_i})(m_i - \overline{m})}{ \frac{1}{n} \sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})}.$$ By weak LLN,
-$$ \hat{\beta}^m_1 \xrightarrow{P} \frac{\textrm{Cov}(\hat{p}_i, m_i)}{\V\left(\hat{p}_i\right)}.$$ To compute this quantity, we first compute several simpler quantities:
-\begin{itemize}
-\item[1.] Expectation of $m_i$: $\E[m_i] = \beta^m_0 + \beta^m_1\pi.$
-\item[2.] Expectation of $\hat{p}_i$: \begin{multline*}
-\E[\hat{p}_i] = \P\left[\hat{p}_i = 1\right] = \P\left[\beta^g_0 + \beta^g_1 p_i + \tau_i \geq c \right] = \\ \textrm{(By LOTP) } \P\left[ \beta^g_0 + \tau_i \geq c \right]\P\left[p_i = 0\right] + \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \P[p_i = 1] \\ = \P\left[ \tau_i \geq c - \beta^g_0\right](1- \pi) + \P\left[ \tau_i \geq c - \beta^g_1 - \beta^g_0 \right](\pi) \\ =  \left(\bar{\Phi}(c - \beta^g_0) \right) (1 - \pi) + \left( \bar{\Phi}(c - \beta^g_1 - \beta^g_0) \right)(\pi) = \\  \Phi(\beta^g_0 - c) (1-\pi) + \Phi(\beta^g_1 + \beta^g_0 - c) \pi = \zeta(1-\pi) + \omega \pi.
-\end{multline*}
-\item[3.] Expectation of $\hat{p}_i p_i$: 
-$$\E\left[ \hat{p}_i p_i \right] = \E\left[\hat{p}_i | p_i = 1 \right] \P\left[ p_i =1 \right] = \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \pi = \omega \pi.$$
-\item[4.] Expectation of $\hat{p}_i m_i$:
-\begin{multline*}
-\E\left[\hat{p}_i m_i\right] = \E[\hat{p}_i (\beta^m_0 + \beta^m_1 p_i + \ep_i)] = \beta^m_0 \E\left[\hat{p}_i\right] + \beta^m_1 \E\left[\hat{p}_i p_i\right] + \E[\hat{p}_i \ep_i] \\ = \beta^m_0 \E[\hat{p}_i] + \beta^m_1 \omega \pi + \E[\hat{p}_i] \E[\ep_i] = \beta^m_0 \E[\hat{p}_i] + \beta^m_1 \omega \pi.
-\end{multline*}
-\item[5.] Variance of $\hat{p}_i$: Because $\hat{p}_i$ is binary, we have that $\V[\hat{p}_i] = \E[\hat{p}_i]\left(1 - \E[\hat{p}_i]\right) .$
-\item[6.] Covariance of $\hat{p}_i, m_i$:
-\begin{multline*}
-\textrm{Cov}\left(\hat{p}_i, m_i\right) = \E\left[\hat{p}_i m_i\right] - \E[\hat{p}_i] \E[m_i] = \beta^m_0 \E[\hat{p}_i] + \beta^m_1 \omega \pi - \E[\hat{p}_i]( \beta^m_0 + \beta^m_1 \pi)\\ = \beta^m_1 \omega \pi - \E[\hat{p}_i] \beta_1^m \pi = \beta^m_1 \pi \left( \omega - \E[\hat{p}_i]\right).
-\end{multline*}
-Combining these expressions, we have that
-$$ \hat{\beta}^m_1 \xrightarrow{P} \frac{\beta^m_1 \pi (\omega - \E[\hat{p}_i])}{\E[\hat{p}_i](1 - \E[\hat{p}_i])} = \beta^m_1 \gamma(\beta^g_1, c, \pi).$$
-\end{itemize}
-
-\subsection{Re-expressing $\gamma$ in a simpler form}\label{sec:simplication}
-We rewrite the attenuation fraction $\gamma$ in a way that makes it more amenable to theoretical analysis. We leverage the fact that $f$ integrates to unity and is even. We have that
-\begin{multline}\label{thm:gamma_expression_1} \E\left[\hat{p}_i\right] = (1 - \pi) \bar{\Phi}(c - \beta_0^g) + \pi \bar{\Phi}(c - \beta^g_0 - \beta^g_1) \\ = (1 - \pi) \Phi(\beta_0^g - c) + \pi\Phi(\beta^g_0 + \beta^g_1 - c), \end{multline}
- and so \begin{multline}\label{thm:gamma_expression_2} 1 - \E\left[\hat{p}_i\right] = (1 - \pi) + \pi - \E[\hat{p}_i]  = (1-\pi) \left(1 - \bar{\Phi}(c - \beta_0^g)\right)  + \pi \left(1 - \bar{\Phi}(c - \beta^g_0 - \beta^g_1) \right) \\ = (1 - \pi)\Phi(c - \beta^g_0) + \pi \Phi(c - \beta_0^g - \beta_1^g).
-\end{multline}
-Next,
-\begin{equation}\label{thm:gamma_expression_3}
-\omega = \Phi(\beta^g_1 + \beta^g_0 - c),\end{equation} and so
-\begin{multline}\label{thm:gamma_expression_4}
-\omega - \E[\hat{p}_i] = \Phi(\beta^g_1 + \beta^g_0 - c) - (1-\pi)\Phi(\beta^g_0 - c) - \pi \Phi(\beta^g_0 + \beta^g_1 - c)  \\ (1-\pi)\Phi(\beta^g_1 + \beta^g_0 - c)  - (1-\pi)\Phi(\beta^g_0 - c).
-\end{multline}
-
-Combining (\ref{thm:gamma_expression_1}, \ref{thm:gamma_expression_2}, \ref{thm:gamma_expression_3}, \ref{thm:gamma_expression_4}), we find that
-\begin{multline}\label{gamma_alternative}
-\gamma(\beta^g_1, c, \pi) = \frac{\pi(\omega - \E[\hat{p}_i])}{\E[\hat{p}_i](1 - \E[\hat{p}_i])} \\ = \frac{\pi \left[(1 - \pi) \Phi(\beta_0^g + \beta_1^g - c) - (1 - \pi) \Phi(\beta_0^g - c)\right]}{\left[(1-\pi)\Phi(\beta_0^g - c) + \pi \Phi(\beta^g_0 + \beta^g_1 - c) \right] \left[(1 - \pi) \Phi(c - \beta^g_0) + \pi\Phi(c - \beta_0^g - \beta_1^g) \right]}.
-\end{multline}
-As a corollary, when $\pi = 1/2$,
-\begin{multline}\label{gamma_alternative_pi_half}
-\gamma(\beta^g_1, c, 1/2)  \\ = \frac{\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c) }{\left[\Phi(\beta_0^g - c) +\Phi(\beta^g_0 + \beta^g_1 - c)\right] \left[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g) \right]}.
-\end{multline}
-Recalling the definitions of $g$ (\ref{def_g}) and $h$ (\ref{def_h}), we can write $\gamma$ as
-$$ \gamma(\beta^g_1, c, \pi) = \frac{\pi g(\beta^g_1, c, \pi)}{h(\beta^g_1, c,\pi)}.$$
-The special case (\ref{gamma_alternative_pi_half}) is identical to
-\begin{equation}\label{gamma_alt2_pi_half}
-\gamma(\beta^g_1, c, 1/2) = \frac{(4)(1/2)g(\beta^g_1, c, 1/2)}{4 h(\beta^g_1, c, 1/2)} = \frac{2 g(\beta^g_1, c, 1/2)}{4h(\beta^g_1, c, 1/2)},
-\end{equation}
-i.e., the numerator and denominator of  (\ref{gamma_alt2_pi_half}) coincide with those of (\ref{gamma_alternative_pi_half}).
-We sometimes will use the notation $2\cdot g$ and $4\cdot h$ to refer to the numerator and denominator of (\ref{gamma_alternative_pi_half}), respectively.
-
-\subsection{Derivatives of  $g$ and $h$ in $c$}\label{sec:derivatives}
-We compute the derivatives of $g$ and $h$ in $c$, which we will need to prove subsequent results. First, by FTC and the evenness of $f$, we have that
-\begin{multline}\label{dg_dc}
-D_2 g(\beta^g_1, c, \pi) = -(1-\pi)f( \beta^g_0 + \beta^g_1 - c ) + (1-\pi) f(\beta^g_0 - c) \\ = (1-\pi) f(c - \beta^g_0) - (1-\pi)f(c - \beta^g_0 - \beta^g_1).
-\end{multline}
-Second, we have that
-\begin{multline}\label{dh_dc}
-D_2 h(\beta^g_1, c, \pi) \\ = -[(1-\pi)f(\beta^g_0 - c) + \pi f( \beta^g_0 + \beta^g_1 - c )]\left[(1-\pi)\Phi(c - \beta^g_0) + \pi \Phi(c - \beta_0^g - \beta_1^g)  \right] \\ + [(1-\pi) f(c - \beta^g_0) +  \pi f(c - \beta^g_0 - \beta^g_1)] \left[(1-\pi) \Phi(\beta_0^g - c) + \pi \Phi(\beta^g_0 + \beta^g_1 - c) \right] \\ = \left[ (1-\pi) f(c - \beta^g_0) +  \pi f(c - \beta^g_0 - \beta^g_1) \right] \cdot \\ \bigg[ (1-\pi) \Phi(\beta_0^g - c) + \pi\Phi(\beta^g_0 + \beta^g_1 - c) \\ - (1-\pi) \Phi(c - \beta^g_0) - \pi \Phi(c - \beta_0^g - \beta_1^g) \bigg].
-\end{multline}
-
-\subsection{Limit of $\gamma$ in $c$}\label{sec:c_limit}
-
-Assume (without loss of generality) that $\beta^g_1 > 0$.
-We compute $\lim_{c \to \infty} \gamma(\beta^g_1, c, \pi)$.
-Observe that $$\lim_{c \to \infty} g(\beta^g_1, c, \pi) = \lim_{c \to \infty} h(\beta^g_1, c, \pi)  = 0.$$ Therefore, we can apply L'H\^{o}pital's rule.
-We have by (\ref{dg_dc}) and (\ref{dh_dc}) that \begin{multline}\label{c_limit_product}
-\lim_{c \to \infty} \gamma(\beta^g_1, c, \pi) = \lim_{c \to \infty} \frac{\pi D_2 g(\beta^g_1, c, \pi)}{D_2h(\beta^g_1, c, \pi)} \\ = \lim_{c \to \infty} \bigg\{ \frac{(1-\pi) f(c - \beta^g_0) + \pi f(c - \beta^g_0 - \beta^g_1)}{\pi (1-\pi) f(c - \beta^g_0) - \pi (1-\pi)f(c - \beta^g_0 - \beta^g_1)} \\ \cdot \bigg[ (1-\pi) \Phi(\beta_0^g - c) + \pi \Phi(\beta^g_0 + \beta^g_1 - c) \\ - (1-\pi) \Phi(c - \beta^g_0) - \pi \Phi(c - \beta_0^g - \beta_1^g) \bigg] \bigg\}^{-1}.
- \end{multline}
- We evaluate the two terms in the product (\ref{c_limit_product}) separately. Dividing by $f(c - \beta^g_0 - \beta^g_1) > 0$, we see that
- \begin{equation}\label{c_limit_product_2}
- \frac{(1-\pi) f(c - \beta^g_0) + \pi f(c - \beta^g_0 - \beta^g_1)}{\pi (1-\pi) f(c - \beta^g_0) - \pi (1-\pi)f(c - \beta^g_0 - \beta^g_1)} = \frac{\frac{(1-\pi) f(c - \beta^g_0)}{ f(c - \beta^g_0 - \beta^g_1)} + \pi}{\frac{ \pi(1-\pi) f(c - \beta^g_0)}{ f(c - \beta^g_0 - \beta^g_1)} - \pi(1-\pi)}.
- \end{equation}
- To evaluate the limit of (\ref{c_limit_product_2}), we first evaluate the limit of
- \begin{multline}\label{c_limit_product_3}
- \frac{f(c - \beta^g_0)}{f(c - \beta^g_0 - \beta^g_1)} = \frac{\exp{[-(1/2)(c - \beta_0^g)^2]}}{\exp{[-(1/2)( c - \beta^g_0 - \beta^g_1)^2]}} \\ = \frac{\exp[ -(1/2)(c^2 - 2 c \beta^g_0 + (\beta^g_0)^2)]}{\exp\left[-(1/2)( c^2 - 2c \beta^g_0 - 2 c \beta^g_1 + (\beta^g_0)^2 + 2( \beta^g_0 \beta^g_1) + (\beta^g_1)^2)\right]} \\ = \exp\big[-c^2/2 + c \beta^g_0 - (\beta^g_0)^2/2 \\ + c^2/2 - c \beta^g_0 - c \beta^g_1 + (\beta^g_0)^2/2 + \beta^g_0 \beta^g_1 + (\beta^g_1)^2/2 \big] \\ = \exp[ -c \beta^g_1 + \beta^g_0 \beta^g_1 + (\beta^g_1)^2/2] = \exp[ \beta^g_0 \beta^g_1 + (\beta^g_1)^2/2]\exp[ -c \beta^g_1].
-\end{multline}
-Taking the limit in (\ref{c_limit_product_3}), we obtain
-$$
-\lim_{c \to \infty} \frac{f(c - \beta^g_0)}{f(c - \beta^g_0 - \beta^g_1)} = \exp[ \beta^g_0 \beta^g_1 + (\beta^g_1)^2/2] \lim_{c \to \infty} \exp[ -c \beta^g_1] = 0
-$$ for $\beta^g_1 > 0$. We now can evaluate the limit of (\ref{c_limit_product_2}):
-$$ \lim_{c \to \infty} \frac{(1-\pi) f(c - \beta^g_0) + \pi f(c - \beta^g_0 - \beta^g_1)}{\pi (1-\pi) f(c - \beta^g_0) - \pi (1-\pi)f(c - \beta^g_0 - \beta^g_1)} = \frac{-\pi}{\pi(1-\pi)} = -\frac{1}{1 -\pi}.$$ Next, we compute the limit of the other term in the product (\ref{c_limit_product}):
-\begin{multline}\label{c_limit_product_4}
-\lim_{c \to \infty} \bigg[ (1-\pi)\Phi(\beta_0^g - c) + \pi \Phi(\beta^g_0 + \beta^g_1 - c) \\ - (1-\pi)\Phi(c - \beta^g_0) - \pi \Phi(c - \beta_0^g - \beta_1^g) \bigg] = -(1-\pi) - \pi = -1.
-\end{multline}
-Combining (\ref{c_limit_product_2}) and (\ref{c_limit_product_4}), the limit (\ref{c_limit_product}) evaluates to
-$$ \lim_{c \to \infty} \gamma(\beta^g_1, c, \pi) = \left(  \frac{ 1 }{ 1 - \pi }\right)^{-1} = 1 - \pi.$$ It follows that the limit in $c$ of the asymptotic relative bias $b$ is
-$$\lim_{c \to \infty} b(\beta^g_1, c, \pi) = 1 - \lim_{c \to \infty} \gamma(\beta^g_1, c, \pi) = \pi.$$
-A corollary is that
-$$\lim_{c \to \infty} b(\beta^g_1, c, 1/2) = 1/2.$$
-
-\subsection{Bayes-optimal decision boundary as a critical value of $\gamma$}\label{sec:bayes_opt}
-Let $c_\textrm{bayes} = \beta^g_0 + (1/2)\beta^g_1.$ We show that $c = c_\textrm{bayes}$ is a critical value of $\gamma$ for $\pi = 1/2$ and given $\beta^g_1$, i.e, $$D_2 \gamma (\beta^g_1, c_\textrm{bayes}, 1/2) = 0.$$ Differentiating (\ref{gamma_alt2_pi_half}), the quotient rule implies that
-\begin{equation}\label{quotient_rule}
-D_2\gamma(\beta^g_1, c, 1/2) \\ = \frac{D_2[2g(\beta^g_1, c, 1/2)] 4h(\beta^g_1, c, 1/2) - 2g(\beta^g_1, c, 1/2) D_2[4h(\beta^g_1, c, 1/2)]}{[4h(\beta^g_1, c, \pi)]^2}.
-\end{equation}
-We have by (\ref{dg_dc}) that
-\begin{equation}\label{dg_dc_bayes}
-D_2[2g(\beta^g_1, c_\textrm{bayes}, 1/2)] = f( \beta^g_1/2) - f( -\beta^g_1/2) = f(\beta^g_1/2) - f(\beta^g_1/2) = 0.
-\end{equation}
-Similarly, we have by (\ref{dh_dc}) that
-\begin{multline}\label{dh_dc_bayes}
-D_2[4 h(\beta^g_1, c_\textrm{bayes}, \pi)] = [f( \beta^g_1/2) + f( -\beta^g_1/2)] \cdot \\ \left[  \Phi(-\beta^g_1/2) + \Phi(\beta^g_1/2) -  \Phi(\beta^g_1/2) - \Phi(-\beta^g_1/2) \right] = 0.
-\end{multline}
-Plugging in (\ref{dh_dc_bayes}) and (\ref{dg_dc_bayes}) to (\ref{quotient_rule}), we find that 
-$$D_2[\gamma(\beta^g_1, c_\textrm{bayes}, 1/2)] = 0.$$ Finally, because
-$$b(\beta^g_1, c, 1/2) = 1 - \gamma(\beta^g_1, c, 1/2),$$ it follows that
-$$D_2[b(\beta^g_1, c_\textrm{bayes}, 1/2)] = -D_2[\gamma(\beta^g_1, c_\textrm{bayes}, 1/2)] = 0.$$
-
-\subsection{Comparing Bayes-optimal decision boundary and large threshold}\label{sec:comparison}
-
-We compare the bias produced by setting the threshold to a large number to the bias produced by setting the threshold to the Bayes-optimal decision boundary.
-Let $r: \R^{\geq 0} \to \R$ be the value of attenuation function evaluated at the Bayes-optimal decision boundary $c_\textrm{bayes} = \beta^g_0 + (1/2) \beta^g_1$, i.e.
-\begin{multline*}
-r(\beta^g_1) = \gamma(\beta^g_1, \beta^g_0 + (1/2)\beta^g_1, 1/2) = \frac{\Phi(\beta^g_1/2) - \Phi(-\beta^g_1/2)}{\left[\Phi(-\beta^g_1/2) + \Phi( \beta^g_1/2) \right] \left[\Phi(\beta^g_1/2) + \Phi( -\beta^g_1/2)\right]} \\ = \frac{\int_{-\beta^g_1/2}^{\beta^g_1/2} f}{\left[ 1 - \Phi(\beta^g_1/2) + \Phi(\beta^g_1/2) \right]\left[ \Phi(\beta^g_1/2) + 1 - \Phi(\beta^g_1/2) \right]} = 2 \int_{0}^{\beta^g_1/2} f = 2 \Phi(\beta^g_1/2) - 1.
-\end{multline*}
-We set $r$ to $1/2$ and solve for $\beta^g_1$:
-\begin{multline*}
-r(\beta^g_1) = 1/2 \iff 2\Phi(\beta^g_1/2) -1 = 1/2 \\ \iff \Phi(\beta^g_1/2) = 3/4 \iff \beta^g_1 = 2 \Phi^{-1}(3/4) \approx 1.35.
-\end{multline*}
-Because $r$ is a strictly increasing function, it follows that $r(\beta^g_1) < 1/2$ for $\beta^g_1 < 2\Phi^{-1}(3/4)$ and $r(\beta^g_1) > 1/2$ for $\beta^g_1 > 2\Phi^{-1}(3/4).$ Next, because $$b(\beta^g_1, c_\textrm{bayes}, 1/2) = 1 - \gamma(\beta^g_1, c_\textrm{bayes}, 1/2) = 1 - r(\beta^g_1),$$ we have that $b(\beta^g_1, c_\textrm{bayes}, 1/2) > 1/2$ for $\beta^g_1 < 2 \Phi^{-1}(3/4)$ and $b(\beta^g_1, c_\textrm{bayes}, 1/2) < 1/2$ for $\beta^g_1 > 2 \Phi^{-1}(3/4)$.
-Recall that the bias induced by sending the threshold to infinity (as stated in Proposition \ref{prop:c_limit_half} and proven in Section \ref{sec:c_limit}) is $1/2$, i.e.
-$$b(\beta^g_1, \infty, 1/2) = 1/2.$$ We conclude that $b(\beta^g_1, c_\textrm{bayes},1/2) > b(\beta^g_1, \infty, 1/2)$ on $\beta^g_1 \in [0, 2\Phi^{-1}(3/4))$; $b(\beta^g_1, c_\textrm{bayes},1/2) = b(\beta^g_1, \infty, 1/2)$ for $\beta^g_1 = 2\Phi^{-1}(3/4)$; and $b(\beta^g_1, c_\textrm{bayes},1/2) < b(\beta^g_1, \infty, 1/2)$ on $\beta^g_1 \in (2\Phi^{-1}(3/4), \infty)$.
-
-\subsection{Monotonicity in $\beta^g_1$}\label{sec:monotone}
-We show that $\gamma$ is monotonically increasing in $\beta^g_1$ for $\pi = 1/2$ and given threshold $c$.
-We begin by stating and proving two lemmas.
-The first lemma establishes an inequality that will serve as the basis for the proof.
-
-\begin{lemma}
-The following inequality holds: 
-\begin{multline}\label{basic_ineq_cp}
-\left[\Phi(\beta^g_0 - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right] \\ \cdot \left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c) + \Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g) \right] \\ \geq \left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c)\right]\left[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g)\right].
-\end{multline}
-\end{lemma}
-\textbf{Proof}: We take cases on the sign on $\beta^g_1$.
-
-\noindent
-\underline{Case 1}: $\beta^1_g < 0$.
-Then $ \beta^g_1 + (\beta^g - c) < (\beta^g_0 - c),$ implying $\Phi(\beta^g_0 + \beta^g_1 - c) < \Phi(\beta^g_0 - c),$ or $[\Phi(\beta^g_0 + \beta^g_1 - c) - \Phi(\beta^g_0 - c)] < 0.$ Moreover, $[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g)]$ is positive.
-Therefore, the right-hand side of (\ref{basic_ineq_cp}) is negative.
-
-Turning our attention of the left-hand side of (\ref{basic_ineq_cp}), we see that
-\begin{equation}\label{basic_ineq_cp_2}
-\Phi(\beta^g_0 + \beta^g_1 - c) + \Phi( c - \beta^g_0 - \beta^g_1) = 1 -\Phi(\beta^g_0 + \beta^g_1 - c) + \Phi( c - \beta^g_0 - \beta^g_1) = 1.
-\end{equation}
-Additionally, $\Phi(\beta^g_0 - c) < 1$ and $ \Phi(c - \beta^g_0) > 0$. Combining these facts with (\ref{basic_ineq_cp_2}), we find that
-$$ \left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c) + \Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g) \right] > 0.
-$$ Finally, because $\left[\Phi(\beta^g_0 - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right] > 0,$ the entire left-hand side of (\ref{basic_ineq_cp}) is positive.
-The inequality holds for $\beta^g_1 < 0$.
-
-\noindent
-\underline{Case 2}: $\beta^1_g \geq 0.$  We will show that the first term on the LHS of (\ref{basic_ineq_cp}) is greater than the first term on the RHS of (\ref{basic_ineq_cp}), and likewise that the second term on the LHS is greater than the second term on the RHS, implying the truth of the inequality.
-Focusing on the first term, the positivity of $\Phi(\beta^g_0 -c)$ implies that
-$$ \Phi(\beta^g_0 - c) \geq - \Phi(\beta^g_0 - c),$$ and so
-$$ \Phi(\beta^g_0 - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \geq \Phi(\beta^g_0 - \beta^g_1 - c) - \Phi(\beta^g_0 - c).$$
-Next, focusing on the second term, $\beta^g_1 \geq 0$ implies that 
-\begin{equation}\label{basic_ineq_cp_3}
-\beta^g_1 + \beta^g_0 - c \geq \beta^g_0 - c \implies \Phi(\beta^g_1 + \beta^g_0 - c) - \Phi(\beta^g_0 - c) \geq 0.
-\end{equation}
-Adding $\Phi(c - \beta^g_0) + \Phi(c - \beta^g_0 - \beta^g_1)$ to both sides of (\ref{basic_ineq_cp_3}) yields
-\begin{multline*}
-\Phi(\beta^g_1 + \beta^g_0 - c) - \Phi(\beta^g_0 - c) + \Phi(c - \beta^g_0) + \Phi(c - \beta^g_0 - \beta^g_1) \\ \geq \Phi(c - \beta^g_0) + \Phi(c - \beta^g_0 - \beta^g_1). \textrm{ }
-\end{multline*}
-The inequality holds for $\beta^g_1 \geq 0$. Combining the cases, the inequality holds for all $\beta^g_1 \in \R$. $\square$
-
-The second lemma establishes the derivatives of the functions $2\cdot g$ and $4 \cdot h$ in $\beta^g_1$.
-\begin{lemma}
-The derivatives in $\beta^g_1$ of $2\cdot g$ and $4\cdot h$ are
-\begin{equation}\label{dg_dbeta}
-\textcolor{violet}{D_1[2g(\beta^g_1, c, 1/2)] = f(\beta^g_0 + \beta^g_1 - c)}
-\end{equation}
-and
-\begin{multline}\label{dh_dbeta}
-\textcolor{teal}{D_1[4h(\beta^g_1, c, 1/2)] = f(\beta^g_0 + \beta^g_1 - c) \left[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g) \right]} \\ \textcolor{teal}{- f(\beta^g_0 + \beta^g_1 - c) \left[\Phi(\beta_0^g - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right]}.\end{multline}
-\end{lemma}
-\textbf{Proof}: Apply FTC and product rule. $\square$
-
-We are ready to prove the monotonicity of $\gamma$ in $\beta^g_1$.
-Subtracting $$\left[\Phi(\beta_0^g - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right]\left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c)\right]$$ from both sides of (\ref{basic_ineq_cp}) and multiplying by $f(\beta^g_0 + \beta^g_1 - c) > 0$ yields
-\begin{multline}\label{basic_ineq_cp_4}
-\textcolor{violet}{f(\beta^g_0 + \beta^g_1 - c)} \textcolor{red}{ \left[\Phi(\beta^g_0 - c) + \Phi\left(\beta^g_0 + \beta^g_1 - c \right) \right] \left[ \Phi(c - \beta^g_0) + \Phi(c - \beta^g_0 - \beta^g_1) \right]}  \\ \geq \textcolor{teal}{f(\beta^g_0 + \beta^g_1 - c) \left[\Phi(c - \beta^g_0) + \Phi(c - \beta_0^g - \beta_1^g)\right]}\textcolor{blue}{\left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c)\right]} \\ -\textcolor{teal}{f(\beta^g_0 + \beta^g_1 - c)   \left[\Phi(\beta_0^g - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right]} \textcolor{blue}{\left[\Phi(\beta_0^g + \beta_1^g - c) - \Phi(\beta_0^g - c)\right]}.
-\end{multline}
-Next, recall that
-\begin{equation}\label{def_2g}
-\textcolor{blue}{2g(\beta^g_1,c,1/2) = \Phi(\beta^g_0 + \beta^g_1 - c) - \Phi(\beta^g_0 - c)}.
-\end{equation}
-and
-\begin{equation}\label{def_4h}
-\textcolor{red}{4h(\beta^g_1, c, 1/2) = \left[ \Phi(\beta^g_0 - c) + \Phi(\beta^g_0 + \beta^g_1 - c) \right] \left[\Phi(c - \beta^g_0) + \Phi( c - \beta^g_0 - \beta^g_1) \right]}.
-\end{equation}
-Substituting (\ref{dg_dbeta}, \ref{dh_dbeta}, \ref{def_2g}, \ref{def_4h}) into (\ref{basic_ineq_cp_4}) produces
-\begin{equation*}
-\textcolor{violet}{D_1[2g(\beta^g_1, c, 1/2)]}\textcolor{red}{4h(\beta^g_1, c, 1/2)} \geq \textcolor{blue}{2g(\beta^g_1, c, 1/2)}\textcolor{teal}{D_1[4h(\beta^g_1, c, 1/2)]},
-\end{equation*}
-or 
-\begin{equation}\label{basic_ineq_cp_5}
-\textcolor{violet}{D_1[2g(\beta^g_1, c, 1/2)]}\textcolor{red}{4h(\beta^g_1, c, 1/2)} - \textcolor{blue}{2g(\beta^g_1, c, 1/2)}\textcolor{teal}{D_1[4h(\beta^g_1, c, 1/2)]} \geq 0.
-\end{equation}
-The quotient rule implies that
-\begin{multline}\label{d_gamma_d_beta}
-D_1 \gamma(\beta^g_1, c, 1/2) \\ = \frac{ \textcolor{violet}{D_1[2g(\beta^g_1, c, 1/2)]}\textcolor{red}{4h(\beta^g_1, c, 1/2)} - \textcolor{blue}{2g(\beta^g_1, c, 1/2)}\textcolor{teal}{D_1[4h(\beta^g_1, c, 1/2)]} }{[4h(\beta^g_1, c, 1/2)]^2}.
-\end{multline}
-We conclude by (\ref{basic_ineq_cp_5}) and (\ref{d_gamma_d_beta}) that $\gamma$ is monotonically increasing in $\beta^g_1$.
-Finally, $b(\beta^g_1, c, \pi) = 1 - \gamma(\beta^g_1, c, \pi)$ is monotonically decreasing in $\beta^g_1$.
-
-\subsection{Strict attenuation bias}\label{sec:att_bias}
-
-We begin by computing the limit of $\gamma$ in $\beta^g_1$ given $\pi = 1/2$.  First,
-\begin{multline*}
-\lim_{\beta^g_1 \to \infty} \gamma(\beta^g_1, c, 1/2) = \frac{1 - \Phi(\beta^g_0 - c)}{\left[1 + \Phi(\beta^g_0 - c) \right] \left[\Phi(c - \beta^g_0) \right]} \\ = \frac{\Phi(c - \beta^g_0)}{ \left[1 + \Phi(\beta^g_0 - c) \right] \left[\Phi(c - \beta^g_0) \right]} = \frac{1}{1 + \Phi(\beta^g_0 - c)} < 1.
-\end{multline*}
-Similarly,
-\begin{multline*}
-\lim_{\beta^g_1 \to -\infty} \gamma(\beta^g_1, c, 1/2) = \frac{ - \Phi(\beta^g_0 - c)}{\left[\Phi(\beta^g_0 - c)\right] \left[\Phi(c - \beta^g_0) + 1 \right]} = \frac{-1}{1 + \Phi(c - \beta^g_0)} > -1.
-\end{multline*}
-The function $\gamma(\beta^g_1, c, 1/2, \beta^g_0)$ is monotonically increasing in $\beta^g_1$ (as stated in Proposition \ref{prop:monotonic} and proven in section \ref{sec:monotone}).
-It follows that
-$$-1 < -\frac{1}{1 + \Phi(c - \beta^g_0)} \leq \gamma(\beta^g_1, c, 1/2, \beta^g_0) \leq \frac{1}{1 - \Phi(\beta^g_0 - c)} < 1$$ for all $\beta^g_1 \in \R$.
-But $\beta^g_0$ and $c$ were chosen arbitrarily, and so
-$$-1 < \gamma(\beta^g_1, c, 1/2, \beta^g_0) < 1$$ for all $(\beta^g_1, c, \beta^g_0) \in \R^3$.
-Finally, because $b(\beta^g_1, c, 1/2, \beta^g_0) = 1 - \gamma(\beta^g_1, c, 1/2, \beta^g_0)$, it follows that
-$$ 0 < b(\beta^g_1, c, 1/2, \beta^g_0) < 2$$ for all $(\beta^g_1, c, \beta^g_0) \in \R^3$
-
-\subsection{Bias-variance decomposition in no-intercept model}\label{sec:bv_decomp}
-
-We prove the bias-variance decomposition for the no-intercept model (\ref{theoretical_model_no_int}). Define $l$ (for ``limit'') by
-$$l = \beta_m \left(\frac{\omega \pi}{\zeta(1-\pi) + \omega \pi}\right),$$ where
-$$
-\begin{cases}
-\omega = \bar{\Phi}(c - \beta_g) = \Phi(\beta_g - c) \\
-\zeta = \bar{\Phi}(c) = \Phi(-c).
-\end{cases}
-$$
-We have that
-\begin{equation*}
-\hat{\beta}_m - l = \frac{\sum_{i=1}^n \hat{p}_i m_i}{ \sum_{i=1}^n \hat{p}^2_i} - l = \frac{\sum_{i=1}^n \hat{p}_i m_i}{ \sum_{i=1}^n \hat{p}^2_i} - \frac{l \sum_{i=1}^n \hat{p}_i^2 }{ \sum_{i=1}^n \hat{p}_i^2} \\ = \frac{\sum_{i=1}^n \hat{p}_i(m_i - l \hat{p}_i)}{ \sum_{i=1}^n \hat{p}_i^2}.
-\end{equation*}
-Therefore,
-\begin{equation}\label{bc_decomp_1}
-\sqrt{n}(\hat{\beta}_m - l) = \frac{(1/\sqrt{n})\sum_{i=1}^n \hat{p}_i(m_i - l \hat{p}_i)}{(1/n)\sum_{i=1}^n \hat{p}_i^2}.
-\end{equation}
-Next, we compute the expectation and variance of $\hat{p}_i(m_i - l\hat{p}_i)$. To do so, we first compute several simpler quantities:
-\begin{enumerate}
-\item Expectation of $\hat{p}_i$: 
-\begin{multline*}
-\E[\hat{p}_i] = \P(p_i\beta_g + \tau_i \geq c) =  \P(\beta_g + \tau_i \geq c)\pi + \P(\tau_i \geq c)(1-\pi) \\ = \pi \omega + (1-\pi)\zeta.
-\end{multline*}
-\item Expectation of $\hat{p}_i p_i$: $$\E\left[\hat{p}_i p_i\right] = \E\left[\hat{p}_i | p_i = 1 \right]\P\left[p_i = 1\right] = \omega \pi.$$
-\item Expectation of $\hat{p}_i m_i$:
-\begin{multline*}
-\E[\hat{p}_i m_i] = \E\left[\hat{p}_i(\beta_m p_i + \ep_i)\right] = \E\left[\beta_m \hat{p}_i p_i + \hat{p}_i \ep_i \right] \\ = \beta_m \E\left[ \hat{p}_i p_i \right] + \E[\hat{p}_i]\E[\ep_i] = \beta_m \omega \pi + 0 = \beta_m \omega \pi.
-\end{multline*}
-\item Expectation of $\hat{p}_i m_i^2$: \begin{multline*}
-\E\left[\hat{p}_i m_i^2\right] = \E \left[ \hat{p}_i( \beta_m p_i + \ep_i )^2 \right] = \E\left[ \hat{p}_i \left( \beta_m^2 p_i^2 + 2 \beta_m p_i \ep_i + \ep_i^2 \right)  \right] \\ = \E\left[ \hat{p}_i p_i \beta^2_m + 2 \beta_m p_i \hat{p}_i \ep_i + \hat{p}_i \ep_i^2 \right] = \beta^2_m \E[ \hat{p}_i p_i] + 2 \beta_m \E[p_i\hat{p}_i] \E[\ep_i] + \E[\hat{p}_i] \E[ \ep^2_i ] \\ = \beta^2_m \E[ \hat{p}_i p_i] + \E[\hat{p}_i] = \beta^2_m \omega \pi + \E[ \hat{p}_i].
-\end{multline*}
-\end{enumerate}
-
-Now, we can compute the expectation and variance of $\hat{p}_i(m_i - l\hat{p}_i)$. First,
-\begin{multline}\label{bv_decomp_2}
-\E\left[\hat{p}_i(m_i - l\hat{p}_i) \right] = \E[\hat{p}_i m_i] - l \E[\hat{p}_i] \\ = \beta_m \omega \pi - \left(\frac{\beta_m \omega \pi}{\zeta (1-\pi) + \omega \pi}\right)[\zeta (1-\pi) + \omega \pi] = 0.
-\end{multline}
-Additionally,
-\begin{multline}\label{bv_decomp_3}
-\V\left[\hat{p}_i(m_i - l\hat{p}_i)\right] = \E\left[\hat{p}_i^2(m_i - l\hat{p}_i)^2\right] - (\E\left[ \hat{p}_i(m_i - l\hat{p}_i)\right])^2 \\ = \E\left[ \hat{p}_i m_i^2\right] - 2l \E[m_i\hat{p}_i] +l^2 \E[\hat{p}_i]= \beta^2_m \omega \pi + \E[ \hat{p}_i] -2l \beta_m \omega \pi + l^2 \E[\hat{p}_i] \\ = \beta_m\omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2).
-\end{multline}
-Therefore, by CLT, (\ref{bv_decomp_2}), and (\ref{bv_decomp_3}),
-\begin{equation}\label{bv_decomp_4}
-(1/\sqrt{n})\sum_{i=1}^n \hat{p}_i(m_i - l \hat{p}_i) \xrightarrow{d} N\left(0, \beta_m\omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2) \right).
-\end{equation}
-Next, by weak LLN,
-\begin{equation}\label{bv_decomp_5}
-(1/n) \sum_{i=1}^n \hat{p}_i^2 = (1/n) \sum_{i=1}^n \hat{p}_i \xrightarrow{P} \E[\hat{p}_i].
-\end{equation}
-Finally, by (\ref{bc_decomp_1}), (\ref{bv_decomp_4}), (\ref{bv_decomp_5}), and Slutsky's Theorem,
-$$ \sqrt{n}(\hat{\beta}_m - l) \xrightarrow{d} N\left(0, \frac{ \beta_m\omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2) }{\left(\E[\hat{p}_i]\right)^2} \right).$$ Thus, for large $n \in \N$, we have that
-$$ \begin{cases}
-\E [\hat{\beta}_m] \approx l, \\
-\V[\hat{\beta}_m] \approx \left[\beta_m\omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2)\right]/[n\E^2[\hat{p}_i]],
-\end{cases}$$
-completing the bias-variance decomposition.
-
-\section{Estimation and inference in the GLM-EIV model}
-\subsection{Estimation}
-We estimate the parameters of the GLM-EIV model using an EM algorithm.
-
-\subsubsection*{E step}
-The E step entails computing the membership probability of each cell.
-Let $\theta^{(t)} = (\beta_m^{(t)}, \beta_g^{(t)}, \pi^{(t)})$ be the parameter estimate at the $t$-th iteration of the algorithm.
-For $k \in \{0,1\}$, let $[\eta^m_i(k)]^{(t)}$ be the $i$th canonical parameter at the $t$-th iteration of the algorithm of the gene expression distribution that results from setting $p_i$ to $k$, i.e.
-$$
-[\eta^m_i(k)]^{(t)} := h_m\left( \langle \tilde{x}_i(k) , \beta_m^{(t)} \rangle + o^m_i \right).
-$$ Similarly, let $\left[\eta^g_i(k)\right]^{(t)}$ be defined by
-$$\left[\eta^g_i(k)\right]^{(t)} :=  h_g\left( \langle \tilde{x}_i(k) , \beta_g^{(t)} \rangle + o^g_i \right).$$
-Next, for $k \in \{0,1\},$ define $\alpha^{(t)}_i(k)$ by
-\begin{multline*}
-\alpha^{(t)}_i(k) := \P\left( M_i = m_i, G_i = g_i | P_i = k, \theta^{(t)} \right) \\ = \P\left( M_i = m_i | P_i = k, \theta^{(t)} \right) \P\left(G_i = g_i | P_i = k, \theta^{(t)} \right) \textrm{ (because $G_i \indep M_i | P_i$)} \\ = f_m\left(m_i; \left[ \eta^m_i(k) \right]^{(t)}\right) f_g\left(g_i; \left[ \eta^g_i(k) \right]^{(t)} \right).
-\end{multline*}
-Finally, let 
-$$
-\begin{cases}
-\pi^{(t)}(1) := \pi^{(t)} = \P\left(P_i = 1 | \theta^{(t)} \right) \\
-\pi^{(t)}(0) := 1 - \pi^{(t)} = \P\left(P_i = 0 | \theta^{(t)} \right).
-\end{cases}
-$$
-The $i$th membership probability $T^{(t)}_i(1)$ is
-\begin{multline}\label{e_step_1}
-T^{(t)}_i(1) = \P(P_i = 1 | M_i = m_i, G_i = g_i, \theta^{(t)})  = \frac{\pi^{(t)}(1) \alpha^{(t)}_i(1)}{ \sum_{k=0}^1 \pi^{(t)}(k) \alpha^{(t)}_i(k)} \textrm{ (by Bayes rule)} \\ = \frac{1}{\frac{ \pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)} + 1} = \frac{1}{ \exp\left(\log\left(\frac{\pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)}\right)\right) + 1} = \frac{ 1 }{ \exp\left(q^{(t)}_i\right) + 1},
-\end{multline}
-where we set 
-\begin{equation}\label{e_step_2}
-q_i^{(t)} := \log\left(\frac{\pi^{(t)}(0) \alpha_i^{(t)}(0)}{\pi^{(t)}(1) \alpha_i^{(t)}(1)}\right).
-\end{equation}
-Next, we have that
-\begin{multline*}
-q^{(t)}_i = \log\left[ \pi^{(t)}(0) \right] + \log\left[ f_m\left(m_i; \left[ \eta^m_i(0) \right]^{(t)}\right) \right] + \log\left[ f_g\left(g_i; \left[ \eta^g_i(0) \right]^{(t)}\right) \right] \\ - \log\left[ \pi^{(t)}(1) \right] - \log\left[ f_m\left(m_i; \left[ \eta^m_i(1) \right]^{(t)}\right) \right] - \log\left[ f_g\left(g_i; \left[ \eta^g_i(1) \right]^{(t)}\right) \right],
-\end{multline*}
-which we can compute. We therefore conclude that
-$$ T_i^{(t)} = \frac{1}{\exp\left(q^{(t)}_i\right) + 1}.$$
-\subsection*{M step}
-Recall that the log-likelihood (\ref{full_log_lik}) of the GLM-EIV model  is
-\begin{multline*}
-\mathcal{L}(\theta; m, g, p) = \sum_{i=1}^n \left[ p_i \log(\pi) + (1-p_i) \log(1-\pi) \right] \\ + \sum_{i=1}^n \log\left( f_m(m_i; \eta^m_i)\right) + \sum_{i=1}^n \log\left( f_g(g_i; \eta_i^g) \right).
-\end{multline*}
-Define $Q(\theta | \theta^{(t)}) = \E_{\left(P |M = m, G = g, \theta^{(t)}\right)}\left[ \mathcal{L}(\theta; m, g, p) \right].$ We have that
-\begin{multline}\label{Q_funct}
-Q(\theta |\theta^{(t)}) = \sum_{i=1}^n \left[T^{(t)}_i(1)\log(\pi) + T_i^{(t)}(0) \log(1 - \pi)\right] \\ + \sum_{k=0}^1 \sum_{i=1}^n T^{(t)}_i(k) \log \left[ f_m(m_i; \eta_i^m(k)) \right] + \sum_{k=0}^1 \sum_{i=1}^n T^{(t)}_i(k) \log \left[ f_g( g_i; \eta^{g,b}_i(k)) \right].
-\end{multline}
-The three terms of (\ref{Q_funct}) are functions of different parameters: the first is a function of $\pi,$ the second is a function of $\beta_m,$ and the third is a function of $\beta_g$.
-Therefore, to find the maximizer $\theta^{(t+1)}$ of (\ref{Q_funct}), we maximize the three terms separately.
-Differentiating the first term with respect to $\pi$, we find that
-	\begin{multline*}
-\frac{ \partial }{\partial \pi } \sum_{i=1}^n \left[ T^{(t)}_i(1)\log(\pi) + T_i^{(t)}(0) \log(1 - \pi)\right]  \\ = \frac{\sum_{i=1}^n T_i^{(t)}(1)}{\pi} - \frac{ \sum_{i=1}^n T_i^{(t)}(0) }{ 1 - \pi}.
-\end{multline*} Setting the derivative equal to $0$ and solving for $\pi$,
-\begin{multline*}
-\frac{\sum_{i=1}^n T_i^{(t)}(1)}{\pi} - \frac{ \sum_{i=1}^n T_i^{(t)}(0) }{ 1 - \pi} = 0 \iff \sum_{i=1}^n T_i^{(t)}(1) - \pi \sum_{i=1}^n T^{(t)}_i(1) = \pi \sum_{i=1}^n T_i(0) \\ \iff \sum_{i=1}^n T^{(t)}_i(1) - \pi\sum_{i=1}^n T_i^{(t)}(1) = \pi n - \pi\sum_{i=1}^n T_i(1) \iff \pi = \frac{ \sum_{i=1}^n T_i^{(t)} (1) }{n}.\end{multline*}
-Thus, the maximizer $\pi^{(t+1)}$ of (\ref{Q_funct}) in $\pi$ is $\pi^{(t+1)} = (1/n)\sum_{i=1}^n T^{(t)}_i(1)$.
-Next, define $w^{(t)} = [T^{(t)}_1(0), \dots, T^{(t)}_n(0), T^{(t)}_1(1), \dots, T^{(t)}_n(1)]^T \in \R^{2n}$.
-We can view the second term of (\ref{Q_funct}) as the log-likelihood of a GLM -- call it $\textrm{GLM}^{(t)}_m$ -- that has exponential family density $f_m$, link function $r_m$, responses $[m,m]^T$, offsets $[o^m, o^m]^T$, weights $w^{(t)}$, and design matrix $\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix}.$ Therefore, the maximizer $\beta^{(t+1)}_m$ of the second term of (\ref{Q_funct}) is the maximizer of $\textrm{GLM}^{(t)}_m$, which we can compute using the iteratively reweighted least squares (IRLS) procedure, as implemented in R's GLM function.
-Similarly, the maximizer $\beta^{(t+1)}_g$ of the third term of (\ref{Q_funct}) is the maximizer of the GLM with exponential family density $f_g$, link function $r_g$, responses $[g,g]^T$, offsets $[o^g, o^g]^T$, weights $w^{(t)}$, and design matrix $\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix}.$
-
-\subsection{Inference}
-We derive the asymptotic observed information matrix of the GLM-EIV log likelihood, enabling us to perform inference on the parameters.
-First, we define some notation.
-For $i \in \{1, \dots, n\}$, $j \in \{0, 1\}$, and $\theta = (\pi, \beta_m, \beta_g),$ let $T^\theta_i(j)$ be defined by
-$$T^\theta_i(j) = \P_\theta\left(P_i = j | M_i = m_i, G_i = g_i\right).$$ Let the $n \times n$ matrix $T^\theta(j)$ be given by $$T^\theta(j) = \textrm{diag}\left\{T^\theta_1(j), \dots, T^\theta_n(j)\right\}.$$
-Next, define the diagonal $n \times n$ matrices $\Delta^m$, $[\Delta']^m$, $V^m$, and $H^m$ by
-$$
-\begin{cases}
-	\Delta^m = \textrm{diag} \{h_m'(l_1^m), \dots, h_m'(l_n^m)\} \\
-	[\Delta']^m = \textrm{diag} \{h_m''(l_1^m), \dots, h_m''(l_n^m) \} \\
-	V^m = \textrm{diag} \{ \psi''_m( \eta^m_1), \dots, \psi''_m( \eta^m_n) \} \\
-	H^m = \textrm{diag} \{ m_1 - \mu_1^m, \dots, m_n - \mu_n^m\}.
-\end{cases} 
-$$ Define the $n \times n$ matrices $\Delta^g, [\Delta']^g, V^g,$ and $H^g$ analogously. These matrices are \textit{unobserved}, as they depend on $\{p_1, \dots, p_n\}$.
-
-For $j \in \{0,1\}$, let the diagonal $n \times n$ matrices $\Delta^m(j), [\Delta']^m(j), V^m(j),$ and $H^m(j)$ be given by
-$$\begin{cases}
-\Delta^m(j) = \textrm{diag} \{ h_m'(l_1^m(j)), \dots, h_m'( l_n^m(j) ) \} \\
-[\Delta']^m(j) = \textrm{diag} \{ h_m''(l_1^m(j)), \dots, h_m''( l_n^m(j)) \} \\
-V^m(j) = \textrm{diag} \{\psi''_m( \eta^m_1(j)), \dots, \psi''_m( \eta^m_n(j))\} \\
-H^m(j) = \textrm{diag} \{m_1 - \mu_1^m(j), \dots, m_n - \mu_n^m(j)\} .
-\end{cases}
-$$
-Define the matrices $\Delta^g(j)$, $[\Delta']^{g}(j)$, $V^g(j),$ and $H^g(j)$ analogously. Finally, define the vectors $s^m(j)$ and $w^m(j)$ in $\R^n$ by 
-$$ \begin{cases}
-s^m(j) = [m_1 - \mu_1^m(j), \dots, m_n - \mu_n^m(j) ]^T \\ w^m(j) = [ T_1(0)T_1(1)\Delta^m_1(j) H^m_1(j), \dots, T_n(0)T_n(1)\Delta_n^m(j) H_n^m(j)]^T,
-\end{cases} $$
-and let the vectors $s^g(j)$ and $w^g(j)$ be defined analogously.
-The quantities $\Delta^m(j), [\Delta']^m(j), V^m(j),$ $H^m(j),$ $s^m(j),$ $w^m(j),$ $\Delta^g(j), [\Delta']^g(j), V^g(j),$ $H^g(j),$ $s^g(j),$ and $w^g(j)$ are all \textit{observed}.
-
-The observed information matrix $J(\theta; m, g)$ evaluated at $\theta = (\pi, \beta_m, \beta_g)$ is the negative Hessian of the marginal log likelihood (\ref{marginal_log_lik}) evaluated at $\theta$, i.e.
-$$J(\theta; m, g) = - \nabla^2\mathcal{L}(\theta; m, g) .$$ This quantity, unfortunately, is hard to compute, as the log likelihood (\ref{marginal_log_lik}) is a complicated mixture.
-Louis \cite{Louis1982} showed that $J(\theta; m, g)$ is equivalent to the following quantity:
-\begin{multline}\label{zero_inf_info_mat}
-J(\theta; m, g) = -\E \left[\nabla^2 \mathcal{L}(\theta; m, g, p) | G = g, M = m \right] \\ + \E\left[\nabla \mathcal{L}(\theta; m, g, p) | G = g, M = m \right] \E\left[\nabla \mathcal{L}(\theta; m, g, p) | G = g, M = m \right]^T \\ - \E\left[ \nabla\mathcal{L}(\theta; m, g, p) \nabla \mathcal{L}(\theta; m, g, p)^T | G = g, M = m \right].
-\end{multline}
-The observed information matrix $J(\theta; m, g)$ has dimension $(2d+1) \times (2d + 1).$ Recall that the log-likelihood (\ref{full_log_lik}) is the sum of three terms.
-The first term depends only on $\pi$, the second on $\beta_m$, and the third on $\beta_g$.
-Therefore, the observed information matrix can be viewed as block matrix consisting of nine submatrices (Figure \ref{infomatrixbackground}; only six submatrices labelled).
-Submatrix I depends on $\pi$, submatrix II on $\beta_m$, submatrix III on $\beta_g$, submatrix IV on $\beta_m$ and $\beta_g$, submatrix V on $\pi$ and $\beta_m$, and submatrix VI on $\pi$ and $\beta_g$.
-We only need to compute these six submatrices to compute the entire matrix, as the matrix is symmetric.
-The following sections derive formulas for submatrices I-VI.
-All expectations are understood to be \textit{conditional} on $m$ and $g$.
-The notation $\nabla_v$  and $\nabla^2_v$  represent the gradient and Hessian, respectively, with respect to the vector $v$.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.45\linewidth]{../../figures/info_matrix/info_matrix_background}
-	\caption{Block structure of the observed information matrix $J(\theta; m, g) = -\nabla^2 \mathcal{L}(\theta; m, g)$.
-The matrix is symmetric, and so we only need to compute submatrices I-VI to compute the entire matrix.}
-	\label{infomatrixbackground}
-\end{figure}
-
-\subsubsection*{Submatrix I}
-Denote submatrix I by $J_{\pi}(\theta; m, g).$ The formula for $J_{\pi}(\theta; m, g)$ is 
-\begin{multline}\label{sub_mat_pi}
-J_{\pi}(\theta; m, g) = -\E\left[\nabla^2_\pi \mathcal{L}(\theta; m, g, p) \right] \\ + \left(\E\left[ \nabla_\pi \mathcal{L}(\theta; m, g, p) \right] \right)^2 - \E\left[(\nabla_\pi \mathcal{L}(\theta; m, g, p))^2 \right].
-\end{multline}
-
-We begin by calculating the first and second derivatives of the log-likelihood $\mathcal{L}$ with respect to $\pi$. The first derivative is
-\begin{multline}\label{d_L_d_pi}
-\nabla_\pi \mathcal{L}(\theta; m, g, p) = \frac{\partial }{\partial \pi } \left( \sum_{i=1}^n p_i \log(\pi) + \sum_{i=1}^n (1 - p_i) \log(1 - \pi) \right) \\ = \frac{ \sum_{i=1}^n p_i }{\pi} - \frac{ \sum_{i=1}^n (1 - p_i) }{ 1 - \pi } = \frac{\sum_{i=1}^n p_i}{\pi} - \frac{n - \sum_{i=1}^n p_i}{1 - \pi} \\ = \left( \frac{1}{\pi} + \frac{1}{1 - \pi} \right) \sum_{i=1}^n p_i - \frac{n}{1-\pi}.
-\end{multline}
-The second derivative is
-\begin{multline*}
-\nabla^2_\pi \mathcal{L}(\theta; m, g, p)  = \frac{\partial^2}{\partial^2\pi} \left( \frac{ \sum_{i=1}^n p_i }{ \pi } - \frac{ n - \sum_{i=1}^n p_i }{1 - \pi}  \right) = \frac{\left( \sum_{i=1}^n p_i \right) - n}{(1 - \pi)^2} - \frac{\sum_{i=1}^n p_i }{ \pi^2 }.
-\end{multline*}
-We compute the expectation of the first term of (\ref{sub_mat_pi}):
-\begin{multline}\label{submat_pi_1}
-\E \left[ -\nabla^2_{\pi} \mathcal{L}(\theta; m, g, p)\right] = - \E\left[\frac{ ( \sum_{i=1}^n p_i) - n}{(1 - \pi)^2} - \frac{\sum_{i=1}^n p_i}{\pi^2} \right] \\ = - \E\left\{\left[\frac{1}{(1-\pi)^2} - \frac{1}{\pi^2} \right] \sum_{i=1}^n p_i - \frac{n}{ (1 - \pi)^2 } \right\} \\ = - \left\{\left[ \frac{1}{(1-\pi)^2} - \frac{1}{\pi^2} \right] \sum_{i=1}^n T^\theta_i(1) - \frac{n}{ (1 - \pi)^2}  \right\} \\ = \left[ \frac{1}{\pi^2} - \frac{1}{(1 - \pi)^2} \right] \sum_{i=1}^n T^\theta_i(1) + \frac{n}{(1-\pi)^2}.
-\end{multline}
-Next, we compute the difference of the second two pieces of (\ref{sub_mat_pi}). To this end, define $$a := \frac{1}{(1-\pi)} + \frac{1}{\pi}$$ and $$b := \frac{n}{(1-\pi)}.$$ We have that
-\begin{multline*}
-\E \left[\nabla_\pi \mathcal{L}(\theta; m, g, p)^2 \right] \\ = \E \left[ \left( a \sum_{i=1}^n p_i - b\right)^2 \right]  =  \E \left[ a^2 \left( \sum_{i=1}^n p_i \right)^2 - 2ab \sum_{i=1}^n p_i + b^2 \right] \\ = a^2 \sum_{i=1}^n \sum_{j=1}^n \E[p_i p_j] -2ab \sum_{i=1}^n \E [p_i] + b^2.
-\end{multline*}
-Next,
-\begin{multline*}
-\left( \E \left[\nabla_\pi \mathcal{L}(\theta; m, g, x) \right] \right)^2 = \left( a \sum_{i=1}^n \E [p_i] - b \right)^2 \\ = a^2 \sum_{i=1}^n \sum_{j=1}^n \E[p_i]  E[p_j] - 2ab \sum_{i=1}^n \E[p_i] + b^2.
-\end{multline*}
-Therefore,
-\begin{multline}\label{submat_pi_2}
-(\E [\nabla_\pi \mathcal{L}(\theta; m, g, p)])^2 - \E \left[\nabla_\pi \mathcal{L}(\theta; m, g, p)^2 \right] \\ = a^2 \sum_{i=1}^n \sum_{j=1}^n \E[p_i] \E[p_j] - a^2 \sum_{i=1}^n \sum_{j=1}^n \E[p_i p_j] = a^2 \left( \sum_{i=1}^n \E[p_i]^2 - \E[p_i^2]\right) \\ = a^2 \left( \sum_{i=1}^n [T^\theta_i(1)]^2 - T^\theta_i(1) \right) = \left( \frac{1}{(1 - \pi)} + \frac{1}{\pi} \right)^2 \left(\sum_{i=1}^n [T^\theta_i(1)]^2 - T^\theta_i(1) \right).
-\end{multline}
-Stringing (\ref{sub_mat_pi}), (\ref{submat_pi_1}) and (\ref{submat_pi_2}) together, we obtain
-\begin{multline}\label{sub_mat_1_formula}
-J_\pi(\theta; m, g) = 
-\left[ \frac{1}{\pi^2} - \frac{1}{(1 - \pi )^2} \right] \sum_{i=1}^n T^\theta_i(1) + \frac{n}{(1-\pi )^2} \\ + \left( \frac{1}{(1 - \pi )} + \frac{1}{\pi} \right)^2 \left( \sum_{i=1}^n [T^\theta_i(1)]^2 - T^\theta_i(1) \right).
-\end{multline}
-
-\subsubsection*{Submatrix II}
-Denote submatrix II by $J_{\beta^m}(\theta; m, g).$ The formula for $J_{\beta^m}(\theta; m, g)$ is
-\begin{multline}\label{sub_mat_2}
-J_{\beta^m}(\theta; m, g) = -\E \left[\nabla_{\beta^m}^2 \mathcal{L}(\theta; m, g, p) \right] \\ + \E\left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] \E\left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T \\ - \E\left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T  \right].
-\end{multline}
-Standard GLM results imply that
-$$ -\nabla_{\beta^m}^2 \mathcal{L}(\theta; m, g, p) = \tilde{X}^T ( \Delta^m V^m \Delta^m - [\Delta']^m H^m ) \tilde{X}$$ and $$\nabla_{\beta^m}\mathcal{L}(\theta; m, g, p) = \tilde{X}^T \Delta^m s^m.$$
-We compute the first term of (\ref{sub_mat_2}). The $(k,l)$th entry of this matrix is
-\begin{multline*}
-\left( \E\left[-\nabla_{\beta^m}^2 \mathcal{L}(\theta; m, g, p)\right]\right)[k,l] = \E \left\{\tilde{X}[,k]^T (\Delta^m V^m \Delta^m - [\Delta']^mH^m) \tilde{X}[,l] \right\} \\ = \sum_{i=1}^n \E \left\{ \tilde{x}_{i,k} (\Delta^m_{i} V^m_{i} \Delta^m_{i} - [\Delta']^m_{i} H^m_{i}) \tilde{x}_{i,l} \right\} \\ = \sum_{i=1}^n \tilde{x}_{i,k}(0) T_i^{\theta}(0) [{\Delta}^m_i(0)  {V}^m_i(0) {\Delta}^m_i(0) - [\Delta']^m_i(0) {H}^m_i(0)] \tilde{x}_{i,l}(0) \\ + \sum_{i=1}^n \tilde{x}_{i,k}(1) T_i^{\theta}(1) [ {\Delta}^m_i(1)  {V}^m_i(1) {\Delta}^m_i(1) - [{\Delta}']^m_i(1) {H}^m_i(1)] \tilde{x}_{i,l}(1) \\ = \sum_{s = 0}^1 \tilde{X}(s)[,k]^T {T}^{\theta}(s) \left[ {\Delta}^m(s) {V}^m(s) {\Delta}^m(s) - [{\Delta}']^m(s) {H}^m(s) \right] \tilde{X}(s)[,l].
-\end{multline*}
-We therefore have that
-\begin{multline}\label{sub_mat_2_1}
-\E\left[-\nabla_{\beta^m}^2 \mathcal{L}(\theta; m, g, p)\right] \\ = \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) \left[ {\Delta}^m(s) {V}^m(s) {\Delta}^m(s) - [{\Delta}']^m(s) {H}^m(s) \right] \tilde{X}(s).
-\end{multline}
-Next, we compute the difference of the last two terms of (\ref{sub_mat_2}). The $(k,l)$th entry is
-\begin{multline*}
-\bigg[ \E \left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] \E \left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T \\ - \E \left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T \right] \bigg] [k,l] \\
-= \left[ \E \left[\tilde{X}^T \Delta^m s^m \right] \E \left[\tilde{X}^T \Delta^m s^m \right]^T\right][k,l] - \E \left[\tilde{X}^T \Delta^m s^m (s^m)^T \Delta^m \tilde{X} \right][k,l] \\
-= \E\left[ \tilde{X}[,k]^T \Delta^m s^m \right] \E \left[ \tilde{X}[,l]^T \Delta^m s^m \right] - \E \left[ \tilde{X}[,k]^T \Delta^m s^m (s^m)^T \Delta^m \tilde{X}[,l ] \right] \\ 
-=\E\left(\sum_{i=1}^n \tilde{x}_{ik} \Delta^m_i s^m_{i} \right) \E \left( \sum_{j=1}^n \tilde{x}_{jl} \Delta^m_j s^m_j \right) - \E \left( \sum_{i=1}^n \sum_{j=1}^n \tilde{x}_{ik} \Delta^m_i s^m_i s^m_j \Delta^m_j \tilde{x}_{jl} \right) \\
-= \sum_{i=1}^n \sum_{j=1}^n \E[ \tilde{x}_{ik} \Delta^m_is^m_i] \E [\tilde{x}_{jl} \Delta^m_j s^m_j]  -  \sum_{i=1}^n \sum_{j=1}^n \E [ \tilde{x}_{ik} \Delta^m_i s^m_i s^m_j \Delta^m_j \tilde{x}_{jl}] \\
-= \sum_{i=1}^n \sum_{j=1}^n \E[ \tilde{x}_{ik} \Delta^m_i s^m_i] \E \left[\tilde{x}_{jl} \Delta^m_j s^m_j \right]  - \sum_{i \neq j} \E [ \tilde{x}_{ik} \Delta^m_i s^m_i] \E[s^m_j \Delta^m_j \tilde{x}_{jl}] \\ - \sum_{i=1}^n \E[\tilde{x}_{ik} \Delta^m_i s^m_i s^m_i \Delta^m_i \tilde{x}_{il}] \\
-= \sum_{i=1}^n \E[\tilde{x}_{ik} \Delta^m_i s^m_i ] \E [ \tilde{x}_{il} \Delta^m_i s^m_i] - \sum_{i=1}^n \E[\tilde{x}_{ik} (\Delta_i^m)^2 (H_i^m)^2 \tilde{x}_{il}] \\ = \sum_{i=1}^n \left[\tilde{x}_{ik}(0) {\Delta}^m_i(0) T_i^{\theta}(0) {H}^m_i(0) + \tilde{x}_{ik}(1) {\Delta}^m_i(1) T_i^{\theta}(1) {H}^m_i(1) \right] \\ \cdot \left[ \tilde{x}_{il}(0) {\Delta}^m_i(0) T_i^{\theta}(0) {H}^m_i(0) + \tilde{x}_{il}(1) {\Delta}^m_i(1) T_i^{\theta}(1) {H}^m_i(1) \right] \\ - \sum_{i=1}^n \left[ \tilde{x}_{ik}(0) T_i^{\theta}(0) (\Delta_i^m(0))^2 ({H}_i^m(0))^2 \tilde{x}_{il}(0)  + \tilde{x}_{ik}(1) T_i^{\theta}(1) ({\Delta}^m_i(1))^2 ({H}_i^m(1))^m \tilde{x}_{il}(1) \right] \\ = \sum_{s=0}^1 \sum_{t=0}^1 \left[ \sum_{i=1}^n \tilde{x}_{ik}(s) T^{\theta}_i(s) {\Delta}^m_i(s) {H}^m_i(t) T_i^{\theta}(t){\Delta}^m_i(t) {H}^m_i(t) \tilde{x}_{il}(t) \right] \\ - \sum_{s=0}^1 \left[\sum_{i=1}^n \tilde{x}_{ik}(s) T_i^\theta(s) ({\Delta}^m_i(s))^2 ({H}_i^m(s))^2 \tilde{x}_{il}(s) \right] \\ = \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)[,k]^T {T}^\theta(s) {\Delta}^m(s) {H}^m(s) {T}^\theta(t) {\Delta}^m(t) {H}^m(t) \tilde{X}(k)[,l] \\ - \sum_{s=0}^1{X}(s)[,k]^T {T}^\theta(s) ({\Delta}^m(s))^2 ({H}^m(s))^2 \tilde{X}(s)[,l].
-\end{multline*}
-The sum of the last two terms on the right-hand side of (\ref{sub_mat_2}) is therefore
-\begin{multline}\label{sub_mat_2_2}
- \E \left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] \E \left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T \\ - \E \left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T \right] \\ =
-\sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T {T}^\theta(s) {\Delta}^m(s) {H}^m(s) T^{\theta}(t) {\Delta}^m(t) {H}^m(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T {T}^\theta(s) ({\Delta}^m(s))^2 ({H}^m(s))^2 \tilde{X}(s).
-\end{multline}
-Combining (\ref{sub_mat_2}), (\ref{sub_mat_2_1}), (\ref{sub_mat_2_2}), we find that
-\begin{multline}\label{sub_mat_2_formula}
-J_{\beta^m}(\theta; m, g) = \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) \left[ {\Delta}^m(s) {V}^m(s) {\Delta}^m(s) - [{\Delta}']^m(s) {H}^m(s) \right] \tilde{X}(s) \\ + \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T {T}^\theta(s) {\Delta}^m(s) {H}^m(s) {T}^\theta(t) {\Delta}^m(t) {H}^m(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) ({\Delta}^m(s))^2 ({H}^m(s))^2 \tilde{X}(s).
-\end{multline}
-
-\subsubsection*{Submatrix III}
-Denote submatrix III by $J_{\beta^g}(\theta; m, g).$ The formula for sub-matrix III is similar to that of sub-matrix II (\ref{sub_mat_2_formula}). Substituting $g$ for $m$ in this equation yields
-\begin{multline}\label{sub_mat_3_formula}
-J_{\beta^g}(\theta; m, g) = \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) \left[{\Delta}^g(s) {V}^g(s) {\Delta}^g(s) - [{\Delta}']^g(s) {H}^g(s) \right] \tilde{X}(s) \\ + \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T {T}^\theta(s) {\Delta}^g(s) {H}^g(s) {T}^\theta(t) {\Delta}^g(t) {H}^g(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) ({\Delta}^g(s))^2 ({H}^g(s))^2 \tilde{X}(s).
-\end{multline}
-
-\subsubsection*{Submatrix IV}
-Denote sub-matrix IV by $J_{(\beta^g, \beta^m)}(\theta; m, g)$. The formula for $J_{(\beta^g, \beta^m)}(\theta; m, g)$ is 
-\begin{multline}\label{sub_mat_4}
-J_{(\beta^g, \beta^m)}(\theta; m,g) = \E \left[-\nabla_{\beta^g} \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] \\ + \E\left[ \nabla_{\beta^{g}}\mathcal{L}(\theta ; m,g,p) \right] \E \left[\nabla_{\beta^m}\mathcal{L} (\theta ; m,g,p)  \right]^T \\ - \E \left[ \nabla_{\beta^{g}}\mathcal{L} (\theta; m,g,p) \nabla_{\beta^m}\mathcal{L}(\theta; m,g,p)^T  \right].
-\end{multline}
-First, we have that
-\begin{equation}\label{sub_mat_4_1}
-\E\left[-\nabla_{\beta^g} \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] = 0,
-\end{equation}
-as differentiating $\mathcal{L}$ with with respect to $\beta^g$ yields a vector that is a function of $\beta^g$, and differentiating this vector with respect to $\beta^m$ yields $0$.
-Next, recall from GLM theory that $$\nabla_{\beta^g} \mathcal{L}(\theta; m, g, p) =   \tilde{X}^T\Delta^g s^g$$ and $$\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) = \tilde{X}^T \Delta^m s^m.$$ The $(k,l)$th entry of the last two terms of (\ref{sub_mat_4}) is
-\begin{multline}\label{sub_mat_4_2}
-\bigg[ \E \left[\nabla_{\beta^g} \mathcal{L}(\theta; m, g, p) \right] \E \left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T \\ - \E \left[ \nabla_{\beta^g} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T \right] \bigg][k,l] \\
-= \left[ \E \left[ \tilde{X}^T \Delta^g s^g \right]\E \left[ \tilde{X}^T \Delta^m s^m \right]^T\right][k,l] - \E \left[ \tilde{X}^T \Delta^g s^g (s^m)^T \Delta^m \tilde{X} \right][k,l] \\ 
-= \E\left[\tilde{X}[,k]^T \Delta^g s^g \right] \E \left[\tilde{X}[,l]^T \Delta^m s^m \right] - \E \left[\tilde{X}[,k]^T \Delta^g s^g (s^m)^T \Delta^m \tilde{X}[,l ] \right] \\
-=\E\left( \sum_{i=1}^n \tilde{x}_{ik} \Delta^g_i s^g_i \right) \E \left( \sum_{j=1}^n \tilde{x}_{jl} \Delta^m_j s^m_j \right) - \E \left( \sum_{i=1}^n \sum_{j=1}^n \tilde{x}_{ik} \Delta^g_i s^g_i s^m_j \Delta^m_j \tilde{x}_{jl} \right) \\
-= \sum_{i=1}^n \sum_{j=1}^n \E[\tilde{x}_{ik} \Delta^g_is^g_i] \E[ \tilde{x}_{jl} \Delta^m_j s^m_j ] - \sum_{i=1}^n \sum_{j=1}^n \E[ \tilde{x}_{ik} \Delta^g_i s^g_i s^m_j \Delta^m_j \tilde{x}_{jl}]  \\
-= \sum_{i=1}^n \sum_{j=1}^n \E[ \tilde{x}_{ik} \Delta^g_i s^g_i] \E \left[\tilde{x}_{jl} \Delta^m_j s^m_j \right]  - \sum_{i \neq j} \E[\tilde{x}_{ik} \Delta^g_i s^g_i] \E[\tilde{x}_{jl}\Delta^m_j  s^m_j ] \\ - \sum_{i=1}^n \E[\tilde{x}_{ik} \Delta^g_i s^g_i s^m_i \Delta^m_i \tilde{x}_{il}] \\
-= \sum_{i=1}^n \E[\tilde{x}_{ik} \Delta^g_i H^g_i] \E[\tilde{x}_{il} \Delta_i^m H^m_i] - \sum_{i=1}^n \E[\tilde{x}_{ik} H_i^g \Delta_i^g \Delta_i^m H_i^m \tilde{x}_{il}] \\ 
-= \sum_{i=1}^n \left[\tilde{x}_{ik}(0) {\Delta}^g_i(0) T^\theta_i(0) {H}^g_i(0) + \tilde{x}_{ik}(1) {\Delta}^g_i(1) T^\theta_i(1) {H}^g_i(1)\right] \\ 
-\cdot \left[\tilde{x}_{il}(0) {\Delta}^m_i(0) T^\theta_i(0) {H}^m_i(0) + \tilde{x}_{il}(1) {\Delta}^m_i(1) T^\theta_i(1) {H}^m_i(1)\right] 
-\\ - \sum_{i=1}^n [\tilde{x}_{ik}(0) T^\theta_i(0) {\Delta}^g_i(0) {H}^g_i(0) {\Delta}^m_i(0) {H}^m_i(0) \tilde{x}_{il}(0) \\ + \tilde{x}_{ik}(1) T^\theta_i(1) {\Delta}^g_i(1) {H}^g_i(1) {\Delta}^m_i(1) {H}^m_i(1) \tilde{x}_{il}(1) ]
-\\ = \sum_{s=0}^1 \sum_{t=0}^1 \left[\sum_{i=1}^n \tilde{x}_{ik}(s) T^\theta_i(s) {\Delta}^g_i(s) {H}^g_i(s) T^\theta_i(t){\Delta}^m_i(t) {H}^m_i(t) \tilde{x}_{il}(t) \right]
-\\ - \sum_{s=0}^1 \left[\sum_{i=1}^n \tilde{x}_{ik}(s) T^\theta_i(s) {\Delta}^g_i(s) {H}^g_i(s) {\Delta}^m_i(s) {H}^m_i(s) \tilde{x}_{il}(s)\right] 
-\\ = \sum_{s=0}^1 \sum_{t=0}^1 \left[ \tilde{X}(s)[,k]^T T^\theta(s) {\Delta}^g(s) {H}^g(s) T^\theta(t){\Delta}^m(t) {H}^m(t) \tilde{X}(t)[,l] \right]\\ - \sum_{s=0}^1 \left[ \tilde{X}[,k]^T T^\theta(s) {\Delta}^g(s) {H}^g(s) {\Delta}^m(s) {H}^m(s) \tilde{X}[,l](s)\right].
-\end{multline}
-Combining (\ref{sub_mat_4}), (\ref{sub_mat_4_1}), and (\ref{sub_mat_4_2}) produces
-\begin{multline}\label{sub_mat_4_formula}
-J_{(\beta^g, \beta^m)}(\theta; m, g) = \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T  T^\theta(s) {\Delta}^g(s) {H}^g(s) T^\theta(t){\Delta}^m(t) {H}^m(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T T^\theta(s) {\Delta}^g(s) {H}^g(s) {\Delta}^m(s) {H}^m(s) \tilde{X}(s).
-\end{multline}
-\\ \noindent
-\subsubsection*{Submatrix V}
-Denote submatrix V by $J_{(\beta^m,\pi)}(\theta; m, g).$ The formula for $J_{(\beta^m,\pi)}(\theta; m, g)$ is
-\begin{multline}\label{sub_mat_5}
-J_{(\beta^m,\pi)}(\theta; m, g) = \E \left[ - \nabla_{\beta^m} \nabla_{ \pi } \mathcal{L}(\theta; m, g, p) \right] \\ + \E\left[ \nabla_{\beta^m}\mathcal{L}(\theta ; m,g,p) \right] \E \left[ \nabla_{\pi}\mathcal{L}(\theta ; m,g,p) \right]^T  - \E \left[ \nabla_{\beta^m}\mathcal{L}(\theta; m,g,p) \nabla_{\pi}\mathcal{L}(\theta; m,g,p)^T \right].
-\end{multline}
-We have that
-\begin{equation}\label{sub_mat_5_1}
-\E \left[ - \nabla_{\beta^m} \nabla_{ \pi } \mathcal{L}(\theta; m, g, p) \right] = 0,
-\end{equation}
-as $\beta^m$ and $\pi$ separate in the log likelihood. Next, set $a := 1/\pi + 1/(1 - \pi)$ and $b := n/(1 - \pi).$ Recall from GLM theory that
-$$\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) = \tilde{X}^T \Delta^m s^m$$ and from (\ref{d_L_d_pi}) that
-$$ a \sum_{i=1}^n p_i - b.$$
-The $k$th entry of the last two terms of (\ref{sub_mat_5}) is
-\begin{multline}\label{sub_mat_5_2}
-\E \left[\nabla_\pi \mathcal{L}(\theta; m, g, p) \right] \E\left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)[k] \right] - \E \left[\nabla_{\pi}\mathcal{L}(\theta; m,g,p) \nabla_{\beta^m}\mathcal{L}(\theta; m,g,p)[k] \right] \\= \left(\E \left[ a \sum_{i=1}^n p_i - b \right] \right) \left(\E\left[ \tilde{X}[,k]^T \Delta^m s^m \right] \right) - \E \left[ \left( a \sum_{i=1}^n p_i - b \right) \tilde{X}[,k]^T \Delta^m s^m \right] \\ = \left( a \sum_{i=1}^n \E[p_i] - b \right) \left( \sum_{j=1}^n \E [ \tilde{x}_{jk}\Delta^m_js^m_j] \right) - \E \left[ \left( a \sum_{i=1}^n p_i - b \right) \left( \sum_{j=1}^n \tilde{x}_{jk} \Delta^m_j s^m_j \right) \right] \\ = a \sum_{i=1}^n \sum_{j=1}^n \E [p_i] \E[ \tilde{x}_{jk} \Delta^m_j s^m_j] - b \sum_{j=1}^n \E[\tilde{x}_{jk} \Delta^m_j s^m_j] \\ - \left[ a \sum_{i=1}^n \sum_{j=1}^n \E [ p_i \tilde{x}_{jk} \Delta^m_j s^m_j] - b \sum_{j=1}^n \E[\tilde{x}_{jk} \Delta^m_j s^m_j] \right] \\ =  a \sum_{i=1}^n \sum_{j=1}^n \E[p_i] \E[\tilde{x}_{jk} \Delta^m_j s^m_j] - a\sum_{i \neq j} \E[p_i] \E[\tilde{x}_{jk} \Delta^m_j s^m_j] - a\sum_{i=1}^n \E[ p_i \tilde{x}_{ik} \Delta^m_i s^m_i ] \\ = a \sum_{i=1}^n \E[p_i] \E[ \tilde{x}_{ik} \Delta^m_i s^m_i] - a \sum_{i=1}^n \E[p_i \tilde{x}_{ik} \Delta^m_i s^m_i] \\ = a \sum_{i=1}^n T^\theta_i(1) [T^\theta_i(0) \Delta^m_i(0) s^m_i(0) \tilde{x}_{ik}(0) + T^\theta_i(1) \Delta^m_i(1) s^m_i(1) \tilde{x}_{ik}(1)] \\ - a \sum_{i=1}^n T^\theta_i(1)\Delta^m_i(1)s^m_i(1)\tilde{x}_{ik}(1) \\ = a \sum_{i=1}^n T^\theta_i(0)T^\theta_i(1) \Delta_i^m(0)H^m_i(0)\tilde{x}_{ik}(0) \\ + a \sum_{i=1}^n \left( [T^\theta_i(1)]^2 \Delta^m_i(1)H^m_i(1) - T^\theta_i(1)\Delta^m_i(1)H^m_i(1) \right) \tilde{x}_{ik}(1)  \\ =a \left[ \sum_{i=1}^n T^\theta_i(0) T^\theta_i(1) \Delta^m_i(0) H^m_i(0) \tilde{x}_{ik}(0) + \sum_{i=1}^n T^\theta_i(1)\Delta^m_i(1)H^m_i(1)[T^\theta_i(1) - 1] \tilde{x}_{ik}(1) \right] \\ = a \left[ \sum_{i=1}^n T^\theta_i(0) T^\theta_i(1) \Delta^m_i(0) H^m_i(0) \tilde{x}_{ik}(0) - \sum_{i=1}^n T^\theta_i(0) T^\theta_i(1) \Delta^m_i(1) H^m_i(1) \tilde{x}_{ik}(1) \right] \\ = a\left(\tilde{X}(0)[,k]^T w^m(0) - \tilde{X}(1)[,k]^T w^m(1)  \right).
-\end{multline}
-Combining (\ref{sub_mat_5}), (\ref{sub_mat_5_1}), and (\ref{sub_mat_5_2}), we conclude that
-\begin{equation}\label{sub_mat_5_formula} J_{(\beta^m, \pi)}(\theta; m, g, p) = \left( \frac{1}{\pi} + \frac{1}{1 - \pi} \right) \left( \tilde{X}(0)^T w^m(0) - \tilde{X}(1)^T w^m(1)\right).
-\end{equation}
-
-\subsubsection*{Submatrix VI}
-Denote submatrix VI by $J_{(\beta^g,\pi)}(\theta; m, g).$ Calculations similar to those for submatrix V show that
-\begin{equation}\label{sub_mat_6_formula} J_{(\beta^g, \pi)}(\theta; m, g, p) = \left(\frac{1}{\pi} + \frac{1}{1 - \pi} \right) \left( \tilde{X}(0)^T w^g(0) - \tilde{X}(1)^T w^g(1)\right).
-\end{equation}
-
-\subsubsection*{Combining submatrices}
-To summarize, the formulas for submatrices I-VI are as follows:
-\begin{itemize}
-\item[I]\begin{multline*}
-J_\pi(\theta; m, g) = 
-\left[ \frac{1}{\pi^2} - \frac{1}{(1 - \pi )^2} \right] \sum_{i=1}^n T^\theta_i(1) + \frac{n}{(1-\pi )^2} \\ + \left( \frac{1}{(1 - \pi )} + \frac{1}{\pi} \right)^2 \left( \sum_{i=1}^n [T^\theta_i(1)]^2 - T^\theta_i(1) \right).
-\end{multline*}
-\item[II] \begin{multline*}
-J_{\beta^m}(\theta; m, g) = \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) \left[ {\Delta}^m(s) {V}^m(s) {\Delta}^m(s) - [{\Delta}']^m(s) {H}^m(s) \right] \tilde{X}(s) \\ + \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T {T}^\theta(s) {\Delta}^m(s) {H}^m(s) {T}^\theta(t) {\Delta}^m(t) {H}^m(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) ({\Delta}^m(s))^2 ({H}^m(s))^2 \tilde{X}(s).
-\end{multline*}
-\item[III] \begin{multline*}
-J_{\beta^g}(\theta; m, g) = \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) \left[{\Delta}^g(s) {V}^g(s) {\Delta}^g(s) - [{\Delta}']^g(s) {H}^g(s) \right] \tilde{X}(s) \\ + \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T {T}^\theta(s) {\Delta}^g(s) {H}^g(s) {T}^\theta(t) {\Delta}^g(t) {H}^g(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) ({\Delta}^g(s))^2 ({H}^g(s))^2 \tilde{X}(s).
-\end{multline*}
-\item[IV] \begin{multline*}
-J_{(\beta^g, \beta^m)}(\theta; m, g) = \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T  T^\theta(s) {\Delta}^g(s) {H}^g(s) T^\theta(t){\Delta}^m(t) {H}^m(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T T^\theta(s) {\Delta}^g(s) {H}^g(s) {\Delta}^m(s) {H}^m(s) \tilde{X}(s).
-\end{multline*}
-\item[V] $$ J_{(\beta^m, \pi)}(\theta; m, g, p) = \left( \frac{1}{\pi} + \frac{1}{1 - \pi} \right) \left( \tilde{X}(0)^T w^m(0) - \tilde{X}(1)^T w^m(1)\right).  $$
-\item[VI] $$ J_{(\beta^g, \pi)}(\theta; m, g, p) = \left(\frac{1}{\pi} + \frac{1}{1 - \pi} \right) \left( \tilde{X}(0)^T w^g(0) - \tilde{X}(1)^T w^g(1)\right).$$
-\end{itemize}
-We stitch these pieces together and transpose submatrices IV, V, and VI to produce the whole information matrix $J(\theta; m, g)$.
-Evaluating this matrix at the EM estimate $\theta^\textrm{EM}$ and inverting yields the asymptotic covariance matrix, which we can use to compute standard errors.
-
-\subsection{Implementation}
-To evaluate the observed information matrix, we need to compute the matrices $\Delta^m(j),$ $[\Delta']^m(j),$ $V^m(j),$ and $H^m(j)$ and the vectors $s^m(j)$ and $w^m(j)$ for $j \in \{0,1\}$.
-We likewise need to compute the analogous gRNA quantities.
-We describe how to compute these quantities in R by extending base family objects.
-We implicitly condition on $p_i$, $z^m_i$, and $o^m_i$.
-
-\begin{algorithm}
-	\caption{Computing the matrices $\Delta^m(j)$, $[\Delta']^m(j)$, $V^m(j)$, $H^m(j)$, and $s^m(j)$ given given $\beta_m$.}\label{algo:computing_info_matrices}
-	\begin{algorithmic}[3]
-		\Require A coefficient vector $\beta_m$; data $[m_1, \dots, m_n]$, $[o^m_1, \dots, o^m_n]$, and $[z_1, \dots, z_n]$; and a family object containing functions \texttt{linkinv}, \texttt{variance}, \texttt{mu.eta}, \texttt{mu.eta.prime}, and \texttt{skewness}.
-		\For{$j \in \{0, 1\}$}
-		\For{$i \in \{1, \dots, n\}$}
-		\State $l^m_i(j) \gets \langle \beta_m, \tilde{x}_i(j) \rangle + o^m_i$
-		\State $\mu^m_i(j) \gets \texttt{linkinv}(l^m_i(j))$
-		\State $[\sigma_i^m(j)]^2 \gets \texttt{variance}(\mu_i^m(j))$
-		\State $h_m'(l_i^m(j)) \gets \texttt{mu.eta}(l_i^m(j))/[\sigma_i^m(j)]^2$
-		\State $\gamma^m_i(j) \gets \texttt{skewness}(\mu^m_i(j))$
-		\State $[r_m^{-1}]''(l_i^m(j)) \gets \texttt{mu.eta.prime}(l^m_i(j))$
-		\State $$h_m''(l_i^m(j)) \gets \frac{[r^{-1}]''(l_i^m(j)) - [([\sigma_i^m(j)]^2)^{3/2}][\gamma^m_i(j)] [h_m'(l_i^m(j))]^2}{[\sigma_i^m(j)]^2}$$
-		\Comment Assign quantities to matrices
-		\State $\Delta_{i}^m(j) \gets h_m'( l_i^m(j))$
-		\State $[\Delta']^m_{i}(j) \gets h''(l^m_i(j))$
-		\State $V^m_{i}(j) \gets [\sigma^m_i(j)]^2$
-		\State $H^m_{i}(j) \gets s^m_i(j) \gets m_i - \mu^m_i(j)$
-		\EndFor
-		\EndFor
-	\end{algorithmic}
-\end{algorithm}
-
-An \texttt{R} family object contains several functions, including \texttt{linkinv}, \texttt{variance}, and \texttt{mu.eta}.
-\texttt{linkinv} is the inverse link function $r_m^{-1}$.
-\texttt{variance} takes as an argument the mean $\mu^m_i$ and returns the variance $[\sigma_i^m]^2$.
-\texttt{mu.eta} is the derivative of the inverse link function $[r^{-1}_m]^{'}$.
-We extend the \texttt{R} family object by adding two additional functions: \texttt{skewness} and \texttt{mu.eta.prime}.
-\texttt{skewness} returns the skewness $\gamma^m_i$ of the distribution as a function of the mean $\mu_i$, i.e.
-$$\texttt{skewness}\left(\mu_i\right) = \E \left[\left(\frac{m_i - \mu_i^m}{ \sigma_i^m}\right)^3\right] := \gamma_i^m.$$ Finally, the \texttt{mu.eta.prime} is the second derivative of the inverse link function $[r^{-1}_m]''.$ Algorithm \ref{algo:computing_info_matrices} computes the matrices $\Delta^m(j)$, $[\Delta']^m(j)$, $V^m(j)$, $H^m(j)$, and vector $s^m(j)$ for given $\beta_m$ and given family object.
-(The vector $w^m(j)$ can be computed in terms of $\Delta^m(j)$ and $H^m(j)$.) We use $\sigma^m_i(j)$ (resp.
-$\gamma^m_i(j)$) to refer to the standard deviation (resp.
-skewness) of the gene expression distribution the $i$th cell when the perturbation $p_i$ is set to $j$.
-
-All steps of the algorithm are obvious except the calculation of $h'_m(l^m_i(j))$ (line 6), $h''(l^m_i(j))$ (line 9), and $V^m_i(j)$ (line 12).
-We omit the $(j)$ notation for compactness.
-First, we prove the correctness of the expression for $h'_m(l^m_i)$.
-Recall the basic GLM identities
-\begin{equation}\label{computing_info_matrix_1}
-\psi_m''(\eta_i^m) =  [\sigma^m_i]^2
-\end{equation}
- and, for all $t \in \R$, 
-\begin{equation}\label{computing_info_matrix_2}
- r_m^{-1}(t) = \psi_m'(h_m(t)).
-\end{equation}
- Differentiating (\ref{computing_info_matrix_2}) in $t$, we find that
-\begin{equation}\label{computing_info_matrix_3}
- (r_m^{-1})'(t) = \psi_m''(h_m(t))h_m'(t), 
-\end{equation}
-or  $$h_m'(t) = \frac{(r_m^{-1})'(t) }{\psi_m''(h_m(t))}.$$ Finally, plugging in $l^m_i$ for $t$,
-$$ h_m'(l_i) = \frac{(r_m^{-1})'(l^m_i)}{\psi''(h_m(l^m_i))} = \frac{(r_m^{-1})'(l^m_i)}{\psi_m''(\eta^m_i)} = \textrm{ by (\ref{computing_info_matrix_1}) } \frac{(r_m^{-1})'(l_i^m)}{[\sigma_i^m]^2}.$$
-
-Next, we prove the correctness for the expression for $h_m''(l_i^m)$. Recall the exponential family identity 
-\begin{equation}\label{computing_info_matrix_4}
-\psi'''_m(\eta^m_i) = \gamma^m_i  ([\sigma^m_i]^2)^{3/2}.
-\end{equation}
-Differentiating (\ref{computing_info_matrix_3}) in $t$, we obtain
-$$ (r_m^{-1})''(t) = \psi_m'''(h_m(t)) [h_m'(t)]^2 + \psi_m''(h_m(t)) h_m''(t),$$ or $$h_m''(t) =\frac{(r_m^{-1})''(t) - \psi'''(h_m(t))[h_m'(t)]^2}{\psi_m''(h_m(t))}.$$ Plugging in $l^m_i$ for $t$, we find that
-\begin{multline*}
-h_m''(l^m_i) = \frac{(r_m^{-1})''(l^m_i) - \psi_m'''(\eta^m_i) [h_m'(l_i^m)]^2}{[\sigma_i^m]^2} \\ = \textrm{ (by \ref{computing_info_matrix_4}) }  \frac{(r_m^{-1})''(l^m_i) - ([\sigma_i^m]^2)^{3/2} (\gamma_i^m) [h_m'(l^m_i)]^2 }{[\sigma_i^m]^2}.
-\end{multline*}
-Finally, the expression for $V^m_i$ follows from (\ref{computing_info_matrix_1}).
-We can apply a similar algorithm to compute the analogous matrices for the gRNA modality.
-Table \ref{family_object_functions} shows the \texttt{linkinv}, \texttt{variance}, \texttt{mu.eta}, \texttt{skewness}, and \texttt{mu.eta.prime} functions for several common family objects (which are defined by a distribution and link function).
-\begin{table}
-\centering
-\caption{\texttt{linkinv}, \texttt{variance}, \texttt{mu.eta}, \texttt{skewness}, \texttt{mu.eta.prime} for common family objects (i.e., pairs of distributions and link functions).}\label{family_object_functions}
-\begin{tabular}{|C{2.5cm}|C{3.5cm}|C{3.4cm}|C{2.5cm}|}
-	\hline 
-	& Gaussian response, identity link & Poisson response, log link & NB response ($\theta > 0$ fixed), log link \\ 
-	\hline 
-	\texttt{linkinv} & $x$ & $\exp(x)$ & $\exp(x)$  \\ 
-	\hline 
-	\texttt{variance} & $x$ & $x$ & $x + x^2/\theta$ \\ 
-	\hline 
-	\texttt{mu.eta} & $1$ & $x$  & $\exp(x)$ \\ 
-	\hline 
-	\texttt{skewness} & $0$ & $x^{-1/2}$ & $\frac{2 x + \theta}{\sqrt{\theta x} \sqrt{x + \theta}}$ \\ 
-	\hline 
-	\texttt{mu.eta.prime} & $0$ & $\exp(x)$ & $\exp(x)$ \\ 
-	\hline 
-\end{tabular}
-\end{table}
-
-\section{Zero-inflated model}\label{sec:zero_inf_model}
-In this section we introduce the ``zero-inflated'' GLM-EIV model.
-The zero-inflated GLM-EIV model is appropriate to use when the unperturbed cells do not transcribe \textit{any} gRNA molecules (i.e., when there are no background reads).
-Let $x_i = [1, z_i]^T \in \R^{d-1}$ be the vector of observed covariates, including an intercept term.
-($x_i$ is the same as $\tilde{x}_i$, but with the perturbation indicator $p_i$ removed.) Let $\beta_{g,z} = [\beta^g_0, \gamma_g] \in \R^{d-1}$ be an unknown coefficient vector.
-($\beta_{g,z}$ is the same as $\beta_g$, but with the perturbation effect $\beta^g_1$ removed).
-Let the linear component $l^{g,z}_i$, mean $\mu^{g,z}_i$, and canonical parameter $\eta^{g,z}_i$ of gRNA count distribution of the $i$th cell be given by
-$$\begin{cases}
-l^{g,z}_i = \langle x_i, \beta_{g,z} \rangle + o^g_i \\
-r_g(\mu^{g,z}_i) = l^{g,z}_i \\
-\eta^{g,z}_i = ([\psi'_g]^{-1} \circ r^{-1}_g)(l^{g,z}_i) := h_g(l^{g,z}_i).
-\end{cases}
-$$
-The density $f_{g,z}$ of gRNA counts in the zero-inflated model is as follows:
-$$f_{g,z}(g_i; \eta^{g,z}_i, p_i) = [f_g(g_i;\eta^{g,z}_i)]^{p_i} \mathbb{I}(g_i = 0)^{1-p_i}.$$
-In other words, when the cell is \textit{perturbed} (i.e., $p_i = 1$), the zero-inflated density $f_{g,z}$ coincides with the background-read density $f_g$; by contrast, when the cell is \textit{unperturbed} (i.e., $p_i = 0$), the zero-inflated density $f_{g,z}$ is a point mass at zero.
-The gene expression density $f_m$ and perturbation indicator density $f_p$ are the same across the background read and zero-inflated models.
-We assume that the gene expression $m_i$ and gRNA count $g_i$ are conditionally independent given the perturbation indicator $p_i$.
-The joint density $f_z$ of $(m_i, p_i, z_i)$ is
-\begin{multline*}
-f_z(m_i, g_i, p_i) = f_m(m_i | p_i) f_{g,z}(g_i | p_i) f_p(p_i) \\ = \pi^{p_i} (1-\pi)^{1-p_i} f_m(m_i; \eta^m_i)[f_g(g_i; \eta_i^{g,z})]^{p_i} \mathbb{I}(g_i = 0)^{1-p_i}.
-\end{multline*}
-The log-likelihood $\mathcal{L}_z$ is
-\begin{multline*}
-\mathcal{L}_z(\theta; m, g, p) = \sum_{i=1}^n \log \left[ \pi^{p_i} (1-\pi)^{1-p_i} \right] + \sum_{i=1}^n \log\left[f_m(m_i; \eta^m_i) \right] \\ + \sum_{i=1}^n p_i \log \left[f_{g}(g_i; \eta_i^{g,z}) \right] + \sum_{i=1}^n (1 - p_i)\log\left[ \mathbb{I}(g_i = 0) \right],
-\end{multline*}
-where $\theta = [\pi, \beta_m, \beta_{g,z}]$ is the vector of unknown parameters.
-Integrating over the unobserved variable $p_i$, the marginal density $f_z$ of $(m_i, g_i)$ is
-$$f_z(m_i, g_i; \theta) = (1-\pi) f_m(m_i;\eta^m_i(0)) \mathbb{I}(g_i = 0) + \pi f_m(m_i; \eta^m_i(1)) f_g(g_i;\eta^{g,z}_i).$$ Finally, the marginal log-likelihood is
-$$
-\mathcal{L}_z(\theta; m_i, g_i) = \sum_{i=1}^n \log\left[ (1-\pi) f_m(m_i;\eta^m_i(0)) \mathbb{I}(g_i = 0) + \pi f_m(m_i; \eta^m_i(1)) f_g(g_i;\eta^{g,z}_i) \right].
-$$
-
-\subsection{Estimation}
-To estimate the parameters of the zero-inflated GLM-EIV model, we use an EM algorithm similar to Algorithm \ref{algo:em_full} but with two changes.
-First, we use a different formula for the $i$th membership probability at the $t$-th step of the algorithm $T^{(t)}_i(1)$.
-(We use $T^{(t)}_i(1)$ to denote the $i$th membership probability in \textit{both} the background read and zero inflated cases; the difference should be clear from context.) Let $\theta^{(t)} = (\pi^{(t)}, \beta^{(t)}_m, \beta^{(t)}_{g,z})$ be the parameter estimate at the $t$-th iteration of the algorithm.
-Arguing in a manner similar to the background read case, we have that $$T^{(t)}_i(1) = \frac{1}{ \exp(q_i^{\left(t,z\right)}) + 1},$$ where
-$$q_i^{(t,z)} = \log \left(\frac{(1 - \pi^{(t)}) \P(M_i = m_i | P_i = 0, \theta^{(t)}) \P(G_i = g_i | P_i = 0, \theta^{(t)})}{(\pi^{(t)}) \P(M_i = m_i | P_i = 1, \theta^{(t)}) \P(G_i = g_i | P_i = 1, \theta^{(t)})}\right).$$
-The expression for $q^{(t,z)}_i$ is
-\begin{multline*}
-q^{(t,z)}_i = \log\left[ 1 - \pi^{(t)} \right] + \log\left[ f_m\left(m_i; \left[ \eta^m_i(0) \right]^{(t)}\right) \right] + \log\left[ \mathbb{I}(g_i = 0) \right] \\ - \log\left[ \pi^{(t)} \right] - \log\left[ f_m\left(m_i; \left[ \eta^m_i(1) \right]^{(t)}\right) \right] - \log\left[ f_g\left(g_i; \left[\eta^{g,z}_i \right]^{(t)}\right) \right],
-\end{multline*}
-where $[\eta^{g,z}_i]^{(t)} = h_g( \langle x_i, \beta^{(t)}_{g,z} \rangle + o^g_i).$ Notice that if $g_i \geq 1$, then $T^{(t)}_i(1) = 1.$ This comports with our intuition that a nonzero gRNA count indicates the presence of a perturbation.
-
-Next, we consider the M step of the EM algorithm, which is similar to the background read case but slightly different.
-Define $Q_z(\theta | \theta^{(t)}) = \E_{(P | M = m, G = g, \theta^{(t)})} \left[ \mathcal{L}_z(\theta; m, g, p) \right].$ We have that
-\begin{multline}\label{q_funct_zero_inf}
-Q_z(\theta | \theta^{(t)}) = \sum_{i=1}^n \left[ T^{(t)}_i(1)\log(\pi) + T_i^{(t)}(0) \log(1 - \pi)\right] + \sum_{i=1}^n \sum_{j=0}^1 T^{(t)}_i(j) \log \left[ f_m( m_i; \eta_i^m(j)) \right] \\ + \sum_{i=1}^n T_i^{(t)}(1) \left[ \log( f_g(g_i; \eta_i^{g, z}))\right] + C.
-\end{multline}
-The three terms of (\ref{q_funct_zero_inf}) are functions of $\pi$, $\beta_m$, and $\beta_{g,z}$, respectively.
-The maximizer $\pi^{(t)}$ and $\beta_m^{(t+1)}$ of the first and second term are the same as in the background read case.
-The maximizer $\beta^{(t+1)}_{g,z}$ of the third term is the maximizer of the GLM with exponential family density $f_g$, link function $r_g$, responses $g$, weights $T^{(t)}(1)$, design matrix $X$, offsets $o^g$.
-
-\subsection{Inference}
-Next, we derive the asymptotic observed information matrix for the zero-inflated model, allowing us to perform inference.
-Again, let $T^{\theta}(1) := \textrm{diag}\{T^{\theta}_1(1), \dots, T^{\theta}_n(1)\},$ but note that $T^\theta_i(1) = \P(P_i = 1 | G_i = g_i, M_i = m_i, \theta)$ is computed differently than in the background read case.
-Define the $n \times n$ matrices $\Delta^{(g,z)}, [\Delta']^{(g,z)}, V^{(g,z)},$ and $H^{(g,z)}$ by
-$$
-\begin{cases}
-\Delta^{(g,z)} = \textrm{diag} \{ h_g'(l_1^{g,z} ), \dots, h_g'( l_n^{g,z}  ) \} \\
-[\Delta']^{(g,z)}  = \textrm{diag} \{ h_g''(l_1^{g,z} ), \dots, h_g''( l_n^{g,z} ) \} \\
-V^{(g,z)}  = \textrm{diag} \{ \psi_g( \eta^{g,z} _1), \dots, \psi_g( \eta^{g,z}_n) \} \\
-H^{(g,z)}  = \textrm{diag} \{ m_1 - \mu_1^{g,z} , \dots, m_n - \mu_n^{g,z} \} .
-\end{cases}
-$$
-Also, define the $\R^n$ vectors $s^{(g,z)}$ and $w^{(g,z)}$ by
-$$s^{(g,z)} = [g_1 - \mu_1^{g,z}, \dots, g_n - \mu_n^{g,z}]^T,$$ and $$w^{(g,z)} = [T^\theta_1(0)T^\theta_1(1) \Delta_1^{(g,z)}H^{(g,z)}_1,\dots, T^\theta_n(0)T^\theta_n(1) \Delta_n^{(g,z)}H^{(g,z)}_n].$$ These quantities are computable, as they do not depend on the unobserved variables $p_1, \dots, p_n$.
-Finally, let the unobserved, $n\times n$ matrix $P$ be defined by $P = \textrm{diag}\{ p_1, \dots, p_n \}$.
-
-The observed information matrix $J_z(\theta; m, g)$ is given by
-$$ J_z(\theta; m, g) = -\nabla^2 \mathcal{L}_z(\theta; m, g).$$ Louis's theorem implies that
-\begin{multline*}
-J_z(\theta; m, g) = -\E \left[\nabla^2 \mathcal{L}_z(\theta; m, g, p) | G = g, M = m \right] \\ + \E\left[\nabla \mathcal{L}_z(\theta; m, g, p) | G = g, M = m \right] \E\left[\nabla \mathcal{L}_z(\theta; m, g, p) | G = g, M = m \right]^T \\ - \E\left[\nabla\mathcal{L}_z(\theta; m, g, p) \nabla \mathcal{L}_z(\theta; m, g, p)^T | G = g, M = m \right].
-\end{multline*}
-The matrix $J_z(\theta; m, g)$ has dimension $d \times d$ and consists of nine submatrices (Figure \ref{infomatrixzeroinf}).
-Three of these submatrices (i.e., I, II, and V) are the same as the corresponding submatrices in the background read case.
-We therefore must compute the remaining submatrices (i.e., III, IV, and VI) to compute the entire matrix $J_z(\theta; m, g)$.
-Again, in the following, all expectations are understood to be conditional on $m$ and $g$.
-
-\begin{figure}
-	\centering
-	\includegraphics[width=0.45\linewidth]{../../figures/info_matrix/info_matrix_zero_inf}
-	\caption{Block structure of the observed information matrix $J_z(\theta; m, g) = -\nabla^2 \mathcal{L}_z(\theta; m, g)$ for the zero-inflated model.
-Submatrices I, II, and VI are the same as in the background read model; therefore, we only need to compute submatrices III, VI, and V.}
-	\label{infomatrixzeroinf}
-\end{figure}
-
-\subsubsection*{Submatrix III (zero-inflated)}
-Denote submatrix III by $J_{\beta_{(g,z)}}(\theta; m, g)$ The formula for $J_{\beta_{(g,z)}}(\theta; m, g)$ is
-\begin{multline}\label{sub_mat_3_zeroinf}
-J_{\beta_{(g,z)}}(\theta; m, g) = -\E \left[\nabla_{\beta_{(g,z)}}^2 \mathcal{L}_z(\theta; m, g, p) \right] \\ + \E\left[ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right] \E\left[\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right]^T \\ - \E\left[ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p)^T \right].
-\end{multline}
-GLM theory indicates that
-$$ -\nabla_{\beta_{(g,z)}}^2 \mathcal{L}_z(\theta; m, g, p) =  X^T P ( \Delta^{(g,z)}V^{(g,z)}\Delta^{(g,z)} - (\Delta')^{(g,z)} H^{(g,z)}) X$$ and
-$$ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) = X^T P \Delta^{(g,z)} s^{(g,z)}.$$
-We begin by computing the first term of (\ref{sub_mat_3_zeroinf}).
-The only random matrix among $X,$ $P$, $\Delta^{(g,z)}$, $V^{(g,z)}$, $(\Delta')^{(g,z)},$ and $H^{(g,z)}$ is $P$.
-Therefore, by the linearity of expectation,
-\begin{multline}\label{sub_mat_3_zeroinf_1}
--\E \left[\nabla_{\beta_{(g,z)}}^2 \mathcal{L}_z(\theta; m, g, p) \right]   = \E \left[ X^T P (\Delta^{(g,z)}V^{(g,z)}\Delta^{(g,z)} - (\Delta')^{(g,z)} H^{(g,z)} \right] \\ = X^T T^{\theta}(1) ( \Delta^{(g,z)}V^{(g,z)}\Delta^{(g,z)} - (\Delta')^{(g,z)} H^{(g,z)}) X.
-\end{multline}
-Next, we compute the difference of the last two terms of (\ref{sub_mat_3_zeroinf}). The $(k,l)$th entry of this matrix is
-\begin{multline*}
-\bigg[ \E \left[\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right] \E \left[\nabla_{\beta{(g,z)}} \mathcal{L}_z(\theta; m, g, p)\right]^T \\ - \E \left[ \nabla_{{\beta_{(g,z)}}} \mathcal{L}_z(\theta; m, g, p) \nabla_{{\beta_{(g,z)}}} \mathcal{L}_z(\theta; m, g, p)^T \right] \bigg] [k,l] \\
-= \left[ \E \left[ {X}^T P \Delta^{(g,z)} s^{(g,z)} \right]\E \left[ {X}^T P \Delta^{(g,z)} s^{(g,z)} \right]^T\right][k,l] - \E \left[ {X}^T P \Delta s^{(g,z)} (s^{(g,z)})^T \Delta^{(g,z)} P {X}^T \right][k,l] \\
-= \E\left[{X}[,k]^T P \Delta^{(g,z)} s^{(g,z)} \right] \E \left[ {X}[,l]^T P \Delta^{(g,z)} s^{(g,z)} \right] - \E \left[ {X}[,k]^T P \Delta^{(g,z)} s^{(g,z)} (s^{(g,z)})^T \Delta^{(g,z)} P {X}[,l ] \right] \\
-=\E\left(\sum_{i=1}^n {x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_{i} \right) \E \left( \sum_{j=1}^n {x}_{jl} P_j \Delta^{(g,z)}_j s^{(g,z)}_j \right) \\ - \E \left( \sum_{i=1}^n \sum_{j=1}^n {x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i s^{(g,z)}_j \Delta^{(g,z)}_j P_j{x}_{jl} \right) \\
-= \sum_{i=1}^n \sum_{j=1}^n \E[ {x}_{ik} P_i \Delta^{(g,z)}_is^{(g,z)}_i] \E [ {x}_{jl} P_j \Delta^{(g,z)}_j s^{(g,z)}_j]  -  \sum_{i=1}^n \sum_{j=1}^n \E [ {x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i s^{(g,z)}_j \Delta^{(g,z)}_jP_j {x}_{jl}  ]  \\
-= \sum_{i=1}^n \sum_{j=1}^n \E[ {x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i ] \E \left[{x}_{jl} P_j \Delta^{(g,z)}_j s^{(g,z)}_j \right]  - \sum_{i \neq j} \E [{x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i] \E[s^{(g,z)}_j P_j \Delta^{(g,z)}_j {x}_{jl}] \\ - \sum_{i=1}^n \E[ {x}_{ik}P_i \Delta^{(g,z)}_i s^{(g,z)}_i s^{(g,z)}_i \Delta^{(g,z)}_i P_i{x}_{il}] \\
-= \sum_{i=1}^n \E[ {x}_{ik} P_i \Delta^{(g,z)}_i H^{(g,z)}_i] \E[{x}_{il}P_i\Delta^{(g,z)}_i H^{(g,z)}_i] - \sum_{i=1}^n \E[{x}_{ik}P_i^2(\Delta^{(g,z)}_i)^2 (H^{(g,z)}_i)^2 {x}_{il}] \\ = \sum_{i=1}^n {x}_{ik} T_i^{\theta}(1)^2(\Delta^{(g,z)}_i)^2 (H_i^{(g,z)})^2{x}_{il} - \sum_{i=1}^n {x}_{ik} T_i^{\theta}(1) (\Delta_i^{(g,z)})^2 (H_i^{(g,z)})^2 {x}_{il} \\ = {X}[,k]^T T^{\theta}(1)^2 (\Delta^{(g,z)})^2 (H^{(g,z)})^2 {X}[,l] - {X}[,k]^T T^{\theta}(1) (\Delta^{(g,z)})^2 (H^{(g,z)})^2 {X}[,l]
-\end{multline*}
-Therefore, we have that
-\begin{multline}\label{sub_mat_3_zeroinf_2}
-\E\left[ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right] \E\left[\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right]^T - \E\left[ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p)^T \right] \\ =
-X^T T^{\theta}(1)^2 (\Delta^{(g,z)})^2(H^{(g,z)})^2{X} - {X}^T T^{\theta}(1) (\Delta^{(g,z)})^2 (H^{(g,z)})^2 {X} \\ = - X^T T^{\theta}(1) \left(\Delta^{(g,z)}\right)^2 \left( H^{(g,z)} \right)^2 \left( I - T^{\theta}(1)  \right) X.
-\end{multline}
-Combining (\ref{sub_mat_3_zeroinf}), (\ref{sub_mat_3_zeroinf_1}), and (\ref{sub_mat_3_zeroinf_2}), we conclude that
-\begin{multline}\label{sub_mat_3_zeroinf_formula}
-J_{\beta_{(g, z)}} = \left( \theta; m, g \right) =  X^T T^{\theta}(1) ( \Delta^{(g,z)}V^{(g,z)}\Delta^{(g,z)} - (\Delta')^{(g,z)} H^{(g,z)}) X \\ - X^T T^{\theta}(1) \left(\Delta^{(g,z)}\right)^2 \left( H^{(g,z)} \right)^2 \left( I - T^{\theta}(1) \right) X.
-\end{multline}
-
-\subsubsection*{Submatrix IV (zero-inflated)}
-Denote submatrix IV by $J_{(\beta_{(g,z)}, \beta_m)}(\theta; m, g)$. The formula for submatrix IV is
-\begin{multline}\label{sub_mat_4_zeroinf}
-J_{(\beta_{(g,z)}, \beta_m)}(\theta; m, g) = - \E \left[\nabla_{\beta_{(g,z)}} \nabla_{\beta_m} \mathcal{L}_z (\theta; m, g, p) \right] \\ + \E\left[ \nabla_{\beta_{(g,z)} }\mathcal{L}_z(\theta ; m,g,p) \right]\E\left[\nabla_{\beta_m}\mathcal{L}_z(\theta; m,g,p)\right]^T \\ - \E \left[\nabla_{\beta_{(g,z)}}\mathcal{L}_z (\theta; m,g,p) \nabla_{\beta_m}\mathcal{L}_z(\theta; m,g,p)]^T \right].
-\end{multline} 
-First, we have that
-\begin{equation}\label{sub_mat_4_zeroinf_1}
-- \E \left[\nabla_{\beta_{(g,z)}} \nabla_{\beta_m} \mathcal{L}_z (\theta; m, g, p) \right] = 0,
-\end{equation}
- as the derivative in $\beta_m$ of $\mathcal{L}_z(\theta; m, g, p)$ is a function of $\beta_m$, and the derivative in $\beta_{(g,z)}$ of this term is $0$.
-Next, we compute the difference of the last two terms of (\ref{sub_mat_4_zeroinf}).
-Entry $(k,l)$ of this matrix is
-\begin{multline}\label{sub_mat_4_zeroinf_2}
-[\E[\nabla_{\beta_{(g,z)}}\mathcal{L}_z(\theta; m, g, p) ] \E [\nabla_{\beta_m} \mathcal{L}_z(\theta; m, g, p) ]^T \\ - \E [\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \nabla_{\beta_m}\mathcal{L}_z(\theta; m, g, p)^T ] ][k,l] \\
-= \left[ \E \left[ {X}^T P \Delta^{(g,z)} s^{(g,z)} \right] \E\left[\tilde{X}^T \Delta^m s^m\right]^T\right][k,l] - \E \left[{X}^T P \Delta^{(g,z)} s^{(g,z)} (s^m)^T \Delta^m \tilde{X} \right][k,l] \\ = \left[\E\left[{X}[,k]^T P \Delta^{(g,z)} s^{(g,z)}\right]\E\left[\tilde{X}[,l]^T \Delta^m s^m\right]^T\right] - \E \left[{X}[,k]^T P \Delta^{(g,z)} s^{(g,z)} (s^m)^T \Delta^m \tilde{X}[,l] \right] \\ = \E\left(\sum_{i=1}^n x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i\right) \E \left(\sum_{j=1}^n \tilde{x}_{jl} \Delta^m_j s^m_j \right) - \E \left(\sum_{i=1}^n \sum_{j=1}^n x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i \Delta^m_j s^m_j \tilde{x}_{jl}  \right) \\
-= \sum_{i=1}^n \sum_{j=1}^n \E [x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i] \E[\Delta^m_js^m_j \tilde{x}_{jl}] - \sum_{i=1}^n \sum_{j=1}^n \E[x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i \Delta^m_j s^m_j \tilde{x}_{jl}] \\
-= \sum_{i=1}^n \sum_{j=1}^n \E [x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i ] \E[\Delta^m_js^m_j \tilde{x}_{jl}] - \sum_{i \neq j} \E [x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i] \E[\Delta^m_js^m_j \tilde{x}_{jl}] \\ - \sum_{i=1}^n \E[x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i \Delta^m_j s^m_j \tilde{x}_{jl}] \\
-= \sum_{i=1}^n \E[x_{ik} P_i \Delta_i^{(g,z)} H^{(g,z)}_i] \E[\tilde{x}_{il} \Delta^m_i H^m_i] - \sum_{i=1}^n \E[x_{ik} P_i \Delta_i^{(g,z)} H_i^{(g,z)} \Delta_i^m H_i^m \tilde{x}_{il}] \\ 
-= \sum_{i=1}^n \left[x_{ik} T_i^\theta(1) \Delta^{(g,z)}_i H^{(g,z)}_i \right] \cdot \left[{\Delta}^m_i(0) T^\theta_i(0) {H}^m_i(0) \tilde{x}_{il}(0) + {\Delta}^m_i(1) T_i^\theta(1) {H}^m_i(1) \tilde{x}_{il}(1)\right]
-\\ - \sum_{i=1}^n \left[x_{ik} T_i^\theta(1) \Delta^{(g,z)}_i H^{(g,z)}_i \Delta^m_i(1) H^m_i(1) \tilde{x}_{il}(1)\right] 
-\\ = \sum_{s=0}^1 \sum_{i=1}^n x_{ik} T_i^\theta(1) H_i^{(g,z)} \Delta_i^{(g,z)} T^\theta_i(s) \Delta_i^m(s) H^m(s) \tilde{x}_{il}(s) \\ - \sum_{i=1}^n \left[x_{il}T_i^\theta(1) \Delta^{(g,z)}_i H^{(g,z)}_i \Delta^m_i(1) H^m_i(1) \tilde{x}_{ik}(1)\right] \\ = \sum_{s=0}^1 X[,k]^T T^\theta(1) H^{(g,z)} \Delta^{(g,z)} T^\theta(s)\Delta^m(s)H^m(s) \tilde{X}(s)[,l] \\ - X[,k]^T \Delta^{(g,z)} H^{(g,z)} T^\theta(1)\Delta^m(1) H^m(1) \tilde{X}[,l].
-\end{multline}
-Combining (\ref{sub_mat_3_zeroinf}), (\ref{sub_mat_3_zeroinf_1}), and (\ref{sub_mat_3_zeroinf_2}) yields
-\begin{multline}\label{sub_mat_4_zeroinf_formula} J_{(\beta_{(g,z)}, \beta_m)}(\theta; m, g) = \left(\sum_{s=0}^1 X^T  T^\theta(1) H^{(g,z)} \Delta^{(g,z)} T^\theta(s) \Delta^m(s) H^m(s) \tilde{X}(s) \right) \\ - X^T \Delta^{(g,z)} H^{(g,z)} T^\theta(1) \Delta^m(1)H^m(1) \tilde{X}(1).
-\end{multline}
-
-\subsubsection*{Submatrix VI (zero-inflated)}
-Denote submatrix VI by $J_{(\beta_{(g,z)},\pi)}(\theta; m, g).$ The formula for $J_{(\beta_{(g,z)},\pi)}(\theta; m, g)$ is
-\begin{multline}\label{sub_mat_6_zeroinf}
-J_{(\beta_{(g,z)},\pi)}(\theta; m, g) = \E \left[-\nabla_{\beta_{(g,z)}} \nabla_{\pi} \mathcal{L}_z(\theta; m, g, p) \right] \\ + \E\left[ \nabla_{\beta_{(g,z)}}\mathcal{L}_z(\theta; m, g ,p)\right] \E \left[ \nabla_{\pi}\mathcal{L}_z(\theta; m,g,p) \right] \\ - \E \left[ \nabla_{\beta_{(g,z)}}\mathcal{L}_z(\theta; m,g,p) \nabla_{\pi}\mathcal{L}_z(\theta; m,g,p) \right].
-\end{multline} Recall that $\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) = X^T P \Delta^{(g,z)}s^{(g,z)}$ and $\nabla_\pi \mathcal{L}_z(\theta; m, g, p) = a \left(\sum_{i=1}^n p_i\right) - b,$ where $$a = \frac{1}{\pi} + \frac{1}{1 - \pi}, \textrm{    } b = \frac{n}{1 - \pi}.$$
-We have that
-\begin{equation}\label{sub_mat_6_zeroinf_1}
-\E \left[-\nabla_{\beta_{(g,z)}} \nabla_{\pi} \mathcal{L}_z(\theta; m, g, p) \right] = 0,
-\end{equation}
-as the derivative in $\pi$ of $\mathcal{L}_z(\theta; m, g, p)$ is a function of $\pi$, and the derivative in $\beta_{(g,z)}$ of this term is $0$.
-Next, we compute the difference of the second two terms of (\ref{sub_mat_6_zeroinf}).
-The $k$th entry of this vector is
-\begin{multline}\label{sub_mat_6_zeroinf_2}
-\E \left[ \nabla_\pi \mathcal{L}_z(\theta; m, g, p) \right] \E\left[\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, x)[k] \right] - \E \left[\nabla_{\pi}\mathcal{L}(\theta; m,g,p) \nabla_{\beta_{(g,z)}} \mathcal{L} (\theta; m,g,p)[k] \right] \\= \left(\E \left[ a \sum_{i=1}^n p_i - b \right] \right)\left( \E \left[{X}[,k]^T P \Delta^{(g,z)} s^{(g,z)} \right]\right) - \E \left[ \left( a \sum_{i=1}^n p_i - b \right) {X}[,k]^T P \Delta^{(g,z)} s^{(g,z)}\right] \\ = \left( a \sum_{i=1}^n \E[p_i] - b \right) \left( \sum_{j=1}^n \E [x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] \right) \\ - \E \left[ \left( a \sum_{i=1}^n p_i - b \right) \left( \sum_{j=1}^n \tilde{x}_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j \right) \right] \\= a \sum_{i=1}^n \sum_{j=1}^n \E [p_i] \E[x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] - b \sum_{j=1}^n \E[x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] \\ - \left[a \sum_{i=1}^n \sum_{j=1}^n \E [ p_i x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] - b \sum_{j=1}^n \E[x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] \right] \\ = a \sum_{i=1}^n \sum_{j=1}^n \E[p_i] \E[{x}_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] - a \sum_{i \neq j} \E[p_i] \E[x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] - a \sum_{i=1}^n \E[ x_{ik} p_i^2 \Delta^{(g,z)}_i s^{(g,z)}_i] \\ = a \sum_{i=1}^n \E[p_i] \E[x_{ik} p_i \Delta^{(g,z)}_i s^{(g,z)}_i] - a \sum_{i=1}^n \E[x_{ik} p_i^2 \Delta^{(g,z)}_i s^{(g,z)}_i] \\ = a \sum_{i=1}^n T^\theta_i(1) x_{ik} T^\theta_i(1) \Delta^{(g,z)}_i s^{(g,z)}_i - a \sum_{i=1}^n x_{ik} T^\theta_i(1) \Delta^{(g,z)}_i s^{(g,z)}_i \\ = a \sum_{i=1}^n \left( {x}_{ik} T^\theta_i(1)^2 \Delta^{(g,z)}_i s^{(g,z)}_i - x_{ik} T^\theta_i(1) \Delta^{(g,z)}_i s^{(g,z)}_i \right) = a \sum_{i=1}^n {x}_{ik}T^\theta_i(1)\Delta^{(g,z)}_i s^{(g,z)}_i \left( T^\theta_i(1) - 1\right) \\ = -a\sum_{i=1}^n x_{ik} T_i(0) T^\theta_i(1)\Delta^{(g,z)}_i H^{(g,z)}_i = -a X[,k]^T w^{(g,z)}.
-\end{multline}
-Combining (\ref{sub_mat_6_zeroinf}), (\ref{sub_mat_6_zeroinf_1}), and (\ref{sub_mat_6_zeroinf_2}), we conclude that
-\begin{equation}\label{sub_mat_6_zeroinf_formula} J_{(\beta_{(g,z)},\pi)}(\theta; m, g) = -\left( \frac{1}{\pi} + \frac{1}{1-\pi} \right)X^T w^{(g,z)}.\end{equation}
-
-\section{Statistical accelerations}
-A key step in the algorithm for computing the pilot parameter estimates (Algorithm \ref{algo:pilot_estimates}) is to fit a weighted, no-intercept, univariate GLM with nonzero offset terms and a binary predictor variable (line 20).
+\subsection*{Intercept-plus-offset models}
+
+A key step in the algorithm for computing the pilot parameter estimates (Algorithm \ref{algo:pilot_estimates_2}) is to fit a weighted, no-intercept, univariate GLM with nonzero offset terms and a binary predictor variable.
 We derive an analytic formula for the MLE of this GLM for three important pairs of response distributions and link functions: Gaussian response with identity link, Poisson response with log link, and negative binomial response with log link.
@@
 
-\subsection*{Intercept-plus-offset models}
 Let $\beta \in \R$ be an unknown constant.
@@
 Let $Y_i|o_i, \dots, Y_n|o_i$ be exponential family-distributed random variables with identity sufficient statistic.
-Suppose the mean $\mu_i$ of $Y_i|o_i$ is given by $$r(\mu_i) = \beta + o_i,$$ where $r: \R \to \R$ is a strictly increasing, differentiable link function.
+Suppose the mean $\mu_i$ of $Y_i|o_i$ is given by $r(\mu_i) = \beta + o_i,$ where $r: \R \to \R$ is a strictly increasing, differentiable link function.
 We call this model the \textit{intercept-plus-offset} model.
@@
  and the density $f$ of $Y_i | \eta_i$ is
-$$f(y_i; \eta_i) = \exp\{y_i \eta_i - \psi(\eta_i) + c(y_i)\}.$$ The weighted log likelihood is
+$f(y_i; \eta_i) = \exp\{y_i \eta_i - \psi(\eta_i) + c(y_i)\}.$ The weighted log likelihood is
 \begin{equation}\label{m_plus_o_mle}
@@
 Therefore,
-\begin{multline}\label{nb_mo_1}
-\eta_i = h(\beta + o_i) = \log\left(\frac{\exp(\beta+o_i)}{\theta + \exp(\beta + o_i)} \right) = \beta + o_i - \log\left(\theta + e^{\beta}e^{o_i}\right) \\ = \beta - \log\left(\theta + e^{\beta}e^{o_i} \right) + C,
-\end{multline}
+\begin{equation}\label{nb_mo_1}
+\eta_i = h(\beta + o_i) = \log\left(\frac{\exp(\beta+o_i)}{\theta + \exp(\beta + o_i)} \right) = \beta + o_i - \log\left(\theta + e^{\beta}e^{o_i}\right) = \beta - \log\left(\theta + e^{\beta}e^{o_i} \right) + C,
+\end{equation}
  and
@@
 
-\section{Real data analysis details}
+\section{Data analysis details}\label{sec:data_analysis_details}
 
@@
 There were $17,028$ candidate \textit{cis} pairs, $97,818$ negative control pairs, and $322$ positive control pairs.
+
+As a point of reference, the Gasperini dataset, which consisted of $>$115,000 pairs, took 2,120 CPU hours to analyze, or equivalently 14 hours distributed across 150 processors.
+
+- construction of negative control pairs
+- cis pairs
+
 \printbibliography

### 2021-12-14 R_scripts/plotting/data_analysis.R

@@ -54,30 +54,33 @@ aggregate_df <- resampling_df %>% filter(contam_level >= 0, parameter == "m_pert
   summarize(m = mean(estimate), m_lower_ci = mean(confint_lower, na.rm = TRUE), m_upper_ci = mean(confint_upper, na.rm = TRUE)) %>% ungroup() %>% group_by(method, pair_id) %>%
   group_modify(.f = function(tbl, key) {
     baseline_est <- tbl$m[tbl$contam_level == 0]
-    p_change <- abs(tbl$m - baseline_est)/baseline_est
+    p_change <- (tbl$m - baseline_est)/baseline_est
     ci_cover <- (tbl$m_lower_ci < baseline_est) & (tbl$m_upper_ci > baseline_est)
     mutate(tbl, p_change = p_change, ci_cover = ci_cover)
   }) %>% ungroup() %>% group_by(contam_level, method) %>% summarize(median_p_change = median(p_change),
+                                                                    p_change_15 =  quantile(p_change, 0.25)[[1]],
+                                                                    p_change_85 =  quantile(p_change, 0.75)[[1]],
                                                                     lower_median_ci = sort(p_change)[qbinom(.025, length(p_change), 0.5)],
                                                                     upper_median_ci = sort(p_change)[qbinom(.975, length(p_change), 0.5)],
                                                                     mean_coverage = mean(ci_cover),
                                                                     lower_cover_ci = mean_coverage - 1.96 * sqrt((1/n_pairs) * mean_coverage * (1 - mean_coverage)),
                                                                     upper_cover_ci = mean_coverage + 1.96 * sqrt((1/n_pairs) * mean_coverage * (1 - mean_coverage))) %>%
-  mutate(method = factor(method, c("glmeiv", "thresholding"), c("GLM-EIV", "Thresholding")))
+  mutate(method = factor(method, c("glmeiv", "thresholding"), c("GLM-EIV", "Thresh.")))
 
 p5 <- ggplot(data = aggregate_df %>% filter(contam_level <= 0.4),
              mapping = aes(x = contam_level, y = median_p_change, col = method)) + 
[…]

### 2021-12-14 R_scripts/plotting/thresholding_theoretical.R

@@ -72,8 +72,9 @@ p_c <- ggplot(data = to_plot_c %>% dplyr::mutate(Metric = factor(Metric, levels 
                                                        panel.grid.minor = element_blank(), axis.line = element_line(colour = "black"))
 
 # combine
-p_bottom <- plot_grid(p_b, p_c, align = "h", ncol = 2, rel_widths = c(1, 2), labels = c("b", "c"))
-p <- plot_grid(p_a, p_bottom, ncol = 1, labels = c("a", ""))
+# p_bottom <- plot_grid(p_b, p_c, align = "h", ncol = 2, rel_widths = c(1, 2), labels = c("b", "c"))
+# p <- plot_grid(p_a, p_bottom, ncol = 1, labels = c("a", ""))
 
 # save
-ggsave(filename = paste0(fig_dir, "/plot.jpeg"), plot = p, device = "jpeg", scale = 1.05, width = 6, height = 4, dpi = 320)
+ggsave(filename = paste0(fig_dir, "/plot.jpeg"), plot = p_c, device = "jpeg", scale = 1.0, width = 4.5, height = 2.5, dpi = 320)
+ggsave(filename = paste0(fig_dir, "/app_plot.jpeg"), plot = p_a, device = "jpeg", scale = 1.5, width = 4.5, height = 1.5, dpi = 320)

### 2021-12-17 text/jsm/jsm_paper.tex
add proof + more citations

@@
 Our focus, however, is on regression rather than dimension reduction.
-We extend the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.
+We extend the classical errors-in-variables model in several directions, most importantly generalizing to exponential family-distributed response distributions and sources of measurement error.
 
@@
 For a given threshold $c \in \R$, the imputed perturbation assignment $\hat{p}_i$ is $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ is the OLS solution, i.e.
-$\hat{\beta}^m_1 = \left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})(m_i - \overline{m})\right]\left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})^2\right]^{-1}.$ We begin by deriving the limit of $\hat{\beta}^m_1$:
+$\hat{\beta}^m_1 = \left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})(m_i - \overline{m})\right]\left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})^2\right]^{-1}.$ We derive the almost sure limit of $\hat{\beta}^m_1$:
 \begin{proposition}\label{prop:convergence}
@@
 
-To answer this question, we introduce GLM-EIV (GLM-based errors-in-variables), a method that models the entire data-generating process.
+To answer this question, we introduce GLM-EIV (GLM-based errors-in-variables), a method that models the entire data-generating process in a single-cell CRISPR screen experiment.
 
@@
 Likewise, the target of inference $\beta^m_1$ is the log fold change in gene expression in response to the perturbation, accounting for technical factors.
-See Appendix \ref{sec:zero_inf_model} for a parallel methodology that we developed in which we model the gRNA distribution as (in a certain sense) zero-inflated counts.
+See Appendix \ref{sec:zero_inf_model} for a parallel methodology that we developed in which we model the gRNA distribution as (in a certain sense) zero-inflated counts that may be applicable in some settings.
 
@@
 where, $x^*_i \sim \textrm{Bern}(\pi), \ep_i, \tau_i \sim N(0,1),$ and $\ep_i$,$\tau_i$, and $x^*_i$ are independent.
-GLM-EIV extends (\ref{classical_eiv}) in at least three directions: first, GLM-EIV allows $y_i$ and $x_i$ to follow (possibly different) exponential family distributions; second, GLM-EIV allows $y_i$ and $x_i$ to be related to $x^*_i$ through (possibly different) nonlinear link functions; and finally, GLM-EIV allows confounders $z_i$ to impact both $x_i$ and $y_i$.
-In such settings $x_i$ and $y_i$ are dependent \textit{even after conditioning on $x^*_i$}, enabling GLM-EIV to capture more complex dependence relationships between $x_i$ and $y_i$ than is possible in (\ref{classical_eiv}) or other standard measurement error models.
+GLM-EIV extends (\ref{classical_eiv}) in at least three directions: first, GLM-EIV allows $y_i$ and $x_i$ to follow exponential family (i.e, not just Gaussian) distributions; second, GLM-EIV allows $y_i$ and $x_i$ to be related to $x^*_i$ through linear \textit{or} nonlinear link functions; and finally, GLM-EIV allows confounders $z_i$ to impact both $x_i$ and $y_i$.
+Therefore, $x_i$ and $y_i$ are dependent \textit{even after conditioning on $x^*_i$}, enabling GLM-EIV to capture more complex dependence relationships between $x_i$ and $y_i$ than is possible in (\ref{classical_eiv}) or other standard measurement error models.
 
@@
 Leveraging our computational infrastructure (Section \ref{sec:computing}), we applied GLM-EIV and the thresholding method to analyze the entire Gasperini and Xie datasets.
-(We report only the most important aspects of the analysis and results in the main text; full details are available in Appendix \ref{sec:data_analysis_details}.) We set the threshold to the approximate Bayes-optimal decision boundary, as our theoretical analyses and simulation studies revealed that the Bayes-optimal decision boundary is a good threshold when the gRNA count distribution is well-separated.
+(We report only the most important aspects of the analysis and results in the main text; full details are available in Appendix \ref{sec:data_analysis_details}.) We set the threshold to the approximate Bayes-optimal decision boundary, as our theoretical analyses and simulation studies revealed that the Bayes-optimal decision boundary is a good choice for the threshold when the gRNA count distribution is well-separated.
 Operating under the assumption that the effect of the perturbation on gRNA expression is similar across pairs, we leveraged the fitted GLM-EIV models to approximate the Bayes boundary in the following way: we (i) sampled several hundred gene-perturbation pairs, (ii) extracted the fitted values $\hat{\beta}_g$ and $\hat{\pi}$ from the GLM-EIV models fitted to these pairs, (iii) computed the median $\overline{\hat{\beta}_g}$ and $\overline{\hat{\pi}}$ across the $\hat{\beta}_g$s and $\hat{\pi}$s, and (iv) used $\overline{\hat{\beta}_g}$ and $\overline{\hat{\pi}}$ to estimate a dataset-wide Bayes-optimal decision boundary.
@@
 We generalized the above analysis to the entire set of positive control pairs.
-First, for each pair we computed the ``relative estimate change'' (REC) as a function of excess background contamination, defined as the magnitude of the relative difference between the estimate at a given level of excess contamination and zero excess contamination (Figure \ref{fig:real_data}d).
-Next, we computed the median REC across all positive control pairs (Figure \ref{fig:real_data}e; bands indicate pointwise 95\% CIs for the population median REC).
+First, for each pair we computed the ``relative estimate change'' (REC) as a function of excess background contamination, defined as the relative difference between the estimate at a given level of excess contamination and zero excess contamination (Figure \ref{fig:real_data}d).
+Next, we computed the median REC across all positive control pairs (Figure \ref{fig:real_data}e; upper and lower bands indicate the pointwise interquartile range of the REC).
 As excess background contamination increased, thresholded regression exhibited severe attenuation bias (as reflected by large median REC values); GLM-EIV, by contrast, remained mostly stable.
@@
 In this work we introduced GLM-EIV (``GLM-based errors in variables''), a new method for single-cell CRISPR screen analysis.
-GLM-EIV extends the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed and (optionally) impacted by the same set of confounding variables.
+GLM-EIV extends the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed and potentially impacted by the same set of confounding variables.
 These extensions enable GLM-EIV to resolve novel analysis challenges posed by single-cell CRISPR screens.
@@
 GLM-EIV intuitively achieves this performance gain by leveraging information from \textit{both} modalities (rather than the gRNA modality alone) to assign perturbation identities to cells.
-In low background contamination settings GLM-EIV and thresholded regression work best in conjunction, with GLM-EIV providing a statistically principled and empirically effective procedure for selecting the threshold.
+On the other hand, in low background contamination settings GLM-EIV and thresholded regression work best in conjunction, with GLM-EIV providing a statistically principled and empirically effective procedure for selecting the threshold.
+GLM-EIV thereby neutralizes a tuning parameter that, until this point, has been selected using heuristic procedures, with little confidence that the choice is near optimal.
 Figure \ref{fig:dec_tree} summarizes how we anticipate GLM-EIV being used in practice.
@@
 This work used the Extreme Science and Engineering Discovery Environment (XSEDE; NSF grant ACI-1548562) and the Bridges-2 system (NSF grant ACI-1928147) at the Pittsburgh Supercomputing Center.
-Code to replicate the analyses is available at \texttt{/timothy-barry/glmeiv-manuscript}.
+Code for the analyses is available at \texttt{/timothy-barry/glmeiv-manuscript}.
 
@@
 \textbf{e-f}, Median relative estimate change (REC; \textbf{e}) and confidence interval coverage rate (\textbf{f}) across \textit{all} 322 positive control pairs, plotted as a function of excess background contamination.
-Colored bands, pointwise 95\% CIs for population quantities.
 Panels (\textbf{c-f}) together illustrate that GLM-EIV demonstrated greater stability than thresholded regression as background contamination increased.}\label{fig:real_data}
@@
 
-We performed quality control.
-We kept cells with percent mitochondrial reads less than 8$\%$.
-For the Gasperini dataset, we furthermore kept cells with UMI and gRNA counts in the 5th to 95th percentiles to reduce the effect of outliers.
-(We did not perform this latter filter on the Xie data, as the latter dataset appeared to be less noisy and contained fewer cells.) Next, we kept genes expressed in at least $10\%$ of cells with a mean expression level of at least $1$.
-We did not QC the gRNAs.
-
-The quality-controlled Xie dataset contained $n = 101,508$ cells, $1,030$ genes, and $516$ gRNAs.
-There were $50,000$ \textit{in silico} negative control pairs and $681$ candidate \textit{cis} pairs.
-The quality-controlled Gasperini dataset contained $n = 170,645$ cells, $2,079$ genes, $6,598$ gRNAs.
-There were $17,028$ candidate \textit{cis} pairs, $97,818$ negative control pairs, and $322$ positive control pairs.
-
-As a point of reference, the Gasperini dataset, which consisted of $>$115,000 pairs, took 2,120 CPU hours to analyze, or equivalently 14 hours distributed across 150 processors.
-
-- construction of negative control pairs
-- cis pairs
+First, we performed quality control on the datasets.
+As is standard in single-cell analysis \cite{Choudhary2021a}, we removed cells with a high fraction ($>8\%$) of mitochondrial reads.
+We additionally excluded genes that were expressed in fewer than $10\%$ of cells or that had a mean expression level of less than $1$.
+We excluded cells in the Gasperini dataset with gene transcript UMI or gRNA counts below the 5th percentile or above the 95th percentile to reduce the effect of outliers.
+We did not repeat this latter quality control step on the Xie data, because the Xie data were less noisy.
+The quality-controlled Gasperini and Xie datasets contained $n = 170,645$ (resp.
+$n = 101,508$) cells, $2,079$ (resp.
+$1,030$) genes, and $6,598$ (resp.
+$516$) gRNAs.
+
+The Gasperini dataset included $17,028$ candidate \textit{cis} pairs, $97,818$ negative control pairs, and $322$ positive control pairs.
+The \textit{cis} pairs consisted of genes paired to nearby enhancers with unknown regulatory effects.
+The negative control pairs consisted of nontargeting gRNAs paired to genes.
+The Xie data contained $681$ candidate \textit{cis} pairs and, unfortunately, no positive control or negative control pairs.
+We therefore constructed a set of $50,000$ \textit{in silico} negative control pairs on the Xie data by pairing perturbations to genes on different chromosomes; see \textit{Methods} section of \cite{Barry2020} for details.
+
+We modeled the gene expression counts using a negative binomial distributions with unknown size parameter $\theta$; we estimated $\theta$ using the {glm.nb} package.
+Choudhary and Satija \cite{Choudhary2021a} report that Poisson models accurately capture highly sparse single-cell data.
+Although Choudhary and Satija did not investigate gRNA data specifically, we modeled the gRNA counts using Poisson distributions, as the gRNA modality exhibited greater sparsity than the gene modality.
+
+We applied GLM-EIV and the thresholding method to analyze the entire set of pairs in both datasets.
+We did not report results on the candidate \textit{cis} pairs in the text because we do not know the ground truth for these pairs, making them less useful for method assessment.
+We focused our attention instead on the negative control pairs in both datasets and the positive control pairs in the Gasperini dataset (Figures \ref{thresholding_empirical} and \ref{fig:real_data}).
+
+Finally, we describe in more detail how we conducted the ``excess background contamination'' analysis (Figure \ref{fig:real_data}, panels c-f).
+For each positive control pair, we varied excess background contamination over the grid $[0.0, 0.05, 0.1, \dots, 0.4].$ For a given level of excess background contamination, we generated $B = 50$ synthetic gRNA datasets, holding fixed the raw gene expressions, covariates, library sizes, and fitted perturbation probabilities.
+We fitted GLM-EIV and the thresholding method to the data, yielding estimates $[\hat{\beta}^m_1]^{(1)}, \dots, [\hat{\beta}^m_1]^{(B)}$.
+Next, we averaged over the $[\hat{\beta}^m_1]^{(i)}$s to obtain the mean estimate for a given pair and level of background contamination, and we calculated the REC using these mean estimates.
+Finally, we computed CI coverage in a somewhat nonstandard way: for a given pair and level of background contamination, we averaged over the upper and lower CIs of the $[\hat{\beta}^m_1]^{(i)}$s and checked for overlap with the raw estimate.
+In other words, we did not check whether the CI of each \textit{individual} $[\hat{\beta}^m_1]^{(i)}$ overlapped the ground truth; this is because we did not generate the data i.i.d., and so we cannot expect standard coverage guarantees to obtain.
+CI coverage (in the sense described here) for excess background contamination equal to $0$ by definitionis $1$, as depicted in Figure \ref{fig:real_data}f.
 

### 2021-12-28 text/jasa/JASA-template.tex
add journal submit folder

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-1in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.7in}%
\addtolength{\topmargin}{-1in}%

\begin{document}

%\bibliographystyle{natbib}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if1\blind
{
  \title{\bf Title}
  \author{Author 1\thanks{
    The authors gratefully acknowledge \textit{please remember to list all relevant funding sources in the unblinded version}}\hspace{.2cm}\\
    Department of YYY, University of XXX\\
    and \\
    Author 2 \\
    Department of ZZZ, University of WWW}
  \maketitle
} \fi

\if0\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf Title}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
The text of your abstract. 200 or fewer words.
\end{abstract}

\noindent%
{\it Keywords:}  3 to 6 keywords, that do not appear in the title
\vfill

\newpage
\spacingset{1.9} % DON'T change the spacing!
\section{Introduction}
\label{sec:intro}

Body of paper.  Margins in this document are roughly 0.75 inches all
around, letter size paper.

\begin{figure}
\begin{center}
\includegraphics[width=3in]{fig1.pdf}
\end{center}
\caption{Consistency comparison in fitting surrogate model in the tidal
power example. \label{fig:first}}
\end{figure}

\begin{table}
\caption{D-optimality values for design $X$ under five different scenarios.  \label{tab:tabone}}
\begin{center}
\begin{tabular}{rrrrr}
one & two & three & four & five\\\hline
1.23 & 3.45 & 5.00 & 1.21 & 3.41 \\
1.23 & 3.45 & 5.00 & 1.21 & 3.42 \\
1.23 & 3.45 & 5.00 & 1.21 & 3.43 \\
\end{tabular}
\end{center}
\end{table}

\begin{itemize}
\item Note that figures and tables (such as Figure~\ref{fig:first} and
Table~\ref{tab:tabone}) should appear in the paper, not at the end or
in separate files.
\item In the latex source, near the top of the file the command
\verb+\newcommand{\blind}{1}+ can be used to hide the authors and
acknowledgements, producing the required blinded version.
\item Remember that in the blind version, you should not identify authors
indirectly in the text.  That is, don't say ``In Smith et. al.  (2009) we
showed that ...''.  Instead, say ``Smith et. al. (2009) showed that ...''.
\item These points are only intended to remind you of some requirements.
Please refer to the instructions for authors
at \url{http://amstat.tandfonline.com/action/authorSubmission?journalCode=uasa20&page=instructions#.VFkk7fnF_0c}
\item For more about ASA\ style, please see \url{http://journals.taylorandfrancis.com/amstat/asa-style-guide/}
\item If you have supplementary material (e.g., software, data, technical
proofs), identify them in the section below.  In early stages of the
submission process, you may be unsure what to include as supplementary
material.  Don't worry---this is something that can be worked out at later stages.
\end{itemize}

\section{Methods}
\label{sec:meth}
Don't take any of these section titles seriously.  They're just for
illustration.

\section{Verifications}
\label{sec:verify}
This section will be just long enough to illustrate what a full page of
text looks like, for margins and spacing.

\addtolength{\textheight}{.5in}%

\cite{Campbell02}, \cite{Schubert13}, \cite{Chi81}

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
{\bf With this spacing we have 25 lines per page.}
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

%new margin nere

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

\addtolength{\textheight}{-.3in}%

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.
The quick brown fox jumped over the lazy dog.

\section{Conclusion}
\label{sec:conc}

\bigskip
\begin{center}
{\large\bf SUPPLEMENTARY MATERIAL}
\end{center}

\begin{description}

\item[Title:] Brief description. (file type)

\item[R-package for  MYNEW routine:] R-package MYNEW containing code to perform the diagnostic methods described in the article. The package also contains all datasets used as examples in the article. (GNU zipped tar file)

\item[HIV data set:] Data set used in the illustration of MYNEW method in Section~ 3.2. (.txt file)

\end{description}

\section{BibTeX}

We hope you've chosen to use BibTeX!\ If you have, please feel free to use the package natbib with any bibliography style you're comfortable with. The .bst file agsm has been included here for your convenience.

\bibliographystyle{Chicago}

\bibliography{Bibliography-MM-MC}
\end{document}

### 2021-12-29 text/jasa/jasa.tex
reformat text for journal

--- 2021-12-17 text/jsm/jsm_paper.tex
@@
-% set document class
-\documentclass[11pt]{article}
-
-% load packages
+\documentclass[12pt]{article}
+\usepackage{amsmath}
+\usepackage{graphicx}
+\usepackage{enumerate}
 \usepackage{setspace}
-\usepackage[margin=1in]{geometry}
 \usepackage{amsfonts}
@@
 
-% set options
-\doublespacing
+% MY COMMANDS
+\renewcommand{\algorithmicrequire}{\textbf{Input:}}
+\renewcommand{\algorithmicensure}{\textbf{Output:}}
 \addbibresource{/Users/timbarry/optionFiles/glmeiv.bib}
+\newtheorem{lemma}{Lemma}
+\newtheorem{proposition}{Proposition}
 \newcommand{\PreserveBackslash}[1]{\let\temp=\\#1\let\\=\temp}
 \newcolumntype{C}[1]{>{\PreserveBackslash\centering}p{#1}}
-\renewcommand{\algorithmicrequire}{\textbf{Input:}}
-\renewcommand{\algorithmicensure}{\textbf{Output:}}
+\newcommand{\indep}{\raisebox{0.05em}{\rotatebox[origin=c]{90}{$\models$}}}
+
 \makeatletter
@@
 \makeatother
-\setcounter{tocdepth}{2}
-\newtheorem{proposition}{Proposition}
-\newtheorem{lemma}{Lemma}
-\newcommand{\blue}[1]{\textcolor{blue}{#1}}
-\allowdisplaybreaks
-\usepackage[caption = false]{subfig}
-\usepackage{/Users/timbarry/optionFiles/mymacros}
-\DeclareMathOperator*{\argmin}{arg\,min}
-\titleformat{\subsection}[runin]
-{\normalfont\large\bfseries}{\thesubsection}{1em}{}
-\titleformat{\subsubsection}[runin]
-{\normalfont\normalsize\bfseries}{\thesubsubsection}{1em}{}
-\newcommand\blankpage{%
-	\null
-	\thispagestyle{empty}%
-	\addtocounter{page}{-1}%
-	\newpage}
+
+% JOURNAL SETTINGS
+% NOTE: To produce blinded version, replace "0" with "1" below.
+\newcommand{\blind}{1}
+\addtolength{\oddsidemargin}{-.5in}
+\addtolength{\evensidemargin}{-1in}
+\addtolength{\textwidth}{1in}
+\addtolength{\textheight}{1.7in}
+\addtolength{\topmargin}{-1in}
+\date{\vspace{-5ex}}
 
 \begin{document}
-\noindent
+\def\spacingset#1{\renewcommand{\baselinestretch}%
+{#1}\small\normalsize} \spacingset{1}
+\if1\blind
+{
+  \title{\bf A new class of measurement error models for CRISPR genome engineering and single-cell RNA sequencing}
+  \author{Timothy Barry$^1$, Eugene Katsevich$^2$, Kathyrn Roeder$^1$ \\
+    $^1$Department of Statistics and Data Science, Carnegie Mellon University \\
+    $^2$Department of Statistics and Data Science, Wharton School}
+  \maketitle
+} \fi
+
+\if0\blind
+{
+  \bigskip
+  \bigskip
+  \bigskip
   \begin{center}
-\textbf{A class of exponential family measurement error models for CRISPR genome engineering and single-cell sequencing} \\
-Timothy Barry, Eugene Katsevich, Kathryn Roeder
+    {\LARGE \bf A new class of measurement error models for CRISPR genome engineering and single-cell RNA sequencing}
 \end{center}
-
+  \medskip
+} \fi
+
+\bigskip
 \begin{abstract}
-CRISPR genome engineering and single-cell sequencing have transformed biological discovery.
+CRISPR genome engineering and single-cell RNA sequencing have transformed biological discovery.
 Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases.
 Despite their promise, single-cell CRISPR screens present substantial statistical challenges.
-We demonstrate through theoretical and real data analyses that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter.
-To overcome these limitations, we introduce GLM-EIV (``GLM-based errors-in-variables''), a new method for single-cell CRISPR screen analysis.
-GLM-EIV extends the classical errors-in-variables model to response distributions and sources of measurement error that are (i) exponential family-distributed and (ii) potentially impacted by the same set of confounding variables.
+We demonstrate through theoretical and real data analyses that a standard method for estimation and inference in single-cell CRISPR screens --``thresholded regression'' -- exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic and challenging-to-select tuning parameter.
+To overcome these difficulties, we introduce GLM-EIV (``GLM-based errors-in-variables''), a new method for single-cell CRISPR screen analysis.
+GLM-EIV extends the classical errors-in-variables model to responses and noisy predictors that are (i) exponential family-distributed and (ii) potentially impacted by the same set of confounding variables.
 We develop a computational infrastructure to deploy GLM-EIV across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters.
@@
 
+\noindent
+{\it Keywords:}  exponential family, errors-in-variables, GLM, CRISPR screen, differential expression
+\spacingset{1.9}
 \section{Introduction}
@@
 The genomics literature has produced several applied methods for single-cell CRISPR screen analysis.
-In a prior work we developed SCEPTRE \cite{Barry2020}, a custom implementation of the conditional randomization test \cite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data.
+Barry et al.\ developed SCEPTRE \cite{Barry2021}, a custom implementation of the conditional randomization test \cite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data.
 SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification.
@@
 
-Let $n \in \N$ be the number of cells assayed in the experiment.
+Let $n \in \mathbb{N}$ be the number of cells assayed in the experiment.
 Consider a single perturbation and a single gene.
-For cell $i \in \{1, \dots, n\}$, let $m_i \in \N$ be the number of gene transcripts sequenced; let $g_i \in \N$ be the number of gRNA transcripts sequenced; let $d^m_i \in \N$ be the number of gene transcripts sequenced across \textit{all} genes (i.e., the library size or sequencing depth); and finally, let $z_i \in \R^{d-1}$ be the cell-specific technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.) The letters ``m,'' ``g'', and ``d'' stand for ``mRNA,'' ``gRNA,'' and ``depth,'' respectively.
+For cell $i \in \{1, \dots, n\}$, let $m_i \in \mathbb{N}$ be the number of gene transcripts sequenced; let $g_i \in \mathbb{N}$ be the number of gRNA transcripts sequenced; let $d^m_i \in \mathbb{N}$ be the number of gene transcripts sequenced across \textit{all} genes (i.e., the library size or sequencing depth); and finally, let $z_i \in \mathbb{R}^{d-1}$ be the cell-specific technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.) The letters ``m,'' ``g'', and ``d'' stand for ``mRNA,'' ``gRNA,'' and ``depth,'' respectively.
 The thresholding method is defined as follows:
 \begin{itemize}
-\item[1.] For a given threshold $c \in \N$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be $$ \hat{p}_i = 0 \textrm{ if } g_i < c; \quad \hat{p}_i = 1 \textrm{ if } g_i \geq c.$$
+	\item[1.] For a given threshold $c \in \mathbb{N}$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be $$ \hat{p}_i = 0 \textrm{ if } g_i < c; \quad \hat{p}_i = 1 \textrm{ if } g_i \geq c.$$
 	%$$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i < c, \\ \hat{p}_i = 1 \textrm{ if } g_i \geq c. \end{cases}$$
@@
 	\end{equation}
-where (i) $NB_{\theta^m}(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta^m$; (ii) $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in  \R^{d-1}$ are unknown parameters; and (iii) $\log(d_i^m)$ is an offset term.
+	where (i) $NB_{\theta^m}(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta^m$; (ii) $\beta^m_0 \in \mathbb{R}, \beta^m_1 \in \mathbb{R},$ and $\gamma_m \in  \mathbb{R}^{d-1}$ are unknown parameters; and (iii) $\log(d_i^m)$ is an offset term.
 
@@
 \begin{equation}\label{theoretical_model}
-m_i = \beta^m_0 + \beta^m_1 p_i + \ep_i; \quad
+m_i = \beta^m_0 + \beta^m_1 p_i + \epsilon_i; \quad
 g_i = \beta^g_0 + \beta^g_1 p_i + \tau_i; \quad
 p_i \sim \textrm{Bern}(\pi); \quad
-\ep_i, \tau_i \sim N(0,1),
-\end{equation}
-where $p_i, \tau_i,$ and $\ep_i$ are independent.
-For a given threshold $c \in \R$, the imputed perturbation assignment $\hat{p}_i$ is $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ is the OLS solution, i.e.
+\epsilon_i, \tau_i \sim N(0,1),
+\end{equation}
+where $p_i, \tau_i,$ and $\epsilon_i$ are independent.
+For a given threshold $c \in \mathbb{R}$, the imputed perturbation assignment $\hat{p}_i$ is $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ is the OLS solution, i.e.
 $\hat{\beta}^m_1 = \left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})(m_i - \overline{m})\right]\left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})^2\right]^{-1}.$ We derive the almost sure limit of $\hat{\beta}^m_1$:
@@
 	\begin{equation}\label{thresh_est_intercepts}
-	\hat{\beta}^m_1 \xrightarrow{a.s.} \beta^m_1 \left(\frac{ \pi( \omega - \E[ \hat{p}_i ])}{ \E[\hat{p}_i] (1 - \E[\hat{p}_i])}\right) \equiv \beta^m_1 \gamma(\beta^g_1, \pi, c, \beta^g_0), 
+	\hat{\beta}^m_1 \xrightarrow{a.s.} \beta^m_1 \left(\frac{ \pi( \omega - \mathbb{E}[ \hat{p}_i ])}{ \mathbb{E}[\hat{p}_i] (1 - \mathbb{E}[\hat{p}_i])}\right) \equiv \beta^m_1 \gamma(\beta^g_1, \pi, c, \beta^g_0),
 	\end{equation} where 
 	$$
-	\E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi; \quad \omega = \Phi\left(\beta_1^g + \beta_0^g -c \right); \quad \zeta = \Phi\left( \beta^g_0 - c \right).$$
+	\mathbb{E}[\hat{p}_i] = \zeta(1-\pi) + \omega\pi; \quad \omega = \Phi\left(\beta_1^g + \beta_0^g -c \right); \quad \zeta = \Phi\left( \beta^g_0 - c \right).$$
 \end{proposition}
-The function $\gamma: \R^4 \to \R$ does not depend on the gene expression parameters $\beta^m_1$ or $\beta^m_0$. The asymptotic relative bias $b: \R^4 \to \R$ of $\hat{\beta}^m_1$ is given by
+The function $\gamma: \mathbb{R}^4 \to \mathbb{R}$ does not depend on the gene expression parameters $\beta^m_1$ or $\beta^m_0$.
+The asymptotic relative bias $b: \mathbb{R}^4 \to \mathbb{R}$ of $\hat{\beta}^m_1$ is given by
 $$b(\beta^g_1, \pi, c, \beta^g_0)  \equiv \frac{1}{\beta^m_1} \left(\beta^m_1 - \lim_\textrm{a.s.} \hat{\beta}^m_1 \right) = 1 - \gamma(\beta^g_1, \pi, c, \beta^g_0).$$
@@
 Second, the magnitude of the bias decreases monotonically in $\beta^g_1$, comporting with the intuition that the problem becomes easier as the gRNA mixture distribution becomes increasingly well-separated.
-Third, the Bayes-optimal decision boundary $c_\textrm{bayes} \in \R$ (i.e., the most accurate decision boundary for classifying cells as perturbed or unperturbed) is a critical value of the bias function.
+Third, the Bayes-optimal decision boundary $c_\textrm{bayes} \in \mathbb{R}$ (i.e., the most accurate decision boundary for classifying cells as perturbed or unperturbed) is a critical value of the bias function.
 Finally, and most subtly, there is no universally applicable rule for selecting a threshold that produces minimal bias: when $\beta^g_1$ is small, setting the threshold to an arbitrarily large number yields smaller bias than setting the threshold to the Bayes decision boundary; when $\beta^g_1$ is large, the reverse is true.
@@
 	The limiting distribution of $\hat{\beta}_m$ is
-	$$\sqrt{n}(\hat{\beta}_m - l) \xrightarrow{d} N\left(0, \frac{ \beta_m \omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2) }{\left(\E[\hat{p}_i]\right)^2} \right),$$ where $$l = \beta_m \omega \pi/[\zeta(1-\pi) + \omega \pi]; \quad
-	\E[\hat{p}_i] = \pi \omega + (1-\pi) \zeta; \quad
+	$$\sqrt{n}(\hat{\beta}_m - l) \xrightarrow{d} N\left(0, \frac{ \beta_m \omega\pi(\beta_m - 2l) + \mathbb{E}[\hat{p}_i](1 + l^2) }{\left(\mathbb{E}[\hat{p}_i]\right)^2} \right),$$ where $$l = \beta_m \omega \pi/[\zeta(1-\pi) + \omega \pi]; \quad
+	\mathbb{E}[\hat{p}_i] = \pi \omega + (1-\pi) \zeta; \quad
 	\omega = \Phi(\beta_g - c); \quad
@@
 \end{equation}
-where $\theta^m > 0$ is a known negative binomial size parameter, and $\beta^m_0 \in \R, \beta^m_1 \in \R,$ and $\gamma_m \in \R^{d - 2}$ are unknown constants.
+where $\theta^m > 0$ is a known negative binomial size parameter, and $\beta^m_0 \in \mathbb{R}, \beta^m_1 \in \mathbb{R},$ and $\gamma_m \in \mathbb{R}^{d - 2}$ are unknown constants.
 The model (\ref{glmeiv_model_1}) is identical to the thresholding model (\ref{thresh_glm}), but the imputed perturbation indicator $\hat{p}_i$ is replaced by the latent perturbation indicator $p_i$.
-Next, let $d^g_i \in \N$ be the number of gRNA transcripts sequenced across \textit{all} gRNAs in cell $i$ (i.e., the gRNA library size).
+Next, let $d^g_i \in \mathbb{N}$ be the number of gRNA transcripts sequenced across \textit{all} gRNAs in cell $i$ (i.e., the gRNA library size).
 The model for the gRNA counts is
@@
 \end{equation}
-where, similar to above, $\theta^g > 0$ is a known negative binomial size parameter, and $\beta^g_0 \in \R, \beta^g_1 \in \R, \gamma_g \in \R^{d - 2}$ are unknown constants.
+where, similar to above, $\theta^g > 0$ is a known negative binomial size parameter, and $\beta^g_0 \in \mathbb{R}, \beta^g_1 \in \mathbb{R}, \gamma_g \in \mathbb{R}^{d - 2}$ are unknown constants.
 We use a negative binomial GLM to model the gRNA counts because the gRNA molecules are transcribed in the cell in the same way as gene transcripts \cite{Datlinger2017,Hill2018}.
@@
 To provide greater modeling flexibility, we generalize the GLM-EIV model to arbitrary exponential family response distributions and link functions.
-To increase notational compactness, let $\tilde{x_i} = [1, p_i, z_i]^T \in \R^d$ be the vector of covariates (including an intercept term) for the $i$th cell.
-(We use the tilde as a reminder that the vector is partially unobserved.) Let $\beta_m = [\beta^m_0, \beta^m_1, \gamma_m]^T \in \R^d$ and $\beta_g = [\beta^g_0, \beta^g_1, \gamma_g]^T \in \R^d$ be the unknown coefficient vectors corresponding to the gene and gRNA expression models, respectively.
+To increase notational compactness, let $\tilde{x_i} = [1, p_i, z_i]^T \in \mathbb{R}^d$ be the vector of covariates (including an intercept term) for the $i$th cell.
+(We use the tilde as a reminder that the vector is partially unobserved.) Let $\beta_m = [\beta^m_0, \beta^m_1, \gamma_m]^T \in \mathbb{R}^d$ and $\beta_g = [\beta^g_0, \beta^g_1, \gamma_g]^T \in \mathbb{R}^d$ be the unknown coefficient vectors corresponding to the gene and gRNA expression models, respectively.
 Finally, let $o^m_i$ and $o^g_i$ be the (possibly zero) offset terms for the gene and gRNA models; in practice, we typically set $o^m_i$ and $o^g_i$ to the log-transformed library sizes (i.e., $\log(d^m_i)$ and $\log(d^g_i)$, respectively).
@@
 We use GLMs to model the gene and gRNA expressions.
-Considering first the gene expression model, let the $i$th linear component $l^m_i$ of the model be $l^m_i = \langle \tilde{x}_i, \beta_m \rangle + o^m_i.$ Next, let the mean $\mu^m_i$ of the $i$th observation be $r_m(\mu^m_i) = l^m_i,$ where $r_m:\R \to \R$ is a strictly increasing, differentiable link function.
-Let $\psi_m: \R \to \R$ be the differentiable, cumulant-generating function of the selected exponential family distribution.
+Considering first the gene expression model, let the $i$th linear component $l^m_i$ of the model be $l^m_i = \langle \tilde{x}_i, \beta_m \rangle + o^m_i.$ Next, let the mean $\mu^m_i$ of the $i$th observation be $r_m(\mu^m_i) = l^m_i,$ where $r_m:\mathbb{R} \to \mathbb{R}$ is a strictly increasing, differentiable link function.
+Let $\psi_m: \mathbb{R} \to \mathbb{R}$ be the differentiable, cumulant-generating function of the selected exponential family distribution.
 We can express the canonical parameter $\eta^m_i$ in terms of $\psi_m$ and $r_m$ by
-$\eta^m_i = \left([\psi'_m]^{-1} \circ r^{-1}_m\right)(l_i^m) := h_m(l_i^m).$ Finally, let $c_m: \R \to \R$ be the carrying density of the selected exponential family distribution.
+$\eta^m_i = \left([\psi'_m]^{-1} \circ r^{-1}_m\right)(l_i^m) := h_m(l_i^m).$ Finally, let $c_m: \mathbb{R} \to \mathbb{R}$ be the carrying density of the selected exponential family distribution.
 The density $f_m$ of $m_i$ conditional on the the canonical parameter $\eta_i$ is
@@
 The unknown parameters in the model are
-$\theta = [\beta_m, \beta_g, \pi]^{T}  \in \R^{2d + 1}.$
+$\theta = [\beta_m, \beta_g, \pi]^{T}  \in \mathbb{R}^{2d + 1}.$
 
@@
 Let the corresponding gRNA quantities $l^g_i(j)$, $\eta_i^g(j)$, and $\mu^g_i(j)$ be defined analogously.
-Let $X \in \R^{n \times {d-1}}$ be the observed design matrix, and let $\tilde{X} \in \R^{n \times d}$ be the augmented design matrix that results from concatenating the column of (unobserved) $p_i$s to $X$, i.e.
+Let $X \in \mathbb{R}^{n \times {d-1}}$ be the observed design matrix, and let $\tilde{X} \in \mathbb{R}^{n \times d}$ be the augmented design matrix that results from concatenating the column of (unobserved) $p_i$s to $X$, i.e.
 $$X := \begin{bmatrix} 
@@
 $$ 
-Furthermore, for $j \in \{0,1\}$, let $\tilde{X}(j) \in \R^{n \times d}$ be the matrix that results from setting $p_i$ to $j$ for all $i \in \{1, \dots, n\}$ in $\tilde{X}$, and let  $[\tilde{X}(0)^T, \tilde{X}(1)^T]^T$ denote the $\R^{2n \times d}$ matrix that results from vertically concatenating $\tilde{X}(0)$ and $\tilde{X}(1)$.
+Furthermore, for $j \in \{0,1\}$, let $\tilde{X}(j) \in \mathbb{R}^{n \times d}$ be the matrix that results from setting $p_i$ to $j$ for all $i \in \{1, \dots, n\}$ in $\tilde{X}$, and let  $[\tilde{X}(0)^T, \tilde{X}(1)^T]^T$ denote the $\mathbb{R}^{2n \times d}$ matrix that results from vertically concatenating $\tilde{X}(0)$ and $\tilde{X}(1)$.
 Furthermore, define $m := [m_1, \dots, m_n]$, and let $g$, $p$, $o^m$, and $o^g$ be defined analogously.
-Finally, let $[m,m]^T \in \R^{2n}$ be the vector that results from concatenating $m$ to itself, i.e.
+Finally, let $[m,m]^T \in \mathbb{R}^{2n}$ be the vector that results from concatenating $m$ to itself, i.e.
 $$ [m,m]^T := [\underbrace{m_1, m_2 \dots, m_{n-1}, m_n}_\textrm{first copy of $m$}, \underbrace{m_1, m_2, \dots, m_{n-1}, m_n}_\textrm{second copy of $m$}],$$ and let $[g,g]^T$, $[o^g,o^g]^T$, and $[o^m,o^m]^T$ be defined similarly.
@@
 \begin{equation}\label{classical_eiv}
-y_i = \beta_0 + \beta_1 x^*_i + \ep_i; \quad
+y_i = \beta_0 + \beta_1 x^*_i + \epsilon_i; \quad
 x_i = x^*_i + \tau_i,
 \end{equation}
-where, $x^*_i \sim \textrm{Bern}(\pi), \ep_i, \tau_i \sim N(0,1),$ and $\ep_i$,$\tau_i$, and $x^*_i$ are independent.
+where, $x^*_i \sim \textrm{Bern}(\pi), \epsilon_i, \tau_i \sim N(0,1),$ and $\epsilon_i$,$\tau_i$, and $x^*_i$ are independent.
 GLM-EIV extends (\ref{classical_eiv}) in at least three directions: first, GLM-EIV allows $y_i$ and $x_i$ to follow exponential family (i.e, not just Gaussian) distributions; second, GLM-EIV allows $y_i$ and $x_i$ to be related to $x^*_i$ through linear \textit{or} nonlinear link functions; and finally, GLM-EIV allows confounders $z_i$ to impact both $x_i$ and $y_i$.
@@
 The membership probability $T_i(1)$ of cell $i \in \{1, \dots, n\}$ given the current parameter estimates $(\beta_m^\textrm{(t)}, \beta_g^\textrm{(t)}, \pi^\textrm{(t)})$ and observed data $(m_i, g_i)$ is
-$$T_i(1) = \P(p_i = 1| M_i = m_i, G_i = g_i, \beta^\textrm{(t)}_m, \beta^\textrm{(t)}_g, \pi^\textrm{(t)}).$$ We can calculate this quantity by applying (i) Bayes rule, (ii) the conditional independence property of $M_i$ and $G_i$, (iii) the density of $M_i$ and $G_i$, and (iv) a log-sum-exp-type trick to ensure numerical stability.
+$$T_i(1) = \mathbb{P}(p_i = 1| M_i = m_i, G_i = g_i, \beta^\textrm{(t)}_m, \beta^\textrm{(t)}_g, \pi^\textrm{(t)}).$$ We can calculate this quantity by applying (i) Bayes rule, (ii) the conditional independence property of $M_i$ and $G_i$, (iii) the density of $M_i$ and $G_i$, and (iv) a log-sum-exp-type trick to ensure numerical stability.
 Next, we produce updated estimates $\pi^{(\textrm{t} +1)}$, $\beta_g^{(\textrm{t}+1)}$, and $\beta_m^{(\textrm{t}+1)}$ of the parameters by maximizing the M step objective function.
@@
 		\For{$i \in \{1, \dots, n\}$} \Comment{E step}
-			\State $T_i(1) \gets \P\left(p_i = 1 |M_i = m_i, G_i = g_i, \beta_m^\textrm{curr}, \beta_g^\textrm{curr}, \pi^\textrm{curr} \right)$
+		\State $T_i(1) \gets \mathbb{P}\left(p_i = 1 |M_i = m_i, G_i = g_i, \beta_m^\textrm{curr}, \beta_g^\textrm{curr}, \pi^\textrm{curr} \right)$
 		\State $T_i(0) \gets 1 - T_i(1)$
@@
 The first step (Algorithm \ref{algo:pilot_estimates_1}) is to obtain good parameter estimates for $[\beta^m_0, \gamma_m]^T$ and $[\beta^g_0, \gamma_g]^T$ via regression.
-Recall that the underlying gene expression parameter vector $\beta_m$ is $\beta_m = [\beta^m_0, \beta^m_1, \gamma_m]^T \in \R^d$, where $\beta^m_0$ is the intercept, $\beta^m_1$ is the effect of the perturbation, and $\gamma_m^T$ is the effect of the technical factors.
+Recall that the underlying gene expression parameter vector $\beta_m$ is $\beta_m = [\beta^m_0, \beta^m_1, \gamma_m]^T \in \mathbb{E}^d$, where $\beta^m_0$ is the intercept, $\beta^m_1$ is the effect of the perturbation, and $\gamma_m^T$ is the effect of the technical factors.
 To produce estimates $[\beta^m_0]^\textrm{pilot}$ and $[\gamma_m^T]^\textrm{pilot}$, we regress the gene expressions $m$ onto the technical factors $X$.
@@
 \clearpage
+
 \begin{appendices}
@@
 		First, the thresholding method incurs strict attenuation bias (i.e., it \textit{under}estimates the true effect size) for all choices of the threshold and over all possible values of the model parameters:
-\begin{proposition}\label{prop:att_bias} Fix $\pi = 1/2$. For all $(\beta^g_1, c, \beta^g_0) \in \R^3$, the asymptotic relative bias is positive, i.e. 
+		\begin{proposition}\label{prop:att_bias} Fix $\pi = 1/2$. For all $(\beta^g_1, c, \beta^g_0) \in \mathbb{E}^3$, the asymptotic relative bias is positive, i.e. 
 			$$b(\beta^g_1, 1/2, c, \beta^g_0) > 0.$$
@@
 		\begin{proposition}\label{prop:bayes_opt}
-		For $\pi = 1/2$ and given $(\beta^g_1, \beta^g_0) \in \R^2$, the Bayes-optimal decision boundary $c_\textrm{bayes}$ is a critical value of the bias function $b$, i.e.
+			For $\pi = 1/2$ and given $(\beta^g_1, \beta^g_0) \in \mathbb{E}^2$, the Bayes-optimal decision boundary $c_\textrm{bayes}$ is a critical value of the bias function $b$, i.e.
 			$$ \frac{\partial b}{\partial c}\left(\beta^g_1, 1/2, c_\textrm{bayes}, \beta^g_0\right) = 0.$$
@@
 		All notation introduced in this subsection (i.e., \ref{sec:notation}) pertains to the Gaussian model with intercepts (\ref{theoretical_model}).
-Recall that the attenuation function $\gamma: \R^4 \to \R$ is defined by
-$$ \gamma(\beta^g_1, c, \pi, \beta^g_0) = \frac{\pi(\omega - \E[\hat{p}_i])}{ \E[\hat{p}_i](1 -\E[\hat{p}_i])},$$ where $$\E[\hat{p}_i] = \zeta(1-\pi) + \omega\pi; \quad \omega = \Phi\left(\beta_1^g + \beta_0^g - c \right); \quad \zeta = \Phi\left( \beta^g_0 - c \right).$$ Additionally, recall that the asymptotic relative bias function $b: \R^4 \to \R$ is
-$$ b(\beta^g_1, c, \pi, \beta^g_0) = 1 - \gamma(\beta^g_1, c, \pi, \beta^g_0).$$ Next, we define the functions $g$ and $h: \R^4 \to \R$ by
+Recall that the attenuation function $\gamma: \mathbb{E}^4 \to \mathbb{E}$ is defined by
+		$$ \gamma(\beta^g_1, c, \pi, \beta^g_0) = \frac{\pi(\omega - \mathbb{E}[\hat{p}_i])}{ \mathbb{E}[\hat{p}_i](1 -\mathbb{E}[\hat{p}_i])},$$ where $$\mathbb{E}[\hat{p}_i] = \zeta(1-\pi) + \omega\pi; \quad \omega = \Phi\left(\beta_1^g + \beta_0^g - c \right); \quad \zeta = \Phi\left( \beta^g_0 - c \right).$$ Additionally, recall that the asymptotic relative bias function $b: \mathbb{E}^4 \to \mathbb{E}$ is
+		$$ b(\beta^g_1, c, \pi, \beta^g_0) = 1 - \gamma(\beta^g_1, c, \pi, \beta^g_0).$$ Next, we define the functions $g$ and $h: \mathbb{E}^4 \to \mathbb{E}$ by
 		\begin{equation}\label{def_g}
@@
 		\end{multline}
-We use $f:\R \to \R$ to denote the $N(0,1)$ density, and we denote the right-tail probability probability of $f$ by $\bar{\Phi}$, i.e.,
+		We use $f:\mathbb{E} \to \mathbb{E}$ to denote the $N(0,1)$ density, and we denote the right-tail probability probability of $f$ by $\bar{\Phi}$, i.e.,
 		$$\bar{\Phi}(x) = \int_{x}^{\infty} f = \Phi(-x).$$
@@
 Therefore, to minimize notation, we typically use $\gamma(\beta^g_1, c, \pi)$ (resp., $b(\beta^g_1, c, \pi),$ $g(\beta^g_1, c, \pi),$ $h(\beta^g_1, c, \pi)$) to refer to the function $\gamma$ (resp., $b, g, h$) evaluated at $(\beta^g_1, c, \pi, \beta^g_0)$.
-Finally, for a given function $r: \R^{p} \to \R$, point $x \in \R^p$, and index $i \in \{1, \dots, p\}$, we use the symbol $D_i r(x)$ to refer to the derivative of the $i$th component of $r$ evaluated at $x$ (\textit{sensu} \cite{fitzpatrick2009}).
+Finally, for a given function $r: \mathbb{E}^{p} \to \mathbb{E}$, point $x \in \mathbb{E}^p$, and index $i \in \{1, \dots, p\}$, we use the symbol $D_i r(x)$ to refer to the derivative of the $i$th component of $r$ evaluated at $x$ (\textit{sensu} \cite{fitzpatrick2009}).
 For example, $D_1 \gamma(\beta^g_1, c, 1/2)$ is the derivative of the first component of $\gamma$ (the component corresponding to $\beta^g_1$) evaluated at $(\beta^g_1, c, 1/2)$.
@@
 		$$ \hat{\beta}^m_1 = \frac{ \frac{1}{n} \sum_{i=1}^n ( \hat{p}_i - \overline{\hat{p}_i})(m_i - \overline{m})}{ \frac{1}{n} \sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})}.$$ By weak LLN,
-$$ \hat{\beta}^m_1 \xrightarrow{P} \frac{\textrm{Cov}(\hat{p}_i, m_i)}{\V\left(\hat{p}_i\right)}.$$ To compute this quantity, we first compute several simpler quantities:
+		$$ \hat{\beta}^m_1 \xrightarrow{P} \frac{\textrm{Cov}(\hat{p}_i, m_i)}{\mathbb{V}\left(\hat{p}_i\right)}.$$ To compute this quantity, we first compute several simpler quantities:
 		\begin{itemize}
-\item[1.] Expectation of $m_i$: $\E[m_i] = \beta^m_0 + \beta^m_1\pi.$
+			\item[1.] Expectation of $m_i$: $\mathbb{E}[m_i] = \beta^m_0 + \beta^m_1\pi.$
 			\item[2.] Expectation of $\hat{p}_i$: \begin{multline*}
-\E[\hat{p}_i] = \P\left[\hat{p}_i = 1\right] = \P\left[\beta^g_0 + \beta^g_1 p_i + \tau_i \geq c \right] = \\ \textrm{(By LOTP) } \P\left[ \beta^g_0 + \tau_i \geq c \right]\P\left[p_i = 0\right] + \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \P[p_i = 1] \\ = \P\left[ \tau_i \geq c - \beta^g_0\right](1- \pi) + \P\left[ \tau_i \geq c - \beta^g_1 - \beta^g_0 \right](\pi) \\ =  \left(\bar{\Phi}(c - \beta^g_0) \right) (1 - \pi) + \left( \bar{\Phi}(c - \beta^g_1 - \beta^g_0) \right)(\pi) = \\  \Phi(\beta^g_0 - c) (1-\pi) + \Phi(\beta^g_1 + \beta^g_0 - c) \pi = \zeta(1-\pi) + \omega \pi.
+			\mathbb{E}[\hat{p}_i] = \mathbb{P}\left[\hat{p}_i = 1\right] = \mathbb{P}\left[\beta^g_0 + \beta^g_1 p_i + \tau_i \geq c \right] = \\ \textrm{(By LOTP) } \mathbb{P}\left[ \beta^g_0 + \tau_i \geq c \right]\mathbb{P}\left[p_i = 0\right] + \mathbb{P}\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \mathbb{P}[p_i = 1] \\ = \mathbb{P}\left[ \tau_i \geq c - \beta^g_0\right](1- \pi) + \mathbb{P}\left[ \tau_i \geq c - \beta^g_1 - \beta^g_0 \right](\pi) \\ =  \left(\bar{\Phi}(c - \beta^g_0) \right) (1 - \pi) + \left( \bar{\Phi}(c - \beta^g_1 - \beta^g_0) \right)(\pi) = \\  \Phi(\beta^g_0 - c) (1-\pi) + \Phi(\beta^g_1 + \beta^g_0 - c) \pi = \zeta(1-\pi) + \omega \pi.
 			\end{multline*}
 			\item[3.] Expectation of $\hat{p}_i p_i$: 
-$\E\left[ \hat{p}_i p_i \right] = \E\left[\hat{p}_i | p_i = 1 \right] \P\left[ p_i =1 \right] = \P\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \pi = \omega \pi.$
+			$\mathbb{E}\left[ \hat{p}_i p_i \right] = \mathbb{E}\left[\hat{p}_i | p_i = 1 \right] \mathbb{P}\left[ p_i =1 \right] = \mathbb{P}\left[ \beta^g_0 + \beta^g_1 + \tau_i \geq c \right] \pi = \omega \pi.$
 			\item[4.] Expectation of $\hat{p}_i m_i$:
 			\begin{multline*}
-\E\left[\hat{p}_i m_i\right] = \E[\hat{p}_i (\beta^m_0 + \beta^m_1 p_i + \ep_i)] = \beta^m_0 \E\left[\hat{p}_i\right] + \beta^m_1 \E\left[\hat{p}_i p_i\right] + \E[\hat{p}_i \ep_i] \\ = \beta^m_0 \E[\hat{p}_i] + \beta^m_1 \omega \pi + \E[\hat{p}_i] \E[\ep_i] = \beta^m_0 \E[\hat{p}_i] + \beta^m_1 \omega \pi.
-\end{multline*}
-\item[5.] Variance of $\hat{p}_i$: Because $\hat{p}_i$ is binary, we have that $\V[\hat{p}_i] = \E[\hat{p}_i]\left(1 - \E[\hat{p}_i]\right) .$
+			\mathbb{E}\left[\hat{p}_i m_i\right] = \mathbb{E}[\hat{p}_i (\beta^m_0 + \beta^m_1 p_i + \epsilon_i)] = \beta^m_0 \mathbb{E}\left[\hat{p}_i\right] + \beta^m_1 \mathbb{E}\left[\hat{p}_i p_i\right] + \mathbb{E}[\hat{p}_i \epsilon_i] \\ = \beta^m_0 \mathbb{E}[\hat{p}_i] + \beta^m_1 \omega \pi + \mathbb{E}[\hat{p}_i] \mathbb{E}[\epsilon_i] = \beta^m_0 \mathbb{E}[\hat{p}_i] + \beta^m_1 \omega \pi.
+			\end{multline*}
+			\item[5.] Variance of $\hat{p}_i$: Because $\hat{p}_i$ is binary, we have that $\mathbb{V}[\hat{p}_i] = \mathbb{E}[\hat{p}_i]\left(1 - \mathbb{E}[\hat{p}_i]\right) .$
 			\item[6.] Covariance of $\hat{p}_i, m_i$:
 			\begin{multline*}
-\textrm{Cov}\left(\hat{p}_i, m_i\right) = \E\left[\hat{p}_i m_i\right] - \E[\hat{p}_i] \E[m_i] = \beta^m_0 \E[\hat{p}_i] + \beta^m_1 \omega \pi - \E[\hat{p}_i]( \beta^m_0 + \beta^m_1 \pi)\\ = \beta^m_1 \omega \pi - \E[\hat{p}_i] \beta_1^m \pi = \beta^m_1 \pi \left( \omega - \E[\hat{p}_i]\right).
+			\textrm{Cov}\left(\hat{p}_i, m_i\right) = \mathbb{E}\left[\hat{p}_i m_i\right] - \mathbb{E}[\hat{p}_i] \mathbb{E}[m_i] = \beta^m_0 \mathbb{E}[\hat{p}_i] + \beta^m_1 \omega \pi - \mathbb{E}[\hat{p}_i]( \beta^m_0 + \beta^m_1 \pi)\\ = \beta^m_1 \omega \pi - \mathbb{E}[\hat{p}_i] \beta_1^m \pi = \beta^m_1 \pi \left( \omega - \mathbb{E}[\hat{p}_i]\right).
 			\end{multline*}
 			Combining these expressions, we have that
-$$ \hat{\beta}^m_1 \xrightarrow{P} \frac{\beta^m_1 \pi (\omega - \E[\hat{p}_i])}{\E[\hat{p}_i](1 - \E[\hat{p}_i])} = \beta^m_1 \gamma(\beta^g_1, c, \pi).$$
+			$$ \hat{\beta}^m_1 \xrightarrow{P} \frac{\beta^m_1 \pi (\omega - \mathbb{E}[\hat{p}_i])}{\mathbb{E}[\hat{p}_i](1 - \mathbb{E}[\hat{p}_i])} = \beta^m_1 \gamma(\beta^g_1, c, \pi).$$
 		\end{itemize}
@@
 		We rewrite the attenuation fraction $\gamma$ in a way that makes it more amenable to theoretical analysis. We leverage the fact that $f$ integrates to unity and is even. We have that
-\begin{equation}\label{thm:gamma_expression_1} \E\left[\hat{p}_i\right] = (1 - \pi) \bar{\Phi}(c - \beta_0^g) + \pi \bar{\Phi}(c - \beta^g_0 - \beta^g_1) = (1 - \pi) \Phi(\beta_0^g - c) + \pi\Phi(\beta^g_0 + \beta^g_1 - c), \end{equation}
- and so \begin{multline}\label{thm:gamma_expression_2} 1 - \E\left[\hat{p}_i\right] = (1 - \pi) + \pi - \E[\hat{p}_i]  = (1-\pi) \left(1 - \bar{\Phi}(c - \beta_0^g)\right)  + \pi \left(1 - \bar{\Phi}(c - \beta^g_0 - \beta^g_1) \right) \\ = (1 - \pi)\Phi(c - \beta^g_0) + \pi \Phi(c - \beta_0^g - \beta_1^g).
+		\begin{equation}\label{thm:gamma_expression_1} \mathbb{E}\left[\hat{p}_i\right] = (1 - \pi) \bar{\Phi}(c - \beta_0^g) + \pi \bar{\Phi}(c - \beta^g_0 - \beta^g_1) = (1 - \pi) \Phi(\beta_0^g - c) + \pi\Phi(\beta^g_0 + \beta^g_1 - c), \end{equation}
+		and so \begin{multline}\label{thm:gamma_expression_2} 1 - \mathbb{E}\left[\hat{p}_i\right] = (1 - \pi) + \pi - \mathbb{E}[\hat{p}_i]  = (1-\pi) \left(1 - \bar{\Phi}(c - \beta_0^g)\right)  + \pi \left(1 - \bar{\Phi}(c - \beta^g_0 - \beta^g_1) \right) \\ = (1 - \pi)\Phi(c - \beta^g_0) + \pi \Phi(c - \beta_0^g - \beta_1^g).
 		\end{multline}
@@
 		\begin{multline}\label{thm:gamma_expression_4}
-\omega - \E[\hat{p}_i] = \Phi(\beta^g_1 + \beta^g_0 - c) - (1-\pi)\Phi(\beta^g_0 - c) - \pi \Phi(\beta^g_0 + \beta^g_1 - c)  \\ (1-\pi)\Phi(\beta^g_1 + \beta^g_0 - c)  - (1-\pi)\Phi(\beta^g_0 - c).
+		\omega - \mathbb{E}[\hat{p}_i] = \Phi(\beta^g_1 + \beta^g_0 - c) - (1-\pi)\Phi(\beta^g_0 - c) - \pi \Phi(\beta^g_0 + \beta^g_1 - c)  \\ (1-\pi)\Phi(\beta^g_1 + \beta^g_0 - c)  - (1-\pi)\Phi(\beta^g_0 - c).
 		\end{multline}
@@
 		\begin{multline}\label{gamma_alternative}
-\gamma(\beta^g_1, c, \pi) = \frac{\pi(\omega - \E[\hat{p}_i])}{\E[\hat{p}_i](1 - \E[\hat{p}_i])} \\ = \frac{\pi \left[(1 - \pi) \Phi(\beta_0^g + \beta_1^g - c) - (1 - \pi) \Phi(\beta_0^g - c)\right]}{\left[(1-\pi)\Phi(\beta_0^g - c) + \pi \Phi(\beta^g_0 + \beta^g_1 - c) \right] \left[(1 - \pi) \Phi(c - \beta^g_0) + \pi\Phi(c - \beta_0^g - \beta_1^g) \right]}.
+		\gamma(\beta^g_1, c, \pi) = \frac{\pi(\omega - \mathbb{E}[\hat{p}_i])}{\mathbb{E}[\hat{p}_i](1 - \mathbb{E}[\hat{p}_i])} \\ = \frac{\pi \left[(1 - \pi) \Phi(\beta_0^g + \beta_1^g - c) - (1 - \pi) \Phi(\beta_0^g - c)\right]}{\left[(1-\pi)\Phi(\beta_0^g - c) + \pi \Phi(\beta^g_0 + \beta^g_1 - c) \right] \left[(1 - \pi) \Phi(c - \beta^g_0) + \pi\Phi(c - \beta_0^g - \beta_1^g) \right]}.
 		\end{multline}
@@
 		We compare the bias produced by setting the threshold to a large number to the bias produced by setting the threshold to the Bayes-optimal decision boundary.
-Let $r: \R^{\geq 0} \to \R$ be the value of attenuation function evaluated at the Bayes-optimal decision boundary $c_\textrm{bayes} = \beta^g_0 + (1/2) \beta^g_1$, i.e.
+Let $r: \mathbb{E}^{\geq 0} \to \mathbb{E}$ be the value of attenuation function evaluated at the Bayes-optimal decision boundary $c_\textrm{bayes} = \beta^g_0 + (1/2) \beta^g_1$, i.e.
 		\begin{multline*}
@@
 		\end{equation*}
-The inequality holds for $\beta^g_1 \geq 0$. Combining the cases, the inequality holds for all $\beta^g_1 \in \R$. $\square$
+		The inequality holds for $\beta^g_1 \geq 0$. Combining the cases, the inequality holds for all $\beta^g_1 \in \mathbb{E}$. $\square$
 
@@
 It follows that
-$$-1 < -\frac{1}{1 + \Phi(c - \beta^g_0)} \leq \gamma(\beta^g_1, c, 1/2, \beta^g_0) \leq \frac{1}{1 - \Phi(\beta^g_0 - c)} < 1$$ for all $\beta^g_1 \in \R$.
+		$$-1 < -\frac{1}{1 + \Phi(c - \beta^g_0)} \leq \gamma(\beta^g_1, c, 1/2, \beta^g_0) \leq \frac{1}{1 - \Phi(\beta^g_0 - c)} < 1$$ for all $\beta^g_1 \in \mathbb{E}$.
 But $\beta^g_0$ and $c$ were chosen arbitrarily, and so
-$$-1 < \gamma(\beta^g_1, c, 1/2, \beta^g_0) < 1$$ for all $(\beta^g_1, c, \beta^g_0) \in \R^3$.
+		$$-1 < \gamma(\beta^g_1, c, 1/2, \beta^g_0) < 1$$ for all $(\beta^g_1, c, \beta^g_0) \in \mathbb{E}^3$.
 Finally, because $b(\beta^g_1, c, 1/2, \beta^g_0) = 1 - \gamma(\beta^g_1, c, 1/2, \beta^g_0)$, it follows that
-$$ 0 < b(\beta^g_1, c, 1/2, \beta^g_0) < 2$$ for all $(\beta^g_1, c, \beta^g_0) \in \R^3$
+		$$ 0 < b(\beta^g_1, c, 1/2, \beta^g_0) < 2$$ for all $(\beta^g_1, c, \beta^g_0) \in \mathbb{E}^3$
 
@@
 			\item Expectation of $\hat{p}_i$: $
-\E[\hat{p}_i] = \P(p_i\beta_g + \tau_i \geq c) =  \P(\beta_g + \tau_i \geq c)\pi + \P(\tau_i \geq c)(1-\pi) = \pi \omega + (1-\pi)\zeta.$
-
-\item Expectation of $\hat{p}_i p_i$: $\E\left[\hat{p}_i p_i\right] = \E\left[\hat{p}_i | p_i = 1 \right]\P\left[p_i = 1\right] = \omega \pi.$
+			\mathbb{E}[\hat{p}_i] = \mathbb{P}(p_i\beta_g + \tau_i \geq c) =  \mathbb{P}(\beta_g + \tau_i \geq c)\pi + \mathbb{P}(\tau_i \geq c)(1-\pi) = \pi \omega + (1-\pi)\zeta.$
+
+			\item Expectation of $\hat{p}_i p_i$: $\mathbb{E}\left[\hat{p}_i p_i\right] = \mathbb{E}\left[\hat{p}_i | p_i = 1 \right]\mathbb{P}\left[p_i = 1\right] = \omega \pi.$
 			\item Expectation of $\hat{p}_i m_i$:
 			\begin{multline*}
-\E[\hat{p}_i m_i] = \E\left[\hat{p}_i(\beta_m p_i + \ep_i)\right] = \E\left[\beta_m \hat{p}_i p_i + \hat{p}_i \ep_i \right] \\ = \beta_m \E\left[ \hat{p}_i p_i \right] + \E[\hat{p}_i]\E[\ep_i] = \beta_m \omega \pi + 0 = \beta_m \omega \pi.
+			\mathbb{E}[\hat{p}_i m_i] = \mathbb{E}\left[\hat{p}_i(\beta_m p_i + \epsilon_i)\right] = \mathbb{E}\left[\beta_m \hat{p}_i p_i + \hat{p}_i \epsilon_i \right] \\ = \beta_m \mathbb{E}\left[ \hat{p}_i p_i \right] + \mathbb{E}[\hat{p}_i]\mathbb{E}[\epsilon_i] = \beta_m \omega \pi + 0 = \beta_m \omega \pi.
 			\end{multline*}
 			\item Expectation of $\hat{p}_i m_i^2$: \begin{multline*}
-\E\left[\hat{p}_i m_i^2\right] = \E \left[ \hat{p}_i( \beta_m p_i + \ep_i )^2 \right] = \E\left[ \hat{p}_i \left( \beta_m^2 p_i^2 + 2 \beta_m p_i \ep_i + \ep_i^2 \right)  \right] \\ = \E\left[ \hat{p}_i p_i \beta^2_m + 2 \beta_m p_i \hat{p}_i \ep_i + \hat{p}_i \ep_i^2 \right] = \beta^2_m \E[ \hat{p}_i p_i] + 2 \beta_m \E[p_i\hat{p}_i] \E[\ep_i] + \E[\hat{p}_i] \E[ \ep^2_i ] \\ = \beta^2_m \E[ \hat{p}_i p_i] + \E[\hat{p}_i] = \beta^2_m \omega \pi + \E[ \hat{p}_i].
+			\mathbb{E}\left[\hat{p}_i m_i^2\right] = \mathbb{E} \left[ \hat{p}_i( \beta_m p_i + \epsilon_i )^2 \right] = \mathbb{E}\left[ \hat{p}_i \left( \beta_m^2 p_i^2 + 2 \beta_m p_i \epsilon_i + \epsilon_i^2 \right)  \right] \\ = \mathbb{E}\left[ \hat{p}_i p_i \beta^2_m + 2 \beta_m p_i \hat{p}_i \epsilon_i + \hat{p}_i \epsilon_i^2 \right] = \beta^2_m \mathbb{E}[ \hat{p}_i p_i] + 2 \beta_m \mathbb{E}[p_i\hat{p}_i] \mathbb{E}[\epsilon_i] + \mathbb{E}[\hat{p}_i] \mathbb{E}[ \epsilon^2_i ] \\ = \beta^2_m \mathbb{E}[ \hat{p}_i p_i] + \mathbb{E}[\hat{p}_i] = \beta^2_m \omega \pi + \mathbb{E}[ \hat{p}_i].
 			\end{multline*}
@@
 		\begin{equation}\label{bv_decomp_2}
-\E\left[\hat{p}_i(m_i - l\hat{p}_i) \right] = \E[\hat{p}_i m_i] - l \E[\hat{p}_i] = \beta_m \omega \pi - \left(\frac{\beta_m \omega \pi}{\zeta (1-\pi) + \omega \pi}\right)[\zeta (1-\pi) + \omega \pi] = 0.
+		\mathbb{E}\left[\hat{p}_i(m_i - l\hat{p}_i) \right] = \mathbb{E}[\hat{p}_i m_i] - l \mathbb{E}[\hat{p}_i] = \beta_m \omega \pi - \left(\frac{\beta_m \omega \pi}{\zeta (1-\pi) + \omega \pi}\right)[\zeta (1-\pi) + \omega \pi] = 0.
 		\end{equation}
@@
 		\begin{multline}\label{bv_decomp_3}
-\V\left[\hat{p}_i(m_i - l\hat{p}_i)\right] = \E\left[\hat{p}_i^2(m_i - l\hat{p}_i)^2\right] - (\E\left[ \hat{p}_i(m_i - l\hat{p}_i)\right])^2 \\ = \E\left[ \hat{p}_i m_i^2\right] - 2l \E[m_i\hat{p}_i] +l^2 \E[\hat{p}_i]= \beta^2_m \omega \pi + \E[ \hat{p}_i] -2l \beta_m \omega \pi + l^2 \E[\hat{p}_i] \\ = \beta_m\omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2).
+		\mathbb{V}\left[\hat{p}_i(m_i - l\hat{p}_i)\right] = \mathbb{E}\left[\hat{p}_i^2(m_i - l\hat{p}_i)^2\right] - (\mathbb{E}\left[ \hat{p}_i(m_i - l\hat{p}_i)\right])^2 \\ = \mathbb{E}\left[ \hat{p}_i m_i^2\right] - 2l \mathbb{E}[m_i\hat{p}_i] +l^2 \mathbb{E}[\hat{p}_i]= \beta^2_m \omega \pi + \mathbb{E}[ \hat{p}_i] -2l \beta_m \omega \pi + l^2 \mathbb{E}[\hat{p}_i] \\ = \beta_m\omega\pi(\beta_m - 2l) + \mathbb{E}[\hat{p}_i](1 + l^2).
 		\end{multline}
@@
 		\begin{equation}\label{bv_decomp_4}
-(1/\sqrt{n})\sum_{i=1}^n \hat{p}_i(m_i - l \hat{p}_i) \xrightarrow{d} N\left(0, \beta_m\omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2) \right).
+		(1/\sqrt{n})\sum_{i=1}^n \hat{p}_i(m_i - l \hat{p}_i) \xrightarrow{d} N\left(0, \beta_m\omega\pi(\beta_m - 2l) + \mathbb{E}[\hat{p}_i](1 + l^2) \right).
 		\end{equation}
@@
 		\begin{equation}\label{bv_decomp_5}
-(1/n) \sum_{i=1}^n \hat{p}_i^2 = (1/n) \sum_{i=1}^n \hat{p}_i \xrightarrow{P} \E[\hat{p}_i].
+		(1/n) \sum_{i=1}^n \hat{p}_i^2 = (1/n) \sum_{i=1}^n \hat{p}_i \xrightarrow{P} \mathbb{E}[\hat{p}_i].
 		\end{equation}
 		Finally, by (\ref{bc_decomp_1}), (\ref{bv_decomp_4}), (\ref{bv_decomp_5}), and Slutsky's Theorem,
-$$ \sqrt{n}(\hat{\beta}_m - l) \xrightarrow{d} N\left(0, \frac{ \beta_m\omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2) }{\left(\E[\hat{p}_i]\right)^2} \right).$$ Thus, for large $n \in \N$, we have that
-$$
-\E [\hat{\beta}_m] \approx l; \quad
-\V[\hat{\beta}_m] \approx \left[\beta_m\omega\pi(\beta_m - 2l) + \E[\hat{p}_i](1 + l^2)\right]/[n\E^2[\hat{p}_i]],$$
+		$$ \sqrt{n}(\hat{\beta}_m - l) \xrightarrow{d} N\left(0, \frac{ \beta_m\omega\pi(\beta_m - 2l) + \mathbb{E}[\hat{p}_i](1 + l^2) }{\left(\mathbb{E}[\hat{p}_i]\right)^2} \right).$$ Thus, for large $n \in \mathbb{N}$, we have that
+		$$
+		\mathbb{E} [\hat{\beta}_m] \approx l; \quad
+		\mathbb{V}[\hat{\beta}_m] \approx \left[\beta_m\omega\pi(\beta_m - 2l) + \mathbb{E}[\hat{p}_i](1 + l^2)\right]/[n\mathbb{E}^2[\hat{p}_i]],$$
 		completing the bias-variance decomposition.
@@
 		\begin{multline*}
-\alpha^{(t)}_i(k) := \P\left( M_i = m_i, G_i = g_i | P_i = k, \theta^{(t)} \right) \\ = \P\left( M_i = m_i | P_i = k, \theta^{(t)} \right) \P\left(G_i = g_i | P_i = k, \theta^{(t)} \right) \textrm{ (because $G_i \indep M_i | P_i$)} \\ = f_m\left(m_i; \left[ \eta^m_i(k) \right]^{(t)}\right) f_g\left(g_i; \left[ \eta^g_i(k) \right]^{(t)} \right).
+		\alpha^{(t)}_i(k) := \mathbb{P}\left( M_i = m_i, G_i = g_i | P_i = k, \theta^{(t)} \right) \\ = \mathbb{P}\left( M_i = m_i | P_i = k, \theta^{(t)} \right) \mathbb{P}\left(G_i = g_i | P_i = k, \theta^{(t)} \right) \textrm{ (because $G_i \indep M_i | P_i$)} \\ = f_m\left(m_i; \left[ \eta^m_i(k) \right]^{(t)}\right) f_g\left(g_i; \left[ \eta^g_i(k) \right]^{(t)} \right).
 		\end{multline*}
@@
 		\begin{cases}
-\pi^{(t)}(1) := \pi^{(t)} = \P\left(P_i = 1 | \theta^{(t)} \right) \\
-\pi^{(t)}(0) := 1 - \pi^{(t)} = \P\left(P_i = 0 | \theta^{(t)} \right).
+		\pi^{(t)}(1) := \pi^{(t)} = \mathbb{P}\left(P_i = 1 | \theta^{(t)} \right) \\
+		\pi^{(t)}(0) := 1 - \pi^{(t)} = \mathbb{P}\left(P_i = 0 | \theta^{(t)} \right).
 		\end{cases}
@@
 		\begin{multline}\label{e_step_1}
-T^{(t)}_i(1) = \P(P_i = 1 | M_i = m_i, G_i = g_i, \theta^{(t)})  = \frac{\pi^{(t)}(1) \alpha^{(t)}_i(1)}{ \sum_{k=0}^1 \pi^{(t)}(k) \alpha^{(t)}_i(k)} \textrm{ (by Bayes rule)} \\ = \frac{1}{\frac{ \pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)} + 1} = \frac{1}{ \exp\left(\log\left(\frac{\pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)}\right)\right) + 1} = \frac{ 1 }{ \exp\left(q^{(t)}_i\right) + 1},
+		T^{(t)}_i(1) = \mathbb{P}(P_i = 1 | M_i = m_i, G_i = g_i, \theta^{(t)})  = \frac{\pi^{(t)}(1) \alpha^{(t)}_i(1)}{ \sum_{k=0}^1 \pi^{(t)}(k) \alpha^{(t)}_i(k)} \textrm{ (by Bayes rule)} \\ = \frac{1}{\frac{ \pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)} + 1} = \frac{1}{ \exp\left(\log\left(\frac{\pi^{(t)}(0) \alpha_i(0)}{\pi^{(t)}(1) \alpha_i(1)}\right)\right) + 1} = \frac{ 1 }{ \exp\left(q^{(t)}_i\right) + 1},
 		\end{multline}
@@
 		\end{equation}
-Define $Q(\theta | \theta^{(t)}) = \E_{\left(P |M = m, G = g, \theta^{(t)}\right)}\left[ \mathcal{L}(\theta; m, g, p) \right].$ We have that
+		Define $Q(\theta | \theta^{(t)}) = \mathbb{E}_{\left(P |M = m, G = g, \theta^{(t)}\right)}\left[ \mathcal{L}(\theta; m, g, p) \right].$ We have that
 		\begin{multline}\label{Q_funct}
@@
 		\begin{multline*}
-\frac{\sum_{i=1}^n T_i^{(t)}(1)}{\pi} - \frac{ \sum_{i=1}^n T_i^{(t)}(0) }{ 1 - \pi} = 0 \iff \sum_{i=1}^n T_i^{(t)}(1) - \pi \sum_{i=1}^n T^{(t)}_i(1) = \pi \sum_{i=1}^n T_i(0) \\ \iff \sum_{i=1}^n T^{(t)}_i(1) - \pi\sum_{i=1}^n T_i^{(t)}(1) = \pi n - \pi\sum_{i=1}^n T_i(1) \iff \pi = \frac{ \sum_{i=1}^n T_i^{(t)} (1) }{n}.\end{multline*}
+		\frac{\sum_{i=1}^n T_i^{(t)}(1)}{\pi} - \frac{ \sum_{i=1}^n T_i^{(t)}(0) }{ 1 - \pi} = 0 \iff \sum_{i=1}^n T_i^{(t)}(1) - \pi \sum_{i=1}^n T^{(t)}_i(1) = \pi \sum_{i=1}^n T_i(0) \\ \iff \sum_{i=1}^n T^{(t)}_i(1) - \pi\sum_{i=1}^n T_i^{(t)}(1) = \pi n - \pi\sum_{i=1}^n T_i(1) \iff \pi = \frac{ \sum_{i=1}^n T_i^{(t)} (1) }{n}.
+	\end{multline*}
+
 	Thus, the maximizer $\pi^{(t+1)}$ of (\ref{Q_funct}) in $\pi$ is $\pi^{(t+1)} = (1/n)\sum_{i=1}^n T^{(t)}_i(1)$.
-Next, define $w^{(t)} = [T^{(t)}_1(0), \dots, T^{(t)}_n(0), T^{(t)}_1(1), \dots, T^{(t)}_n(1)]^T \in \R^{2n}$.
+Next, define $w^{(t)} = [T^{(t)}_1(0), \dots, T^{(t)}_n(0), T^{(t)}_1(1), \dots, T^{(t)}_n(1)]^T \in \mathbb{R}^{2n}$.
 We can view the second term of (\ref{Q_funct}) as the log-likelihood of a GLM -- call it $\textrm{GLM}^{(t)}_m$ -- that has exponential family density $f_m$, link function $r_m$, responses $[m,m]^T$, offsets $[o^m, o^m]^T$, weights $w^{(t)}$, and design matrix $\begin{bmatrix} \tilde{X}(0) \\ \tilde{X}(1) \end{bmatrix}.$ Therefore, the maximizer $\beta^{(t+1)}_m$ of the second term of (\ref{Q_funct}) is the maximizer of $\textrm{GLM}^{(t)}_m$, which we can compute using the iteratively reweighted least squares (IRLS) procedure, as implemented in R's GLM function.
@@
 For $i \in \{1, \dots, n\}$, $j \in \{0, 1\}$, and $\theta = (\pi, \beta_m, \beta_g),$ let $T^\theta_i(j)$ be defined by
-$$T^\theta_i(j) = \P_\theta\left(P_i = j | M_i = m_i, G_i = g_i\right).$$ Let the $n \times n$ matrix $T^\theta(j)$ be given by $$T^\theta(j) = \textrm{diag}\left\{T^\theta_1(j), \dots, T^\theta_n(j)\right\}.$$
+		$$T^\theta_i(j) = \mathbb{P}_\theta\left(P_i = j | M_i = m_i, G_i = g_i\right).$$ Let the $n \times n$ matrix $T^\theta(j)$ be given by $$T^\theta(j) = \textrm{diag}\left\{T^\theta_1(j), \dots, T^\theta_n(j)\right\}.$$
 		Next, define the diagonal $n \times n$ matrices $\Delta^m$, $[\Delta']^m$, $V^m$, and $H^m$ by
@@
 		$$
-Define the matrices $\Delta^g(j)$, $[\Delta']^{g}(j)$, $V^g(j),$ and $H^g(j)$ analogously. Finally, define the vectors $s^m(j)$ and $w^m(j)$ in $\R^n$ by 
+		Define the matrices $\Delta^g(j)$, $[\Delta']^{g}(j)$, $V^g(j),$ and $H^g(j)$ analogously. Finally, define the vectors $s^m(j)$ and $w^m(j)$ in $\mathbb{E}^n$ by 
 		$$ \begin{cases}
@@
 		\begin{multline}\label{zero_inf_info_mat}
-J(\theta; m, g) = -\E \left[\nabla^2 \mathcal{L}(\theta; m, g, p) | G = g, M = m \right] \\ + \E\left[\nabla \mathcal{L}(\theta; m, g, p) | G = g, M = m \right] \E\left[\nabla \mathcal{L}(\theta; m, g, p) | G = g, M = m \right]^T \\ - \E\left[ \nabla\mathcal{L}(\theta; m, g, p) \nabla \mathcal{L}(\theta; m, g, p)^T | G = g, M = m \right].
+		J(\theta; m, g) = -\mathbb{E} \left[\nabla^2 \mathcal{L}(\theta; m, g, p) | G = g, M = m \right] \\ + \mathbb{E}\left[\nabla \mathcal{L}(\theta; m, g, p) | G = g, M = m \right] \mathbb{E}\left[\nabla \mathcal{L}(\theta; m, g, p) | G = g, M = m \right]^T \\ - \mathbb{E}\left[ \nabla\mathcal{L}(\theta; m, g, p) \nabla \mathcal{L}(\theta; m, g, p)^T | G = g, M = m \right].
 		\end{multline}
@@
 		\begin{equation}\label{sub_mat_pi}
-J_{\pi}(\theta; m, g) = -\E\left[\nabla^2_\pi \mathcal{L}(\theta; m, g, p) \right] + \left(\E\left[ \nabla_\pi \mathcal{L}(\theta; m, g, p) \right] \right)^2 - \E\left[(\nabla_\pi \mathcal{L}(\theta; m, g, p))^2 \right].
+		J_{\pi}(\theta; m, g) = -\mathbb{E}\left[\nabla^2_\pi \mathcal{L}(\theta; m, g, p) \right] + \left(\mathbb{E}\left[ \nabla_\pi \mathcal{L}(\theta; m, g, p) \right] \right)^2 - \mathbb{E}\left[(\nabla_\pi \mathcal{L}(\theta; m, g, p))^2 \right].
 		\end{equation}
@@
 		\begin{multline}\label{submat_pi_1}
-\E \left[ -\nabla^2_{\pi} \mathcal{L}(\theta; m, g, p)\right] = - \E\left[\frac{ ( \sum_{i=1}^n p_i) - n}{(1 - \pi)^2} - \frac{\sum_{i=1}^n p_i}{\pi^2} \right] \\ = - \E\left\{\left[\frac{1}{(1-\pi)^2} - \frac{1}{\pi^2} \right] \sum_{i=1}^n p_i - \frac{n}{ (1 - \pi)^2 } \right\} \\ = - \left\{\left[ \frac{1}{(1-\pi)^2} - \frac{1}{\pi^2} \right] \sum_{i=1}^n T^\theta_i(1) - \frac{n}{ (1 - \pi)^2}  \right\} \\ = \left[ \frac{1}{\pi^2} - \frac{1}{(1 - \pi)^2} \right] \sum_{i=1}^n T^\theta_i(1) + \frac{n}{(1-\pi)^2}.
+		\mathbb{E} \left[ -\nabla^2_{\pi} \mathcal{L}(\theta; m, g, p)\right] = - \mathbb{E}\left[\frac{ ( \sum_{i=1}^n p_i) - n}{(1 - \pi)^2} - \frac{\sum_{i=1}^n p_i}{\pi^2} \right] \\ = - \mathbb{E}\left\{\left[\frac{1}{(1-\pi)^2} - \frac{1}{\pi^2} \right] \sum_{i=1}^n p_i - \frac{n}{ (1 - \pi)^2 } \right\} \\ = - \left\{\left[ \frac{1}{(1-\pi)^2} - \frac{1}{\pi^2} \right] \sum_{i=1}^n T^\theta_i(1) - \frac{n}{ (1 - \pi)^2}  \right\} \\ = \left[ \frac{1}{\pi^2} - \frac{1}{(1 - \pi)^2} \right] \sum_{i=1}^n T^\theta_i(1) + \frac{n}{(1-\pi)^2}.
 		\end{multline}
@@
 		\begin{multline*}
-\E \left[\nabla_\pi \mathcal{L}(\theta; m, g, p)^2 \right] \\ = \E \left[ \left( a \sum_{i=1}^n p_i - b\right)^2 \right]  =  \E \left[ a^2 \left( \sum_{i=1}^n p_i \right)^2 - 2ab \sum_{i=1}^n p_i + b^2 \right] \\ = a^2 \sum_{i=1}^n \sum_{j=1}^n \E[p_i p_j] -2ab \sum_{i=1}^n \E [p_i] + b^2.
+		\mathbb{E} \left[\nabla_\pi \mathcal{L}(\theta; m, g, p)^2 \right] \\ = \mathbb{E} \left[ \left( a \sum_{i=1}^n p_i - b\right)^2 \right]  =  \mathbb{E} \left[ a^2 \left( \sum_{i=1}^n p_i \right)^2 - 2ab \sum_{i=1}^n p_i + b^2 \right] \\ = a^2 \sum_{i=1}^n \sum_{j=1}^n \mathbb{E}[p_i p_j] -2ab \sum_{i=1}^n \mathbb{E} [p_i] + b^2.
 		\end{multline*}
@@
 		\begin{equation*}
-\left( \E \left[\nabla_\pi \mathcal{L}(\theta; m, g, x) \right] \right)^2 = \left( a \sum_{i=1}^n \E [p_i] - b \right)^2 = a^2 \sum_{i=1}^n \sum_{j=1}^n \E[p_i]  E[p_j] - 2ab \sum_{i=1}^n \E[p_i] + b^2.
+		\left( \mathbb{E} \left[\nabla_\pi \mathcal{L}(\theta; m, g, x) \right] \right)^2 = \left( a \sum_{i=1}^n \mathbb{E} [p_i] - b \right)^2 = a^2 \sum_{i=1}^n \sum_{j=1}^n \mathbb{E}[p_i]  E[p_j] - 2ab \sum_{i=1}^n \mathbb{E}[p_i] + b^2.
 		\end{equation*}
@@
 		\begin{multline}\label{submat_pi_2}
-(\E [\nabla_\pi \mathcal{L}(\theta; m, g, p)])^2 - \E \left[\nabla_\pi \mathcal{L}(\theta; m, g, p)^2 \right] \\ = a^2 \sum_{i=1}^n \sum_{j=1}^n \E[p_i] \E[p_j] - a^2 \sum_{i=1}^n \sum_{j=1}^n \E[p_i p_j] = a^2 \left( \sum_{i=1}^n \E[p_i]^2 - \E[p_i^2]\right) \\ = a^2 \left( \sum_{i=1}^n [T^\theta_i(1)]^2 - T^\theta_i(1) \right) = \left( \frac{1}{(1 - \pi)} + \frac{1}{\pi} \right)^2 \left(\sum_{i=1}^n [T^\theta_i(1)]^2 - T^\theta_i(1) \right).
+		(\mathbb{E} [\nabla_\pi \mathcal{L}(\theta; m, g, p)])^2 - \mathbb{E} \left[\nabla_\pi \mathcal{L}(\theta; m, g, p)^2 \right] \\ = a^2 \sum_{i=1}^n \sum_{j=1}^n \mathbb{E}[p_i] \mathbb{E}[p_j] - a^2 \sum_{i=1}^n \sum_{j=1}^n \mathbb{E}[p_i p_j] = a^2 \left( \sum_{i=1}^n \mathbb{E}[p_i]^2 - \mathbb{E}[p_i^2]\right) \\ = a^2 \left( \sum_{i=1}^n [T^\theta_i(1)]^2 - T^\theta_i(1) \right) = \left( \frac{1}{(1 - \pi)} + \frac{1}{\pi} \right)^2 \left(\sum_{i=1}^n [T^\theta_i(1)]^2 - T^\theta_i(1) \right).
 		\end{multline}
@@
 		\begin{multline}\label{sub_mat_2}
-J_{\beta^m}(\theta; m, g) = -\E \left[\nabla_{\beta^m}^2 \mathcal{L}(\theta; m, g, p) \right] \\ + \E\left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] \E\left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T - \E\left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T  \right].
+		J_{\beta^m}(\theta; m, g) = -\mathbb{E} \left[\nabla_{\beta^m}^2 \mathcal{L}(\theta; m, g, p) \right] \\ + \mathbb{E}\left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] \mathbb{E}\left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T - \mathbb{E}\left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T  \right].
 		\end{multline}
@@
 		\begin{multline*}
-\left( \E\left[-\nabla_{\beta^m}^2 \mathcal{L}(\theta; m, g, p)\right]\right)[k,l] = \E \left\{\tilde{X}[,k]^T (\Delta^m V^m \Delta^m - [\Delta']^mH^m) \tilde{X}[,l] \right\} \\ = \sum_{i=1}^n \E \left\{ \tilde{x}_{i,k} (\Delta^m_{i} V^m_{i} \Delta^m_{i} - [\Delta']^m_{i} H^m_{i}) \tilde{x}_{i,l} \right\} \\ = \sum_{i=1}^n \tilde{x}_{i,k}(0) T_i^{\theta}(0) [{\Delta}^m_i(0)  {V}^m_i(0) {\Delta}^m_i(0) - [\Delta']^m_i(0) {H}^m_i(0)] \tilde{x}_{i,l}(0) \\ + \sum_{i=1}^n \tilde{x}_{i,k}(1) T_i^{\theta}(1) [ {\Delta}^m_i(1)  {V}^m_i(1) {\Delta}^m_i(1) - [{\Delta}']^m_i(1) {H}^m_i(1)] \tilde{x}_{i,l}(1) \\ = \sum_{s = 0}^1 \tilde{X}(s)[,k]^T {T}^{\theta}(s) \left[ {\Delta}^m(s) {V}^m(s) {\Delta}^m(s) - [{\Delta}']^m(s) {H}^m(s) \right] \tilde{X}(s)[,l].
+		\left( \mathbb{E}\left[-\nabla_{\beta^m}^2 \mathcal{L}(\theta; m, g, p)\right]\right)[k,l] = \mathbb{E} \left\{\tilde{X}[,k]^T (\Delta^m V^m \Delta^m - [\Delta']^mH^m) \tilde{X}[,l] \right\} \\ = \sum_{i=1}^n \mathbb{E} \left\{ \tilde{x}_{i,k} (\Delta^m_{i} V^m_{i} \Delta^m_{i} - [\Delta']^m_{i} H^m_{i}) \tilde{x}_{i,l} \right\} \\ = \sum_{i=1}^n \tilde{x}_{i,k}(0) T_i^{\theta}(0) [{\Delta}^m_i(0)  {V}^m_i(0) {\Delta}^m_i(0) - [\Delta']^m_i(0) {H}^m_i(0)] \tilde{x}_{i,l}(0) \\ + \sum_{i=1}^n \tilde{x}_{i,k}(1) T_i^{\theta}(1) [ {\Delta}^m_i(1)  {V}^m_i(1) {\Delta}^m_i(1) - [{\Delta}']^m_i(1) {H}^m_i(1)] \tilde{x}_{i,l}(1) \\ = \sum_{s = 0}^1 \tilde{X}(s)[,k]^T {T}^{\theta}(s) \left[ {\Delta}^m(s) {V}^m(s) {\Delta}^m(s) - [{\Delta}']^m(s) {H}^m(s) \right] \tilde{X}(s)[,l].
 		\end{multline*}
@@
 		\begin{multline}\label{sub_mat_2_1}
-\E\left[-\nabla_{\beta^m}^2 \mathcal{L}(\theta; m, g, p)\right] = \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) \left[ {\Delta}^m(s) {V}^m(s) {\Delta}^m(s) - [{\Delta}']^m(s) {H}^m(s) \right] \tilde{X}(s).
+		\mathbb{E}\left[-\nabla_{\beta^m}^2 \mathcal{L}(\theta; m, g, p)\right] = \sum_{s=0}^1 \tilde{X}(s)^T T^{\theta}(s) \left[ {\Delta}^m(s) {V}^m(s) {\Delta}^m(s) - [{\Delta}']^m(s) {H}^m(s) \right] \tilde{X}(s).
 		\end{multline}
@@
 		\begin{multline*}
-\bigg[ \E \left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] \E \left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T \\ - \E \left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T \right] \bigg] [k,l] \\
-= \left[ \E \left[\tilde{X}^T \Delta^m s^m \right] \E \left[\tilde{X}^T \Delta^m s^m \right]^T\right][k,l] - \E \left[\tilde{X}^T \Delta^m s^m (s^m)^T \Delta^m \tilde{X} \right][k,l] \\
-= \E\left[ \tilde{X}[,k]^T \Delta^m s^m \right] \E \left[ \tilde{X}[,l]^T \Delta^m s^m \right] - \E \left[ \tilde{X}[,k]^T \Delta^m s^m (s^m)^T \Delta^m \tilde{X}[,l ] \right] \\ 
-=\E\left(\sum_{i=1}^n \tilde{x}_{ik} \Delta^m_i s^m_{i} \right) \E \left( \sum_{j=1}^n \tilde{x}_{jl} \Delta^m_j s^m_j \right) - \E \left( \sum_{i=1}^n \sum_{j=1}^n \tilde{x}_{ik} \Delta^m_i s^m_i s^m_j \Delta^m_j \tilde{x}_{jl} \right) \\
-= \sum_{i=1}^n \sum_{j=1}^n \E[ \tilde{x}_{ik} \Delta^m_is^m_i] \E [\tilde{x}_{jl} \Delta^m_j s^m_j]  -  \sum_{i=1}^n \sum_{j=1}^n \E [ \tilde{x}_{ik} \Delta^m_i s^m_i s^m_j \Delta^m_j \tilde{x}_{jl}] \\
-= \sum_{i=1}^n \sum_{j=1}^n \E[ \tilde{x}_{ik} \Delta^m_i s^m_i] \E \left[\tilde{x}_{jl} \Delta^m_j s^m_j \right]  - \sum_{i \neq j} \E [ \tilde{x}_{ik} \Delta^m_i s^m_i] \E[s^m_j \Delta^m_j \tilde{x}_{jl}] \\ - \sum_{i=1}^n \E[\tilde{x}_{ik} \Delta^m_i s^m_i s^m_i \Delta^m_i \tilde{x}_{il}] \\
-= \sum_{i=1}^n \E[\tilde{x}_{ik} \Delta^m_i s^m_i ] \E [ \tilde{x}_{il} \Delta^m_i s^m_i] - \sum_{i=1}^n \E[\tilde{x}_{ik} (\Delta_i^m)^2 (H_i^m)^2 \tilde{x}_{il}] \\ = \sum_{i=1}^n \left[\tilde{x}_{ik}(0) {\Delta}^m_i(0) T_i^{\theta}(0) {H}^m_i(0) + \tilde{x}_{ik}(1) {\Delta}^m_i(1) T_i^{\theta}(1) {H}^m_i(1) \right] \\ \cdot \left[ \tilde{x}_{il}(0) {\Delta}^m_i(0) T_i^{\theta}(0) {H}^m_i(0) + \tilde{x}_{il}(1) {\Delta}^m_i(1) T_i^{\theta}(1) {H}^m_i(1) \right] \\ - \sum_{i=1}^n \left[ \tilde{x}_{ik}(0) T_i^{\theta}(0) (\Delta_i^m(0))^2 ({H}_i^m(0))^2 \tilde{x}_{il}(0)  + \tilde{x}_{ik}(1) T_i^{\theta}(1) ({\Delta}^m_i(1))^2 ({H}_i^m(1))^m \tilde{x}_{il}(1) \right] \\ = \sum_{s=0}^1 \sum_{t=0}^1 \left[ \sum_{i=1}^n \tilde{x}_{ik}(s) T^{\theta}_i(s) {\Delta}^m_i(s) {H}^m_i(t) T_i^{\theta}(t){\Delta}^m_i(t) {H}^m_i(t) \tilde{x}_{il}(t) \right] \\ - \sum_{s=0}^1 \left[\sum_{i=1}^n \tilde{x}_{ik}(s) T_i^\theta(s) ({\Delta}^m_i(s))^2 ({H}_i^m(s))^2 \tilde{x}_{il}(s) \right] \\ = \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)[,k]^T {T}^\theta(s) {\Delta}^m(s) {H}^m(s) {T}^\theta(t) {\Delta}^m(t) {H}^m(t) \tilde{X}(k)[,l] \\ - \sum_{s=0}^1{X}(s)[,k]^T {T}^\theta(s) ({\Delta}^m(s))^2 ({H}^m(s))^2 \tilde{X}(s)[,l].
+		\bigg[ \mathbb{E} \left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] \mathbb{E} \left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T \\ - \mathbb{E} \left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T \right] \bigg] [k,l] \\
+		= \left[ \mathbb{E} \left[\tilde{X}^T \Delta^m s^m \right] \mathbb{E} \left[\tilde{X}^T \Delta^m s^m \right]^T\right][k,l] - \mathbb{E} \left[\tilde{X}^T \Delta^m s^m (s^m)^T \Delta^m \tilde{X} \right][k,l] \\
+		= \mathbb{E}\left[ \tilde{X}[,k]^T \Delta^m s^m \right] \mathbb{E} \left[ \tilde{X}[,l]^T \Delta^m s^m \right] - \mathbb{E} \left[ \tilde{X}[,k]^T \Delta^m s^m (s^m)^T \Delta^m \tilde{X}[,l ] \right] \\
+		=\mathbb{E}\left(\sum_{i=1}^n \tilde{x}_{ik} \Delta^m_i s^m_{i} \right) \mathbb{E} \left( \sum_{j=1}^n \tilde{x}_{jl} \Delta^m_j s^m_j \right) - \mathbb{E} \left( \sum_{i=1}^n \sum_{j=1}^n \tilde{x}_{ik} \Delta^m_i s^m_i s^m_j \Delta^m_j \tilde{x}_{jl} \right) \\
+		= \sum_{i=1}^n \sum_{j=1}^n \mathbb{E}[ \tilde{x}_{ik} \Delta^m_is^m_i] \mathbb{E} [\tilde{x}_{jl} \Delta^m_j s^m_j]  -  \sum_{i=1}^n \sum_{j=1}^n \mathbb{E} [ \tilde{x}_{ik} \Delta^m_i s^m_i s^m_j \Delta^m_j \tilde{x}_{jl}] \\
+		= \sum_{i=1}^n \sum_{j=1}^n \mathbb{E}[ \tilde{x}_{ik} \Delta^m_i s^m_i] \mathbb{E} \left[\tilde{x}_{jl} \Delta^m_j s^m_j \right]  - \sum_{i \neq j} \mathbb{E} [ \tilde{x}_{ik} \Delta^m_i s^m_i] \mathbb{E}[s^m_j \Delta^m_j \tilde{x}_{jl}] \\ - \sum_{i=1}^n \mathbb{E}[\tilde{x}_{ik} \Delta^m_i s^m_i s^m_i \Delta^m_i \tilde{x}_{il}] \\
+		= \sum_{i=1}^n \mathbb{E}[\tilde{x}_{ik} \Delta^m_i s^m_i ] \mathbb{E} [ \tilde{x}_{il} \Delta^m_i s^m_i] - \sum_{i=1}^n \mathbb{E}[\tilde{x}_{ik} (\Delta_i^m)^2 (H_i^m)^2 \tilde{x}_{il}] \\ = \sum_{i=1}^n \left[\tilde{x}_{ik}(0) {\Delta}^m_i(0) T_i^{\theta}(0) {H}^m_i(0) + \tilde{x}_{ik}(1) {\Delta}^m_i(1) T_i^{\theta}(1) {H}^m_i(1) \right] \\ \cdot \left[ \tilde{x}_{il}(0) {\Delta}^m_i(0) T_i^{\theta}(0) {H}^m_i(0) + \tilde{x}_{il}(1) {\Delta}^m_i(1) T_i^{\theta}(1) {H}^m_i(1) \right] \\ - \sum_{i=1}^n \left[ \tilde{x}_{ik}(0) T_i^{\theta}(0) (\Delta_i^m(0))^2 ({H}_i^m(0))^2 \tilde{x}_{il}(0)  + \tilde{x}_{ik}(1) T_i^{\theta}(1) ({\Delta}^m_i(1))^2 ({H}_i^m(1))^m \tilde{x}_{il}(1) \right] \\ = \sum_{s=0}^1 \sum_{t=0}^1 \left[ \sum_{i=1}^n \tilde{x}_{ik}(s) T^{\theta}_i(s) {\Delta}^m_i(s) {H}^m_i(t) T_i^{\theta}(t){\Delta}^m_i(t) {H}^m_i(t) \tilde{x}_{il}(t) \right] \\ - \sum_{s=0}^1 \left[\sum_{i=1}^n \tilde{x}_{ik}(s) T_i^\theta(s) ({\Delta}^m_i(s))^2 ({H}_i^m(s))^2 \tilde{x}_{il}(s) \right] \\ = \sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)[,k]^T {T}^\theta(s) {\Delta}^m(s) {H}^m(s) {T}^\theta(t) {\Delta}^m(t) {H}^m(t) \tilde{X}(k)[,l] \\ - \sum_{s=0}^1{X}(s)[,k]^T {T}^\theta(s) ({\Delta}^m(s))^2 ({H}^m(s))^2 \tilde{X}(s)[,l].
 		\end{multline*}
@@
 		\begin{multline}\label{sub_mat_2_2}
- \E \left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] \E \left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T - \E \left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T \right] \\ =
+		\mathbb{E} \left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] \mathbb{E} \left[ \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T - \mathbb{E} \left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T \right] \\ =
 		\sum_{s=0}^1 \sum_{t=0}^1 \tilde{X}(s)^T {T}^\theta(s) {\Delta}^m(s) {H}^m(s) T^{\theta}(t) {\Delta}^m(t) {H}^m(t) \tilde{X}(t) \\ - \sum_{s=0}^1 \tilde{X}(s)^T {T}^\theta(s) ({\Delta}^m(s))^2 ({H}^m(s))^2 \tilde{X}(s).
@@
 		\begin{multline}\label{sub_mat_4}
-J_{(\beta^g, \beta^m)}(\theta; m,g) = \E \left[-\nabla_{\beta^g} \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] \\ + \E\left[ \nabla_{\beta^{g}}\mathcal{L}(\theta ; m,g,p) \right] \E \left[\nabla_{\beta^m}\mathcal{L} (\theta ; m,g,p)  \right]^T - \E \left[ \nabla_{\beta^{g}}\mathcal{L} (\theta; m,g,p) \nabla_{\beta^m}\mathcal{L}(\theta; m,g,p)^T  \right].
+		J_{(\beta^g, \beta^m)}(\theta; m,g) = \mathbb{E} \left[-\nabla_{\beta^g} \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] \\ + \mathbb{E}\left[ \nabla_{\beta^{g}}\mathcal{L}(\theta ; m,g,p) \right] \mathbb{E} \left[\nabla_{\beta^m}\mathcal{L} (\theta ; m,g,p)  \right]^T - \mathbb{E} \left[ \nabla_{\beta^{g}}\mathcal{L} (\theta; m,g,p) \nabla_{\beta^m}\mathcal{L}(\theta; m,g,p)^T  \right].
 		\end{multline}
@@
 		\begin{equation}\label{sub_mat_4_1}
-\E\left[-\nabla_{\beta^g} \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] = 0,
+		\mathbb{E}\left[-\nabla_{\beta^g} \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right] = 0,
 		\end{equation}
@@
 		\begin{multline}\label{sub_mat_4_2}
-\bigg[ \E \left[\nabla_{\beta^g} \mathcal{L}(\theta; m, g, p) \right] \E \left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T \\ - \E \left[ \nabla_{\beta^g} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T \right] \bigg][k,l] \\
-= \left[ \E \left[ \tilde{X}^T \Delta^g s^g \right]\E \left[ \tilde{X}^T \Delta^m s^m \right]^T\right][k,l] - \E \left[ \tilde{X}^T \Delta^g s^g (s^m)^T \Delta^m \tilde{X} \right][k,l] \\ 
-= \E\left[\tilde{X}[,k]^T \Delta^g s^g \right] \E \left[\tilde{X}[,l]^T \Delta^m s^m \right] - \E \left[\tilde{X}[,k]^T \Delta^g s^g (s^m)^T \Delta^m \tilde{X}[,l ] \right] \\
-=\E\left( \sum_{i=1}^n \tilde{x}_{ik} \Delta^g_i s^g_i \right) \E \left( \sum_{j=1}^n \tilde{x}_{jl} \Delta^m_j s^m_j \right) - \E \left( \sum_{i=1}^n \sum_{j=1}^n \tilde{x}_{ik} \Delta^g_i s^g_i s^m_j \Delta^m_j \tilde{x}_{jl} \right) \\
-= \sum_{i=1}^n \sum_{j=1}^n \E[\tilde{x}_{ik} \Delta^g_is^g_i] \E[ \tilde{x}_{jl} \Delta^m_j s^m_j ] - \sum_{i=1}^n \sum_{j=1}^n \E[ \tilde{x}_{ik} \Delta^g_i s^g_i s^m_j \Delta^m_j \tilde{x}_{jl}]  \\
-= \sum_{i=1}^n \sum_{j=1}^n \E[ \tilde{x}_{ik} \Delta^g_i s^g_i] \E \left[\tilde{x}_{jl} \Delta^m_j s^m_j \right]  - \sum_{i \neq j} \E[\tilde{x}_{ik} \Delta^g_i s^g_i] \E[\tilde{x}_{jl}\Delta^m_j  s^m_j ] \\ - \sum_{i=1}^n \E[\tilde{x}_{ik} \Delta^g_i s^g_i s^m_i \Delta^m_i \tilde{x}_{il}] \\
-= \sum_{i=1}^n \E[\tilde{x}_{ik} \Delta^g_i H^g_i] \E[\tilde{x}_{il} \Delta_i^m H^m_i] - \sum_{i=1}^n \E[\tilde{x}_{ik} H_i^g \Delta_i^g \Delta_i^m H_i^m \tilde{x}_{il}] \\ 
+		\bigg[ \mathbb{E} \left[\nabla_{\beta^g} \mathcal{L}(\theta; m, g, p) \right] \mathbb{E} \left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p) \right]^T \\ - \mathbb{E} \left[ \nabla_{\beta^g} \mathcal{L}(\theta; m, g, p) \nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)^T \right] \bigg][k,l] \\
+		= \left[ \mathbb{E} \left[ \tilde{X}^T \Delta^g s^g \right]\mathbb{E} \left[ \tilde{X}^T \Delta^m s^m \right]^T\right][k,l] - \mathbb{E} \left[ \tilde{X}^T \Delta^g s^g (s^m)^T \Delta^m \tilde{X} \right][k,l] \\
+		= \mathbb{E}\left[\tilde{X}[,k]^T \Delta^g s^g \right] \mathbb{E} \left[\tilde{X}[,l]^T \Delta^m s^m \right] - \mathbb{E} \left[\tilde{X}[,k]^T \Delta^g s^g (s^m)^T \Delta^m \tilde{X}[,l ] \right] \\
+		=\mathbb{E}\left( \sum_{i=1}^n \tilde{x}_{ik} \Delta^g_i s^g_i \right) \mathbb{E} \left( \sum_{j=1}^n \tilde{x}_{jl} \Delta^m_j s^m_j \right) - \mathbb{E} \left( \sum_{i=1}^n \sum_{j=1}^n \tilde{x}_{ik} \Delta^g_i s^g_i s^m_j \Delta^m_j \tilde{x}_{jl} \right) \\
+		= \sum_{i=1}^n \sum_{j=1}^n \mathbb{E}[\tilde{x}_{ik} \Delta^g_is^g_i] \mathbb{E}[ \tilde{x}_{jl} \Delta^m_j s^m_j ] - \sum_{i=1}^n \sum_{j=1}^n \mathbb{E}[ \tilde{x}_{ik} \Delta^g_i s^g_i s^m_j \Delta^m_j \tilde{x}_{jl}]  \\
+		= \sum_{i=1}^n \sum_{j=1}^n \mathbb{E}[ \tilde{x}_{ik} \Delta^g_i s^g_i] \mathbb{E} \left[\tilde{x}_{jl} \Delta^m_j s^m_j \right]  - \sum_{i \neq j} \mathbb{E}[\tilde{x}_{ik} \Delta^g_i s^g_i] \mathbb{E}[\tilde{x}_{jl}\Delta^m_j  s^m_j ] \\ - \sum_{i=1}^n \mathbb{E}[\tilde{x}_{ik} \Delta^g_i s^g_i s^m_i \Delta^m_i \tilde{x}_{il}] \\
+		= \sum_{i=1}^n \mathbb{E}[\tilde{x}_{ik} \Delta^g_i H^g_i] \mathbb{E}[\tilde{x}_{il} \Delta_i^m H^m_i] - \sum_{i=1}^n \mathbb{E}[\tilde{x}_{ik} H_i^g \Delta_i^g \Delta_i^m H_i^m \tilde{x}_{il}] \\ 
 		= \sum_{i=1}^n \left[\tilde{x}_{ik}(0) {\Delta}^g_i(0) T^\theta_i(0) {H}^g_i(0) + \tilde{x}_{ik}(1) {\Delta}^g_i(1) T^\theta_i(1) {H}^g_i(1)\right] \\ 
@@
 		\begin{multline}\label{sub_mat_5}
-J_{(\beta^m,\pi)}(\theta; m, g) = \E \left[ - \nabla_{\beta^m} \nabla_{ \pi } \mathcal{L}(\theta; m, g, p) \right] \\ + \E\left[ \nabla_{\beta^m}\mathcal{L}(\theta ; m,g,p) \right] \E \left[ \nabla_{\pi}\mathcal{L}(\theta ; m,g,p) \right]^T  - \E \left[ \nabla_{\beta^m}\mathcal{L}(\theta; m,g,p) \nabla_{\pi}\mathcal{L}(\theta; m,g,p)^T \right].
+		J_{(\beta^m,\pi)}(\theta; m, g) = \mathbb{E} \left[ - \nabla_{\beta^m} \nabla_{ \pi } \mathcal{L}(\theta; m, g, p) \right] \\ + \mathbb{E}\left[ \nabla_{\beta^m}\mathcal{L}(\theta ; m,g,p) \right] \mathbb{E} \left[ \nabla_{\pi}\mathcal{L}(\theta ; m,g,p) \right]^T  - \mathbb{E} \left[ \nabla_{\beta^m}\mathcal{L}(\theta; m,g,p) \nabla_{\pi}\mathcal{L}(\theta; m,g,p)^T \right].
 		\end{multline}
@@
 		\begin{equation}\label{sub_mat_5_1}
-\E \left[ - \nabla_{\beta^m} \nabla_{ \pi } \mathcal{L}(\theta; m, g, p) \right] = 0,
+		\mathbb{E} \left[ - \nabla_{\beta^m} \nabla_{ \pi } \mathcal{L}(\theta; m, g, p) \right] = 0,
 		\end{equation}
@@
 		\begin{multline}\label{sub_mat_5_2}
-\E \left[\nabla_\pi \mathcal{L}(\theta; m, g, p) \right] \E\left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)[k] \right] - \E \left[\nabla_{\pi}\mathcal{L}(\theta; m,g,p) \nabla_{\beta^m}\mathcal{L}(\theta; m,g,p)[k] \right] \\= \left(\E \left[ a \sum_{i=1}^n p_i - b \right] \right) \left(\E\left[ \tilde{X}[,k]^T \Delta^m s^m \right] \right) - \E \left[ \left( a \sum_{i=1}^n p_i - b \right) \tilde{X}[,k]^T \Delta^m s^m \right] \\ = \left( a \sum_{i=1}^n \E[p_i] - b \right) \left( \sum_{j=1}^n \E [ \tilde{x}_{jk}\Delta^m_js^m_j] \right) - \E \left[ \left( a \sum_{i=1}^n p_i - b \right) \left( \sum_{j=1}^n \tilde{x}_{jk} \Delta^m_j s^m_j \right) \right] \\ = a \sum_{i=1}^n \sum_{j=1}^n \E [p_i] \E[ \tilde{x}_{jk} \Delta^m_j s^m_j] - b \sum_{j=1}^n \E[\tilde{x}_{jk} \Delta^m_j s^m_j] \\ - \left[ a \sum_{i=1}^n \sum_{j=1}^n \E [ p_i \tilde{x}_{jk} \Delta^m_j s^m_j] - b \sum_{j=1}^n \E[\tilde{x}_{jk} \Delta^m_j s^m_j] \right] \\ =  a \sum_{i=1}^n \sum_{j=1}^n \E[p_i] \E[\tilde{x}_{jk} \Delta^m_j s^m_j] - a\sum_{i \neq j} \E[p_i] \E[\tilde{x}_{jk} \Delta^m_j s^m_j] - a\sum_{i=1}^n \E[ p_i \tilde{x}_{ik} \Delta^m_i s^m_i ] \\ = a \sum_{i=1}^n \E[p_i] \E[ \tilde{x}_{ik} \Delta^m_i s^m_i] - a \sum_{i=1}^n \E[p_i \tilde{x}_{ik} \Delta^m_i s^m_i] \\ = a \sum_{i=1}^n T^\theta_i(1) [T^\theta_i(0) \Delta^m_i(0) s^m_i(0) \tilde{x}_{ik}(0) + T^\theta_i(1) \Delta^m_i(1) s^m_i(1) \tilde{x}_{ik}(1)] \\ - a \sum_{i=1}^n T^\theta_i(1)\Delta^m_i(1)s^m_i(1)\tilde{x}_{ik}(1) \\ = a \sum_{i=1}^n T^\theta_i(0)T^\theta_i(1) \Delta_i^m(0)H^m_i(0)\tilde{x}_{ik}(0) \\ + a \sum_{i=1}^n \left( [T^\theta_i(1)]^2 \Delta^m_i(1)H^m_i(1) - T^\theta_i(1)\Delta^m_i(1)H^m_i(1) \right) \tilde{x}_{ik}(1)  \\ =a \left[ \sum_{i=1}^n T^\theta_i(0) T^\theta_i(1) \Delta^m_i(0) H^m_i(0) \tilde{x}_{ik}(0) + \sum_{i=1}^n T^\theta_i(1)\Delta^m_i(1)H^m_i(1)[T^\theta_i(1) - 1] \tilde{x}_{ik}(1) \right] \\ = a \left[ \sum_{i=1}^n T^\theta_i(0) T^\theta_i(1) \Delta^m_i(0) H^m_i(0) \tilde{x}_{ik}(0) - \sum_{i=1}^n T^\theta_i(0) T^\theta_i(1) \Delta^m_i(1) H^m_i(1) \tilde{x}_{ik}(1) \right] \\ = a\left(\tilde{X}(0)[,k]^T w^m(0) - \tilde{X}(1)[,k]^T w^m(1)  \right).
+		\mathbb{E} \left[\nabla_\pi \mathcal{L}(\theta; m, g, p) \right] \mathbb{E}\left[\nabla_{\beta^m} \mathcal{L}(\theta; m, g, p)[k] \right] - \mathbb{E} \left[\nabla_{\pi}\mathcal{L}(\theta; m,g,p) \nabla_{\beta^m}\mathcal{L}(\theta; m,g,p)[k] \right] \\= \left(\mathbb{E} \left[ a \sum_{i=1}^n p_i - b \right] \right) \left(\mathbb{E}\left[ \tilde{X}[,k]^T \Delta^m s^m \right] \right) - \mathbb{E} \left[ \left( a \sum_{i=1}^n p_i - b \right) \tilde{X}[,k]^T \Delta^m s^m \right] \\ = \left( a \sum_{i=1}^n \mathbb{E}[p_i] - b \right) \left( \sum_{j=1}^n \mathbb{E} [ \tilde{x}_{jk}\Delta^m_js^m_j] \right) - \mathbb{E} \left[ \left( a \sum_{i=1}^n p_i - b \right) \left( \sum_{j=1}^n \tilde{x}_{jk} \Delta^m_j s^m_j \right) \right] \\ = a \sum_{i=1}^n \sum_{j=1}^n \mathbb{E} [p_i] \mathbb{E}[ \tilde{x}_{jk} \Delta^m_j s^m_j] - b \sum_{j=1}^n \mathbb{E}[\tilde{x}_{jk} \Delta^m_j s^m_j] \\ - \left[ a \sum_{i=1}^n \sum_{j=1}^n \mathbb{E} [ p_i \tilde{x}_{jk} \Delta^m_j s^m_j] - b \sum_{j=1}^n \mathbb{E}[\tilde{x}_{jk} \Delta^m_j s^m_j] \right] \\ =  a \sum_{i=1}^n \sum_{j=1}^n \mathbb{E}[p_i] \mathbb{E}[\tilde{x}_{jk} \Delta^m_j s^m_j] - a\sum_{i \neq j} \mathbb{E}[p_i] \mathbb{E}[\tilde{x}_{jk} \Delta^m_j s^m_j] - a\sum_{i=1}^n \mathbb{E}[ p_i \tilde{x}_{ik} \Delta^m_i s^m_i ] \\ = a \sum_{i=1}^n \mathbb{E}[p_i] \mathbb{E}[ \tilde{x}_{ik} \Delta^m_i s^m_i] - a \sum_{i=1}^n \mathbb{E}[p_i \tilde{x}_{ik} \Delta^m_i s^m_i] \\ = a \sum_{i=1}^n T^\theta_i(1) [T^\theta_i(0) \Delta^m_i(0) s^m_i(0) \tilde{x}_{ik}(0) + T^\theta_i(1) \Delta^m_i(1) s^m_i(1) \tilde{x}_{ik}(1)] \\ - a \sum_{i=1}^n T^\theta_i(1)\Delta^m_i(1)s^m_i(1)\tilde{x}_{ik}(1) \\ = a \sum_{i=1}^n T^\theta_i(0)T^\theta_i(1) \Delta_i^m(0)H^m_i(0)\tilde{x}_{ik}(0) \\ + a \sum_{i=1}^n \left( [T^\theta_i(1)]^2 \Delta^m_i(1)H^m_i(1) - T^\theta_i(1)\Delta^m_i(1)H^m_i(1) \right) \tilde{x}_{ik}(1)  \\ =a \left[ \sum_{i=1}^n T^\theta_i(0) T^\theta_i(1) \Delta^m_i(0) H^m_i(0) \tilde{x}_{ik}(0) + \sum_{i=1}^n T^\theta_i(1)\Delta^m_i(1)H^m_i(1)[T^\theta_i(1) - 1] \tilde{x}_{ik}(1) \right] \\ = a \left[ \sum_{i=1}^n T^\theta_i(0) T^\theta_i(1) \Delta^m_i(0) H^m_i(0) \tilde{x}_{ik}(0) - \sum_{i=1}^n T^\theta_i(0) T^\theta_i(1) \Delta^m_i(1) H^m_i(1) \tilde{x}_{ik}(1) \right] \\ = a\left(\tilde{X}(0)[,k]^T w^m(0) - \tilde{X}(1)[,k]^T w^m(1)  \right).
 		\end{multline}
@@
 \texttt{skewness} returns the skewness $\gamma^m_i$ of the distribution as a function of the mean $\mu_i$, i.e.
-$$\texttt{skewness}\left(\mu_i\right) = \E \left[\left(\frac{m_i - \mu_i^m}{ \sigma_i^m}\right)^3\right] := \gamma_i^m.$$ Finally, the \texttt{mu.eta.prime} is the second derivative of the inverse link function $[r^{-1}_m]''.$ Algorithm \ref{algo:computing_info_matrices} computes the matrices $\Delta^m(j)$, $[\Delta']^m(j)$, $V^m(j)$, $H^m(j)$, and vector $s^m(j)$ for given $\beta_m$ and given family object.
+$$\texttt{skewness}\left(\mu_i\right) = \mathbb{E} \left[\left(\frac{m_i - \mu_i^m}{ \sigma_i^m}\right)^3\right] := \gamma_i^m.$$ Finally, the \texttt{mu.eta.prime} is the second derivative of the inverse link function $[r^{-1}_m]''.$ Algorithm \ref{algo:computing_info_matrices} computes the matrices $\Delta^m(j)$, $[\Delta']^m(j)$, $V^m(j)$, $H^m(j)$, and vector $s^m(j)$ for given $\beta_m$ and given family object.
 (The vector $w^m(j)$ can be computed in terms of $\Delta^m(j)$ and $H^m(j)$.) We use $\sigma^m_i(j)$ (resp.
@@
 		\end{equation}
- and, for all $t \in \R$, 
+		and, for all $t \in \mathbb{R}$, 
 		\begin{equation}\label{computing_info_matrix_2}
@@
 The zero-inflated GLM-EIV model is appropriate to use when the unperturbed cells do not transcribe \textit{any} gRNA molecules (i.e., when there are no background reads).
-Let $x_i = [1, z_i]^T \in \R^{d-1}$ be the vector of observed covariates, including an intercept term.
-($x_i$ is the same as $\tilde{x}_i$, but with the perturbation indicator $p_i$ removed.) Let $\beta_{g,z} = [\beta^g_0, \gamma_g] \in \R^{d-1}$ be an unknown coefficient vector.
+Let $x_i = [1, z_i]^T \in \mathbb{R}^{d-1}$ be the vector of observed covariates, including an intercept term.
+($x_i$ is the same as $\tilde{x}_i$, but with the perturbation indicator $p_i$ removed.) Let $\beta_{g,z} = [\beta^g_0, \gamma_g] \in \mathbb{R}^{d-1}$ be an unknown coefficient vector.
 ($\beta_{g,z}$ is the same as $\beta_g$, but with the perturbation effect $\beta^g_1$ removed).
@@
 Arguing in a manner similar to the background read case, we have that $$T^{(t)}_i(1) = \frac{1}{ \exp(q_i^{\left(t,z\right)}) + 1},$$ where
-$$q_i^{(t,z)} = \log \left(\frac{(1 - \pi^{(t)}) \P(M_i = m_i | P_i = 0, \theta^{(t)}) \P(G_i = g_i | P_i = 0, \theta^{(t)})}{(\pi^{(t)}) \P(M_i = m_i | P_i = 1, \theta^{(t)}) \P(G_i = g_i | P_i = 1, \theta^{(t)})}\right).$$
+		$$q_i^{(t,z)} = \log \left(\frac{(1 - \pi^{(t)}) \mathbb{P}(M_i = m_i | P_i = 0, \theta^{(t)}) \mathbb{P}(G_i = g_i | P_i = 0, \theta^{(t)})}{(\pi^{(t)}) \mathbb{P}(M_i = m_i | P_i = 1, \theta^{(t)}) \mathbb{P}(G_i = g_i | P_i = 1, \theta^{(t)})}\right).$$
 		The expression for $q^{(t,z)}_i$ is
@@
 		Next, we consider the M step of the EM algorithm, which is similar to the background read case but slightly different.
-Define $Q_z(\theta | \theta^{(t)}) = \E_{(P | M = m, G = g, \theta^{(t)})} \left[ \mathcal{L}_z(\theta; m, g, p) \right].$ We have that
+Define $Q_z(\theta | \theta^{(t)}) = \mathbb{E}_{(P | M = m, G = g, \theta^{(t)})} \left[ \mathcal{L}_z(\theta; m, g, p) \right].$ We have that
 		\begin{multline}\label{q_funct_zero_inf}
@@
 		Next, we derive the asymptotic observed information matrix for the zero-inflated model, allowing us to perform inference.
-Again, let $T^{\theta}(1) := \textrm{diag}\{T^{\theta}_1(1), \dots, T^{\theta}_n(1)\},$ but note that $T^\theta_i(1) = \P(P_i = 1 | G_i = g_i, M_i = m_i, \theta)$ is computed differently than in the background read case.
+Again, let $T^{\theta}(1) := \textrm{diag}\{T^{\theta}_1(1), \dots, T^{\theta}_n(1)\},$ but note that $T^\theta_i(1) = \mathbb{P}(P_i = 1 | G_i = g_i, M_i = m_i, \theta)$ is computed differently than in the background read case.
 Define the $n \times n$ matrices $\Delta^{(g,z)}, [\Delta']^{(g,z)}, V^{(g,z)},$ and $H^{(g,z)}$ by
@@
 		$$
-Also, define the $\R^n$ vectors $s^{(g,z)}$ and $w^{(g,z)}$ by
+		Also, define the $\mathbb{R}^n$ vectors $s^{(g,z)}$ and $w^{(g,z)}$ by
 		$$s^{(g,z)} = [g_1 - \mu_1^{g,z}, \dots, g_n - \mu_n^{g,z}]^T,$$ and $$w^{(g,z)} = [T^\theta_1(0)T^\theta_1(1) \Delta_1^{(g,z)}H^{(g,z)}_1,\dots, T^\theta_n(0)T^\theta_n(1) \Delta_n^{(g,z)}H^{(g,z)}_n].$$ These quantities are computable, as they do not depend on the unobserved variables $p_1, \dots, p_n$.
@@
 		\begin{multline*}
-J_z(\theta; m, g) = -\E \left[\nabla^2 \mathcal{L}_z(\theta; m, g, p) | G = g, M = m \right] \\ + \E\left[\nabla \mathcal{L}_z(\theta; m, g, p) | G = g, M = m \right] \E\left[\nabla \mathcal{L}_z(\theta; m, g, p) | G = g, M = m \right]^T \\ - \E\left[\nabla\mathcal{L}_z(\theta; m, g, p) \nabla \mathcal{L}_z(\theta; m, g, p)^T | G = g, M = m \right].
+		J_z(\theta; m, g) = -\mathbb{E} \left[\nabla^2 \mathcal{L}_z(\theta; m, g, p) | G = g, M = m \right] \\ + \mathbb{E}\left[\nabla \mathcal{L}_z(\theta; m, g, p) | G = g, M = m \right] \mathbb{E}\left[\nabla \mathcal{L}_z(\theta; m, g, p) | G = g, M = m \right]^T \\ - \mathbb{E}\left[\nabla\mathcal{L}_z(\theta; m, g, p) \nabla \mathcal{L}_z(\theta; m, g, p)^T | G = g, M = m \right].
 		\end{multline*}
@@
 		\begin{multline}\label{sub_mat_3_zeroinf}
-J_{\beta_{(g,z)}}(\theta; m, g) = -\E \left[\nabla_{\beta_{(g,z)}}^2 \mathcal{L}_z(\theta; m, g, p) \right] + \E\left[ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right] \E\left[\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right]^T \\ - \E\left[ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p)^T \right].
+		J_{\beta_{(g,z)}}(\theta; m, g) = -\mathbb{E} \left[\nabla_{\beta_{(g,z)}}^2 \mathcal{L}_z(\theta; m, g, p) \right] + \mathbb{E}\left[ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right] \mathbb{E}\left[\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right]^T \\ - \mathbb{E}\left[ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p)^T \right].
 		\end{multline}
@@
 		\begin{multline}\label{sub_mat_3_zeroinf_1}
--\E \left[\nabla_{\beta_{(g,z)}}^2 \mathcal{L}_z(\theta; m, g, p) \right]   = \E \left[ X^T P (\Delta^{(g,z)}V^{(g,z)}\Delta^{(g,z)} - (\Delta')^{(g,z)} H^{(g,z)} \right] \\ = X^T T^{\theta}(1) ( \Delta^{(g,z)}V^{(g,z)}\Delta^{(g,z)} - (\Delta')^{(g,z)} H^{(g,z)}) X.
+		-\mathbb{E} \left[\nabla_{\beta_{(g,z)}}^2 \mathcal{L}_z(\theta; m, g, p) \right]   = \mathbb{E} \left[ X^T P (\Delta^{(g,z)}V^{(g,z)}\Delta^{(g,z)} - (\Delta')^{(g,z)} H^{(g,z)} \right] \\ = X^T T^{\theta}(1) ( \Delta^{(g,z)}V^{(g,z)}\Delta^{(g,z)} - (\Delta')^{(g,z)} H^{(g,z)}) X.
 		\end{multline}
@@
 		\begin{multline*}
-\bigg[ \E \left[\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right] \E \left[\nabla_{\beta{(g,z)}} \mathcal{L}_z(\theta; m, g, p)\right]^T \\ - \E \left[ \nabla_{{\beta_{(g,z)}}} \mathcal{L}_z(\theta; m, g, p) \nabla_{{\beta_{(g,z)}}} \mathcal{L}_z(\theta; m, g, p)^T \right] \bigg] [k,l] \\
-= \left[ \E \left[ {X}^T P \Delta^{(g,z)} s^{(g,z)} \right]\E \left[ {X}^T P \Delta^{(g,z)} s^{(g,z)} \right]^T\right][k,l] - \E \left[ {X}^T P \Delta s^{(g,z)} (s^{(g,z)})^T \Delta^{(g,z)} P {X}^T \right][k,l] \\
-= \E\left[{X}[,k]^T P \Delta^{(g,z)} s^{(g,z)} \right] \E \left[ {X}[,l]^T P \Delta^{(g,z)} s^{(g,z)} \right] - \E \left[ {X}[,k]^T P \Delta^{(g,z)} s^{(g,z)} (s^{(g,z)})^T \Delta^{(g,z)} P {X}[,l ] \right] \\
-=\E\left(\sum_{i=1}^n {x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_{i} \right) \E \left( \sum_{j=1}^n {x}_{jl} P_j \Delta^{(g,z)}_j s^{(g,z)}_j \right) \\ - \E \left( \sum_{i=1}^n \sum_{j=1}^n {x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i s^{(g,z)}_j \Delta^{(g,z)}_j P_j{x}_{jl} \right) \\
-= \sum_{i=1}^n \sum_{j=1}^n \E[ {x}_{ik} P_i \Delta^{(g,z)}_is^{(g,z)}_i] \E [ {x}_{jl} P_j \Delta^{(g,z)}_j s^{(g,z)}_j]  -  \sum_{i=1}^n \sum_{j=1}^n \E [ {x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i s^{(g,z)}_j \Delta^{(g,z)}_jP_j {x}_{jl}  ]  \\
-= \sum_{i=1}^n \sum_{j=1}^n \E[ {x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i ] \E \left[{x}_{jl} P_j \Delta^{(g,z)}_j s^{(g,z)}_j \right]  - \sum_{i \neq j} \E [{x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i] \E[s^{(g,z)}_j P_j \Delta^{(g,z)}_j {x}_{jl}] \\ - \sum_{i=1}^n \E[ {x}_{ik}P_i \Delta^{(g,z)}_i s^{(g,z)}_i s^{(g,z)}_i \Delta^{(g,z)}_i P_i{x}_{il}] \\
-= \sum_{i=1}^n \E[ {x}_{ik} P_i \Delta^{(g,z)}_i H^{(g,z)}_i] \E[{x}_{il}P_i\Delta^{(g,z)}_i H^{(g,z)}_i] - \sum_{i=1}^n \E[{x}_{ik}P_i^2(\Delta^{(g,z)}_i)^2 (H^{(g,z)}_i)^2 {x}_{il}] \\ = \sum_{i=1}^n {x}_{ik} T_i^{\theta}(1)^2(\Delta^{(g,z)}_i)^2 (H_i^{(g,z)})^2{x}_{il} - \sum_{i=1}^n {x}_{ik} T_i^{\theta}(1) (\Delta_i^{(g,z)})^2 (H_i^{(g,z)})^2 {x}_{il} \\ = {X}[,k]^T T^{\theta}(1)^2 (\Delta^{(g,z)})^2 (H^{(g,z)})^2 {X}[,l] - {X}[,k]^T T^{\theta}(1) (\Delta^{(g,z)})^2 (H^{(g,z)})^2 {X}[,l]
+		\bigg[ \mathbb{E} \left[\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right] \mathbb{E} \left[\nabla_{\beta{(g,z)}} \mathcal{L}_z(\theta; m, g, p)\right]^T \\ - \mathbb{E} \left[ \nabla_{{\beta_{(g,z)}}} \mathcal{L}_z(\theta; m, g, p) \nabla_{{\beta_{(g,z)}}} \mathcal{L}_z(\theta; m, g, p)^T \right] \bigg] [k,l] \\
+		= \left[ \mathbb{E} \left[ {X}^T P \Delta^{(g,z)} s^{(g,z)} \right]\mathbb{E} \left[ {X}^T P \Delta^{(g,z)} s^{(g,z)} \right]^T\right][k,l] - \mathbb{E} \left[ {X}^T P \Delta s^{(g,z)} (s^{(g,z)})^T \Delta^{(g,z)} P {X}^T \right][k,l] \\
+		= \mathbb{E}\left[{X}[,k]^T P \Delta^{(g,z)} s^{(g,z)} \right] \mathbb{E} \left[ {X}[,l]^T P \Delta^{(g,z)} s^{(g,z)} \right] - \mathbb{E} \left[ {X}[,k]^T P \Delta^{(g,z)} s^{(g,z)} (s^{(g,z)})^T \Delta^{(g,z)} P {X}[,l ] \right] \\
+		=\mathbb{E}\left(\sum_{i=1}^n {x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_{i} \right) \mathbb{E} \left( \sum_{j=1}^n {x}_{jl} P_j \Delta^{(g,z)}_j s^{(g,z)}_j \right) \\ - \mathbb{E} \left( \sum_{i=1}^n \sum_{j=1}^n {x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i s^{(g,z)}_j \Delta^{(g,z)}_j P_j{x}_{jl} \right) \\
+		= \sum_{i=1}^n \sum_{j=1}^n \mathbb{E}[ {x}_{ik} P_i \Delta^{(g,z)}_is^{(g,z)}_i] \mathbb{E} [ {x}_{jl} P_j \Delta^{(g,z)}_j s^{(g,z)}_j]  -  \sum_{i=1}^n \sum_{j=1}^n \mathbb{E} [ {x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i s^{(g,z)}_j \Delta^{(g,z)}_jP_j {x}_{jl}  ]  \\
+		= \sum_{i=1}^n \sum_{j=1}^n \mathbb{E}[ {x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i ] \mathbb{E} \left[{x}_{jl} P_j \Delta^{(g,z)}_j s^{(g,z)}_j \right]  - \sum_{i \neq j} \mathbb{E} [{x}_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i] \mathbb{E}[s^{(g,z)}_j P_j \Delta^{(g,z)}_j {x}_{jl}] \\ - \sum_{i=1}^n \mathbb{E}[ {x}_{ik}P_i \Delta^{(g,z)}_i s^{(g,z)}_i s^{(g,z)}_i \Delta^{(g,z)}_i P_i{x}_{il}] \\
+		= \sum_{i=1}^n \mathbb{E}[ {x}_{ik} P_i \Delta^{(g,z)}_i H^{(g,z)}_i] \mathbb{E}[{x}_{il}P_i\Delta^{(g,z)}_i H^{(g,z)}_i] - \sum_{i=1}^n \mathbb{E}[{x}_{ik}P_i^2(\Delta^{(g,z)}_i)^2 (H^{(g,z)}_i)^2 {x}_{il}] \\ = \sum_{i=1}^n {x}_{ik} T_i^{\theta}(1)^2(\Delta^{(g,z)}_i)^2 (H_i^{(g,z)})^2{x}_{il} - \sum_{i=1}^n {x}_{ik} T_i^{\theta}(1) (\Delta_i^{(g,z)})^2 (H_i^{(g,z)})^2 {x}_{il} \\ = {X}[,k]^T T^{\theta}(1)^2 (\Delta^{(g,z)})^2 (H^{(g,z)})^2 {X}[,l] - {X}[,k]^T T^{\theta}(1) (\Delta^{(g,z)})^2 (H^{(g,z)})^2 {X}[,l]
 		\end{multline*}
@@
 		\begin{multline}\label{sub_mat_3_zeroinf_2}
-\E\left[ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right] \E\left[\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right]^T - \E\left[ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p)^T \right] \\ =
+		\mathbb{E}\left[ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right] \mathbb{E}\left[\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \right]^T - \mathbb{E}\left[ \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p)^T \right] \\ =
 		X^T T^{\theta}(1)^2 (\Delta^{(g,z)})^2(H^{(g,z)})^2{X} - {X}^T T^{\theta}(1) (\Delta^{(g,z)})^2 (H^{(g,z)})^2 {X} \\ = - X^T T^{\theta}(1) \left(\Delta^{(g,z)}\right)^2 \left( H^{(g,z)} \right)^2 \left( I - T^{\theta}(1)  \right) X.
@@
 		\begin{multline}\label{sub_mat_4_zeroinf}
-J_{(\beta_{(g,z)}, \beta_m)}(\theta; m, g) = - \E \left[\nabla_{\beta_{(g,z)}} \nabla_{\beta_m} \mathcal{L}_z (\theta; m, g, p) \right] \\ + \E\left[ \nabla_{\beta_{(g,z)} }\mathcal{L}_z(\theta ; m,g,p) \right]\E\left[\nabla_{\beta_m}\mathcal{L}_z(\theta; m,g,p)\right]^T - \E \left[\nabla_{\beta_{(g,z)}}\mathcal{L}_z (\theta; m,g,p) \nabla_{\beta_m}\mathcal{L}_z(\theta; m,g,p)]^T \right].
+		J_{(\beta_{(g,z)}, \beta_m)}(\theta; m, g) = - \mathbb{E} \left[\nabla_{\beta_{(g,z)}} \nabla_{\beta_m} \mathcal{L}_z (\theta; m, g, p) \right] \\ + \mathbb{E}\left[ \nabla_{\beta_{(g,z)} }\mathcal{L}_z(\theta ; m,g,p) \right]\mathbb{E}\left[\nabla_{\beta_m}\mathcal{L}_z(\theta; m,g,p)\right]^T - \mathbb{E} \left[\nabla_{\beta_{(g,z)}}\mathcal{L}_z (\theta; m,g,p) \nabla_{\beta_m}\mathcal{L}_z(\theta; m,g,p)]^T \right].
 		\end{multline} 
@@
 		\begin{equation}\label{sub_mat_4_zeroinf_1}
-- \E \left[\nabla_{\beta_{(g,z)}} \nabla_{\beta_m} \mathcal{L}_z (\theta; m, g, p) \right] = 0,
+		- \mathbb{E} \left[\nabla_{\beta_{(g,z)}} \nabla_{\beta_m} \mathcal{L}_z (\theta; m, g, p) \right] = 0,
 		\end{equation}
@@
 		\begin{multline}\label{sub_mat_4_zeroinf_2}
-[\E[\nabla_{\beta_{(g,z)}}\mathcal{L}_z(\theta; m, g, p) ] \E [\nabla_{\beta_m} \mathcal{L}_z(\theta; m, g, p) ]^T \\ - \E [\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \nabla_{\beta_m}\mathcal{L}_z(\theta; m, g, p)^T ] ][k,l] \\
-= \left[ \E \left[ {X}^T P \Delta^{(g,z)} s^{(g,z)} \right] \E\left[\tilde{X}^T \Delta^m s^m\right]^T\right][k,l] - \E \left[{X}^T P \Delta^{(g,z)} s^{(g,z)} (s^m)^T \Delta^m \tilde{X} \right][k,l] \\ = \left[\E\left[{X}[,k]^T P \Delta^{(g,z)} s^{(g,z)}\right]\E\left[\tilde{X}[,l]^T \Delta^m s^m\right]^T\right] - \E \left[{X}[,k]^T P \Delta^{(g,z)} s^{(g,z)} (s^m)^T \Delta^m \tilde{X}[,l] \right] \\ = \E\left(\sum_{i=1}^n x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i\right) \E \left(\sum_{j=1}^n \tilde{x}_{jl} \Delta^m_j s^m_j \right) - \E \left(\sum_{i=1}^n \sum_{j=1}^n x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i \Delta^m_j s^m_j \tilde{x}_{jl}  \right) \\
-= \sum_{i=1}^n \sum_{j=1}^n \E [x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i] \E[\Delta^m_js^m_j \tilde{x}_{jl}] - \sum_{i=1}^n \sum_{j=1}^n \E[x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i \Delta^m_j s^m_j \tilde{x}_{jl}] \\
-= \sum_{i=1}^n \sum_{j=1}^n \E [x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i ] \E[\Delta^m_js^m_j \tilde{x}_{jl}] - \sum_{i \neq j} \E [x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i] \E[\Delta^m_js^m_j \tilde{x}_{jl}] \\ - \sum_{i=1}^n \E[x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i \Delta^m_j s^m_j \tilde{x}_{jl}] \\
-= \sum_{i=1}^n \E[x_{ik} P_i \Delta_i^{(g,z)} H^{(g,z)}_i] \E[\tilde{x}_{il} \Delta^m_i H^m_i] - \sum_{i=1}^n \E[x_{ik} P_i \Delta_i^{(g,z)} H_i^{(g,z)} \Delta_i^m H_i^m \tilde{x}_{il}] \\ 
+		[\mathbb{E}[\nabla_{\beta_{(g,z)}}\mathcal{L}_z(\theta; m, g, p) ] \mathbb{E} [\nabla_{\beta_m} \mathcal{L}_z(\theta; m, g, p) ]^T \\ - \mathbb{E} [\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) \nabla_{\beta_m}\mathcal{L}_z(\theta; m, g, p)^T ] ][k,l] \\
+		= \left[ \mathbb{E} \left[ {X}^T P \Delta^{(g,z)} s^{(g,z)} \right] \mathbb{E}\left[\tilde{X}^T \Delta^m s^m\right]^T\right][k,l] - \mathbb{E} \left[{X}^T P \Delta^{(g,z)} s^{(g,z)} (s^m)^T \Delta^m \tilde{X} \right][k,l] \\ = \left[\mathbb{E}\left[{X}[,k]^T P \Delta^{(g,z)} s^{(g,z)}\right]\mathbb{E}\left[\tilde{X}[,l]^T \Delta^m s^m\right]^T\right] - \mathbb{E} \left[{X}[,k]^T P \Delta^{(g,z)} s^{(g,z)} (s^m)^T \Delta^m \tilde{X}[,l] \right] \\ = \mathbb{E}\left(\sum_{i=1}^n x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i\right) \mathbb{E} \left(\sum_{j=1}^n \tilde{x}_{jl} \Delta^m_j s^m_j \right) - \mathbb{E} \left(\sum_{i=1}^n \sum_{j=1}^n x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i \Delta^m_j s^m_j \tilde{x}_{jl}  \right) \\
+		= \sum_{i=1}^n \sum_{j=1}^n \mathbb{E} [x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i] \mathbb{E}[\Delta^m_js^m_j \tilde{x}_{jl}] - \sum_{i=1}^n \sum_{j=1}^n \mathbb{E}[x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i \Delta^m_j s^m_j \tilde{x}_{jl}] \\
+		= \sum_{i=1}^n \sum_{j=1}^n \mathbb{E} [x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i ] \mathbb{E}[\Delta^m_js^m_j \tilde{x}_{jl}] - \sum_{i \neq j} \mathbb{E} [x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i] \mathbb{E}[\Delta^m_js^m_j \tilde{x}_{jl}] \\ - \sum_{i=1}^n \mathbb{E}[x_{ik} P_i \Delta^{(g,z)}_i s^{(g,z)}_i \Delta^m_j s^m_j \tilde{x}_{jl}] \\
+		= \sum_{i=1}^n \mathbb{E}[x_{ik} P_i \Delta_i^{(g,z)} H^{(g,z)}_i] \mathbb{E}[\tilde{x}_{il} \Delta^m_i H^m_i] - \sum_{i=1}^n \mathbb{E}[x_{ik} P_i \Delta_i^{(g,z)} H_i^{(g,z)} \Delta_i^m H_i^m \tilde{x}_{il}] \\
 		= \sum_{i=1}^n \left[x_{ik} T_i^\theta(1) \Delta^{(g,z)}_i H^{(g,z)}_i \right] \cdot \left[{\Delta}^m_i(0) T^\theta_i(0) {H}^m_i(0) \tilde{x}_{il}(0) + {\Delta}^m_i(1) T_i^\theta(1) {H}^m_i(1) \tilde{x}_{il}(1)\right]
@@
 		\begin{multline}\label{sub_mat_6_zeroinf}
-J_{(\beta_{(g,z)},\pi)}(\theta; m, g) = \E \left[-\nabla_{\beta_{(g,z)}} \nabla_{\pi} \mathcal{L}_z(\theta; m, g, p) \right] + \E\left[ \nabla_{\beta_{(g,z)}}\mathcal{L}_z(\theta; m, g ,p)\right] \E \left[ \nabla_{\pi}\mathcal{L}_z(\theta; m,g,p) \right] \\ - \E \left[ \nabla_{\beta_{(g,z)}}\mathcal{L}_z(\theta; m,g,p) \nabla_{\pi}\mathcal{L}_z(\theta; m,g,p) \right].
+		J_{(\beta_{(g,z)},\pi)}(\theta; m, g) = \mathbb{E} \left[-\nabla_{\beta_{(g,z)}} \nabla_{\pi} \mathcal{L}_z(\theta; m, g, p) \right] + \mathbb{E}\left[ \nabla_{\beta_{(g,z)}}\mathcal{L}_z(\theta; m, g ,p)\right] \mathbb{E} \left[ \nabla_{\pi}\mathcal{L}_z(\theta; m,g,p) \right] \\ - \mathbb{E} \left[ \nabla_{\beta_{(g,z)}}\mathcal{L}_z(\theta; m,g,p) \nabla_{\pi}\mathcal{L}_z(\theta; m,g,p) \right].
 		\end{multline} Recall that $\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, p) = X^T P \Delta^{(g,z)}s^{(g,z)}$ and $\nabla_\pi \mathcal{L}_z(\theta; m, g, p) = a \left(\sum_{i=1}^n p_i\right) - b,$ where $$a = \frac{1}{\pi} + \frac{1}{1 - \pi}, \textrm{    } b = \frac{n}{1 - \pi}.$$
@@
 		\begin{equation}\label{sub_mat_6_zeroinf_1}
-\E \left[-\nabla_{\beta_{(g,z)}} \nabla_{\pi} \mathcal{L}_z(\theta; m, g, p) \right] = 0,
+		\mathbb{E} \left[-\nabla_{\beta_{(g,z)}} \nabla_{\pi} \mathcal{L}_z(\theta; m, g, p) \right] = 0,
 		\end{equation}
@@
 		\begin{multline}\label{sub_mat_6_zeroinf_2}
-\E \left[ \nabla_\pi \mathcal{L}_z(\theta; m, g, p) \right] \E\left[\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, x)[k] \right] - \E \left[\nabla_{\pi}\mathcal{L}(\theta; m,g,p) \nabla_{\beta_{(g,z)}} \mathcal{L} (\theta; m,g,p)[k] \right] \\= \left(\E \left[ a \sum_{i=1}^n p_i - b \right] \right)\left( \E \left[{X}[,k]^T P \Delta^{(g,z)} s^{(g,z)} \right]\right) - \E \left[ \left( a \sum_{i=1}^n p_i - b \right) {X}[,k]^T P \Delta^{(g,z)} s^{(g,z)}\right] \\ = \left( a \sum_{i=1}^n \E[p_i] - b \right) \left( \sum_{j=1}^n \E [x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] \right) \\ - \E \left[ \left( a \sum_{i=1}^n p_i - b \right) \left( \sum_{j=1}^n \tilde{x}_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j \right) \right] \\= a \sum_{i=1}^n \sum_{j=1}^n \E [p_i] \E[x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] - b \sum_{j=1}^n \E[x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] \\ - \left[a \sum_{i=1}^n \sum_{j=1}^n \E [ p_i x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] - b \sum_{j=1}^n \E[x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] \right] \\ = a \sum_{i=1}^n \sum_{j=1}^n \E[p_i] \E[{x}_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] - a \sum_{i \neq j} \E[p_i] \E[x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] - a \sum_{i=1}^n \E[ x_{ik} p_i^2 \Delta^{(g,z)}_i s^{(g,z)}_i] \\ = a \sum_{i=1}^n \E[p_i] \E[x_{ik} p_i \Delta^{(g,z)}_i s^{(g,z)}_i] - a \sum_{i=1}^n \E[x_{ik} p_i^2 \Delta^{(g,z)}_i s^{(g,z)}_i] \\ = a \sum_{i=1}^n T^\theta_i(1) x_{ik} T^\theta_i(1) \Delta^{(g,z)}_i s^{(g,z)}_i - a \sum_{i=1}^n x_{ik} T^\theta_i(1) \Delta^{(g,z)}_i s^{(g,z)}_i \\ = a \sum_{i=1}^n \left( {x}_{ik} T^\theta_i(1)^2 \Delta^{(g,z)}_i s^{(g,z)}_i - x_{ik} T^\theta_i(1) \Delta^{(g,z)}_i s^{(g,z)}_i \right) = a \sum_{i=1}^n {x}_{ik}T^\theta_i(1)\Delta^{(g,z)}_i s^{(g,z)}_i \left( T^\theta_i(1) - 1\right) \\ = -a\sum_{i=1}^n x_{ik} T_i(0) T^\theta_i(1)\Delta^{(g,z)}_i H^{(g,z)}_i = -a X[,k]^T w^{(g,z)}.
+		\mathbb{E} \left[ \nabla_\pi \mathcal{L}_z(\theta; m, g, p) \right] \mathbb{E}\left[\nabla_{\beta_{(g,z)}} \mathcal{L}_z(\theta; m, g, x)[k] \right] - \mathbb{E} \left[\nabla_{\pi}\mathcal{L}(\theta; m,g,p) \nabla_{\beta_{(g,z)}} \mathcal{L} (\theta; m,g,p)[k] \right] \\= \left(\mathbb{E} \left[ a \sum_{i=1}^n p_i - b \right] \right)\left( \mathbb{E} \left[{X}[,k]^T P \Delta^{(g,z)} s^{(g,z)} \right]\right) - \mathbb{E} \left[ \left( a \sum_{i=1}^n p_i - b \right) {X}[,k]^T P \Delta^{(g,z)} s^{(g,z)}\right] \\ = \left( a \sum_{i=1}^n \mathbb{E}[p_i] - b \right) \left( \sum_{j=1}^n \mathbb{E} [x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] \right) \\ - \mathbb{E} \left[ \left( a \sum_{i=1}^n p_i - b \right) \left( \sum_{j=1}^n \tilde{x}_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j \right) \right] \\= a \sum_{i=1}^n \sum_{j=1}^n \mathbb{E} [p_i] \mathbb{E}[x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] - b \sum_{j=1}^n \mathbb{E}[x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] \\ - \left[a \sum_{i=1}^n \sum_{j=1}^n \mathbb{E} [ p_i x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] - b \sum_{j=1}^n \mathbb{E}[x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] \right] \\ = a \sum_{i=1}^n \sum_{j=1}^n \mathbb{E}[p_i] \mathbb{E}[{x}_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] - a \sum_{i \neq j} \mathbb{E}[p_i] \mathbb{E}[x_{jk} p_j \Delta^{(g,z)}_j s^{(g,z)}_j] - a \sum_{i=1}^n \mathbb{E}[ x_{ik} p_i^2 \Delta^{(g,z)}_i s^{(g,z)}_i] \\ = a \sum_{i=1}^n \mathbb{E}[p_i] \mathbb{E}[x_{ik} p_i \Delta^{(g,z)}_i s^{(g,z)}_i] - a \sum_{i=1}^n \mathbb{E}[x_{ik} p_i^2 \Delta^{(g,z)}_i s^{(g,z)}_i] \\ = a \sum_{i=1}^n T^\theta_i(1) x_{ik} T^\theta_i(1) \Delta^{(g,z)}_i s^{(g,z)}_i - a \sum_{i=1}^n x_{ik} T^\theta_i(1) \Delta^{(g,z)}_i s^{(g,z)}_i \\ = a \sum_{i=1}^n \left( {x}_{ik} T^\theta_i(1)^2 \Delta^{(g,z)}_i s^{(g,z)}_i - x_{ik} T^\theta_i(1) \Delta^{(g,z)}_i s^{(g,z)}_i \right) = a \sum_{i=1}^n {x}_{ik}T^\theta_i(1)\Delta^{(g,z)}_i s^{(g,z)}_i \left( T^\theta_i(1) - 1\right) \\ = -a\sum_{i=1}^n x_{ik} T_i(0) T^\theta_i(1)\Delta^{(g,z)}_i H^{(g,z)}_i = -a X[,k]^T w^{(g,z)}.
 		\end{multline}
@@
 				\For{$i \in \{1, \dots, n\}$} \Comment{E step}
-		\State $T_i(1) \gets \P(P_i = 1 | M_i = m_i, G_i = g_i, \pi^\textrm{curr}, [\beta^g_1]^\textrm{curr}, [\beta_1^m]^\textrm{curr})$
+				\State $T_i(1) \gets \mathbb{P}(P_i = 1 | M_i = m_i, G_i = g_i, \pi^\textrm{curr}, [\beta^g_1]^\textrm{curr}, [\beta_1^m]^\textrm{curr})$
 				\State $T_i(0) \gets 1 - T_i(1)$
@@
 
-Let $\beta \in \R$ be an unknown constant.
+		Let $\beta \in \mathbb{R}$ be an unknown constant.
 Let $o_1, \dots, o_n \sim \mathcal{P}_1$, where $\mathcal{P}_1$ is a distribution.
 Let $Y_i|o_i, \dots, Y_n|o_i$ be exponential family-distributed random variables with identity sufficient statistic.
-Suppose the mean $\mu_i$ of $Y_i|o_i$ is given by $r(\mu_i) = \beta + o_i,$ where $r: \R \to \R$ is a strictly increasing, differentiable link function.
+Suppose the mean $\mu_i$ of $Y_i|o_i$ is given by $r(\mu_i) = \beta + o_i,$ where $r: \mathbb{R} \to \mathbb{R}$ is a strictly increasing, differentiable link function.
 We call this model the \textit{intercept-plus-offset} model.
@@
 (A special case, which corresponds to no weights, is $w_i = 1$ for all $i \in \{1, \dots, n\}$.) Throughout, we assume that $y_iw_i$ and $\exp(o_i)w_i$ have finite first moment.
-Suppose the cumulant-generating function and carrying density of the exponential family distribution are $\psi:\R \to \R$ and $c: \R \to \R$, respectively.
+Suppose the cumulant-generating function and carrying density of the exponential family distribution are $\psi:\mathbb{R} \to \mathbb{R}$ and $c: \mathbb{R} \to \mathbb{R}$, respectively.
 The canonical parameter $\eta_i$ of the $i$th observation is
@@
 		$\psi(\eta) = -\theta \log(1 - e^\eta).$ The derivative $\psi'$ of $\psi$ is
-$$ \psi'(t) = \theta \left(\frac{e^t}{1 - e^t}\right) = \frac{\theta}{e^{-t} - 1}.$$ Define the function $\delta: \R \to \R$ by $\delta(t) = -\log\left(\theta/t + 1 \right).$ We see that
+		$$ \psi'(t) = \theta \left(\frac{e^t}{1 - e^t}\right) = \frac{\theta}{e^{-t} - 1}.$$ Define the function $\delta: \mathbb{R} \to \mathbb{R}$ by $\delta(t) = -\log\left(\theta/t + 1 \right).$ We see that
 		$$\psi'(\delta(t)) = \frac{\theta}{\exp\left(\log(\theta/t + 1 )\right) - 1} = t,$$ implying $\delta = [\psi']^{-1}.$ By (\ref{can_param}), we have that
@@
 		We cannot solve for $\beta$ in (\ref{nb_mle}) analytically. However, we can derive an asymptotically exact solution. By the law of total expectation,
-$$ \E \left[\frac{w_i  e^{o_i} (y_i + \theta)}{e^{\beta + o_i} + \theta} \right] = \E \left[\E\left[\frac{w_i e^{o_i} (y_i + \theta) }{e^{\beta + o_i} + \theta} \bigg| (o_i, w_i) \right] \right] = \E \left[\frac{w_i e^{o_i} (e^{\beta + o_i} + \theta)}{ e^{\beta + o_i} + \theta} \right] = \E [w_i e^{o_i}];
-$$
-the second equality holds because $\E[y_i | o_i ] = \mu_i = e^{\beta + o_i}.$ Dividing by $n$ on both sides of (\ref{nb_mle}) and rearranging,
+		$$ \mathbb{E} \left[\frac{w_i  e^{o_i} (y_i + \theta)}{e^{\beta + o_i} + \theta} \right] = \mathbb{E} \left[\mathbb{E}\left[\frac{w_i e^{o_i} (y_i + \theta) }{e^{\beta + o_i} + \theta} \bigg| (o_i, w_i) \right] \right] = \mathbb{E} \left[\frac{w_i e^{o_i} (e^{\beta + o_i} + \theta)}{ e^{\beta + o_i} + \theta} \right] = \mathbb{E} [w_i e^{o_i}];
+		$$
+		the second equality holds because $\mathbb{E}[y_i | o_i ] = \mu_i = e^{\beta + o_i}.$ Dividing by $n$ on both sides of (\ref{nb_mle}) and rearranging,
 		\begin{equation}\label{nb_mo_3}
@@
 		\begin{equation}
-\hat{\beta}^{\textrm{NB}} \xrightarrow{P} \log\left(\frac{\E[w_i y_i]}{\E[w_i e^{o_i}]} \right).
+		\hat{\beta}^{\textrm{NB}} \xrightarrow{P} \log\left(\frac{\mathbb{E}[w_i y_i]}{\mathbb{E}[w_i e^{o_i}]} \right).
 		\end{equation}
 		But the Poisson MLE $\hat{\beta}^{\textrm{Pois}}$ (\ref{pois_mle}) converges in probability to the same limit:
-$$ \hat{\beta}^{\textrm{pois}} =  \log \left(\frac{ (1/n) \sum_{i=1}^n w_i y_i}{(1/n)\sum_{i=1}^n w_i e^{o_i}} \right) \xrightarrow{P} \log \left(\frac{\E[w_i y_i]}{ \E[w_i e^{o_i}]} \right).$$ Therefore, for large $n$, we can approximate $\hat{\beta}^{\textrm{NB}}$ by $\hat{\beta}^{\textrm{pois}}$.
+		$$ \hat{\beta}^{\textrm{pois}} =  \log \left(\frac{ (1/n) \sum_{i=1}^n w_i y_i}{(1/n)\sum_{i=1}^n w_i e^{o_i}} \right) \xrightarrow{P} \log \left(\frac{\mathbb{E}[w_i y_i]}{ \mathbb{E}[w_i e^{o_i}]} \right).$$ Therefore, for large $n$, we can approximate $\hat{\beta}^{\textrm{NB}}$ by $\hat{\beta}^{\textrm{pois}}$.
 
@@
 We present an alternate (and simpler, more general) derivation of the approximate formula for the MLE.
-Let $y_i$ be a random variable with finite first moment and conditional mean $\E[y_i | o_i] = \exp(\beta + o_i).$ By the law of total expectation,
-$$ \E[y_i] = \E[\E[ y_i | o_i]] = \E[\exp(\beta + o_i)] = \exp(\beta) \E(e^{o_i}),$$ or
+Let $y_i$ be a random variable with finite first moment and conditional mean $\mathbb{E}[y_i | o_i] = \exp(\beta + o_i).$ By the law of total expectation,
+		$$ \mathbb{E}[y_i] = \mathbb{E}[\mathbb{E}[ y_i | o_i]] = \mathbb{E}[\exp(\beta + o_i)] = \exp(\beta) \mathbb{E}(e^{o_i}),$$ or
 		\begin{equation}\label{nb_mo_4}
-\beta = \log\left(\frac{\E[y_i]}{E[e^{o_i}]} \right).
+		\beta = \log\left(\frac{\mathbb{E}[y_i]}{E[e^{o_i}]} \right).
 		\end{equation}
@@
 		\begin{equation}\label{nb_mo_5}
-\log\left( \frac{(1/n) \sum_{i=1}^n y_i }{(1/n) \sum_{i=1}^n e^{o_i}} \right) \xrightarrow{P}  \log\left(\frac{ \E[y_i] }{ \E[e^{o_i}] } \right) = \beta.
+		\log\left( \frac{(1/n) \sum_{i=1}^n y_i }{(1/n) \sum_{i=1}^n e^{o_i}} \right) \xrightarrow{P}  \log\left(\frac{ \mathbb{E}[y_i] }{ \mathbb{E}[e^{o_i}] } \right) = \beta.
 		\end{equation}

### 2021-12-29 text/jsm/jsm_paper.tex
reformat text for journal

@@
 The genomics literature has produced several applied methods for single-cell CRISPR screen analysis.
-In a prior work we developed SCEPTRE \cite{Barry2020}, a custom implementation of the conditional randomization test \cite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data.
+In a prior work we developed SCEPTRE \cite{Barry2021}, a custom implementation of the conditional randomization test \cite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data.
 SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification.

### 2022-01-04 text/GLM-EIV/jasa.tex
fix reproducibility

--- 2021-12-29 text/jasa/jasa.tex
@@
 backend=biber,
-style=numeric-comp,
+style=authoryear,
 maxcitenames=1,
 maxbibnames=3,
+uniquelist=false,
 backref=true,
@@
 \renewcommand{\algorithmicensure}{\textbf{Output:}}
-\addbibresource{/Users/timbarry/optionFiles/glmeiv.bib}
+\addbibresource{glmeiv.bib}
 \newtheorem{lemma}{Lemma}
@@
 \newcommand{\indep}{\raisebox{0.05em}{\rotatebox[origin=c]{90}{$\models$}}}
-
+\allowdisplaybreaks
 \makeatletter
@@
 \addtolength{\topmargin}{-1in}
-\date{\vspace{-5ex}}
 
 \begin{document}
+
 \def\spacingset#1{\renewcommand{\baselinestretch}%
@@
 {
-  \title{\bf A new class of measurement error models for CRISPR genome engineering and single-cell RNA sequencing}
-  \author{Timothy Barry$^1$, Eugene Katsevich$^2$, Kathyrn Roeder$^1$ \\
-    $^1$Department of Statistics and Data Science, Carnegie Mellon University \\
-    $^2$Department of Statistics and Data Science, Wharton School}
+	%  \title{\bf FDR control following filtering for hierarchically structured hypotheses}
+	\title{\bf A class of exponential family measurement error models for singe-cell CRISPR screens}
+	%	Eugene Katsevich, Chiara Sabatti, Marina Bogomolov}
+	\author{Timothy Barry\\
+		Department of Statistics and Data Science,\\Carnegie Mellon University\\
+		and\\
+		Eugene Katsevich\\
+		Department of Statistics and Data Science,\\Wharton School, University of Pennsylvania\\
+		and\\
+		Kathryn Roeder\thanks{ FUNDING INFORMATION GOES HERE.}\hspace{.2cm}\\
+		Department of Statistics and Data Science,\\Computational Biology Department,\\ 
+		Carnegie Mellon University}
 	\maketitle
@@
 	\begin{center}
-    {\LARGE \bf A new class of measurement error models for CRISPR genome engineering and single-cell RNA sequencing}
+		{\LARGE\bf Title}
 	\end{center}
@@
 Despite their promise, single-cell CRISPR screens present substantial statistical challenges.
-We demonstrate through theoretical and real data analyses that a standard method for estimation and inference in single-cell CRISPR screens --``thresholded regression'' -- exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic and challenging-to-select tuning parameter.
+We demonstrate through theoretical and real data analyses that a standard method for estimation and inference in single-cell CRISPR screens --``thresholded regression'' -- exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic, challenging-to-select tuning parameter.
 To overcome these difficulties, we introduce GLM-EIV (``GLM-based errors-in-variables''), a new method for single-cell CRISPR screen analysis.
-GLM-EIV extends the classical errors-in-variables model to responses and noisy predictors that are (i) exponential family-distributed and (ii) potentially impacted by the same set of confounding variables.
+GLM-EIV extends the classical errors-in-variables model to responses and noisy predictors that are exponential family-distributed and potentially impacted by the same set of confounding variables.
 We develop a computational infrastructure to deploy GLM-EIV across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters.
@@
 
-\noindent
-{\it Keywords:}  exponential family, errors-in-variables, GLM, CRISPR screen, differential expression
+\noindent%
+{\it Keywords:} genome engineering, cloud computing, exponential family, GLM, latent variable model, scRNA-seq
+
 \spacingset{1.9}
+
 \section{Introduction}
-CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new medical therapies \cite{Rothgangl2021,Musunuru2021} and transforming basic biology research \cite{Przybyla2021}.
-Recently, scientists have paired CRISPR genome engineering with single-cell sequencing \cite{Dixit2016,Datlinger2017}.
-The resulting assays, known as a ``single-cell CRISPR screens,'' link genetic perturbations in individual cells to changes in gene expression, illuminating regulatory networks underlying human diseases and other traits \cite{Morris2021a}.
+
+CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new medical therapies \parencite{Rothgangl2021,Musunuru2021} and transforming biological discovery \parencite{Przybyla2021}.
+Recently, scientists have paired CRISPR genome engineering with single-cell RNA sequencing \parencite{Dixit2016,Datlinger2017}.
+The resulting assays, known as a ``single-cell CRISPR screens,'' link genetic perturbations in individual cells to changes in gene expression, illuminating regulatory networks underlying human diseases and other traits \parencite{Morris2021a}.
 
 Despite their promise, single-cell CRISPR screens present substantial statistical challenges.
-A major difficulty is that CRISPR perturbations are assigned stochastically to cells and cannot be observed directly.
+A major difficulty is that the ``treatment'' -- i.e., the presence or absence of a CRISPR perturbation -- is assigned randomly to cells and is not directly observable.
 As a consequence, one cannot know with certainty which cells were perturbed.
-Instead, one must leverage an indirect, noisy proxy of perturbation presence or absence -- namely, transcribed guide RNA counts -- to ``guess'' which cells were perturbed.
-Using these imputed perturbation assignments, one can attempt to estimate the effect of the perturbation on gene expression.
-The standard approach, which we call ``thresholded regression'' or the ``thresholding method,'' is to assign perturbation identities to cells by simply thresholding the guide RNA counts.
+Instead, one must leverage an indirect, quantitative proxy of perturbation presence or absence to ``guess'' which cells received a perturbation.
+This indirect proxy takes the form of a so-called guide RNA count, with higher counts indicating that a cell is more likely to have been perturbed.
+The standard approach to single-cell CRISPR screen analysis is to impute perturbation assignments onto the cells by simply thresholding the guide RNA counts; using these imputations, one can attempt to estimate the effect of the perturbation on gene expression.
+We call this standard approach ``thresholded regression'' or the ``thresholding method.''
 
@@
 We assume that the response variable $y$ is a GLM of an underlying predictor variable $x^*$ and vector of confounders $z$.
-We do not observe $x^*$ directly; rather, we observe a noisy version $x$ of $x^*$ that itself is a GLM of $x^*$ and $z$.
+We do not observe $x^*$ directly; rather, we observe a noisy version $x$ of $x^*$ that itself is a GLM of $x^*$ and the same set of confounders $z$.
 The goal of the analysis is to estimate the effect of $x^*$ on $y$ using the observed data $(x, y, z)$ only.
@@
 Second, we introduce a new method, GLM-EIV (``GLM-based errors-in-variables''), for single-cell CRISPR screen analysis.
-GLM-EIV extends the classical errors-in-variables model to response distributions and sources of measurement error that are (i) exponential family-distributed and (ii) potentially impacted by the same set of confounding variables.
-GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic.
-% Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.
+GLM-EIV extends the classical errors-in-variables model \parencite{carroll} to responses and noisy predictors that are exponential family-distributed and potentially impacted by the same set of confounding variables.
+GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding.
 We implement several statistical accelerations (that possibly are of independent utility) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method.
@@
 \section{Background and analysis challenges}
+\iffalse
+\textcolor{red}{=====================================}
+\textcolor{blue}{I don't think this section works as is.
+I think the whole section should be relocated and revised, but let me say what you might do, if you keep it here.
+First, as much as I like the eSVD paragraph, I would argue that we aren't the only ones using negative binomial models for CRISPR screen analysis.
+For one, SCEPTRE did!
+So quite honestly I don't think this paragraph is appropriate (but see note below).
+With a heading like "Related work" people might expect to know what is being done in practice.
+Your SCEPTRE paragraph is the place to start, but I think we need to expand the lead sentence.
+Cite them all and then give a bit about at least one method available in the literature (very very little detail).
+I'd pick the Monacle based concept and just go with the idea that it is a negative binomial model that uses thresholding and that it sometimes doesn't work and results in a lot of false signals.
+Next you can the remainder of your SCEPTRE paragraph, but again I think it makes sense to say something about the basic model from which the CRT is built.
+In other words you want to convey that his method still uses thresholds, but solves the false positive problem.
+Now you are ready for your final few sentences about how you'll move beyond this.
+NOTE: Regarding your eSVD paragraph, you could use something like this paragraph in the Discussion, where you could make the point that it is a good thing when models are built on the generating distribution of the data (rather than forcing a normal by transformation).
+Then you could say this paper is like the others you cite in doing this.
+I make a try below, see the blue paragraph.\\
+Here is the alternative idea.
+Since my suggestion causes this section to be only one paragraph, you might consider having Assay Overview first and then putting this material at the lead of analysis challenges without a major heading.
+If you moved it to there you could do a better job of citing other works for analysis of CRISPR screens.
+I am a fan of giving credit to everyone, but doing so with terse sentences that say something like "Several methods have appeared in the literature, each with some shortcomings (ref, ref, ref).".
+Anyway, this section could lead people to believe we have very big egos, basically citing ourselves and almost no one else.}
+\textcolor{blue}{THIS IS HOW YOU MIGHT REWRITE THE SCEPTRE PARAGRAPH.The genomics literature has produced a few applied methods for single-cell CRISPR screen analysis, each with some shortcomings (cite Gasp, Xie, etc + Barry).
+The idea is to test whether a given perturbation is associated with the change in expression of a given gene, while ideally adjusting for sources of confounding and ensuring robustness to expression model misspecification.
+Most relevant is SCEPTRE, a custom implementation of the conditional randomization test \parencite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data \parencite{Barry2021}.
+SCEPTRE successfully handles the challenges of poor calibration and low power shared by other methods.
+In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE.
+Most importantly,  we account for the fact that the perturbation is measured with noise, an issue that available methods circumvent via thresholding.
+Additionally, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is challenging to attain within the nonparametric hypothesis testing framework of SCEPTRE.}
+\fi
 
@@
 Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions.
-For example, Lin et al.\ \cite{Lin2021} and Townes et al.\ \cite{Townes2019}  (separately) developed eSVD and GLM-PCA, generalizations of SVD and PCA, respectively, to exponential family response distributions.
-Unlike their vanilla counterparts, eSVD and GLM-PCA can model gene expression counts directly, thereby capturing the relationship between the mean and variance of a gene's expression level \cite{Lause2021} and improving performance on dimension reduction tasks.
+For example, Lin et al.\ \parencite{Lin2021} and Townes et al.\ \parencite{Townes2019}  (separately) developed eSVD and GLM-PCA, generalizations of SVD and PCA, respectively, to exponential family response distributions.
+Unlike their vanilla counterparts, eSVD and GLM-PCA can model gene expression counts directly, thereby capturing the relationship between the mean and variance of a gene's expression level \parencite{Lause2021} and improving performance on dimension reduction tasks.
 We see our work as a continuation of this effort to ``port'' common statistical methods and models to single-cell count data.
-Our focus, however, is on regression rather than dimension reduction.
-We extend the classical errors-in-variables model in several directions, most importantly generalizing to exponential family-distributed response distributions and sources of measurement error.
+Our focus, however, is on measurement error regression models rather than dimension reduction.
 
@@
 
-The genomics literature has produced several applied methods for single-cell CRISPR screen analysis.
-Barry et al.\ developed SCEPTRE \cite{Barry2021}, a custom implementation of the conditional randomization test \cite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data.
+The genomics literature has produced a few applied methods for single-cell CRISPR screen analysis.
+Most relevantly, Barry et al.\ developed SCEPTRE, a custom implementation of the conditional randomization test \parencite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data \parencite{Barry2021}.
 SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification.
@@
 
-%Other applied methods for single-cell CRISPR screen analysis include MIMOSCA \cite{Dixit2016} and scMAGeCK \cite{Yang2019}.
-% These methods, like SCEPTRE, focus on hypothesis testing (rather than estimation), but unlike SCEPTRE, they ignore the countedness of the data and are unable to handle confounders.
-
-\subsection{Assay overview}
-
-There are several broad classes of single-cell CRISPR screen assays, each suited to answer a different set of biological questions \cite{Gasperini2019,Datlinger2021,Mimitou2019}.
-In this work we focus on high-multiplicity of infection (MOI) single-cell CRISPR screens.
-We expect the ideas that we develop for this assay to apply (with some effort) to other classes of single-cell CRISPR screens as well.
-In this section we motivate high MOI single-cell screens, overview the experimental protocol, and present relevant analysis challenges.
-
-The human genome consists of genes, enhancers (segments of DNA that regulate the expression of one or more genes), and other genomic elements (that are not of importance to the current work).
-Genome-wide association studies (GWAS) have revealed that the majority ($>90\%$) of variants associated with diseases lie outside genes and (very likely) inside enhancers \cite{Gallagher2018}.
+\subsection{Assay background}
+There are several broad classes of single-cell CRISPR screen assays, each suited to answer a different set of biological questions \parencite{Gasperini2019,Datlinger2021,Mimitou2019}.
+In this work we focus on high-multiplicity of infection (MOI) single-cell CRISPR screens, which we motivate and describe here.
+The human genome consists of genes, enhancers (segments of DNA that regulate the expression of one or more genes), and other genomic elements.
+Genome-wide association studies (GWAS) have revealed that the majority ($>90\%$) of variants associated with diseases lie outside genes and (very likely) inside enhancers \parencite{Gallagher2018}.
 These noncoding variants are thought to contribute to disease by modulating the expression of one or more disease-relevant genes.
 Scientists do not know the gene (or genes) through which most noncoding variants exert their effect, limiting the interpretability of GWAS results.
-A central open challenge in genetics, therefore, is to link enhancers that harbor GWAS variants to the genes that they target at genome-wide scale \cite{Gasperini2020}.
+A central open challenge in genetics, therefore, is to link enhancers that harbor GWAS variants to the genes that they target at genome-wide scale \parencite{Gasperini2020}.
 
@@
 The perturbations assort into the cells randomly, with each cell receiving on average 10-40 distinct perturbations.
-Conversely, a given perturbation enters about 0.1-2\% of cells.
+Conversely, a given perturbation enters about 0.1-2\% of cells (this work).
 
@@
 
-\subsection{Analysis challenges}
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=1\linewidth]{figures/analysis_challenges.pdf}
+	\caption{\textbf{Experimental design and analysis challenges}: \textbf{a,} Experimental design.
+For a given perturbation (e.g., the perturbation indicated in blue), we partition the cells into two groups: perturbed and unperturbed.
+Next, for a given gene, we conduct a differential expression analysis across the two groups, yielding an estimate of the impact of the given perturbation on the given gene.
+\textbf{b,} DAG representing all variables in the system.
+The perturbation (latent) impacts both gene expression and gRNA expression; technical factors act as confounders, also impacting gene and gRNA expression.
+The target of estimation is the effect of the perturbation on gene expression.
+\textbf{c,} Schematic illustrating the ``background read'' phenomenon.
+Due to errors in the sequencing and alignment processes, unperturbed cells exhibit a nonzero gRNA count distribution (bottom).
+The target of estimation is the change in mean gene expression in response to the perturbation (top).
+\textbf{d}, Example data on four cells for a given perturbation-gene pair.
+Note that (i) the perturbation is unobserved, and (ii) the gene and gRNA data are discrete counts.}
+	\label{analysis_challenges}
+\end{figure}
+
 High MOI single-cell CRISPR screens present several statistical challenges, four of which we highlight here.
@@
 Finally, and most subtly, sequenced gRNAs sometimes map to cells that have not received a perturbation.
-This phenomenon, which we ``background contamination,'' results from errors in the sequencing and alignment processes \cite{Replogle2020}.
+This phenomenon, which we ``background contamination,'' results from errors in the sequencing and alignment processes \parencite{Replogle2020}.
 The marginal distribution of the gRNA counts is best conceptualized as a mixture model (Figure \ref{analysis_challenges}c; Gaussian distributions used for illustration purposes only).
@@
 
-In this work we analyze two large-scale, high MOI, single-cell CRISPR screen datasets published by Gasperini et al.\ \cite{Gasperini2019} and Xie et al.\ \cite{Xie2019a}.
+In this work we analyze two large-scale, high MOI, single-cell CRISPR screen datasets published by \parencite{Gasperini2019} and \parencite{Xie2019a}.
 Gasperini (resp., Xie) targeted approximately 6,000 (resp., 500) candidate enhancers in a population of approximately 200,000 (resp., 100,000) cells.
@@
 \section{Thresholding method}
-We study thresholding from empirical and theoretical perspectives, highlighting several inherent limitations of the approach.
+We study thresholding from empirical and theoretical perspectives, highlighting several limitations of the approach.
 Gasperini and Xie imputed perturbation assignments onto the cells via thresholding, but they carried out the subsequent differential expression analysis in different ways: Gasperini fitted negative binomial regression models to the data, whereas Xie applied nonparametric tests of independence.
-We study Gasperini's variant of the thresholding method, as it relates more closely to GLM-EIV and (in our view) is the more natural method.
+We study Gasperini's variant of the thresholding method, as it relates more closely to GLM-EIV.
 
@@
 \begin{itemize}
-	\item[1.] For a given threshold $c \in \mathbb{N}$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be $$ \hat{p}_i = 0 \textrm{ if } g_i < c; \quad \hat{p}_i = 1 \textrm{ if } g_i \geq c.$$
-	%$$\begin{cases} \hat{p}_i = 0 \textrm{ if } g_i < c, \\ \hat{p}_i = 1 \textrm{ if } g_i \geq c. \end{cases}$$
+	\item[1.] For a given threshold $c \in \mathbb{N}$, let the imputed perturbation assignment $\hat{p}_i \in \{0, 1\}$ be given by $\hat{p}_i = 0$ if  $g_i < c$ and $\hat{p}_i = 1$ otherwise.
 	\item[2.] Assume that $m_i$ is related to $\hat{p}_i, d^m_i,$ and $z_i$ through the following GLM:
@@
 
-	\item[3.]  Fit a GLM to obtain an estimate and confidence interval for the target of inference $\beta^m_1$.
+	\item[3.] Fit a GLM to obtain an estimate, CI, and $p$-value for the target of inference $\beta^m_1$.
 \end{itemize}
 The sequencing depth $d^m_i$ is included as an offset term in (\ref{thresh_glm}) so that $\beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i$ can be interpreted as a relative expression.
-Exponentiating both sides of (\ref{thresh_glm}) yields
-$$\mu_i = \exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m.$$ We see that $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration.
-
-Fold change in this context is defined as the ratio of mean gene expression in perturbed cells to mean gene expression in unperturbed cells.
-The target of inference $\beta^m_1$ is the \textit{log} fold change in expression in response to the perturbation, controlling for the technical factors.
-Therefore, $\exp(\beta^m_1) = 1$ (eq., $\beta^m_1 = 0$) indicates no change in expression, whereas $\exp(\beta^m_1) > 1$ (eq., $\beta^m_1 > 0$) and $\exp(\beta^m_1) < 1$ (eq., $\beta^m_1 < 0$) indicate an increase and decrease in expression, respectively.
+Exponentiating both sides of (\ref{thresh_glm}) shows that the mean $\mu_i$ of the $i$th cell is equal to
+$\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m;$ this quantity is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration.
+The target of inference $\beta^m_1$ is the log fold change in expression in response to the perturbation, controlling for the technical factors.
+Fold change in this context is defined as the ratio of the mean gene expression in perturbed cells to the mean gene expression in unperturbed cells.
+Therefore, $\exp(\beta^m_1) = 1$ (i.e., $\beta^m_1 = 0$) indicates no change in expression, whereas $\exp(\beta^m_1) > 1$ (i.e., $\beta^m_1 > 0$) and $\exp(\beta^m_1) < 1$ (i.e., $\beta^m_1 < 0$) indicate an increase and decrease in expression, respectively.
 
 \subsection{Empirical challenges of the thresholding method}\label{sec:thresholding_empirical}
-
-We examined the behavior of the thresholding method on real data and uncovered attenuation bias and bias-variance tradeoff effects.
-We applied the thresholding method to analyze the set of 322 positive control perturbation-gene pairs in the Gasperini dataset.
+To shed light on empirical challenges of the thresholding method, we applied the thresholding method to analyze the set of positive control perturbation-gene pairs in the Gasperini dataset.
 The positive control pairs consisted of perturbations that targeted gene transcription start sites (TSSs) for inhibition.
 Repressing the TSS of a given gene decreases its expression; therefore, the positive control pairs \textit{a priori} are expected to exhibit a strong decrease in expression.
-
-To investigate the sensitivity of the thresholding method to the selected threshold, we deployed the thresholding method to the positive control data using three different choices for the threshold: 1, 5, and 20.
-We found that the chosen threshold substantially impacted the results (Figure \ref{thresholding_empirical}a-b).
-Estimates for fold change produced by threshold = 1 were smaller in magnitude (i.e., closer to the baseline of $1$) than those produced by threshold = 5.
-(Figure \ref{thresholding_empirical}a.) Estimates produced by threshold = 5 and threshold = 20 were more concordant (Figure \ref{thresholding_empirical}b).
-
-We reasoned that thresholded regression systematically underestimated true effect sizes, especially for small thresholds.
+To investigate sensitivity to the choice of threshold, we deployed the thresholding method using three different thresholds: 1, 5, and 20.
+We found that the chosen threshold substantially impacted the results (Figure \ref{thresholding_empirical}a-b): estimates for fold change produced by threshold = 1 were smaller in magnitude (i.e., closer to the baseline of $1$) than those produced by threshold = 5.
+(Figure \ref{thresholding_empirical}a.) On the other hand, estimates produced by threshold = 5 and threshold = 20 were more concordant (Figure \ref{thresholding_empirical}b).
+
+We reasoned that thresholded regression systematically underestimated true effect sizes on the positive control pairs, especially for threshold = 1.
 For a given perturbation, the majority ($>98\%$) of cells are unperturbed.
@@
 
+% 2. Emprical challenges
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=1\linewidth]{figures/thresholding_empirical.jpeg}
+	\caption{\textbf{Empirical challenges of thresholded regression.} \textbf{a-b,} Fold change estimates produced by threshold = 1 versus threshold = 5 (a) and threshold = 20 versus threshold = 5 (b).
+The selected threshold substantially impacts the results.
+\textbf{c-d,} $p$-values (c) and CI widths (d) produced by threshold = 20 versus threshold = 5.
+The latter threshold yields more confident estimates.
+\textbf{e-f}, Empirical distribution of randomly-selected gRNA from Gasperini (e) and Xie (f) data (0 counts not shown).
+The gRNA data do not appear to imply an obvious threshold selection strategy.}
+	\label{thresholding_empirical}
+\end{figure}
+
 Finally, to determine whether there is an ``obvious'' location at which to draw the threshold, we examined the empirical gRNA count distributions and checked for bimodality.
@@
 
-Next, we studied the thresholding method from a theoretical perspective, recovering in precise mathematical terms phenomena revealed in the empirical analysis (plus several others).
-We worked in a simplified Gaussian setting.
+Next, we studied the thresholding method from a theoretical perspective, recovering in a simplified Gaussian setting phenomena revealed in the empirical analysis.
 Suppose we observe gRNA expression and gene expression data $(g_1, m_1), \dots, (g_n, m_n)$ on $n$ cells from the following linear model:
@@
 For a given threshold $c \in \mathbb{R}$, the imputed perturbation assignment $\hat{p}_i$ is $\hat{p}_i = \mathbb{I}(g_i \geq c).$ The thresholding estimator $\hat{\beta}^m_1$ is the OLS solution, i.e.
-$\hat{\beta}^m_1 = \left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})(m_i - \overline{m})\right]\left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})^2\right]^{-1}.$ We derive the almost sure limit of $\hat{\beta}^m_1$:
+$\hat{\beta}^m_1 = \left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})^2\right]^{-1}\left[\sum_{i=1}^n (\hat{p}_i - \overline{\hat{p}})(m_i - \overline{m})\right].$ We derive the almost sure limit of $\hat{\beta}^m_1$:
 \begin{proposition}\label{prop:convergence}
@@
 	\end{equation} where 
-	$$
-	\mathbb{E}[\hat{p}_i] = \zeta(1-\pi) + \omega\pi; \quad \omega = \Phi\left(\beta_1^g + \beta_0^g -c \right); \quad \zeta = \Phi\left( \beta^g_0 - c \right).$$
+	$
+	\mathbb{E}[\hat{p}_i] = \zeta(1-\pi) + \omega\pi$, $\omega \equiv \Phi\left(\beta_1^g + \beta_0^g -c \right)$, and $\zeta \equiv \Phi\left( \beta^g_0 - c \right).$
 \end{proposition}
@@
 Having derived an exact expression for the asymptotic relative bias of $\hat{\beta}^m_1$, we can prove several results about this quantity.
-For the sake of simplicity, we consider a slightly specialized setting in which we set $\pi$ to to a specific value (see Appendix \ref{sec:appendix_theory} for details; proofs and detailed statements likewise deferred to Appendix \ref{sec:appendix_theory}).
-
-First, the thresholding estimator strictly \textit{underestimates} (in absolute value) the true value of $\beta^m_1$ over all choices of the threshold and over all possible values of the model parameters.
-This phenomenon, called attenuation bias, is a common attribute of estimators that ignore measurement in errors-in-variables models \cite{Stefanski2000a}.
+We fix $\pi$ to $1/2$ for simplicity.
+(In reality, $\pi$ is smaller, but the relevant statistical phenomena emerge for $\pi = 1/2$.) First, the thresholding estimator strictly \textit{underestimates} (in absolute value) the true value of $\beta^m_1$ over all choices of the threshold $c$ and over all values of the regression coefficients $(\beta^m_0, \beta^m_1)$ and $(\beta^g_0, \beta^g_1)$.
+This phenomenon, called attenuation bias, is a common attribute of estimators that ignore measurement in errors-in-variables models \parencite{Stefanski2000a}.
 Second, the magnitude of the bias decreases monotonically in $\beta^g_1$, comporting with the intuition that the problem becomes easier as the gRNA mixture distribution becomes increasingly well-separated.
-Third, the Bayes-optimal decision boundary $c_\textrm{bayes} \in \mathbb{R}$ (i.e., the most accurate decision boundary for classifying cells as perturbed or unperturbed) is a critical value of the bias function.
-Finally, and most subtly, there is no universally applicable rule for selecting a threshold that produces minimal bias: when $\beta^g_1$ is small, setting the threshold to an arbitrarily large number yields smaller bias than setting the threshold to the Bayes decision boundary; when $\beta^g_1$ is large, the reverse is true.
+Third, the Bayes-optimal decision boundary $c_\textrm{bayes} \in \mathbb{R}$ (i.e., the most accurate decision boundary for classifying cells) is a critical value of the bias function.
+Finally, and most subtly, there is no universally applicable rule for selecting a threshold that yields minimal bias: when $\beta^g_1$ is small, setting the threshold to an arbitrarily large number yields smaller bias than setting the threshold to the Bayes decision boundary; when $\beta^g_1$ is large, the reverse is true.
+Appendix \ref{sec:appendix_theory} contains detailed proposition statements and proofs of these results.
 
@@
 Suppose the intercepts in (\ref{theoretical_model}) are fixed at $0$ (i.e., $\beta^m_0$ = $\beta^g_0$ = 0).
-For notational simplicity write $\beta_m = \beta^m_1$ and $\beta_g = \beta^g_1.$ The thresholding estimator $\hat{\beta}_m$ is the no-intercept OLS solution $\hat{\beta}_m = \left[\sum_{i=1}^n \hat{p}_i m_i \right]\left[\sum_{i=1}^n \hat{p}_i^2 \right]^{-1}.$ We have the following limiting distributional result.
+For notational simplicity we write $\beta_m = \beta^m_1$ and $\beta_g = \beta^g_1.$ The thresholding estimator $\hat{\beta}_m$ is the no-intercept OLS solution $\hat{\beta}_m = \left[\sum_{i=1}^n \hat{p}_i^2 \right]^{-1}\left[\sum_{i=1}^n \hat{p}_i m_i \right].$ The following proposition derives the scaled, asymptotic distribution of $\hat{\beta}_m:$
 \begin{proposition}\label{prop:bv_decomp}
 	The limiting distribution of $\hat{\beta}_m$ is
-	$$\sqrt{n}(\hat{\beta}_m - l) \xrightarrow{d} N\left(0, \frac{ \beta_m \omega\pi(\beta_m - 2l) + \mathbb{E}[\hat{p}_i](1 + l^2) }{\left(\mathbb{E}[\hat{p}_i]\right)^2} \right),$$ where $$l = \beta_m \omega \pi/[\zeta(1-\pi) + \omega \pi]; \quad
+	$$\sqrt{n}(\hat{\beta}_m - l) \xrightarrow{d} N\left(0, \frac{ \beta_m \omega\pi(\beta_m - 2l) + \mathbb{E}[\hat{p}_i](1 + l^2) }{\left(\mathbb{E}[\hat{p}_i]\right)^2} \right),$$ where $$l \equiv \beta_m \omega \pi/[\zeta(1-\pi) + \omega \pi]; \quad
 	\mathbb{E}[\hat{p}_i] = \pi \omega + (1-\pi) \zeta; \quad
-	\omega = \Phi(\beta_g - c); \quad
-	\zeta = \Phi(-c).$$
+	\omega \equiv \Phi(\beta_g - c); \quad
+	\zeta \equiv \Phi(-c).$$
 \end{proposition}
-This proposition yields an asymptotically exact bias-variance decomposition for $\hat{\beta}_m$ (Figure \ref{thresholding_theoretical}): as the threshold tends to infinity, the bias decreases and the variance increases, consistent with empirical observations that we made in Section \ref{sec:thresholding_empirical}.
-Overall, our empirical and theoretical analyses reveal that thresholded regression poses several difficulties: thresholding incurs strict attenuation bias (that in some settings is large), and selecting a good threshold is challenging.
-These considerations motivate our core research question: \textit{does modeling the gRNA counts directly, thereby circumventing thresholding, facilitate estimation and inference in single-cell CRISPR screen analysis}?
-
-\section{GLM-EIV}
-
-To answer this question, we introduce GLM-EIV (GLM-based errors-in-variables), a method that models the entire data-generating process in a single-cell CRISPR screen experiment.
+This proposition yields an asymptotically exact bias-variance decomposition for $\hat{\beta}_m$: as the threshold tends to infinity, the bias decreases and the variance increases (Appendix \ref{sec:appendix_theory}).
+Overall, our empirical and theoretical analyses indicate that thresholded regression poses a basic difficulty: selecting a good threshold is challenging, and even if the threshold is optimal, the estimator could suffer from high bias or high variance.
+These considerations motivate our research question: does modeling the gRNA counts directly, thereby circumventing thresholding, facilitate estimation and inference in single-cell CRISPR screens?
+
+\section{GLM-based errors-in-variables (GLM-EIV)}
+
+To answer this question, we introduce the GLM-EIV model.
 
 \subsection{Model}\label{sec:model}
-Building on the work of several previous authors \cite{Townes2019,Svensson2020,Hafemeister2019}, Sarkar and Stephens \cite{Sarkar2021} proposed a simple strategy for modeling single-cell gene expression data, which, in the framework of negative binomial GLMs, is equivalent to using the log-transformed library size as an offset term (as in (\ref{thresh_glm})).
+Building on the work of several previous authors \parencite{Townes2019,Svensson2020,Hafemeister2019}, Sarkar and Stephens proposed a simple strategy for modeling single-cell gene expression data, which, in the framework of negative binomial GLMs, is equivalent to using the log-transformed library size as an offset term (as in (\ref{thresh_glm}); \cite{Sarkar2021}).
 We generalize Sarkar and Stephens' approach to model \textit{both} gene and gRNA modalities.
@@
 where, similar to above, $\theta^g > 0$ is a known negative binomial size parameter, and $\beta^g_0 \in \mathbb{R}, \beta^g_1 \in \mathbb{R}, \gamma_g \in \mathbb{R}^{d - 2}$ are unknown constants.
-We use a negative binomial GLM to model the gRNA counts because the gRNA molecules are transcribed in the cell in the same way as gene transcripts \cite{Datlinger2017,Hill2018}.
+We use a negative binomial GLM to model the gRNA counts because the gRNA molecules are transcribed in the cell in the same way as gene transcripts \parencite{Datlinger2017,Hill2018}.
 Finally, we model the marginal perturbation probability as $
@@
 Likewise, the target of inference $\beta^m_1$ is the log fold change in gene expression in response to the perturbation, accounting for technical factors.
-See Appendix \ref{sec:zero_inf_model} for a parallel methodology that we developed in which we model the gRNA distribution as (in a certain sense) zero-inflated counts that may be applicable in some settings.
-
-\subsubsection*{General model}
-To provide greater modeling flexibility, we generalize the GLM-EIV model to arbitrary exponential family response distributions and link functions.
+See Appendix \ref{sec:zero_inf_model} for a parallel methodology in which we instead model the gRNA distribution using a zero-inflated GLM.
+
+\paragraph{General model} To provide greater modeling flexibility, we generalize the GLM-EIV model to arbitrary exponential family response distributions and link functions.
 To increase notational compactness, let $\tilde{x_i} = [1, p_i, z_i]^T \in \mathbb{R}^d$ be the vector of covariates (including an intercept term) for the $i$th cell.
@@
 
-Let the terms $l^g_i, o^g_i, \mu^g_i, \eta^g_i, \psi_g, r_g, h_g$ and $c_g$ be defined in an analogous way for the gRNA model:
-$$
-l^g_i = \langle \tilde{x}_i, \beta_g \rangle + o^g_i; \quad
-r_g(\mu^g_i) = l^g_i; \quad
-\eta^g_i = \left([\psi'_g]^{-1} \circ r^{-1}_g\right)(l_i^g) := h_g(l_i^g).
-$$
-The density $f_g$ of $g_i$ given the canonical parameter is $f_g(m_i; \eta^g_i) = \exp\left\{g_i \eta^g_i - \psi_g(\eta^g_i) + c_g(g_i)\right\}.$
+Let the terms $l^g_i, o^g_i, \mu^g_i, \eta^g_i, \psi_g, r_g, h_g$ and $c_g$ be defined in an analogous way for the gRNA model, i.e.
+$l^g_i = \langle \tilde{x}_i, \beta_g \rangle + o^g_i$, $r_g(\mu^g_i) = l^g_i$, and $\eta^g_i = \left([\psi'_g]^{-1} \circ r^{-1}_g\right)(l_i^g) = h_g(l_i^g).$ The density $f_g$ of $g_i$ given the canonical parameter is $f_g(m_i; \eta^g_i) = \exp\left\{g_i \eta^g_i - \psi_g(\eta^g_i) + c_g(g_i)\right\}.$
 Finally, the unobserved variable $p_i$ is assumed to follow a Bernoulli distribution with mean $\pi \in (0, 1/2]$. Its marginal density $f_p$ is given by $f_p(p_i) = \pi^{p_i}(1-\pi)^{1 - p_i}.$
@@
 
-\subsubsection*{Notation}
-We briefly introduce notation that we will use throughout.
+\paragraph{Notation} We briefly introduce notation that we will use throughout.
 For $j \in \{0,1\}$, let $\tilde{x}_i(j) := [1, j, z_i]^T$ denote the value of $\tilde{x}_i$ that results from setting $p_i$ to $j$.
 Next, let  $l^m_i(j)$, $\eta^m_i(j),$ and $\mu^m_i(j)$ be the values of $l^m_i$, $\eta^m_i$, and $\mu^m_i$, respectively, that result from setting $p_i$ to $j$, i.e.,
-$$
-l^m_i(j) := \langle \tilde{x}_i(j), \beta_m \rangle + o^m_i; \quad \eta^m_i(j) := h_m(l^m_i(j)); \quad
-\mu_i^m(j) = r_m^{-1}(l^m_i(j)).
-$$
-Let the corresponding gRNA quantities $l^g_i(j)$, $\eta_i^g(j)$, and $\mu^g_i(j)$ be defined analogously.
-Let $X \in \mathbb{R}^{n \times {d-1}}$ be the observed design matrix, and let $\tilde{X} \in \mathbb{R}^{n \times d}$ be the augmented design matrix that results from concatenating the column of (unobserved) $p_i$s to $X$, i.e.
-$$X := \begin{bmatrix} 
+$l^m_i(j) = \langle \tilde{x}_i(j), \beta_m \rangle + o^m_i$, $\eta^m_i(j) = h_m(l^m_i(j))$, and
+$\mu_i^m(j) = r_m^{-1}(l^m_i(j)).$ Let the corresponding gRNA quantities $l^g_i(j)$, $\eta_i^g(j)$, and $\mu^g_i(j)$ be defined analogously.
+Next, let $X \in \mathbb{R}^{n \times {d-1}}$ be the observed design matrix, and let $\tilde{X} \in \mathbb{R}^{n \times d}$ be the augmented design matrix that results from concatenating the column of (unobserved) $p_i$s to $X$, i.e.
+$$ X := \begin{bmatrix} 
 1 & z_1 \\
-1 & z_2 \\
 \vdots & \vdots \\
@@
 1 & p_1 & z_1 \\
-1 & p_2 & z_2 \\ 
 \vdots & \vdots & \vdots \\
@@
 \end{bmatrix} = \begin{bmatrix}
-\tilde{x}_1^T \\ \tilde{x}_2^T \\ \vdots \\ \tilde{x}_n^T
+\tilde{x}_1^T \\ \vdots \\ \tilde{x}_n^T
 \end{bmatrix}.
@@
 Finally, let $[m,m]^T \in \mathbb{R}^{2n}$ be the vector that results from concatenating $m$ to itself, i.e.
-$$ [m,m]^T := [\underbrace{m_1, m_2 \dots, m_{n-1}, m_n}_\textrm{first copy of $m$}, \underbrace{m_1, m_2, \dots, m_{n-1}, m_n}_\textrm{second copy of $m$}],$$ and let $[g,g]^T$, $[o^g,o^g]^T$, and $[o^m,o^m]^T$ be defined similarly.
-
-\subsubsection*{Log likelihood and model properties}
-We derive the log-likelihood of the GLM-EIV model.
+$[m,m]^T := [m_1, \dots, m_n, m_1, \dots, m_n],$ and let $[g,g]^T$, $[o^g,o^g]^T$, and $[o^m,o^m]^T$ be defined similarly.
+
+\paragraph{Log likelihood and model properties} We derive the log-likelihood of the GLM-EIV model.
 We conduct estimation and inference conditional on the library sizes and technical factors $l^m_i, l^g_i,$ and $z_i$; therefore, we treat these quantities as fixed constants.
@@
 GLM-EIV extends (\ref{classical_eiv}) in at least three directions: first, GLM-EIV allows $y_i$ and $x_i$ to follow exponential family (i.e, not just Gaussian) distributions; second, GLM-EIV allows $y_i$ and $x_i$ to be related to $x^*_i$ through linear \textit{or} nonlinear link functions; and finally, GLM-EIV allows confounders $z_i$ to impact both $x_i$ and $y_i$.
-Therefore, $x_i$ and $y_i$ are dependent \textit{even after conditioning on $x^*_i$}, enabling GLM-EIV to capture more complex dependence relationships between $x_i$ and $y_i$ than is possible in (\ref{classical_eiv}) or other standard measurement error models.
+Therefore, $x_i$ and $y_i$ are conditionally \textit{dependent} given $x^*_i$, enabling GLM-EIV to capture more complex dependence relationships between $x_i$ and $y_i$ than is possible in (\ref{classical_eiv}) or other standard measurement error models.
 
@@
 We iterate through the E and M steps until the log likelihood (\ref{marginal_log_lik}) converges (see Appendix \ref{sec:glmeiv_details}).
-Our EM algorithm is reminiscent of (but distinct from) that of Ibrahim \cite{Ibrahim1990}, who also applied weighted GLM solvers to carry out the M step.
+Our EM algorithm is reminiscent of (but distinct from) that of Ibrahim  \parencite{Ibrahim1990}, who also applied weighted GLM solvers to carry out the M step.
 
@@
 This strategy, however, is prohibitively slow, as the data are large and the EM algorithm is iterative.
-Therefore, we derive an analytic formula for the asymptotic observed information matrix using Louis's Theorem \cite{Louis1982} (Appendix \ref{sec:glmeiv_details}).
+Therefore, we derive an analytic formula for the asymptotic observed information matrix using Louis's Theorem (\cite{Louis1982}; Appendix \ref{sec:glmeiv_details}).
 Leveraging this analytic formula, we can calculate standard errors quickly, enabling us to perform inference in practice on real, large-scale data.
 
-\subsection{Statistical accelerations}
+\subsection{Statistical accelerations and computational infrastructure}
 A downside of the the EM algorithm (Algorithm \ref{algo:em_full}) is that it requires fitting many GLMs.
 Assuming that we run the algorithm 15 times using randomly-generated pilot estimates, and assuming that the algorithm iterates through E and M steps about 10 times per run, we must fit approximately 300 GLMs.
-(These numbers are based on exploratory applications of the method to real and simulated data.) We devised a strategy to produce a highly accurate pilot estimate $(\pi^\textrm{pilot}, \beta_m^\textrm{pilot}, \beta_g^\textrm{pilot})$ of the true parameters, enabling us to run the algorithm once and converge upon the MLE within a few iterations.
-The strategy involves layering several statistical ``tricks'' (some new, some old) on top of one another.
-We expect these tricks to be independently useful for accelerating other single-cell methods.
-
-\begin{algorithm}
-	\caption{Computing $[\beta_0^m]^\textrm{pilot}$, $[\gamma^T_m]^\textrm{pilot}$, $[\beta_0^g]^\textrm{pilot}$, and $[\gamma^T_g]^\textrm{pilot}$ .}\label{algo:pilot_estimates_1}
-	\begin{algorithmic}[2]
-		\Require Data $m$, $g$, $o^m$, $o^g$, and $X$; gene expression distribution $f_m$ and link function $r^{-1}_m$; gRNA expression distribution $f_g$ and link function $r^{-1}_g$; number of EM starts $B$.
-		\For {$k \in \{m,g\}$}
-		\State \multiline{Fit a GLM $GLM_k$ with responses $k$, offsets $o^k$, design matrix $X$, distribution $f_k$, and link function $r^{-1}_k$.}
-		\State Set $[\beta_0^k]^\textrm{pilot}$ and $[\gamma_k^T]^\textrm{pilot}$ to the fitted coefficients of $GLM_k$.
-		\For{$i \in \{1, \dots, n\}$}
-		\State $\hat{f}^k_i \gets [\beta_0^k]^\textrm{pilot} + \langle [\gamma_k^T]^\textrm{pilot} , z_i \rangle + o_i^k$ \Comment{untransformed fitted values}
-		\EndFor
-		\EndFor
-		\State \textbf{return} $([\beta_0^m]^\textrm{pilot}$, $\hat{f}^m$, $[\gamma^T_m]^\textrm{pilot}$, $[\beta_0^g]^\textrm{pilot}$, $[\gamma^T_g]^\textrm{pilot}$, $\hat{f}^g)$
-	\end{algorithmic}
-\end{algorithm}
-
-The first step (Algorithm \ref{algo:pilot_estimates_1}) is to obtain good parameter estimates for $[\beta^m_0, \gamma_m]^T$ and $[\beta^g_0, \gamma_g]^T$ via regression.
-Recall that the underlying gene expression parameter vector $\beta_m$ is $\beta_m = [\beta^m_0, \beta^m_1, \gamma_m]^T \in \mathbb{E}^d$, where $\beta^m_0$ is the intercept, $\beta^m_1$ is the effect of the perturbation, and $\gamma_m^T$ is the effect of the technical factors.
-To produce estimates $[\beta^m_0]^\textrm{pilot}$ and $[\gamma_m^T]^\textrm{pilot}$, we regress the gene expressions $m$ onto the technical factors $X$.
-The intuition for this procedure is as follows: the probability of perturbation $\pi$ is very small.
-Therefore, the true log likelihood is approximately equal to the log likelihood that results from omitting $p_i$ from the model:
-\begin{multline*}
-\sum_{i=1}^n f_m(m_i; \eta^m_i) = \underbrace{\sum_{i : p_i =1} f_m(m_i; h_m(\beta_0 + \beta_1 + \gamma^T z_i + o^m_i))}_\textrm{few terms}  + \underbrace{\sum_{i : p_i = 0} f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i))}_\textrm{many terms} \\ \approx \sum_{i = 1}^n f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i)).
-\end{multline*}
-We similarly can obtain pilot estimates $[\beta^g_0]^\textrm{pilot}$ and $[\gamma^T_g]^\textrm{pilot}$ by regressing the gRNA counts $g$ onto the technical factors $X$.
-We extract the fitted values (on the scale of the linear component) for use in a subsequent step: $\hat{f}^k_i = [\beta^k_0]^\textrm{pilot} + \langle [\gamma^T_k]^\textrm{pilot}, z_i \rangle + o^k_i,$ for $k \in \{m,g\}$.
-
-Next, we obtain estimates $[\beta_1^m]^\textrm{pilot},$ $[\beta_1^g]^\textrm{pilot},$ and $\pi^\textrm{pilot}$ for $\beta^m_1$, $\beta^g_1$, and $\pi$ by fitting a ``reduced'' GLM-EIV.
-We describe the reduced GLM-EIV procedure in text; the algorithm box (Algorithm \ref{algo:pilot_estimates_2}) is deferred to Appendix \ref{sec:statistical_accelerations}.
-The log likelihood of the no-intercept, univariate GLM with predictor $p_i$ and offset $\hat{f}^m_i$ is approximately equal to the true log likelihood:
-$$ \sum_{i=1}^n f_m(m_i; \eta^m_i) =
-\sum_{i=1}^n f_m(m_i; h_m(\beta_0 + \beta_1 p_i + \gamma^T z_i + o^m_i)) \approx \sum_{i=1}^n f_m(m_i; h_m( \beta_1 p_i + \hat{f}^m_i)).
-$$
-Therefore, to estimate $\beta^m_1$, $\beta^g_1$, and $\pi$, we fit a GLM-EIV model with gene expressions $m$, gRNA counts $g$, gene offsets $\hat{f}^m := [\hat{f}^m_1, \dots, \hat{f}^m_n]^T$, gRNA offsets $\hat{f}^g := [\hat{f}^g_1, \dots, \hat{f}^g_n]^T$, and \textit{no} intercept or covariate terms.
-Intuitively, we ``encode'' all information about technical factors, library sizes, and baseline expression levels into $\hat{f}^m$ and $\hat{f}^g$.
-We run the algorithm $B \approx 15$ times over randomly-selected starting values for $\beta^m$, $\beta^g$, and $\pi$ and select the solution with greatest the log likelihood.
-
-The M step of the reduced GLM-EIV algorithm requires fitting two no-intercept, univariate GLMs with offsets.
-We derive analytic formulas for the MLEs of these GLMs in the three most important cases: Gaussian response with identity link, Poisson response with log link, and negative binomial response with log link (see Appendix \ref{sec:statistical_accelerations}; the latter formula is asymptotically exact).
-Consequently, we do not need to run the relatively slow IRLS procedure to carry out the M step of the reduced GLM-EIV algorithm.
-Overall, the proposed method for obtaining the full set of pilot parameter estimates requires fitting only two GLMs (via IRLS).
-
-\subsection{Computing}\label{sec:computing}
-
-We developed a computational infrastructure to apply GLM-EIV to large-scale, single-cell CRISPR screen data.
-The infrastructure leverages Nextflow, a programming language that facilitates building data-intensive pipelines \cite{DITommaso2017}, and ondisc, an R package that we developed (in a separate project) to facilitate large-scale computing on single-cell data.
-Nextflow and ondisc together enable the construction of highly portable single-cell pipelines: one can analyze data \textit{out-of-memory} on a laptop or in a \textit{distributed} fashion across tens or hundreds of nodes on a cloud (e.g., Microsoft Azure, Google Cloud) or high-performance cluster.
-
-\begin{algorithm}
-	\caption{Applying GLM-EIV at scale.}\label{algo:at_scale}
-	\begin{algorithmic}
-		\State $G \gets \{\textrm{gene}_1, \dots, \textrm{gene}_{d_g}\}; P \gets \{\textrm{perturbation}_1, \dots, \textrm{perturbation}_{d_p}\}$
-
-		\For{gene $\in G$}
-		\State Run precomputation (Algorithm \ref{algo:pilot_estimates_1}) on gene; save $\hat{f}^m$, $[\beta^m_0]^\textrm{pilot}$ and $[\gamma^T_m]^\textrm{pilot}$.
-		\EndFor
-		\For{perturbation $\in P$}
-		\State Run precomputation  (Algorithm \ref{algo:pilot_estimates_1}) on perturbation; save $\hat{f}^g$, $[\beta^g_0]^\textrm{pilot}$ and $[\gamma^T_g]^\textrm{pilot}$.
-		\EndFor
-		\For{(gene, perturbation) $\in G \times P$}
-		\State Load $\hat{f}^m, \hat{f}^g,$ $[\beta^m_0]^\textrm{pilot}$ $[\gamma^T_m]^\textrm{pilot}$, $[\beta^g_0]^\textrm{pilot}$ and $[\gamma^T_g]^\textrm{pilot}$.
-		\State Compute $[\beta^m_1]^\textrm{pilot}, [\beta^g_1]^\textrm{pilot}, \pi^\textrm{pilot}$ by fitting a reduced GLM-EIV (Algorithm \ref{algo:pilot_estimates_2}).
-		\State Run GLM-EIV using the pilot parameters (Algorithm \ref{algo:em_full}).
-		\EndFor
-	\end{algorithmic}
-\end{algorithm}
-
-Leveraging these technologies, we develop a scalable and efficient pipeline for GLM-EIV (Algorithm \ref{algo:at_scale}).
-First, we run a round of ``precomputations'' on all $d_g$ genes and $d_p$ perturbations.
-The precomputations involve regressing the gene expressions (or gRNA counts) onto the technical factors, thereby ``factoring out'' Algorithm \ref{algo:pilot_estimates_1}.
-Next, we run differential expression analyses on the full set of gene-perturbation pairs; for a given pair, this amounts to obtaining the complete set of pilot parameters (by running a reduced GLM-EIV), fitting the GLM-EIV model (Algorithm \ref{algo:em_full}), and performing inference.
-The three loops in Algorithm \ref{algo:at_scale} are embarassingly parallel and therefore can be massively parallelized.
-
-% \subsection{Zero-inflated model}
-%The GLM-EIV model assumes that unperturbed cells exhibit a nonzero ``background read'' gRNA count distribution due to errors in the sequencing and alignment processes.
-This assumption very likely is appropriate for most current single-cell CRISPR screen datasets \cite{Replogle2020}.
-However, background contamination may be negligible for some datasets, in which case the gRNA counts of unperturbed cells should be modeled as point masses at zero.
-We introduce such an extension -- which we call the ``zero-inflated GLM-EIV model'' -- in Appendix \ref{sec:zero_inf_model}.
-We additionally derive efficient methods for estimation and inference in the zero-inflated model.
-Given the ubiquity of background contamination in current single-cell CRISPR screen datasets, we focus our attention on the standard GLM-EIV model for the remainder of the main text.
+(These numbers are based on exploratory applications of the method to real and simulated data.) We devised a strategy to produce a highly accurate pilot estimate of the true parameters, enabling us to run the algorithm once and converge upon the MLE within a few iterations.
+The strategy involves layering on top of one another several statistical tricks (that we expect to be independently useful for accelerating other single-cell methods).
+Details are deferred to Appendix \ref{sec:statistical_accelerations}.
+Overall, the statistical accelerations reduce the number of GLMs that we must fit to $<10$ in most cases.
+
+Next, we developed a computational infrastructure to apply GLM-EIV to large-scale, single-cell CRISPR screen data.
+The infrastructure leverages \texttt{Nextflow}, a programming language that facilitates building data-intensive pipelines \parencite{DITommaso2017}, and \texttt{ondisc}, an \texttt{R} package that we developed (in a separate project) to facilitate large-scale computing on single-cell data.
+\texttt{Nextflow} and \texttt{ondisc} together enable the construction of highly portable single-cell pipelines: one can analyze data \textit{out-of-memory} on a laptop or in a \textit{distributed} fashion across tens or hundreds of nodes on a cloud (e.g., Microsoft Azure, Google Cloud) or high-performance cluster.
+Leveraging these technologies, we constructed a pipeline for deploying GLM-EIV at-scale.
+The pipeline aggressively recycles computation when possible, saving a considerable amount of compute; see Appendix \ref{sec:computing} for details.
+Overall, the statistical accelerations and computational infrastructure together enable us to deploy GLM-EIV to large-scale single-cell CRISPR screen data without much hassle.
+
 \section{Simulation studies}\label{sec:simulation}
@@
 
-First, we observed that GLM-EIV dominated thresholded regression on all statistical metrics: GLM-EIV exhibited lower bias (row 1) and mean squared error (row 2) than thresholded regression; additionally, GLM-EIV had superior confidence interval coverage (row 3) despite having produced generally narrower confidence intervals (row 4).
-Intuitively, GLM-EIV outperformed the thresholding method because (i) GLM-EIV leveraged information from \textit{both} modalities (rather than the gRNA modality alone) to assign perturbation identities to cells, and (ii) GLM-EIV produced soft rather than hard assignments, capturing the inherent uncertainty in whether a perturbation occurred.
-We additionally found that accelerated GLM-EIV performed as well as vanilla GLM-EIV on all statistical metrics (rows 1-4) despite having substantially lower computational cost (bottom row).
-In fact, the execution time of accelerated GLM-EIV was almost within an order of magnitude of that of the thresholding method (bottom row).
-
-Interestingly, thresholded regression exhibited better confidence interval coverage under estimated $\theta$ than under known $\theta$ (row 3).
-Estimating $\theta$ leads to slight inflation bias (i.e., overestimating the true effect size), whereas, as we showed previously, thresholding leads to attenuation bias (i.e., underestimating the true effect size).
-These phenomena partially cancel, yielding less biased estimates.
-GLM-EIV exhibited worse performance under unknown $\theta$ than known $\theta$, likely due to poor $\theta$ estimation.
-We note that GLM-EIV and the thresholding method in principal are compatible with \textit{any} $\theta$ estimation procedure, including those based on more sophisticated techniques, such as regularization \cite{Hafemeister2019}.
-We defer rigorous investigation of the impact of different $\theta$ estimation strategies on these methods to future work.
-
-\section{Data analysis}
-
-Leveraging our computational infrastructure (Section \ref{sec:computing}), we applied GLM-EIV and the thresholding method to analyze the entire Gasperini and Xie datasets.
-(We report only the most important aspects of the analysis and results in the main text; full details are available in Appendix \ref{sec:data_analysis_details}.) We set the threshold to the approximate Bayes-optimal decision boundary, as our theoretical analyses and simulation studies revealed that the Bayes-optimal decision boundary is a good choice for the threshold when the gRNA count distribution is well-separated.
-Operating under the assumption that the effect of the perturbation on gRNA expression is similar across pairs, we leveraged the fitted GLM-EIV models to approximate the Bayes boundary in the following way: we (i) sampled several hundred gene-perturbation pairs, (ii) extracted the fitted values $\hat{\beta}_g$ and $\hat{\pi}$ from the GLM-EIV models fitted to these pairs, (iii) computed the median $\overline{\hat{\beta}_g}$ and $\overline{\hat{\pi}}$ across the $\hat{\beta}_g$s and $\hat{\pi}$s, and (iv) used $\overline{\hat{\beta}_g}$ and $\overline{\hat{\pi}}$ to estimate a dataset-wide Bayes-optimal decision boundary.
-We repeated this procedure on both datasets, yielding a threshold of $3$ for Gasperini and $7$ for Xie.
-These thresholds were close to the original thresholds, which were selected in a more heuristic way.
-
-We compared GLM-EIV to thresholded regression on the real data, focusing specifically on the negative control pairs (i.e., gene-perturbation pairs that, by design, are expected to exhibit a fold change of $1$, or no association).
-We found that GLM-EIV and the thresholding method produced similar results (Figure \ref{fig:real_data}a-b): estimates, CI coverage rates, and CI widths were concordant.
-The estimated effect of the perturbation on gene expression $\exp(\hat{\beta}_1^g)$ was unexpectedly large: the 95\% CI for this parameter was $[4306, 5186]$ and $[300, 316]$ on the Gasperini and Xie data, respectively.
-We reasoned that the datasets lay in an ``easy'' region of the parameter space, making thresholding a tenable strategy (provided the threshold is selected well).
-However, this was not obvious \textit{a priori} and may not be the case for other datasets.
-We note that GLM-EIV produced outlier estimates (likely due to non-global EM convergence) on a small ($< 2.5 \%$ on Gasperini, $<0.05\%$ on Xie) number of pairs consisting of a handful of genes (not plotted).
-
-We artificially increased the difficulty of the perturbation assignment problem by generating partially-synthetic datasets.
-First, for a given pair, we sampled gRNA counts directly from the fitted GLM-EIV model.
-Next, to simulate elevated background contamination, we sampled gRNA counts from a slightly modified version of the fitted model in which we increased the mean gRNA expression of \textit{unperturbed} while holding constant the mean gRNA expression of \textit{perturbed} cells.
-We defined a parameter called ``excess background contamination'' (normed to take values in $[0,1]$) to quantify the relative distance between the unperturbed and perturbed gRNA count distributions.
-We held fixed the real-data gene expressions, library sizes, covariates, and fitted perturbation probabilities in all settings.
-
-We generated partially-synthetic data in the above manner for each of the 322 positive control pairs in the Gasperini dataset, varying excess background contamination over the interval $[0,0.4].$ We then applied GLM-EIV and the thresholding method to analyze the data.
-We present results on two example pairs (the pair containing gene \textit{LRIF1} and the pair containing gene \textit{NDUFA2}) in Figures \ref{fig:real_data}c-d.
-We observed that the estimate produced by the methods on the raw data (depicted as a horizontal black line) coincided almost exactly with the estimate produced by the methods on the partially-synthetic data generated by setting excess background contamination to zero (This result replicated across nearly all pairs; average relative difference $0.003$.) We additionally observed that as excess background contamination increased, the performance of thresholded regression degraded considerably while that of GLM-EIV remained stable.
-
-We generalized the above analysis to the entire set of positive control pairs.
-First, for each pair we computed the ``relative estimate change'' (REC) as a function of excess background contamination, defined as the relative difference between the estimate at a given level of excess contamination and zero excess contamination (Figure \ref{fig:real_data}d).
-Next, we computed the median REC across all positive control pairs (Figure \ref{fig:real_data}e; upper and lower bands indicate the pointwise interquartile range of the REC).
-As excess background contamination increased, thresholded regression exhibited severe attenuation bias (as reflected by large median REC values); GLM-EIV, by contrast, remained mostly stable.
-Finally, taking the estimate obtained on the raw data as ``ground truth,'' we computed CI coverage across pairs as a function of excess contamination.
-GLM-EIV exhibited significantly higher CI coverage than thresholded regression as the data became increasingly contaminated (Figure \ref{fig:real_data}f; bands indicate $95\%$ pointwise CIs for population coverage).
-
-\section{Discussion}
-
-In this work we introduced GLM-EIV (``GLM-based errors in variables''), a new method for single-cell CRISPR screen analysis.
-GLM-EIV extends the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed and potentially impacted by the same set of confounding variables.
-These extensions enable GLM-EIV to resolve novel analysis challenges posed by single-cell CRISPR screens.
-We demonstrated through simulation studies, real data analyses, and theory that GLM-EIV outperforms thresholded regression by a considerable margin in high background contamination settings.
-GLM-EIV intuitively achieves this performance gain by leveraging information from \textit{both} modalities (rather than the gRNA modality alone) to assign perturbation identities to cells.
-On the other hand, in low background contamination settings GLM-EIV and thresholded regression work best in conjunction, with GLM-EIV providing a statistically principled and empirically effective procedure for selecting the threshold.
-GLM-EIV thereby neutralizes a tuning parameter that, until this point, has been selected using heuristic procedures, with little confidence that the choice is near optimal.
-Figure \ref{fig:dec_tree} summarizes how we anticipate GLM-EIV being used in practice.
-
-To our knowledge this is the first single-cell CRISPR screen paper oriented toward a statistical audience.
-We hope that this work helps to introduce the broader statistics community to an emerging class functional genomics assays that likely will exert a major impact on biology research in the coming years \cite{Przybyla2021}.
-Additionally, this is the first work to leverage the ondisc-Nextflow-cloud/HPC technology stack, a tightly-integrated, user-friendly, and powerful set of computational tools for large-scale single-cell analysis that we expect to be of interest to other researchers.
-
-We anticipate that GLM-EIV could be applied to other types of single-cell CRISPR screen and multimodal single-cell data.
-For example, we likely (with some effort) could extend GLM-EIV to ``low-multiplicity of infection'' screens \cite{Schraivogel2020} in which each cell receives one or two perturbations rather than dozens (as is the case in ``high multiplicity screens,'' studied in this work).
-We also likely could apply GLM-EIV to analyze multimodal single-cell chromatin accessibility assays.
-A question of interest in such experiments is whether chromatin state (i.e., closed or open) is associated with the expression of a gene (or abundance of a protein) \cite{Mimitou2021}.
-We do not directly observe the chromatin state of a cell; instead, we observe tagged DNA fragments that serve as proxies for whether a given region of chromatin is open or closed.
-These tagged DNA fragments come in the form of discrete counts; therefore, GLM-EIV might be applied in such experiments to aid in the selection of thresholds or to analyze whole datasets.
-
-\section{Acknowledgments and code availability}
-We thank Xuran Wang helping to process the Xie dataset.
-We also thank Songcheng Dai for helping to deploy the GLM-EIV pipeline on Azure.
-This work used the Extreme Science and Engineering Discovery Environment (XSEDE; NSF grant ACI-1548562) and the Bridges-2 system (NSF grant ACI-1928147) at the Pittsburgh Supercomputing Center.
-Code for the analyses is available at \texttt{/timothy-barry/glmeiv-manuscript}.
-
-\printbibliography
-
-\clearpage
-\section*{Figures}
-% 1. Challenges
 \begin{figure}[h!]
 	\centering
-	\includegraphics[width=1\linewidth]{../../figures/analysis_challenges/plot.pdf}
-	\caption{\textbf{Experimental design and analysis challenges}: \textbf{a,} Experimental design.
-For a given perturbation (e.g., the perturbation indicated in blue), we partition the cells into two groups: perturbed and unperturbed.
-Next, for a given gene, we conduct a differential expression analysis across the two groups, yielding an estimate of the impact of the given perturbation on the given gene.
-\textbf{b,} DAG representing all variables in the system.
-The perturbation (latent) impacts both gene expression and gRNA expression; technical factors act as confounders, also impacting gene and gRNA expression.
-The target of estimation is the effect of the perturbation on gene expression.
-\textbf{c,} Schematic illustrating the ``background read'' phenomenon.
-Due to errors in the sequencing and alignment processes, unperturbed cells exhibit a nonzero gRNA count distribution (bottom).
-The target of estimation is the change in mean gene expression in response to the perturbation (top).
-\textbf{d}, Example data on four cells for a given perturbation-gene pair.
-Note that (i) the perturbation is unobserved, and (ii) the gene and gRNA data are discrete counts.}
-	\label{analysis_challenges}
-\end{figure}
-
-\clearpage
-% 2. Emprical challenges
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=1\linewidth]{../../figures/thresholding_empirical/plot}
-	\caption{\textbf{Empirical challenges of thresholded regression.} \textbf{a-b,} Fold change estimates produced by threshold = 1 versus threshold = 5 (a) and threshold = 20 versus threshold = 5 (b).
-The selected threshold substantially impacts the results.
-\textbf{c-d,} $p$-values (c) and CI widths (d) produced by threshold = 20 versus threshold = 5.
-The latter threshold yields more confident estimates.
-\textbf{e-f}, Empirical distribution of randomly-selected gRNA from Gasperini (e) and Xie (f) data (0 counts not shown).
-The gRNA data do not appear to imply an obvious threshold selection strategy.}
-	\label{thresholding_empirical}
-\end{figure}
-
-\clearpage
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=0.7\linewidth]{../../figures/thresholding_theoretical/plot}
-	\caption{\textbf{Thresholding method bias-variance decomposition.} Bias decreases and variance increases as the threshold tends to infinity.
-$\beta^g_1 = 1, \beta^m_1 = 1,$ and $\pi = 0.1$ in this plot.}
-	\label{thresholding_theoretical}
-\end{figure}
-
-\clearpage
-% Theory
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=0.9\linewidth]{../../figures/main_text_sim/plot}
+	\includegraphics[width=0.9\linewidth]{figures/main_txt_sim.pdf}
 	\caption{\textbf{Simulation study}.
@@
 
-\clearpage
-\begin{figure}
+First, we observed that GLM-EIV dominated thresholded regression on all statistical metrics: GLM-EIV exhibited lower bias (row 1) and mean squared error (row 2) than thresholded regression; additionally, GLM-EIV had superior confidence interval coverage (row 3) despite having produced generally narrower confidence intervals (row 4).
+Intuitively, GLM-EIV outperformed the thresholding method because (i) GLM-EIV leveraged information from \textit{both} modalities (rather than the gRNA modality alone) to assign perturbation identities to cells, and (ii) GLM-EIV produced soft rather than hard assignments, capturing the inherent uncertainty in whether a perturbation occurred.
+We additionally found that accelerated GLM-EIV performed as well as vanilla GLM-EIV on all statistical metrics (rows 1-4) despite having substantially lower computational cost (bottom row).
+In fact, the execution time of accelerated GLM-EIV was almost within an order of magnitude of that of the thresholding method (bottom row).
+
+Interestingly, thresholded regression exhibited better confidence interval coverage under estimated $\theta$ than under known $\theta$ (row 3).
+Estimating $\theta$ leads to slight inflation bias (i.e., overestimating the true effect size), whereas, as we showed previously, thresholding leads to attenuation bias (i.e., underestimating the true effect size).
+These phenomena partially cancel, yielding less biased estimates.
+GLM-EIV exhibited worse performance under unknown $\theta$ than known $\theta$, likely due to poor $\theta$ estimation.
+We note that GLM-EIV and the thresholding method in principal are compatible with \textit{any} $\theta$ estimation procedure, including those based on more sophisticated techniques, such as regularization \cite{Hafemeister2019}.
+We defer rigorous investigation of the impact of different $\theta$ estimation strategies on these methods to future work.
+
+\section{Data analysis}
+
+Leveraging our computational infrastructure, we applied GLM-EIV and the thresholding method to analyze the entire Gasperini and Xie datasets.
+(We report only the most important aspects of the analysis and results in the main text; full details are available in Appendix \ref{sec:data_analysis_details}.) We set the threshold to the approximate Bayes-optimal decision boundary, as our theoretical analyses and simulation studies revealed that the Bayes-optimal decision boundary is a good choice for the threshold when the gRNA count distribution is well-separated.
+Operating under the assumption that the effect of the perturbation on gRNA expression is similar across pairs, we leveraged the fitted GLM-EIV models to approximate the Bayes boundary in the following way: we (i) sampled several hundred gene-perturbation pairs, (ii) extracted the fitted values $\hat{\beta}_g$ and $\hat{\pi}$ from the GLM-EIV models fitted to these pairs, (iii) computed the median $\overline{\hat{\beta}_g}$ and $\overline{\hat{\pi}}$ across the $\hat{\beta}_g$s and $\hat{\pi}$s, and (iv) used $\overline{\hat{\beta}_g}$ and $\overline{\hat{\pi}}$ to estimate a dataset-wide Bayes-optimal decision boundary.
+We repeated this procedure on both datasets, yielding a threshold of $3$ for Gasperini and $7$ for Xie.
+These thresholds were close to the thresholds used in the original publications, which were selected in a more heuristic way.
+
+We compared GLM-EIV to thresholded regression on the real data, focusing specifically on the negative control pairs (i.e., gene-perturbation pairs that, by design, are expected to exhibit a fold change of $1$, or no association).
+We found that GLM-EIV and the thresholding method produced similar results (Figure \ref{fig:real_data}a-b): estimates, CI coverage rates, and CI widths were concordant.
+CI coverage rates, which ranged from 87.7\%-91.2\%, were slightly below the nominal rate of 95\%, likely due to mild model misspecification.
+The estimated effect of the perturbation on gene expression $\exp(\hat{\beta}_1^g)$ was unexpectedly large: the 95\% CI for this parameter was $[4306, 5186]$ and $[300, 316]$ on the Gasperini and Xie data, respectively.
+We reasoned that the datasets lay in an ``easy'' region of the parameter space, making thresholding a tenable strategy (provided the threshold is selected well).
+However, this was not obvious \textit{a priori} and may not be the case for other datasets.
+We note that GLM-EIV produced outlier estimates (likely due to non-global EM convergence) on a small ($< 2.5 \%$ on Gasperini, $<0.05\%$ on Xie) number of pairs consisting of a handful of genes.
+These outliers were not plotted in Figures \ref{fig:real_data}a-b but were used to compute the CI coverage.
+
+To evaluate performance of GLM-EIV versus thresholding in more challenging scenarios, we artificially increased the difficulty of the perturbation assignment problem by generating partially-synthetic datasets.
+First, for a given pair, we sampled gRNA counts directly from the fitted GLM-EIV model.
+Next, to simulate elevated background contamination, we sampled gRNA counts from a slightly modified version of the fitted model in which we increased the mean gRNA expression of \textit{unperturbed} while holding constant the mean gRNA expression of \textit{perturbed} cells.
+We defined a parameter called ``excess background contamination'' (normed to take values in $[0,1]$) to quantify the relative distance between the unperturbed and perturbed gRNA count distributions.
+We held fixed the real-data gene expressions, library sizes, covariates, and fitted perturbation probabilities in all settings.
+
+\begin{figure}[h!]
 	\centering
-	\includegraphics[width=0.9\linewidth]{../../figures/data_analysis/annotated_plot_crop}
+	\includegraphics[width=0.9\linewidth]{figures/data_analysis.pdf}
 	\caption{\textbf{Data analysis}.
@@
 
-\clearpage
-% Dec tree
-\begin{figure}[h!]
+We generated partially-synthetic data in the above manner for each of the 322 positive control pairs in the Gasperini dataset, varying excess background contamination over the interval $[0,0.4].$ We then applied GLM-EIV and the thresholding method to analyze the data.
+We present results on two example pairs (the pair containing gene \textit{LRIF1} and the pair containing gene \textit{NDUFA2}) in Figures \ref{fig:real_data}c-d.
+We observed that the estimate produced by the methods on the raw data (depicted as a horizontal black line) coincided almost exactly with the estimate produced by the methods on the partially-synthetic data generated by setting excess background contamination to zero (This result replicated across nearly all pairs; average relative difference $0.003$.) We additionally observed that as excess background contamination increased, the performance of thresholded regression degraded considerably while that of GLM-EIV remained stable.
+
+We generalized the above analysis to the entire set of positive control pairs.
+First, for each pair we computed the ``relative estimate change'' (REC) as a function of excess background contamination, defined as the relative difference between the estimate at a given level of excess contamination and zero excess contamination (Figure \ref{fig:real_data}d).
+Next, we computed the median REC across all positive control pairs (Figure \ref{fig:real_data}e; upper and lower bands indicate the pointwise interquartile range of the REC).
+As excess background contamination increased, thresholded regression exhibited severe attenuation bias (as reflected by large median REC values); GLM-EIV, by contrast, remained mostly stable.
+Finally, taking the estimate obtained on the raw data as ``ground truth,'' we computed CI coverage across pairs as a function of excess contamination.
+GLM-EIV exhibited significantly higher CI coverage than thresholded regression as the data became increasingly contaminated (Figure \ref{fig:real_data}f; bands indicate $95\%$ pointwise CIs for population coverage).
+
+\section{Discussion}
+
+In this work we introduced GLM-EIV (``GLM-based errors in variables''), a new method for single-cell CRISPR screen analysis.
+GLM-EIV extends the classical errors-in-variables model to responses and noisy predictors that are exponential family-distributed and potentially impacted by the same set of confounding variables.
+These extensions enable GLM-EIV to resolve novel analysis challenges posed by single-cell CRISPR screens.
+We demonstrated through simulation studies, real data analyses, and theory that GLM-EIV outperforms thresholded regression by a considerable margin in high background contamination settings.
+GLM-EIV intuitively achieves this performance gain by leveraging information from \textit{both} modalities (rather than the gRNA modality alone) to assign perturbation identities to cells.
+On the other hand, in low background contamination settings GLM-EIV and thresholded regression work best in conjunction, with GLM-EIV providing a statistically principled and empirically effective procedure for selecting the threshold.
+GLM-EIV thereby neutralizes a tuning parameter that, until this point, has been selected using heuristic procedures, with little confidence that the choice is near optimal.
+Figure \ref{fig:dec_tree} summarizes how we anticipate GLM-EIV being used in practice.
+
+To our knowledge this is the first single-cell CRISPR screen paper oriented toward a statistical audience.
+We hope that this work helps to introduce the broader statistics community to an emerging class of functional genomics assays that likely will exert a major impact on biological research in the coming years \parencite{Przybyla2021}.
+Additionally, this is the first work to leverage the \texttt{ondisc}-\texttt{Nextflow}-cloud technology stack, a tightly-integrated, user-friendly, and powerful set of tools for large-scale single-cell analysis.
+We expect this technology stack to be of interest to other researchers.
+
+\begin{figure}[h]
 	\centering
-	\includegraphics[width=0.9\linewidth]{../../figures/dec_tree/dec_tree_crop}
+	\includegraphics[width=0.9\linewidth]{figures/dec_tree.pdf}
 	\caption{\textbf{Use of GLM-EIV in practice}.
@@
 \end{figure}
+
+We anticipate that GLM-EIV could be applied to other types of single-cell CRISPR screen and multimodal single-cell data.
+For example, GLM-EIV likely could be extended (with effort) to ``low-multiplicity of infection'' screens  \parencite{Schraivogel2020} in which each cell receives one or two perturbations rather than dozens (as is the case in ``high multiplicity screens,'' studied in this work).
+We also could apply GLM-EIV to analyze multimodal single-cell chromatin accessibility assays.
+A question of interest in such experiments is whether chromatin state (i.e., closed or open) is associated with the expression of a gene (or abundance of a protein) \parencite{Mimitou2021}.
+We do not directly observe the chromatin state of a cell; instead, we observe tagged DNA fragments that serve as count-based proxies for whether a given region of chromatin is open or closed.
+GLM-EIV might be applied in such experiments to aid in the selection of thresholds or to analyze whole datasets.
+
+Finally, GLM-EIV might be extended to areas beyond genomics, such as psychology.
+Many psychological constructs (e.g., presence or absence of a social media addiction) are latent and can be assessed only through an imperfect proxy (e.g., the number of times one has checked social media).
+Researchers might use GLM-EIV to regress an outcome variable (e.g., self-reported well-being) onto the latent construct via the imperfect proxy, potentially resolving challenges related to attenuation bias and threshold selection.
+Applications to psychology and other areas are a topic of future investigation.
+
+\section{Acknowledgments and resource availability}
+We thank Eric Tchetgen Tchetgen for helpful conversations, Xuran Wang for helping to process the Xie dataset, and Songcheng Dai for helping to deploy the GLM-EIV pipeline on Azure.
+This work used the Extreme Science and Engineering Discovery Environment (XSEDE; NSF grant ACI-1548562) and the Bridges-2 system (NSF grant ACI-1928147) at the Pittsburgh Supercomputing Center.
+Data and results are deposited at [LINK].
+Github repositories containing manuscript replication code, the \texttt{glmeiv} R package, and the cloud/HPC-scale GLM-EIV pipeline are available at \url{github.com/timothy-barry/glmeiv-manuscript}, \url{github.com/timothy-barry/glmeiv}, and \url{github.com/timothy-barry/glmeiv-pipeline}, respectively.
+
+\printbibliography
 \clearpage
-
 \begin{appendices}
@@
 			\centering
-			\includegraphics[width=1\linewidth]{../../figures/thresholding_theoretical/app_plot}
+			\includegraphics[width=1\linewidth]{figures/threshold_vs_bias.jpeg}
 			\caption{\textbf{Bias as a function of threshold}.
@@
 		\mathbb{V}[\hat{\beta}_m] \approx \left[\beta_m\omega\pi(\beta_m - 2l) + \mathbb{E}[\hat{p}_i](1 + l^2)\right]/[n\mathbb{E}^2[\hat{p}_i]],$$
-		completing the bias-variance decomposition.
+		completing the bias-variance decomposition. Figure \ref{thresholding_theoretical} plots the bias-variance decomposition as a function of the threshold.
+
+\begin{figure}[h]
+	\centering
+	\includegraphics[width=0.7\linewidth]{figures/bv_decomp.jpeg}
+	\caption{\textbf{Thresholding method bias-variance decomposition.} Bias decreases and variance increases as the threshold tends to infinity.
+$\beta^g_1 = 1, \beta^m_1 = 1,$ and $\pi = 0.1$ in this plot.}
+	\label{thresholding_theoretical}
+\end{figure}
 
@@
 			\centering
-			\includegraphics[width=0.45\linewidth]{../../figures/info_matrix/info_matrix_background}
+			\includegraphics[width=0.45\linewidth]{figures/info_matrix_background}
 			\caption{Block structure of the observed information matrix $J(\theta; m, g) = -\nabla^2 \mathcal{L}(\theta; m, g)$.
@@
 			\centering
-			\includegraphics[width=0.45\linewidth]{../../figures/info_matrix/info_matrix_zero_inf}
+			\includegraphics[width=0.45\linewidth]{figures/info_matrix_zero_inf}
 			\caption{Block structure of the observed information matrix $J_z(\theta; m, g) = -\nabla^2 \mathcal{L}_z(\theta; m, g)$ for the zero-inflated model.
@@
 
-		\section{Statistical accelerations}\label{sec:statistical_accelerations}
-
-		\subsection*{Computing $\pi^\textrm{pilot}$, $[\beta^m_1]^\textrm{pilot}$, $[\beta^m_1]^\textrm{pilot}$}
-
-		We present the algorithm (Algorithm \ref{algo:pilot_estimates_2}) for computing the pilot estimates $\pi^\textrm{pilot}$, $[\beta^m_1]^\textrm{pilot}$, $[\beta^m_1]^\textrm{pilot}$.
-This algorithm, alongside Algorithm \ref{algo:pilot_estimates_1}, is used to compute pilot estimates for use in Algorithm \ref{algo:em_full}.
+\section{Statistical accelerations and computing}\label{sec:statistical_accelerations}
+
+\subsection{Statistical accelerations}
+
+We describe in detail the procedure for obtaining the pilot parameter estimates $(\pi^\textrm{pilot}, \beta_m^\textrm{pilot}, \beta_g^\textrm{pilot})$.
+This procedure consists of two sub-routines, which we label Algorithm \ref{algo:pilot_estimates_1} and Algorithm \ref{algo:pilot_estimates_2}.
+The first step (Algorithm \ref{algo:pilot_estimates_1}) is to obtain good parameter estimates for $[\beta^m_0, \gamma_m]^T$ and $[\beta^g_0, \gamma_g]^T$ via regression.
+Recall that the underlying gene expression parameter vector $\beta_m$ is $\beta_m = [\beta^m_0, \beta^m_1, \gamma_m]^T \in \mathbb{R}^d$, where $\beta^m_0$ is the intercept, $\beta^m_1$ is the effect of the perturbation, and $\gamma_m^T$ is the effect of the technical factors.
+To produce estimates $[\beta^m_0]^\textrm{pilot}$ and $[\gamma_m^T]^\textrm{pilot}$, we regress the gene expressions $m$ onto the technical factors $X$.
+The intuition for this procedure is as follows: the probability of perturbation $\pi$ is very small.
+Therefore, the true log likelihood is approximately equal to the log likelihood that results from omitting $p_i$ from the model:
+\begin{multline*}
+\sum_{i=1}^n f_m(m_i; \eta^m_i) = \underbrace{\sum_{i : p_i =1} f_m(m_i; h_m(\beta_0 + \beta_1 + \gamma^T z_i + o^m_i))}_\textrm{few terms}  + \underbrace{\sum_{i : p_i = 0} f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i))}_\textrm{many terms} \\ \approx \sum_{i = 1}^n f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i)).
+\end{multline*}
+We similarly can obtain pilot estimates $[\beta^g_0]^\textrm{pilot}$ and $[\gamma^T_g]^\textrm{pilot}$ by regressing the gRNA counts $g$ onto the technical factors $X$.
+We extract the fitted values (on the scale of the linear component) for use in a subsequent step: $\hat{f}^k_i = [\beta^k_0]^\textrm{pilot} + \langle [\gamma^T_k]^\textrm{pilot}, z_i \rangle + o^k_i,$ for $k \in \{m,g\}$.
+
+\begin{algorithm}
+	\caption{Computing $[\beta_0^m]^\textrm{pilot}$, $[\gamma^T_m]^\textrm{pilot}$, $[\beta_0^g]^\textrm{pilot}$, and $[\gamma^T_g]^\textrm{pilot}$ .}\label{algo:pilot_estimates_1}
+	\begin{algorithmic}[2]
+		\Require Data $m$, $g$, $o^m$, $o^g$, and $X$; gene expression distribution $f_m$ and link function $r^{-1}_m$; gRNA expression distribution $f_g$ and link function $r^{-1}_g$; number of EM starts $B$.
+		\For {$k \in \{m,g\}$}
+		\State \multiline{Fit a GLM $GLM_k$ with responses $k$, offsets $o^k$, design matrix $X$, distribution $f_k$, and link function $r^{-1}_k$.}
+		\State Set $[\beta_0^k]^\textrm{pilot}$ and $[\gamma_k^T]^\textrm{pilot}$ to the fitted coefficients of $GLM_k$.
+		\For{$i \in \{1, \dots, n\}$}
+		\State $\hat{f}^k_i \gets [\beta_0^k]^\textrm{pilot} + \langle [\gamma_k^T]^\textrm{pilot} , z_i \rangle + o_i^k$ \Comment{untransformed fitted values}
+		\EndFor
+		\EndFor
+		\State \textbf{return} $([\beta_0^m]^\textrm{pilot}$, $\hat{f}^m$, $[\gamma^T_m]^\textrm{pilot}$, $[\beta_0^g]^\textrm{pilot}$, $[\gamma^T_g]^\textrm{pilot}$, $\hat{f}^g)$
+	\end{algorithmic}
+\end{algorithm}
+
+Next, we obtain estimates $[\beta_1^m]^\textrm{pilot},$ $[\beta_1^g]^\textrm{pilot},$ and $\pi^\textrm{pilot}$ for $\beta^m_1$, $\beta^g_1$, and $\pi$ by fitting a ``reduced'' GLM-EIV (Algorithm \ref{algo:pilot_estimates_2}).
+The log likelihood of the no-intercept, univariate GLM with predictor $p_i$ and offset $\hat{f}^m_i$ is approximately equal to the true log likelihood:
+$$ \sum_{i=1}^n f_m(m_i; \eta^m_i) =
+\sum_{i=1}^n f_m(m_i; h_m(\beta_0 + \beta_1 p_i + \gamma^T z_i + o^m_i)) \approx \sum_{i=1}^n f_m(m_i; h_m( \beta_1 p_i + \hat{f}^m_i)).
+$$
+Therefore, to estimate $\beta^m_1$, $\beta^g_1$, and $\pi$, we fit a GLM-EIV model with gene expressions $m$, gRNA counts $g$, gene offsets $\hat{f}^m := [\hat{f}^m_1, \dots, \hat{f}^m_n]^T$, gRNA offsets $\hat{f}^g := [\hat{f}^g_1, \dots, \hat{f}^g_n]^T$, and \textit{no} intercept or covariate terms.
+Intuitively, we ``encode'' all information about technical factors, library sizes, and baseline expression levels into $\hat{f}^m$ and $\hat{f}^g$.
+We run the algorithm $B \approx 15$ times over randomly-selected starting values for $\beta^m$, $\beta^g$, and $\pi$ and select the solution with greatest the log likelihood.
+
+The M step of the reduced GLM-EIV algorithm requires fitting two no-intercept, univariate GLMs with offsets.
+We derive analytic formulas for the MLEs of these GLMs in the three most important cases: Gaussian response with identity link, Poisson response with log link, and negative binomial response with log link (see section \ref{sec:int_plus_offset}; the latter formula is asymptotically exact).
+Consequently, we do not need to run the relatively slow IRLS procedure to carry out the M step of the reduced GLM-EIV algorithm.
+Overall, the proposed method for obtaining the full set of pilot parameter estimates requires fitting only two GLMs (via IRLS).
 
@@
 
-		\subsection*{Intercept-plus-offset models}
+\subsection{Intercept-plus-offset models}\label{sec:int_plus_offset}
 
@@
 		\begin{multline}\label{stat_acc_1}
-		$$\mathcal{L}(\beta_1; m) = \sum_{i=1}^n T_i(0) f_m(m_i; h_m(\beta_1 + \hat{f}^m_i )) + \sum_{i=1}^n T_i(1) f_m(m_i; h_m(\hat{f}^m_i)) \\ = \sum_{i=1}^n T_i(1) f_m(m_i; h_m(\beta_1 + \hat{f}^m_i )) + C.$$
+		\mathcal{L}(\beta_1; m) = \sum_{i=1}^n T_i(0) f_m(m_i; h_m(\beta_1 + \hat{f}^m_i )) + \sum_{i=1}^n T_i(1) f_m(m_i; h_m(\hat{f}^m_i)) \\ = \sum_{i=1}^n T_i(1) f_m(m_i; h_m(\beta_1 + \hat{f}^m_i )) + C.
 		\end{multline}
@@
 
-		Let $\beta \in \mathbb{R}$ be an unknown constant.
+\paragraph{General formulation}	Let $\beta \in \mathbb{R}$ be an unknown constant.
 Let $o_1, \dots, o_n \sim \mathcal{P}_1$, where $\mathcal{P}_1$ is a distribution.
@@
 
-		\subsubsection*{Gaussian}
+		\paragraph{Gaussian}
 		First, consider a Gaussian response distribution and identity link function $r(\mu) = \mu$. The cumulant-generating function $\psi$ is $\psi(\eta) = \eta^2/2$, and so, by (\ref{can_param}),
@@
 
-		\subsubsection*{Poisson}
+		\paragraph{Poisson}
 		Next, consider a Poisson response distribution and log link function $r(\mu) = \log(\mu).$ The cumulant-generating function $\psi$ is $\psi(\eta) = e^\eta.$ Therefore, by (\ref{can_param}),
@@
 
-		\subsubsection*{Negative binomial}
+		\paragraph{Negative binomial}
 		Finally, we consider a negative binomial response distribution (with fixed size parameter $\theta > 0$) and log link function $r(\mu) = \log(\mu)$. The cumulant-generating function $\psi$ is
@@
 
-		\subsection*{Application to GLM-EIV}
+		\paragraph{Application to GLM-EIV}
 		The GLM that we seek to estimate (\ref{stat_acc_1}) is an approximate intercept-plus-offset model: $T_1(1), \dots, T_n(1)$ are the weights $w_1,\dots, w_n$, and $\hat{f}^m_1, \dots, \hat{f}^m_n$ are the offsets $o_1, \dots, o_m$.
@@
 In practice, we find that the intercept-plus-offset model is very good approximation to the GLM (\ref{stat_acc_1}), especially when the number of cells $n$ is large.
-Additionally, we note that the GLM (\ref{stat_acc_1}) is fit as a subroutine of the algorithm for producing pilot parameter estimates (Algorithm \ref{algo:pilot_estimates}).
-The quality of the pilot parameter estimates does not affect the validity of the estimation and inference procedures (Algorithm \ref{algo:em_full}), barring issues related to convergence to local optima (which are outside the scope fo the current work).
+Additionally, we note that the GLM (\ref{stat_acc_1}) is fit as a subroutine of the algorithm for producing pilot parameter estimates (Algorithm \ref{algo:pilot_estimates_2}).
+The quality of the pilot parameter estimates does not affect the validity of the estimation and inference procedures (Algorithm \ref{algo:em_full}), barring issues related to convergence to local optima.
+
+\subsection{Computing}\label{sec:computing}
+
+We describe in detail the at-scale GLM-EIV pipeline.
+First, we run a round of ``precomputations'' on all $d_g$ genes and $d_p$ perturbations.
+The precomputations involve regressing the gene expressions (or gRNA counts) onto the technical factors, thereby ``factoring out'' Algorithm \ref{algo:pilot_estimates_1}.
+Next, we run differential expression analyses on the full set of gene-perturbation pairs; for a given pair, this amounts to obtaining the complete set of pilot parameters (by running a reduced GLM-EIV), fitting the GLM-EIV model (Algorithm \ref{algo:em_full}), and performing inference.
+The three loops in Algorithm \ref{algo:at_scale} are embarassingly parallel and therefore can be massively parallelized.
+
+\begin{algorithm}
+	\caption{Applying GLM-EIV at scale.}\label{algo:at_scale}
+	\begin{algorithmic}
+		\State $G \gets \{\textrm{gene}_1, \dots, \textrm{gene}_{d_g}\}; P \gets \{\textrm{perturbation}_1, \dots, \textrm{perturbation}_{d_p}\}$
+
+		\For{gene $\in G$}
+		\State Run precomputation (Algorithm \ref{algo:pilot_estimates_1}) on gene; save $\hat{f}^m$, $[\beta^m_0]^\textrm{pilot}$ and $[\gamma^T_m]^\textrm{pilot}$.
+		\EndFor
+		\For{perturbation $\in P$}
+		\State Run precomputation  (Algorithm \ref{algo:pilot_estimates_1}) on perturbation; save $\hat{f}^g$, $[\beta^g_0]^\textrm{pilot}$ and $[\gamma^T_g]^\textrm{pilot}$.
+		\EndFor
+		\For{(gene, perturbation) $\in G \times P$}
+		\State Load $\hat{f}^m, \hat{f}^g,$ $[\beta^m_0]^\textrm{pilot}$ $[\gamma^T_m]^\textrm{pilot}$, $[\beta^g_0]^\textrm{pilot}$ and $[\gamma^T_g]^\textrm{pilot}$.
+		\State Compute $[\beta^m_1]^\textrm{pilot}, [\beta^g_1]^\textrm{pilot}, \pi^\textrm{pilot}$ by fitting a reduced GLM-EIV (Algorithm \ref{algo:pilot_estimates_2}).
+		\State Run GLM-EIV using the pilot parameters (Algorithm \ref{algo:em_full}).
+		\EndFor
+	\end{algorithmic}
+\end{algorithm}
 
@@
 			\centering
-			\includegraphics[width=1.0\linewidth]{../../figures/supplement_sim/gaussian}
+			\includegraphics[width=1.0\linewidth]{figures/sup_sim.pdf}
 			\caption{Additional simulation results on Gaussian data.
@@
 CI coverage (in the sense described here) for excess background contamination equal to $0$ by definitionis $1$, as depicted in Figure \ref{fig:real_data}f.
-
 \printbibliography
+
 \end{refsection}

### 2022-01-04 text/GLM-EIV/jasa_v2.tex
fix reproducibility

--- 2022-01-04 text/GLM-EIV/jasa.tex
@@
 GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding.
-We implement several statistical accelerations (that possibly are of independent utility) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method.
+We implement several statistical accelerations (that likely are of independent utility) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method.
 
@@
 
-\section{Background and analysis challenges}
 \iffalse
-\textcolor{red}{=====================================}
-\textcolor{blue}{I don't think this section works as is.
-I think the whole section should be relocated and revised, but let me say what you might do, if you keep it here.
-First, as much as I like the eSVD paragraph, I would argue that we aren't the only ones using negative binomial models for CRISPR screen analysis.
-For one, SCEPTRE did!
-So quite honestly I don't think this paragraph is appropriate (but see note below).
-With a heading like "Related work" people might expect to know what is being done in practice.
-Your SCEPTRE paragraph is the place to start, but I think we need to expand the lead sentence.
-Cite them all and then give a bit about at least one method available in the literature (very very little detail).
-I'd pick the Monacle based concept and just go with the idea that it is a negative binomial model that uses thresholding and that it sometimes doesn't work and results in a lot of false signals.
-Next you can the remainder of your SCEPTRE paragraph, but again I think it makes sense to say something about the basic model from which the CRT is built.
-In other words you want to convey that his method still uses thresholds, but solves the false positive problem.
-Now you are ready for your final few sentences about how you'll move beyond this.
-NOTE: Regarding your eSVD paragraph, you could use something like this paragraph in the Discussion, where you could make the point that it is a good thing when models are built on the generating distribution of the data (rather than forcing a normal by transformation).
-Then you could say this paper is like the others you cite in doing this.
-I make a try below, see the blue paragraph.\\
-Here is the alternative idea.
-Since my suggestion causes this section to be only one paragraph, you might consider having Assay Overview first and then putting this material at the lead of analysis challenges without a major heading.
-If you moved it to there you could do a better job of citing other works for analysis of CRISPR screens.
-I am a fan of giving credit to everyone, but doing so with terse sentences that say something like "Several methods have appeared in the literature, each with some shortcomings (ref, ref, ref).".
-Anyway, this section could lead people to believe we have very big egos, basically citing ourselves and almost no one else.}
-\textcolor{blue}{THIS IS HOW YOU MIGHT REWRITE THE SCEPTRE PARAGRAPH.The genomics literature has produced a few applied methods for single-cell CRISPR screen analysis, each with some shortcomings (cite Gasp, Xie, etc + Barry).
-The idea is to test whether a given perturbation is associated with the change in expression of a given gene, while ideally adjusting for sources of confounding and ensuring robustness to expression model misspecification.
-Most relevant is SCEPTRE, a custom implementation of the conditional randomization test \parencite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data \parencite{Barry2021}.
-SCEPTRE successfully handles the challenges of poor calibration and low power shared by other methods.
-In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE.
-Most importantly,  we account for the fact that the perturbation is measured with noise, an issue that available methods circumvent via thresholding.
-Additionally, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is challenging to attain within the nonparametric hypothesis testing framework of SCEPTRE.}
-\fi
-
 \subsection{Related work}
@@
 Additionally, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is challenging to attain within the nonparametric hypothesis testing framework of SCEPTRE.
-
-\subsection{Assay background}
+\fi
+
+\section{Assay background}
 There are several broad classes of single-cell CRISPR screen assays, each suited to answer a different set of biological questions \parencite{Gasperini2019,Datlinger2021,Mimitou2019}.
 In this work we focus on high-multiplicity of infection (MOI) single-cell CRISPR screens, which we motivate and describe here.
-The human genome consists of genes, enhancers (segments of DNA that regulate the expression of one or more genes), and other genomic elements.
-Genome-wide association studies (GWAS) have revealed that the majority ($>90\%$) of variants associated with diseases lie outside genes and (very likely) inside enhancers \parencite{Gallagher2018}.
+The human genome consists of genes, enhancers (segments of DNA that regulate the expression of one or more genes), and other genomic elements (that are not of relevance to the current work).
+Genome-wide association studies (GWAS) have revealed that the majority ($>90\%$) of variants associated with diseases lie outside genes and inside enhancers \parencite{Gallagher2018}.
 These noncoding variants are thought to contribute to disease by modulating the expression of one or more disease-relevant genes.
@@
 The enhancer-by-enhancer approach is valid because the perturbations assort into cells approximately independently of one another.
+
+The genomics literature has produced a few applied methods for single-cell CRISPR screen analysis \parencite{Gasperini2019,Xie2019a,Barry2021}.
+Gasperini et al.\ applied negative binomial GLMs (as implemented in the Monocle software; \cite{Trapnell2014}) to carry out the differential expression analysis described above.
+Xie et al., by contrast, applied chi-squared-like tests of independence for this purpose.
+Both of these approaches have limitations: the former is not robust to misspecification of the gene expression model, and the latter is unable to correct for the presence of technical confounders.
+Recently, Barry et al.\ introduced SCEPTRE, a custom implementation of the conditional randomization test \parencite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data \parencite{Barry2021}.
+SCEPTRE simultaneously adjusts for confounder presence and ensures robustness to expression model misspecification, overcoming limitations of the prior methods and demonstrating state-of-the-art sensitivity and specificity on single-cell CRISPR screen data.
+In this work we tackle a set of analysis challenges that are complimentary to those addressed by SCEPTRE.
+Most importantly, we seek to account for the fact that the perturbation is measured with noise, an issue that all available methods (including SCEPTRE) circumvent via thresholding.
+Additionally, we seek to \textit{estimate} (with confidence) the effect size of a perturbation on gene expression change, an objective that is challenging to attain within the nonparametric hypothesis testing framework of SCEPTRE.
 
@@
 
+\section{Analysis challenges and proposed statistical model}
+
 High MOI single-cell CRISPR screens present several statistical challenges, four of which we highlight here.
@@
 
+We propose to model the single-cell CRISPR screen data-generating process using a pair of GLMs.
+Let $n \in \mathbb{N}$ be the number of cells assayed in the experiment.
+Consider a single perturbation and a single gene.
+For cell $i \in \{1, \dots, n\}$, let $p_i \in \{0,1\}$ indicate perturbation presence or absence; let $m_i \in \mathbb{N}$ be the number of gene transcripts sequenced; let $g_i \in \mathbb{N}$ be the number of gRNA transcripts sequenced; let $d^m_i \in \mathbb{N}$ be the number of gene transcripts sequenced across \textit{all} genes (i.e., the library size or sequencing depth); let $d^g_i$ be the gRNA library size; and finally, let $z_i \in \mathbb{R}^{d-1}$ be the cell-specific technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.) The letters ``m,'' ``g'', and ``d'' stand for ``mRNA,'' ``gRNA,'' and ``depth,'' respectively.
+
+Building on the work of several previous authors \parencite{Townes2019,Svensson2020,Hafemeister2019}, Sarkar and Stephens proposed a simple strategy for modeling single-cell gene expression data, which, in the framework of negative binomial GLMs, is equivalent to using the log-transformed library size as an offset term (\cite{Sarkar2021}).
+Sarkar and Stephens' framework enjoys strong theoretical and empirical support; therefore, we generalize their approach to model \textit{both} gene and gRNA modalities in single-cell CRISPR screen experiments.
+To this end, we assume that the gene expression counts are given by
+\begin{equation}\label{glmeiv_model_1}
+m_i |(p_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i^m); \quad \log(\mu^m_i) = \beta^m_0 + \beta^m_1 p_i + \gamma_m^T z_i + \log(d^m_i),
+\end{equation}
+where (i) $\textrm{NB}_{\theta^m}(\mu^m_i)$ is a negative binomial distribution with mean $\mu^m_i$ and known size parameter $\theta^m$; (ii) $\beta^m_0 \in \mathbb{R}, \beta^m_1 \in \mathbb{R},$ and $\gamma_m \in  \mathbb{R}^{d-1}$ are unknown parameters; and (iii) $\log(d_i^m)$ is an offset term.
+Similarly, we model the gRNA counts by
+\begin{equation}\label{glmeiv_model_2}
+g_i | (p_i, z_i, d^g_i) \sim \textrm{NB}_{\theta^g}\left(\mu_i^g\right); \quad \log(\mu_i^g) = \beta^g_0 + \beta^g_1p_i + \gamma^T_g z_i + \log(d^g_i),
+\end{equation}
+where $\mu^g_i$, $\theta^g$, $\beta^g_0$, $\beta^g_1$, $\gamma_g$, and $d^g_i$ are analogous.
+We use a negative binomial GLM to model the gRNA counts as well as the gene expressions because the gRNA transcripts are generated via the same biological mechanism as the gene transcripts \parencite{Datlinger2017,Hill2018}.
+Finally, we model the marginal perturbation probability as
+\begin{equation}\label{glmeiv_model_3}
+ p_i \sim \textrm{Bern}(\pi); \quad \pi \in (0,1/2],
+\end{equation}
+where $p_i$ is unobserved. Together, (\ref{glmeiv_model_1} - \ref{glmeiv_model_3}) define the negative binomial GLM-EIV model.
+
+The log-transformed sequencing depth $\log(d^m_i)$ is included as an offset term in (\ref{glmeiv_model_1}) so that $\beta^m_0 + \beta^m_1 p_i + \gamma^T_m z_i$ can be interpreted as a relative expression.
+Exponentiating both sides of (\ref{glmeiv_model_1}) reveals that the mean gene expression $\mu_i^m$ of the $i$th cell is
+$\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m.$ Because $d^m_i$ is the sequencing depth, $\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right)$ is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration.
+The target of inference $\beta^m_1$ is the log fold change in expression in response to the perturbation, controlling for the technical factors.
+Fold change in this context is the ratio of the mean gene expression in perturbed cells to the mean gene expression in unperturbed cells.
+Hence, $\exp(\beta^m_1) = 1$ (i.e., $\beta^m_1 = 0$) indicates no change in expression, whereas $\exp(\beta^m_1) > 1$ (i.e., $\beta^m_1 > 0$) and $\exp(\beta^m_1) < 1$ (i.e., $\beta^m_1 < 0$) indicate an increase and decrease in expression, respectively.
+
 In this work we analyze two large-scale, high MOI, single-cell CRISPR screen datasets published by \parencite{Gasperini2019} and \parencite{Xie2019a}.
@@
 
-\section{Thresholding method}
+\section{Analysis of the thresholding method}
+
 We study thresholding from empirical and theoretical perspectives, highlighting several limitations of the approach.
-Gasperini and Xie imputed perturbation assignments onto the cells via thresholding, but they carried out the subsequent differential expression analysis in different ways: Gasperini fitted negative binomial regression models to the data, whereas Xie applied nonparametric tests of independence.
-We study Gasperini's variant of the thresholding method, as it relates more closely to GLM-EIV.
-
-Let $n \in \mathbb{N}$ be the number of cells assayed in the experiment.
-Consider a single perturbation and a single gene.
-For cell $i \in \{1, \dots, n\}$, let $m_i \in \mathbb{N}$ be the number of gene transcripts sequenced; let $g_i \in \mathbb{N}$ be the number of gRNA transcripts sequenced; let $d^m_i \in \mathbb{N}$ be the number of gene transcripts sequenced across \textit{all} genes (i.e., the library size or sequencing depth); and finally, let $z_i \in \mathbb{R}^{d-1}$ be the cell-specific technical factors (e.g., sequencing batch, percent mitochondrial reads, etc.) The letters ``m,'' ``g'', and ``d'' stand for ``mRNA,'' ``gRNA,'' and ``depth,'' respectively.
-The thresholding method is defined as follows:
+In the context of the negative bionomial GLM-EIV model introduced above (\ref{glmeiv_model_1}-\ref{glmeiv_model_3}), thresholding methods use the gRNA counts (\ref{glmeiv_model_2})
+to impute the latent perturbation indicator (\ref{glmeiv_model_3}), thereby reducing the full data generating process to a single, gene expression model (\ref{glmeiv_model_1}).
+We study Gasperini et al.'s  variant of the thresholding method (i.e., thresholded negative binomial regression), as this version of the thresholding method relates most closely to GLM-EIV.
+The method is defined as follows:
+
 \begin{itemize}
@@
 	\begin{equation}\label{thresh_glm}
-	m_i | (\hat{p}_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i); \quad
-	\log(\mu_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left(d_i^m\right),
-	\end{equation}
-	where (i) $NB_{\theta^m}(\mu_i)$ is a negative binomial distribution with mean $\mu_i$ and known size parameter $\theta^m$; (ii) $\beta^m_0 \in \mathbb{R}, \beta^m_1 \in \mathbb{R},$ and $\gamma_m \in  \mathbb{R}^{d-1}$ are unknown parameters; and (iii) $\log(d_i^m)$ is an offset term.
-
-	\item[3.] Fit a GLM to obtain an estimate, CI, and $p$-value for the target of inference $\beta^m_1$.
+	m_i | (\hat{p}_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu^m_i); \quad
+	\log(\mu^m_i) = \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i + \log\left(d_i^m\right).
+	\end{equation}
+	The model (\ref{thresh_glm}) is equivalent to the model (\ref{glmeiv_model_2}), but the latent perturbation indicator $p_i$ has been replaced by the imputed perturbation indicator $\hat{p}_i.$
+
+	\item[3.] Fit a GLM to (\ref{thresh_glm}) to obtain an estimate and CI for the target of inference $\beta^m_1$.
 \end{itemize}
-The sequencing depth $d^m_i$ is included as an offset term in (\ref{thresh_glm}) so that $\beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i$ can be interpreted as a relative expression.
-Exponentiating both sides of (\ref{thresh_glm}) shows that the mean $\mu_i$ of the $i$th cell is equal to
-$\exp \left( \beta^m_0 + \beta^m_1 \hat{p}_i + \gamma^T_m z_i \right) d_i^m;$ this quantity is the \textit{fraction} of all transcripts sequenced in the cell produced by the gene under consideration.
-The target of inference $\beta^m_1$ is the log fold change in expression in response to the perturbation, controlling for the technical factors.
-Fold change in this context is defined as the ratio of the mean gene expression in perturbed cells to the mean gene expression in unperturbed cells.
-Therefore, $\exp(\beta^m_1) = 1$ (i.e., $\beta^m_1 = 0$) indicates no change in expression, whereas $\exp(\beta^m_1) > 1$ (i.e., $\beta^m_1 > 0$) and $\exp(\beta^m_1) < 1$ (i.e., $\beta^m_1 < 0$) indicate an increase and decrease in expression, respectively.
 
 \subsection{Empirical challenges of the thresholding method}\label{sec:thresholding_empirical}
-To shed light on empirical challenges of the thresholding method, we applied the thresholding method to analyze the set of positive control perturbation-gene pairs in the Gasperini dataset.
+To shed light on empirical challenges of the thresholding method, we applied thresholded negative binomial regression to analyze the set of positive control perturbation-gene pairs in the Gasperini dataset.
 The positive control pairs consisted of perturbations that targeted gene transcription start sites (TSSs) for inhibition.
 Repressing the TSS of a given gene decreases its expression; therefore, the positive control pairs \textit{a priori} are expected to exhibit a strong decrease in expression.
-To investigate sensitivity to the choice of threshold, we deployed the thresholding method using three different thresholds: 1, 5, and 20.
+
+To investigate the sensitivity of the thresholding method to threshold choice, we deployed the method using three different choices for the threshold: 1, 5, and 20.
 We found that the chosen threshold substantially impacted the results (Figure \ref{thresholding_empirical}a-b): estimates for fold change produced by threshold = 1 were smaller in magnitude (i.e., closer to the baseline of $1$) than those produced by threshold = 5.
 (Figure \ref{thresholding_empirical}a.) On the other hand, estimates produced by threshold = 5 and threshold = 20 were more concordant (Figure \ref{thresholding_empirical}b).
-
-We reasoned that thresholded regression systematically underestimated true effect sizes on the positive control pairs, especially for threshold = 1.
-For a given perturbation, the majority ($>98\%$) of cells are unperturbed.
-This imbalance leads to an asymmetry: misclassifying \textit{unperturbed} cells as \textit{perturbed} is intuitively ``worse'' than misclassifying \textit{perturbed} cells as \textit{unperturbed}.
-Misclassified unperturbed cells contaminate the set of truly perturbed cells, leading to attenuation bias; by contrast, misclassified perturbed cells are swamped in number and ``neutralized'' by the truly unperturbed cells.
-Setting the threshold to a large number reduces the unperturbed-to-perturbed misclassification rate, decreasing bias.
-
-We hypothesized, however, that the reduction in bias obtained by selecting a large threshold causes the variance of the estimator to increase.
-To investigate, we compared $p$-values and confidence intervals produced by threshold = 5 and threshold = 20 for the target of inference $\beta^m_1$.
-We found that threshold = 5 yielded smaller (i.e., more significant) $p$-values and narrower confidence intervals than did threshold = 20 (Figures \ref{thresholding_empirical}c-d).
-We concluded that the threshold controls a bias-variance tradeoff: as the threshold increases, the bias of the estimator decreases and the variance increases.
 
@@
 \end{figure}
+
+We reasoned that thresholded regression systematically underestimated true effect sizes on the positive control pairs, especially for threshold = 1.
+For a given perturbation, the majority ($>98\%$) of cells are unperturbed.
+This imbalance leads to an asymmetry: misclassifying \textit{unperturbed} cells as \textit{perturbed} is intuitively ``worse'' than misclassifying \textit{perturbed} cells as \textit{unperturbed}.
+Misclassified unperturbed cells contaminate the set of truly perturbed cells, leading to attenuation bias; by contrast, misclassified perturbed cells are swamped in number and ``neutralized'' by the truly unperturbed cells.
+Setting the threshold to a large number reduces the unperturbed-to-perturbed misclassification rate, decreasing bias.
+
+We hypothesized, however, that the reduction in bias obtained by selecting a large threshold causes the variance of the estimator to increase.
+To investigate, we compared $p$-values and confidence intervals produced by threshold = 5 and threshold = 20 for the target of inference $\beta^m_1$.
+We found that threshold = 5 yielded smaller (i.e., more significant) $p$-values and narrower confidence intervals than did threshold = 20 (Figures \ref{thresholding_empirical}c-d).
+We concluded that the threshold controls a bias-variance tradeoff: as the threshold increases, the bias of the estimator decreases and the variance increases.
 
@@
 We fix $\pi$ to $1/2$ for simplicity.
-(In reality, $\pi$ is smaller, but the relevant statistical phenomena emerge for $\pi = 1/2$.) First, the thresholding estimator strictly \textit{underestimates} (in absolute value) the true value of $\beta^m_1$ over all choices of the threshold $c$ and over all values of the regression coefficients $(\beta^m_0, \beta^m_1)$ and $(\beta^g_0, \beta^g_1)$.
+(In reality, $\pi$ is smaller, but the relevant statistical phenomena emerge for $\pi = 1/2$.) First, the thresholding estimator strictly underestimates (in absolute value) the true value of $\beta^m_1$ over all choices of the threshold $c$ and over all values of the regression coefficients $(\beta^m_0, \beta^m_1)$ and $(\beta^g_0, \beta^g_1)$.
 This phenomenon, called attenuation bias, is a common attribute of estimators that ignore measurement in errors-in-variables models \parencite{Stefanski2000a}.
@@
 This proposition yields an asymptotically exact bias-variance decomposition for $\hat{\beta}_m$: as the threshold tends to infinity, the bias decreases and the variance increases (Appendix \ref{sec:appendix_theory}).
-Overall, our empirical and theoretical analyses indicate that thresholded regression poses a basic difficulty: selecting a good threshold is challenging, and even if the threshold is optimal, the estimator could suffer from high bias or high variance.
-These considerations motivate our research question: does modeling the gRNA counts directly, thereby circumventing thresholding, facilitate estimation and inference in single-cell CRISPR screens?
+Overall, our empirical and theoretical analyses indicate that thresholded regression poses a basic difficulty: selecting a good threshold is challenging, and even if we have selected a threshold that is (in some sense) optimal, the resulting estimator still may suffer from high bias or high variance.
+These considerations motivate our research question: does modeling the gRNA counts directly, thereby circumventing thresholding, facilitate estimation and inference in single-cell CRISPR screen analysis?
 
@@
 
-To answer this question, we introduce the GLM-EIV model.
+To answer this question, we introduce the general GLM-EIV model, which generalizes the negative binomial GLM-EIV model (\ref{glmeiv_model_1}-\ref{glmeiv_model_3}) to arbitrary exponential family response distributions and link functions, thereby providing much greater modeling flexibility.
+We derive efficient methods for estimation and inference in this model and develop a pipeline to deploy the model at-scale.
+Appendix \ref{sec:zero_inf_model} develops a parallel methodology in which we model gRNA counts as (in some sense) zero-inflated.
 
 \subsection{Model}\label{sec:model}
-Building on the work of several previous authors \parencite{Townes2019,Svensson2020,Hafemeister2019}, Sarkar and Stephens proposed a simple strategy for modeling single-cell gene expression data, which, in the framework of negative binomial GLMs, is equivalent to using the log-transformed library size as an offset term (as in (\ref{thresh_glm}); \cite{Sarkar2021}).
-We generalize Sarkar and Stephens' approach to model \textit{both} gene and gRNA modalities.
-To this end, let the latent variable $p_i \in \{0,1\}$ indicate whether cell $i \in \{1, \dots, n\}$ was perturbed.
-We model the gene expression counts according to
-\begin{equation}\label{glmeiv_model_1}
-m_i |(p_i, z_i, d^m_i) \sim \textrm{NB}_{\theta^m}(\mu_i^m); \quad \log(\mu^m_i) = \beta^m_0 + \beta^m_1 p_i + \gamma_m^T z_i + \log(d^m_i),
-\end{equation}
-where $\theta^m > 0$ is a known negative binomial size parameter, and $\beta^m_0 \in \mathbb{R}, \beta^m_1 \in \mathbb{R},$ and $\gamma_m \in \mathbb{R}^{d - 2}$ are unknown constants.
-The model (\ref{glmeiv_model_1}) is identical to the thresholding model (\ref{thresh_glm}), but the imputed perturbation indicator $\hat{p}_i$ is replaced by the latent perturbation indicator $p_i$.
-Next, let $d^g_i \in \mathbb{N}$ be the number of gRNA transcripts sequenced across \textit{all} gRNAs in cell $i$ (i.e., the gRNA library size).
-The model for the gRNA counts is
-\begin{equation}\label{glmeiv_model_2}
-g_i | (p_i, z_i, d^g_i) \sim \textrm{NB}_{\theta^g}\left(\mu_i^g\right); \quad \log(\mu_i^g) = \beta^g_0 + \beta^g_1p_i + \gamma^T_g z_i + \log(d^g_i),
-\end{equation}
-where, similar to above, $\theta^g > 0$ is a known negative binomial size parameter, and $\beta^g_0 \in \mathbb{R}, \beta^g_1 \in \mathbb{R}, \gamma_g \in \mathbb{R}^{d - 2}$ are unknown constants.
-We use a negative binomial GLM to model the gRNA counts because the gRNA molecules are transcribed in the cell in the same way as gene transcripts \parencite{Datlinger2017,Hill2018}.
-Finally, we model the marginal perturbation probability as $
-p_i \sim \textrm{Bern}(\pi),$ where $\pi \in (0,1/2]$. 
-Together, (\ref{glmeiv_model_1}), (\ref{glmeiv_model_2}), and the marginal distribution of $p_i$ define the negative binomial GLM-EIV model.
-The terms $(\beta^m_0 + \beta^m_1p_i + \gamma^T_m z_i)$ and $(\beta^g_0 + \beta^g_1 p_i + \gamma^T_g z_i)$ can be interpreted as relative gene and gRNA expressions, similar to the analogous term in the thresholding model.
-Likewise, the target of inference $\beta^m_1$ is the log fold change in gene expression in response to the perturbation, accounting for technical factors.
-See Appendix \ref{sec:zero_inf_model} for a parallel methodology in which we instead model the gRNA distribution using a zero-inflated GLM.
-
-\paragraph{General model} To provide greater modeling flexibility, we generalize the GLM-EIV model to arbitrary exponential family response distributions and link functions.
-To increase notational compactness, let $\tilde{x_i} = [1, p_i, z_i]^T \in \mathbb{R}^d$ be the vector of covariates (including an intercept term) for the $i$th cell.
+
+\paragraph{General model} This section is more technical than the previous ones; a first-time reader can skip to Section \ref{sec:estimation_inference} without loss of information relevant to the narrative.
+Let $\tilde{x_i} = [1, p_i, z_i]^T \in \mathbb{R}^d$ be the vector of covariates (including an intercept term) for the $i$th cell.
 (We use the tilde as a reminder that the vector is partially unobserved.) Let $\beta_m = [\beta^m_0, \beta^m_1, \gamma_m]^T \in \mathbb{R}^d$ and $\beta_g = [\beta^g_0, \beta^g_1, \gamma_g]^T \in \mathbb{R}^d$ be the unknown coefficient vectors corresponding to the gene and gRNA expression models, respectively.
@@
 
-We use GLMs to model the gene and gRNA expressions.
-Considering first the gene expression model, let the $i$th linear component $l^m_i$ of the model be $l^m_i = \langle \tilde{x}_i, \beta_m \rangle + o^m_i.$ Next, let the mean $\mu^m_i$ of the $i$th observation be $r_m(\mu^m_i) = l^m_i,$ where $r_m:\mathbb{R} \to \mathbb{R}$ is a strictly increasing, differentiable link function.
+We use a pair of GLMs to model the gene and gRNA expressions.
+Considering first the gene expression model, let the $i$th linear component $l^m_i$ of the model be $l^m_i \equiv \langle \tilde{x}_i, \beta_m \rangle + o^m_i.$ Next, let the mean $\mu^m_i$ of the $i$th observation be $r_m(\mu^m_i) \equiv l^m_i,$ where $r_m:\mathbb{R} \to \mathbb{R}$ is a strictly increasing, differentiable link function.
 Let $\psi_m: \mathbb{R} \to \mathbb{R}$ be the differentiable, cumulant-generating function of the selected exponential family distribution.
 We can express the canonical parameter $\eta^m_i$ in terms of $\psi_m$ and $r_m$ by
-$\eta^m_i = \left([\psi'_m]^{-1} \circ r^{-1}_m\right)(l_i^m) := h_m(l_i^m).$ Finally, let $c_m: \mathbb{R} \to \mathbb{R}$ be the carrying density of the selected exponential family distribution.
+$\eta^m_i = \left([\psi'_m]^{-1} \circ r^{-1}_m\right)(l_i^m) \equiv h_m(l_i^m).$ Finally, let $c_m: \mathbb{R} \to \mathbb{R}$ be the carrying density of the selected exponential family distribution.
 The density $f_m$ of $m_i$ conditional on the the canonical parameter $\eta_i$ is
@@
 Let the terms $l^g_i, o^g_i, \mu^g_i, \eta^g_i, \psi_g, r_g, h_g$ and $c_g$ be defined in an analogous way for the gRNA model, i.e.
-$l^g_i = \langle \tilde{x}_i, \beta_g \rangle + o^g_i$, $r_g(\mu^g_i) = l^g_i$, and $\eta^g_i = \left([\psi'_g]^{-1} \circ r^{-1}_g\right)(l_i^g) = h_g(l_i^g).$ The density $f_g$ of $g_i$ given the canonical parameter is $f_g(m_i; \eta^g_i) = \exp\left\{g_i \eta^g_i - \psi_g(\eta^g_i) + c_g(g_i)\right\}.$
+$l^g_i \equiv \langle \tilde{x}_i, \beta_g \rangle + o^g_i$, $r_g(\mu^g_i) \equiv l^g_i$, and $\eta^g_i = \left([\psi'_g]^{-1} \circ r^{-1}_g\right)(l_i^g) \equiv h_g(l_i^g).$ The density $f_g$ of $g_i$ given the canonical parameter is $f_g(m_i; \eta^g_i) = \exp\left\{g_i \eta^g_i - \psi_g(\eta^g_i) + c_g(g_i)\right\}.$
 Finally, the unobserved variable $p_i$ is assumed to follow a Bernoulli distribution with mean $\pi \in (0, 1/2]$. Its marginal density $f_p$ is given by $f_p(p_i) = \pi^{p_i}(1-\pi)^{1 - p_i}.$
@@
 \paragraph{Notation} We briefly introduce notation that we will use throughout.
-For $j \in \{0,1\}$, let $\tilde{x}_i(j) := [1, j, z_i]^T$ denote the value of $\tilde{x}_i$ that results from setting $p_i$ to $j$.
+For $j \in \{0,1\}$, let $\tilde{x}_i(j) \equiv [1, j, z_i]^T$ denote the value of $\tilde{x}_i$ that results from setting $p_i$ to $j$.
 Next, let  $l^m_i(j)$, $\eta^m_i(j),$ and $\mu^m_i(j)$ be the values of $l^m_i$, $\eta^m_i$, and $\mu^m_i$, respectively, that result from setting $p_i$ to $j$, i.e.,
-$l^m_i(j) = \langle \tilde{x}_i(j), \beta_m \rangle + o^m_i$, $\eta^m_i(j) = h_m(l^m_i(j))$, and
-$\mu_i^m(j) = r_m^{-1}(l^m_i(j)).$ Let the corresponding gRNA quantities $l^g_i(j)$, $\eta_i^g(j)$, and $\mu^g_i(j)$ be defined analogously.
+$l^m_i(j) \equiv \langle \tilde{x}_i(j), \beta_m \rangle + o^m_i$, $\eta^m_i(j) \equiv h_m(l^m_i(j))$, and
+$\mu_i^m(j) \equiv r_m^{-1}(l^m_i(j)).$ Let the corresponding gRNA quantities $l^g_i(j)$, $\eta_i^g(j)$, and $\mu^g_i(j)$ be defined analogously.
 Next, let $X \in \mathbb{R}^{n \times {d-1}}$ be the observed design matrix, and let $\tilde{X} \in \mathbb{R}^{n \times d}$ be the augmented design matrix that results from concatenating the column of (unobserved) $p_i$s to $X$, i.e.
-$$ X := \begin{bmatrix} 
+$$ X \equiv \begin{bmatrix} 
 1 & z_1 \\
@@
 \end{bmatrix}; \quad 
-\tilde{X}  := 
+\tilde{X} \equiv 
 \begin{bmatrix}
@@
 Finally, let $[m,m]^T \in \mathbb{R}^{2n}$ be the vector that results from concatenating $m$ to itself, i.e.
-$[m,m]^T := [m_1, \dots, m_n, m_1, \dots, m_n],$ and let $[g,g]^T$, $[o^g,o^g]^T$, and $[o^m,o^m]^T$ be defined similarly.
+$[m,m]^T \equiv [m_1, \dots, m_n, m_1, \dots, m_n],$ and let $[g,g]^T$, $[o^g,o^g]^T$, and $[o^m,o^m]^T$ be defined similarly.
 
@@
 
-\subsection{Estimation and inference}
-
+\subsection{Estimation and inference}\label{sec:estimation_inference}
 We derive an EM algorithm (Algorithm \ref{algo:em_full}) to estimate the parameters of the GLM-EIV model.
@@
 Assuming that we run the algorithm 15 times using randomly-generated pilot estimates, and assuming that the algorithm iterates through E and M steps about 10 times per run, we must fit approximately 300 GLMs.
-(These numbers are based on exploratory applications of the method to real and simulated data.) We devised a strategy to produce a highly accurate pilot estimate of the true parameters, enabling us to run the algorithm once and converge upon the MLE within a few iterations.
-The strategy involves layering on top of one another several statistical tricks (that we expect to be independently useful for accelerating other single-cell methods).
-Details are deferred to Appendix \ref{sec:statistical_accelerations}.
+(These numbers are based on exploratory applications of the method to real and simulated data.) We devise a strategy to produce a highly accurate pilot estimate of the true parameters, enabling us to run the algorithm once and converge upon the MLE within a few iterations.
+The strategy involves layering several statistical ``tricks'' (that we expect to be independently useful for accelerating other single-cell methods) on top of one another; details are deferred to Appendix \ref{sec:statistical_accelerations}.
 Overall, the statistical accelerations reduce the number of GLMs that we must fit to $<10$ in most cases.
 
-Next, we developed a computational infrastructure to apply GLM-EIV to large-scale, single-cell CRISPR screen data.
+Next, we develop a computational infrastructure to apply GLM-EIV to large-scale, single-cell CRISPR screen data.
 The infrastructure leverages \texttt{Nextflow}, a programming language that facilitates building data-intensive pipelines \parencite{DITommaso2017}, and \texttt{ondisc}, an \texttt{R} package that we developed (in a separate project) to facilitate large-scale computing on single-cell data.
@@
 The pipeline aggressively recycles computation when possible, saving a considerable amount of compute; see Appendix \ref{sec:computing} for details.
-Overall, the statistical accelerations and computational infrastructure together enable us to deploy GLM-EIV to large-scale single-cell CRISPR screen data without much hassle.
-
-\section{Simulation studies}\label{sec:simulation}
+Overall, the statistical accelerations and computational infrastructure make the deployment of GLM-EIV to large-scale single-cell CRISPR screen quite feasible.
+
+\subsection{Connections to other methods}\label{sec:statistical_lit_review}
+The closest parallels to GLM-EIV in the statistical methodology literature are \parencite{Grun2008} and \parencite{Ibrahim1990}.
+Gr\"{u}n \& Leisch derived a method for estimation and inference in a $k$-component mixture of GLMs.
+While we prefer to view GLM-EIV as a generalized errors-in-variables method,  the GLM-EIV model is equivalent to a two-component mixture of products of GLM densities.
+Ibrahim proposed a procedure for fitting GLMs in the presence of missing-at-random covariates.
+Our method, by contrast, involves fitting two conditionally independent GLMs in the presence of a totally latent covariate.
+Thus, while Ibrahim and Gr\"{u}n \& Leisch are helpful references, our estimation and inference tasks are more complex than theirs.
+Next, \parencite{Aigner1973} and \parencite{Savoca2000} proposed measurement error models that consist of unobserved \textit{binary} rather than \textit{continuous} predictors; the latter are more commonly used in measurement error models.
+GLM-EIV likewise consists of a latent binary predictor, but our statistical context is more GLM-oriented than theirs.
+
+\section{Simulation study}\label{sec:simulation}
 
@@
 We included ``sequencing batch'' (modeled as a Bernoulli-distributed variable) as a covariate and sequencing depth (modeled as a Poisson-distributed variable) as an offset.
-We varied the log-fold change in gRNA expression, $\beta^g_1,$ over a grid on the interval $[\log(1), \log(4)];$ $\beta^g_1$ controls problem difficulty, with higher values corresponding to easier problems.
+We varied the log-fold change in gRNA expression, $\beta^g_1,$ over a grid on the interval $[\log(1), \log(4)];$ $\beta^g_1$ controls problem difficulty, with higher values corresponding to easier problem settings.
 Finally, we generated the gene expression and gRNA count data from two response distributions: Poisson and negative binomial (size parameter fixed at $\theta = 20$ for the latter).
@@
 We ran all methods on the negative binomial data twice: once treating the size parameter $\theta$ as a known constant and once treating $\theta$ as unknown.
-In the latter case we used the \texttt{glm.nb} function from the \texttt{MASS} package to estimate $\theta$ before applying the methods \cite{Ripley2013}.
+In the latter case we used the \texttt{glm.nb} function from the \texttt{MASS} package to estimate $\theta$ before applying the methods \parencite{Ripley2013}.
 We display the results of the simulation study in Figure \ref{main_text_sim}.
-Columns correspond to distributions (i.e., Poisson, NB with known $\theta$, and NB with unknown $\theta$), rows correspond to performance metrics (i.e., bias, mean squared error, CI coverage rate (nominal rate $95\%$), CI width, and method execution time).
+Columns correspond to distributions (i.e., Poisson, NB with known $\theta$, and NB with unknown $\theta$), and rows correspond to performance metrics (i.e., bias, mean squared error, CI coverage rate (nominal rate $95\%$), CI width, and method execution time).
 The problem difficulty parameter $\beta^g_1$ is plotted on the horizontal axis, and the methods are depicted in different colors (GLM-EIV masked by accelerated GLM-EIV in several panels).
@@
 GLM-EIV demonstrated superior statistical performance to the thresholding method on all metrics (rows 1-4).
-Accelerated GLM-EIV had substantially lower computational cost than ``vanilla'' GLM-EIV (bottom row) despite demonstrating identical statistical performance (rows 1-4).}
+Accelerated GLM-EIV had substantially lower computational cost than ``vanilla'' GLM-EIV (bottom row), despite demonstrating identical statistical performance (rows 1-4).}
 	\label{main_text_sim}
@@
 Estimating $\theta$ leads to slight inflation bias (i.e., overestimating the true effect size), whereas, as we showed previously, thresholding leads to attenuation bias (i.e., underestimating the true effect size).
-These phenomena partially cancel, yielding less biased estimates.
+These phenomena partially canceled, yielding less biased estimates.
 GLM-EIV exhibited worse performance under unknown $\theta$ than known $\theta$, likely due to poor $\theta$ estimation.
-We note that GLM-EIV and the thresholding method in principal are compatible with \textit{any} $\theta$ estimation procedure, including those based on more sophisticated techniques, such as regularization \cite{Hafemeister2019}.
+We note that GLM-EIV and the thresholding method in principal are compatible with \textit{any} $\theta$ estimation procedure, including those based on more sophisticated techniques, such as regularization \parencite{Hafemeister2019}.
 We defer rigorous investigation of the impact of different $\theta$ estimation strategies on these methods to future work.
@@
 Leveraging our computational infrastructure, we applied GLM-EIV and the thresholding method to analyze the entire Gasperini and Xie datasets.
-(We report only the most important aspects of the analysis and results in the main text; full details are available in Appendix \ref{sec:data_analysis_details}.) We set the threshold to the approximate Bayes-optimal decision boundary, as our theoretical analyses and simulation studies revealed that the Bayes-optimal decision boundary is a good choice for the threshold when the gRNA count distribution is well-separated.
+We report only the most important aspects of the analysis and results in the main text; full details are available in Appendix \ref{sec:data_analysis_details}.
+We set the threshold in the thresholding method to the approximate Bayes-optimal decision boundary, as our theoretical analyses and simulation studies indicated that the Bayes-optimal decision boundary is a good choice for the threshold when the gRNA count distribution is well-separated.
 Operating under the assumption that the effect of the perturbation on gRNA expression is similar across pairs, we leveraged the fitted GLM-EIV models to approximate the Bayes boundary in the following way: we (i) sampled several hundred gene-perturbation pairs, (ii) extracted the fitted values $\hat{\beta}_g$ and $\hat{\pi}$ from the GLM-EIV models fitted to these pairs, (iii) computed the median $\overline{\hat{\beta}_g}$ and $\overline{\hat{\pi}}$ across the $\hat{\beta}_g$s and $\hat{\pi}$s, and (iv) used $\overline{\hat{\beta}_g}$ and $\overline{\hat{\pi}}$ to estimate a dataset-wide Bayes-optimal decision boundary.
@@
 However, this was not obvious \textit{a priori} and may not be the case for other datasets.
-We note that GLM-EIV produced outlier estimates (likely due to non-global EM convergence) on a small ($< 2.5 \%$ on Gasperini, $<0.05\%$ on Xie) number of pairs consisting of a handful of genes.
-These outliers were not plotted in Figures \ref{fig:real_data}a-b but were used to compute the CI coverage.
+We note that GLM-EIV produced outlier estimates (defined as estimated fold change $<0.75$ or $>1.25$) on a small ($< 2.5 \%$ on Gasperini, $<0.05\%$ on Xie) number of pairs consisting of a handful of genes, likely due to non-global EM convergence.
+These outliers are not plotted in Figures \ref{fig:real_data}a-b but were used to compute the CI coverage reported in the inset tables.
 
@@
 In this work we introduced GLM-EIV (``GLM-based errors in variables''), a new method for single-cell CRISPR screen analysis.
-GLM-EIV extends the classical errors-in-variables model to responses and noisy predictors that are exponential family-distributed and potentially impacted by the same set of confounding variables.
-These extensions enable GLM-EIV to resolve novel analysis challenges posed by single-cell CRISPR screens.
+GLM-EIV extends the classical errors-in-variables model to responses and noisy predictors that are exponential family-distributed and potentially impacted by the same set of confounding variables, enabling GLM-EIV to resolve novel analysis challenges posed by single-cell CRISPR screens.
 We demonstrated through simulation studies, real data analyses, and theory that GLM-EIV outperforms thresholded regression by a considerable margin in high background contamination settings.
 GLM-EIV intuitively achieves this performance gain by leveraging information from \textit{both} modalities (rather than the gRNA modality alone) to assign perturbation identities to cells.
-On the other hand, in low background contamination settings GLM-EIV and thresholded regression work best in conjunction, with GLM-EIV providing a statistically principled and empirically effective procedure for selecting the threshold.
+On the other hand, in low background contamination settings, GLM-EIV and thresholded regression work best in conjunction, with GLM-EIV providing a statistically principled and empirically effective procedure for selecting the threshold.
 GLM-EIV thereby neutralizes a tuning parameter that, until this point, has been selected using heuristic procedures, with little confidence that the choice is near optimal.
@@
 We hope that this work helps to introduce the broader statistics community to an emerging class of functional genomics assays that likely will exert a major impact on biological research in the coming years \parencite{Przybyla2021}.
-Additionally, this is the first work to leverage the \texttt{ondisc}-\texttt{Nextflow}-cloud technology stack, a tightly-integrated, user-friendly, and powerful set of tools for large-scale single-cell analysis.
-We expect this technology stack to be of interest to other researchers.
-
-\begin{figure}[h]
+Additionally, this is the first work to leverage the \texttt{ondisc}-\texttt{Nextflow}-HPC/cloud technology stack, a tightly-integrated, user-friendly, and powerful set of tools for large-scale single-cell analysis.
+We expect this technology stack to be of interest to other single-cell researchers.
+
+\begin{figure}
 	\centering
@@
 We anticipate that GLM-EIV could be applied to other types of single-cell CRISPR screen and multimodal single-cell data.
-For example, GLM-EIV likely could be extended (with effort) to ``low-multiplicity of infection'' screens  \parencite{Schraivogel2020} in which each cell receives one or two perturbations rather than dozens (as is the case in ``high multiplicity screens,'' studied in this work).
+For example, GLM-EIV might be extended (with some effort) to ``low-multiplicity of infection'' screens  \parencite{Schraivogel2020} in which each cell receives one or two perturbations rather than dozens (as is the case in ``high multiplicity screens,'' studied in this work).
 We also could apply GLM-EIV to analyze multimodal single-cell chromatin accessibility assays.
-A question of interest in such experiments is whether chromatin state (i.e., closed or open) is associated with the expression of a gene (or abundance of a protein) \parencite{Mimitou2021}.
+A question of interest in such experiments is whether chromatin state (i.e., closed or open) is associated with the expression of a gene or abundance of a protein \parencite{Mimitou2021}.
 We do not directly observe the chromatin state of a cell; instead, we observe tagged DNA fragments that serve as count-based proxies for whether a given region of chromatin is open or closed.
@@
 
-Finally, GLM-EIV might be extended to areas beyond genomics, such as psychology.
+Several authors working on statistical methods for single-cell data recently have extended models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions.
+For example, \parencite{Lin2021} and \parencite{Townes2019} (separately) developed eSVD and GLM-PCA, generalizations of SVD and PCA, respectively, to exponential family response distributions.
+Unlike their vanilla counterparts, eSVD and GLM-PCA can model gene expression counts directly, improving performance on dimension reduction tasks.
+We see our work (in part) as a continuation of this broad effort to ``port'' common statistical methods and models to single-cell count data.
+Our focus, however, is on regression rather than dimension reduction: we extend the classical errors-in-variables model in several key directions (see above), enabling its direct and natural application to multimodal single-cell data.
+
+Finally, GLM-EIV might be applied to areas beyond genomics, such as psychology.
 Many psychological constructs (e.g., presence or absence of a social media addiction) are latent and can be assessed only through an imperfect proxy (e.g., the number of times one has checked social media).
@@
 We likewise need to compute the analogous gRNA quantities.
-We describe how to compute these quantities in R by extending base family objects.
+The procedure that we propose for this purpose is general, but for concreteness, we describe how to implement this procedure in R by extending base family objects.
 We implicitly condition on $p_i$, $z^m_i$, and $o^m_i$.
+
+		An \texttt{R} family object contains several functions, including \texttt{linkinv}, \texttt{variance}, and \texttt{mu.eta}.
+\texttt{linkinv} is the inverse link function $r_m^{-1}$.
+\texttt{variance} takes as an argument the mean $\mu^m_i$ of the $i$th example and returns its variance $[\sigma_i^m]^2$.
+\texttt{mu.eta} is the derivative of the inverse link function $[r^{-1}_m]^{'}$.
+We extend the \texttt{R} family object by adding two additional functions: \texttt{skewness} and \texttt{mu.eta.prime}.
+\texttt{skewness} returns the skewness $\gamma^m_i$ of the distribution as a function of the mean $\mu_i$, i.e.
+$$\texttt{skewness}\left(\mu_i\right) = \mathbb{E} \left[\left(\frac{m_i - \mu_i^m}{ \sigma_i^m}\right)^3\right] := \gamma_i^m.$$ Finally, the \texttt{mu.eta.prime} is the second derivative of the inverse link function $[r^{-1}_m]''.$ Algorithm \ref{algo:computing_info_matrices} computes the matrices $\Delta^m(j)$, $[\Delta']^m(j)$, $V^m(j)$, and $H^m(j)$ and vector $s^m(j)$ for given $\beta_m$ and given family object.
+(The vector $w^m(j)$ can be computed in terms of $\Delta^m(j)$ and $H^m(j)$.) We use $\sigma^m_i(j)$ (resp.
+$\gamma^m_i(j)$) to refer to the standard deviation (resp.
+skewness) of the gene expression distribution the $i$th cell when the perturbation $p_i$ is set to $j$.
 
@@
 
-		An \texttt{R} family object contains several functions, including \texttt{linkinv}, \texttt{variance}, and \texttt{mu.eta}.
-\texttt{linkinv} is the inverse link function $r_m^{-1}$.
-\texttt{variance} takes as an argument the mean $\mu^m_i$ and returns the variance $[\sigma_i^m]^2$.
-\texttt{mu.eta} is the derivative of the inverse link function $[r^{-1}_m]^{'}$.
-We extend the \texttt{R} family object by adding two additional functions: \texttt{skewness} and \texttt{mu.eta.prime}.
-\texttt{skewness} returns the skewness $\gamma^m_i$ of the distribution as a function of the mean $\mu_i$, i.e.
-$$\texttt{skewness}\left(\mu_i\right) = \mathbb{E} \left[\left(\frac{m_i - \mu_i^m}{ \sigma_i^m}\right)^3\right] := \gamma_i^m.$$ Finally, the \texttt{mu.eta.prime} is the second derivative of the inverse link function $[r^{-1}_m]''.$ Algorithm \ref{algo:computing_info_matrices} computes the matrices $\Delta^m(j)$, $[\Delta']^m(j)$, $V^m(j)$, $H^m(j)$, and vector $s^m(j)$ for given $\beta_m$ and given family object.
-(The vector $w^m(j)$ can be computed in terms of $\Delta^m(j)$ and $H^m(j)$.) We use $\sigma^m_i(j)$ (resp.
-$\gamma^m_i(j)$) to refer to the standard deviation (resp.
-skewness) of the gene expression distribution the $i$th cell when the perturbation $p_i$ is set to $j$.
-
 		All steps of the algorithm are obvious except the calculation of $h'_m(l^m_i(j))$ (line 6), $h''(l^m_i(j))$ (line 9), and $V^m_i(j)$ (line 12).
@@
 		\begin{equation}\label{computing_info_matrix_3}
-		(r_m^{-1})'(t) = \psi_m''(h_m(t))h_m'(t), 
-		\end{equation}
-		or  $$h_m'(t) = \frac{(r_m^{-1})'(t) }{\psi_m''(h_m(t))}.$$ Finally, plugging in $l^m_i$ for $t$,
+		(r_m^{-1})'(t) = \psi_m''(h_m(t))h_m'(t) \iff h_m'(t) = \frac{(r_m^{-1})'(t) }{\psi_m''(h_m(t))}.
+		\end{equation}
+	Finally, plugging in $l^m_i$ for $t$,
 		$$ h_m'(l_i) = \frac{(r_m^{-1})'(l^m_i)}{\psi''(h_m(l^m_i))} = \frac{(r_m^{-1})'(l^m_i)}{\psi_m''(\eta^m_i)} = \textrm{ by (\ref{computing_info_matrix_1}) } \frac{(r_m^{-1})'(l_i^m)}{[\sigma_i^m]^2}.$$
@@
 		Differentiating (\ref{computing_info_matrix_3}) in $t$, we obtain
-		$$ (r_m^{-1})''(t) = \psi_m'''(h_m(t)) [h_m'(t)]^2 + \psi_m''(h_m(t)) h_m''(t),$$ or $$h_m''(t) =\frac{(r_m^{-1})''(t) - \psi'''(h_m(t))[h_m'(t)]^2}{\psi_m''(h_m(t))}.$$ Plugging in $l^m_i$ for $t$, we find that
+		$$(r_m^{-1})''(t) = \psi_m'''(h_m(t)) [h_m'(t)]^2 + \psi_m''(h_m(t)) h_m''(t) \iff h_m''(t) =\frac{(r_m^{-1})''(t) - \psi'''(h_m(t))[h_m'(t)]^2}{\psi_m''(h_m(t))}.$$ Plugging in $l^m_i$ for $t$, we find that
 		\begin{equation*}
@@
 
-		First, we performed quality control on the datasets.
-As is standard in single-cell analysis \cite{Choudhary2021a}, we removed cells with a high fraction ($>8\%$) of mitochondrial reads.
+		First, we performed quality control on both datasets.
+As is standard in single-cell analysis \parencite{Choudhary2021a}, we removed cells with a high fraction ($>8\%$) of mitochondrial reads.
 We additionally excluded genes that were expressed in fewer than $10\%$ of cells or that had a mean expression level of less than $1$.
@@
 $1,030$) genes, and $6,598$ (resp.
-$516$) gRNAs.
+$516$) distinct perturbations.
 
@@
 The \textit{cis} pairs consisted of genes paired to nearby enhancers with unknown regulatory effects.
-The negative control pairs consisted of nontargeting gRNAs paired to genes.
-The Xie data contained $681$ candidate \textit{cis} pairs and, unfortunately, no positive control or negative control pairs.
-We therefore constructed a set of $50,000$ \textit{in silico} negative control pairs on the Xie data by pairing perturbations to genes on different chromosomes; see \textit{Methods} section of \cite{Barry2020} for details.
+The negative control pairs consisted of non-targeting gRNAs paired to genes.
+The Xie data contained $681$ candidate \textit{cis} pairs and.
+Unfortunately, it did not come with positive control or negative control pairs.
+We therefore constructed a set of $50,000$ \textit{in silico} negative control pairs on the Xie data by pairing perturbations to genes on different chromosomes; see \textit{Methods} section of \parencite{Barry2021} for details.
 
 		We modeled the gene expression counts using a negative binomial distributions with unknown size parameter $\theta$; we estimated $\theta$ using the {glm.nb} package.
-Choudhary and Satija \cite{Choudhary2021a} report that Poisson models accurately capture highly sparse single-cell data.
+Choudhary and Satija \parencite{Choudhary2021a} report that Poisson models accurately capture highly sparse single-cell data.
 Although Choudhary and Satija did not investigate gRNA data specifically, we modeled the gRNA counts using Poisson distributions, as the gRNA modality exhibited greater sparsity than the gene modality.

### 2022-01-04 text/jasa/jasa.tex
fix reproducibility

@@
 backend=biber,
-style=numeric-comp,
+style=chicago-authordate,
 maxcitenames=1,
@@
 {
-  \title{\bf A new class of measurement error models for CRISPR genome engineering and single-cell RNA sequencing}
-  \author{Timothy Barry$^1$, Eugene Katsevich$^2$, Kathyrn Roeder$^1$ \\
-    $^1$Department of Statistics and Data Science, Carnegie Mellon University \\
-    $^2$Department of Statistics and Data Science, Wharton School}
+	%  \title{\bf FDR control following filtering for hierarchically structured hypotheses}
+	\title{\bf A class of measurement error models for singe-cell CRISPR screens}
+	%	Eugene Katsevich, Chiara Sabatti, Marina Bogomolov}
+	\author{Timothy Barry\thanks{ FUNDING INFORMATION GOES HERE.}\hspace{.2cm}\\
+		Department of Statistics and Data Science,\\Carnegie Mellon University\\
+		and\\
+		Eugene Katsevich\\
+		Department of Statistics and Data Science,\\Wharton School, University of Pennsylvania\\
+		and\\
+		Kathryn Roeder\\
+		Department of Statistics and Data Science,\\Computational Biology Department,\\ 
+		Carnegie Mellon University}
 	\maketitle
@@
 Despite their promise, single-cell CRISPR screens present substantial statistical challenges.
-We demonstrate through theoretical and real data analyses that a standard method for estimation and inference in single-cell CRISPR screens --``thresholded regression'' -- exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic and challenging-to-select tuning parameter.
+We demonstrate through theoretical and real data analyses that a standard method for estimation and inference in single-cell CRISPR screens --``thresholded regression'' -- exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic, challenging-to-select tuning parameter.
 To overcome these difficulties, we introduce GLM-EIV (``GLM-based errors-in-variables''), a new method for single-cell CRISPR screen analysis.
@@
 
-\noindent
-{\it Keywords:}  exponential family, errors-in-variables, GLM, CRISPR screen, differential expression
+\noindent%
+{\it Keywords:} exponential family, GLM, latent variable model, RNA-seq, genome engineering, cloud computing
+
 \spacingset{1.9}
+
 \section{Introduction}
-CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new medical therapies \cite{Rothgangl2021,Musunuru2021} and transforming basic biology research \cite{Przybyla2021}.
-Recently, scientists have paired CRISPR genome engineering with single-cell sequencing \cite{Dixit2016,Datlinger2017}.
-The resulting assays, known as a ``single-cell CRISPR screens,'' link genetic perturbations in individual cells to changes in gene expression, illuminating regulatory networks underlying human diseases and other traits \cite{Morris2021a}.
+CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new medical therapies \parencite{Rothgangl2021,Musunuru2021} and transforming basic biology research \parencite{Przybyla2021}.
+Recently, scientists have paired CRISPR genome engineering with single-cell RNA sequencing \parencite{Dixit2016,Datlinger2017}.
+The resulting assays, known as a ``single-cell CRISPR screens,'' link genetic perturbations in individual cells to changes in gene expression, illuminating regulatory networks underlying human diseases and other traits \parencite{Morris2021a}.
 
 Despite their promise, single-cell CRISPR screens present substantial statistical challenges.
-A major difficulty is that CRISPR perturbations are assigned stochastically to cells and cannot be observed directly.
+A major difficulty is that the ``treatment'' -- i.e., the presence of absence of a CRISPR perturbation -- is assigned randomly to cells and is not directly observable.
 As a consequence, one cannot know with certainty which cells were perturbed.
-Instead, one must leverage an indirect, noisy proxy of perturbation presence or absence -- namely, transcribed guide RNA counts -- to ``guess'' which cells were perturbed.
-Using these imputed perturbation assignments, one can attempt to estimate the effect of the perturbation on gene expression.
-The standard approach, which we call ``thresholded regression'' or the ``thresholding method,'' is to assign perturbation identities to cells by simply thresholding the guide RNA counts.
+Instead, one must leverage an indirect, quantitative proxy of perturbation presence or absence to ``guess'' which cells received a perturbation.
+This indirect proxy takes the form of so-called guide RNA counts, with higher counts indicating that a cell is more likely to have been perturbed.
+The standard approach to single-cell CRISPR screen analysis is to impute perturbation assignments onto the cells by simply thresholding the gRNA counts; using these imputations, one can attempt to estimate the effect of the perturbation on gene expression.
+We call this standard approach ``thresholded regression'' or the ``thresholding method.''
 
@@
 Second, we introduce a new method, GLM-EIV (``GLM-based errors-in-variables''), for single-cell CRISPR screen analysis.
-GLM-EIV extends the classical errors-in-variables model to response distributions and sources of measurement error that are (i) exponential family-distributed and (ii) potentially impacted by the same set of confounding variables.
+GLM-EIV extends the classical errors-in-variables model to responses and noisy predictors that are (i) exponential family-distributed and (ii) potentially impacted by the same set of confounding variables.
 GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic.
-% Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.
 We implement several statistical accelerations (that possibly are of independent utility) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method.
@@
 Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions.
-For example, Lin et al.\ \cite{Lin2021} and Townes et al.\ \cite{Townes2019}  (separately) developed eSVD and GLM-PCA, generalizations of SVD and PCA, respectively, to exponential family response distributions.
-Unlike their vanilla counterparts, eSVD and GLM-PCA can model gene expression counts directly, thereby capturing the relationship between the mean and variance of a gene's expression level \cite{Lause2021} and improving performance on dimension reduction tasks.
+For example, Lin et al.\ \parencite{Lin2021} and Townes et al.\ \parencite{Townes2019}  (separately) developed eSVD and GLM-PCA, generalizations of SVD and PCA, respectively, to exponential family response distributions.
+Unlike their vanilla counterparts, eSVD and GLM-PCA can model gene expression counts directly, thereby capturing the relationship between the mean and variance of a gene's expression level \parencite{Lause2021} and improving performance on dimension reduction tasks.
 We see our work as a continuation of this effort to ``port'' common statistical methods and models to single-cell count data.
 Our focus, however, is on regression rather than dimension reduction.
-We extend the classical errors-in-variables model in several directions, most importantly generalizing to exponential family-distributed response distributions and sources of measurement error.
-
-The closest parallels to our work in the statistical methodology literature are Gr\"{u}n \& Leisch \cite{Grun2008} and Ibrahim \cite{Ibrahim1990}.
+We extend the classical errors-in-variables model in several directions, most importantly generalizing to responses and noisy predictors that are exponential family-distributed.
+
+The closest parallels to our work in the statistical methodology literature are Gr\"{u}n \& Leisch \parencite{Grun2008} and Ibrahim \parencite{Ibrahim1990}.
 Gr\"{u}n \& Leisch derived a method for estimation and inference in a $k$-component mixture of GLMs.
@@
 Thus, while Ibrahim and Gr\"{u}n \& Leisch are helpful references, our estimation and inference tasks are more complex than theirs.
-Other relevant references include Aigner \cite{Aigner1973} and Savoca \cite{Savoca2000}.
-
-The genomics literature has produced several applied methods for single-cell CRISPR screen analysis.
-Barry et al.\ developed SCEPTRE \cite{Barry2021}, a custom implementation of the conditional randomization test \cite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data.
+Next, we focus in this work on noisy \textit{binary} predictors rather than noisy \textit{continuous} predictors, the latter of which are more commonly used in measurement error models.
+\parencite{Aigner1973} and \parencite{Savoca2000} also developed measurement error models for noisy binary predictors, albeit in a different statistical context than ours.
+
+The genomics literature has produced a few applied methods for single-cell CRISPR screen analysis.
+Most relevantly, Barry et al.\ developed SCEPTRE, a custom implementation of the conditional randomization test \parencite{Candes2018, Liu2021} tailored to single-cell CRISPR screen data \parencite{Barry2021}.
 SCEPTRE tests whether a given perturbation is associated with the change in expression of a given gene, adjusting for sources of confounding and ensuring robustness to expression model misspecification.
@@
 
-There are several broad classes of single-cell CRISPR screen assays, each suited to answer a different set of biological questions \cite{Gasperini2019,Datlinger2021,Mimitou2019}.
+There are several broad classes of single-cell CRISPR screen assays, each suited to answer a different set of biological questions \parencite{Gasperini2019,Datlinger2021,Mimitou2019}.
 In this work we focus on high-multiplicity of infection (MOI) single-cell CRISPR screens.
@@
 The human genome consists of genes, enhancers (segments of DNA that regulate the expression of one or more genes), and other genomic elements (that are not of importance to the current work).
-Genome-wide association studies (GWAS) have revealed that the majority ($>90\%$) of variants associated with diseases lie outside genes and (very likely) inside enhancers \cite{Gallagher2018}.
+Genome-wide association studies (GWAS) have revealed that the majority ($>90\%$) of variants associated with diseases lie outside genes and (very likely) inside enhancers \parencite{Gallagher2018}.
 These noncoding variants are thought to contribute to disease by modulating the expression of one or more disease-relevant genes.
 Scientists do not know the gene (or genes) through which most noncoding variants exert their effect, limiting the interpretability of GWAS results.
-A central open challenge in genetics, therefore, is to link enhancers that harbor GWAS variants to the genes that they target at genome-wide scale \cite{Gasperini2020}.
+A central open challenge in genetics, therefore, is to link enhancers that harbor GWAS variants to the genes that they target at genome-wide scale \parencite{Gasperini2020}.
 
@@
 
-In this work we analyze two large-scale, high MOI, single-cell CRISPR screen datasets published by Gasperini et al.\ \cite{Gasperini2019} and Xie et al.\ \cite{Xie2019a}.
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=1\linewidth]{../../figures/analysis_challenges/plot.pdf}
+	\caption{\textbf{Experimental design and analysis challenges}: \textbf{a,} Experimental design.
+For a given perturbation (e.g., the perturbation indicated in blue), we partition the cells into two groups: perturbed and unperturbed.
+Next, for a given gene, we conduct a differential expression analysis across the two groups, yielding an estimate of the impact of the given perturbation on the given gene.
+\textbf{b,} DAG representing all variables in the system.
+The perturbation (latent) impacts both gene expression and gRNA expression; technical factors act as confounders, also impacting gene and gRNA expression.
+The target of estimation is the effect of the perturbation on gene expression.
+\textbf{c,} Schematic illustrating the ``background read'' phenomenon.
+Due to errors in the sequencing and alignment processes, unperturbed cells exhibit a nonzero gRNA count distribution (bottom).
+The target of estimation is the change in mean gene expression in response to the perturbation (top).
+\textbf{d}, Example data on four cells for a given perturbation-gene pair.
+Note that (i) the perturbation is unobserved, and (ii) the gene and gRNA data are discrete counts.}
+	\label{analysis_challenges}
+\end{figure}
+
+In this work we analyze two large-scale, high MOI, single-cell CRISPR screen datasets published by (\cite{Gasperini2019}) and (\cite{Xie2019a}).
 Gasperini (resp., Xie) targeted approximately 6,000 (resp., 500) candidate enhancers in a population of approximately 200,000 (resp., 100,000) cells.
@@
 These extensions enable GLM-EIV to resolve novel analysis challenges posed by single-cell CRISPR screens.
-We demonstrated through simulation studies, real data analyses, and theory that GLM-EIV outperforms thresholded regression by a considerable margin in high background contamination settings.
+We demonstrated through simulation studies, real data analyses, and mathematical theory that GLM-EIV outperforms thresholded regression by a considerable margin in high background contamination settings.
 GLM-EIV intuitively achieves this performance gain by leveraging information from \textit{both} modalities (rather than the gRNA modality alone) to assign perturbation identities to cells.
@@
 
-\clearpage
 \section*{Figures}
 % 1. Challenges
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=1\linewidth]{../../figures/analysis_challenges/plot.pdf}
-	\caption{\textbf{Experimental design and analysis challenges}: \textbf{a,} Experimental design.
-For a given perturbation (e.g., the perturbation indicated in blue), we partition the cells into two groups: perturbed and unperturbed.
-Next, for a given gene, we conduct a differential expression analysis across the two groups, yielding an estimate of the impact of the given perturbation on the given gene.
-\textbf{b,} DAG representing all variables in the system.
-The perturbation (latent) impacts both gene expression and gRNA expression; technical factors act as confounders, also impacting gene and gRNA expression.
-The target of estimation is the effect of the perturbation on gene expression.
-\textbf{c,} Schematic illustrating the ``background read'' phenomenon.
-Due to errors in the sequencing and alignment processes, unperturbed cells exhibit a nonzero gRNA count distribution (bottom).
-The target of estimation is the change in mean gene expression in response to the perturbation (top).
-\textbf{d}, Example data on four cells for a given perturbation-gene pair.
-Note that (i) the perturbation is unobserved, and (ii) the gene and gRNA data are discrete counts.}
-	\label{analysis_challenges}
-\end{figure}
 
@@
 
+%\clearpage
+%\begin{figure}[h!]
+%	\centering
+%	\includegraphics[width=0.7\linewidth]{../../figures/thresholding_theoretical/plot}
+%	\caption{\textbf{Thresholding method bias-variance decomposition.} Bias decreases and variance increases as the threshold tends to infinity.
+$\beta^g_1 = 1, \beta^m_1 = 1,$ and $\pi = 0.1$ in this plot.}
+%	\label{thresholding_theoretical}
+% \end{figure}
+
 \clearpage
-\begin{figure}[h!]
-	\centering
-	\includegraphics[width=0.7\linewidth]{../../figures/thresholding_theoretical/plot}
-	\caption{\textbf{Thresholding method bias-variance decomposition.} Bias decreases and variance increases as the threshold tends to infinity.
-$\beta^g_1 = 1, \beta^m_1 = 1,$ and $\pi = 0.1$ in this plot.}
-	\label{thresholding_theoretical}
-\end{figure}
-
-\clearpage
-% Theory
 \begin{figure}[h!]

### 2022-01-04 R_scripts/analysis/approx_bayes_bdy.R

@@ -1,5 +1,5 @@
 library(magrittr)
-load_all("~/research_code/glmeiv/")
+library(glmeiv)
 library(dplyr)
 
 ########

### 2022-01-04 R_scripts/processing/qc_gasperini.R

@@ -89,6 +89,7 @@ set.seed(11)
 sample_pairs <- subsetted_pairs %>% dplyr::filter(gene_id %in% sample(x = subsetted_pairs$gene_id, size = 4, replace = FALSE) &
                                   gRNA_id %in% sample(x = subsetted_pairs$gRNA_id, size = 4, replace = FALSE)) %>%
   dplyr::slice_sample(n = 15)
+pc_pairs <- subsetted_pairs %>% dplyr::filter(site_type == "selfTSS")
 sample_pairs_pc <- subsetted_pairs %>% dplyr::filter(site_type == "selfTSS") %>% dplyr::slice_sample(n = 15)
 
 ########################################################
@@ -106,6 +107,9 @@ saveRDS(object = covariate_matrix_to_save,
 # subsetted pairs
 saveRDS(object = subsetted_pairs,
         paste0(glmeiv_offsite_dir_gasp_data, "gRNA_gene_pairs.rds"))
+# subsetted pc pairs
+saveRDS(object = subsetted_pairs,
+        paste0(glmeiv_offsite_dir_gasp_data, "gRNA_gene_pairs_pc.rds"))
 # sample pairs
 saveRDS(sample_pairs,
         paste0(glmeiv_offsite_dir_gasp_data, "gRNA_gene_pairs_sample.rds"))

### 2022-01-04 bash_scripts/run_all.sh

@@ -8,37 +8,54 @@
 bash setup.sh
 source ~/.research_config
 
-# set a couple variables
-sim_dir=$LOCAL_GLMEIV_DATA_DIR"public/simulations/"
-
-#######################
-# 1. Gasperini analysis
-#######################
+################
+# 1. QC datasets
+################
 # QC Gasperini data
 Rscript ../R_scripts/processing/qc_gasperini.R
-# run glmeiv at scale on Gasperini data
-bash run_gasp_glmeiv.sh
-# run thresholding method at scale on Gasperini data
-bash run_gasp_thresh.sh
-
-#################
-# 2. Xie analysis
-#################
 # QC Xie data
 Rscript ../R_scripts/processing/qc_xie.R
+
+#####################
+# 2. GLM-EIV analysis
+#####################
+# run glmeiv at scale on Gasperini data
+bash run_gasp_glmeiv.sh
 # run glmeiv at scale on Xie data
 bash run_xie_glmeiv.sh
-# run thresholding method at scale on Xie data
+
+##########################
+# 3. Thresholding analysis
+##########################
+# Get the optimal threshold from the GLM-EIV results
+Rscript ../R_scripts/analysis/approx_bayes_bdy.R
+# run thresholding on gasperini data
+bash run_gasp_thresholding.sh
+# run thresholding on xie data
 bash run_xie_thresh.sh
+# run thresholding on positive control Gasperini pairs, setting thresh to 1,5,20.
+bash run_gasp_pc_thresholding.sh
 
 ################
-# 3. Simulations
+# 4. Simulations
 ################
+sim_dir=$LOCAL_GLMEIV_DATA_DIR"public/simulations/"
 Rscript ../R_scripts/simulations/create_simspec_objects.R
 $SIMULATR -f $sim_dir"spec_objects/sim_spec_0.rds" -r $sim_dir"results/raw_result_0.rds"
 $SIMULATR -f $sim_dir"spec_objects/sim_spec_1.rds" -r $sim_dir"results/raw_result_1.rds"
 $SIMULATR -f $sim_dir"spec_objects/sim_spec_2.rds" -r $sim_dir"results/raw_result_2.rds"
 
+##################################
+# 5. Gasperini resampling analysis
+##################################
+bash run_gasp_resampling.sh
+
 #############
-# 4. Plotting
+# 6. Plotting
 #############
+Rscript ../R_scripts/plotting/analysis_challenges.R
[…]

### 2022-01-04 bash_scripts/run_gasp_thresholding_pc.sh

@@ -1,0 +1,26 @@
+source set_gasp_args.sh
+
+#######################
+# Run Nextflow pipeline
+#######################
+if [ $trial = true ]
+then
+  gRNA_gene_pairs_pc=$processed_data_dir"gRNA_gene_pairs_sample_pc.rds"
+else
+  gRNA_gene_pairs_pc=$processed_data_dir"gRNA_gene_pairs_pc.rds"
+fi
+
+rm -f trace.txt
+nextflow run $thresholding_nf_pipeline --pairs $gRNA_gene_pairs_pc \
+--covariate_matrix $covariate_matrix \
+--pair_pod_size 50 \
+--result_dir $result_dir \
+--out_file_name "thresholding_result_pc.rds" \
+--gene_pod_size 20 \
+--gene_odm $gene_odm \
+--gene_metadata $gene_metadata \
+--m_offsets $m_offsets \
+--gRNA_odm $gRNA_odm \
+--gRNA_metadata $gRNA_metadata \
+--threshold "1,5,20" \
+-bg > $PWD/log -ansi-log false -with-trace -w $work_dir

### 2022-01-04 bash_scripts/set_gasp_args.sh

@@ -40,7 +40,7 @@ g_fam="poisson"
 # vi. pairs to analyze and pod sizes
 if [ $trial = true ]
 then
-  gRNA_gene_pairs=$processed_data_dir"gene_gRNA_pairs_problem.rds"
+  gRNA_gene_pairs=$processed_data_dir"gRNA_gene_pairs_sample.rds "
   gene_pod_size=10 #500
   gRNA_pod_size=10 #500
   pair_pod_size=50 #500

### 2022-01-04 nextflow_scripts/bin/collect_results.R

@@ -5,7 +5,9 @@ library(magrittr)
 args <- commandArgs(trailingOnly = TRUE)
 n_args <- length(args)
 pairs_fp <- args[1L]
-raw_result_fps <- args[seq(2L, n_args)]
+fp_name <- args[2L]
+raw_result_fps <- args[seq(3L, n_args)]
+
 # load data
 pairs_df <- readRDS(pairs_fp)
 combined_result <- do.call(rbind, lapply(raw_result_fps, function(fp) readRDS(fp)))
@@ -22,4 +24,4 @@ if (ncol(pairs_df) >= 3) {
   out <- combined_result
 }
 
-saveRDS(out, "result_thresholding.rds")
+saveRDS(out, fp_name)

### 2022-01-04 nextflow_scripts/bin/run_analysis.R

@@ -11,7 +11,7 @@ gene_metadata_fp <- args[3L]
 m_offsets_fp <- args[4L]
 gRNA_odm_fp <- args[5L]
 gRNA_metadata_fp <- args[6L]
-threshold <- as.numeric(args[7L])
+thresholds <- as.numeric(unlist(strsplit(args[7L], ",")))
 other_args <- args[seq(8L, n_args)]
 
 #########################################
@@ -44,19 +44,17 @@ for (i in seq(1L, n_pairs)) {
     g <- as.numeric(gRNA_odm[[gRNA,]])
   }
 
-  # loop through different thresholds, fitting model to each
-  phat <- as.integer(g >= threshold)
-  # run thresholding method
-  fit <- glmeiv::run_thresholding_method(phat = phat, m = m, m_fam = m_fam,
-                                         m_offset = m_offset, covariate_matrix = covariate_matrix,
-                                         n_examples_per_param = 5) %>% dplyr::mutate(gene_id = gene, gRNA_id = gRNA)
-  out_l[[i]] <- fit
+  coef_df <- lapply(X = thresholds, FUN = function(threshold) {
+    # perform threshold
+    phat <- as.integer(g >= threshold)
+    # run thresholding method
+    fit <- glmeiv::run_thresholding_method(phat = phat, m = m, m_fam = m_fam,
+                                           m_offset = m_offset, covariate_matrix = covariate_matrix,
+                                           n_examples_per_param = 5) %>% dplyr::mutate(threshold = threshold)
+    return(fit)
+  }) %>% do.call(rbind, .) %>% dplyr::mutate(gene_id = gene, gRNA_id = gRNA)
+  out_l[[i]] <- coef_df
 }
 
 out <- do.call(rbind, out_l) %>% dplyr::mutate_at(c("parameter", "target", "gene_id", "gRNA_id"), factor)
 saveRDS(object = out, file = "raw_result.rds")
-
-
-# leftover code
-# g_prime <- g[g >= 1]
-# curr_threshold <- mean(g_prime)

### 2022-01-04 nextflow_scripts/bin/collect_results.R
add lib paths to nextflow pipeline

@@ -1,3 +1,5 @@
+if (!("/jet/home/timbar/R/x86_64-redhat-linux-gnu-library/4.0" %in% .libPaths())) .libPaths("/jet/home/timbar/R/x86_64-redhat-linux-gnu-library/4.0")
+
 #######################
 # 1. Set fps, load data
 #######################

### 2022-01-04 nextflow_scripts/bin/run_precomp.R

@@ -1,3 +1,5 @@
+if (!("/jet/home/timbar/R/x86_64-redhat-linux-gnu-library/4.0" %in% .libPaths())) .libPaths("/jet/home/timbar/R/x86_64-redhat-linux-gnu-library/4.0")
+
 ########################################
 # 1. Load packages and command-line args
 ########################################

### 2023-07-27 presentation/glmeiv_talk.tex
start text outline
update sim spec object for experiment 0 to be compatible with new simulatr

@@
 
-\begin{frame}
-\frametitle{(1) The perturbation is unobserved, and (2) technical factors are present.}
-\begin{figure}
-	\centering
-	\includegraphics[width=1\linewidth]{../figures/fig1/dag}
-	\label{dag}
-\end{figure}
-\end{frame}
+% \begin{frame}
+% \frametitle{(1) The perturbation is unobserved, and (2) technical factors are present.}
+% \begin{figure}
+%	\centering
+%	\includegraphics[width=1\linewidth]{../figures/fig1/dag}
+%	\label{dag}
+%\end{figure}
+% \end{frame}
 
-\begin{frame}
-\frametitle{(3) Unperturbed cells exhibit background gRNA reads.}
-\begin{figure}
-	\centering
-	\includegraphics[width=0.75\linewidth]{../figures/fig1/density_plot_annotated.png}
-\end{figure}
-\end{frame}
+%\begin{frame}
+%\frametitle{(3) Unperturbed cells exhibit background gRNA reads.}
+%\begin{figure}
+%	\centering
+%	\includegraphics[width=0.75\linewidth]{../figures/fig1/density_plot_annotated.png}
+%\end{figure}
+% \end{frame}
 
-\begin{frame}
-\frametitle{(4) The count data are highly discrete.}
-\begin{figure}
-	\centering
-	\includegraphics[width=0.75\linewidth]{../figures/fig1/mRNA_count_hist_annotated.png}
-\end{figure}
-\end{frame}
+% \begin{frame}
+%  \frametitle{(4) The count data are highly discrete.}
+% \begin{figure}
+%	\centering
+%	\includegraphics[width=0.75\linewidth]{../figures/fig1/mRNA_count_hist_annotated.png}
+% \end{figure}
+% \end{frame}
 
@@
 
-\begin{frame}
-\frametitle{Problem 1: There is no clear location in the data at which to draw the threshold.}
-\begin{figure}
-	\centering
-	\includegraphics[width=1\linewidth]{../figures/fig2/histograms}
-\end{figure}
-
-\end{frame}
+%\begin{frame}
+%\frametitle{Problem 1: There is no clear location in the data at which to draw the threshold.}
+%\begin{figure}
+%	\centering
+%	\includegraphics[width=1\linewidth]{../figures/fig2/histograms}
+%\end{figure}
+%\end{frame}
 
@@
 
-\begin{frame}
-\frametitle{The thresholding method shows clear attenuation bias.}
-\begin{itemize}
-\item We set $\beta_m$ to 1.
-\end{itemize}
+%\begin{frame}
+%\frametitle{The thresholding method shows clear attenuation bias.}
+%\begin{itemize}
+%\item We set $\beta_m$ to 1.
+%\end{itemize}
 
-\begin{figure}
-	\centering
-	\includegraphics[width=0.75\linewidth]{../figures/fig2/att_bias}
-\end{figure}
-
-\end{frame}
+%\begin{figure}
+%	\centering
+%	\includegraphics[width=0.75\linewidth]{../figures/fig2/att_bias}
+%\end{figure}
+%\end{frame}
 
@@
 
-\begin{frame}
-\begin{figure}
-	\centering
-	\includegraphics[width=1\linewidth]{../figures/fig3/simulation_warmup_arm_g}
-\end{figure}
-\end{frame}
+%\begin{frame}
+%\begin{figure}
+%	\centering
+%	\includegraphics[width=1\linewidth]{../figures/fig3/simulation_warmup_arm_g}
+%\end{figure}
+%\end{frame}
 
@@
 
-\begin{frame}
-\frametitle{Estimates and CIs produced by GLM-EIV and the thresholding method were similar.}
-\begin{figure}
-	\centering
-	\includegraphics[width=1\linewidth]{../figures/fig5/comparison}
-\end{figure}
-\begin{center}
-Candidate \textit{cis} pairs in Xie et al.\ data.
-\end{center}
-\end{frame}
+%\begin{frame}
+%\frametitle{Estimates and CIs produced by GLM-EIV and the thresholding method were similar.}
+%\begin{figure}
+%	\centering
+%	\includegraphics[width=1\linewidth]{../figures/fig5/comparison}
+%\end{figure}
+%\begin{center}
+%Candidate \textit{cis} pairs in Xie et al.\ data.
+%\end{center}
+%\end{frame}
 
@@
 \begin{itemize}
-\item Xuran Wang helped with the Xie data preprocessing.
+\item Thanks to Xuran Wang for helping with the Xie data preprocessing.
 \item All analyses were run on Pittsburgh Supercomputer.

### 2023-07-27 presentation/glmeiv_talk_updated.tex
update sim spec object for experiment 0 to be compatible with new simulatr

--- 2023-07-27 presentation/glmeiv_talk.tex
@@
 %Information to be included in the title page:
-\title{A generalized errors-in-variables model, with application to single-cell CRISPR screens}
-\author{Tim Barry$^1$, Eugene Katsevich$^2$, Kathyrn Roeder$^1$}
+\title{Exponential family measurement error models for single-cell CRISPR screens}
+\author{Timothy Barry$^1$, Eugene Katsevich$^2$, Kathyrn Roeder$^1$}
 \institute{$^1$CMU Statistics and Data Science\\ $^2$Wharton Statistics and Data Science}
-\date{September 2021}
 \DeclareMathOperator*{\argmin}{arg\,min}
@@
 
-\bibliography{/Users/timbarry/optionFiles/glmeiv.bib}
+\bibliography{glmeiv.bib}
 \bibliographystyle{apalike}

### 2023-07-27 R_scripts/plotting/thresholding_empirical.R

@@ -23,19 +23,23 @@ thresh_res_pc_wide <- thresh_res_pc %>% dplyr::select(target, value, threshold, 
 
 to_plot <- thresh_res_pc_wide %>% dplyr::filter(target == "estimate") %>% dplyr::select(threshold_1, threshold_5)
 p1 <- ggplot(data = to_plot, mapping = aes(x = threshold_5, y = threshold_1)) + geom_point(alpha = 0.7, col = my_cols[2]) +
-  xlab("Thresh = 5 (est)") + ylab("Thresh = 1 (est)") + geom_abline(slope = 1, intercept = 0) + theme_bw(base_size = 10) + my_theme
+  xlab("Fold change (thresh = 5)") + ylab("Fold change (thresh = 1)") + 
+  geom_abline(slope = 1, intercept = 0) + theme_bw(base_size = 10) + my_theme +
+  theme(plot.margin = margin(0.1, 0.5, 0.1, 0.5, "cm"))
 
 # threshold = 1 ests are bigger than (i.e., closer to one than) threshold = 5; this is attenuation bias due to the fact that excessively many unperturbed cells have been classified as perturbed.
 
 # a) est. vs. est.
 to_plot <- thresh_res_pc_wide %>% dplyr::filter(target == "estimate") %>% dplyr::select(threshold_5, threshold_20)
 p2 <- ggplot(data = to_plot, mapping = aes(x = threshold_5, y = threshold_20)) + geom_point(alpha = 0.9, col = my_cols[2]) +
-  xlab("Thresh = 5 (est)") + ylab("Thresh = 20 (est)") + geom_abline(slope = 1, intercept = 0, lwd = 0.75) + theme_bw(base_size = 10) + my_theme
+  xlab("Fold change (thresh = 5)") + ylab("Fold change (thresh = 20)") + geom_abline(slope = 1, intercept = 0, lwd = 0.75) +
+  theme_bw(base_size = 10) + my_theme + theme(plot.margin = margin(0.1, 0.5, 0.1, 0.5, "cm"))
 
 # b) p-value vs. p-value
 to_plot <- thresh_res_pc_wide %>% dplyr::filter(target == "p_value") %>% dplyr::select(threshold_5, threshold_20)
[…]

### 2023-07-27 R_scripts/simulations/create_simspec_objects.R

@@ -15,24 +15,21 @@ save_obj <- function(obj, file_path, overwrite) {
 }
 sim_dir <- paste0(.get_config_path("LOCAL_GLMEIV_DATA_DIR"), "public/simulations/spec_objects")
 
-one_rep_times <- list(generate_data_function = NA_integer_,
-                      thresholding = NA_integer_,
-                      glmeiv_fast = NA_integer_,
-                      glmeiv_slow = NA_integer_)
-
-##############################
-# Experiment 0
+######################################
+# Experiment 0 (NOT REPORTED IN PAPER)
 # Vary g_pert
 # Fix distribution to Poisson
 # All three methods
 # One covariate (batch)
-##############################
+######################################
 set.seed(4)
+m_perturbation <- log(0.25)
 theta <- 20
-n <- 150000
+n <- 15000 # 150000
 g_perturbation_grid <- log(seq(1, 4, 0.5))
 param_grid <- expand.grid(g_perturbation = g_perturbation_grid)
 param_grid$grid_id <- seq(1, nrow(param_grid))
+param_grid$ground_truth <- m_perturbation
 
 fixed_params <- list(
   m_fam = poisson() %>% augment_family_object(),
@@ -41,7 +38,7 @@ fixed_params <- list(
   n = n,
   B = 500,
   m_intercept = log(0.01),
-  m_perturbation = log(0.25),
+  m_perturbation = m_perturbation,
   g_intercept = log(0.005),
   covariate_matrix = data.frame(batch = rbinom(n = n, size = 1, prob = 0.5)),
   m_covariate_coefs = log(0.9),
@@ -49,7 +46,7 @@ fixed_params <- list(
   n_processors = 20,
   alpha = 0.95,
   n_em_rep = 15,
-  save_membership_probs_mult = 500,
+  save_membership_probs_mult = 1000L,
   pi = 0.02,
   m_offset = log(rpois(n = n, lambda = 10000)),
   g_offset = log(rpois(n = n, lambda = 5000)),
@@ -60,15 +57,19 @@ fixed_params <- list(
   g_intercept_guess_range = log(c(1e-4, 1e-1)),
   m_covariate_coefs_guess_range = log(c(0.25, 2)),
   g_covariate_coefs_guess_range = log(c(0.25, 2)),
-  run_unknown_theta_precomputation = FALSE)
+  run_unknown_theta_precomputation = FALSE,
+  exponentiate_coefs = FALSE,
[…]
